### 2011-12-12 AMH3 - Interest Rate Models/AMH3 Lecture Notes.tex
Initial Commit

% Created by Andrew Tulloch

%!TEX TS-program = xelatex
%!TEX encoding = UTF-8 Unicode

\documentclass[10pt, oneside, reqno]{amsbook}
% \chapterstyle{bianchi}
% 
% \makeatletter
% \newcommand\thickhrulefill{\leavevmode \leaders \hrule height 1ex \hfill \kern \z@}
% \setlength\midchapskip{10pt}
% \makechapterstyle{VZ14}{
% \renewcommand\chapternamenum{}
% \renewcommand\printchaptername{}
% \renewcommand\chapnamefont{\Large\scshape}
% \renewcommand\printchapternum{%
% \chapnamefont\null\thickhrulefill\quad
% \@chapapp\space\thechapter\quad\thickhrulefill}
% \renewcommand\printchapternonum{%
% \par\thickhrulefill\par\vskip\midchapskip
% \hrule\vskip\midchapskip
% }
% \renewcommand\chaptitlefont{\Huge\scshape\centering}
% \renewcommand\afterchapternum{%
% \par\nobreak\vskip\midchapskip\hrule\vskip\midchapskip}
% \renewcommand\afterchaptertitle{%
% \par\vskip\midchapskip\hrule\nobreak\vskip\afterchapskip}
% }
% \makeatother
% \chapterstyle{VZ14}

% \OnehalfSpacing

% \makechapterstyle{AlexanderGrebenkov}{%
% \renewcommand{\chapterheadstart}{\vspace*{\beforechapskip}\hrule\medskip}
% \renewcommand{\chapnamefont}{\normalfont\large\scshape}
% \renewcommand{\chapnumfont}{\normalfont\large\scshape}
% \renewcommand{\chaptitlefont}{\normalfont\large\scshape}
% \renewcommand{\printchaptername}{\S}
% \renewcommand{\chapternamenum}{ }
% \renewcommand{\printchapternum}{\chapnumfont \thechapter}
% \renewcommand{\afterchapternum}{. }
% \renewcommand{\afterchaptertitle}{\par\nobreak\medskip\hrule\vskip
% \afterchapskip}
% }
% \chapterstyle{AlexanderGrebenkov}
\usepackage{amsthm, amsmath, amssymb}
\usepackage{setspace, graphicx, enumerate}
\onehalfspacing    
% \usepackage[landscape, scale=0.90]{geometry}       
% \usepackage{fullpage}      
\usepackage{fontspec,xltxtra,xunicode}
\defaultfontfeatures{Mapping=tex-text}

%\setromanfont[Mapping=tex-text,Contextuals= 
%{NoWordInitial,NoWordFinal,NoLineInitial,NoLineFinal}]{Hoefler Text}
%\setsansfont[Scale=MatchLowercase,Mapping=tex-text]{Hoefler Text}
%\setmonofont[Scale=MatchLowercase]{Andale Mono}

	% AMS Theorems
	\theoremstyle{plain}% default 
	\newtheorem{thm}{Theorem}[chapter] 
	\newtheorem{lem}[thm]{Lemma} 
	\newtheorem{prop}[thm]{Proposition} 
	\newtheorem{exer}[thm]{Exercise} 
	\newtheorem*{cor}{Corollary}

\newcommand{\res}[2]{\text{Res}(#1,#2)}
	\theoremstyle{definition} 
		\newtheorem{defn}[thm]{Definition}
		\newtheorem{conj}[thm]{Conjecture}
		\newtheorem{exmp}[thm]{Example}

	\theoremstyle{remark} 
		\newtheorem*{rem}{Remark} 
		\newtheorem*{note}{Note} 
		\newtheorem{case}{Case}

		\newcommand{\expc}[1]{\mathbb{E}\left[#1\right]}
		\newcommand{\var}{\text{Var}}
		\newcommand{\cov}[1]{\text{Cov}\left(#1\right)}
		\newcommand{\prob}[1]{\mathbb{P}(#1)}
		\newcommand{\given}{ \, | \,}
		\newcommand{\us}{0 \leq u \leq s}
		\newcommand{\ts}[1]{\{ #1 \}}

\renewcommand{\phi}{\varphi}
\newcommand{\sigf}{\mathcal{F}}

\newcommand{\dzz}{\, dz}
\newcommand{\bigo}[1]{\mathcal{O}(#1)}

\newcommand{\al}{\alpha}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\N}{\mathbb{N}}

\newcommand{\I}{\mathbb{I}}

\renewcommand{\P}{\mathbb{P}}

\newcommand{\F}{\mathbb{F}}
\newcommand{\Ga}{\mathbb{G}}

\newcommand{\aut}[1]{\text{Aut}{(#1)}}

\newcommand{\gener}[1]{\langle #1 \rangle}
\newcommand{\charr}[1]{\text{char}(#1)}
\newcommand{\nth}{n\textsuperscript{th}}

% \newcommand{\limsup}{\text{limsup}}

\newcommand{\tworow}[2]{\genfrac{}{}{0pt}{}{#1}{#2}}
\newcommand{\xdeg}[2]{[#1 : #2]}
\newcommand{\gal}[2]{\text{Gal}(#1/#2)}
\newcommand{\minpoly}[2]{m_{#1, #2}(x)}

\newcommand{\mapping}[5]{\begin{align*}
	#1 : \quad     #2 &\rightarrow #3 \\
			#4  &\mapsto #5
\end{align*}	
}

\def\cip{\,{\buildrel p \over \rightarrow}\,} 
\def\cid{\,{\buildrel d \over \rightarrow}\,} 
\def\cas{\,{\buildrel a.s. \over \rightarrow}\,}

\def\clp{\,{\buildrel L^p \over \rightarrow}\,}

\def\eqd{\,{\buildrel d \over =}\,} 
\def\eqas{\,{\buildrel a.s. \over =}\,}
\newcommand{\sigg}{\mathcal{G}}		
\newcommand{\indic}[1]{\mathbf{1}_{\{ #1 \}} }
\newcommand{\ito}{It\^o\ }		
\newcommand{\itos}{It\^o's\ }

\newcommand{\doleans}[1]{\mathcal E_t \left(\int_0^\cdot #1 \right)}

\numberwithin{equation}{chapter}

\usepackage{hyperref}

\title{AMH3a - Interest Rate Modelling}								% Document Title
\author{Andrew Tulloch}
%\date{}                                           % Activate to display a given date or no date

\begin{document}
% \maketitle
% \chapter*{AMH3 - Interest Rate Modelling} % (fold)
% \label{cha:}
% \clearpage
% % chapter  (end)

\maketitle

\tableofcontents

\chapter{Preliminaries} % (fold)
\label{cha:preliminaries}

% chapter preliminaries (end)
\section{Introduction to Interest Rate Modelling} % (fold)
\label{sec:introduction_to_interest_rate_modelling}

% section introduction_to_interest_rate_modelling (end)
There is a one-to-one correspondence between the class $Q$ of all probability measures equivalent to $\P$ and the class $\Lambda$ of all $\F$-adapted (or $\F$-predictable) process $\lambda_t$ satisfying \[
	\P\left(\int_0^{T^\star} |\lambda_u|^2 \, du < \infty \right) = 1
\]
and \[
	\E_\P\left( \mathcal{E}_{T^\star} \left(\int_0^\cdot \lambda_u \, dW_u\right) \right) = 1
\] 
Thus our correspondence is \[Q \ni \P^\lambda \iff \lambda \in \Lambda.\]
Consequently, \begin{enumerate}[(i)]
	\item $\frac{d\Q}{d\P} = \eta_{T^\star}$
	\item \begin{align*}
		\frac{d\Q}{d\P} \given \sigf_t 	&= \eta^\Q_t \\
										&= \E_\P\left(\eta_{T^\star} \given \sigf_t \right) \\
										&= \doleans{\lambda_u \, dW_u}
	\end{align*}
\end{enumerate}

\begin{thm}[Abstract Bayes formula]
	Let $\Q \sim \P$ with $\frac{d \Q}{d \P} = \eta$.  Suppose that $\sigg \subset \sigf$.  We then have \[
		\E_\Q(X \given \sigg) = \frac{\E_\P(\eta X \given \sigg)}{\E_\P(\eta \given \sigg)}.
	\]  Note that is $\sigg = \{ \emptyset, \Omega \}$ then the formula reduces to \[
		\E_\Q(X) = \E_\P(\eta X).
	\]

	If $Q \sim \P$ with $\frac{d\Q}{d\P} |_{\sigf_t} = \eta_t$, for all $t \in [0, T^\star]$, then \[
		\E_\Q(X \given \sigf_t) = \frac{\E_\P(\eta_{T^\star} X \given \sigf_t)}{\E_\P(\eta_{T^\star} \given \sigf_t)}.
	\]

	Hence if $X$ is $\sigf_t$ measurable for some $T \in [0, T^\star]$ then \[
		\E_\Q(X \given \sigf_t) = \frac{\E_\P(\eta_T X \given \sigf_t)}{\eta_t} = \E_\P(\eta_t^{-1} \eta_T X \given \sigf_t)
	\]
\end{thm}

\begin{exmp}
	If $\eta_t = \doleans{\lambda_u \, dW_u}$, then \[
		\E_\Q(X \given \sigf_t) = \E_\P(e^{\int_t^T \lambda_u \, dW_u - \frac{1}{2} \int_0^t |\lambda_u|^2 \, du} X \given \sigf_t)
	\] 
\end{exmp}

\begin{lem}
	\label{lem:martingale_equivalence_measures}
	A $\F$-adapted and $\Q$-integrable process $M$ is a $(\Q, \F)$-martingale if and only if the product $M \eta$ is a $(\P, \F)$-martingale.
\end{lem}

\begin{proof}
	$\E_\Q(M_t \given \sigf_s) = M_s$, $s \leq t$, so \[
		M_s = \E_\Q(M_t \given \sigf_s) = \frac{\E_\P(\eta_t M_t \given \sigf_s)}{\eta_s}
	\]
\end{proof}

\begin{lem}
	If $X$ and $Y$ are two processes of the form \begin{align*}
		dX_t &= \alpha_t \, dt + \beta_t \, dW_t \\
		dY_t &= \tilde \alpha_t \, dt + \tilde \beta_t \, dW_t
	\end{align*} then the product satisfies the \ito product formula \begin{align*}
		d(X_t Y_t) = X_t \, dY_t + Y_t \, dX_t + d \langle X, Y \rangle_t
	\end{align*}	
\end{lem}

If $X$ is of the form $dX_t = \alpha_t \, dt + \beta_t \, dW_t$ and $f$ is of class $C^2(\R)$, then the continuos martingale part of $Y_t = f(X_t)$ is given as \[
	\int_0^t f'(X_u) \beta_u \, dW_u
\]

\begin{prop}
	\label{prop:proposition_1.1}
	% TODO MUST COMPLETE - PROPOSITION 1.1 in course notes.
\end{prop}

\begin{proof}[Proof of Proposition 1.1]
	Let $\P^\lambda$ be equivalent to $\P$, so that \[
		d\eta_t = \eta_t \lambda_t \, dW_t
	\]  and \[
		\frac{d\P^\lambda}{d \P} = \eta_t
	\] on $(\Omega, \sigf_t)$, $t \in [0, T^\star]$.

	Define $B(t, T)$ as follows, for all $t \in [0, T]$, \begin{align*}
		B(t, T) &= B_t \E_{\P^\lambda} \left( \frac{1}{B_T} \given \sigf_t \right) \\
		&= \E_{\P^\lambda} \left(e^{-\int_t^T r_u \, du} \given \sigf_t \right)
	\end{align*}

	For i), we simply apply Girsanov's theorem, replacing $dW_t$ by $dW_t = dW^\lambda_t - \lambda_t \, dt$ in the dynamics of $r$ under $\P$.

	For ii), we first recall that $Z(t, T) = \frac{B(t, T)}{B_t}$ is given by \[
		Z(t, T) = \E_{\P^\lambda} \left( \frac{1}{B_T} \given \sigf_t \right)
	\] is a $(\P^\lambda, \F)$-martingale.

	Note that $\F^{\lambda} \neq \F$ in general.  From Lemma \ref{lem:martingale_equivalence_measures}, we know that $\eta_t Z(t, T)$ is a $(\P, \F)$-martingale.  Thus applying the predictable representation property, there exists an $\F$-adapted process $\gamma_t$ such that \[
		M_t \equiv \eta_t Z(t, T) = Z(0, T) + \int_0^t \gamma_u \, dW_u
	\] for all $t \in [0, T]$.  Consequently, $dM_t = \gamma_t \, dW_t$ and hence \[
		dZ(t, T) = d(\eta^{-1}_t M_t) = M_t \, d\eta_t^{-1} + \eta_t^{-1}\, dM_t + d \langle \eta^{-1}, M \rangle_t
	\]  where \begin{align*}
		d \eta_t^{-1} = - \eta_t^{-1} \lambda_t \, dW_t^\lambda.
	\end{align*}  We obtain \begin{align*}
		dZ(t, T) &= \eta_t Z(t, T) \left( - \eta_t^{-1} \lambda_t \, dW_t^\lambda \right) + \eta_t^{-1} \gamma_t \left(dW_t^\lambda + \lambda_t \, dt \right) + \left(- \eta_t^{-1} \lambda_t \gamma_t \right) \, dt \\
		&= \eta_t^{-1} \left(\gamma_t - M_t \lambda_t \right) \, dW_t^\lambda
	\end{align*} so that \[
		dZ(t, T) = \tilde b^\lambda(t, T) \, dW_t^\lambda
	\]  Since $B(t, T) = B_t Z(t, T)$, using again the \ito formula we have \begin{align*}
		dB(t, T) 	&= B_t \, dZ(t, T) + Z(t, T) \, dB_t  \\
					&= \frac{B(t, T)}{B_t} r_t B_t \, dt + B_t \tilde b^\lambda(t, T) \, dW_t^\lambda \\
					&= r_t B(t, T) \, dt + B(t, T) \underbrace{\frac{B_t \tilde b^\lambda(t, T)}{B(t, T)}}_{b^\lambda(t, T)} \, dW^\lambda_t.
	\end{align*}   We conclude that for all $T \in [0, T^\star]$, there exists an $\F$-adapted process $b^\lambda(t, T)$, $t \in [0, T]$ called the volatility of the bond
, such that \[
	dB(t, T) = B(t, T)(r_t \, dt + b^\lambda(t, T) \, dW^\lambda_t).  
\]	  In fact, it does not depend on the choice of $\lambda$.  For simplicity, we can write $b(t, T) \equiv b^\lambda(t, T)$.

The final formula is a special case of the well known result:
\begin{align*}
	dX_t 	&= X_t (\alpha_t \, dt + \beta_t \, dW_t) \\
				\Updownarrow  \\
		X_t	&= X_0 e^{\int_0^t \alpha_u \, du} \doleans {\beta_u \, dW_u }  \\
			&= X_0 e^{\int_0^t \alpha_u \, du} e^{\int_0^t \beta_u \, dW_u - \frac{1}{2} \int_0^t |\beta_u|^2 \, du}
		\end{align*}

	This completes our proof of Proposition 1.1, under the assumption that $\frac{1}{B_T}$ is $\P^\lambda$-integrable.
	% We have \[
	% 	dM_t = \eta_t \lambda_t dW_t
	% \] and \[
	% 	dM_t^{-1} = - \eta_t^{-1} \lambda_t dW^\lambda_t.
	% \]  This is true as \[
	% 	\frac{d\P}{d\Q}
	% \]
\end{proof}

There are still several issues given this pricing formula. 
\begin{enumerate}[(i)]
	\item How to compute $b(t, T)$ explicitly in terms of $\mu$ and $\sigma$ under the assumptions that \[
		dr_t = \mu(r_t, t) \, dt + \sigma(r_t, t) \, dW_t
	\] and $\lambda_t = \lambda(r_t, t)$ is the risk premium.  
	\item How can we calibrate our short-term rate model, meaning that \[
		\E_{\P^\lambda}\left( \frac{1}{B_T} \right) = B(0, T) = P(0, T).
	\]
\end{enumerate}

The issue of pricing bonds is related to solving a backward stochastic differential equation (BSDE).  The general form is \[
	X_t = X_0 + \int_0^t \mu (X_u, u) \, du + \int_0^t \xi_u \, dW_u \tag{$\star$}
\]   where $\mu : \R \times \R^+ \rightarrow \R$ is some function and $\xi$ is some $\F$-adapted process.  We also fix $T > 0$ and postulate that $X_T$ is a \textbf{known } $\sigf_T$-measurable random variable.

\begin{defn}
	We say that $(X, \xi)$ solves the BSDE with terminal condition with terminal condition $Y$ ($\sigf_T$-measurable) if: 
	\begin{enumerate}[(i)]
		\item $(X, \xi)$ satisfies ($\star$), 
		\item $X_T = Y$.  
	\end{enumerate}

	This can be extended to cases where $\mu : \R \times \R^+ \times \Omega \rightarrow \R$ is $\F$-adapted.
\end{defn}

\chapter{Markovian Models of the Short Rate} % (fold)
\label{cha:markovian_models_of_the_short_rate}

	Let $\P^\star$ be a martingale measure in the sense that \[
		B(t, T) = \E_{\P^\star} \left( e^{-\int_t^T r_u \, du} \given \sigf_t \right).
	\]  In particular, \[
		B(0, T) = \E_{\P^\star} \left( e^{-\int_0^T r_u \, du}  \right).  
	\]  We postulate that \begin{equation}
		dr_t = \mu(r_t, t) \, dt + \sigma(r_t, t) \, dW^\star_t,
		\label{eq:generic_short_rate}
	\end{equation} where $W^\star$ is a Brownian motion under $\P^\star$. The filtration $\F$ is any filtration such that $W^\star$ is a BM with respect to $\F$ .  We assume that \eqref{eq:generic_short_rate} has a unique (strong) solution.

	Then it known that $r_t$ has the Markov property with respect to $\F$, meaning that for any bounded continuous function $h : \R \rightarrow \R$, \[
 	\E_{\P^\star} \left(h(r_t) \given \sigf_s \right) = \E_{\P^\star}\left(h(r_t) \given r_s \right)
 \] for all $s \leq t$.

Hence \[
	\E_{\P^\star} \left(e^{-\int_t^T r_u \, du} \given \sigf_t \right) = v(r_t, t, T) = \tilde v(r_t, t) 
\] suppressing the dependence on $T$.

Goals:
\begin{enumerate}[(i)]
	\item Compute explicitly $v(r_t, t, T)$ for some classical models \begin{enumerate}[(a)]
		\item Merton's model 
		\item Vasicek's model
		\item CIR model (Bessel process)
	\end{enumerate}
	using either the probabilistic approach (martingale measure) or the  analytic approach (PDEs).
	\item Represent the price of the bond as follows \[
		B(t, T) = \exp \left( m(t, T) - n(t, T) r_t \right)
	\]

	For a fixed maturity $T$, \[
		m(\cdot, T), n(\cdot, T) : [0, T] \rightarrow \R
	\] can also be computed using the second method by separating variables in the PDE.  Note that $m(T, T), n(T, T) = 0$.
	\item Compute explicitly the volatility $b(t, T)$ of the bond by applying the \ito formula to the function $v(r_t, t, T)$.
	\item Extend the model to the time-inhomogenous case in order to ensure that $B(0, T) = P(0, T)$ for all $T \in [0, T^\star]$.
\end{enumerate}

\section{Merton's model} % (fold)
\label{sub:merton_s_model}
	Assure \[
		r_t = r_0 + at + \sigma W_t^\star 
	\] where $W^\star = W^\lambda$ for some $\lambda$.  Hence
	\begin{equation}
		dr_t = a \, dt + \sigma \, dW^\star_t, \quad r_0 > 0.
	\end{equation} 
	\begin{note}
		The generator of the time homogenous Markov diffusion can be represented as \[
			A_r = a \frac{\partial }{\partial r} + \frac{1}{2} r^2 \frac{\partial^2}{\partial r^2}.  
		\]
	\end{note}

	\begin{prop}
		The price $B(t, T)$ is given by \begin{equation}
			\label{eq:merton_short_rate_price}
			B(t, T) = e^{-r_t(T-t) - \frac{1}{2}a(T-t)^2 + \frac{1}{6} \sigma^2(T-t)^3}.  
		\end{equation}  Hence \[
			dB(t, T) = B(t, T)\left(r_t \, dt - \sigma(T-t) dW^\star_t \right).
		\]  Thus we have the volatility of the bond $b(t, T) = -\sigma(T-t)$.
	\end{prop}

	\begin{proof}
		It is enough to calculate $B(0, T)$ and then establish the general formula for $B(t, T)$ using the property that $r_t$ is a time-homogenous Markov process, thus \[
			B(0, T) = v(r_0, T) \Rightarrow B(t, T) =v(r_t, T-t)
		\]
		Computation of $B(0, T)$ is done as follows:\[
			B(0, T) = \E_{\P^\star} \left(e^{-\int_0^T r_u \, du} \right) = \E_{\P^\star} \left( e^{-\xi_T} \right)
		\] where the distribution of $\xi_T$ can be found explicitly.  We argue that \[
			\xi_T \sim N\left(r_0 T + \frac{1}{2} a T^2, \frac{1}{3} \sigma^2 T^3 \right)
		\]  We have \begin{align*}
			\xi_T 	&= \int_0^T r_u \, du \\
					&= \int_0^T \left( r_0 + au + \sigma W^\star_u \right) \, du \\
					&= \int_0^T (r_u + au)  \, du + \sigma \int_0^T W^\star_u \, du
		\end{align*}
		Th rest proceeds quite simply.

		We then derive the dynamics of $B(t, T)$.  By the \ito formula, we have that since $B(t, T) = v(r_t, t, T)$, we must have \[
			dB(t, T) = r_t B(t, T) \, dt + b(t, T) B(t, T) \, dW^\star_t.  
		\] Note that the martingale component comes from \[
			\frac{\partial v}{\partial r} dr_t 
		\] and \[
			\frac{\partial}{\partial r}v(r_t, t, T) = -(T- t) v(r_t, t, T)
		\] so that \begin{align*}
			\frac{\partial}{\partial r}v(r_t, t, T) \, dr_t &= -(T-t) v(r_t t, T) (a \, dt + \sigma \, dW^\star_t) \\
			&\sim - \sigma(T- t)B(t, T) \, dW^\star_t
		\end{align*}  We then obtain the equality $B(t, T) = -\sigma(T-t)$.  In particular, $B(t, T) = 0$.  
	\end{proof}

	\begin{exer}
		Apply the PDE approach to obtain \eqref{eq:merton_short_rate_price}.
	\end{exer}

\section{Vasicek's Model} % (fold)
\label{sub:vasicek_s_model}
Consider the dynamics \begin{equation}
	dr_t = (a - br_t) \, dt + \sigma \, dW^\star_t.
	\label{eq:vasicek_short_rate_dynamics}
\end{equation}

\begin{lem}
	The unique solution to Vasicek's equation is \begin{equation}
		r_t = r_0 e^{-bt} + \frac{a}{b}\left(1-e^{-bt} \right) + \sigma \int_0^t e^{-b(t-u)} \, dW^\star_u.
	\end{equation}
\end{lem}

\begin{prop}
	The bond price in the Vasicek model is given by \begin{align*}
		B(t, T) &= \exp(m(t, T) - n(t, T) r_t) \\
		n(t, T) &= \frac{1}{b}\left(1- e^{-b(T-t)} \right) \\
	\end{align*} and $m(t, T)$ is also known explicitly.

	The volatility of the bond satisfies \[
		b(t, T) = -\sigma n(t, T) = -\frac{\sigma}{b}\left(1 -e^{-b(T-t)} \right)
	\] and \[
		dB(t, T) = B(t, T) \left(r_t \, dt - \sigma n(t, T) \, dW^\star_t \right).
	\]
\end{prop}

\begin{thm}[Stochastic Fubini's theorem]
	In the computation above, we obtain the following double integral \[
		\int_0^T \int_0^t e^{-b(t-u)} \, dW^\star_u \, dt = \frac{1}{b} \int_0^T \left(1-e^{-b(T-u)} \right) \, dW^\star_u.
	\]   To obtain this result, we must use the stochastic Fubini theorem \[
		\int_0^T \int_0^t f(t, u) \, dW^\star_u \, dt = \int_0^T \int_u^T f(t, u) \, dt \, dW^\star_u
	\] where $f$ is a continuous function.

\end{thm}

\subsection{PDE Approach to Vasicek's model} % (fold)
\label{sub:pde_approach_to_vasicek_s_model}
We can either use some known results or provide some simple arguments.

We start by postulating that $B(t, T) = v(r_t, t, T)$ where $v \in C^{2,1}(\R \times [0, T\star], \R)$.  On the other hand, we may apply the \ito formula and obtain \begin{align*}
	dv(r_t, t, T) = \left( \frac{\partial r}{\partial t} + \mu(r_t, t) \frac{\partial v}{\partial r} + \frac{1}{2} \sigma^2(r_t, t) \frac{\partial^2 v}{\partial r^2} \right) \, dt + \sigma(r_t, t) \frac{\partial v}{\partial r} dW^\star_t.
\end{align*}

 On the other hand, from Proposition \ref{prop:proposition_1.1} we have \[
 	dB(t, T) = dv(r_t, t, T) = r_t v(r_t, t, T) \, dt + b(t, T)v(r_t, t, T) \, dW^\star_t.
 \]  This means that \begin{align*}
 	\underbrace{\left( \frac{\partial r}{\partial t} + \mu \frac{\partial v}{\partial r} + \frac{1}{2} \sigma^2 \frac{\partial^2 v}{\partial r^2} - r_t v \right) \, dt}_{A_t} &= \underbrace{\left( b(t, T) v - \sigma \frac{\partial v}{\partial r} \right) \, dW^\star_t}_{M_t}.
 \end{align*}

\begin{lem}
	If $(M_t)_{t \in [0, T^\star]}$ is a continuous local martingale and a process of finite variation then $M_t = M_0$ for $t \in [0, T^\star]$.
\end{lem}

Since $r_t$ is a Gaussian process, we note that the unknown function should necessarily satisfy the following pricing PDE for $v = v(r_t, t, T)$, \[
	\begin{cases}
		 \frac{\partial r}{\partial t} + \mu \frac{\partial v}{\partial r} + \frac{1}{2} \sigma^2 \frac{\partial^2 v}{\partial r^2} - r_t v = 0 \\
		v(r_t, T, T) = h(r_t).
	\end{cases}
\] For the bond maturing at $T$, we set $h(r) = 1$.

To solve this PDE in the Vasicek case, we postulate that \[
	v(r_t, t, T) = e^{m(t, T) - n(t, T) r_t}
\] and derive a system of two ODEs satisfied by the function $m$ and $n$. 
% subsection pde_approach_to_vasicek_s_model (end)
% subsection vasicek_s_model (end)
	% 
	% Recall that \begin{align*}
	% 	B(t, T) = \E_{\P^\star} \left( \frac{B_t}{B_T} \given \sigf_t \right) = f(r_t, t, T) \\
	% 	dB_t = r_t B_t \, dt, \quad B_0 = 1.
	% \end{align*}  
	% 
	% Computations of $f$ will be mainly based on simple properties of the Gaussian distribution. 
	% 
	% \begin{lem}
	% 	Assume that $\xi \sim N(\mu, \sigma^2)$.  Then $\zeta = e^{\xi}$ has the expected value \[
	% 		\E(\zeta) = e^{\mu + \frac{1}{2}\sigma^2}, \quad \var(\zeta) = e^{2 \mu + \sigma^2}(e^{\sigma^2} - 1).
	% 	\]
	% \end{lem}
	% 
	% \begin{proof}
	% 	We have \begin{align*}
	% 		\E(\zeta) = \E(e^{\xi}) = \E(e^{\mu + \sigma Z}) = e^\mu \E(e^{\sigma Z}) = e^{\mu} \E(e^{\sigma Z - \frac{1}{2} \sigma^2 + \frac{1}{2} \sigma^2}) = e^{\mu} e^{\frac{1}{2} \sigma^2} \underbrace{\E(e^{\sigma Z - \frac{1}{2} \sigma^2 t})}_{=1} = e^{\mu + \frac{1}{2}\sigma^2}
	% 	\end{align*}
	% \end{proof}

% subsection merton_s_model (end)

% lecture_4 (end)

\section{Valuation of Bond Options} % (fold)
\label{sub:valuation_of_bond_optiosn}

Consider a European call option on a $U$-maturity zero-coupon bond with expiry $T$ and strike $K$ where $t \leq T < U$ and $K > 0$.  The payoff at time $T$ equals \[
	C_T = \left(B(T, U) - K \right)^+ = \left(B(T, U) - KB(T, T) \right)^+
\]  We postulate that \begin{align*}
	C_t 	&= B_t \E_{\P^\star} \left(B_T^{-1} C_T \given \sigf_t \right) \\
			&= \E_{\P^\star} \left( e^{-\int_t^T r_u \, du} \left(v\left(r_T, T, U \right) - K\right)^+ \right)	
\end{align*}

The idea is to change the martingale measure $\P^\star$ to another probability measure $\Q$ such that \begin{align*}
	C_t &= B(t, T) \E_{\Q} \left(C_T \given \sigf_t \right) \\
	&= B(t, T) \E_{\Q} \left( \left(F_t \xi - K \right)^+ \given \sigf_t \right)
\end{align*} where $F_t = \frac{B(t, U)}{B(t, T)}$.  The measure $\Q$ is equivalent to $\P$ on $(\Omega, \sigf_T)$ and it is chosen in such a way such that $(F_t)_{t \in [0, T]}$ is a $\Q$-martingale.

Alternatively, consider a claim $X = C_T$ maturing at time $T$.  Then \begin{align*}
	C_t &= B_t \E_{\P^\star} \left( \frac{C_T}{B_T} \given \sigf_t \right) \\
	\Phi_t(X) &= B_t \E_{\P^\star} \left( \frac{X}{B_T} \given \sigf_t \right)
\end{align*}
\begin{exmp}
	In the context of equity options this approach yields the following representation of the price of a call option: \[
		C_t = S_t \hat P \left(S_T > K \given \sigf_t \right) - K B(t, T) \P^\star \left( S_T > K \given \sigf_t \right)
	\] 
	where \begin{align*}
		\frac{B_t}{S_t} &\text{ is a $\hat P$-martingale} \\
		\frac{S_t}{B_t} &\text{ is a $P^\star$-martingale}
	\end{align*}  If $B_t = e^{rt}$ (deterministic) then $\hat P =  P^\star$.
\end{exmp}

\section{The CIR Model} % (fold)
\label{sub:the_cir}
We postulate that \[
	dr_t = (a- b r_t) \, dt + \sigma \sqrt{r_t} \, dW^\star_t,
\]  where $a, b \sigma$ are positive constants.  Using Yamada-Watanabe theorem, we obtain uniqueness and existence of solutions.  A suitable comparison theorem tells us that if $r_0 > 0$ then $r_t \geq 0$ for $ t \in [ 0, T]$.  It is known that the solution $r$ to the CIR equation is related to the Bessel process.  It is known that \begin{enumerate}[(i)]
	\item $B(t, T) = e^{m(t, T) - n(t, T) r_t}$ where $m$ and $n$ can be computed explicitly using the PDE approach.  
	\item The price of a call option can be computed explicitly using the probabilistic approach.
\end{enumerate}

One can prove that \[
	C_t = B(t, U) \Phi_1(B(t, U), B(t, T), t, T, U) - K B(t, T) \Phi_2(B(t, U), B(t, T), t, T, U)
\] where $\Phi_1, \Phi_2$ are given explicitly in terms of the distribution of a Bessel process.

\section{Calibration} % (fold)
\label{sub:calibration}

We denote by $\hat B(0, T)$ the market price of a zero coupon bond with maturity $T$.  We assume that \[
	\hat B(0, T) = e^{- \int_0^T \hat f(0, u) \, du}
\] where the instantaneous forward rate  is a differentiable function such that \[
	\hat f_T(0, t) 
\] exists for $t \in 0, T$.  In general, we can fit to market data a model of the form \[
	dr_t = (a(t) - br_t) \, dt + \sigma r_t^\beta \, W^\star_t
\] for $\beta \in [0, 1]$.

\begin{prop}
	Let $\beta = 0$. Then the model fits the market data if and only if $a(t) = \hat f_T(0, t) + h'(t) + b(\hat f(0, t) + h(t))$ where \[
		h(t) = \frac{\sigma^2\left(1-e^{-bt}\right)^2}{2b^2}.
	\]
\end{prop}
It is essential here to assume that the function $\hat f(0, T)$ is differentiable with respect to $T$.  If we wish to produce a model such that $f(0, T) = \hat f(0, T)$.

% subsection calibration (end)
% subsection the_cir (end)
% subsection valuation_of_bond_optiosn (end)

\chapter{The HJM Approach to Modelling Bond Prices} % (fold)
\label{cha:the_hjm_approach_to_modelling_bond_prices}

\section{Introduction} % (fold)
\label{sec:introduction_to_hjm}

% section introduction (end)
Take as inputs the following objects \begin{enumerate}[(i)]
	\item $(\Omega, \F, \P)$, $W$, a $d$-dimensional Brownian motion.
	\item The dynamics of a family of processes \[
		\{ f(t, T), t \in [0, T] \}, T \in [0, T^\star]
	\] where $f(\cdot, T)$ is an $\F$-adapted process such that \[
		df(t, T) = \alpha(t, T) \, dt + \sigma(t, T) \cdot dW_t
	\] with some initial condition $f(0, \cdot) : [0, T^\star] \rightarrow \R$.  
\end{enumerate} As an output, we obtain the family of bond prices \[
	\{ B(t, T), t \in [0, T] \}, T \in [0, T^\star]
\] given by \[
	B(t, T) = \exp \left(-\int_t^T f(t, u) \, du \right)
\]  We must first derive the dynamics of $B(\cdot, T)$ under $\P$ for any maturity $T$ in the following form \[
	dB(t, T) = B(t, T) \left( a(t, T) \, dt + b(t, T) dW^\star_t \right)
\] where $a$ and $b$ are given in terms of $\alpha$ and $\beta$.

Next, we will find out under which assumptions on $\alpha$ and $\beta$ the HJM model admits a spot martingale measure $\P^\star$ or equivalently, a forward martingale measure $\P_{T^\star}$.

By definition, $\P^\star$ is any probability measure on $(\Omega, \sigf_{T^\star})$ such that $\P^\star \sim \P$ and the processes \[
	Z_t = \frac{B(t, T)}{B_t} = \frac{B(t, T)}{\exp\left(\int_0^t f(u, u) \, du \right)}
\] are $\P^\star$-(local) martingales.  Similarly, $\P_{T^\star} \sim \P$ and the processes \[
	F(t, T, T^\star) = \frac{B(t, T)}{B(t, T^\star)} 
\] are $\P_{T^\star}$-(local) martingales.

\begin{note}
	Let $F(t, T, U) = \frac{B(t, T)}{B(t, U)}$.

	\begin{enumerate}[(i)]
		\item If $U \leq T$, then $F(t, T, U)$ is the forward price of a $T$-maturity bond for the settlement date at time $U$.
		\item If $U \geq T$ then $F(t, T, U)$ represents the forward rate in the FRA initiated at time $t$ for the future time interval $[T,U]$. 
	\end{enumerate}  
\end{note}

\begin{defn}[HJM approach]
	Assume that \[
		df(t, T) = \alpha(t, T) \, dt + \sigma(t, T) \cdot dW_t
	\] with $W$ a $d$-dimensional Brownian motion and \[
		\sigma(t, T) \cdot dW_t = \sum_{i=1}^d \sigma^{i}(t, T) dW_t^{i}. 
	\]  All processes are specified under $\P$.

	We define $B(t, T) = e^{-\int_t^T f(t, u) \, du}$.  
\end{defn}

\begin{lem}
	Let $\alpha^\star(t, T) = \int_t^T \alpha(t, u) \, du$, and $\sigma^\star(t, T) = \int_t^T \sigma(t, u) \, du$.  These are $\F$-adapted processes.

	Then we claim that \[
		dB(t, T) = B(t, T) \left( a(t, T) \, dt + b(t, T) \cdot dW_t \right)
	\] where \begin{align*}
		a(t, T) &= f(t, t) - \alpha^\star(t, T) + \frac{1}{2} \left(\sigma\star(t, T)\right)^2 \\
		b(t, T) &= - \sigma^\star(t, T).
	\end{align*}
\end{lem}

Let $Z(t, T) = \frac{B(t, T)}{B_t}$, with $B_t = e^{\int_0^t f(u, u) \, du}$, so that \[
	dZ(t, T) = Z(t, T) \left( \left(\frac{1}{2} \left( \sigma(t, T) \right)^2 - \alpha^\star(t, T) \right) \, dt - \sigma^\star(t, T) \cdot dW_t \right) 
\]

Under which assumptions on $\alpha$ and $\sigma$ does there exists a probability measure $\Q \sim \P$ on $(\Omega, \sigf_{T^\star})$ such that $Z(t, T), t \in [0, T]$ is a $\Q$-martingale for every $T \in [0, T^\star]$.

We can also form process \[
	F_B(t, T, T^\star) = F(t, T, T^\star) = \frac{B(t, T)}{B(t, T^\star)}.
\]

\section{Trading Strategies} % (fold)
\label{sec:trading_strategies}
We first choose $\tau = \{ T_1 < T_2 < \dots < T_k \leq T^\star \}$ and take some $\F$-adapted process $\phi = (\phi^1, \dots, \phi^k)$.  $\tau$ represents the maturities of traded bonds.  $\phi^i$ represents the number of shares of $\tau_i$-maturity bonds.

Then the wealth process of $(\phi, \tau)$ equals \[
	V_t(\phi) = \sum_{i=1}^k \phi^i_t B(t, T_i).
\]

\begin{defn}[Self-financing]
	We say that $\phi$ is self financing if \[
		dV_t(\phi) = \sum_{i=1}^k \phi_t^i \, dB(t, T_i).
	\]
\end{defn}

\begin{lem}{\ }
	\begin{enumerate}[(i)]
		\item 
	Let $V_t^\star(\phi) = \frac{V_t(\phi)}{B_t}$.  Then $\phi$ is self-financing if and only if \[
		dV^\star_t(\phi) = \sum_{i=1}^k \phi_t^i d Z(t, T_i).
 	\] 
		\item 
		Let $F_v(t, T) = \frac{V_t(\phi)}{B(t, T)}$ for some $0 < T \leq T^\star$. Then $\phi$ is self-financing if and only if \[
			dF_v(t, T) = \sum_{i=1}^k \phi^i_t d \left( \frac{B(t, T_i)}{B(t, T)} \right) = \sum_{i=1}^k  \phi^i_t dF(t, T_i, T)
		\] where we assume $T \geq T_k$.
	\end{enumerate}

\end{lem}
% section trading_strategies (end)

\section{Martingale Measures} % (fold)
\label{sec:martingale_measures}
We will first address the issue of existence of the so-called \emph{forward martingale measure}, that is, a martingale measure for processes $\frac{V_t(\phi)}{B(t, T^\star)}$ or equivalently, a martingale measure for processes \[
	F_B(t, T, T^\star) = \frac{B(t, T)}{B(t, T^\star)}, t \in [0,T], T \in [0, T^\star].
\]

\begin{lem}
	For any $T \in [0, T^\star]$, \begin{align*}
		dF_B(t, T, T^\star) = F_B(t, T, T^\star) \left( \tilde a(t, T) \, dt +  (b(t, T) - b(t, T^\star) ) \, dW_t \right) 
	\end{align*} where \[
		\tilde a(t, T) = a(t, T) - a(t, T^\star) - b(t, T^\star) \left( b(t, T) - b(t, T^\star) \right)
	\]

	We denote by $\hat \P = \P^\star$ the martingale equivalent to $\P$ on $(\Omega, \sigf_{T^\star})$ by \[
		\frac{d\hat P}{d \P} = \doleans{h_u \, dW^\star_u}
	\]  If $h$ is such that \[
		\E\left(\mathcal E_{T^\star} \left( \int_0^\cdot h_u \, dW_u \right) \right) = 1
	\] the $\hat \P$ is well defined and we can compute the dynamics of $F_B(t, T, T^\star)$ under $\hat P$ with respect to $\hat W$, where \[
		\hat W = W_t - \int_0^t h_u \, du, t \in [0, T^\star]
	\]

	Assume that \begin{equation}
		a(t, T) - a(t, T^\star) = (b(t, T^\star) - h_t) \cdot (b(t, T) - b(t, T^\star))
		\label{eq:3.10_in_lecture_notes}
	\end{equation}  Condition \eqref{eq:3.10_in_lecture_notes}  in the lecture notes ensures that there is no drift term in the dynamics of $F_B(t, T, T^\star)$ under $\hat P$ for all maturities $T$.  After some computations, \eqref{eq:3.10_in_lecture_notes} can be represented as follows
	\[
	\alpha(t, T) + \sigma(t, T) \left(h_t + \int_T^{T^\star} \sigma(t, u) \, du \right) = 0.
	\]
\end{lem}

Later on we will denote by $\P_T$ the forward measure for the date $T$.  Thus $\hat P = \P_{T^\star}$.

% section martingale_measures (end)

\subsection{Spot Martingale Measure} % (fold)
\label{sec:spot_martingale_measure}
We know that \[
	dZ(t, T) = - Z(t, T) \left( \left(\alpha^\star(t, T) - \frac{1}{2} \left| \sigma^\star(t, T) \right|^2 \right) \, dt + \sigma^\star(t, T) \, dW_t \right)
\]

Now, the conditions for the drift term in $dZ(t, T)$ disappearing reads \begin{align*}
	\alpha^\star(t, T) &= \frac{1}{2} |\sigma^\star(t, T) |^2 - \sigma^\star(t, T) \lambda_t \\
	\Updownarrow  \\
	\alpha(t, T) &= \sigma(t, T) \left(\sigma^\star(t, T) - \lambda_t \right)
\end{align*}

The last formula can be seen as a tool for simple derivations of processes of interest interest under the measure $\P^\star$ (setting $\lambda = 0$).  We denote \[
	W^\star_t - W_t - \int_0^t \lambda_u \, du
\] 
% subsection spot_martingale_measure (end)

\subsection{Forward Measure} % (fold)
\label{sub:forward_measure}
We are going to examine the relationship between $\P^\star$ and $\P_T$ in a general term structure model.

\begin{note}
	Define the following.
	\begin{align*}
		dB(t, T) &= B(t, T) \left(r_t \, dt + b(t, T) \, dW^\star_t \right) \\
		d \zeta^i_t &= \zeta^i_t \left(r_t \, dt + \sigma^i_t \, dW^\star_t \right)
	\end{align*}
\end{note}

By definition, \[
	\pi_t(X) = B_t \E_{\P^\star} \left( \frac{X}{B_T} \given \sigf_t \right)
\]  Can we fine $\Q$ such that $\Q \sim \P^\star$ and \[
	B_t \E_{\P^\star} \left( \frac{X}{B_T} \given \sigf_t \right) = B(t, T) \E_\Q \left( X \given \sigf_t \right)
\] for any claim $X \in \sigf_T$ where $B(t, T) = B_t\E_{\P^\star} \left( \frac{1}{B_T} \given \sigf_t \right)$.  Formally, \[
	\E_{\Q} \left( X \given \sigf_t \right) = \frac{\E_{\P^\star} \left( \frac{X}{B_T} \given \sigf_t \right)}{\E_{\P^\star} \left( \frac{1}{B_T} \given \sigf_t \right)}
\]

We are guessing that $\Q \sim \P^\star$ with density on $(\Omega, \sigf_t)$ \[
	\frac{d\Q}{d \P} = \frac{1}{B(0, T) B_T}, \P^\star-a.s.
\]  $\E_{\P^\star} \left(\frac{1}{B_T} \right) = B(0, T)$.

\begin{defn}
	Suppose that $\P^\star$ is a spot martingale measure for our model.  Then for any maturity $T \in [0, T^\star]$, we define the forward martingale measure for the date $T$ by setting on $(\Omega, \sigf_{T^\star})$ \[
		\frac{d\P_T}{d\P^\star} = \frac{1}{B(0, T) B_T}, \P^\star-a.s.
	\]
\end{defn}

\begin{prop}{\ }
	\begin{enumerate}[(i)]
		\item \begin{align*}
			\frac{d\P_T}{d\P^\star} \given \sigf_t &= \E_{\P^\star} \left( \frac{d\P_T}{d\P^\star} \given \sigf_t \right) \\
				&= \E_{\P^\star} \left( \frac{B_0 B(T, T)}{B(0, T) B_T} \given \sigf_t \right) \\
				&= \frac{B_0}{B(0, T)} \E_{\P^\star} \left( \frac{B(T, T)}{B_T} \given \sigf_t \right) \\
				&= \frac{B_0}{B(0, T)} \frac{B(t, T)}{B_t}, \P^\star-a.s.
		\end{align*}
	\end{enumerate} 
\end{prop}

Recall that $\frac{\pi_t(X)}{B_t}$ is a $\P^\star$-martingale. Similarly, $\frac{\pi_t(X)}{B(t, T)}$ is a $\P_T$-martingale.  If $\eta_t = \frac{d \P_T}{d\P^\star} \given \sigf_t$ then $M$ is a $\P_T$-martingale if and only if $M \eta$ is a $\P^\star$-martingale.

\begin{exer}
	If we know that under $\P$ processes $\frac{X_t}{Z_t}$ are martingales where $Z$ is a fixed, positive process and under $\Q$ process $\frac{X_t}{Y_t}$ are martingales for a fixed positive process $Y$ then we can find a density of $\Q$ with respect to $\P$ in terms of $Z$ and $Y$. 
\end{exer}

We consider an arbitrage free model of bond prices and stock prices in which the spot martingale measure $\P^\star$ exists, such that $\frac{B(t, T)}{B_t}$ and $\frac{S^i_t}{B_t}$ are $\P^\star$-martingales.

We do not postulate that our model is complete.

Assume that $X$ is an attainable claim in this model.  We know that the arbitrage price $\pi_t(X)$ is unique and it can be computed using the risk-neutral valuation formula \[
	\pi_t(X) = B_t \E_{\P^\star} \left( \frac{X}{B_T} \given \sigf_t \right).
\]

\begin{rem}
	How do we find the forward price of $X$ at the time $t$ in the forward contract with settlement date $T$.
\end{rem}

\begin{defn}[Forward contract]
	The forward contract written at time $t$ on a time $T$ contingent claim is represented by the time $T$ contingent claim \[
		G_T = X - F_X(t, T)
	\] such that \begin{enumerate}[(i)]
		\item $F_X(t, T)$ is an $\sigf_t$-measurable random variable, 
		\item the arbitrage price at time $t$ on a contingent clam $G_T$ equals zero, that is, $\pi_t(G_T) = 0$.
	\end{enumerate}
\end{defn}

To compute $F_X(t, T)$, we will use the risk-neutral formula

\begin{align*}
	\pi_t(G_T) &= B_t \E_{\P^\star} \left( \frac{G_T}{B_T} \given \sigf_t \right) \\
		&= B_t \E_{\P^\star} \left( \frac{X}{B_T} \given \sigf_t \right) - F_X(t, T) B_t \E_{\P^\star} \left(\frac{1}{B_T} \given \sigf_t \right) \\
		&= \pi_t(X) - F_X(t, T) B(t, T) \\
		&= 0
\end{align*} and so \[
	F_X(t, T) = \frac{\pi_t(X)}{B(t, T)}.
\]

Define \begin{align*}
	F_Z(t, T) &= \frac{Z_t}{B(t, T)} \qquad \text{$Z_t = S_t$ or $B(t, T)$} \\
	F_S(t, T) &= \frac{S_t}{B(t, T)} \qquad \text{forward price of stock $S$}\\
	F_B(t, U, T) &= \frac{B(t, U)}{B(t, T)} \qquad \text{forward price of $U$-maturity bond}.
\end{align*}

\begin{defn}[Forward measure]

	We assume that $\P^\star$ is given.  The corresponding forward measure for the date $T, T \in [0, T^\star]$ is defined by \[
		\frac{d\P_T}{d\P^\star} = \frac{1}{B(0,T) B_T}, \quad \P^\star-a.s.
	\] so that \[
		\frac{d\P_T}{d\P^\star} \, |_{\sigf_t} = \frac{B_0}{B(0, T)} \frac{B(t, T)}{B_t}
	\] for every $t \in [0, T]$.
\end{defn}

\begin{lem}
	Assume that $W_t^\star$ is a Brownian motion under $\P^\star$ and \[
		dB(t, T) = B(t, T) \left( r_t \, dt + b(t, T) \, dW^\star_t \right)
	\]  Then $\eta_t \equiv \frac{d\P_T}{d\P^\star} \, |_{\sigf_t}$ equals \[
		\eta_t = \exp \left( \int_0^t b*u, T) \, dW^\star_u - \frac{1}{2} \int_0^t \left| b(u, T) \right|^2 \, du \right).
	\]  That is, \[
		\eta_t = \doleans{b(u, T) \, dW^\star_u}. 
		\tag{$\star$}
	\]

	It then follows that \[
		d \eta_t = \eta_t b(t, T) \, dW^\star_t, \quad \eta_0 = 1.
	\] and \[
		W_t^T = W^\star_t - \int_0^t b(u, T) \, du
	\] is a Brownian motion under $\P_T$.
\end{lem}

\begin{proof}
	Equation $(\star)$ follows from \[
		\frac{d\P_T}{d\P^\star} \, |_{\sigf_t} = \frac{B_0}{B(0, T)} \frac{B(t, T)}{B_t}
	\]

	The corollaries follow from differentiation and Girsanov's theorem, respectively.
\end{proof}

\begin{exer}
	Let $T \leq U$.  Find the dynamics of the forward price $F_B(t, U , T)$ under $\P_T$.  Apply the \ito formula under $\P^\star$, use Girsanov's theorem to express the dynamics of $F_B(t, U, T)$ in terms of $b(t, T)$, $b(t, U)$ and $W^T$.  Compute the volatility $\gamma(t, U ,T)$ of $F_B(t, U, T)$.  Apply the above the the HJM model $(\alpha(t, T), \sigma(t, T), W)$.
\end{exer}

\subsection{Applications of forward measures} % (fold)
\label{sub:applications_of_forward_measures}
\begin{enumerate}[(i)]
	\item Valuation of contingent claims.
	\item Construction of models for market rates.
\end{enumerate}  Application (i) is based on the following equality \[
	B_t \E_{\P^\star} \left( \frac{X}{B_T} \given \sigf_t \right) = B(t, T) \E_{\P_T} \left( X \given \sigf_t \right).
\]

\begin{lem}
	If $X$ is an attainable claim and settles at time $T$, then \[
		\pi_t(X) = B(t, T) \E_{\P_T} \left( X \given \sigf_t \right)
	\]
\end{lem}

\subsubsection{Valuation of claims with maturity $U \neq T$.}
Assume that $U \leq T$. Then the payoff $X$ at $U$ is equivalent to the payoff $Y = \frac{X}{B(U, T)}$ at time $T$.  Equivalence is understood in the sense that \[
	\text{$X$ at $U$} \sim \text{$Y$ at $T$} \iff \pi_t(X) = \pi_t(Y), t \in [0, U].
\]  So \[
	\pi_t(X) = B(t, U) \E_{\P_U} \left( X \given \sigf_t \right) = \pi_t(Y) = B(t, T) \E_{\P_T} \left( \frac{X}{B(U, T)} \given \sigf_t \right).
\]

To establish this equality, observe that for $t \in [0, U]$, \begin{align*}
	\frac{d \P_U}{d \P_T} \, |_{\sigf_t} = \frac{ \frac{d\P_U}{d\P^\star} |_{\sigf_t}}{\frac{d\P_T}{d\P^\star} |_{\sigf_t}} = \frac{ \frac{B_0}{B(0, U)} \frac{B(t, U)}{B_t}}{\frac{B_0}{B(0, T)}\frac{B(t, T)}{B_t}} = \frac{B(0, T)}{B(0, U)}\frac{B(t, U)}{B(t, T)}.
\end{align*}  We then need only apply the Bayes formula and apply the previous result.

Assume now that $U \geq T$. We postulate that $X$ is $\sigf_T$-measurable. Then the claim $Y = B(T, U) X$ is equivalent to $X$, in the sense that $\pi_t(X) = \pi_t(Y)$.

\begin{enumerate}[(i)]
	\item $U \leq T$.  Then $\pi_t(X) = B(t, T) = \E_{\P_T} \left( \frac{X}{B(U, T)} \given \sigf_t \right)$.  
	\item $U \geq T$ and $X \in \sigf_T$.  Then $\pi_t(X) = B(t, T) \E_{\P_T} \left( B(T, U) X \given \sigf_t \right)$.
\end{enumerate}

% subsection applications_of_forward_measures (end)

\section{The Gaussian HJM Model} % (fold)
\label{sec:examples_of_hjm_models}
Under $\P^\star$, \begin{equation}
	\label{eq:hjm_1}
	dB(t, T) = B(t, T) \left(r_t \, dt - \sigma^\star(t, T) \, dW^\star_t \right)
\end{equation} where \begin{equation}
	\label{eq:hjm_2}
	-\sigma^\star(t, T) = \int_t^T \sigma(t, u) \, du = b(t, T).
\end{equation}  Moreover, \begin{equation}
	\label{eq:hjm_3}
	df(t, T) = \sigma(t, T) \sigma^\star(t, T) \, dt  + \sigma(t, T) \, dW^\star_t
\end{equation} and \begin{equation}
	\label{eq:hjm_4}
	r_t = f(0, t) + \int_0^t \sigma(u, t) \sigma^\star(u, t) \, du + \int_0^t \sigma(u, t) \, dW^\star_u
\end{equation}

\begin{rem}
	From \eqref{eq:hjm_1} and \eqref{eq:hjm_3}, we see that for any fixed $T$, processes $B(t, T)$ and $f(t, T)$ are continuous semimartingales.  In \eqref{eq:hjm_4}, we integrate a different process for each $t$.  Also, as an additional input we take some function $f(0, t)$.

	Can we then compute $dr_t$?  The answer to this question is positive in some special cases.
\end{rem}

We now always postulate that $\sigma(t, T)$ is deterministic.  Then we say that we deal with the \emph{Gaussian HJM model} since $r_t$ has a normal distribution for any $t \in [0, T^\star]$.

Several examples of the Gaussian HJM model include: 
\begin{enumerate}[(i)]
	\item The Ho-Lee model.  We take $d = 1$ and $\sigma(t, T) = \sigma$.  Since $b(t, T) = - \sigma(T-t)$, it can also be seen as a counterpart to Merton's model.
	\item The bond price satisfies under $\P^\star$, \begin{align*}
		dB(t, T) = B(t, T) \left( r_t \, dt - \sigma(T- t) \, dW^\star_t \right).
	\end{align*}  The short term rate equals\[
		r_t = f(0, t) + \frac{1}{2} \sigma^2 t^2 + \sigma W^\star_t,
	\] so that \[
		dr_t = \underbrace{\left( f_T(0, t) + \sigma^2 t \right)}_{a(t)} \, dt + \sigma dW^\star_t .
	\] where the function $a : [0, T^\star] \rightarrow \R$ can also be derived if we start from the extended merton model $dr_t = a(t) \, dt + \sigma dW^\star_t$ and we fit this model to the yield curve $\E_{\P^\star} \left( e^{-\int_0^T r_t \, dt} \right) = e^{-\int_0^T f(0, t) \, dt}$.  We also need to show that $r_0 = f(0, 0)$.  To solve this problem, we need to assume that $f_T(0, t)$ exists.
	\item Vasicek's model.  Take $d = 1$ and $\sigma(t, T) = \sigma e^{-b(T-t)}$ where $\sigma, b$ are positive numbers.  Then \[
		b(t, T) = -\sigma^\star(t, T) = - \frac{\sigma}{b} \left(e^{-b(T- t)} - 1\right),
	\]  and other computations are given in the course notes.
\end{enumerate}

% section examples_of_hjm_models (end)

\chapter{Valuation of Options in Gaussian Models} % (fold)
\label{sec:valuation_of_options_in_gaussian_models}
\section{Options on Bonds} % (fold)
\label{sec:options_on_bonds}

% section options_on_bonds (end)
Consider any term structure in which at least some bonds are traded.  If the short term rate process is given then under $\P^\star$, \[
	dB(t, T_i) = B(t, T_i) \left( r_t \, dt + b(t, T_i) \, dW^\star_t \right)
\] where $b(t, T_i)$ is a deterministic function and $0 < T_1 < \dots < T_m$.  If $r$ is not explicitly specified then we should focus on the dynamics of the forward prices, for example \[
	F_B(t, T_i, T_j) = \frac{B(t, T_i)}{B(t, T_j)}, \quad i = 1, \dots, m
\] under the forward measure $\P_{T_j}$.

	How do we value and hedge European bond options with maturity $T$ and the underlying zero coupon bond maturing at $U > T$.  The payoff at $T$ equals \begin{align*}
		C_T &= \left( B(T, U) - K \right)^+ \\
		P_T &= \left( K - B(T, U) \right)^+ 
	\end{align*} so that \[
		C_T - P_T = B(T, U) - K
	\] and thus for $t \in [0, T]$, \[
		C_t - P_t = B(t, U) - KB(t, T). 
	\]  Instead of computing the expectation under $\P^\star$, \[
		C_t = B_t \E_{\P^\star} \left( \frac{C_T}{B_T} \given \sigf_t \right),
	\] we will compute the equivalent probability measure $P_T$ \[
		C_t = B(t, T) \E_{\P_T} \left( C_T \given \sigf_t \right).
	\]  Let $D = \left\{ B(T, U) > K \in \sigf_T \right\}$.  Then \[
		C_T = B(T, U) \mathbf{1}_D - K \mathbf{1}_D = X_1 - X_2
	\] So that \[
		C_t = \pi_t(X_1) - \pi_t(X_2) = I_1 - I_2.
	\]  For $I_2$, we compute \begin{align*}
		I_2 = \pi_t(K \mathbf{1}_D) = K B(t, T) \P_T\left(D \given \sigf_T \right).  
	\end{align*} We observe that \[
		B(T, U) = \frac{B(T, U)}{B(T, T)} = F_B(T, U , T)
	\] where under $\P_T$ the forward price $F_B(t, U, T), [t \in [0, T]]$ satisfies \[
		dF_B(t, U , T) = F_B(t, U, T) \left( b(t, U) - b(t, T) \right) \, dW^T_t
	\] so that $F_t = F_B(t, U, T)$ satisfies \[
		F_T = F_t \exp \left( \zeta(t, T) - \frac{1}{2} v^2( t, T) \right)
	\] where \[
		\zeta(t, T) = \int_0^T \gamma(u, U , T) \, dW^T_u, \quad v^2(t, T) = \int_t^T \left| \gamma(u, U, T) \right|^2 \, du
	\] where $\gamma(u, U, T) = b(u, U) - b(u, T)$.

	We need to compute \begin{align*}
		\P_T(D \given \sigf_t) &= P_T(B(T, U) > K  \given \sigf_t) \\
						&= \P_T ( F_B(T, U, T) > K \given \sigf_t) \\
						&= \P_T \left(F_t e^{\zeta(t, T) - \frac{1}{2} v^2(t, T)} \given \sigf_t \right),
	\end{align*} where $\zeta(t, T)$ is independent of $\sigf_t$ and $\zeta(t, T) \sim N(0, v^2(t, T))$.  Hence \begin{align*}
		\P_T(D \given \sigf_t) &= \P_T \left( F e^{\zeta(t, T) - \frac{1}{2} v^2(t, T)} \given F = F_t \right)  \\
		&= \P_T \left( \frac{\zeta(t, T)}{v(t, T)} > \ln \frac{K}{F} + \frac{1}{2} v^2(t, T) \given F = F_t \right) \\
		&= N( \tilde d_{-}(F_t, t, T))
	\end{align*} where $\tilde d_2(F_t, t, T) = \frac{\ln \frac{F}{K} \pm \frac{1}{2} v^2(t, T)}{v(t, T)}$.

	For $I_1,$ we need to compute the conditional expectation \[
		I_1 = B(t, T) \E_{\P_T} \left( B(T, U) \mathbf{1}_D \given \sigf_t \right)
	\] where \[
		\frac{B(T, U)}{C} = \frac{F_B(T, U, T)}{C} = \frac{d\tilde \P_T}{d\P_T}.
	\] so that \begin{align*}
		\frac{d \tilde \P_T}{d \P_T} \given \sigf_t &= \frac{F_B(t, U, T)}{C} \\
			&= \exp \left( \int_0^t \gamma(u, U, T) \, dW^T_u - \frac{1}{2} \int_0^t \left| \gamma(u, U, T) \right|^2 \, du \right) \\
			&= \tilde \eta_t
	\end{align*} for $t \in [0, T]$.  Note also that \begin{align*}
		\E_{\tilde \P_T} (X \given \sigf_t) = \frac{\E_{\P_T}(X \tilde \eta_t \given \sigf_t)}{\tilde \eta_t} \\
		\frac{F_B(t, U, T)}{c} \E_{\tilde \P_T}(\mathbf{1}_D \given \sigf_t) &= \E_{\P_T} \left( \mathbf{1}_D \frac{B(T, U)}{C} \given \sigf_t \right)
	\end{align*} and \begin{align*}
		\E_{\P_T} \left(B(T, U) \mathbf{1}_D \given \sigf_t \right) = \frac{B(t, U)}{B(t, T)} \tilde P_T (D \given \sigf_t) 
	\end{align*} and thus \[
		I_1 = B(t, U) \tilde P_T(D \given \sigf_t)
	\] 
	and since $dF_t = F_t \gamma(t, U , T) dW^T_t$ and \[
		\tilde W_t^T - \int_0^t \gamma(u, U, T) \, du
	\] is a $\tilde P_T$-Brownian motion, we obtain \[
		dF_t = F_t \left( \left|\gamma(t, U, T) \right|^2 \, dt + \gamma(t, U, T) \, d\tilde W^T_t \right)
	\] under $\tilde P_T$.  Solving this equation, we obtain \[
		F_T = F_t \exp \left( \int_t^T \gamma(u, U, T) \, d\tilde W^T_t + \frac{1}{2} \int_0^T \left| \gamma(t, U, T) \right|^2 \, du \right).
	\] and so \[
		\tilde \P_T(D \given \sigf_t) = N(\tilde d_{+}(F_t, t, T)).
	\]

	We conclude that \begin{align*}
		I_1 &= B(t, U) N(\tilde d_+(F_t, t, T)), \\
		I_2 &= K B(t, T) N(d_-(F_t, t, T)).
	\end{align*} so that the price of the call bond option is now known explicitly.  It remains to find out whether the call option can be replicated, for instance, by a trading strategy $\phi = (\phi^1, \phi^2)$ with the wealth process $V(\phi)$, \begin{align*}
		V_t(\phi) 	&= \phi_t^1 B(t, U) + \phi_t^2 B(t, T) \\
		dV_t(\phi) 	&= \phi_t^1 dB(t, U) + \phi^2_t dB(t, T) \\
		V_T(\phi) 	&= C_T = (B(T, U) - K)^+.
	\end{align*}  Equivalently,  \begin{align*}
		\frac{V_t(\phi)}{B(t, T)} &= \phi^1_t F_B(t, U, T) + \phi_t^2 \\
		d \left( \frac{V_t(\phi)}{B(t, T)} \right) &= \phi^1_t dF_B(t, U, T) \\
		\frac{V_T(\phi)}{B(T, T)} &= (F_B(T, U, T) - K)^+.  
	\end{align*}

	Let $F_V(t, T) = \frac{V_t(\phi)}{B(t, T)}$.  Then we need to solve the following problem \begin{align*}
		dF_V(t, T) = \phi_t^1 d F_B(t, U, T) \\
		F_V(T, T) = (F_B(T, U, T) - K)^+
	\end{align*} where \[
		dF_B(t, U, T) = F_B(t, U, T) \gamma(t, U, T) dW^T_t.
	\]  To solve this equation, observe that \[
		\frac{C_t}{B(t, T)} = \frac{B(t, U)}{B(t, T)}\left( N( \tilde d_+ (F_t, t, T)) - K N (\tilde d_-(F_t, t, T)) \right), 		
	\] and \[
		F_C(t, T) = F_t \left( N( \tilde d_+ (F_t, t, T)) - K N (\tilde d_-(F_t, t, T)) \right).
	\]

\begin{lem}
	Let $(Y_t)$ be given by \begin{align*}
		Y_t &= X_t \left( N( \tilde d_+ (X_t, t, T)) - K N (\tilde d_-(X_t, t, T)) \right) \\
		dX_t &= X_t \sigma(t) \, dW_t \\
		\tilde d_{\pm}(x, t, T) &= \frac{ \ln \frac{x}{K} \pm ^2(t, T)}{v(t, T)}. 
	\end{align*}

	Then \[
		dY_t = N(d_{-}(X_t, t, T)) \, dX_t.
	\]
\end{lem}

\begin{proof}
	Apply the \ito formula.  Assume here that $\sigma$ is deterministic. 
\end{proof}

 If we apply the lemma to $F_c(t, T)$, we obtain \begin{align*}
	dF_c(t, T) &= N(\tilde d_+(F_t, t, T)) \, dF_t \\
				&= \phi^1_t \, dF_t.
\end{align*} so that \[
	\phi^1_t = N(\tilde d_1(F_t, t, T))
\] and \[
	\phi^2_t = \frac{C_t - \phi^1_t B(t, U)}{B(t, T)}
\] Then \begin{align*}
	V_t(\phi) 	&= C_t = \phi^1_tB(t, U) + \phi^2_t B(t, T). \\
	dV_t(\phi) 	&= dC_t = \phi^1_t dB(t, U) + \phi^2-t dB(t, T).
\end{align*}

In the future, we will deal with more general options of the form \[
	C_T = (Z^1_T - K Z^2_T)^+
\]  where $Z^i$ is some portfolio of bonds.  Then the choice of a natural hedging strategy depends on the choice of traded assets.

\begin{lem}
	The price $C_t$ of a call option equals \[
		C_t = B(t, U) \P_U(D \given \sigf_t) - K B(t, T) \P_T(D \given \sigf_t)
	\]
\end{lem}

\begin{proof}
	\begin{align*}
		C_T = B(T, U) \mathbf{1}_D - K \mathbf{1}_D = X_1 - X_2 \\
		\pi_t(X_2) = B(t, T) \E_{\P_T} K \mathbf{1}_D \given \sigf_t = K B(t, T) \P_T (D \given \sigf_t)
	\end{align*} and $X_1 = B(T, U) \mathbf{1}_D$ is equivalent to $Y_1 = \mathbf{1}_D$ at time $U$, so that \begin{align*}
		\pi_t(X_1) = \pi_t(Y_1) = B(t, U) \P_{U}(D \given \sigf_t)
	\end{align*} for $t \in [0, T]$.
\end{proof}

\section{Options on Coupon Bonds} % (fold)
\label{sec:options_on_coupon_bonds}
Let $T_1 < T_2 < \dots < T_n \leq T^\star$ be coupon dates and $c_1 \dots, c_2$ he corresponding deterministic coupons.  Then the price $Z_t = B_c(t, T)$ of the coupon bond equals \[
	Z_t = \sum_{j=1}^n c_j B(t, T_j).
\]  We consider the call option with maturity $T < T_1$ and the payoff \[
	C_T = (Z_T - K)^+ = \sum_{j=1}^m c_j B(t, T_j) \mathbf{1}_D - K \mathbf{1}_D
\]   where \[
	D = \{ Z_T > K \}.
\]

One possible way of pricing this is to represent $C_t$ as follows: \begin{align*}
	C_t = \sum_{j=1}^n c_j B(t, T_j) \P_{T_j}(D \given \sigf_t) - K B(t, T)P_T(D \given \sigf_t).
\end{align*}

\begin{rem}[On the proof of Proposition 4.3]
	We know that if we set $D = \{ Z_T > K \}$, then \[
		C_t = \sum_{j=1}^m c_j B(t, T_j) \P_{T_j}(D \given \sigf_t) - K B(t, T) \P_T(D \given \sigf_t).
	\]
	For simplicity, we may set $t = 0$ - then we need to compute $\P_{T_j}(D)$ and $\P_T(D)$.  Recall that $T = T_0 < T_1 \dots < T_m$.  Then \[
		D = \left\{ \sum_{j=1}^m c_j \underbrace{F_B(T, T, T_j)}_{F^j_B(T)} > K \right\}
	\]       where $dF^j_B(t) = F^j_B(t)(b(t, T_j) - b(t, T)) dW^T_t$, and hence \[
		\P_T(D) = \P_T \left( \sum_{j=1}^m c_j F^j_B(0) e^{\int_0^T \gamma(t, T_j, T) \, dW^T_t - \frac{1}{2} \int_0^T \left| \gamma(t, T_j, T) \right|^2 \, dt} > K \right)
 	\]
If we denote $\zeta_j = \int_0^T \gamma(t, T_j, T) \, dW^T_t$, then the vector $\zeta = (\zeta_1, \dots, \zeta_m)$ has a normal distribution under $\P_T$, with mean $(0, 0, \dots, 0)$ and covariance $(\nu_{kl})$ where \[
	\nu_{kl} = \int_0^T \gamma(t, T_k, T) \cdot \gamma(t, T_l, T) \, dt.
\]

To compute $\P_{T_j}(D)$, we need to know the distribution of $\zeta$ under $\P_{T_j}$.  Since $W^{T_j}_t = W^T_t - \int_0^t \gamma(u, T_j, T) \, du$ it is clear that under $\P_{T_j}$, the forward price $F^l_B(t) = F_B(t, T_l, T)$ \[
	dF^l_B(t) = F^l_B(t) \gamma(t, T_l, T) \, dW^T_t + F_B(t) \gamma(t, T_l, T) \gamma(t, T_j, T) \, dt
\] so that the joint distribution of $\zeta_1, \dots, \zeta_m$ under each forward measure $\P_{T_j}$ can also be computed.  The joint distribution is Gaussian with the same covariance matrix but with means $v_{lj}$

\end{rem}

\section{Pricing of General Contingent Claims} % (fold)
\label{sec:pricing_of_general_contingent_claims}
Let $\zeta_i(t, T) = \int_t^T \gamma_i(u, T) \, dW^T_u$.  Then under $\P_T$ the random variables $\zeta_i(t, T), \dots, \zeta_n(t, T)$ are normally distributed with mean $(0, \dots, 0)$ and covariance matrix ($\gamma_{ij}$) given by \[
	\gamma_{ij} = \int_t^T \gamma_i(u, T) \gamma_j(u, T) \, du.
\]

\begin{prop}
	Let $X = g(Z^1_T, \dots, Z^n_T)$ at time $T$.  Then the price of $X$ at time $t \in [0, T)$ is given by \[
		\pi_t(X) = B(t, T) \int_{\R^k} g\left( \frac{Z^1_t}{B(t, T)} \frac{n_k(x+ \theta_1)}{n_k(x)}, \dots, \frac{Z^n_t}{B(t, T)} \frac{n_k(x+\theta_n)}{n_k(x)}\right) n_k(x) \, dx
	\] where $n_k$ is the standard $n$-dimensional Gaussian density on $\R^k$ and $(\theta_i)$ are elements of $\R^k$ such that \[
		\theta_i \theta_j = \gamma_{ij}
	\] for all $i, j$.   This follows from the Cholesky decomposition of the covariance matrix $(\gamma_{ij})$
\end{prop}

\begin{proof}
	\begin{align*}
		\pi_t(X) = B(t, T) \E_{\P_T} \left( g(F_{Z^1}(T, T)), \dots, F_{Z^n}(T, T) \given \sigf_t \right) \\
		F_{Z^i}(T, T) = F_{Z^i}(t, T) e^{\zeta_i(t, T) - \frac{1}{2} \gamma_{ii}} \\
		\pi_t(X) = B(t, T) \E_\Q \left( g\left( F_{Z_t^i} e^{\theta_i \eta - \frac{1}{2} \gamma_{ii}} \right) \given \sigf_t \right) \\
		&= B(t, T) \int_{\R^k} g\left( \frac{Z^i_t}{B(t, T)} e^{\theta_i \cdot x - \frac{1}{2} \underbrace{\gamma_{ii}}_{|\theta_i|^2}} \right) n_k(x) \, dx.
	\end{align*}  Since $\frac{n_k(x + \theta_i)}{n_k(x)} = e^{\theta_i \cdot x - \frac{1}{2}|\theta_i|^2}$, we obtain our result.
\end{proof}
% section pricing_of_general_contingent_claims (end)
% section options_on_coupon_bonds (end)
% section valuation_of_options_in_gaussian_models (end)
% chapter the_hjm_approach_to_modelling_bond_prices (end)

\chapter{Modelling of Forward LIBORs} % (fold)
\label{cha:modelling_of_forward_libors}
\section{Introduction to LIBOR} % (fold)
\label{sec:introduction}
Let $\delta$ equal 3 months.  If $L(0) = 10\%$ then if we borrow $N$ at time $0$, we will pay back after three months the amount $N(1 + \delta L(0))$ where the unit is one year so that $\delta = \frac{1}{4}$.

\begin{enumerate}[(i)]
	\item Spot LIBOR is (or was) the most commonly used rate for interbank funding and as an underlying for interest rate derivatives such as caps and floors.  
	\item By convention, the pricing formula for caplets and floorlets was a version of the Black formula which reads \[
		C_T = F_t N(d_+) - K N(d_-)
	\] where $F_t$ is the forward price of the underlying asset. 
\end{enumerate}

Let us consider a caplet with maturity $T$ and settlement date $T + \delta$.  Here, a caplet is a call option on LIBOR, in the sense that it pays the amount $C_P = (L(T) - K)^+ \delta N$ at time $T + \delta$ where $T$ is the maturity date, $N$ is the nominal value, and $x$ the strike level.

\begin{defn}[Cap]
	 A cap is a portfolio of caplets over non-overlapping periods \[
	 	0 < T_0 < T_1 < \dots < T_n 
	 \] so we have $n$ caplets, struck at $T_{i}$ for the period $[T_i, T_{i+1}]$ and paying $(L(T_i) - K)^+ N \delta_{i+1}$ at $T_{i+1}$, where $\delta_{i+1} = T_{i+1} - T_i$.
\end{defn}

By convention, the price of a caplet over $[T, T+\delta]$ equals \[
	\textsc{Cpl}_t = B(t, T+\delta) \left( L(t) N(d_+) - KN(d_-)\right)
\] where \[
	d_\pm = \frac{\ln \frac{L(t)}{K} \pm \frac{1}{2} \sigma^2(T-t)}{\sigma \sqrt{T-t}}.
\]

% section introduction (end)

\section{Caps and Floors in the LIBOR Market Model} % (fold)
\label{sec:caps_and_floors_in_the_libor_market_model}

A caplet (floorlet) is a protection against the rise (fall) in the LIBOR rate. The caplet (floorlet) pays off: \begin{align*}
 	\mathbf{Cpl}^j_{T_j} 	N\left( L(T_{j-1}) - \kappa\right)^+ \delta_j \\
	\mathbf{Frl}^j_{T_j} 	N\left(  \kappa - L(T_{j-1}) \right)^+ \delta_j \\
\end{align*} paid at time $T_j$.

We clearly have the \emph{cap-floor put call parity}, \[
	\mathbf{Cpl}^j_{T_j} - \mathbf{Frl}^j_{T_j} = N_p \left(L(T_{j-1}) - \kappa \right) \delta_j. 
\]

\begin{exer}
	Using this relationship, find the difference $\mathbf{CPl}_t^j - \mathbf{Frl}^j_t$ for any $t \in [0, T_{j-1}]$.
\end{exer}

Recall that \[
	1 + \delta_j L(T_{j-1}) = \frac{1}{B(T_{j-1}, T_j)}
\]  Hence \begin{align*}
	\mathbf{Cpl}^j_{T_j} = N \left( \frac{1}{B(T_{j-1}, T_j)} - \underbrace{(1 + \delta_j \kappa)}_{\tilde \delta_j} \right)^+ \delta_j
\end{align*}  An equivalent payoff at time $T_{j-1}$ equals \begin{align*}
	\tilde{\mathbf{Cpl}}^j_{T_{j-1}} &= B(T_{j-1}, T_j) \mathbf{Cpl}_{T_j}^j \\
				&= \tilde \delta_j N \left( \frac{1}{\tilde \delta_j} - B(T_{j-1}, T_j) \right)^+.
\end{align*}

% section caps_and_floors_in_the_libor_market_model (end)
% chapter modelling_of_forward_libors (end)

\begin{defn}
	The forward swap rate $\kappa(t, T_0, T_1, \dots, T_n) = \kappa(t, T, n)$ where $T_0 = T$ is the $\sigf_t$-measurable random variable such that $\mathbf{FS}_t \left( \kappa(t, T, n) \right) = 0$.  
\end{defn}

\begin{lem}
	The forward swap rate equals \begin{align*}
		\kappa(t, T, n) = \frac{B(t, T_0) - B(t, T_n)}{\sum_{j=1}^n \delta_j B(t, T_j)}
	\end{align*}
\end{lem}

\chapter{Modelling of Forward Swap Rates} % (fold)
\label{cha:modelling_of_forward_swap_rates}
\begin{enumerate}[(i)]
	\item Definition and payoffs of an $n$-period forward swap.
	\item Valuation formula for a forward swap (6.4)
	\item Definition and formula for forward swap rates (6.5)
	\item Definition and equivalent representations for a swaption (Lemma 6.5)
	\item Postulates of Jamshidian's model of co-terminal forward swap rates
	\item Valuation of a swaption (Proposition 6.3)
	\item Choice of a numeraire portfolio
\end{enumerate}

Consider the family of co-terminal swap rates \begin{align*}
	\kappa(t, T_0; n) &= \frac{B(t, T_0) - B(t, T_n)}{\sum_{k=1}^n \delta_k B(t, T_k)} \\
	\kappa(t, T_1,; n-1) &= \frac{B(t, T_1) - B(t, T_n)}{\sum_{k=2}^n \delta_k B(t, T_k)} \\
	&\downarrow \\
	\kappa(t, T_{n-1}; 1) &= \frac{B(t, T_{n-1}) - B(t, T_n)}{\delta_n B(t, T_n)} = L(t, T_{n-1}) 
\end{align*}  For ease of notation, we let $\kappa(t, T_; n-j) = \tilde \kappa(t, T_j)$.

\section{Payer Swaptions} % (fold)
\label{sec:payer_swaptions}
Let us take $j = 0$ so that the underlying forward swap has $n$ periods.  Let $\mathbf{FS}_t(\kappa)$ denote the value of the forward swap. We know that \[
	\mathbf{FS}_t(\kappa) = B(t, T_0) - \sum_{j=1}^n c_j B(t, T_j)
\] where $c_j = \kappa \delta_j$, $j = 1, \dots, n-1$, $c_n = (1 + \kappa \delta_n)$.

\begin{lem}
	The price $\mathbf{FS}_t(\kappa)$ can be represented as follows: \begin{align*}
		\mathbf{FS}_t(\kappa) &= \mathbf{FS}_t(\kappa) - \mathbf{FS}_t(\kappa(t, T_0; n)) \\
					&= \sum_{j=1}^n \left( \kappa(t, T_0; n) - \kappa \right) \delta_j B(t, T_j) \\
					&= G_t(n)
	\end{align*}  where \[
		G_t(n) = \sum_{\delta_k B(t, T_k)}, \quad G_t(n-j) = \sum_{k=j+1}^n \delta_k B(t, T_k)
	\]  
\end{lem}

A payer swaption with a fixed rate $\kappa$, maturing date $T = T_0$ and the underlying $n$-period fixed-for-floating forward swap can be identified with the payoff $\left(\mathbf{FS}_T(\kappa) \right)^+$ at time $T$.  A receiver swaption pays $\left(-\mathbf{FS}_T(\kappa) \right)^+$ at time $T$.  Of course, we have a put call parity relationship \[
	\mathbf{PS}_t(\kappa) - \mathbf{RS}_t(\kappa) = \mathbf{FS}_t(\kappa)
\]

The inequality $\mathbf{FS}_t(\kappa) > 0$ holds if and only if $\kappa(T, T; n) > \kappa$ where $\kappa(T, T; n)$ is the spot swap rate at time $T_0$. Hence if $\kappa(T, T; n) \leq \kappa$ the swaption expires worthless, but it is still possible to enter at $T$ a forward swap with fixed rate $\kappa(T, T; n) \leq \kappa$.

If we define \[
	Y_k = \delta_k \left( \kappa(T, T; n) - \kappa \right)^+,
\] we know that \begin{align*}
	\left( \mathbf{FS}_t(\kappa) \right)^+ &= \sum_{k=1}^n \delta_k B(T, T_k) \left(\kappa(T, T; n) - \kappa \right)^+ \\
	&= \sum_{k=1}^n B(T, T_k) Y_k
\end{align*} which is equivalent to a sequence of payoffs $Y_1, \dots, Y_n$ at times $T_1, \dots, T_n$.  Also for $j = 0, 1, \dots, n-1$, \begin{align*}
	\left( \mathbf{FS}_{T_0}^0 (\kappa) \right)^+ = G_{T_0}(n) \left( \kappa(T_0, T_0; n) - \kappa \right)^+ \\
	\left( \mathbf{FS}_{T_j}^j (\kappa) \right)^+ = G_{T_j} (n-j) \left( \kappa(T_j, T_j) ;n) - \kappa \right)^+
\end{align*}

We now seek to construct a model for the joint dynamics of a co-terminal family of forward swap rates \[
	\kappa(t, T_j; n-j) = \tilde \kappa(t, T_j), t \in [0, T_j] 
\] such that the volatility $\nu(t, T_j)$ is given in advance by a deterministic function and the model is driven by a $d$-dimensional Brownian motion.\footnote{Any process that we can apply Girsanov's theorem to will be sufficient.}

We expect that each process $\tilde \kappa(t, T_j)$ will be a martingale under some probability measure $\tilde P_{T_{j+1}}$ so that \[
	d\tilde \kappa(t, T_j) = \tilde \kappa(t, T_j) \nu(t, T_j) \, d\tilde W^{T_{j+1}}_t
\] where $\tilde W_t^{T_{j+1}}$ is a Brownian motion under $\tilde P_{T_{j+1}}$ and the Radon-Nikodym densities for $j=0, \dots, n-1$ should be given by \[
	\frac{d \tilde P_{T_{j+1}}}{d \tilde \P_{T_n}} = \mathbf{?}
\] which should be expressed in terms of $\tilde W^{T_n}, \tilde \kappa(t, T_k), \nu(t, T_k)$ for $k = n-j+1, \dots, n$.

\section{Valuation of Swaptions in Jamshidian's Model} % (fold)
\label{sec:valuation_of_swaptions_in_jamshidian_s_model}
Let us assume that the model is well defined.  We will value the $j$-th swaption for $j = 0, \dots, n-1$.  Suppose that it is attainable, so that the price can be computed using the martingale method, meaning here that \[
	\pi_t(X) = G_t(n-j) \E_{\tilde \P_{T_{j+1}}} \left( \frac{X}{G_{T_j}(n-j)} \given \sigf_t \right)
\] where $X$ is any attainable claim in Jamshidian's model with maturity $T$.  Observe that only a finite family of forward swaps are traded in this model.  In our case, $X = G_{T_j}(n-j) \left( \tilde \kappa(T_j, T_j) - \kappa \right)^+$, and thus \[
	\mathbf{PS}_t^i(\kappa) = G_t(n-j) \E_{\tilde \P_{T_{j+1}}} \left( \left( \tilde \kappa(T_j, T_j) - \kappa \right)^+ \given \sigf_t \right).
\]  Since $\eta(t, T_j): [0, T_j] \rightarrow \R^d$ is deterministic, we can evaluated this expression using the Black formula, and obtain \[
	\tilde \kappa(t, T_j) \Phi \left( \tilde d_+^j(\tilde \kappa(t, T_j), t, T_j) \right) - \kappa \Phi \left(\tilde d_-^j (\tilde \kappa(t, T_j), t, T_j) \right)
\]

 where \begin{align*}
	\tilde d_{\pm}(x, t, T_j) &= \frac{\ln \frac{x}{\kappa} \pm \frac{1}{2} v^2_j(t, T_j)}{v_j(t, T_j)} \\
	v_j(t, T_j) &= \int_t^{T_j} \left| v(u, T_j) \right|^2 \, du.
\end{align*}

For replication of a swaption, we formally define the relative price \[
	\mathbf{F}_{S_j, G}(t, T_j) = \frac{\mathbf{PS}_t^j}{G_t(n-j)} = \tilde \kappa(t, T_j) \Phi \left( \tilde d_+^j(t) \right) - \kappa \Phi(\tilde d^j_-(t)).
\]  In this case, \[
	dF_{S_j, G}(t, T_j) = \Phi(\tilde d_+^j(t)) \, d \tilde \kappa(t, T_j).
\]

It is possible to then hedge this option using forward swaps in discrete time.

Let $\psi^j$ be any trading strategy in the $j$-th forward swap.  At time $0$ the value of our strategy is zero.  Then the trading strategy: \begin{align*}
	t = 0 & \quad \text{$\psi^j_0$ positions in market forward swap with rate $\tilde \kappa(0, T_j)$} \\
	t = t_1  & \quad \text{$\phi^j_{t_1}$ positions in market forward swap with rate $\tilde \kappa(t_1, T_j)$} \\
		&\downarrow
	t = t_n = T_j
\end{align*}  Then gains and losses can be conveniently expressed in units of $G_t(n-j)$.  For instance, the value of our $\psi^j_0$ positions at time $t_1$ equals \begin{align*}
	\mathbf{PL}_{t_1} = G_{t_1}(n-j) \psi^j_0 \left( \tilde \kappa(t_1, T_j) - \tilde \kappa(0, T_j) \right) \\
	\tilde{\mathbf{PL}}_{t_1} = \underbrace{\psi^j_0 \left( \tilde \kappa(t_1, T_j) - \tilde \kappa(0, T_j) \right)}_{\text{paid in installments at times $T_{j+1}, \dots, T_n$.}}
\end{align*}  After $n$ steps, \begin{align*}
	\tilde{\mathbf{PL}}_{T_j} = \sum_{k=0}^{n-1} \psi_{t_k}^j \left( \kappa(t_{k+1}, T_j) - \tilde \kappa(t_k, T_j) \right) \\
	&\rightarrow_{\substack{n \rightarrow \infty \\ t_k = \frac{k}{n} T_j}} \int_0^{T_j} \psi^j_u \, d\kappa(u, T_j) 
\end{align*}

The premium $\mathbf{PS}_0^j$ is totlaly invested in the level porftolio $G(n-j)$ so that the totla value of the profit and loss at time $T_j$ equals \[
	\frac{\mathbf{PS}^j_0}{G_0(n-j)} + \int_0^{T_j} \psi^j_u \, d \kappa(u, T_j)
\]  Taking derivatives, we can show that by setting $\psi^j_t = \Phi \left( \tilde d_+^j(t) \right)$ we obtain the replicating strategy for the $j$-th swaption.
% section valuation_of_swaptions_in_jamshidian_s_model (end)
% section payer_swaptions (end)
% chapter modelling_of_forward_swap_rates (end)
\end{document}

### 2011-12-12 AMH4 - Advanced Option Pricing/AMH4 Lecture Notes.tex
Initial Commit

% Created by Andrew Tulloch

%!TEX TS-program = xelatex
%!TEX encoding = UTF-8 Unicode

\documentclass[10pt, oneside, reqno]{amsart}
\usepackage{amsthm, amsmath, amssymb}
\usepackage{setspace, graphicx, enumerate}
% \usepackage{fullpage}
% \usepackage[landscape, scale=0.85]{geometry}
\onehalfspacing

% \usepackage{mathspec}
% \setallmainfonts{Minion Pro}

\usepackage{fontspec,xltxtra,xunicode}
\defaultfontfeatures{Mapping=tex-text}

% \setlength{\columnsep}{50pt}
% \setlength{\columnseprule}{2pt}

%\setromanfont[Mapping=tex-text,Contextuals= 
%{NoWordInitial,NoWordFinal,NoLineInitial,NoLineFinal}]{Hoefler Text}
%\setsansfont[Scale=MatchLowercase,Mapping=tex-text]{Hoefler Text}
%\setmonofont[Scale=MatchLowercase]{Andale Mono}

	% AMS Theorems
	\theoremstyle{plain}% default 
	\newtheorem{thm}{Theorem}[section] 
	\newtheorem{lem}[thm]{Lemma} 
	\newtheorem{prop}[thm]{Proposition} 
	\newtheorem*{cor}{Corollary}

\newcommand{\res}[2]{\text{Res}(#1,#2)}
	\theoremstyle{definition} 
		\newtheorem{defn}[thm]{Definition}
		\newtheorem{conj}[thm]{Conjecture}
		\newtheorem{exmp}[thm]{Example}

	\theoremstyle{remark} 
		\newtheorem*{rem}{Remark} 
		\newtheorem*{note}{Note} 
		\newtheorem{case}{Case}

		\newcommand{\expc}[1]{\mathbb{E}\left[#1\right]}
		\newcommand{\var}{\text{Var}}
		\newcommand{\cov}{\text{Cov}}
		\newcommand{\prob}[1]{\mathbb{P}(#1)}
		\newcommand{\given}{ \, | \,}
		\newcommand{\us}{0 \leq u \leq s}
		\newcommand{\ts}[1]{\{ #1 \}}

\renewcommand{\phi}{\varphi}
\newcommand{\sigf}{\mathcal{F}}

\newcommand{\dzz}{\, dz}
\newcommand{\bigo}[1]{\mathcal{O}(#1)}

\newcommand{\al}{\alpha}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\N}{\mathbb{N}}

\newcommand{\I}{\mathbb{I}}

\renewcommand{\P}{\mathbb{P}}

\newcommand{\F}{\mathbb{F}}
\newcommand{\Ga}{\mathbb{G}}

\newcommand{\aut}[1]{\text{Aut}{(#1)}}

\newcommand{\gener}[1]{\langle #1 \rangle}
\newcommand{\charr}[1]{\text{char}(#1)}
\newcommand{\nth}{n\textsuperscript{th}}

% \newcommand{\limsup}{\text{limsup}}

\newcommand{\tworow}[2]{\genfrac{}{}{0pt}{}{#1}{#2}}
\newcommand{\xdeg}[2]{[#1 : #2]}
\newcommand{\gal}[2]{\text{Gal}(#1/#2)}
\newcommand{\minpoly}[2]{m_{#1, #2}(x)}

\newcommand{\mapping}[5]{\begin{align*}
	#1 : \quad     #2 &\rightarrow #3 \\
			#4  &\mapsto #5
\end{align*}	
}

\def\cip{\,{\buildrel p \over \rightarrow}\,} 
\def\cid{\,{\buildrel d \over \rightarrow}\,} 
\def\cas{\,{\buildrel a.s. \over \rightarrow}\,}

\def\clp{\,{\buildrel L^p \over \rightarrow}\,}

\def\eqd{\,{\buildrel d \over =}\,} 
\def\eqas{\,{\buildrel a.s. \over =}\,}
\newcommand{\sigg}{\mathcal{G}}		
\newcommand{\indic}[1]{\mathbf{1}_{\{ #1 \}} }

\title{AMH4 - Advanced Option Pricing}								% Document Title
\author{Andrew Tulloch}
%\date{}                                           % Activate to display a given date or no date

\begin{document}
\maketitle
\section{Theory of Option Pricing} % (fold)
\label{sec:theory_of_option_pricing}
\begin{defn}[Brownian motion]
	A process $W_t$ is a $\P$-Brownian motion if it satisfies
	\begin{enumerate}
		\item $W_t$ is continuous with $W_0 = 0$ (a.s.)
		\item $W_t$ has stationary and independent increments.
		\item For any $t > 0$, $W_t \sim N(0, t)$ under the probability measure $\P$.
	\end{enumerate}
\end{defn}

\begin{thm}[Properties of conditional expectation]
	Assume we have a probability space $(\Omega, \P)$ and $\sigma$-algebras $\sigg, \sigg_1, \sigg_2$.  Assume that $\sigg_2 \subset \sigg_1$.  Then 
	\begin{enumerate}
		\item If $X$ is a random variable, then \[
			\E(X \given \sigg_2) = \E( \E(X \given \sigg_1) \given \sigg_2)
		\]
		\item If $Y$ is a $\sigg$-measurable random variable, then \[
			\E(XY \given \sigg) = Y \E(X \given \sigg)
		\]
	\end{enumerate}
\end{thm}
\begin{defn}[Martingale]
	A stochastic process $X_t$ is a $\sigf_t$-martingale if $\E(|X_t|) < \infty$ and \[
		X_s = \E(X_t \given \sigf_s)
	\] for all $s \leq t$.
\end{defn}

\begin{thm}[It\^o's lemma]
	If $F(X_t, t)$ is $C_{2,1}$ and $dX_t = \alpha_t \, dt + \beta_t \, dW_t$, then \[
		dF = (F_t + \alpha F_x + \frac{1}{2}\beta^2 F_{xx}) \, dt + \beta F_x \, dW_t 
	\]
\end{thm}

\begin{lem}[Product and Quotient rule]
	Let $X_t$ be an It\^o processes, so that \[
		dX_t = \alpha dt + \beta dW_t.
	\]  Let $F(X_t, t), G(X_t, t)$ be $C_{2, 1}$. Then \begin{align*}
		d(FG) = (F\, dG + G \, dF) + \beta^2 F_x G_x \, dt \\
		d(F/G) = \frac{G \, dF - F \, dG}{G^2} + \frac{\beta^2 G_x}{G^3}(FG_x - GF_x) \, dt
	\end{align*}
\end{lem}

\begin{lem}[It\^o isometry]
	If $\sigma_s \in L^2$, then \[
		\E(\int_0^t \sigma_s \, dW_s)^2 = \E(\int_0^t \sigma^2 \, ds)
	\]
\end{lem}

\begin{defn}[Local martingale]
	$X_t$ is a local martingale if there exists a sequence of stopping times $\nu_n$ such that for every $n$, the process $X^n_t = X_{\min(\nu_n, t)}$ is a martingale.  
\end{defn}

\begin{thm}[Martingale representation theorem]
	Let $\sigf_t$ be the natural filtration of a Brownian motion.  \begin{enumerate}
		\item Any progressively measurable process $\sigma_t$ satisfying \[\P(\int_0^t \sigma_s^2 \, ds) < \infty = 1 \quad \forall t \] the process \[
			t \mapsto \int_0^t \sigma_s \, dW_s
		\] is a local martingale. 
		\item If $X_t$ is an $L^2$ martingale, then there exists a progressively measurable process $\sigma_s$ such that \[
			X_t = \int_0^t \sigma_s \, dW_s
		\] 
	\end{enumerate}

	Hence the Brownian martingales (martingales with respect to the Brownian filtration) are essentially the It\^o integrals.
\end{thm}

\begin{thm}[Girsanov]
	Let $\lambda_t$ be progressively measurable with \[
		\E\exp(\frac{1}{2} \int_0^T \lambda^2(t) \, dt) < \infty
	\] Then there exists a measure $\P^\star$ such that \begin{enumerate}
		\item $\P^\star$ is equivalent to $\P$, 
		\item \[
		\frac{dP^\star}{dP} = \exp( - \int_0^t \lambda_t \, dW_t - \frac{1}{2} \int_0^t	\lambda^2_t \, dt )
		\]
		\item $W^\star_t = W_t + \int_0^t \lambda_s \, ds$ is a $\P^\star$-Brownian motion
	\end{enumerate}

	As a partial corollary, if $\P^\star$ is equivalent to $\P$ then there exists a progressively measurable process $\lambda_t$ such that \[
		W^\star_t = W_t + \int_0^t \lambda_s \, ds
	\] is a Brownian motion under $\P^\star$.  
\end{thm}

\begin{cor}
	We can then use Girsanov's theorem to transform a Brownian motion with drift to a martingale.  e.g.  Under $\P$, \begin{align*}
		dX_t 	&= \mu_t \, dt + \sigma_t \, dW_t \\
				&= \sigma_t d(W_t + \int_0^t \sigma_s^{-1} \mu_s \, ds) \\
				&= \sigma_t dW^\star_t
	\end{align*} where we set $\lambda_s = \sigma_s^{-1} \mu_s$ in Girsanov's theorem.
\end{cor}

\begin{thm}[Multivariate It\^o's lemma]
	Let $dX_{i,t} = \alpha_i \, dt + \beta_i \, dW_{i,t}$ with $W_{i,t}$ correlated Brownian motions.  Then if $F(X_{1,t}, \dots, X_{n,t}, t)$ is $C_{2,1}$, then \[
		dF = \left( F_t + \sum_{i=1}^n \alpha_i F_i + \frac{1}{2} \sum_{i=1}^n \sum_{j=1}^n \beta_i \beta_j \rho_{ij} F_{ij} \right) \, dt + \sum_{i=1}^n \beta_i F_i \, dW_i(t)
	\]
\end{thm}
% section theory_of_option_pricing (end)

\section{Black-Scholes PDE Method} % (fold)
\label{sec:black_scholes_pde_method}

\begin{thm}[Black-Scholes PDE]
	Let $f(X_t, t)$ represent the price of a contingent claim on an asset $X_t$, where $X_t$ is assumed to follow geometric Brownian motion.  Under certain assumptions, we can derive the Black-Scholes PDE, \[
		f_t = rf - r x f_x - \frac{1}{2} \sigma^2 x^2 f_{xx}
	\]  
\end{thm}

Solving the Black-Scholes PDE along with initial conditions and payoff at expiration yields the function $f(X_t, t)$ which gives the option value at any time $t$ and any underlying value $X_t$.

% section black_scholes_pde_method (end)

\section{Martingale method} % (fold)
\label{sec:martinag}
Consider a market with risky security $X_t$ and riskless security $B_t$.  
\begin{defn}[Contingent claim]
	A random variable $C_T : \Omega \rightarrow \R$, $\sigf_T$-measurable is called a contingent claim.  If $C_T$ is $\sigma(X_T)$-measurable it is \textbf{path-independent.}
\end{defn}
\begin{defn}[Strategy]
	Let $\alpha_t$ represent number of units of $X_t$, and $\beta_t$ represent number of units of $B_t$.  If $\alpha_t, \beta_t$ are $\sigf_t$-adapted, then they are strategies in our market model.  Our strategy value $V_t$ at time $t$ is \[
		V_t = \alpha_t X_t + \beta_t B_t
	\]
\end{defn}

\begin{defn}[Self-financing strategy]
	A strategy $(\alpha_t, \beta_t)$ is self financing if \[
		dV_t = \alpha_t \, dX_t + \beta_t \, dB_t
	\]  The intuition is that we make one investment at $t = 0$, and after that only rebalance between $X_t$ and $B_t$.  
\end{defn}

\begin{defn}[Admissible strategy]
	$(\alpha_t, \beta_t)$ is an \textbf{admissible strategy} if it is self financing and $V_t \geq 0$ for all $0 \leq t \leq T$.  
\end{defn}

\begin{defn}[Arbitrage]
	An arbitrage is an admissible strategy such that $V_0 = 0, V_T \geq 0$ and $\P(V_T > 0) > 0$.  
\end{defn}

\begin{defn}[Attainable claim]
	A contingent claim $C_T$ is said to be attainable if there exists an admissible strategy $(\alpha_t, \beta_t)$ such that $V_T = C_T$.  In this case, the portfolio is said to replicate the claim.  By the law of one price, $C_t = V_t$ at all $t$.
\end{defn}

\begin{defn}[Complete]
	The market is said to be \textbf{complete} if every contingent claim is attainable
\end{defn}

\begin{thm}[Harrison and Pliska]
	Let $\P$ denote the real world measure of the underlying asset price $X_t$. If the market is arbitrage free, there exists an equivalent measure $\P^\star$, such that the discounted asset price $\hat X_t$ and every discounted attainable claim $\hat C_t$ are $\P^\star$-martingales.  Further, if the market is complete, then $\P^\star$ is unique.  In mathematical terms,\[
		C_t = B_t \E^\star(B_T^{-1} C_T \given \sigf_t).
	\]

	$\P^\star$ is called the equivalent martingale measure (EMM) or the risk-neutral measure.  
\end{thm}
% section martinag (end)
\section{Monte Carlo methods} % (fold)
\label{sec:monte_carlo_methods}
\subsection{Method of antithetic variances} % (fold)
\label{sub:method_of_antithetic_variances}
Instead of simulating $X$, also simulate a random variable $Z$ with the same variance and expectation as $X$, but is negatively correlated with $X$.  Then take as $Y$ the random variable \[
	Y = \frac{X + Z}{2}
\] Obviously $\E(Y) = \E(X)$.  On the other side, we have \begin{align*}
\var(Y) &= \cov\left( \frac{X+Z}{2}, \frac{X + Z}{2} \right) \\
		&= \frac{1}{4} \var(X) + 2 \cov(X, Z) + \var(Z) \leq \frac{1}{2} \var(X)	
\end{align*}
  So we can reduce variance by a factor of two.

\subsection{Control variate method} % (fold)
\label{sub:control_variate}
\begin{thm}
	Suppose we seek to estimate $\theta = \E(Y)$ where $Y=h(X)$ is the outcome of a simulation.  Suppose that $Z$ is also an output of the simulation, and assume that $\E(Z)$ is known.  Let \[
	c = \frac{\text{Cov}(Y, Z)}{\text{Var}(Z)}.\tag{$\ddag$}
	\] Then \[
		\hat \theta_c = Y + c(\E(Z) - Z) \tag{$\dagger$}
	\] is an unbiased estimator of $\theta$, and if $\text{Cov}(Y, Z) \neq 0$, $\hat \theta_c$ has a lower variance than $\hat\theta = Y$, and indeed has the lowest variance for all estimators of the form \[
		\hat \theta_\gamma = Y + \gamma(\E(Z) - Z) 
	\]
\end{thm} 
\begin{proof}
	We have \[
		\text{Var}(\hat \theta_c) = \text{Var}(Y) + c^2 \text{Var}(Z) - 2c \, \text{Cov}(Y, Z). \tag{$\star$}
	\]  From elementary methods of calculus, we see that $\var{\hat\theta_c}$ is minimised at \[
		c = \frac{\text{Cov}(Y, Z)}{\text{Var}(Z)}
	\]  Substituting in this value for $c$ in ($\star$), we obtain \begin{align*}
		\text{Var}(\hat \theta_c) &= \var(Y)  - \frac{\cov(Y,Z)^2}{\var(Z)} \\
									&= \var(\hat \theta)  - \frac{\cov(Y,Z)^2}{\var(Z)}
	\end{align*}
	and thus we only need $\cov(Y,Z) \neq 0$ to obtain our variance reduction.

	In practice, we do not know $\text{Cov}(Y,Z)$.  Thus, we have to do a number of \emph{burn-in} simulations to generate $Y$ and $Z$, and then compute an estimate $\hat c$ to use in the full simulation. 
\end{proof}

% subsection control_variate (end)
% subsection method_of_antithetic_variances (end)
% section monte_carlo_methods (end)

\section{Numerical Simulation of Stochastic Differential Equations} % (fold)
\label{sec:numerical_simulation_of_stochastic_differential_equations}
\begin{thm}
	Let \[
		dX_t = a(t, X_t) \, dt + b(t, X_t) \, dB_t
	\]
	Assume $\E X_0 < \infty$.  $X_0$ is independent of $B_s$ and there exists a constant $c > 0$ such that \begin{enumerate}
		\item $|a(t, x)| + |b(t, x)| \leq C(1 + |x|)$.  
		\item $a(t, x), b(t, x)$ satisfy the Lipschitz condition in $x$, i.e.\[
			|a(t,x)- a(t, y)| + |b(t, x) - b(t, y)| \leq C|x - y|
		\] for all $t \in (0, T).  $
	\end{enumerate}  Then there exists a \textbf{unique (strong) solution}. 
\end{thm}

\begin{defn}[Strong convergence]
	A numberical scheme for solving an SE is said to converge with strong order $\gamma$, if for sufficiently small $\Delta$, we have \[
		\E(|X(T) - X_N|) \leq K_T \Delta^\gamma
	\]  This implies that the generated paths approximate the true paths of the SDE - and so one calls this path-wise convergence or strong convergence.  
\end{defn}

\begin{defn}[Weak convergence]
	A numerical scheme for solving an SDE is said to converge with weak order $\beta$ if for sufficiently small $\Delta$ and each polynomial $g$, we have \[
		|\E(g(X_T)) - \E(g(X_N))| \leq K_{g, T} \Delta^\beta
	\]  Note that strong convergence always implies weak convergence.

	Note also that strong convergence implies pathwise convergence.  This is true by Markov's inequality, we have \begin{align*}
		\P(|X_n - X(T)| \geq \Delta^{\beta/2}) &o\leq \frac{\E(|X_n - X(T)|)}{\Delta^{\beta/2}} \\
				&\leq C \frac{\Delta^{\beta}}{\Delta^{\beta/2}}
	\end{align*}
\end{defn}

\begin{note}
	\begin{enumerate}
		\item Weak convergence is basically convergence in distribution, but it has no path-wise properties.
		\item If terms like $\E(h(X_T))$ are computed via Monte Carlo, then the weak convergence concept is sufficient.  
		\item If the option is a path dependent option, then strong convergence is the right concept, as the payoff depends on the whole path, rather than the distribution of the terminal value of the stock.  
	\end{enumerate}
\end{note}

\begin{thm}[Euler-Maruyama scheme]
	\begin{align*}
		X_0 &= X(0) \\
		X_{n+1} = X_{n} + a(t_n, X_n) \Delta t_n + b(t_n, X_n) \Delta W_n
	\end{align*} where \begin{align*}
		\Delta t_n &= t_{n+1} - t_n \\
		\Delta W_n = W_{t_{n+1}} - W_{t_n}l
	\end{align*}

	Euler-Maruyama has \textbf{strong convergence order $\gamma = \frac{1}{2}$} and \textbf{weak convergence order $\beta = 1$}.
\end{thm}

\begin{thm}[Milstein scheme]
	Consider the homogenous scalar stochastic differential equation \[
		dX_t = a(X_t) \, dt + b(X_t) \, dW_t
	\]  
	\begin{align*}
		X_0 &= X(0) \\
		X_{n+1} &= X_n + a(X_n) \Delta t_n + b(X_n) \Delta W_n + \frac{1}{2} b'(X_n)b(X_n)((\Delta W_n)^2 - \Delta t_n)
	\end{align*}   
	One can prove that the Milsten scheme has \textbf{strong and weak convergence order $\gamma = 1$.}
\end{thm}
% section numerical_simulation_of_stochastic_differential_equations (end)
\section{Stochastic Optimal Control} % (fold)
\label{sec:stochastic_optimal_control}

\begin{defn}[Controlled stochastic differential equation]
	\begin{align*}
		dx(t) = f(t, x(t), u(t)) \, dt + \sigma(t,x(t), u(t)) \, dW(t)
	\end{align*} where $u(t, \omega) = u(t, x(t, \omega))$ is a stochastic process, known as the \textbf{control}.  
\end{defn}

\begin{defn}[Admissible control]
	A control $u$ is called admissible for the constraints if for every initial value $x_0 \in S$ the corresponding stochastic differential equation has a unique solution with $x(0) = x_0$ and $u(t, \omega) \in \mathcal U$ for all $t \in [0, \infty]$.  We denote the set of admissible controls with $\mathcal A$.  
\end{defn}

\begin{defn}[Stochastic optimal control problem]
	We seek to solve \[
		\max_{u \in \mathcal A} \E \left[ \int_0^T e^{-rt} B(t, x(t), u(t)) \, dt + e^{-rT}S(x(T)) \cdot \mathbf{1}_{T < \infty} \, dt\right]
	\] under the dynamic constraint \[
		dx(t) = f(t, x(t), u(t)) \, dt + \sigma(t, x(t), u(t)) \, dW(t)
	\] with initial condition $x(0) = x_0$, and discount rate $r > 0$.

	$B$ is called the benefit function, $S$ is called the final payoff, and the control $u$ is called the optimal control, and the optimal value is called the value of the problem.
\end{defn}

\begin{defn}[Value function] \[
	\max_{u \in \mathcal A} \E \left[ \int_t^T e^{-r(s-t)} B(s, x(s), u(s)) \, ds + e^{-r(T-t)}S(x(T)) \cdot \mathbf{1}_{T < \infty} \, dt \given x(t) = x \right]
	\] subject to \begin{align*}
		dx(s) &= f(s, x(s), u(s)) \, ds + \sigma(s, x(s), u(s)) \, dW(s) \\
		x(t) &= x
	\end{align*}

	Note that $V(0, x_0)$ is the value of the optimal control problem. $V(t, x)$ is the value of the problem, if we started at time $t$ with initial state $x$.  
\end{defn}

\begin{thm}[Hamilton-Jacobi-Bellman equation]
	Assume $T < \infty$.  Let $V: [0, T] \times S \rightarrow R$ be a $C_{1, 2}$ function and assume it satisfies the HJB equation \begin{align*}
		rV(t, x) - V_t(t, x) &= \max_{u \in \mathcal A} \left(B(t, x, u) + V_x(t, x) f(t, x, u(t)) + \frac{1}{2} \text{tr}(V_{xx}(t, x) \sigma(t, x, u) \sigma(t, x, u)^T) \right) \\
		V(T, x) &= S(x).
	\end{align*}  Let $\phi(t, x)$ be the set of maximisers of the right hand side and let $u^* \in \mathcal A$ such that $u^*(t, \omega) \in \phi(t, x(t, \omega))$ for all $t \in [ 0, T], \omega \in \Omega$. Then $u^*$ is the optimal control and $V$ is the value function for the stochastic optimal control problem.  
\end{thm}

\begin{thm}[Hamilton-Jacobi-Bellman equation, infinite time]
	Consider the time homogenous, infinite time horizon problem \[
		\max_{u \in \mathcal A} \E \left[ \int_0^\infty e^{-rt} B(x(t), u(t)) \, dt \right] 
	\] subject to \[
		dx(t) = f(x(t), u(t)) \, dt + \sigma(x(t), u(t)) \, dW_t.
	\] Then the value function is independent of $t$, and so $V(t,x) = V(x)$, and the optimal control is of the type $u(t, x) = u(x)$.  The HBJ equation in this case becomes the ODE \[
		rV(x) = \max_u \left( B(x, u) + V'(x) f(x, u) + \frac{1}{2} V''(x) \sigma(x, u)^2 \right)
	\] 
\end{thm}
% section stochastic_optimal_control (end)

\end{document}

### 2011-12-12 COMP 2907 - Algorithms and Complexity/COMP 2907 Lecture Notes.tex
Initial Commit

%  
%  COMP 2907 Exam Notes
%
%  Created by Andrew Tulloch on 2009-11-04.
%!TEX TS-program = xelatex
%!TEX encoding = UTF-8 Unicode

% \documentclass[10pt]{amsart}
\documentclass[10pt, oneside, reqno]{amsart}
\usepackage{amsthm, amsmath, amssymb}

\usepackage{geometry, setspace, enumerate}
\onehalfspacing                 
\usepackage[ruled,section]{algorithm}
\usepackage{algpseudocode}

% \usepackage{fontspec,xltxtra,xunicode}
% \defaultfontfeatures{Mapping=tex-text}

%\setromanfont[Mapping=tex-text,Contextuals= 
%{NoWordInitial,NoWordFinal,NoLineInitial,NoLineFinal}]{Hoefler Text}
%\setsansfont[Scale=MatchLowercase,Mapping=tex-text]{Hoefler Text}
%\setmonofont[Scale=MatchLowercase]{Andale Mono}

	% AMS Theorems
	\theoremstyle{plain}% default 
	\newtheorem{thm}{Theorem}[section] 
	\newtheorem{prob}[thm]{Problem}
	\newtheorem{question}[thm]{Question}

	\newtheorem{lem}[thm]{Lemma} 
	\newtheorem{prop}[thm]{Proposition} 
	\newtheorem*{cor}{Corollary}

	\theoremstyle{definition} 
		\newtheorem{defn}[thm]{Definition}
		\newtheorem{conj}[thm]{Conjecture}
		\newtheorem{exmp}[thm]{Example}

	\theoremstyle{remark} 
		\newtheorem*{rem}{Remark} 
		\newtheorem*{note}{Note} 
		\newtheorem{case}{Case}

\newcommand{\expc}[1]{\mathbb{E}\left[#1\right]}

\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\F}{\mathbb{F}}
\newcommand{\Ga}{\mathbb{G}}

\newcommand{\nth}{n\textsuperscript{th}}
\newcommand{\bigo}[1]{\mathcal{O}(#1)}

\setlength{\topmargin}{0.2cm}
\setlength{\footskip}{0.2cm}
\setlength{\hoffset}{-1cm}
\setlength{\voffset}{-2cm}

\title{COMP 2907 - Algorithms and Complexity \\  Exam Notes}								% Document Title
\author{Andrew Tulloch and Giles Gardam}
%\date{}                                           % Activate to display a given date or no date

\begin{document}
\maketitle

\section{Introduction} % (fold)
\label{sec:introduction}
\begin{algorithm}[H]
	\label{alg:dynamic_rising_trend}
	\caption{Propose-and-reject algorithm for the stable matching problem}
	\begin{algorithmic}[1]
		\Procedure{StableMatching}{$M,W$}
		\While{some man is free and hasn't proposed to every woman}
			\State Choose such a man $m$
			\State $w \leftarrow $ first woman on $m$'s list to whom $m$ has not yet proposed 
			\If{w is free}
				\State assign $m$ and $w$ to be engaged
			\ElsIf{w preferms $m$ to her fiance $m'$}
				\State assign $m$ and $w$ to be engaged, and $m'$ to be free
			\Else
				\State $w$ rejects $m$
			\EndIf
		\EndWhile
	\EndProcedure
	\end{algorithmic}
\end{algorithm}

\begin{thm}
	The Gale-Shapley algorithm yields a stable matching in $\bigo{n^2}$ time. The matching it returns is man-optimal (all men simultaneously get matched to the best possible woman over all stable matchings), and is hence independent of the order in which free men are chosen to propose.
\end{thm}

\begin{defn}[Efficieny]
	An algorithm is \emph{efficient} if its worst-case running time is polynomial.	
\end{defn}

\begin{defn}[$\mathcal{O}$, $\Theta$, $\Omega$]
	We have the following bounds on the running time of an algorithm $T(n)$.
	\begin{itemize}
		\item $T(n)$ is $\bigo{f(n)}$ if there exists $c > 0$ such that $T(n) \leq c f(n)$ for $n$ sufficiently large.
		\item $T(n)$ is $\Omega{f(n)}$ if there exists $c > 0$ such that $T(n) \geq c f(n)$ for $n$ sufficiently large.

		\item $T(n)$ is $\Theta(f(n))$ if it is both $\bigo{f(n)}$ and $\Omega(f(n))$.
	\end{itemize}
\end{defn}

\subsection{Graphs} % (fold)
\label{sub:graphs}

\begin{defn}[Adjacency list]
	An adjacency list is a node indexed array of lists.  It contains two representations of each edge.  Checking if $(u,v)$ is an edge takes $\bigo{\deg u}$ time, identifying all edges and space used are both $\bigo{n+m}$.  (Convention is that a graph has $n$ vertices and $m$ edges.)
\end{defn}

\begin{thm}[Trees]
	An undirected graph is a \textbf{tree} if it is connected and does not contain a cycle.

	Let $G$ be an undirected graph on $n$ nodes. Any two of the following statements imply the third.
	\begin{itemize}
		\item $G$ is connected.
		\item $G$ does not contain a cycle.
		\item $G$ has $n-1$ edges.
	\end{itemize}
\end{thm}

\begin{thm}[\textsc{bfs} algorithm]
	Explore outward from $s$ in all possible directions, adding nodes one ``layer'' at a time.  
	\begin{itemize}
		\item $L_0 = \{s\}$
		\item $L_{i} = $ all nodes that do not belong to an earlier layer, and that have an edge to a node in $L_{i-1}$.
	\end{itemize}
\end{thm}

\begin{thm}
	\textsc{bfs} runs in $\bigo{m+n}$ time if the graph is given by its adjacency representation.
\end{thm}

\begin{defn}[Bipartite graph]
	A graph is bipartite if the nodes can be coloured red and blue such that every edge has one red and one blue end.
\end{defn}

\begin{thm}
	A graph is bipartite if and only if it does not contain an odd length cycle.
\end{thm}

\begin{defn}[Strongly connected]
	A (directed) graph is strongly connected if for every pair of nodes $(u,v)$, there is a path from $u$ to $v$ and a path from $v$ to $u$.  Can be tested by \textsc{bfs} in $\bigo{m+n}$ time (exanding along `forward' edges in one pass and then along `backward' edges in the other).
\end{defn}

\begin{defn}[Directed acyclic graph]
	A directed acyclic graph is a directed graph that contains no directed cycles.  
\end{defn}

\begin{lem}
	A topological order is total ordering such that if an edge joins $u$ to $v$ then $u < v$.
	$G$ has a topological order if and only if $G$ is a directed acyclic graph.
\end{lem}

% subsection graphs (end)

% section introduction (end)

\section{Greedy Algorithm} % (fold)
\label{sec:greedy_algorithm}
\subsection{Interval scheduling} % (fold)
\label{sub:interval_scheduling}

\begin{algorithm}[H]
	\label{alg:dynamic_rising_trend}
	\caption{Greedy algorithm for interval scheduling \textbf{P}}
	\begin{algorithmic}[1]
		\Procedure{IntervalScheduling}{$J$}
			\State Sort jobs by finish times so that $f_1 \leq f_2 \leq \dots \leq f_n$.
			\State $A \gets \emptyset$
			\For{$j = 1$ to $n$}
				\If{job $j$ compatible with $A$}
					\State $A \gets A \cup \{j\}$
				\EndIf
			\EndFor
			\State \textbf{return} $A$
	\EndProcedure
	\end{algorithmic}
\end{algorithm}

\begin{thm}
	\textsc{IntervalScheduling} runs in $\bigo{n \log n}$ time.
\end{thm}
% subsection interval_scheduling (end)

\subsection{Interval Partitioning} % (fold)
\label{sub:interval_partitioning}

\begin{algorithm}[H]
	\label{alg:dynamic_rising_trend}
	\caption{Greedy algorithm for interval partitioning}
	\begin{algorithmic}[1]
		\Procedure{IntervalPartitioning}{$J$}
			\State Sort intervals by starting times so that $s_1 \leq s_2 \leq \dots \leq s_n$.
			\State $d \gets 0$
			\For{$j = 1$ to $n$}
				\If{lecture $j$ compatible with some classroom $k$}
					\State schedule lecture $j$ in classroom $k$
				\Else
					\State allocate a new classroom $d+1$
					\State schedule lecture $j$ in classroom $d+1$
					\State $d \gets d+1$
				\EndIf
			\EndFor
	\EndProcedure
	\end{algorithmic}
\end{algorithm}

% subsection interval_partitioning (end)

\subsection{Some simple greedy algorithms} % (fold)
\begin{itemize}
\item For jobs with fixed deadlines and durations to be scheduled on one machine, we can minimize the maximum lateness going by earliest deadline first.
\item For optimal caching, we evict the item that is requested farthest in the future.
\item To minimize the number of fuel stops, the truck driver's algorithm chooses the furthest possible stop at each step.
\item To find the best $k$-clustering, i.e. partition of nodes into $k$ sets which maximizes the minimum distance between nodes in different clusters, we can apply Kruskal's algorithm, ending when there are $k$ connected compnents.
\end{itemize}
% subsection some_simple_greedy_algorithms (end)
\subsection{Dijkstra's algorithm} % (fold)
\label{sub:dijkstra_s_algorithm}

Given a directed graph $G = (V,E)$ with non-negative edge weights, finds the shortest path from a node $s$ to a target node $t$ (or any arbitrary node).
\begin{itemize}
	\item Maintain a set of \textbf{explored nodes} $S$ for which we have determined the shortest path distance $d(u)$ from $s$ to $u$.
	\item Initialise $S = \{s\}, d(s) = 0$.
	\item Repeatedly choose unexplored node $v$ which minimises \[
		\pi(v) = \min_{(u,v): u \in S} d(u) + w_{uv},
	\]
	add $v$ to $S$, and set $d(v) = \pi(v)$.
\end{itemize}

\begin{thm}
	\textsc{dijkstra} runs in $\bigo{m \log n}$ with a binary heap.
\end{thm}
% subsection dijkstra_s_algorithm (end)

\subsection{Minimum spanning trees} % (fold)
\label{sub:minimum_spanning_trees}

\begin{defn}[Minimum spanning tree]
	Given a connected graph $G = (V,E)$ with real-valued edge weights $w_{uv}$, an \textsc{mst} is a subset of the edges $T \subseteq E$ such that $T$ is a spanning tree whose sum of edge weights is minimised.
\end{defn}

\begin{thm}[Kruskal's algorithm]
	Start with $T = \emptyset$.  Consider edges in ascending order of cost.  Insert edge $e$ in $T$ unless doing so would create a cycle.
\end{thm}

\begin{thm}[Reverse-Delete algorithm]
	Start with $T = E$.  Consider edges in descending order of cost.  Delete edge $e$ from $T$ unless doing so would disconnect $T$.
\end{thm}

\begin{thm}[Prim's algorithm]
	Start with some root node $s$ and greedily grow a tree from $s$ outward.  At each step, add the cheapest edge $e$ to $T$ that has exactly one endpoint in $T$.
\end{thm}

\begin{prop}[Cut property]
	Let $S$ be any subset of nodes, and let $e$ be the minimum cost edge with exactly one endpoint in $S$.  Then the \textsc{mst} contains $e$.
\end{prop}

\begin{proof} Assume the contrary. Adding $e$ creates a cycle. Some other edge on the cut is more expensive than $e$, and removing it leaves a cheaper spanning tree.
\end{proof}

\begin{prop}[Cycle property]
	Let $C$ be any cycle, and let $f$ be the maximum cost edge belonging to $C$.  Then the \textsc{mst} does not contain $f$.
\end{prop}

\begin{proof} Assume the contrary. Removing $f$ leaves two connected components, with one end of $f$ in each. Some other edge in the cycle is on the cut, and adding it gives us a cheaper spanning tree. 
\end{proof}

\begin{algorithm}[H]
	\label{alg:dynamic_rising_trend}
	\caption{Prim's algorithm for minimal spanning tree}
	\begin{algorithmic}[1]
		\Procedure{Prim}{$G, c$}
			\ForAll{$v \in V$} $a[v] \gets \infty$ \EndFor
			\State Initialise an empty priority queue $Q$
			\ForAll{$v \in V$} Insert $v$ onto $Q$ \EndFor
			\State Initialise a set of explored nodes $S \gets \phi$

			\While{$Q$ is not empty}
				\State $u \gets $ delete minimum element from $Q$.
				\State $S \gets S \cup \{u \}$
				\ForAll{edges $e = (u,v)$ incident to $u$}
					\If{$ v \notin S$ and $c_e < a[v]$}
						\State decrease priority $a[v]$ to $c_e$
					\EndIf

				\EndFor
			\EndWhile
	\EndProcedure
	\end{algorithmic}
\end{algorithm}

\begin{thm}
	\textsc{Prim} runs in $\bigo{n^2}$ with an array, $\bigo{m \log n}$ with a binary heap. 
\end{thm}

\begin{algorithm}[H]
	\label{alg:dynamic_rising_trend}
	\caption{Kruskal's algorithm for minimal spanning tree}
	\begin{algorithmic}[1]
		\Procedure{Kruskal}{$G, c$}
			\State Sort edge weights so that $c_1 \leq c_2 \leq \dots \leq c_m$.
			\State $T \gets \phi$
			\ForAll{$u \in V$} make a set containing singleton $u$.
			\EndFor

			\For{$i = 1$ to $m$}
				\State $(u,v) = e_i$
				\If{$u$ and $v$ are in different sets}
					\State $T \gets T \cup \{e_i \}$
					\State merge the sets containing $u$ and $v$
				\EndIf
			\EndFor
			\State \textbf{return} $T$
		\EndProcedure
	\end{algorithmic}
\end{algorithm}

% TODO Add a summary of the union-find data structure.

% subsection minimum_spanning_trees (end)

% section greedy_algorithm (end)

\section{Divide \& Conquer} % (fold)
\label{sec:divide_&_conquer}

\begin{thm}[Master theorem]
	If $T(n) \leq a T(n/b) + \bigo{n^d}$, then \[
		T(n) = \begin{cases}
			\bigo{n^d} & \text{if $a < b^d$}\\
			\bigo{n^d \log n} &\text{if $a = b^d$}\\
			\bigo{n^{\log_b a}} & \text{if $ a  > b^d$}
		\end{cases}
	\]
\end{thm}
\begin{algorithm}[H]
	\label{alg:dynamic_rising_trend}
	\caption{Counting inversions}
	\begin{algorithmic}[1]
		\Procedure{SortAndCount}{$L$}
			\If{list $L$ has one element}
				\State \textbf{return} 0 and the list $L$
			\EndIf

			\State Divide $L$ into two halves $A$ and $B$
			\State $(r_A, A) \gets \textsc{SortAndCount}(A)$
			\State $(r_B, A) \gets \textsc{SortAndCount}(B)$
			\State $(r, L) \gets \textsc{MergeAndCount}(A,B)$

			\State
			\State \textbf{return} $r = r_A + r_B + r$ and the sorted list $L$
	\EndProcedure
	\end{algorithmic}
\end{algorithm}

\subsection{Closest pair of points} % (fold)
\label{sub:closest_pair_of_points}

\begin{algorithm}[H]
	\label{alg:dynamic_rising_trend}
	\caption{Finding the closest pair of points in a plane}
	\begin{algorithmic}[1]
		\Procedure{ClosestPair}{$L$}
			\State \textbf{Compute} separation line $X$ such that half the points are on one side and half on the other side.
			\State 
			\State $\delta_1 = \textsc{ClosestPair}(\text{left half})$
			\State $\delta_2 = \textsc{ClosestPair}(\text{right half})$
			\State $\delta = \min(\delta_1, \delta_2)$

			\State
			\State \textbf{Delete} all points further than $\delta$ from separation line $X$
			\State \textbf{Sort} remaining points by $y$-coordinate

			\State \textbf{Scan} points in $y$ order and compare distance between each point and next 11 neighbours.  If any of these distances is less than $\delta$, update $\delta$

			\State 
			\State \textbf{return} $\delta$
	\EndProcedure
	\end{algorithmic}
\end{algorithm}

\begin{thm}
	Running time is $T(n) \leq 2 T(n/2) + \bigo{n \log n} \Rightarrow T(n) = \bigo{n \log^2 n}$
\end{thm}

\begin{rem}
	Can be improved to run in $\bigo{n \log n}$ by pre-sorting and merging lists.
\end{rem}
% subsection closest_pair_of_points (end)

\subsection{Multiplication} % (fold)
\begin{thm} We can multiply two $n$-bit integers in $O(n^{\log_2 3})$ bit operations.
\end{thm}
\begin{rem} This is achieved recursively, putting $x=2^{n/2}x_1+x_0$, $y=2^{n/2}y_1+y_0$ and computing $xy = 2^n x_1 y_1 + 2^{n/2} \left( (x_1+x_0)(y_1+y_0) - x_1y_1 - x_0y_0) \right) + x_0 y_0$, so that the expensive computation is computing $3$ multiplications of two order $n/2$ integers; the shifts and additions take linear time. Note that $\log_2 3 =1.58496...$
\end{rem}
\label{sub:multiplication}

\begin{thm}
	Naive matrix multiplication is $\bigo{n^3}$.  Can be improved to $\bigo{n^{2.81}}$ by Strassen, and has been improved to $\bigo{n^{2.376}}$ by Coppersmith-Winograd.
\end{thm}
% subsection multiplication (end)

% section divide_&_conquer (end)

\section{Dynamic Programming} % (fold)
\label{sec:dynamic_programming}

\subsection{Weighted interval scheduling} % (fold)
\label{sub:weighted_interval_scheduling}
The following algorithm solves the weighted interval scheduling problem.
The jobs must be sorted in ascending order by finishing time.
We use the recurrence \[
	f(j) = \begin{cases}
		0 &\text{if $j = 0$} \\
		\max \{v_j + f(p(j)), f(j-1) \} &\text{otherwise}
	\end{cases}
\] where $p(j)$ is the largest index $i < j$ such that job $i$ is compatible with job $j$, $0$ if there is no such $i$.

\begin{algorithm}[H]
	\label{alg:dynamic_rising_trend}
	\caption{Weighted interval scheduling using dynamic programming}
	\begin{algorithmic}[1]
		\Procedure{WeightedInterval}{$L$}
			\State Sort jobs by finish times so that $f_1 \leq f_2 \leq \dots \leq f_n$
			\ForAll{$i = 1$ to $n$} 
			\State Compute $p(i)$ 
			\EndFor

			\State

			\ForAll{$j = 1$ to $n$}
				\State $M[j] \gets \textsc{null}$
			\EndFor

			\Procedure{ComputeOpt}{$j$}
			\If{$M[j]$ is \textsc{null}}
				\State $M[j] = \max(w_j + \textsc{ComputeOpt}(p(j)), \textsc{ComputeOpt}(j-1))$
			\EndIf
			\State \textbf{return} $M[j]$
			\EndProcedure

		\State \textbf{return} \textsc{ComputeOpt}($n$)
		\EndProcedure
	\end{algorithmic}
\end{algorithm}

\begin{thm}
	\textsc{WeightedInterval} runs in $\bigo{n \log n}$ time.
\end{thm}

\begin{rem}
The runtime of \textsc{WeightedInterval} is $\bigo{n}$ if the jobs are already sorted by finish time and start time. We need them sorted by start time to compute the $p(j)$ in $\bigo{n}$, so we can just do one pass over the lists. Otherwise we get a $n \log n$ from doing repeated binary searches.
\end{rem}
% subsection weighted_interval_scheduling (end)

\subsection{Segmented least squares} % (fold)
\label{sub:segmented_least_squres}

We use the recursion 
\[
	f(j) = \begin{cases}
		0 &\text{if $j = 0$} \\
	\displaystyle c + \min_{1 \leq i \leq j} \left\{ e(i,j) + f(i-1)\right\} &\text{otherwise}			\end{cases}
\] where $f(j)$ is the minimum cost for points $p_1, p_2, \dots, p_j$, and $e(i,j)$ is the minimum sum of squares for points $p_i, p_{i+1},\dots, p_j$.

\begin{thm}
	Computing segmented least squares runs in $\bigo{n^3}$ time.  The bottleneck is computing $e(i,j)$ for all $n^2$ pairs. This can in fact be done in $\bigo{n^2}$ by trickery (it involves evaluating the $e(i,j)$ in increasing order of $j - i$), and brings the whole algorithm down to $\bigo{n^2}$.
\end{thm}
% subsection segmented_least_squres (end)

\subsection{Knapsack problem} % (fold)
\label{sub:knapsack_problem}

\begin{defn}
	$f(i,w)$ is the maximum profit subset of items $1, \dots, i$ with weight limit $w$.  
\end{defn}

We then have the following recursion:
\[
	f(i,w) = \begin{cases}
		0 		&\text{if $i = 0$}\\
		f(i-1,w)&\text{if $w_i > w$}\\
		\max \left\{ f(i-1,w), v_i + f(i-1, w - w_i)\right\} &\text{otherwise}
	\end{cases}
\]

\begin{thm}
	The dynamic programming algorithm for the knapsack problems runs in time $\bigo{nW}$ (only pseudo-polynomial).
\end{thm}

\begin{rem}
The decision version of knapsack is NP-complete. There is however a polynomial-time approximation algorithm that finds a feasible solution within $0.01\%$ of the optimum.
\end{rem}
% subsection knapsack_problem (end)

\subsection{RNA secondary structure} % (fold)
\label{sub:}

\begin{defn}
	$f(i,j)$ is the maximum number of base pairs in a secondary structure of the substring $x_i x_2\dots x_j$. Let $g(i, j)$ be the maximum number of base pairs in the case that $x_j$ is aligned with some element. Then $$g(i, j) = 1 + \max_{\substack{t:i \leq t < j-4 \\ b_t, b_j \text{complementary}}} f(i,t-1) + f(t+1, j-1).$$
\end{defn}

We use the recursion:
\[
	f(i,j) = \begin{cases}
		0 &\text{if $i \geq j-4$}\\
		\max \{f(i,j-1), g(i, j) \}
	\end{cases}
\]

% subsection  (end)

\subsection{Sequence alignment} % (fold)
\label{sub:sequence_alignment}

\begin{defn}
	$f(i,j)$ is the minimum cost of aligning strings $x_1 x_2\dots x_i$ and $y_1 y_2 \dots y_j$
\end{defn}

We then have the recursion:
\[
	f(i,j) = \begin{cases}
		j \delta 		&\text{if $i = 0$}\\
		i \delta 		&\text{if $j = 0$}\\
		\min \left\{ \alpha_{x_i,y_j} + f(i-1,j-1), \right. & \\
		\ \ \left. \delta + f(i-1,j), \delta + f(i,j-1) \right\} &\text{otherwise} % apologies for that horrific formatting...
	\end{cases}
\]

\begin{thm}
	The naive implementation of the above algorithm can be implemented in $\Theta(mn)$ time and space  ($m$ and $n$ are lengths of given strings).
\end{thm}

\begin{lem}
	By calculating the recurrence iteratively (not recursively), iterating over $i$ and only storing values of $f(i, j)$ for the current $i$ and $i-1$ , we can find the optimal \textbf{value} in $\bigo{m+n}$ space and $\Theta(m n)$ time. It is not possible to reconstruct the alignment this way however.
\end{lem}

\begin{rem}
	By divide and conquer, we can find the optimal alignment in linear space as well. Essentially, we determine where the shortest path in the edit distance graph intersects the middle column by iterating over the elements $(i, \lfloor \frac{n}{2} \rfloor)$ in it, summing shortest distances to the two corners $(0, 0)$ and $(m, n)$. We then recurse on the left and right subproblems.
\end{rem}
% subsection sequence_alignment (end)

\subsection{The Bellman-Ford algorithm for shortest paths} % (fold)
\label{sub:the_bellman_ford_algorithm_for_shortest_paths}

Dijkstra's algorithm failed in graphs with negative edge costs.  To remedy this, we introduce the Bellman-Ford algorithm.

\begin{defn}
	$f(i,v)$ is the length of the shortest $v-t$ path $P$ using at most $i$ edges.
\end{defn}

Then we have the recursion:
\[
	f(i,v) = \begin{cases}
		0 						&\text{if $i = 0$ and $v=t$}\\
		\infty 					&\text{if $i = 0$ and $v \neq t$}\\
		\min \left\{ f(i-1,v), \displaystyle\min_{(v,w)\in E} \left\{f(i-1,w) + c_{vw}\right\} \right\} &\text{otherwise}
	\end{cases}
\]

By only maintaining one array $M[v]$ equal to the shortest $v-t$ path we have found so far, we can reduce the running time to $\bigo{mn}$ and space to $\bigo{m+n}$.
No need to check edges of the form $(v,w)$ unless $M[w]$ changed in the previous iteration.

% subsection the_bellman_ford_algorithm_for_shortest_paths (end)

\subsection{Negative cycles in a graph} % (fold)
\label{sub:negative_cycles_in_a_graph}

\begin{lem}
	If $f(n,v) = f(n-1,v)$ for all $v$, then there are no negative cycles.
\end{lem}

\begin{lem}
	If $f(n,v) < f(n-1,v)$ for some node $v$, then any shortest path from $v$ to $t$ contains a cycle $W$.  Moreover, $W$ has negative cost.
\end{lem}

\begin{thm}
	We can detect negative cost cycles in $\bigo{mn}$ time. 
	\begin{itemize}
		\item Add a new node $t$ and connect all nodes to $t$ with a zero-cost edge.
		\item Check if $f(n,v) = f(n-1,v)$ for all nodes $v$.
		\begin{itemize}
			\item If yes, then there are no negative cycles.
			\item If no, then extract a cycle from shortest path from $v$ to $t$.
		\end{itemize}
	\end{itemize}
\end{thm}
% subsection negative_cycles_in_a_graph (end)

% section dynamic_programming (end)

\section{Network Flow} % (fold)
\begin{defn}[$s-t$ cut]
	An $s-t$ cut is a partition ($A,B$) of $V$ with $s \in A$ and $t \in B$.
\end{defn}

\begin{defn}
The capacity of a cut $(A,B)$ is $\text{cap}(A,B) = \displaystyle\sum_{\text{$e$ out of $A$}}c(e)$
\end{defn}

\begin{prob}[Min $s-t$ cut problem]
Find an $s-t$ cut of minimum capacity.
\end{prob}

\begin{defn}[Flow]
	An $s-t$ flow is a function that satisfies:
	\begin{itemize}
		\item For each $e \in E: 0 \leq f(e) \leq c(e)$
		\item For each $v \in V - \{s,t\}:  \displaystyle \sum_{\text{$e$ into $v$}}f(e)  = \displaystyle \sum_{\text{$e$ out of  $v$}}f(e)$
	\end{itemize}

	The value of a flow $f$ is $v(f) = \displaystyle \sum_{\text{$e$ out of $s$}}f(e)$
\end{defn}

\begin{prob}[Max flow problem]
	Find the $s-t$ flow of maximum value.
\end{prob}

\newcommand{\sumsub}[1]{\displaystyle\sum_{\text{#1}}}
\begin{lem}[Flow value lemma]
	Let $f$ be any flow, and let $(A,B)$ be any $s-t$ cut.  Then, the net flow sent across the cut is equal to the amount leaving $s$.\[
		\sumsub{$e$ out of $A$} f(e) - \sumsub{$e$ into $A$}f(e) = v(f)
	\]
\end{lem}

\begin{thm}[Weak duality]
	Let $f$ be any flow, and let $(A,B)$ be any $s-t$ cut.  Then the value of the flow is at most the capacity of the cut, that is $v(f) \leq \text{cap}(A,B)$. 
\end{thm}

\begin{cor}
	Let $F$ be any flow, and let $(A,B)$ be any cut.  If $v(f) = \text{cap}(A,B)$, then $f$ is a max flow and $(A,B)$ is a min cut.
\end{cor}

\begin{thm}[Max-flow Min-cut theorem]
	The value of a max flow is equal to the value of the min cut.
\end{thm}

\begin{defn}[Augmenting path]
	\textbf{Original edge} $e = (u,v) \in E$.  Flow $f(e)$, capacity $c(e)$.

	\textbf{Residual edge}
	\begin{itemize}
		\item ``Undo'' flow sent. 
		\item $e = (u,v)$ and $e^R = (v,u)$
		\item Residual capacity given by \[
			c_f(e) = \begin{cases}
				c(e) - f(e) &\text{if $e \in E$},\\
				f(e) 		&\text{if $e^R \in E$}
			\end{cases}
		\]
	\end{itemize}

	\textbf{Residual graph} $G_f = (V, E_f)$.  Consists of the residual edges with positive residual capacity.\[
		E_f = \{ e:f(e) < c(e)\} \cup \{ e^R : c(e) > 0\}
	\]
\end{defn}

	\begin{algorithm}[H]
		\label{alg:dynamic_rising_trend}
		\caption{Augments a path \textbf{P}}
		\begin{algorithmic}[1]
					\Procedure{Augment}{$f, c, P$}
			\State $b \leftarrow \text{bottleneck}(P)$			
			\ForAll{$e \in P$}
				\If{$e \in E$}  $f(e) \leftarrow f(e) + b$
				\Else{}  $f(e^R) \leftarrow f(e) - b$
				\EndIf
			\EndFor

			\State \textbf{return} $f$
		\EndProcedure
		\end{algorithmic}
	\end{algorithm}

\begin{algorithm}[H]
	\label{alg:dynamic_rising_trend}
	\caption{Finds the maximum flow}
	\begin{algorithmic}[1]
	\Procedure{FordFulkerson}{$G, s, t, c$}

		\ForAll{$e \in E$} $f(e) \leftarrow 0$
		\EndFor
		\State $G_f \leftarrow \text{residual graph}$
		\While{there exists an augmenting path \textbf{P}}
			\State $f \leftarrow \textsc{Augment}(f, c, P)$
			\State update $G_f$
		\EndWhile
		\State \textbf{return} $f$
	\EndProcedure
	\end{algorithmic}
\end{algorithm}

\begin{thm}[Running time of the above algorithm]
	Under the assumption that all capacities are integers between $1$ and $C$, we have that the algorithm terminates in at most $v(f^\star) \leq nC$ iterations.   
\end{thm}

\begin{cor}
	If $C = 1$, \textsc{FordFulkerson} runs in $\bigo{mn}$ time
\end{cor}

\begin{cor}
	The generic \textsc{FordFulkerson} algorithm is not polynomial in input size ($m,n, \log C$). Instances can be constructed which do take $\Theta(nC)$ iterations.
\end{cor}

\begin{thm}[Capacity Scaling algorithm for max-flow]
	The following algorithm finds the max flow in $\bigo{m \log C}$ augmentations.

	The intuition is as follows.
	\begin{itemize}
		\item Don't worry about finding exact highest bottleneck path.
		\item Maintain a scaling parameter $\Delta$.  
		\item Let $G_f(\Delta)$ be the subgraph of the residual graph consisting of only arcs with capacity at least $\Delta$.
	\end{itemize}

\end{thm}
\begin{algorithm}[H]
	\label{alg:dynamic_rising_trend}
	\caption{Capacity scaling algorithm}
	\begin{algorithmic}[1]
	\Procedure{ScalingMaxFlow}{$G, s, t, c$}

		\ForAll{$e \in E$} $f(e) \leftarrow 0$
		\EndFor
		\State $\Delta \leftarrow \text{smallest power of $2$ greater than or equal to $C$}$
		\State $G_f \leftarrow \text{residual graph}$

		\While{$\Delta \geq 1$}
			\State $G_f(\Delta) \leftarrow \Delta-\text{residual graph}$
			\While{ there exists an augmenting path \textbf{P} in $G_f(\Delta)$}
				\State $f \leftarrow \textsc{Augment}(f, c, P)$
				\State update $G_f(\Delta)$
			\EndWhile
			\State $\Delta \leftarrow \frac{\Delta}{2}$
		\EndWhile
		\State \textbf{return} $f$
	\EndProcedure
	\end{algorithmic}
\end{algorithm}

\begin{lem}
	The outer loop repeats $1 + \log_2{C}$ times, as $\Delta$ decreases by a factor of $2$ each iteration.
\end{lem}

\begin{lem}
	Let $f$ be the flow at the end of a $\Delta$ scaling phase.  Then the value of the maximum flow is at most $v(f) + m \Delta$. 
\end{lem}

\begin{lem}
	There are at most $2m$ augmentations per scaling phase.
\end{lem}
\begin{thm}
	The \textsc{ScalingMaxFlow} algorithm finds a max flow in $\bigo{m \log C}$ augmentations.  It can be implemented to run in $\bigo{m^2 \log C}$ time.
\end{thm}

\subsection{Bipartite matching} % (fold)
\label{sub:bipartite_matching}
\begin{prob}[Bipartite matching problem]
	Consider the following.
	\begin{itemize}
		\item Input: undirected, \textbf{bipartite} graph $G = (L \cup R, E)$.
		\item $M \subseteq E$ is a \textbf{matching} if each node appears in at most one edge in $M$.
		\item Max matching: find a maximum cardinality matching.
	\end{itemize}
\end{prob}

Convert the problem to a maximum flow problem.
\begin{itemize}
	\item Create digraph $G' = (L \cup R \cup \{s,t\}), E')$.
	\item Direct all edges from $L$ to $R$, and assign unit capacity.
	\item Add sources $s$ and unit capacity edges from each node in $R$ to $t$.
\end{itemize}

Then we have the result:
\begin{thm}
	The maximum cardinality matching in $G$ is equal to the value of the maximum flow in $G'$.
\end{thm}

\begin{defn}[Perfect matching]
	A matching $M \subseteq E$ is \textbf{perfect} if each node appears in exactly one edge in $M$.
\end{defn}

\begin{thm}
	Let $G = (L \cup R, E)$ be a bipartite graph with $|L| = |R|$.  Then $G$ has a perfect matching if and only if $|N(S)| \geq |S|$ for all subsets $S \subseteq L$ ($N(S)$ is the set of nodes adjacent to nodes in $S$).
\end{thm}
% subsection bipartite_matching (end)

\subsection{Disjoint paths} % (fold)
\label{sub:disjoint_paths}

\begin{prob}[Edge disjoint paths]
	Given a directed graph $G = (V,E)$ and two nodes $s$ and $t$, find the maximum number of edge disjoint $s-t$ paths
\end{prob}

\begin{defn}
	Two paths are \textbf{edge-disjoint} if they have no edge in common.
\end{defn}

\begin{thm}[Solution]
	Assign unit capacity to every edge.  Then the maximum number of edge disjoint $s-t$ paths equals the maximum flow value.
\end{thm}

\begin{prob}[Network connectivity]
	Given a directed graph $G = (V,E)$ and two nodes $s$ and $t$, find the minimum number of edges whose removal disconnects $t$ from $s$.
\end{prob}

\begin{defn}
	A set of edges $F \subseteq E$ disconnects $t$ from $s$ if all $s-t$ paths uses at least one edge in $F$.
\end{defn}

\begin{thm}
	The maximum umber of edge disjoint $s-t$ paths is equal to the minimum number of edges whose removal disconnects $t$ from $S$.  
\end{thm}

% subsection disjoint_paths (end)

\subsection{Circulation with demands and lower bounds} % (fold)
\label{sub:circulation_with_demands_and_lower_bounds}

\begin{defn}[Circulation]
	A \textbf{circulation} is a function that satisfies 
	\begin{itemize}
		\item For each $e \in E: 0 \leq f(e) \leq c(e)$,

		\item For each $v \in V: \sumsub{$e$ into $v$} f(e) - \sumsub{$e$ out of $v$}f(e) = d(v)$.
	\end{itemize}
\end{defn}

\begin{thm}
	Necessary condition - sum of supplies equals the sum of demands.
	\[
		\sumsub{$v: d(v) > 0$} d(v) = \sumsub{$v: d(v) < 0$} -d(v) = D
	\]
\end{thm}

\begin{thm}
	Formulate as a circulation problem.
	\begin{itemize}
		\item Add new sources $s$ and sink $t$.
		\item For every $v$ with $d(v) > 0$, add edge $(s,v)$ with capacity $-d(v)$.
		\item For every $v$ with $d(v) < 0$, add edge $(v,t)$ with capacity $d(v)$
	\end{itemize}

	Then we have the following theorem.
	\begin{thm}
		$G$ has a circulation if and only if $G'$ has a maximum flow over value $D$.
 	\end{thm}
\end{thm}

\begin{cor}
	Given $G = (V,E,c,d)$, there does not exist a circulation if and only if there exists a partition $(A,B)$ of $G$ such that $\sum_{v \in B} d(v) > \text{cap}(A,B)$.
\end{cor}

\begin{thm}
	When formulating a circulation problem with lower bounds, we do the following.
	\begin{itemize}
		\item For every $e \in E$, end $l(e)$ units of flow along edge $e$.
		\item Update demands of both endpoints.
	\end{itemize}
\end{thm}
\begin{thm}
	There exists a circulation in $G$ if and only if there exists a circulation in $G'$.  If all demands, capacities and lower bounds in $G$ are integers, then there is a circulation in $G$ that is integer valued.
\end{thm}

% subsection circulation_with_demands_and_lower_bounds (end)
\subsection{Survey design} % (fold)
\label{sub:survey_design}

\begin{prob}[Survey design] We must
	\begin{itemize}
		\item Design survey asking $n_1$ customers about $n_2$ products.
		\item Can only survey customer $i$ about product $j$ if the own it.
		\item Ask customer $i$ between $c_i$ and $c_i'$ questions.
		\item Ask between $p_j$ and $p_j'$ customers about product $j$.
	\end{itemize}

	Design a survey that meets these specifications, if possible.
\end{prob}

\begin{thm}
	Formulate as a circulation problem with lower bounds.
	\begin{itemize}
		\item Include an edge $i_j$ if customer owns product $i$.  
		\item Set $L = \text{set of customers}, R = \text{set of products}$.
		\item Set $s \rightarrow i \in L$ with $f(e) = [c_i, c_i']$.
		\item Set $j \in R \rightarrow t$ with $f(e) = [p_j, p_j']$
	\end{itemize}
\end{thm}

% subsection survey_design (end)

\subsection{Project selection} % (fold)
\label{sub:project_selection}

\begin{prob}[Project selection]
	Consider the following.
	\begin{itemize}
		\item A set \textbf{P} of possible projects.  Project $v$ has revenue $p_v$.
		\item Set of prerequisites $E$.  If $(v,w) \in E$, then we must do project $w$ to do project $v$.  
	\end{itemize}

	Choose a feasible subset of projects to maximise revenue.
\end{prob}

\begin{defn}[Prerequisite graph]
 	A graph on the nodes of projects, with the following properties:
\begin{itemize}
	\item Include an edge from $v$ to $w$ if we must do project $w$ to do project $v$.  
	\item Assign capacity $\infty$ to all prerequisite edges.
	\item Add an edge $(s,v)$ with capacity $p_v$ if $p_v > 0$.
	\item Add an edge $(v,t)$ with capacity $-p_v$ if $p_v < 0$.
	\item Define $p_s = p_t = 0$.  
\end{itemize}
\end{defn}

\begin{thm}
	$(A,B)$ is a minimum cut if and only if $A - \{s\}$ is the optimal set of projects.
\end{thm}

% subsection project_selection (end)

% section network_flow (end)

\section{Reductions} % (fold)
\label{sec:reductions}
\begin{defn}[Polynomial time reduction]
	Problem $X$ reduces to problem $Y$ if arbitrary instances of problem $X$ can be solved using:
	\begin{itemize}
		\item Polynomial number of standard computational steps, plus
		\item Polynomial number of calls to an oracle that solves problem $Y$.
	\end{itemize}

	We denote this as $X \leq_p Y$
\end{defn}

\begin{note}
	We can use this to classify problems according to \textbf{relative} difficulty.  
\end{note}

\begin{lem}
	If $X \leq_p Y$ and $Y$ can be solved in polynomial time, then $X$ can also be solved in polynomial time.
\end{lem}

\begin{note}
	If $X \leq_p Y$ and $Y \leq_p X$, then we use the notation $X \equiv_p Y$.  In fact the binary relation $\leq_p$ is a partial order on the set of all problems, because it is reflexive, anti-symmetric, and transitive.
\end{note}

There are three basic strategies for polynomial time reduction.

\begin{itemize}
	\item Reduction by simple equivalence.
	\item Reduction from special case to general case.
	\item Reduction by encoding with gadgets.
\end{itemize}

\begin{prob}[\textsc{IndependentSet}]
	Given a graph $G = (V,E)$ and an integer $k$, is there a subset of vertices $S \subseteq V$ such that $|S| \geq k$ and for each edge, at most one of its endpoints is in $S$?
\end{prob}

\begin{prob}[\textsc{VertexCover}]
	Given a graph $G = (V,E)$ and an integer $k$, is there a subset of vertices $S \subseteq V$ such that $|S| \leq k$ and for each edge, at least one of its endpoints is in $S$?
\end{prob}

\begin{thm}
	We claim \textsc{VertexCover} $\equiv_p$ \textsc{IndependentSet}. 
\end{thm}

\begin{proof}
	$S$ is an independent set if and only if $V - S$ is a vertex cover.
\end{proof}

\begin{prob}[\textsc{SetCover}]
	Given a set $U$ of elements, a collection $S_1, S_2,\dots, S_m$ of subsets of $U$, and an integer $k$, does there exists a collection of at most $\leq k$ of these sets whose union is equal to $U$?
\end{prob}

\begin{thm}
	\textsc{VertexCover} $\leq_p$ \textsc{SetCover}
\end{thm}
\begin{proof}
	Given a \textsc{VertexCover} instance $G = (V,E), k$, we can construct a set cover instance whose size equals the size of the vertex cover instance. Set $k = k$, $U = E, S_v = \{e \in E : e \text{ incident to } v\}$.
	Then there is a set cover of size $\leq k$ if and only if there is a vertex cover of size $\leq k$.
\end{proof}

\begin{thm}
	\textsc{3-sat} $\leq_p$ \textsc{IndependentSet}
\end{thm}
\begin{proof}
	Given an instance $\Phi$ of \textsc{3-sat}, we construct an instance $(G,k)$ of \textsc{IndependentSet} that has an independent set of size $k$ if and only if $\Phi$ is satisfiable.  
	\begin{itemize}
		\item In each clause, connect all literals in a triangle.  
		\item Connect literal to each of its negations.
	\end{itemize}
\end{proof}

\begin{thm}[Self-reducibility]Consider the following.
	\begin{itemize}
		\item \textbf{Decision problem.}  Does there exist a vertex cover of size $\leq k$?

		\item \textbf{Search problem.}  Find vertex cover of minimum cardinality.

	\end{itemize}

		By self-reducibility, we have that the search problem reduces to the decision problem. That is, search problem $\leq_p$ decision problem.  This applies to all (\textbf{NP}-complete) problems in this chapter. This is dependent on the number of possible values of the optimal (you cannot binary search an infinite set).
\end{thm}

% section reductions (end)

\section{Complexity} % (fold)
\label{sec:complexity}

\begin{defn}[Complexity class \textbf{P}]
	The class of decision problems for which there is a polynomial time algorithm.
\end{defn}

\begin{defn}[Certifier]
	An algorithm $C(s,t)$ is a certifier for a problem $X$ if for every string $s$, $s \in X$ if and only if there exists a string $t$ such that $C(s,t)$ outputs \textsc{true}.  The string $t$ is known as a \emph{certificate} or \emph{witness}.  
\end{defn}

\begin{defn}[Complexity class \textbf{NP}]
	The class of decision problems for which there is a polynomial time certifier - an algorithm $C(s,t)$ that is polynomial in time and $|t| \leq p(|s|)$ for some polynomial \textbf{P}.
\end{defn}

\begin{rem}
	\textbf{NP} stands for \textbf{nondeterministic} polynomial time.
\end{rem}

\begin{exmp}[Certifiers and Certificates - \textsc{sat}]
	Consider the problem \textsc{sat} - given a CNF (conjunctive normal form - the `\emph{and} combination of many \emph{or} clauses') formula $\Phi$, is there a satisfying assignment?
	\begin{itemize}
		\item \textbf{Certificate.} An assignment of truth values to the $n$ boolean variables.
		\item \textbf{Certifier.} Check that each clause in $\Phi$ has at least one true literal.
	\end{itemize}

	Conclusion: \textsc{sat} is in \textbf{NP}.
\end{exmp}

\begin{exmp}[Certifiers and Certificates - \textsc{HamiltonianCycle}]
	Consider the problem \textsc{HamiltonianCycle} - given an undirected graph $G = (V,E)$, does there exist a simple cycle $C$ that visits every node?
	\begin{itemize}
		\item \textbf{Certificate.} An permutation of the $n$ nodes.
		\item \textbf{Certifier.} Check that the permutation contains each node in $V$ exactly once, and that there is an edge between each pair of adjacent nodes in the permutation.
	\end{itemize}

	Conclusion: \textsc{HamiltonianCycle} is in \textbf{NP}.
\end{exmp}

\begin{defn}[Complexity class \textbf{EXP}]
	The class of decision problems for which there is an exponential time algorithm
\end{defn}

\begin{thm}
\[
	\mathbf{P} \subseteq \mathbf{NP} \subseteq \mathbf{EXP}
\]\end{thm}

\subsection{\textbf{NP}-Completeness} % (fold)
\label{sub:np_completeness}
\begin{defn}[Polynomial transformation]
A problem $X$ \textbf{polynomial transforms} to problem $Y$ if given any  input $x$ to $X$, we can construct an input $y$ in polynomial time such that $x$ is a \textsc{true} instance of $X$ if and only if $y$ is a \textsc{true} instance of $Y$. $|y|$ must be polynomial in $|x|$ as it is constructed in polynomial time. The definition of transformation is due to Karp, reduction to Cook.
\end{defn}

\begin{rem}
	A polynomial transformation is a polynomial reduction with just one call to oracle for $Y$, exactly at the end of the algorithm for $X$. Almost all previous reductions were of this form.
\end{rem}

\begin{defn}[\textbf{NP}-complete]
	A problem $Y$ in \textbf{NP} is \textbf{NP}-complete if for every problem $X$ in \textbf{NP}, $X \leq_p Y$.
\end{defn}

\begin{thm}
	Suppose $Y$ is \textbf{NP}-complete.  Then $Y$ is solvable in polynomial time if and only if \textbf{P} = \textbf{NP}.
\end{thm}

\begin{exmp}[Establishing a problem $Y$ is \textbf{NP}-complete]
	The following is sufficient.
	\begin{itemize}
		\item Show that $Y$ is in \textbf{NP}.
		\item Choose an \textbf{NP}-complete problem $X$.
		\item Show that $X \leq_p Y$. 
	\end{itemize}
\end{exmp}

\begin{thm}
	\textsc{3-sat} is \textbf{NP}-complete
\end{thm}

\begin{proof}
	\textsc{circuit-sat} $\leq_p$ \textsc{3-sat}
\end{proof}

\begin{cor}
	\textsc{VertexCover}, \textsc{IndependentSet}, and \textsc{SetCover} are all \textbf{NP}-complete.
\end{cor}

\begin{proof}
	By reductions in the first part, we had that \textsc{3-sat} reduces to all of the above problems.  Hence, they are all \textbf{NP}-complete.
\end{proof}
% subsection np_completeness (end)

\subsection{co-\textbf{NP} and the Asymmetry of \textbf{NP}} % (fold)
\label{sub:coNP}

\begin{defn}
	Given a decision problem $X$, its complement $\overline{X}$ is the same problem with the \textsc{true} and \textsc{false} answers reversed.
\end{defn}

\begin{defn}[Complexity class \textbf{co-}\textbf{NP}]
	Complements of decision problems in \textbf{NP}.
\end{defn}

\begin{question}
	Does \textbf{NP} $=$ \textbf{co-\textbf{NP}}?
\end{question}

\begin{thm}
	If \textbf{NP $\neq$ co-NP}, then \textbf{P $\neq$ NP}.
\end{thm}

\begin{thm}
	\textbf{P $\subseteq$ NP $\cap$ \textbf{co-NP}}
\end{thm}

\begin{thm}
	\textsc{primes} is in \textbf{NP $\cap$ co-NP}. In fact \textsc{primes} is in \textbf{P} (AKS 2002).
\end{thm}

\begin{thm}[\textsc{factor} is in \textbf{NP $\cap$ co-NP}]Consider the following problems.

	\textsc{factorize}.  Given an integer $x$, find its prime factorisation.

	\textsc{factor}.  Given two integers $x$ and $y$, does $x$ have a nontrivial factor less than $y$?

\begin{thm}
	\textsc{factor} $\equiv_p$ \textsc{factorize}
\end{thm}

Then we have \textsc{factor} is in \textbf{NP $\cap$ co-NP}
\end{thm}

We have established the following sequence.\[
	\textsc{primes} \leq_p \textsc{composites} \leq_p \textsc{factor}
\]

\begin{question}
	Does \textsc{factor} $\leq_p$ \textsc{primes}?
\end{question}
% subsection co-\textbf{NP} and the Asymmetry of \textbf{NP} (end)

\begin{defn}[\textbf{NP}-hard]
	A decision problem such that every problem in \textbf{NP} reduces to it.  This problem is not necessarily in \textbf{NP}.
\end{defn}

% section complexity (end)

\section{Dealing with Intractability} % (fold)
\label{sec:dealing_with_intractability}
Here, we attempt to solve special cases of \textbf{NP}-complete problems that arise in practice.

\subsection{Vertex Cover} % (fold)
\label{sub:vertex_cover}

% subsection vertex_cover (end)
\begin{thm}
	The following algorithm determines if $G$ has a vertex cover of size less than or equal to $k$ in $\bigo{2^k k n}$ time.
\end{thm}

\begin{algorithm}[H]
	\label{alg:Vecter Cover}
	\caption{Small vertex covers}
	\begin{algorithmic}[1]
		\Procedure{VertexCover}{$G, k$}
			\If{$G$ contains no edges} {\textbf{return} \textsc{True}}
			\EndIf
			\If{$G$ contains more than $k |G|$ edges}
			 \textbf{return} \textsc{False}
			\EndIf
			\State
			\State Let $(u,v)$ be any edge of $G$
			\State $a$ = \textsc{VertexCover}($G-\{u\}, k-1$)
			\State $a$ = \textsc{VertexCover}($G-\{v\}, k-1$)
			\State \textbf{return} $a$ \textbf{or} $b$
		\EndProcedure
	\end{algorithmic}
\end{algorithm}

\begin{thm}
	The above algorithm runs in time $\bigo{2^k k n}$.
\end{thm}

\subsection{Independent set on trees} % (fold)
\label{sub:independent_set_on_trees}

% subsection independent_set_on_trees (end)
\begin{thm}
	The following greedy algorithm finds a maximum cardinality independent set in forests (and hence trees).
\end{thm}

\begin{algorithm}[H]
	\label{alg:Vecter Cover}
	\caption{Finds an independent set in a tree or forest.}
	\begin{algorithmic}[1]
		\Procedure{IndependentSet}{$F$}
			\State $S \gets \phi$
			\While{$F$ has at least one edge}
				\State Let $e = (u,v)$ be an edge with $v$ a leaf.
				\State $S \gets S \cup \{ v\}$
				\State $F \gets F - \{ u, v \}$
			\EndWhile
		\State \textbf{return} $S$
		\EndProcedure
	\end{algorithmic}
\end{algorithm}

\begin{thm}
	The above algorithm can run in time $\bigo{n}$ by considering nodes in postorder.
\end{thm}

\subsection{Weighted independent set on trees} % (fold)
\label{sub:weighted_independent_set_on_trees}
\begin{prob}
	Given a tree and node weights $w_v > 0$, find an independent set $S$ that maximises $\sum_{v \in S} w_v$.
\end{prob}

\begin{thm}
	Let $f_{in}(u)$ be the maximum weight independent set rooted at $u$ containing $u$, and let $f_{out}(u)$ be the maximum weight independent set rooted at $u$ not containing $u$.  Then we have the following:
	\begin{align*}
		f_{in}(u) &= w_u + \sum_{v \in \textsc{child}(u)} f_{out}(v) \\
		f_{out}(u) &= \sum_{v \in \textsc{child}(u)} \max \{f_{in}(v), f_{out}(v)\}		
	\end{align*}
\end{thm}

We can find the maximum weighted set by rooting the tree at a node $r$ and considering each node in postorder, and calculating $f_{in}$ and $f_{out}$ at each node $v$. This takes $\bigo{n}$ time.

% subsection weighted_independent_set_on_trees (end)

% section dealing_with_intractability (end)

\section{Randomized and approximation algorithms} % (fold)
\label{sec:probabilistic_algorithms}

\subsection{Probability results} % (fold)
The union bound is $P(\cup_i X_i) \leq \sum_i P(x_i)$.

The Markov inequality for a non-negative $X$ is $P(X \geq a) \leq \frac{\expc X}{a}$.

Chebyshev's inequality is $P(\left | X - \expc X \right | \geq a) \leq \frac {\textrm{Var}(X)}{a^2}$.

The horrific Chernoff bound is $\sum\limits_{i = \lfloor \frac{n}{2} \rfloor + 1}^n \binom{n}{i}p^i (1 - p)^{n - i} \geq  1-e^{- 2n \left( {p - \frac{1}{2}} \right)^2}$.

\subsection{Randomized algorithm for global minimum cut} % (fold)
\label{sub:probabilistic_algorithm_for_global_minimum_cut}
The following algorithm, known as the Contraction algorithm, finds the global minimum cut in an undirected graph.

\begin{defn}[Global minimum cut]
	The size of a cut $(A,B)$ of a graph $G$ is the number of edges with one end in $A$ and one end in $B$.
\end{defn}

\begin{algorithm}[H]
	\label{alg:dynamic_rising_trend}
	\caption{The Contraction algorithm for finding the global minimum cut}
	\begin{algorithmic}[1]
	\Procedure{Contraction}{$G$}
		\ForAll{$v \in V$} $S[v] \gets v$ \EndFor

		\If{$|G| = 2$} \textbf{return} the cut $(S[v_1], S[v_2])$

		\Else
			\State choose an edge $e = (u,v)$ from $G$ uniformly at random.
			\State Set $G'$ equal to the graph resulting from the contraction of $e$, with a new node $w$ replacing $u$ and $v$.
			\State $S[w] \gets S[u] \cup S[v]$ 
			\State \textbf{return} \textsc{Contraction}($G'$)
		\EndIf
	\EndProcedure
	\end{algorithmic}
\end{algorithm}

\begin{thm}
	The algorithm returns the global minimum cut with probability ${n \choose 2}^{-1}$.  Running the algorithm ${n \choose 2} \log n$ returns a global minimum cut with probability greater than $1 - \frac{1}{n}$
\end{thm}

% subsection probabilistic_algorithm_for_global_minimum_cut (end)

\subsection{Approximation algorithm for makespan scheduling} % (fold)
\label{sub:approximation_algorithm_for_makespan_scheduling}
\begin{prob}[Makespan scheduling]
	We have $n$ jobs, each of which takes time $t_i$ to process, and $m$ machines.  Let $A_j$ be the set of jobs assigned to machine $j$.  Let $L_j = \sum_{i \in A_j} t_i$ be the \textbf{load} of machine $j$.  The \emph{makespan} of an assignment is the maximum load on any machine.  
\end{prob}

\begin{algorithm}[H]
	\label{alg:dynamic_rising_trend}
	\caption{Longest Processing Time (\textsc{lpt}) makespan approximation}
	\begin{algorithmic}[1]
	\Procedure{lpt}{$J$}
		\State Sort jobs $J$ so that $t_1 \geq t_2 \geq \dots \geq t_n$.
		\State $A_j \gets \emptyset$
		\State $L_j \gets 0$.
		\For{$i = 1$ to $n$}
			\State $j \gets \text{argmin}_k L_k$
			\State $A_j \gets A_j \cup i$
			\State $L_j \gets L_j + t_i$
		\EndFor
	\EndProcedure
	\end{algorithmic}
\end{algorithm}

\begin{thm}
	If there are at most $m$ jobs, \textsc{lpt} scheduling is optimal.
\end{thm}
\begin{thm}
	\textsc{lpt} scheduling is a $\frac{3}{2}$-approximation.  Careful analysis shows that \textsc{lpt} scheduling is a $\frac{4}{3}$-approximation
\end{thm}

% subsection approximation_algorithm_for_makespan_scheduling (end)

\subsection{Randomized algorithm for \textsc{max-3-sat}} % (fold)
\label{sub:probablistic_algorithm_for_}

\begin{prob}[\textsc{max-3-sat}]
	Given a \textsc{3-sat} instance, find a truth assignment that satisfies as many clauses as possible.
\end{prob}

\begin{lem}
	As \textsc{3-sat} is \textbf{NP}-complete, this is an \textbf{NP}-hard search problem. 
\end{lem}

\begin{thm}
	We can find an approximate solution by setting each variable \textsc{true} with probability $\frac{1}{2}$.  The expected number of clauses satisfied by a random assignment is within a factor of $\frac{7}{8}$ of the optimal solution.
\end{thm}

\begin{proof}
	Each clause is satisfied with probability $1 - \left(\frac{1}{2}\right)^3 = \frac{7}{8}$.  Let $Z_i$ be a random variable equal to 1 if clause $i$ is satisfied, and $0$ otherwise.  Then, by linearity of expectation, we have\[
		\expc{Z} = \expc{Z_1} + \expc{Z_2} + \dots + \expc{Z_k} = \frac{7}{8}k
	\]  Since no assignment can satisfy more than $k$ clauses. 
\end{proof}
% subsection probablistic_algorithm_for_ (end)
\subsection{Randomized algorithm for database access} % (fold)
\label{sub:randomized_algorithm_for_database_access}

% subsection randomized_algorithm_for_database_access (end)
\begin{prob}
	Suppose we have $n$ processes $P_1, \dots, P_n$ attempting to access a single shared database.  The database has the property that it can be accessed by at most one process in a single time period, and if two or more processes attempt to access the database, all processes are locked out for the period.
\end{prob}

\begin{thm}
	Using a randomized algorithm where a process attempts to access the database with probability $\frac{1}{n}$ at each time step, we have the following: with probability at least $1-\frac{1}{n}$, all processes succeed in accessing the database at least once within $t = e \lceil e n \rceil \ln n$ rounds. 
\end{thm}

\subsection{Approximation algorithm for the travelling salesman problem} % (fold)
\label{sub:approximation_algorithm_for_the_travelling_salesman_problem}
In the case of the metric-\textsc{tsp}, where the edge weights satisfy the triangle inequality, we have the following simple algorithm for a 2-approximation for \textsc{tsp}.

\begin{algorithm}[H]
	\label{alg:dynamic_rising_trend}
	\caption{2-Approximation for \textsc{Metric-tsp}}
	\begin{algorithmic}[1]
	\Procedure{Approx-Metric-tsp}{$G$}
		\State Compute a minimal spanning tree $T$ of $G$
		\State Root $T$ arbitraryily and traverse in pre-order (i.e. \textsc{dfs}): $v_1, v_2, \dots, v_n$.
		\State \textbf{return} \textsc{tour: $v_1 \rightarrow v_2 \rightarrow \dots \rightarrow v_n \rightarrow v_1$}
	\EndProcedure
	\end{algorithmic}
\end{algorithm}

\begin{lem}
	Let $\textsc{A}(G)$ be the approximation returned by the above algorithm on the graph $G$.  Then $\textsc{A}(G) \leq  2 \times \textsc{MST}(G)$.
\end{lem}

\begin{lem}
	Let $\textsc{OPT}(G)$ be the optimal solution for our \textsc{metric-tsp} instance.  Then  $\textsc{MST}(G) \leq \textsc{opt}(G)$. 
\end{lem}
\begin{thm}
	We then have $\textsc{A}(G) \leq 2 \times \textsc{OPT}(G)$ - that is, the above algorithm is a 2-approximation for \textsc{Metric-tsp}
\end{thm}

% subsection approximation_algorithm_for_the_travelling_salesman_problem (end)

% section probabilistic_algorithms (end)

\subsection{Some NP-complete problems}

The following are some decision problems which are NP-complete.
\linebreak
\begin{tabular}{p{2in}p{2in}p{2in}}
\begin{itemize}
\item 3SAT
\item Vertex cover
\item Independent set
\item Clique
\end{itemize}
&
\begin{itemize}
\item Travelling salesman
\item Hamiltonian path
\item Hamiltonian circuit
\item Max cut
\end{itemize}
&
\begin{itemize}
\item Longest path
\item Knapsack
\item Subset sum
\end{itemize}
\end{tabular}

\end{document}

### 2011-12-12 MATH 3961 - Metric Spaces/MATH 3961 Lecture Notes.tex
Initial Commit

%
%  untitled
%
%  Created by Andrew Tulloch on 2009-11-04.
%  Copyright (c) 2009 __MyCompanyName__. All rights reserved.
%!TEX TS-program = xelatex
%!TEX encoding = UTF-8 Unicode

\documentclass[10pt, oneside, reqno]{amsart}
\usepackage{geometry, setspace, graphicx, enumerate}
\onehalfspacing                 
\usepackage{fontspec,xltxtra,xunicode}
% \defaultfontfeatures{Mapping=tex-text}
% 
% \setromanfont[Mapping=tex-text,Contextuals= 
% {NoWordInitial,NoWordFinal,NoLineInitial,NoLineFinal}]{Hoefler Text}
% \setsansfont[Scale=MatchLowercase,Mapping=tex-text]{Hoefler Text}
% \setmonofont[Scale=MatchLowercase]{Andale Mono}

	% AMS Theorems
	\theoremstyle{plain}% default 
	\newtheorem{thm}{Theorem}[section] 
	\newtheorem{lem}[thm]{Lemma} 
	\newtheorem{prop}[thm]{Proposition} 
	\newtheorem{cor}[thm]{Corollary}

	\theoremstyle{definition} 
		\newtheorem{defn}[thm]{Definition}
		\newtheorem{conj}[thm]{Conjecture}
		\newtheorem{exmp}[thm]{Example}

	\theoremstyle{remark} 
		\newtheorem*{rem}{Remark} 
		\newtheorem*{note}{Note} 
		\newtheorem{case}{Case}

\newcommand{\al}{\alpha}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\F}{\mathbb{F}}

\newcommand{\Ga}{\mathbb{G}}

\newcommand{\iprod}[2]{\langle #1, #2 \rangle}

\newcommand{\met}{(X,d)}
\newcommand{\topo}{(X,\tau)}

\newcommand{\aut}[1]{\text{Aut}{(#1)}}

\newcommand{\gener}[1]{\langle #1 \rangle}
\newcommand{\charr}[1]{\text{char}(#1)}
\newcommand{\nth}{n\textsuperscript{th}}

\newcommand{\tworow}[2]{\genfrac{}{}{0pt}{}{#1}{#2}}
\newcommand{\xdeg}[2]{[#1 : #2]}
\newcommand{\gal}[2]{\text{Gal}(#1/#2)}
\newcommand{\minpoly}[2]{m_{#1, #2}(x)}

\newcommand{\mapping}[5]{\begin{align*}
	#1 : \quad     #2 &\rightarrow #3 \\
			#4  &\mapsto #5
\end{align*}}

\newcommand{\intr}{\text{Int}\,} 
\newcommand{\ol}[1]{\overline{#1}} 
\newcommand{\xbf}{X \,\backslash \, F}

\title{MATH 3961 - Metric Spaces \\ Exam Notes}								% Document Title
\author{Andrew Tulloch}
%\date{}                                           % Activate to display a given date or no date

\begin{document}
\maketitle

\section{Metric Spaces} % (fold)
\label{sec:metric_spaces}

\begin{defn}[Metric]
	A metric, or distance function, on a set $X$ is a mapping $d : X \times X \rightarrow \R$ such that 
	\begin{itemize}
		\item $d(x,y) \geq 0$ for all $x,y \in X$, and $d(x,y) = 0$ if and only if $x = y$.
		\item $d(x,y) = d(y,x)$ for all $x,y \in X$.
		\item $d(x,z) \leq d(x,y) + d(y,z)$ for all $x,y,z \in X$.
	\end{itemize}

	We call $(X,d)$ a \textbf{metric space}.
\end{defn}

\begin{defn}[Open ball]
	Let $(X,d)$ be a metric space.  For $x \in X$ and $\epsilon > 0$, the set $B_d(x, \epsilon)$ defined by \[
		B_d(x,\epsilon) = \{ y \in X \, | \, d(x,y) < \epsilon \}
	\]
	is callen an \textbf{open ball} in the set $X$.  
\end{defn}

\begin{defn}[Open sets in metric spaces]
	Let $(X,d)$ be a metric space and let $U$ be any subset of $X$.  Then $U$ is called an \textbf{open set} in $X$ if every point of $U$ is an interior point of $U$; that is, for any $a \in U$, there is an open ball $B(a, \epsilon)$ such that $B(a, \epsilon) \subseteq U$.
\end{defn}

\begin{defn}[Properties of open sets]
	Let $\met$ be a metric space.
	\begin{itemize}
		\item $\emptyset$ and $X$ are open.
		\item The union of an arbitrary collection of open sets is open.
		\item The intersection of a \textbf{finite} number of open sets is open.
	\end{itemize}
\end{defn}

\begin{defn}[Closed set]
	A subset $A$ of a metric space $\met$ is \textbf{closed} if it's complement $X \backslash A$ is open in $X$.
\end{defn}

\begin{defn}[Properties of closed sets]
	Let $\met$ be a metric space.
	\begin{itemize}
		\item $\emptyset$ and $X$ are closed.
		\item The union of an finite collection of closed sets is closed.
		\item The intersection of an arbitrary number of closed sets is closed.
	\end{itemize}
\end{defn}

\begin{defn}[Limit point of a subset]
	Let $\met$ be a metric space and let $A$ be a subset of $X$.  Then a point $x$ in $X$ is a \textbf{limit point} of $A$ if every open ball $B(x, \epsilon)$ contains at least one point of $A$.

	The set of all limit points of $A$ is called the \textbf{derived set} $A'$.
\end{defn}

\begin{defn}[Closure of a set]
	Let $\met$ be a metric space and let $A \subseteq X$.  Then the set consisting of $A$ and its limit points is called the \textbf{closure} of $A$, denoted $\ol{A}$.  \[
		\ol{A} = A \cup A'
	\]
\end{defn}

\begin{thm}
	The closure of a set is a closed set, and a set is closed if and only if it is equal to its closure.
\end{thm}

\begin{defn}[Interior of a set]
	Let $\met$ be a metric space and let $A \subseteq X$.  A point $a \in A$ is an \textbf{interior point} of $A$ if there exists $\epsilon > 0$ such that \[
		B(a, \epsilon) \subseteq A
	\]

	The set of interior points of $A$ is called $\intr A$, the \textbf{interior} of $A$.
\end{defn}

\begin{thm}
	The set $\intr A$ is open, and a set $A$ is open if and only if $\intr A = A$.
\end{thm}

\begin{thm}[Properties of interior and closure]
	The interior of a set $A$ is the largest open subset contained in $A$, and the closure of $A$ is the smallest closed set containing $A$.
\end{thm}

\begin{defn}[Isolated point]
	Let $\met$ be a metric space and let $A$ be a subset of $X$.  A point $x \in A$ is called an \textbf{isolated point} if there exists an $\epsilon > 0$ such that \[
		B(x, \epsilon) \backslash \{ x \} \cap A = \emptyset
	\]
\end{defn}

\begin{defn}[Boundary of a subset]
Let $\met$ be a metric space and $A \subseteq X$.  Then the \textbf{boundary} of $A$ is defined as \[
	\partial A = \ol{A} \cap \ol{X \backslash A} = \ol{A} \backslash \intr A
\]
\end{defn}
\begin{thm}[Properties of the boundary]
	Let $\met$ be a metric space and $A \subset X$. Then we have \begin{itemize}
		\item $\ol A = \intr A \cup \partial A$.
		\item $A$ is closed if and only if $\partial A \subseteq A$.
		\item $A$ is open if and only if $\partial A \subseteq X \backslash A$.
		\item $\partial (X \backslash A) = \partial A$.
	\end{itemize}
\end{thm}

\begin{defn}[Diameter of a set]
	The diameter of a subset $A$ of a metric space $\met$, $\delta(A)$ is defined as \[
		\delta(A) = \sup_{x,y \in A} d(x,y)
	\]
\end{defn}

\begin{defn}[Bounded set]
	A subset $A$ of a metric space $\met$ is \textbf{bounded} if its diameter is finite.  Alternatively, a subset is bounded if it is contained in a large enough open set - i.e., there exists $x \in X$ and $\epsilon > 0$ such that $ A \subseteq B(x, \epsilon)$ 
\end{defn}

\subsection{Separable metric spaces} % (fold)
\label{sub:separable_metric_spaces}

\begin{defn}[Separable metric space]
	Let $\met$ be a metric space.  Then a subset $A$ of $X$ is said to be \textbf{dense} in $X$ if $\ol A = X$.  A metric space $\met$ is said to be \textbf{separable} if $X$ has a countable dense subset.
\end{defn}

\begin{cor}
	We note that $A$ is dense in $X$ if and only if for any $x \in X$ and $\epsilon > 0$, there is a point $a \in A$ such that $d(x,a) < \epsilon$.
\end{cor}

% subsection separable_metric_spaces (end)

\subsection{Subspaces} % (fold)
\label{sec:subspaces}

\begin{defn}[Open sets in a subspace]
	Let $\met$ be a metric space and $(Y, d_Y)$ be a metric subspace of $\met$.  Let $G$ be a subset of $Y$.  Then $G$ is open in $Y$ if and only if, for any $x \in G$, there is an open ball $B(x, \epsilon)$ in $X$ such that \[
		B(x, \epsilon) \cap Y \subseteq G
	\]
	A subset $H$ of $Y$ is closed in $Y$ if its complement $G = Y \backslash H$ of $H$ is open in $Y$.

\end{defn}
\begin{thm}[Open sets in a metric subspace]
	Let $(Y, d_Y)$ be a metric subspace of a metric space $\met$, and let $G \subseteq Y$.  Then $G$ is open in $Y$ if and only if there exists an open subset $U$ in $X$ such that $G = U \cap Y$.
\end{thm}

% subsection subspaces (end)

\subsection{Convergence in a Metric Space} % (fold)
\label{sub:convergence_in_a_metric_space}

\begin{defn}[Convergence]
	A sequence $(x_n)$ in a metric space $\met$ is said to \textbf{converge} to a point $x \in X$ if for any $\epsilon > 0$, there exists $N$ such that \[
		n > N \quad \text{implies} \quad d(x_n, x) < \epsilon
	\]
	The point $x$ is called a \textbf{limit} of the sequence $(x_n)$
\end{defn}

\begin{cor}
	A sequence $(x_n)$ in a metric space $\met$ is said to \textbf{converge} to a point $x \in X$ if any open ball $B(x, \epsilon)$ contains almost all $x_n$.
\end{cor}

\begin{thm}[Connection between closed sets and convergent sequences]
	Let $\met$ be a metric space, $A \subseteq X$ and $x \in X$.  Then 
	\begin{itemize}
		\item $x \in \ol{A}$ if and only if there is a sequence $(x_n)$ in $A$ such that $x_n \rightarrow x$.
		\item $A$ is closed if and only if $A$ contains all the limits of convergent sequences in $A$.
	\end{itemize}
\end{thm}

\begin{defn}[Uniform convergence]
Let $(f_n)$ be a sequence of real-valued functions defined on a set $S$ and let $f$ be a function defined on $S$.  Then we say that the sequence $(f_n)$ converges to $f$ uniformly if for any $\epsilon > 0$, there exists $N$ such that \[
	\sup_{x \in S} d(f_n(x), f(x)) < \epsilon
\] for all $n > N$, and where $N$ is independent of $x$.
\end{defn}

\begin{defn}[Cauchy Sequences]
	A sequence $(x_n)$ in a metric space $\met$ is said to be \textbf{Cauchy} in $X$ if for any $\epsilon > 0$, there exists $N$ such that \[
		m,n > N \Rightarrow d(x_m, x_n) < \epsilon
	\]
\end{defn}

\begin{defn}[Completeness in Metric Spaces]
A space $X$ is said to be complete if every Cauchy sequence in $X$ converges in $X$.
\end{defn}

\begin{prop}
	Every convergent sequence $(x_n)$ in a metric space $\met$ is a Cauchy sequence.
\end{prop}

\begin{cor}
	Let $\met$ be a complete metric space.  Then a closed metric subspace $Y = (Y, d_Y)$ of $X$ is complete.  
\end{cor}

\begin{prop}
	Let $\met$ be a metric space.  If a Cauchy sequence $(x_n)$ in $\met$ has a subsequence converging to $x$, then $(x_n)$ converges to $x$.
\end{prop}
% subsection convergence_in_a_metric_space (end)

% section metric_spaces (end)

\section{Continuous Mappings} % (fold)
\label{sec:continuous_mappings}

\begin{defn}[Continuous mapping between metric spaces]
	Let $\met$ and $(Y, d_Y)$ be two metric spaces.  Then a mapping $f : X \rightarrow Y$ is said to be \textbf{continuous} at a point $a \in X$ if for any $\epsilon > 0$, there exists $\delta$ such that \[
		d_X(x,a) < \delta \Rightarrow d_Y(f(x), f(a)) < \epsilon
	\]
\end{defn}

\begin{thm}[Topological characterisations of continuity]
	A funtion $f : X \rightarrow Y$ is continuous at $a \in X$ if and only if for any open set $W$ containing $f(a)$, there exists an open set $G$ containing $A$ such that $f(G) \subseteq W$.  
\end{thm}

\begin{thm}[Sequential characterisation of continuity]
	A function $f: X \rightarrow Y$ is continuous at $a \in X$ if and only if for any sequence $(x_n)$ which converges to $a$ in $X$, the corresponding sequence $(f(x_n))$ converges to $f(a)$ in $Y$.
\end{thm}

\begin{defn}[Continuous mapping]
	A map is continuous on $X$ if any only if it is continuous at every point in $X$.
\end{defn}

\begin{thm}[Topological definition of continuous mapping]
	A mapping $f: X \rightarrow Y$ is continuous on $X$ if and only if for any open set $W$ in $Y$, the set $f^{-1}(W)$ is open in $X$. Alternatively, a function is continuous if the preimage of open sets are open in $X$.
\end{thm}

\begin{thm}[Continuity in terms of closed sets]
	A mapping $f: X \rightarrow Y$ is continuous on $X$ if and only if the preimage of closed sets are closed in $X$.
\end{thm}

% section continuous_mappings (end)

\section{Homeomorphism and Equivalent Metrics} % (fold)
\label{sec:homeomorphism_and_equivalent_metrics}
\begin{defn}[Homeomorphism]
	Let $X$ and $Y$ be metric spaces and let $f: X \rightarrow Y$ be a map between them.  Then $f$ is a \textbf{homeomorphism} from $X$ to $Y$ if we have \begin{itemize}
		\item $f$ is a bijection.
		\item $f$ and $f^{-1}$ are continuous
	\end{itemize}

	If a homeomorphism exists between $X$ and $Y$, we say that $X$ and $Y$ are \textbf{homeomorphic}, and that $X \simeq Y$.
\end{defn}

\begin{defn}[Characterisations of homeomorphism]
Let $f: X \rightarrow Y$ be a bijective mapping.  Then the following are equivalent.
\begin{itemize}
	\item $f$ is a homeomorphism;
	\item for any $U \subseteq X$, $U$ is open in $X$ if and only if $f(U)$ is open in $Y$;
	\item for any $G \subseteq X$, $G$ is closed in $X$ if and only if $f(G)$ is closed in $Y$;
	\item for any $A \subseteq X$, $f(\ol{A}) = \ol{f(A)}$;
	\item for any $B \subseteq Y$, $\ol{f^{-1}(B)} = f^{-1}(\ol{B})$
	\item for any $B \subseteq Y$, $f^{-1}(\intr B) = \intr f^{-1}(B)$
\end{itemize}
\end{defn}

\begin{defn}[Isometric mappings]
	Let $\met$ and $(Y, d_Y)$ be two metric spaces and $f: X \rightarrow Y$ a mapping.  Then $f$ is said to be \textbf{isometric} or an \textbf{isometry} if $f$ preserves distances; that is, for all $x,y \in X$, \[
		d_Y(f(x), f(y)) = d_X(x,y)
	\]
	The space $X$ is said to be isometric with the space $Y$ if there exists a bijective isometry of $X$ onto $Y$.  The spaces $X$ and $Y$ are isometric spaces
\end{defn}

\begin{thm}
	Any isometric mapping from $X$ onto $Y$ is a homeomorphism.  Moreover, if $X$ is complete and $Y$ is isometric with $X$, then $Y$ is also complete.
	\end{thm}

\begin{defn}[Equivalent metrics]
	Let $(X, d_1)$ and $(X, d_2)$ be two metric spaces.  If the identity mapping $\text{id} : (X, d_1) \rightarrow (X, d_2)$ is a homeomorphism, then the metrics $d_1$ and $d_2$ are said to be equivalent on $X$.
\end{defn}

\begin{thm}[Characterisations of equivalent metrics]
	Let $(X, d_1)$ and $(X, d_2)$ be two metric spaces.  Then the following are equivalent.
	\begin{itemize}
		\item The metrics $d_1$ and $d_2$ are equivalent on $X$;
		\item for any $U \subseteq X$, $U$ is open in $(X, d_1)$ if and only if $U$ is open in $(X, d_2)$;
		\item for any $G \subseteq X$, $G$ is closed in $(X, d_1)$ if and only if $G$ is closed in $(X, d_2)$;
		\item The sequence $(x_n)$ converges to $a$ in $(X, d_1)$ if and only if it converges to $a$ in $(X, d_2)$.
	\end{itemize}
\end{thm}

\begin{thm}[Equivalent metrics]
	Let $(X, d_1)$ and $(X, d_2)$ be two metric spaces.  If there exist strictly positive numbers $c$ and $C$ such that \[
		c d_1(x,y) \leq d_2(x,y) \leq C d_1(x,y)
	\] for all $x,y \in X$, then the metrics $d_1, d_2$ are equivalent on $X$.
	\end{thm}
% section homeomorphism_and_equivalent_metrics (end)

\section{Contraction Mapping Theorem} % (fold)
\label{sec:contraction_mapping_theorem}

\begin{defn}[Uniformly continuous]
	A function $f: X \rightarrow Y$ is uniformly continuous if for any $\epsilon > 0$ there exists $\delta > 0$ such that \[
		d_X(x,y) < \delta \Rightarrow d_Y(f(x), f(y)) < \epsilon
	\] where $\delta$ is independent of $x,y$.
\end{defn}

\begin{defn}[Contraction mapping]
	Let $f$ be a mapping from a metric space $\met$ to itself.  Then $f$ is called a \textbf{contraction mapping} if there exists a constant $K > 1$ such that for all $x,y \in X$, \[
		d(f(x), f(y)) \leq K (x,y)
	\] 
\end{defn}

\begin{prop}
	If $f: X \rightarrow X$ is a contraction mapping, then $f$ is \textbf{uniformly continuous}, and hence continuous, on $X$
\end{prop}

\begin{thm}[Banach fixed point theorem]
	Let $\met$ be a complete metric space and let $f: X \rightarrow X$ be a contraction mapping.  Then $f$ has a unique fixed point $p$ in $X$
\end{thm}

\begin{proof}
	\textbf{Existence:} Show that the sequence $(x_n)$, defined as $x_n = f^n(x_0)$ for some $x_0 \in X$ is Cauchy.

	\textbf{Uniqueness:} If $p$ and $q$ are fixed points of $f$, the we have that \[
			d(p,q) = d(f(p), f(q)) \leq K d(p,q)
			\] and so $p = q$.
\end{proof}

\begin{cor}[Application of Banach fixed point theorem to ordinary differential equations]
	We seek to solve the differential equation \[
		\frac{dx}{dt} = f(t,x)
	\] given the initial condition $x(t_0) = x_0$.

	Define $Y$ be the subspace of the set of all continuous functions on $[t_0 - \beta, t_0 + \beta]$ with the supremum metric, satisfying $d(x(t), x_0) < c \beta$
	We claim that the mapping $F: Y \rightarrow Y$ defined by\[
		F(x(t)) = x_0 + \int_{t_0}^t f(s, x(s)) \, ds
	\] is a contraction mapping.  As $F$ is a contraction mapping on a complete metric space, we must have that it has a unique fixed point, which satisfies the differential equation above.
\end{cor}

% section contraction_mapping_theorem (end)

\section{Completeness} % (fold)
\label{sec:completeness}

We recall the definition of completeness in metric spaces.

\begin{defn}[Completeness in Metric Spaces]
A space $X$ is said to be complete if every Cauchy sequence in $X$ converges in $X$.
\end{defn}

We now state the theorem that every metric space can be completed. The space $\hat{X}$ in the theorem is called the completion of the given space $X$.

\begin{thm}[Completion of a metric spaces]
	Let $\met$ be a metric space.  Then there exists a complete metric space $\hat{X} = (\hat{X}, \hat{d})$ which has a subspace $W$ that is isometric with $X$ and is dense in $\hat{X}$
\end{thm}

\begin{proof}
	TO DO
\end{proof}
% section completeness (end)

\section{Connectedness} % (fold)
\label{sec:connectedness}

\begin{defn}[Disconnected]
Let $X$ be a metric space (or topological space). Then $X$ is said to be \textbf{disconnected} if there exist two non-empty subsets $A_1, A_2$ of $X$ such that \[
	X= A_1 \cup A_2 \text{ and } A_1 \cap \ol A_2 , \ol A_1 \cap A_2 = \emptyset
\]

If no two sets $A_1, A_2$ exist, we say that $X$ is \textbf{connected}
\end{defn}

\begin{thm}[Characterisations of connectedness]
	Let $\met$ be a metric space. Then the following statements are equivalent:
	\begin{itemize}
		\item $X$ is disconnected;
		\item There exist two non-empty disjoint open subsets $A_1, A_2$ in $X$ such that $X = A_1 \cup A_2$;
		\item There exist two non-empty disjoint closed subsets $A_1, A_2$ in $X$ such that $X = A_1 \cup A_2$; 
		\item There exists are proper subset of $X$ which is both open and closed in $X$
	\end{itemize}
\end{thm}

\begin{defn}[Connected subspace]
	Let $\met$ be a metric space and $A$ a non-empty subset of $X$.  Then $A$ is said to be a connected subst of $X$ if $A$ is connected as a metric subspace and to be a \textbf{disconnected} subset of $X$ if $A$ is disconnected as a metric subspace.
\end{defn}

\begin{thm}[Intervals in $\R$]
	A subset $A$ of $\R$ containing at least two points is connected if and only if $A$ is a interval.
\end{thm}

\begin{thm}[Characterisations of connectedness]
	Let $\mathcal{S}(2)$ be the two point discrete metric space.
	If $A$ is connected then any continuous mapping $f: A \rightarrow \mathcal{S}(2)$ is a constant mapping.  Alternatively, if $f : A \rightarrow \mathcal{S}(2)$ is continuous, then $f(A) = \{0\}$ or $\{1\}$.
\end{thm}

\begin{thm}[Connectedness is a topological property]
	Let $X$ and $Y$ be two metric spaces, and let $f: X \rightarrow Y$ be a continuous mapping.  Then if $A \subseteq X$ is connected in $X$, then the image $f(A)$ is connected in $Y$.
\end{thm}

\begin{defn}[Path-connected]
	Let $X$ be a metric space and $A$ a subset of $X$. Then $A$ is said to be \textbf{path-connected} if for any $a,b \in A$, there is a path joining $a$ and $b$, that is, a continuous mapping $f: [0,1] \rightarrow A$ such that $f(0) = a, f(1) = b$.
\end{defn}

\begin{thm}[Path-connectedness implies connectedness]
	Let $X$ be a metric space and $A$ a subset of $X$.  If $A$ is path-connected, then $A$ is connected. 
\end{thm}
We note that the converse is not necessarily true - that is, there exist connected sets that are not path connected.  However, in $\R^n$, we have the following.

\begin{thm}[For open sets in $\R^n$, path-connectedness is equivalent to connectedness]
	Let $X$ be any open set in $\R^n$.  Then $X$ is connected if and only if $X$ is path-connected.
\end{thm}
% section connectedness (end)

\section{Compactness} % (fold)
\label{sec:compactness}
Compactness in metric spaces.

\begin{defn}[Open covering of a set]
	Let $X$ be a metric space (or any topological space), and let $A \subseteq X$.  Then a family $\mathcal{U}$ of open sets in $X$, is called an \textbf{open covering} of $A$ if \[
		A \subseteq \cup_{U \in \mathcal{U}} U
	\]
A subset $\mathcal{V}$ of $\mathcal{U}$ is called a \textbf{finite subcovering} if $\mathcal{V}$ covers $A$ and has a finite number of elements.
\end{defn}

\begin{defn}[Compact subset of a topological space]
	let $X$ be a metric space (or any topological space), and let $A \subseteq X$.  Then $A$ is called a \textbf{compact subset} of $X$ if \textbf{every} open covering $\mathcal{U}$ of $A$ has a finite subcovering $\mathcal{V}$ of $A$. 
\end{defn}

In metric spaces, we have the following useful theorem.

\begin{thm}[Implications of compactness in metric spaces]
	Let $\met$ be a metric space.  If $A$ is a compact subset in $X$, then $A$ is closed and bounded in $X$.  
\end{thm}

In $\R^n$, we have the following, more general, results.  These are key results in characterising compact subsets of Euclidean space.

\begin{thm}[Heine-Borel]
	Every closed and bounded interval in $\R$ is compact.
\end{thm}

\begin{thm}[Compactness in $\R^n$]
	Let $A$ be a subset of $\R^n$.  Then $A$ is compact if and only if $A$ is closed and bounded.
\end{thm}

\subsection{Properties of compact sets} % (fold)
\label{sub:properties_of_compact_sets}
\begin{thm}
	A subset $A$ in $\R^n$ is compact if and only if every sequence in $A$ has a convergent subsequence with limit in $A$.
\end{thm}

\begin{defn}[Compactness is a topological property]
	Let $X$ and $Y$ be topological spaces, and let $f : X \rightarrow Y$ be a continuous mapping.  If a subset $A$ in $X$ is compact, then the image $f(A)$ is compact in $Y$.

	That is, continuous images of compact sets are compact.
\end{defn}

\begin{cor}
	The following are true in an arbitrary metric space.
	\begin{itemize}
		\item A continuous image of a compact subset is closed.
		\item A continuous image of a compact subset is bounded.
	\end{itemize}
\end{cor}

\begin{thm}
	Any closed subspace $A$ of a compact space $X$ is compact.
\end{thm}

\begin{proof}
	TO DO
\end{proof}
% subsection properties_of_compact_sets (end)
% section compactness (end)

\section{Applications to Continuous Functions $f : [a,b] \rightarrow \R$} % (fold)
\label{sec:applications_to_f_a_b_rightarrow_r_}
\begin{thm}[Intermediate Value Theorem]
	Let $f: [a,b] \rightarrow \R$ be a continuous function.  Let $K$ be a number lying between $f(a)$ and $f(b)$.  Then there exists a point $c \in [a,b]$ such that $f(c) = K$.
\end{thm}

\begin{proof}
	Using connectedness of $[a,b]$, we have that $f([a,b])$ is connected.  Thus, if $f(a) < K < f(b)$, then $K \in f([a,b])$.  Hence, there exists $c \in [a,b]$ such that $f(c) = K$.
\end{proof}
% section applications_to_f_a_b_rightarrow_r_ (end)

\section{Topological Spaces} % (fold)
\label{sec:topological_spaces}

A topological space is defined as follows.  Let $X$ be a non-empty set.  Then a family $\tau$ of subsets of $X$ is called a topology for $X$ if $\tau$ satisfies 
\begin{itemize}
	\item $\emptyset, X \in \tau$;
	\item The union of any subfamily of members of $\tau$ is in $\tau$;
	\item The intersection of any \textbf{finite} subfamily of members of $\tau$ is in $\tau$.
\end{itemize}

The pair $(X, \tau)$ is called a \textbf{topological space}, and the members of $\tau$ are called the \textbf{open sets} in $(X, \tau)$.

\begin{defn}[Interior of a subset]
	Let $\topo$ be any topological space and $A \subseteq X$.  Then $a \in A$ is called an \textbf{interior point} of $A$ if there exists an open set $U$ containing $a$ such that $U \subseteq A$.  We denote by $\intr A$ the set of all interior points of $A$.
\end{defn}

\begin{thm}[Properties of the interior]
	Let $\topo$ be any topological space and let $A \subseteq X$.  Then $\intr A$ is the \textbf{largest} open subset contained in $A$.
\end{thm}

\begin{defn}[Closed subsets]
	Let $A$ be a subset of a topological space $\topo$.  Then $A$ is called a \textbf{closed set} in $X$ if the complement $X \backslash A$ is open in $X$, that is, if $X \backslash A \in \tau$.
\end{defn}

\begin{thm}
	In a topological space $\topo$,
	\begin{itemize}
		\item $\emptyset$ and $X$ are closed;
		\item The intersection of any collection of closed sets is closed;
		\item The union of any \textbf{finite} collection of closed sets are closed.
	\end{itemize}
\end{thm}

\begin{defn}[Limit point of a subset]
	Let $\topo$ be a topological space and let $A \subseteq X$.  Then a point $x \in X$ is called a \textbf{limit point} or \textbf{accumulation point} of $A$ if every open set $G$ containing $x$ contains a point of $A$ different from $x$, i.e.,\[
		G \in \tau, x \in G \Rightarrow (G \backslash \{x \}) \cap A \neq \emptyset
	\]
We denote by $A'$ the set of all limit points of $A$, and is called the derived set of $A$.
\end{defn}

\begin{thm}[Properties of closed sets]
	Let $\topo$ be a topological space and let $A \subseteq X$.  Then $A$ is closed if and only if $A' \subseteq A$.
\end{thm}

\begin{defn}[Closure of a subset]
	Let $\topo$ be a topological space and $A \subseteq X$.  Then the set consisting of $A$ together with all its limit points is called the \textbf{closure} of $A$, and is denoted by $\ol A$.  Thus,\[
		\ol A = A \cup A'
 	\]
\end{defn}

\begin{prop}
	Let $\topo$ be a topological space and $A \subseteq X$.  Then \[
		\ol A = \{ x \in X \, | \, \text{for every open set $U$ containing $x$, $U \cap A \neq \emptyset$} \}.
	\]
\end{prop}

\begin{thm}
		Let $\topo$ be a topological space and $A \subseteq X$.  Then $\ol A$ is the smallest closed set containing $A$.
\end{thm}

\begin{defn}[Dense subset]
		Let $\topo$ be a topological space and $A \subseteq X$.  Then a subset $A$ of $X$ is said to be \textbf{dense} in $X$ if $\ol A = X$.
\end{defn}

\begin{defn}[Nowhere dense]
		Let $\topo$ be a topological space and $A \subseteq X$. Then $A$ is \textbf{nowhere dense} in $X$ if and only if the interior of the closure of $A$ is empty.  That is, $\intr (\ol A) = \emptyset$.

		Alternatively, a set is nowhere dense if and only if $X \backslash \ol A$ is dense in $X$.
\end{defn}

\begin{defn}[Boundary of a subset]
	Let $\topo$ be a topological space and $A \subseteq X$.  Then the \textbf{boundary} of $A$, denoted by $\partial A$, is defined as \[
		\partial A = \ol A \cap \ol{X \backslash A}
	\]
\end{defn}

\begin{thm}[Characterisation of the boundary]
		Let $\topo$ be a topological space and $A \subseteq X$. Then \[
			\ol A = \intr A \cup \partial A
		\]
\end{thm}

\begin{defn}[Convergence in topological spaces]
	Let $\topo$ be a topological space.  Then a sequence $(x_n)$ of points in $X$ is said to \textbf{converge} to a point $x \in X$ if for any open set $U$ containing $x$, there exists a positive integer $N$ such that \[
		n > N \Rightarrow x_n \in U
	\]
	That is, if any open set $U$ containing $x$ contains almost all of the terms of the sequence.
\end{defn}

\begin{defn}[Induced or relative topology]
		Let $\topo$ be a topological space and $Y \subseteq X$. Let \[
			\tau_Y = \{ G \subseteq Y \, | \, G = U \cap Y \text{ for some $U \in \tau$} \}
		\]
	Then $\tau_Y$ is a topology for $Y$, called the \textbf{induced} or \textbf{relative topology} on $Y$ and the space $(Y, \tau_Y)$ is called a subspace of $\topo$
\end{defn}

\begin{defn}[Bases for a topology]
	Let $\topo$ be a topological space.  Then a subfamily $\mathcal{B}$ of $\tau$ is called a \textbf{base} for the topology if for every open set $U$ in $\tau$ is the union of members of $\mathcal{B}$.  Equivalently, $\mathcal{B} \subseteq \tau$ is a basis for $\tau$ if and only if for any point $a$ in an open set $U \in \tau$, there exists $V \in \mathcal{B}$ such that $a \in V \subseteq U$.
\end{defn}

\begin{thm}[Characterisation of a basis for a topology]
	A family of nonempty subsets of a set $X$ is a base for some topology $\tau$ on $X$ if and only if it satisfies the following properties.
	\begin{itemize}
		\item $X = \cup_{B \in \mathcal{B}} B$
		\item For any $B_1, B_2 \in \mathcal{B}$, $B_1 \cap B_2$ is the union of members of $\mathcal{B}$.  Equivalently, if $b \in B_1 \cap B_2$, then there exists $B_b \in \mathcal{B}$ such that $b \in B_b \subseteq B_1 \cap B_2$.
	\end{itemize}
\end{thm}

\begin{thm}[Continuity in terms of a basis]
	Let $(X, \tau_X)$ and $(Y, \tau_Y)$ be topological spaces and $f : X \rightarrow Y$ a mapping.  Let $\mathcal{B}_Y$ be a basis for $\tau_Y$.  Then $f$ is continuous if and only if for any $B \in \mathcal{B}_Y$, $f^{-1}(B)$ is open in $X$ - i.e., is in $\tau_X$.
\end{thm}

\begin{thm}[Product spaces]
	Let $(X, \tau_X)$ and $(Y, \tau_Y)$ be topological spaces.  Then the family $\mathcal{B}$ given by \[
		\mathcal{B} = \{ U \times V \, | \, U \in \tau_X, V \in \tau_Y \}
	\]
	is a base for a topology on $X \times Y$.
\end{thm}

\subsection{Compactness} % (fold)
\label{sub:compactness}

\begin{defn}[Compactness in terms of a basis]
	A topological space $\topo$ is compact if and only if there exists a base $\mathcal{B}$ for $\tau$ such that every open covering of $X$ be members of $\mathcal{B}$ has a finite subcovering. 
\end{defn}

\begin{thm}
	The product space of two compact topological spaces is compact.
\end{thm}

% subsection compactness (end)

\subsection{Connectedness} % (fold)
\label{sub:connectedness}

\begin{thm}
	Let $X$ and $Y $ be topological spaces and let $f : X \rightarrow Y$ be a homeomorphism.  Then $X$ is connected if and only if $Y$ is connected.  	
\end{thm}

\begin{thm}
	Let $X$ be a topological space.  Let $\{ A_i \}$be a family of connected subsets in $X$ and suppose that for all $i,j$, $A_i \cap A_j \neq \emptyset$.  Then the union $A = \cup_i A_i$ is connected.
\end{thm}

% subsection connectedness (end)

% section topological_spaces (end)

\section{Separation Properties} % (fold)
\label{sec:spearation_properties}

\begin{defn}[$T_0$-spaces]
	A topological space $\topo$ is called a \textbf{$T_0$-space} if for any pair of distinct points $a,b$ of $X$, either there exists an open set $U$ containing $a$ and not $b$ or an open set $V$ containing $b$ and not $a$.
\end{defn}

\begin{defn}[$T_1$-spaces]
A topological space $\topo$ is called a \textbf{$T_1$-space} if for any pair of distinct points $a,b$ of $X$, there exists an open set $U $ in $X$ with $a \in U$ and $b \notin U$.
\end{defn}

Every $T_1$-space is $T_0$, but not the reverse.

\begin{thm}[Characterisation of $T_1$-spaces]
	AA topological space $\topo$ is called a \textbf{$T_1$-space} if and only if eery singleton set $\{ a \}$ of $X$ is closed (and so every finite subset of $X$ is closed).
\end{thm}

\begin{defn}[$T_2$-spaces or Hausdorff spaces]
	A topological space $\topo$ is called a \textbf{$T_2$-space} or a \textbf{Hausdorff space} if for any pair of distinct points $a,b$ of $X$, there are disjoint open sets $U$ and $V$ in $X$ such that $a \in U$ and $b \in V$.
\end{defn}

Every $T_2$-space is $T_1$, but not the reverse.

\begin{exmp}
	Any metric space is a $T_2$-space
\end{exmp}

\begin{thm}
	Every subspace of a $T_1$- or $T_2$-space is a $T_1$- or $T_2$-space. 
\end{thm}

\begin{thm}
	Every product space of a $T_1$- or $T_2$-space is a $T_1$- or $T_2$-space. 
\end{thm}

\subsection{Regular Spaces and $T_3$-spaces} % (fold)
\label{sub:regular_spaces}

\begin{defn}[Regular space]
	A topological space $\topo$ is called a \textbf{regular} space if for any closed set $F$ in $X$ and $a \in X \backslash F$, there are disjoint open sets $U$ and $V$ in $X$ such that $F \subseteq U$ and $a \in V$.  A regular $T_1$-space is called a $T_3$-space.
\end{defn}

\begin{thm}
	A topological space $X$ is regular if and only if for any point $a$ in $X$ and any open set $U$ containing $A$, there is an open set $W$ containing $a$ such that $\ol W \subseteq U$.

\end{thm}

Every $T_3$-space is $T_2$, but not the reverse.

% subsection regular_spaces (end)
\subsection{Normal Spaces and $T_4$-spaces} % (fold)
\label{sub:normal_spaces_and_t_4_spaces}

\begin{defn}[Normal spaces and $T_4$-spaces]
	A topological space $\topo$ is called a \textbf{normal space} if for any two disjoint closed sets $A$ and $B$ in $X$, there are disjoint open sets $U$ and $V$ in $X$ such that $A \subseteq U$ and $B \subseteq V$.  A normal $T_1$-space is called a $T_4$-space.
\end{defn}

Clearly every $T_4$-space is a $T_3$-space.  However, a normal space may not be a $T_1$-space or a regular space and a regular space may not be normal.

\begin{thm}
	Every metric space $\met$ is a $T_4$-space
\end{thm}

\begin{thm}
	Every compact Hausdorff space is normal.  Additionally, any compact subset $A$ of a Hausdorff space $X$ is closed.
\end{thm}
% subsection normal_spaces_and_t_4_spaces (end)

Our final theorem is \textbf{Urysohn's Lemma}.
\begin{thm}[Urysohn's Lemma]
	Let $X$ be a normal space.  Then, for any disjoint closed sets $A$ and $B$ in $X$, there exists a continuous function $f: X \rightarrow [0,1]$ such that $F(A) = \{0 \}$ and $F(B) = \{1 \}.$
\end{thm}

\begin{thm}[Tietze Extension Theorem]
	Let $X$ be a normal space, $A$ a closed subset of $X$, and $f : A \rightarrow \R$ continuous.  Then there is a continuous function $g: X \rightarrow R$ such that $g |_A = f$ - that is, $g$ restricted to $A$ is $f$. 
\end{thm}

% section spearation_properties (end)

\section{Hilbert Spaces} % (fold)
\label{sec:hilbert_spaces}
Let $V$ be a vector space over a field $\F$, where $\F$ is either $\R$ or $\C$.

\begin{defn}[Inner product space]
	A function $\langle \cdot, \cdot \rangle: V \times V \rightarrow \F$ is called an inner product if 
	\begin{itemize}
		\item $\iprod{u}{v} = \ol{\iprod{u}{v}}$ for all $u,v \in V$.
		\item $\iprod{u}{u} \geq 0$ for all $u \in V$ with equality if and only if $u = 0$.
		\item $\iprod{\alpha u + \beta v}{w} = \alpha \iprod{u}{w} + \beta \iprod{v}{w}$ for all $u,v,w \in V$ and $\alpha, \beta \in \F$.
	\end{itemize}
	We say that $V$ equipped with $\iprod{\cdot}{\cdot}$ is an \textbf{inner product space}.
\end{defn}

\begin{defn}[Induced norm]
	If $V$ is an inner product space with $\iprod{\cdot}{\cdot}$, we define \[
		\| u \| := \sqrt{\iprod{u}{u}}
	\] for all $u \in E$.  The operation $\| \cdot \|$ is the \textbf{induced norm}.
\end{defn}

\begin{thm}[Cauchy-Schwarz inequality]
	Let $V$ be an inner product space.  The \[
		|\iprod{u}{v}| \leq \|u \| \| v \|
	\]
\end{thm}

\begin{defn}[Hilbert space]
	An inner product space which is complete with respect to the induced norm is called a \textbf{Hilbert space}.
\end{defn}

\begin{prop}[Continuity of the inner product]
	Let $V$ be an inner product space.  Then the inner product is continuous with respect to the induced norm.  
\end{prop}

\begin{prop}[Parallelogram identity]
	Let $V$ be an inner product space and $\| \cdot \|$ the induced norm.  Then \[
		\| u + v \|^2 + \| u - v \|^2 = 2 \|u \|^2 + 2 \| v \|^2
	\] for all $u,v \in E$
\end{prop}

\subsection{Projections and orthogonal complements} % (fold)
\label{sub:projections_and_orthogonal_complements}

\begin{defn}[Projection]
	Let $V$ be a normed space and $M$ a non-empty closed subset of $V$.  We define the \textbf{set of projections of $x$ onto $M$} by \[
		P_M(x) = \{ m \in M \, | \, \|x - m \| = d(x, M) \}
	\]
	This set is non-empty as $M$ is closed.
\end{defn}
\begin{defn}[Orthogonal complement]
	For an arbitrary non-empty subset $M$ of an inner product space $H$ we set \[
		M^{\perp} := \{ x \in H \, | \, \iprod{x}{m} = 0 \text{ for all $m \in M$} \}
	\]
	We call $M^\perp$ the orthogonal complement of $M$ in $H$.
\end{defn}

\begin{lem}[Properties of the orthogonal complement]
	Suppose $M$ is a non-empty subset of the inner product space $H$.  Then $M^\perp$ is a closed subspace of $H$ and $M^\perp = \ol M^\perp = (\text{span $M$})^\perp = (\text{span $\ol M$})\perp$
\end{lem}

\begin{thm}[Key properties of the orthogonal complement]
	Suppose that $M$ is a closed subspace of a Hilbert space $H$. Then 
	\begin{itemize}
		\item $H = M \oplus M^\perp$ 
	\end{itemize}

	\end{thm}

\begin{cor}[Dense subspace of a Hilbert space]
	A subspace $M$ of a Hilbert space $H$ is dense in $H$ if and only if $M^\perp = \{0 \}$.
\end{cor}

% subsection projections_and_orthogonal_complements (end)

\subsection{Orthogonal systems} % (fold)
\label{sub:orthogonal_systems}

\begin{defn}[Orthogonal systems]  
	Let $H$ be an inner product space with inner product $\iprod{\cdot}{\cdot}$ and induced norm $\| \cdot \|$.  Let $M \subset H$ be a non-empty subset.
	\begin{itemize}
		\item $M$ is called an \textbf{orthogonal system} if $\iprod{u}{v} = 0$ for all $u,v \in M$ with $u \neq v$.
		\item $M$ is called an \textbf{orthonomal system} if it is orthogonal and $\| u \| = 1$.
		\item $M$ is called a \textbf{complete orthonormal sytem} or \textbf{orthonormal basis} of $H$ if it is an orthonormal system and $\ol{\text{span $M$}} = H$.
	\end{itemize}

\end{defn}

\begin{thm}[Pythagoras's Theorem]
	Suppose that $H$ is an inner product space and $M$ an orthogonal system in $H$.  Then the following assertions are true:
	\begin{itemize}
		\item $M \backslash \{0\}$ is linearly independent.
		\item If $(x_n)$ is a sequence in $M$ with distinct terms and $H$ is complete, then $\sum x_k$ converges if and only if $\sum \| x_k \|^2$ converges.  In that case, \[
			\left\| \sum x_k \right\|^2 = \sum \|x_k \|^2
		\]
	\end{itemize}
\end{thm}

\begin{thm}[Bessel's Inequality]
	Let $H$ be an inner product space and $M$ an orthonormal system in $H$.  Then \[
		\sum_{m \in M} | \iprod{x}{m}|^2 \leq \| x \|^2
	\]
	for all $x \in H$.  Moreover, the set $\{ m \in M \, | \, \iprod{x}{m} \neq 0 \}$ is at most countable for all $x \in H$.
\end{thm}

% subsection orthogonal_systems (end)
% section hilbert_spaces (end)

\end{document}

### 2011-12-12 MATH 3962 - Rings, Fields, and Galois Theory/MATH 3962 Lecture Notes.tex
Initial Commit

%
%  untitled
%
%  Created by Andrew Tulloch on 2009-11-04.
%  Copyright (c) 2009 __MyCompanyName__. All rights reserved.
%!TEX TS-program = xelatex
%!TEX encoding = UTF-8 Unicode

\documentclass[10pt, oneside, reqno]{amsart}
\usepackage{geometry, setspace, graphicx, enumerate}
\onehalfspacing                 
\usepackage{fontspec,xltxtra,xunicode}
\defaultfontfeatures{Mapping=tex-text}

%\setromanfont[Mapping=tex-text,Contextuals= 
%{NoWordInitial,NoWordFinal,NoLineInitial,NoLineFinal}]{Hoefler Text}
%\setsansfont[Scale=MatchLowercase,Mapping=tex-text]{Hoefler Text}
%\setmonofont[Scale=MatchLowercase]{Andale Mono}

	% AMS Theorems
	\theoremstyle{plain}% default 
	\newtheorem{thm}{Theorem}[section] 
	\newtheorem{lem}[thm]{Lemma} 
	\newtheorem{prop}[thm]{Proposition}

	\theoremstyle{definition} 
		\newtheorem{defn}[thm]{Definition}
		\newtheorem{conj}[thm]{Conjecture}
		\newtheorem{exmp}[thm]{Example}
		\newtheorem{cor}[thm]{Corollary}

	\theoremstyle{remark} 
		\newtheorem*{rem}{Remark} 
		\newtheorem*{note}{Note} 
		\newtheorem{case}{Case}

\newcommand{\al}{\alpha}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{Q}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\F}{\mathbb{F}}
\newcommand{\Ga}{\mathbb{G}}

\newcommand{\aut}[1]{\text{Aut}{(#1)}}

\newcommand{\gener}[1]{\langle #1 \rangle}
\newcommand{\charr}[1]{\text{char}(#1)}
\newcommand{\nth}{n\textsuperscript{th}}

\newcommand{\tworow}[2]{\genfrac{}{}{0pt}{}{#1}{#2}}
\newcommand{\xdeg}[2]{[#1 : #2]}
\newcommand{\gal}[2]{\text{Gal}(#1/#2)}
\newcommand{\minpoly}[2]{m_{#1, #2}(x)}

\newcommand{\mapping}[5]{\begin{align*}
	#1 : \quad     #2 &\rightarrow #3 \\
			#4  &\mapsto #5
\end{align*}	
}

\title{MATH 3962 - Rings, Fields and Galois Theory \\ Exam Notes}								% Document Title
\author{Andrew Tulloch}
%\date{}                                           % Activate to display a given date or no date

\begin{document}
\maketitle

\section{Background Theory} % 
\label{cha:background_theory}

\begin{defn}[Monoids]
	A monoid is a set $S$ equipped with a single operation $\cdot$ obeying the following axioms
	\begin{itemize}
		\item \textbf{Closure} For all $a,b \in S$, $a \cdot b \in S$ 
		\item \textbf{Associativity} For all $a,b,c \in S$, $(a \cdot b) \cdot c = a \cdot (b \cdot c)$.
		\item \textbf{Identity}  There exists an element $e$ in $S$ such that $e \cdot a = a = a \cdot e$ for all $a \in S$. 
	\end{itemize}

\end{defn}

\begin{defn}[Groups]
	A group is a set equipped with an operation $\cdot$ obeying the axioms of associativity, existence of inverses, and existence of an identity.
\end{defn}

\begin{defn}[Abelian Groups]
	An abelian group is a group where the operation $\cdot$ is commutative.
\end{defn}

\begin{defn}[Cyclic Groups]
	A cyclic group is a group that can be generated by a single element $\langle x \rangle = \{ x^n \, | \, x \in \mathbb{Z}\}$
\end{defn}

\begin{defn}[Subgroup]
	A subset $H$ of a group $G$ is a subgroup of $G$ if and only if $H$ is non-empty and for all $x,y \in G$ 
	\begin{itemize}
		\item If $x,y \in H$ then $x \cdot y \in H$ 
		\item If $x \in H$ then $x^{-1} \in H$.
	\end{itemize}
\end{defn}

\begin{defn}[Normal Subgroup]
	A subgroup $K$ of a group $G$ is said to be \textbf{normal} in $G$ if $g^{-1} k g \in K$ for all $k \in K$ and $g \in G$.  Equivalently, the subgroup $K$ is normal in $G$ if $g^{-1}Kg = K$, or $gK = Kg$ for all $g \in G$.
\end{defn}

\begin{defn}[Quotient Group]
	If $G$ is a group and $H$ is a subgroup, we can form the \textbf{quotient group} $G/H$ as follows.
	Defining the equivalence relation $~$ as follows: for all $x,y \in G$, \[
		x \sim y if x = yh
	\] for some $h \in H$.  The set \[
		xH = \{xh \, | \, h \in H \}
	\] is called the left coset containing $x$.  These cosets partition $G$, and the number of cosets of $H$ in $G$ is denoted by $\xdeg{G}{H}$.  By \textbf{Lagrange's Theorem}, we have \[
		\xdeg{G}{H} = \frac{|G|}{|H|}
	\]

	Now, letting $K$ be a normal subgroup of $G$, we have the following.
	THe set of al cosets in $G$ forms a group, with multiplication satisfying $(xK)(yK) = xyK$ for all $x,y \in G$.
\end{defn}

\begin{defn}[Solvable Groups]
	A group $G$ is \textbf{solvable} if there is a chain of subgroups\[
		1 = G_0 \lhd G_1 \lhd G_2 \lhd \cdots \lhd G_s = G
	\] where each $G_i$ is normal in $G_{i+1}$ and the quotient groups $G_{i+1}/G_i$ is abelian for all $i$.
\end{defn}

\begin{cor}
	The finite group $G$ is solvable if and only if for every divisor $n$ of $|G|$ with $\text{gcd}(n, \frac{|G|}{n}) = 1$, $G$ has a subgroup of order $n$.
\end{cor}

\begin{cor} Let $N$ be normal in $G$.
	If $N$ and $G/N$ are solvable, then $G$ is solvable.
\end{cor}

\begin{thm}[Subgroups of Cyclic Groups]
	Let $G = \langle x \rangle$ be a cyclic group.  Then we have the following.
	\begin{itemize}
		\item Every subgroup of $H$ is cyclic.   More precisely, if $K \leq H$, then either $K = \{1\}$ or $K = \langle x^d \rangle$, where $d$ is the smallest positive integer such that $x^d \in K$.
		\item If $|H| = \infty$, then if $a \neq b,$ then $\gener{x^a} \neq \gener{x^b}$. 
		\item If $|H| = n < \infty$, then for each positive integer $a$ dividing $n$ there is a unique subgroup of $H$ of order $a$.  This subgroup is the cyclic group $\gener{x^d}$, where $d = \frac{n}{a}$.  Furthermore, the subgroups of $H$ correspond bijectively with the positive divisors of $n$.
	\end{itemize}
\end{thm}

\begin{defn}[Homomorphism of Groups]
	A map $\varphi: G \rightarrow H$ is a homomorphism if and only if
	\begin{itemize}
		\item $\varphi(xy) = \varphi(x) \varphi(y)$ for all $x,y \in G$
	\end{itemize}
\end{defn}

\begin{defn}[Isomorphism of Groups]
	A map $\varphi: G \rightarrow H$ is an isomorphism of groups if and only if 
	\begin{itemize}
		\item $\varphi$ is a homomorphism.
		\item $\varphi$ is a bijection.
	\end{itemize}
\end{defn}

\begin{defn}[Symmetric Groups]
	The symmetric group of order $n$ is the set of all permutations of the finite set $\{1,2,\dots, n\}$, with the operation being composition of permutations.
\end{defn}

\begin{prop}[Properties of the Symmetric Groups]
	Let $S_n$ be the symmetric group of order $n$.  Then we have 
	\begin{itemize}
		\item  $| S_n | = n!$
		\item $S_n$ is non-abelian for all $n \geq 3$.
	\end{itemize}

\end{prop}

\begin{defn}[Elementary Symmetric Functions]
	Let $x_1, x_2, \dots, x_n$ be indeterminates.  Then the \textbf{elementary symmetric functions} $s_1, x_2, \dots, s_n$ are defined by \begin{align*}
		s_1 &= x_1 + x_2 + \cdots  + x_n\\
		s_2 &= x_1 x_2 + x_1 x_3 + \cdots + x_2 x_3 + x_2 x_4 + \cdots x_{n-1} x_n \\
		s_n &= x_1 x_2 \cdots x_n
	\end{align*}
\end{defn}

\begin{defn}[Symmetric functions]
A function $f(x_1, x_2, \dots, x_n)$ is called \textbf{symmetric} if it is not changed by any permutation of the variables $x_1, x_2, \dots, x_n$. 
\end{defn}

\subsection{Solving Polynomial Equations} % 
\label{sub:solving_polynomial_equations}
We have explicit solutions for solving polynomials of degrees two and three. Polynomials of degree two are solved using the quadratic equation.  Polynomials of degree three are solvable using \textbf{Cardano's Method}

% subsection solving_polynomial_equations

% section background_theory

\section{General Ring Theory} % 
\label{cha:general_ring_theory}

\begin{defn}[Rings]
	A \textbf{ring} $R$ is a set equipped with two binary operations $+$ and $\times$ satisfying the following axioms
	\begin{itemize}
		\item (R,+) is an \textbf{abelian group} -  impyling the existence of negatives, a zero element, and commutative addition.
		\item $\times$ is associative: $(a \times b) \times c = a \times (b \times c)$ for all $a,b,c \in R$
		\item Multiplication distributes over addition: \[
			(a + b) \times c  = (a \times c) + (b \times c) \qquad c \times (a + b) = (c \times a) + (c \times b)
		\]
	\end{itemize}

	A ring is \textbf{commutative} if multiplication is commutative.

	A ring is said to contain an identity if there is an alement $1 \in R$ such that $1 \times a = a \times 1 = a$ for all $a \in R$.
\end{defn}

\begin{defn}[Zero divisors]
	A non-zero element $a \in R$ is called a zero divisor if there is a nonzero element $b \in R$ such that either $ab = 0$ or $ba = 0$. 
\end{defn}

\begin{defn}[Field]
	A field can be defined in several ways.
	\begin{itemize}
		\item A field is a commutative ring $F$ with identity $1 \neq 0$ such that every non-zero element $a \in F$ has a multiplicative inverse.
		\item A field is a commutative ring $F$ with identiy $1 \neq 0$ in which every nonzero element is a unit, i.e. $F^{\times} = F - \{0\}$.
	\end{itemize}
\end{defn}

\begin{defn}[Unit]
	Assume a ring $R$ has identity $1 \neq 0$.  Then an element $u$ of $R$ is called a \textbf{unit} if there is some $v \in R$ such that $uv = vu = 1$.  The set of units in $R$ is denoted $R^{\times}$.  Te set of units in $R$ form a group under multiplication, denoted the \textbf{group of units} of $R$.
\end{defn}

\begin{defn}[Integral Domain]
	An integral domain is a commutative ring with identity $1 \neq 0$ with no zero divisors.
\end{defn}

\begin{cor}[Cancellation property]
	Let $R$ be an integral domain.  Then for any $a,b,c \in R$, if $ab = ac$, then either $a = 0$ or $b = c$. 
		ab = ac 
\end{cor}

\begin{cor}
	Any finite integral domain is a field.
\end{cor}

\begin{defn}[Subring]
	A subring of a ring $R$ is a subgroup of $R$ that is closed under multiplication.  Alternatively, a subset $S$ of a ring $R$ is a subring if the operations of addition and multiplication in $R$ when restricted to $S$ gives $S$ the structure of a ring.    
\end{defn}

\begin{cor}
	To show a subset of a ring $R$ is a subring it suffices to check that it is \textbf{nonempty} and \textbf{closed under subtraction and under multiplication}.
\end{cor}

\begin{defn}[Characteristic of a Ring]
	The characteristic of a ring, $\charr{R}$, is defined as the smallest positive number $n$ such that $ n \times 1 = 0$
\end{defn}

\begin{exmp}[Examples of Rings]
	The set of all $n \times n$ matrices over a ring $R$ is a ring, denoted by $\text{Mat}_n(R)$.

	Let $x$ be indeterminate, and let $R$ be a commutative ring with identity $1 \neq 0$.  The set of all formal sums \[
		a_n x^n + a_{n-1} x^{n-1} + \cdots + a_1 x + a_0 
	\] with $a_i \in R$ is the \textbf{polynomial ring} $R[x]$.
\end{exmp}

\subsection{Homomorphisms, kernels, images.} % 
\label{sub:homomorphisms_kernels_images_}
This section deals with maps between rings $R$ and $S$.

\begin{defn}[Homomorphisms of Rings]
	Let $\varphi: R \rightarrow S$ be a map between two rings $R$ and $S$.  Then $\varphi$ is a ring homomorphism if and only if
	\begin{itemize}
		\item $\varphi(a + b) = \varphi(a) + \varphi(b)$
		\item $\varphi(ab) = \varphi(a) \varphi(b)$
		\item $f(1) = 1$
	\end{itemize}
\end{defn}

\begin{cor}
	The image of a ring homomorphism $\varphi$ is a subring of $S$.
\end{cor}
\begin{defn}[Kernel of a Homomorphism]
	The kernel of a ring homomorphism $\varphi$, denoted $\ker \varphi$, is the set of elements $R$ that maps to $0 \in S$, i.e. the set of all $a \in R$ such that $\varphi(a) = 0$.
\end{defn}

\begin{cor}
	The kernel of a homomorphism $\varphi$ is a subring of $R$.  Furthermore, if $\al \in \ker \varphi$, then $r \al$ an $\al r \in \ker \varphi$ for all $r \in R$, i.e, $\ker \varphi$ is closed under multiplication by elements in $R$. 
\end{cor}

\begin{exmp}
	Let $R$ be a subring of a commutative ring $T$, and let $\al \in T$.  Then the function $\text{eval}_\alpha : R[x] \rightarrow T$ is a homomorphism.
\end{exmp}

% subsection homomorphisms_kernels_images_

\subsection{Ideals, Quotient Rings, and Isomorphisms} % 
\label{sub:ideals_quotient_rings_and_isomorphisms}

\begin{defn}[Ideal of a Ring]
	A subset $I$ of a ring $R$ is an ideal of $R$ if and only if the following conditions all hold.
	\begin{itemize}
		\item $I$ is nonempty
		\item $a + b \in I$ for al $a, b \in I$
		\item $-x \in I$ for all $x \in I$
		\item $ax, xa \in I$ for all $x \in I$ and $a \in R$.
	\end{itemize}
\end{defn}

\begin{cor}
	The kernel of a homomorphism $\varphi$ is an ideal of $R$.
\end{cor}

\begin{prop}[Cosets of an Ideal]
	Let $I$ e an ideal in a ring $R$.  The equivalence relation $\equiv$ defined by $a \equiv b$ if and only if $a - b \in I$ is an equivalence relation on the ring $R$, partioning the ring into a set of equivalence classes $r + I$ with $r \in R$ called the \textbf{cosets} of $I$ in $R$
\end{prop}

\begin{defn}[Quotient Ring]
	Let $R$ be a ring and let $I$ be an ideal of $R$.  Then the additive quotient group $R / I$ is a ring under the binary operations:
	\begin{itemize}
		\item $(r + I) + (s+I) = (r+s) + I$
		\item $(r + I) \times (s + I) = (rs) + I$
	\end{itemize}

	The elements of $R/I$ are precisely the cosets of $I$ in $R$.
\end{defn}

\begin{thm}
	Let $I$ be an ideal in the ring $R$.  Then the mapping $\varphi: R \rightarrow R/I$ given by $\varphi(\alpha) = \al + I$ is a surjective homomorphism with kernel $I$.
\end{thm}

We can collect these results into the following theorem, known as the \textbf{First Isomorphism Theorem}.

\begin{thm}[First Isomorphism Theorem]
	Let $R$ and $S$ be rings, and $\varphi: R \rightarrow S$ a homomorphism of rings.  Then the kernel of $\varphi$ is an ideal of $R$, the image of $\varphi$ is a subring of $S$, and there is an isomorphism $\psi: R / \ker \varphi \rightarrow \varphi(R)$ such that $\psi(r + \ker \varphi) = \varphi(r)$. 
\end{thm}

\begin{thm}[Second Isomorphism Theorem]
	Let $R$ be a ring. Let $S$ be a subring and let $I$ be an ideal of $R$.  Then $S + I = \{ s + i \, | \, s \in S, i \in I \}$ is a subring of $R$, $S \cap I$ is an ideal of $S$, and $(S + I)/I$ is isomorphic to $S/(S \cap I)$.
\end{thm}

\begin{thm}[Third Isomorphism Theorem]
	Let $I$ and $J$ be ideas of $R$ with $I \subseteq J$.  Then $J/I$ is an ideal of $R/I$ and $(R/I)/(J/I)$ is isomorphic to $R/J$. 
\end{thm}

% subsection ideals_quotient_rings_and_isomorphisms

\subsection{Classification of Ideals} % 
\label{sub:classification_of_ideals}

\begin{prop}[Sum, Product, Intersection of Ideals]
	Let $I$ and $J$ be ideals of $R$.  Then
	\begin{itemize}
		\item The sum of $I$ and $J$, $I + J$, is equal to $\{ a + b \, | \, a \in I, b \in J \}$.
		\item The product of $I$ and $J$, $IJ$, is equal to the set of all finite sums of elements of the form $ab$ with $a \in I$, $b \in J$.
		\item The intersection of ideals, $I \cap J$, is defined simply as $I \cap J$.
	\end{itemize}

	It can be shown that the sum $I + J$ of ideal $I$ and $J$ is the smallest ideal of $r$ containing both $I$ and $J$, and the product $IJ$ is an ideal contained in $I \cap J$, but can be strictly smaller.
\end{prop}

\begin{cor}
	Let $I = a\Z, J = b\Z$.  Then we have $I + J = d\Z$, where $d  = \text{GCD}(a,b)$.  We also have that $IJ = ab \Z$, and $I \cap J = \text{LCM}(a,b)$ 
\end{cor}
\begin{defn}[Principal Ideal]
	Let $R$ be a commutative ring.  An ideal that can be generated by a single element, of the form $aR$ for some $a \in R$, is called a \textbf{principal ideal}. 
\end{defn}

\begin{cor}
	Every ideal in the ring $\Z$ is principal.
\end{cor}

\begin{prop}
	Let $I$ be an ideal of a ring $R$.  Then $I = R$ if and only if $I$ contains a unit
\end{prop}

\begin{prop}
	Let $I$ be an ideal of a commutative ring $R$.  Then $R$ is a field if and only if its only ideals are $0$ and $R$.
\end{prop}

\begin{defn}[Maximal Ideal]
	An ideal $M$ in an arbitary ring $S$ is called a \textbf{maximal ideal} if $M \neq S$ and the only ideals containing $M$ are $M$ and $S$.
\end{defn}

\begin{prop}
	Asume $R$ is commutative.  Then the idela $M$ is a maximal ideal if and only if the quotient ring $R/M$ is a field.
\end{prop}

\begin{defn}[Prime Ideal]
	Assume $R$ is commutative.  An ideal $P$ is called a \textbf{prime ideal} if $P \neq R$ and whenever the product $ab$ of two elements $a,b \in R$ is an element of $P$, then at least one of $a$ and $b$ is an element of $P$.

	The definition is motivated by the following example.  Let $n$ be a nonnegative integer.  Then $n\Z$ is a \textbf{prime} ideal provided $n \neq 1$ and every time the product $ab$ of two integers is an element of $n\Z$, at least one of $a,b$ is an element of $n\Z$. This is equivalent to stating that whenever $n$ divides $ab$, $n$ must divide $a$ or divide $b$.  Thus, $n$ must be prime.  Thus, \textbf{the prime ideals of $\Z$ are simply the ideal $p\Z$ of $\Z$ generated by prime numbers $p$ together with the ideal $0$}.
\end{defn}

\begin{prop}
	Assume $R$ is commutative.  Then the ideal $P$ is a prime ideal in $R$ if and only if the quotient ring $R/P$ is an integral domain.
\end{prop}

\begin{cor}
	Assume $R$ is commutative.  Then every maximal ideal of $R$ is a prime ideal.
\end{cor}
\begin{proof}
	If $M$ is a maximal ideal then $R/M$ is a field.  As a field is an integral domain, we thus have that $R/M$ is an integral domain, and thus $M$ is a prime ideal.
\end{proof}

% subsection classification_of_ideals 
% section general_ring_theory

\section{Integral Domains} % 
\label{cha:integral_domains}

\begin{defn}[Field of Fractions]
	Let $R$ be a commutative ring.  Let $D$ be any nonempty subset of $R$ that does not contain 0, does not contain any zero divisors, and is closed under multiplication.  Then there is a commutative ring $Q$ with $1$ such that $Q$ contains $R$ as a subring and every element of $D$ is a unit of $Q$.  The ring $Q$ has the following additional properties.
	\begin{itemize}
		\item Every element of $Q$ is of the form $r d^{-1}$ for some $r \in R$ and $d \in D$.   In particular, if $D = R - \{0\}$, then $Q$ is a field.
		\item The ring $Q$ is the \textbf{smallest} ring containing $R$ in which all element of $D$ become units,
 in the following sense - Any ring containing an isomorphic copy of $R$ in which all the elements of $D$ become units must also contain an isomorphic copy of $Q$.	
	\end{itemize}
\end{defn}

\begin{defn}[Divisibility]
	Let $R$ is a commutative ring.  Let $a,b \in R$.  Then we say that $a$ divides $b$ if and only if $b = ca$ for some $c \in R$.  We write $a | b$ if $a$ divides $b$. 
\end{defn}

\begin{defn}[Units, Irreducibles, Primes, Associates]
	Let $R$ be an integral domain - a commutative ring with $1 \neq 0$ with no zero divisors.  Then we have the following.
	\begin{itemize}
		\item An element $a \in R$ is a  \textbf{unit} in $R$ is an element such that there exists $b \in R$ where $ab = ba = 1$.
		\item Suppose $r \in R$ is nonzero and is not a unit.  Then $r$ is called \textbf{irreducible } in $R$ if whenever $r = ab$ with $a,b \in R$, at least one of $a$ or $b$ must be a unit in $R$.  Otherwise, $r$ is said to be \textbf{reducible.}
		\item The nonzero element $p \in R$ is called \textbf{prime} in $R$ if the ideal $(p)$ generated by $p$ is a prime ideal.  Alternatively, if $R$ is a commutative ring, and $p \in R$.  We say that $p$ is \textbf{prime} if it is nonzero and not a unit, and the following condition holds: for all $a,b \in R$, if $p | ab$ then either $p | a$ or $p | b$.
		\item Two elements $a$ and $b$ differing by a unit are said to be \textbf{associate} in $R$ (i.e., $a = ub$ for some unit $u$ in $R$).
	\end{itemize}
\end{defn}

\begin{prop}
	In an integral domain a prime element is always irreducible.
\end{prop}
\begin{proof}
	Suppose $(p)$ is a nonzero prime ideal and $p = ab$.  Then $ab = p \in (p)$, so by the definition of prime ideal one of $a$ or $b$, say $a$, is in $(p)$.  Thus $a = pr$ for some $r$.  This implies $p = a = prb$, and so $rb = 1$.  Thus $b$ is a unit.  This shows that $p$ is irreducible.
\end{proof}

\begin{defn}[Principle Ideal Domains]
	A \textbf{Principle Ideal Domains} (PID) is an integral domain in which every ideal is principal. 
\end{defn}

\begin{defn}[Unique Factorisation Domains]
	A \textbf{Unique Factorisation Domain} (UFD) is an integral domain $R$ in which every nonzero element $r \in R$ which is not a unit has the following two properties:
	\begin{itemize}
		\item $r$ can be written as a finite product of irreducibles $p_i$ of $R$ (not necessarily distinct)
		\item The decomposition above is \textbf{unique up to associates}: if $r = q_1 q_2 \dots q_m$ is another factorisation of $r$ into irreducibles, then $m = n$ and there is a renumbering of the factors so that $p_i$ is associate to $q_i$. 
		\end{itemize}
\end{defn}

\begin{thm}
	Every principle ideal domain is a unique factorisation domain.
\end{thm}

\subsection{Greatest Common Divisor, Euclidean Algorithm} % 
\label{sub:greatest_common_divisor_euclidean_algorithm}
\begin{defn}[Greatest common divisor]
	Let $R$ be a principal ideal domain and $a,b$ nonzero elements of $R$.  An element $d \in R$ is called a \textbf{greatest common divisor} of $a$ and $b$ if 
	\begin{enumerate}
		\item $d | a$ and $d|b$, and
		\item for all $e \in R$, if $e | a$ and $e | b$ then $e | d$.
	\end{enumerate}
\end{defn}

\begin{prop}[Existence and properties of the GCD]
	Let $R$ be a principal ideal domain, and $a,b \in R$ nonzero elements.  Then
	\begin{itemize}
		\item There is an element $d \in R$ which is a greatest common divisor of $a$ and $b$, and every associate of $d$ is also a greatest common divisor of $a$ and $b$.
		\item The greatest common divisor is unique up to associates.
		\item An element $d \in R$ is a greatest common divisor of $a$ and $b$ if and only if $d | a$, $d|b$ and there exist $r,s \in R$ such that $d = ar + bs$.  
		\item An element $d \in R$ is a greatest common divisor of $a$ and $b$ if and only if $aR + bR = dR$. 
	\end{itemize}
\end{prop}

\begin{defn}[Euclidean Algorithm]
	This operation works in Euclidean Domains - domains where we can define a degree function measuring the size of each element.
	Given $a,b \in R$, calculates the GCD of $a$ and $b$.  Operates as follows:

	While $b \neq 0$ - set $a,b$ = $b, \text{Rem}(a,b)$

	where Rem($a,b$) is the remainder of $a$ upon division by $b$.
\end{defn}

\begin{thm}
	Every Principle Ideal Domain and Unique Factorisation Domain is a Euclidean Domain.
\end{thm}

% subsection greatest_common_divisor_euclidean_algorithm

\begin{defn}[Gaussian Integers]
	The Gaussian Integers $\Ga$ are defined as $\Z[i]$, the set $\{ a + b i \, | \, a, b \in Z \}$.
\end{defn}

Let $\alpha = a + bi \in \Ga$. Define the norm $N(\alpha) = \al \overline{\al} = a^2 + b^2$.

We have the following theorem, due to Fermat.

\begin{thm}
	The prime $p$ is the sum of two integer squares, $p = a^2 + b^2, a,b \in \Z$, if and only if $p = 2$ or $p \equiv 1 \bmod 4$.  This representation is essentially unique up to signs and interchanging elements.

	Secondly, the irreducible element in in the Gaussian integers $\Ga$ are as follows.
	\begin{itemize}
		\item 1 + i (with norm 2)
		\item The primes $p \in \Z$ with $p \equiv 3 \bmod 4$ (with norm $p^2$)
		\item $a + bi, a-bi$, the distinct irreducible factors of $p = a^2 + b^2$ for primes $p$ with $p \equiv 1 \bmod 4$.
	\end{itemize} 
\end{thm}

\begin{proof}
	COMPLETE THIS!
\end{proof}

\subsection{Polynomial Rings} % 
\label{sub:polynomial_rings}

\begin{defn}[Polynomial Ring]
	Let $x$ be indeterminate, and let $R$ be a commutative ring with identity $1 \neq 0$.  The set of all formal sums \[
		a_n x^n + a_{n-1} x^{n-1} + \cdots + a_1 x + a_0 
	\] with $a_i \in R$ is the \textbf{polynomial ring} $R[x]$.
\end{defn}

\begin{prop}
	Let $R$ be an integral domain.  Then
	\begin{itemize}
		\item $\deg p(x) q(x) = \deg p(x) + \deg q(x)$ if $p(x), q(x)$ are non-zero.
		\item The units of $R[x]$ are the units of $R$.
		\item $R[x]$ is an integral domain.
	\end{itemize}
\end{prop}

\begin{prop}
	Let $I$ be an ideal of the ring $R$, and let $(I) = I[x]$ denote the ideal of $R[x]$ generated by $I$ (the set of all polynomials with coefficients in $I$).  Then we have\[
		R[x]/(I) \simeq (R/I)[x]
	\] 
\end{prop}

\begin{thm}
	Let $F$ be a field.  The polynomial ring $F[x]$ is a Euclidean Domain.  Specifically, if $a(x)$ and $b(x)$  are two polynomials in $F[x]$ with $b(x)$ nonzero, then there are unique $q(x)$ and $r(x)$ in $F[x]$ such that \[
		a(x) = q(x) b(x) + r(x)
	\] with $r(x) = 0$ or $\deg r(x) < \deg b(x)$.
\end{thm}

\begin{thm}
	If $F$ is a field, then $F[x]$ is a Principal Ideal Domain and a Unique Factorisation Domain.
\end{thm}

\begin{thm}[Gauss's Lemma]
	Let $R$ be a UFD with field of fractions $F$ and let $p(x) \in R[x]$.  If $p(x)$ is reducible in $F[x]$ then $p(x)$ is reducible in $R[x]$.
\end{thm}

\begin{cor}
	$R$ is a UFD if and only if $R[x]$ is a UFD.
\end{cor}

\begin{prop}
	Let $F$ be a field and let $p(x) \in F[x]$.  Then $p(x)$ has a factor of degree one if and only if $p(x)$ has a root in $F$, i.e., there exists $\al \in F$ with $p(\al) = 0$.
\end{prop}

\begin{cor}
	A quadratic or cubic in $F[x]$ is reducible if and only if it has a root in $F$.
\end{cor}

Our next theorem gives us conditions on the roots of polynomials with integer coefficients.

\begin{thm}[Rational Roots Theorem]
	Let $p(x) = a_n x^n + a_{n-1} x^{n-1} + \cdots + a_1 x + a_0 \in \Z[x]$.  If $\frac{r}{s} \in \Q$ is a root of $p(x)$ and $r,s$ are relatively prime, then $r | a_0$ and $s | a_n$.  In particular, if $p(x)$ is \textbf{monic} and $p(d) = 0$ for all integers $d$ dividing the constant term $a_0$ of $p(x)$, then $p(x)$ has no roots in $\Q$.
\end{thm}

The following theorem gives conditions on the reducibility of a polynomial modulo some proper ideal.

\begin{thm}
	Let $I$ be a proper ideal in the integral domain $R$ and let $p(x)$ be a non-constant monic polynomial in $R[x]$.  If the image of $p(x)$ in $(R/I)[x]$ cannot be factored in $(R/I)[x]$ into two polynomials of smaller degree, then $p(x)$ is irreducible in $R[x]$.
\end{thm}

\begin{exmp}
	Consider the polynomial $p(x) = x^2 + x + 1 \in \Z[x]$.  Then, reducing modulo 2, we see that $p(x)$ is irreducible in $\Z[x]$.
\end{exmp}

Our next theorem, Eisenstein's Irreducibility Criterion, applied to the ring $\Z[x]$ is stated below.

\begin{thm}[Eisenstein's Criterion for $\Z$]

	Let $p$ be a prime in $\Z$ and let $f(x) = a_n x^n + a_{n-1}x^{n-1} + \cdots + a_1 x + a_0 \in \Z[x]$. Suppose $p$ divides $a_i$ for all $a_i, i \in \{0, 1, \dots, n-1 \}$, $p$ does not divide $a_n$, and $p^2$ does not divide $a_0$.  Then $f(x)$ is irreducible in $\Q[x]$.
\end{thm}
% subsection polynomial_rings

% section integral_domains

\section{Fields} %

\label{sec:fields}

\begin{defn}[Field Extension]
	If $K$ is a field containing the subfield $F$, then $K$ is said to be an \textbf{extension field}, or simply and \textbf{extension}, of $F$, denoted $K/F$.  
\end{defn}

\begin{defn}[Degree of an Extension]
	The degree of a field extension $K/F$, denoted $\xdeg{K}{F}$, is the dimension of $K$ as a vector space over $F$.  The extension is said to be finite if $\xdeg{K}{F}$ is finite, and \textbf{infinite} otherwise
\end{defn}

\begin{thm}
	Let $F$ be a field and let $p(x) \in F[x]$ be an irreducible polynomial.  Then there exists a field $K$ containing an isomorphic copy of $F$ in which $p(x)$ has a root.  identifying $F$ with this isomorphic copy shows that there exists an extension of $F$ in which $p(x)$ has a root. 
\end{thm}
\begin{proof}
	 Consider the quotient  \[
		K = F[x]/(p(x))
	\] of the polynomial ring $F[x]$ by the ideal generated by $p(x)$.  As $p(x)$ is irreducible in the PID $F[x]$, the ideal generated by $p(x)$ is a \textbf{maximal} ideal.  Thus, the quotient $F[x]/(p(x))$ is a field.  The projection $\pi$ of $F[x]$ to the quotient $F[x]/(p(x))$ restricted to $F \subset F[x]$ gives a homomorphism $\varphi = \pi |_{F} : F \rightarrow K$ which is not identically zero, and hence $\varphi(F) \simeq F$.

	If $\overline{x} = \pi(x)$ denotes the image of $x$ in the quotient $K$, then we have \begin{align*}
		p(\overline{x}) &= \overline{p(x)}  &\text{(since $\pi$ is a homomorphism)} \\
					&=  p(x) \bmod p(x)	 &\text{in $F[x]/(p(x))$} \\
					&= 0
	\end{align*}

	Thus $K$ contains a root of the polynomial $p(x)$.  Hence, $K$ is an extension of $F$ in which the polynomial $p(x)$ has a root. 
\end{proof}

Our next theorem allows us to understand the field $K = F[x]/(p(x))$ more fully, by having a simple representation for the elements of this field.  Since $F$ is a subfield of $K$, we might ask in particular for a basis for $K$ as a vector space over $F$.

\begin{thm}
	Let $p(x) \in F[x]$ be an irreducible polynomial of degree $n$ over the field $F$, and let $K$ be the field $F[x]/(p(x))$.  Let $\theta = x \bmod (p(x)) \in K$.  Then the elements \[
		1, \theta, \theta^2, \dots, \theta^{n-1}
	\]
	are a basis for $K$ as a vector space over $F,$ so the degree of the extension is $n$, i.e., $\xdeg{K}{F} = n$. Hence, \[
		K = \{ a_0 + a_1 \theta + a_2 \theta^2  + \cdots + a_{n-1} \theta^{n-1} \,| \, a_i \in F \}
	\] consists of all polynomials of degree less than or equal to $n$ in $\theta$.
\end{thm}

\begin{proof}
	Let $a(x) \in F[x]$ be any polynomial with coefficients in $F$.  Since $F[x]$ is a Euclidean Domain, we may divide $a(x)$ by $p(x)$:\[
		a(x) = q(x)p(x) + r(x)
	\] 
	It thus follows that $a(x) \equiv r(x) \bmod (p(x))$, which shows that every residue class in $F[x]/(p(x))$ is represented by a polynomial of degree less than $n$. Hence the images $1, \theta, \theta^2,\dots, \theta^{n-1}$ of $1,x,x^2,\dots$ in the quotient \textbf{span} the quotient as a vector space over $F$.   We now show these elements are linearly independent, and so form a basis for the quotient over $F$.
	If the elements $1, \theta, \theta^2, \dots, \theta^{n-1}$ were not linearly independent in $K$, then there would be a linear combination\[
		b_0 + b_1 \theta + b_2 \theta^2 + \cdots + b_{n-1} \theta^{n-1} = 0
	\] in $K$, with $b_i \in F$ not all equal to zero.  This is equivalent to \[
	b_0 + b_1 \theta + b_2 \theta^2 + \cdots + b_{n-1} \theta^{n-1} \equiv 0 \bmod (p(x))
	\] i.e., $p(x)$ divides the above polynomial in $x$.  But $\deg p(x) > \deg{\sum^{n-1} b_i x^i}$, and so by contradiction we have the above elements are a basis for $K$ over $F$.  Thus $\deg{K}{F} = n$.
\end{proof}

\begin{prop}
	The above theorem gives us a formula for elements of the field $K$.  Let $K$ be an extension of $F$, and $a(\theta), b(\theta) \in K$.  Then addition is defined as usual, and multiplication in $K$ is defined as \[
		a(\theta) b(\theta) = r(\theta)
	\]  where $r(x)$ is the remainder obtained upon dividing the polynomial $a(x) b(x)$ by $p(x)$ in $F[x]$
\end{prop}

\begin{defn}[Simple Extension]
	If the field $K$ is generated by a single element $\al$ over $F,$ then $K = F(\al)$, then $K$ is said to be a \textbf{simple} extension of $F$ and the element $\al$ is called a \textbf{primitive element} for the extension.
\end{defn}

\begin{thm}
	Let $F$ be a field and $p(x) \in F[x]$ be an irreducible polynomial. Suppose $K$ is an extension field of $F$ containing a root $\al$ of $p(x)$, thus $p(\al) = 0$.  Let $F(\al)$ denote the subfield of $K$ generated over $F$ by $\al$.  Then \[
		F(\al) \simeq F[x]/(p(x))
	\]
\end{thm}
\begin{proof}
	Consider the natural homomorphism
	\mapping{\varphi}{F[x]}{F(\alpha) \subseteq K}{a(x)}{a(\alpha)}
	Since $p(\alpha) = 0$ by assumption, we have that the element $p(x)$ is in the kernel of $\varphi$, and so we obtain an induced homomorphism \[
		\varphi : F[x]/(p(x)) \rightarrow F(\al)
	\]
	Since $p(x)$ is irreducible, we have that the quotient ring is a field, and as $\varphi$ is not identically zero, we must have $\varphi$ is an isomorphism.
\end{proof}

We now prove a theorem regarding the different roots of an irreducible polynomial.  Consider the equation $p(x) = x^3 - 2$.  Adjoining any of the three roots produces the same field extension (up to isomorphism).    This is known as the \textbf{Isomorphism Extension Theorem}.

\begin{thm}[Isomorphism Extension Theorem]
	Let $\varphi : F \mapsto F'$ be an isomorphism of fields.  Let $p(x) \in F[x]$ be irreducible and let $p'(x) \in F'[x]$ be the irreducible polynomial obtained by applying the map $\varphi$ to the coefficients of $p(x)$.  Let $\alpha$ be a root of $p(x)$ (in some extension of $F$), and let $\beta$ be a root of $p'(x)$ in some extension of $F'$.  Then, there is an isomorphism
	\mapping{\sigma}{F(\alpha)}{F'(\beta)}{\alpha}{\beta}
	mapping $\alpha$ to $\beta$ and extending $\varphi$, i.e., such that $\sigma$ restricted to $F$ is the isomorphism $\varphi$.
\end{thm}
\begin{proof}
	The isomorphism $\varphi$ induces a natural isomorphism from $F[x]$ to $F'[x]$ which maps the maximal ideal $(p(x))$ to the maximal ideal $(p'(x))$.  Taking quotients by these ideals, we have the following isomorphism of fields\[
		F[x]/(p(x)) \rightarrow F'[x]/(p'(x))
	\]
	and as the above fields are isomorphic to $F(\alpha)$ and $F'(\beta)$, respectively.
\end{proof}

In the following, let $K$ be an extension of $F$.
\begin{defn}[Algebraic Elements and Algebraic Extensions]
	An element $\alpha$ in $K$ is said to be \textbf{algebraic} over $K$ if $\alpha$ is a root of some nonzero polynomial $f(x) \in F[x]$.  If $\alpha$ is not algebraic then it is \textbf{transcendental} over $F$.  The extension $K/F$ is said to be \textbf{algebraic} if every element of $K$ is algebraic over $F$.
\end{defn}

\begin{prop}[Minimal polynomials]
	Let $\alpha$ be algebraic over $F$.  Then there is a unique monic irreducible polynomial $\minpoly{\alpha}{F} \in F[x]$ which has $\alpha$ as a root. A polynomial $f(x) \in F[x]$ has $\al$ as a root if and only if $\minpoly{\al}{F}$ divides $f(x)$ in $F[x]$.
\end{prop}
\begin{proof}
	Let $g(x) \in F[x]$ be a polynomial of minimal degree having $\alpha$ as a root.  Multiplying $g(x)$ by a constant, we have $g(x)$ is monic.  Supposing the $g(x)$ were reducible in $F[x]$, then \[
		g(x) = a(x)b(x)
	\] with $a(x), b(x)$ having degrees less than $\deg g(x)$.  Yet as $g(\alpha) = 0$, then either $a(\al)$ or $b(\al)$ are zero, contradicting the minimal degree of $g(x)$.

	Suppose now that $f(x) \in F[x]$ is a polynomial having $\al$ as a root.  By the Euclidean Algorithm in $F[x]$, there are polynomials $q(x), r(x) \in F[x]$ such that \[
		f(x) = q(x) g(x) + r(x)
	\] with $\deg r(x) < \deg g(x)$.  Then $f(\al) = g(\al) q(\al) + r(\al) = r(\al) = 0$, and thus $r(x) = 0$ by minimality of $g(x)$.  Thus, any polynomial $f(x) \in F[x]$ with root $\al$ is divisible by $g(x)$.  
	This proves that $\minpoly{\alpha}{F} = g(x)$, completing the proof.
\end{proof}

\begin{cor}
	If $L/F$ is an extension of fields and $\al$ is algebraic over both $F$ and $L$, then $\minpoly{\al}{L}$ divides $\minpoly{\al}{F}$ in $L[x]$.
\end{cor}

\begin{prop}
	Let $\al$ be algebraic over $F$, and let $F(\al)$ be the field generated by $\al$ over $F$. Then\[
		F(\al) \simeq F[x]/(\minpoly{\al}{F})
	\]
	so that in particular,\[
		\xdeg{F(\al)}{F} = \deg \minpoly{\al}{F} = \deg \al
	\]
\end{prop}

\begin{thm}[Tower Theorem]
	Let $F \subseteq K \subseteq L$ be fields. Then\[
		\xdeg{L}{F} = \xdeg{L}{K} \xdeg{K}{F}
	\]
\end{thm}
\begin{proof}
	The proof proceeds as follows.  Let $(\alpha_i)$ be a basis for $L$ over $K$, and let $(\beta_i)$ be a basis for $K$ over $F$. The elements of $L$ are of the form $\sum a_i \alpha_i$, with $a_i \in K$.  Similarly, the elements $a_i$ are of the form $\sum b_i \beta_i$ with $b_i \in F$.  Thus, elements of $L$ are of the form $\sum c_{ij} \alpha_i \beta_j$ - thus $\alpha_i \beta_j$ span $L$.  Now, consider the linear relation $\sum c_{ij} \al_i \beta_j = 0$.  As the elements $\beta_j$ and $\alpha_i$ are both basis, it can be shown that $c_{ij} = 0$, thus elements $\alpha_i \beta_j$ are a basis for $L$ over $F$, and the theorem follows.
\end{proof}

\begin{defn}[Splitting Field]
	Let $F$ be a field.  The extension field $K$ of $F$ is called splitting field for the polynomial $f(x) \in F[x]$ if $f(x)$ factors completely into linear factors (or \textbf{splits completely}) in $K[x]$ and $f(x)$ does not factor completely into linear factors over any proper subfield of $K$ containing $F$.
\end{defn}

\begin{thm}
	For any field $F$, if $f(x) \in F[x]$, then there exists and extension $K$ of $F$ which is a splitting field for $f(x)$.
\end{thm}

\begin{prop}
	The splitting field for a polynomial of degree $n$ over $F$ is of degree at most $n!$ over $F$.
\end{prop}

\begin{prop}[Uniqueness of splitting fields]
	Any two splitting fields for a polynomial $f(x) \in F[x]$ over a field $F$ are isomorphic.
\end{prop}

\begin{proof}
	Take $\varphi$ to be the identity mapping from $F$ to itself and $E, E'$ in the isomorphism extension theorem to be the two spitting fields for $f(x)$.

	The proof proceeds by inducting on $n$, the degree of the extension of the splitting field.  
\end{proof}

\begin{defn}[Separable Polynomial]
	A polynomial over $F$ is called \textbf{separable} if it has no multiple roots.  A polynomial which is not separable is inseparable.
\end{defn}

\begin{cor}
	Every irreducible polynomial over a field of characteristic 0 (e.g. $\Q, \Z, \R$) is separable.  A polynomial over such a field is separable if and only if it is the product of distinct irreducible polynomials.
\end{cor}

\subsection{Finite Fields} % 
\label{sub:finite_fields}

A finite field  $\F$ is a field with a finite number of elements.  A finite field has characteristic $p$ for some prime $p$, and so is a finite dimensional vector space over $\F_p$. If the dimension of the extension $\xdeg{\F_p}{\F} = n$, then the finite field has $p^n$ elements.
\begin{prop}
	Let $F$ be a field of characteristic $p$. Then for any $a,b \in F$,\[
		(a+b)^p = a^p + b^p \quad \text{and} \quad (ab)^p = a^p b^p
	\]
	Alternatively, the map $\varphi(a) = a^p$ is an injective field homomorphism from $F$ to $F$
\end{prop}
\begin{proof}
	Use the binomial theorem, and note that ${p \choose i}, i = 1,2,\dots,p-1$ is zero in characteristic $p$.
\end{proof}
\begin{defn}
	The map $\varphi(a) = a^p$ is called the \textbf{Frobenius endomorphism} of $F$.
\end{defn}

\begin{cor}
	If $\F$ is finite of characteristic $p$, then every element of $\F$ is a $p$\textsuperscript{th} power in $\F$ - notationally, $\F = \F^p$
\end{cor}
\begin{proof}
	This follows from the injectivity of the Frobenius endomorphism - as $\F$ is finite, an injective function is surjective.
\end{proof}

The field $K$ is said to be \textbf{separable} over $F$ if eery element of $K$ is the root of a separable polynomial over $F$.  Equivalently, the minimal polynomial over $F$ of every element of $K$ is separable.

% subsection finite_fields

\subsection{Cyclotomic Extensions} % 
\label{sub:cyclotomic_extensions}
\begin{defn}[Cyclotomic Extensions]
	The extension $\Q(\zeta_n)/Q$ generated by the $n$\textsuperscript{th} roots of unit over $\Q$ is called a cyclotomic extension.
\end{defn}

\begin{defn}[Cyclotomic Polynomial]
	The $\nth$ cyclotomic polynomial $\varphi_n(x)$ is defined as the polynomial whose roots are the primitive $\nth$ roots of unity, which is of degree $\varphi(n)$.  
\end{defn}
\begin{exmp}
	For $p$ prime, the $p$\textsuperscript{th} cyclotomic polynomial $\Phi_p(x)$ is given by \[
		\Phi_p(x) = \frac{x^p - 1}{x-1}
	\] 
	For example, $\Phi_5(x) = x^4 + x^3 + x^2 + x + 1$.
\end{exmp}
\begin{cor}
	We have that \[
		\Phi_n(x) = \frac{x^n - 1}{\Pi_{d|n} \Phi_d(x)}
	\]
\end{cor}

\begin{thm}
	The cyclotomic polynomial $\Phi_n(x)$ is an irreducible monic polynomial in $\Z[x]$ of degree $\varphi(n)$.
\end{thm}
\begin{proof}
	If $\Phi_n(x)$ is reducible, there exist $f(x), g(x) \in \Z[x]$ such that \[
		\Phi_n(x) = f(x) g(x)
	\] where we take $f(x)$ to be an irreducible factor of $\Phi_n(x)$.  Let $\zeta$ be a primitive $\nth$ root of 1 which is a root of $f(x)$ (so that $f(x)$ is the minimal polynomial of $\zeta$ over $\Q$) and let $p$ denote \textbf{any} prime not dividing $n$.  Then $\zeta^p$ is also a primitive $\nth$ root of 1, and hence is a root of either $f(x)$ or $g(x)$.

	Suppose $g(\zeta^p) = 0$.  Then $\zeta$ is a root of $g(x^p)$, and since $f(x)$ is the minimal polynomial for $\zeta$, $f(x)$ must divide $g(x^p)$ in $\Z[x]$, say \[
		g(x^p) = f(x) h(x)
	\] Reducing modulo $p$ gives \[
		\overline{g}(x^p) = \overline{f}(x^p) \overline{h}(x^p)
	\] in $\F_p[x]$.

	By the remarks of polynomials over finite fields, we have\[
		\overline{g}(x^p) = (\overline{g}(x))^p
	\] so we have the equation \[
		(\overline{g}(x))^p = \overline{f}(x)\overline{h}(x)
	\] in the UFD $\F_p[x]$.  it follows that $\overline{f}(x)$ and $\overline{g}(x)$ have a factor in common in $\F_p[x]$.

	Now, from $\Phi_n(x) = f(x)g(x)$ we see by reducing modulo $p$ that $\overline{\Phi_n}(x) = \overline{f}(x) \overline{g}(x)$, and so we have that $\overline{\Phi_n}(x)$ has a multiple root.  But then also $x^n - 1$ would have a multiple root over $\F_p$ since it has $\overline{\Phi_n}(x)$ as a factor.  This is a contradiction since we have that $x^n - 1$ has $n$ distinct roots over any field of characteristic not dividing $n$.

	Hence $\zeta^p$ must be a root of $f(x)$.  Since this applies to every root $\zeta$ of $f(x)$, we must have that $\zeta^a$ is a root of $f(x)$ for every integer $a$ relatively prime to $n$.  This means that \textbf{every} primitive $\nth$ root of unity is a root of $f(x)$, and thus $f(x) = \Phi_n(x)$, showing the $\Phi_n(x)$ is irreducible.
\end{proof}

% subsection cyclotomic_extensions

\subsection{Constructible Numbers} % 
\label{sub:constructible_number}

We now explore the set of numbers that can be constructed by a ruler and compass.  Constructible numbers are completely classified by the following theorem.

\begin{prop}
	If the element $\alpha \in \R$ is obtained from a field $F \subset \R$ by a series of compass and straightedge constructions then $\xdeg{F(\alpha)}{F} = 2^k$ for some integer $k \geq 0$.
\end{prop}

\begin{proof}
	If a number is constructible, then there is a chain of subfields $\Q \subset F_0 \subseteq F_1 \subseteq \cdots \subseteq F_m$ such that each field $F_i$ is constructed by adjoining the square root of an element $a_{i-1} \in F_{i-1}$, i.e. $F_{i} = F_{i-1}(\sqrt{a_{i-1}})$.  This is an extension of degree one (if $a_{i-1}$ is a square in $F_{i-1}$) or of degree two. Thus, by the Tower Theorem, we have that the overall extension degree is a power of two, and the proposition follows.
\end{proof}
% subsection constructible_number

% section fields

\section{Galois Theory} % 
\label{cha:galois_theory}

\begin{defn}[Automorphism]
	Let $K$ be a field.  
	\begin{itemize}
		\item An isomorphism $\sigma$ of $K$ with itself is called an \textbf{automorphism} of $K$.  The collection of automorphisms of $K$ is denoted $\aut{K}$.  
		\item An automorphism $\sigma \in \aut{K}$ is said to \textbf{fix} an element $\alpha \in K$ if $\sigma(\alpha) = \alpha$.  If $F$ is a subset of $K$, then an automorphism $\sigma$ is said to \textbf{fix} $F$ if it fixes all the elements of $F$. 
	\end{itemize}
\end{defn}

\begin{defn}[]
	Let $K/F$ be an extension of fields.  Let $\aut{K/F}$ be the collection of automorphisms of $K$ which fix $F$.
\end{defn}

\begin{prop}
	$\aut{K}$ is a group under composition and $\aut{K/F}$ is a subgroup.
\end{prop}

\begin{prop}
	Let $K/F$ be a field extension and let $\al \in K$ be algebraic over $F$.  Then for any $\sigma \in \aut{K/F}$,  $\sigma (\al)$ is a root of the minimal polynomial for $\al$ over $F$, i.e., $\aut{K/F}$ permutes the roots of irreducible polynomials.  Equivalently, any polynomial with coefficients in $F$ with $\al$ as a roots has $\sigma(\alpha)$ as a root.
\end{prop}
\begin{proof}
	The proof is simple by noting that $\sigma$ is a homomorphism fixing $F$.
\end{proof}

\begin{prop}
	Let $H \leq \text{Aut}(K)$ be a subgroup of the group of automorphisms of $K$.  Then the collection $F$ of elements of $K$ fixed by all the elements of $H$ is a subfield of $K$.
\end{prop}

\begin{prop}
	Let $E$ be splitting field over $F$ of the polynomial $f(x) \in F[x]$.  Then \[
		|\aut{E/F}| \leq \xdeg{E}{F}
	\] with equality if $f(x)$ is separable over $F$.
\end{prop}

\begin{defn}[Galois Extension]
	Let $K/F$ be a finite extension.  Then $K$ is said to be \textbf{Galois} over $F$ and $K/F$ is a \textbf{Galois} extension if $|\aut{K/F}| = \xdeg{K}{F}$.  If $K/F$ is Galois, then the group of automorphism $\aut{K/F}$ is called the \textbf{Galois group of $K/F$}, denoted $\gal{K}{F}$
\end{defn}

\begin{prop}
	If $K$ is the splitting field over $F$ of a separable polynomial $f(x)$ then $K/F$ is Galois.
\end{prop}

\begin{cor}
	The splitting field of any polynomial over $\Q$ is Galois, since the splitting field of $f(x)$ is clearly the same as the splitting field of the product of the irreducible factors of $f(x)$.
\end{cor}

\begin{defn}[Galois group of a polynomial]
	If $f(x)$ is a separable polynomial over $F$, then the \textbf{Galois group of $f(x)$ over $F$} is the Galois group of the splitting field of $f(x)$ over $F$.
\end{defn}

We now prove a fundamental relation between the orders of subgroups of the automorphism group of a field$K$ and the degrees of the extensions over their fixed fields.
\begin{thm}
	Let $G$ be a subgroup of the automorphisms of a field $K$ and let $F$ be the fixed field.  Then \[
		\xdeg{K}{F} = |G|
	\]

\end{thm}
\begin{prop}
	Let $K/F$ be any finite extension.  Then \[
		| \aut{K/F} \leq \xdeg{K}{F}
	\] with equality if and only if $F$ is the fixed field of $\aut{K/F}$.  Alternatively, $K/F$ is Galois if and only if $F$ is the fixed field of $\aut{K/F}$.
\end{prop}

\begin{thm}
	The extension $K/F$ is Galois if and only if $K$ is the splitting field of some separable polynomial over $F$. Furthermore, if this is the case then every irreducible polynomial with coefficients in $F$ which has a root in $K$ is separable and has all its roots in $K$.
\end{thm}

\begin{thm}[Fundamental Theorem of Galois Theory] Let $K/F$ be a Galois extension and set $G = \gal{K}{F}$.  Then there is a bijection from (subfields $E$ of $K$ containing $F$) and (subgroups $H$ of $G$), given by the correspondences (the field $E$ mapping to the elements of $G$ fixing $E$) and ($H$ mapping to the fixed field of $H$) which are inverse to each other.

	Let $F \subseteq E \subseteq K$ and $1 \leq H \leq G = \gal{K}{F}$, where $E$ is the fixed field of $H$. Under this correspondence,
	\begin{itemize}
		\item $\xdeg{K}{E} = |H|$ and $\xdeg{E}{F} = |G:H|$, the index of $H$ in $G$.
		\item $K/E$ is always Galois, with Galois group $\gal{K}{E} = H$.
		\item $E$ is Galois over $F$ if and only if $H$ is a normal subgroup in $G$.  If this is the case, then the Galois group is isomorphic to the quotient group \[
			\gal{E}{F} \simeq G/H
		\]
	\end{itemize}
\end{thm}

\begin{defn}[Cyclic Extensions]
	An extension $K/F$ is said to be \textbf{cyclic} if it is Galois with a cyclic Galois group.
\end{defn}

\begin{prop}
	Let $F$ be a field of characteristic not dividing $n$ which contains the $n$\textsuperscript{th} roots of unity.  Then the extension $F(\sqrt[n]{a})$ for $a \in F$ is cyclic over $F$ of degree dividing $n$.
\end{prop}
\begin{proof}
	The extension is Galois over $F$ if $F$ contains the $n$\textsuperscript{th} roots of unity since it is the splitting field for $x^n - a$.  For any $\sigma \in \gal{K}{F}$, $\sigma(\sqrt[n]{a})$ is another root of this polynomial, hence $\sigma(\sqrt[n](a)) = \zeta_\sigma \sqrt[n]{a}$ for some root of unity $\zeta_\sigma$.  This gives a map \mapping{\varphi}{\gal{K}{F}}{\mu_n}{\sigma}{\zeta_\sigma}
	where $\mu_n$ denotes the group of $n$\textsuperscript{th} roots of unity.  Since $F$ contains $\mu_n$, every $n$\textsuperscript{th} root of unity is fixed by every element of $\gal{K}{F}$.  Hence \begin{align*}
		\sigma \tau (\sqrt[n]{a}) &= \sigma(\zeta_\tau \sqrt[n]{a}) \\
									&= \zeta_\tau \sigma(\sqrt[n]{a}) \\
									&= \zeta_\tau \zeta_\sigma \sqrt[n]{a} \\
									&= \zeta_\sigma \zeta_\tau \sqrt[n]{a}
	\end{align*}
	which shows that $\zeta_{\sigma \tau} = \zeta_\sigma \zeta_\tau$, so the map above is a homomorphism.  The kernel consists of precisely of the automorphism which fix $\sqrt[n]{a}$, namely the identity.  This gives an injection of $\gal{K}{F}$ into the cyclic group $\mu_n$ of order $n$, which proves the proposition.
\end{proof}

\begin{thm}
	Any cyclic extension of degree $n$ over a field $F$ of characteristic not dividing $n$ which contains the $n$\textsuperscript{th} roots of unity is of the form $F(\sqrt[n]{a})$ for some $a \in F$.
\end{thm}

\begin{defn}[Solvable by radicals]
	An element $\al$ which is algebraic over $F$ can be \textbf{expressed by radicals} or \textbf{solved for in terms of radicals} if $\alpha$ is an element of a field $K$ which can be obtained by a succession of simple radical extensions
\[
	F = K_0 \subset K_1 \subset \cdots \subset K
\]
where $K_{i+1} = K_i(\sqrt[n_i]{a_i})$ for some $a_i \in K_i$, and $\sqrt[n_i]{a_i}$ is some root of the polynomial $x^{n_i} - a_i$.  Such a field $K$ is a \textbf{root extension} of $F$.

A polynomial $f(x) \in F[x]$ can be \textbf{solved by radicals} if all its roots can be solved for in terms of radicals.
\end{defn}

\begin{defn}[Solvable Groups]
	A group $G$ is \textbf{solvable} if there is a chain of subgroups\[
		1 = G_0 \lhd G_1 \lhd G_2 \lhd \cdots \lhd G_s = G
	\] where each $G_i$ is normal in $G_{i+1}$ and the quotient groups $G_{i+1}/G_i$ is abelian for all $i$.
\end{defn}

\begin{cor}
	The finite group $G$ is solvable if and only if for every divisor $n$ of $|G|$ with $\text{gcd}(n, \frac{|G|}{n}) = 1$, $G$ has a subgroup of order $n$.
\end{cor}

\begin{cor} Let $N$ be normal in $G$.
	If $N$ and $G/N$ are solvable, then $G$ is solvable.
\end{cor}

\begin{thm}[Solvability of a polynomial by radicals]
	A polynomial $f(x)$ is solvable by radicals if and only if its Galois group is a solvable group.
\end{thm}

\begin{proof}
	IMPORTANT PROOF
\end{proof}

\begin{cor}
	The general equation of degree $n$ cannot be solved by radicals for $n \geq 5$.  For $n \geq 5$ the group $S_n$ is not solvable.
\end{cor}
% section galois_theory

\end{document}

### 2011-12-12 MATH 3964 - Complex Analysis/MATH 3964 Lecture Notes.tex
Initial Commit

%
%
%  Created by Andrew Tulloch and Giles Gardam on 2009-11-04.
%  Copyright (c) 2009 __MyCompanyName__. All rights reserved.
%!TEX TS-program = xelatex
%!TEX encoding = UTF-8 Unicode

\documentclass[10pt, oneside, reqno]{amsart}
\usepackage{geometry, setspace, graphicx, enumerate}
\onehalfspacing                 
\usepackage{fontspec,xltxtra,xunicode}
\defaultfontfeatures{Mapping=tex-text}

	% AMS Theorems
	\theoremstyle{plain}% default 
	\newtheorem{thm}{Theorem}[section] 
	\newtheorem{lem}[thm]{Lemma} 
	\newtheorem{prop}[thm]{Proposition} 
	\newtheorem*{cor}{Corollary}

	\theoremstyle{definition} 
		\newtheorem{defn}[thm]{Definition}
		\newtheorem{conj}[thm]{Conjecture}
		\newtheorem{exmp}[thm]{Example}

	\theoremstyle{remark} 
		\newtheorem*{rem}{Remark} 
		\newtheorem*{note}{Note} 
		\newtheorem{case}{Case}

		\newcommand{\expc}[1]{\mathbb{E}\left[#1\right]}
		\newcommand{\var}[1]{\text{Var}\left(#1\right)}
		\newcommand{\cov}[1]{\text{Cov}\left(#1\right)}
		\newcommand{\prob}[1]{\mathbb{P}(#1)}
		\newcommand{\given}{ \, | \,}
		\newcommand{\us}{0 \leq u \leq s}
		\newcommand{\ts}[1]{\{ #1 \}}

\newcommand{\dzz}{\, dz}
\newcommand{\bigo}[1]{\mathcal{O}(#1)}
\newcommand{\res}[2]{\text{Res}(#1,#2)}

\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\F}{\mathbb{F}}
\newcommand{\Ga}{\mathbb{G}}

\newcommand{\nth}{n\textsuperscript{th}}

\title{MATH 3964 - Exam Notes}								% Document Title
\author{Andrew Tulloch and Giles Gardam}
%\date{}                                           % Activate to display a given date or no date

\begin{document}
\maketitle

% \tableofcontents

\section{Contour Integration and Cauchy's Theorem} % (fold)

\subsection{Analytic functions} % (fold)
\label{sec:analytic_functions}

\begin{defn}
	A function $f(z)$ is \textbf{differentiable} at $z_0$ if the limit \[
		f'(z_0) = \lim_{\zeta \rightarrow z_0} \frac{f(\zeta) - f(z_0)}{\zeta - z_0}
	\] exists independently of the path of approach.
\end{defn}

\begin{defn}
	A function $f(z)$ is \textbf{analytic} on a region $D$ if it is differentiable everywhere on $D$.  Thus the derivative $f'(z)$ is a function defined on $D$.  a function is analytic at a particular point $z_0 \in \C$ if it is differentiable on some open neighbourhood of $z_0$.
\end{defn}

\begin{thm}
	A necessary condition for $f(z)$ to be analytic is that if $f(z) = u + iv$, then 
	\begin{align*}
		v_y = u_x & v_x = -u_y
	\end{align*}
\end{thm}

\begin{defn}
	A function that is analytic throughout the whole complex plain is called an entire function.
\end{defn}

\begin{defn}
	A point at which a locally analytic function $f(z)$ fails to be analytic is a \textbf{singularity} of $f(z)$.
\end{defn}

The easiest way to construct analytic functions is as sums of convergent power series.  Any power series in the complex domain \[
	f(z) = \sum_{k=0}^\infty a_k (z-z_0)^k
\] having positive or infinite radius of convergence $R$ converges to an analytic function in the interior of its disc of convergence, $|z-z_0| < R$.

\begin{thm}
	Every convergence power series is differentiable term by term in the interior of its disc.
\end{thm}

\begin{prop}
	The $\nth$ derivative of $f(z)$ at $z = z_0$ is \[
		f^{(n)}(z_0)  =n! a_n
	\]
\end{prop}

The radius of convergence is given by either of the equivalent exact formulae:\[
	R = \liminf_{n \rightarrow \infty} |a_n|^{-\frac{1}{n}} 
\]
or \[
	\frac{1}{R} = \limsup_{n \rightarrow \infty} |a_n|^{\frac{1}{n}} 
\]

\begin{thm}
	Every power series with a positive or infinite radius of convergence is differentiable term by term to all orders in the interior of its disc of convergence.
\end{thm}

% section analytic_functions (end)

\subsection{Contour integration} % (fold)
\label{sec:contour_integration}

A contour in the complex-plane is just a curve, finite or infinite, which has an arrow or \textbf{orientation}.  We wish to assign a meaning to the \textbf{contour integral},\[
	\int_C f(z) \dzz
\]
where $C$ is a contour and $f(z)$ is a function which is defined and piecewise continuous along  $C$

\begin{lem}[Triangle inequality for contour integrals]
	\[
	\left| \int_C f(z) \dzz \right| \leq \int_C |f(z)| \, |dz|	
	\]
\end{lem}

\begin{lem}[The ML formula]
	If a contour $C$ has length $L$ and if $|f(z)| \leq M$ on $C$, then \[
		\left| \int_C f(z) \dzz \right| \leq ML
	\]
\end{lem}

\begin{lem}[Jordan's lemma]
	Let $C_R$ be all or part of the semicircular contour $Re^{i \theta}$, where $\theta$ runs from $0$ to $\pi$. Suppose that $|f(z)| \leq M(R)$ on $C_R$ and $\lambda$ is a positive real number.  Let \[
		I(R) = \int_{C_R} f(z) e^{i \lambda z} \dzz
	\]
	Then, we have the bound $|I(R)| = \bigo{M(R)}$
\end{lem}
% section contour_integration (end)

\subsection{Cauchy's theorem and extensions} % (fold)
\label{sec:cauchy_s_theorem_and_extensions}

\begin{thm}[Cauchy's theorem]
	If $f(z)$ is analytic on a simply connected region $D$ and if $C$ is any rectifiable closed contour or cycle in $D$, then \[
		\int_C f(z) \dzz = 0.
	\]
\end{thm}

% section cauchy_s_theorem_and_extensions (end)

\subsection{Cauchy's integral formula} % (fold)
\label{sec:cauchy_s_integral_formula}

\begin{thm}[Cauchy's integral formula]
	Suppose $f(z)$ is analytic in a simply connected region $D$ and that $C$ is a positively oriented rectifiable Jordan curve in $D$.  Then \[
		\frac{1}{2 \pi i} \int_C \frac{f(z)}{z - z_0} \dzz = \begin{cases}
			f(z_0), & z_0 \in C \\
			0,		& z_0 \notin C
		\end{cases}
	\]
\end{thm}

\begin{thm}[Analyticity of Cauchy integrals]
	The function \[
		g(z) = 	\frac{1}{2 \pi i} \int_C \frac{f(\zeta)}{\zeta- z} \, d \zeta
	\]
	is differentiable to all orders in $D$ and is therefore analytic in $D$.
	It's $\nth$ derivative is \[
		g^{(n)}(z) = 	\frac{n!}{2 \pi i} \int_C \frac{f(\zeta)}{(\zeta- z)^{n+1}} \, d \zeta	
	\]
\end{thm}

\begin{thm}
	Suppose that $f(z)$ is analytic in a simply connected region $D$.  Then \[
		f(z) = 	\frac{1}{2 \pi i} \int_C \frac{f(\zeta)}{\zeta- z} \, d \zeta	
	\] and \[
		f^{(n)}(z) = 	\frac{n!}{2 \pi i} \int_C \frac{f(\zeta)}{(\zeta- z)^{n+1}} \, d \zeta	
	\]
\end{thm}

\begin{cor}
	A function $f(z)$ has an antiderivative in a simply connected region $D$ if and only if $f(z)$ is analytic in $D$.
\end{cor}

\begin{thm}[Removable singularities theorem]
	Suppose that $f(z)$ is analytic in a region $D$ except possibly at the point $z_1 \in D$.  At $z_1$, suppose that \[
		\lim_{z \rightarrow z_1} (z - z_1) f(z) = 0
	\]
Then a value of $f(z_1)$ can be assigned so that f(z) becomes analytic at $z_1$.  
\end{thm}

\begin{defn}[Analyticitiy at infinity]
	Suppose that $f(z)$ is analytic on an unbounded set and let $g(z) = f(1/z)$.  Then the point $\infty \in \C^{\star}$ is  point of analyticity of $f(z)$ if $z=0$ is a point of analyticity of $g(z)$.  Similarly, $z = \infty$ is a singularity of a particular type of $f(z)$ if $g(z)$ has a singularity of that same type at $z=0$. 
\end{defn}

\begin{defn}
	 function $f(z)$ which has a singularity at $z_0$ and is analytic in a deleted neighbourhood of $z_0$ has a \textbf{pole} at $z_0$, or more specifically, a \textbf{pole of order $k$}, if $(z-z_0)^k f(z)$ is analytic and nonzero at $z_0$, where $k$ must be a positive integer according to the removable singularities theorem.  A pole of order one is a \textbf{simple pole}, a pole of order two is a \textbf{double pole}, and so on.
\end{defn}

\begin{defn}
	A function is \textbf{meromorphic} if it is analytic in the whole complex plane $\C$ except for poles.  
\end{defn}

% section cauchy_s_integral_formula (end)

\subsection{The Cauchy-Taylor theorem and analytic continuation} % (fold)
\label{sec:the_cauchy_taylor_theorem_and_analytic_continuation}

\begin{thm}[Cauchy-Taylor theorem]
	Suppose that $f(z)$ is analytic at $z_0$ and the disc $D(R) = B(z_0,R)$ is the largest open disc on which $f(z)$ is analytic.  Then the Taylor series, \[
		f(z) = \sum_{k=0}^\infty \frac{f^{(k)}(z_0)}{k!} (z - z_0)^k
	\]
	converges absolutely to $f(z)$ on $D(R)$ and uniformly on compact subsets.  
\end{thm}

\begin{thm}
	Let $f(z) = \sum_{k=0}^\infty a_k (z-z_0)^k$ have a finite radius of convergence $R$.  The radius of convergence of a power series is the distance from the centre to the nearest singularity of its sum function $f(z)$.  
\end{thm}

\begin{thm}[Cauchy's inequality]
	Let $f(z)$ analytic on an open disc $D(\rho)$ with centre $z_0$.  Then, if $|f(z)| \leq M(\rho)$, we have \[
		\left|f^{(n)}(z_0) \right| \leq \frac{n! M(\rho)}{\rho^n}
	\]
	If a power series $\sum_{k=0}^\infty a_k (z-z_0)^k$ converges to $f(z)$, then \[
		|a_n| \leq \frac{M(\rho)}{\rho^n}
	\]
\end{thm}[Liouville's theorem]
If an entire function is bounded, or if it possibly grows at a rate such that $f(z)/z \rightarrow 0$ uniformly as $z \rightarrow \infty$, then $f(z)$ is constant.

\begin{thm}[Uniqueness of analytic continuation]
	Suppose that $f(z), g(z)$ are analytic in a common region $D$.  Let $H$ be a subset of $D$ that contains a convergent subsequence $\{z_k \}$ whose limit is in the interior of $D$.  If $f(z) = g(z), z \in H$, then $f(z) = g(z)$ everywhere in $D$.  
\end{thm}

% section the_cauchy_taylor_theorem_and_analytic_continuation (end)

\subsection{Laurent's theorem and the residue theorem} % (fold)
\label{sec:laurent_s_theorem_and_the_residue_theorem}

\begin{thm}[Laurent's theorem]
	Suppose that $f(z)$ is analytic in the open circular annulus $R_1 \leq |x - x_0| \leq R_2$.  Then $f(z)$ admits a power series expansion in both positive and negative powers (a \textbf{Laurent series}), \[
		f(z) = \sum_{k=-\infty}^\infty a_k (z-z_0)^k
	\] which is absolutely convergent in $D$ and uniformly convergent on compact subsets.

	A formula for the coefficient $a_n$ is \[
		a_n = \frac{1}{2\pi i} \int_{C(r)} \frac{f(\zeta)}{(\zeta - z_0)^{n+1}} \, d \zeta
	\]
\end{thm}

\begin{defn}
	If $z_0$ is a pole or isolated essential singularity of $f(z)$, then the \textbf{residue} of $f(z)$ at $z_0$ is the coefficient $a_{-1}$ of $(z-z_0)^{-1}$ in the Laurent expansion of $f(z)$ about $z_0$.  The notation for the residue is $\res{f}{z_0}$.
\end{defn}

\begin{thm}[Picard's first theorem]
	A non-constant entire function has an image either the whole of the complex plane, with at most one exception.
\end{thm}

\begin{lem}
	If $f(z)$ has a simple pole at $z_0$, then \[
		\res{f}{z_0} = \lim_{z \rightarrow z_0} (z - z_0) f(z) = \left[(z-z_0) f(z) \right]_{z = z_0}
	\]
\end{lem}

\begin{lem}
	If $f(z)$ has a pole of order $k$ at $z_0$, then \[
		\res{f}{z_0} = \frac{1}{(k-1)!}\left[\frac{d^{k-1}}{dz^{k-1}} \left((z-z_0)^k f(z)\right) \right]_{z = z_0}
	\]
\end{lem}

\begin{lem}
	If $f(z)$ and $h(z)$ are analytic at $z_0$ and $h(z)$ has a simple zero at $z_0$, then \[
		\res{g/h}{z_0} = \frac{g(z_0)}{h'(z_0)}
	\]
\end{lem}

\begin{thm}[Residue theorem]
	Suppose that $f(z)$ is analytic inside an on a curve $C$ except for a finite number of poles or isolated essential singularities $z_i$ inside $C$.  Then \[
		\int_C f(z) \dzz = 2 \pi i \sum_{i} \res{f}{z_i}
	\]
\end{thm}

\begin{defn}[Residue at infinity]
	Suppose that $f(z)$ is analytic everywhere outside of a bounded region, in which it admits a convergent Laurent series.  Then the residue of $f(z)$ at infinity is minus the coefficient $a_{-1}$, that is,\[
		\res{f}{\infty} = -a_{-1}
	\]
\end{defn}
% section laurent_s_theorem_and_the_residue_theorem (end)

\subsection{The Gamma function $\Gamma(z)$} % (fold)
\label{sec:the_gamma_function_gamma_z_}

\begin{defn}
	For $\Re(z)> 0$, define \[
		\Gamma(z) = \int_0^\infty t^{z-1}e^{-t} dt
	\]
	The integral converges for $\Re(z)> 0$, and uniformly for $\Re(z) \leq \delta > 0$.  Hence $\Gamma(z)$ is analytic at least in the half-plane $\Re(z) > 0$.  
\end{defn}

\begin{lem}[Recurrence relation]
	Integration by parts gives $\Gamma(z+1) = z \Gamma(z)$. This gives the analytic continuation to $\Re(z) > -1$.  Repeated application provides the analytic continuation to the whole plane except for simple poles at $z = 0, -1, -2, \dots$.
\end{lem}

\begin{defn}[Beta function]
	For $\Re(\alpha), \Re(\beta) > 0$, define the function \[
		B(\alpha, \beta) = \int_0^1 t^{\alpha - 1} (1-t)^{\beta - 1} \, dt
	\]
\end{defn}

\begin{thm}
We have the following relation
\[ B(\alpha, \beta) = \frac{\Gamma(\alpha) \Gamma(\beta)}{\Gamma(\alpha + \beta)}
\]
\end{thm}

\begin{lem}[Duplication formula]
	\[
		\Gamma(\alpha + \frac{1}{2}) = \frac{\Gamma(2\alpha) \sqrt(\pi)}{2^{2\alpha - 1} \Gamma(\alpha)}
	\]
\end{lem}

\begin{lem}[Functional relation]
	\[
		\Gamma(z) \Gamma(1-z) = \frac{\pi}{\sin \pi z}
	\]
\end{lem}

% section the_gamma_function_gamma_z_ (end)

\subsection{The residue theorem} % (fold)
\label{sec:the_residue_theorem}

\begin{lem}
	In all cases, including isolated essential singularities, \[
		\res{f}{z_0}
	\] is the coefficient of $\epsilon^{-1}$ in the Laurent expansion \[
		f(z_0 + \epsilon) = \sum_{n \in \Z} a_n \epsilon^n
	\]
\end{lem}

\begin{defn}[Residue at infinity]
	Suppose that $f(z)$ is analytic everywhere outside of a bounded region, in which it admits a convergent Laurent series.  Then the residue of $f(z)$ at infinity is minus the coefficient $a_{-1}$, that is,\[
		\res{f}{\infty} = -a_{-1}
	\]

	If $f(z) \sim \frac{K}{|z^2|}$ as $|z| \rightarrow \infty$, then the residue at $\infty$ vanishes.
\end{defn}

\begin{thm}[Argument principle]
	Suppose that $f(z)$ is analytic on and inside a positively oriented simple closed contour $C$.  Suppose also that $f(z) \neq 0$ on $C$.  Then \[
		\frac{1}{2\pi} \Delta_C \arg f(z) = N - P
	\] where $N$ is the total number of \textbf{zeroes} and $P$ is the total number of \textbf{poles} inside $C$, counting multiplicities.
\end{thm}

\begin{exmp}[Integration of rational functions]
	Let $P(x)$ and $Q(x)$ be polynomials with $\deg P(x) = \deg Q(x) - 1$.  Then \[
		\int_\R \frac{P(x)}{Q(x)} \, dx 
 	\] does not exist in the usual sense, but the limit \[
 		\lim_{R \rightarrow \infty} \int_{-R}^R \frac{P(x)}{Q(x)} \, dx
 	\] does exist.  This limit is called the principle value of the integral.

 	We then have 
\[
	P \int_{\R} \frac{P(x)}{Q(x)} \, dx = \pi i \,(\text{residue at $\infty$}) + 2 \pi i \sum (\text{residues in the upper half-plane})
\]
\end{exmp}

\begin{exmp}[Poles on the real axis]
	Let $f(x)$ be analytic with a pole on the real axis at $z = x_0$.  Then we have \[
		P \int_a^b f(x) = \int_{C^+} f(z) dz + \pi i \res{f}{z_0}
	\]
\end{exmp}

\begin{exmp}[Integrals of trigonometric functions]
	Let $R(x,y)$ be a rational function bounded on the circle $|z| = 1$.  Then \[
		\int_0^{2 \pi} R(\cos \theta, \sin \theta) \, d \theta = \int_{|z| = 1} \frac{1}{iz} R\left( \frac{z^2 + 1}{2z}, \frac{z^2 - 1}{2 i z} \right) \, dz
	\] 
\end{exmp}

\begin{exmp}
	The integrals \begin{align*}
		\int_\R f(x) \cos \alpha x \, dx \\
		\int_\R f(x) \sin \alpha x \, dx
	\end{align*}
	are the real and imaginary parts of \[
		I = \int_\R f(x) e^{i \alpha x} \, dx
	\]

\end{exmp}

% section the_residue_theorem (end)

% chapter contour_integration (end)
\section{Analytic Theory of Differential Equations} % (fold)
\label{cha:analytic_theory_of_differential_equations}

\subsection{Existence and uniqueness} % (fold)
\label{sec:existence_and_uniqueness}

\begin{thm}[Existence and uniqueness of first order differential equations]
	Let $f$ be an analytic function of two complex variables in the open polydisc $|z - z_0| < a$, $|w - w_0| < b$.
	The first order differential equation,
	\[
		\frac{dw}{dz} = f(z,w)
	\] has a unique analytic solution $w = w(z)$ such that $w = w_0$ when $ z = z_0$ in some disc $|z - z_0| < h, 0 < h \leq a$
\end{thm}

\begin{proof}
	By repeated differentiation, we can constrict the formal Taylor series \[
		w(z) = w_0 + \sum_{n=1}^\infty \frac{w^{(n)}(z_0)}{n!} (z - z_0)^n
	\] in terms of $f$ and its derivatives, which satisfies the differential equation.  We can then show that this series has a positive radius of convergence.
\end{proof}

\subsection{Singular point analysis of nonlinear differential equations} % (fold)
\label{sec:singular_point_analysis_of_nonlinear_differential_equations}

\begin{defn}[Singular points of differential equations and their solutions]
	Consider the $\nth$-order differential equation,\[
		w^{(n)} = f(z, w, w', w'', \dots, w^{(n-1)})
	\] where $f$ is a locally analytic function of its $n$ complex arguments.  A singular point of the differential equation is a point \[
		(z_0, w_0, w'_0, \dots, w_0^{(n-1)}) \in \C^n
	\]
	at which $f$ is \textbf{not analytic} or a point where one or more of the arguments is \textbf{infinite}.

	A \textbf{fixed} singularity of a differential equation is a singular hyperplane $z = z_0$ or $z = \infty$
\end{defn}

\begin{defn}[Classification of singularities of solutions $w(z)$]
	Singularities of the solutions are classified as \textbf{fixed} or \textbf{movable}.  A \textbf{movable singular point} is a singular point of $w(z)$ depending on one or more integration constants. 
	A \textbf{fixed singular point} of $w(z)$ is independent of the integration constants and is included among the fixed singularities of the differential equation.
\end{defn}

\begin{exmp}
	\begin{enumerate}[(i)]
		\item \[
			w' = w^2
		\] with solution $w = -\frac{1}{z - C}$.  This has a movable pole at $z = C$.

	\item \[
		w' = \frac{1}{w}
	\] with solution $w = \pm \sqrt{2(z-C)}$.  This has a moveable quadratic branch point at $z = C$.

	\item \[
		w'' = \left(w'\right)^2
	\] with solution $w = -\log (z - C_1) + C_2$, movable logarithmic branch point at $z = C_1$.

	\end{enumerate}

\end{exmp}

% section existence_and_uniqueness (end)
\subsection{Painlev\'e transcendents} % (fold)
\label{sec:painlev'e_transcendents}

Consider the differential equation \[
	w'' = 6 w^2 + g(z)
\]
Attempting a Laurent-type expansion about a movable singularity $z = z_0$ leads us to find that the movable terms balance if and only if $p = 2$, $a_0 = 1$.  We then find the recurrence relation for $a_n$.  We find that it is of the form \[
	(n+1)(n-6) a_n = f_n(a_0, a_1, \dots)
\] and so there is a possible obstruction at $n = 6$ - a \textbf{resonance number}.  To resolve this, we introduce \textbf{logarithmic terms}.

This introduces a \textbf{log-pole} at $z_0$, a logarithmic branch point.  To avoid this, we find that we must set $g''(z_0) = 0$, and hence our original differential equation is \[
	w'' = 6 w^2 + \alpha z + \beta
\] called the Painlev\'e-I transcendent.  When $\alpha = 0$, the system admits the first integral $(w')^2 = 4w^3 + 2 \beta w + K$, which is solved with a \textbf{Weierstrass elliptic function}, having one double pole in each period parallelogram.

Consider the differential equation \[
	w'' = 2w^3 + C(z) w + D(z).
\]
Similar analysis yields that $C''(z) = 0$, $D'(z) = 0$.  Hence, the differential equation \[
	w'' = 2w^3 + (\alpha z + \beta)w + \gamma
\]  When $\alpha = 0$, it has the first integral \[
	(w')^2 = w^4 + \beta w^2 + 2 \gamma w + K
\] which is solved by \textbf{Jacobi elliptic functions}.  Each period parallelogram has two simple poles, one with residue $a_0 = 1$, the other with $a_0 =  -1$.  When $\alpha \neq 0$, the DE above defines a new function known as the Painlev\'e-II transcendent.

% section singular_point_analysis_of_nonlinear_differential_equations (end)

% section painlev'e_transcendents (end)

\subsection{Fuchsian theory} % (fold)
\label{sec:fuchsian_theory}
\[
	w^{(n)} + p_1(z) w^{(n-1)} + \dots + p_n(z) w = R(z)
\]
The \textbf{fixed singularities} of the solution $w(z)$ are included among the singularities of the $p_i(z), i = 1, 2, \dots, n$, and $R(z)$ and possibly $z = \infty$.  Linear DE's \textbf{cannot have movable singularities}.

\begin{thm}
	Suppose $z_0$ is not a singularity of the $p_i(z)$ or $R(z)$.  Then the solution of the above linear DE satisfying initial conditions $w^(i) = w^(i)_0$ is analytic in a disc centred at $z_0$ with radius equal to the distance between $z_0$ and the nearest singularity of the $p_i(z)$ and $R(z)$. 
\end{thm}

\begin{defn}[Regular singular points]
	Consider the linear homogenous DE \[
		w^{(n)} + p_1(z) w^{(n-1)} + \dots + p_n(z) w = 0
	\] where the $p_i(z)$ are rational functions.  The DE has a regular singular point at $t = z_1$ if 
	\begin{itemize}
		\item at least one of the $p_i$ has a pole at $z = z_1$
		\item the \textbf{order} of the pole of $p_i(z)$ at $z = z_1$ is at most $i$ for all $i$.
	\end{itemize}
	We have $z = \infty$ is a \textbf{regular singular point} if \[
		p_i(z) = \bigo{\frac{1}{z^i}}
	\]  as $z \rightarrow \infty$ for all $i$.
\end{defn}

Near a regular singular point at $z = z_1,$ one or more particular solutions can be constructed as a power of $z - z_1$ times a convergent power series;
\[
	(z - z_0)^p \left\{ a_0 + a_1 (z - z_0) + \dots \right\}
\]
The leading powers $p$ satisfy the \textbf{indicinal equation} \[
	p(p-1)(p-2)\dots(p-2+1) + q_1 p(p-1)(p-2)\dots(p-n+2) + q_{n-1}p + q_n = 0
\] where $q_i = \lim_{z \rightarrow z_1}(z - z_1)^i p_i (z)$

\begin{defn}[Fuchsian differential equation]
	A Fuchsian DE is a linear homogenous DE  all of whose singular points are regular. 
\end{defn}

Hence \begin{itemize}
	\item all the $p_i(z)$ are rational functions,
	\item If $z_1$ is a pole of any of the $p_i$ then the order of the pole in $p_i$ is less than $p_i$ for all $i$,
	\item as $z \rightarrow \infty$, \[
		p_i(z) = \bigo{\frac{1}{z^i}}
	\]
\end{itemize}

\begin{defn}[M\"obius transformations]
	Fuchsian character is preserved under $\bar z = \frac{az + b}{cz + d}$ with $ad - bc \neq 0$.  The singular points change position, but their exponents are not affected
\end{defn}

If $z_1$ is a regular singular point, the transformation \[
	\bar w = (z - z_1)^\lambda w
\] has the following effect: 
\begin{itemize}
	\item all the exponents at $z_1$ are raised by $\lambda$,
	\item all the exponents at $\infty$ are lowered by $\lambda$,
	\item exponents are other points are not affected.
\end{itemize}

\begin{thm}[Quadratic transformation]
	The transformation $z = \bar z ^2$ has the following effect:
	\begin{itemize}
		\item Exponents at $z = 0$ and $z = \infty$ are doubled.
		\item A singular point at $z+1 \neq -$ splits into a pair of singular points at $\pm \sqrt{z_1}$ with the \textbf{same exponents}
	\end{itemize}
\end{thm}

% section fuchsian_theory (end)

\subsection{Hypergeometric functions} % (fold)
\label{sec:hypergeometric_functions}
The hypergeometric differential equation \[
	z(1-z) w'' + (\gamma - (1 + \alpha + \beta)z)w' - \alpha \beta w = 0
\] is a Fuchsian DE with regular points at 0, 1 and $\infty$, with exponents
\begin{itemize}
	\item[$z = 0$] exponents $0, 1 - \gamma$,
	\item[$z = 1$] exponents $0, \gamma - \alpha - \beta$,
	\item[$z = \infty$] exponents $\alpha, \beta$
\end{itemize}

The \textbf{hypergeometric function} $F(\alpha, \beta; \gamma; z)$ is the particular solution \[
	F(\alpha, \beta; \gamma, z) = \sum_{n = 0}^\infty \frac{(\alpha)_n (\beta)}{(\gamma)_n n!} z^n 
\] where $(\alpha)_n = \alpha (\alpha + 1)\dots(\alpha + n - 1) = \frac{\Gamma(\alpha + n)}{\Gamma(\alpha)}$.

The general solution of the hypergeometric equation is \[
	y  = C_1 F(\alpha, \beta; \gamma, z) + C_2 z^{1 - \gamma} F(1 + \alpha - \gamma, 1 + \beta - \gamma; 2 -\gamma; z)
\]

We have an integral formula \[
	F(\alpha, \beta; \gamma; z) = \frac{\Gamma(\gamma)}{\Gamma(\gamma - \beta)\Gamma(\beta)} \int_0^1 t^{\beta - 1} (1-t)^{\gamma - \beta - 1} (1- zt)^{- \alpha} \,dt
\] valid for $\Re \gamma > \Re \beta > 0$, $|z| < 1$. but can be extended to $z \in \C = [1,\infty)$.

\subsection{Gauss' formula at $z = 1$} % (fold)
\label{sub:gauss_formula_at_z_1_}
	\[
		F(\alpha, \beta; \gamma; 1) = \frac{\Gamma(\gamma)\Gamma(\gamma - \alpha - \beta)}{\Gamma(\gamma - \alpha)\Gamma(\gamma - \beta)}
	\] 
\subsection{Kummer's formula at $z = -1$} % (fold)
\label{sub:kummer_s_formula_at_z_1_}
\[
	F(\alpha, \beta; 1 + \beta - \alpha; -1) = \frac{1}{2} \frac{\Gamma(\frac{1}{2}\beta)\Gamma(1 + \beta - \alpha)}{\Gamma(\beta) \Gamma(1 + \frac{1}{2}\beta - \alpha)}
\]

\subsection{Evaluation at $z = \frac{1}{2}$} % (fold)
\label{sub:evaluation_at_z_1_2_}
	\[
		F(\alpha, \beta; \gamma; \frac{1}{2}) = 2^\alpha F(\alpha, \gamma - \beta; \gamma; -1)
	\]
	\[
		F(\alpha, \beta; \gamma; \frac{1}{2}) = \frac{\Gamma(\frac{1}{2}\gamma) \Gamma(\frac{1}{2}\gamma + \frac{1}{2})}{\Gamma(\frac{1}{2}\alpha + \frac{1}{2}\gamma)\Gamma(\frac{1}{2} - \frac{1}{2}\alpha + \frac{1}{2}\gamma)}
	\]
	\[
		F(\alpha, \beta; \frac{1}{2}(\alpha + \beta +1); \frac{1}{2}) = \frac{ \sqrt{\pi} \Gamma(\frac{1}{2}\alpha + \frac{1}{2}\beta + \frac{1}{2})}{\Gamma(\frac{1}{2}\alpha + \frac{1}{2})\Gamma(\frac{1}{2}\beta + \frac{1}{2})}
	\]

% subsection evaluation_at_z_1_2_ (end)

% subsection kummer_s_formula_at_z_1_ (end)
% subsection gauss_formula_at_z_1_ (end)

\subsection{Connection formulae} % (fold)
\label{sub:connection_formulae}
We have \begin{align*}
	F(\alpha, \beta; 1 + \alpha + \beta - \gamma; 1 - z) = \\
	 A \,F(\alpha, \beta; \gamma ;z) + \\
	 B \, z^{1 - \gamma} F(1 + \alpha - \gamma, 1+ \beta - \gamma; 2 - \gamma; z)
\end{align*}
where \[
	A = \frac{\Gamma(1 - \gamma)\Gamma(1 + \alpha + \beta - \gamma)}{\Gamma(1 + \alpha - \gamma)\Gamma(1 + \beta - \gamma)}
\]\[
	B = \frac{\Gamma(\gamma - 1)\Gamma(1 + \alpha + \beta - \gamma)}{\Gamma(\alpha)\Gamma(\beta)}
\]

\begin{align*}
	F(\alpha, \beta; \gamma ; z) = \\
	A_1 F(\alpha, \beta; 1 + \alpha + \beta - \gamma; 1 -z) + \\
	 B_1 (1 - z)^{\gamma - \alpha - \beta} F(\gamma - \alpha, \gamma - \beta; 1 - \alpha - \beta + \gamma; 1 - z)	
\end{align*}
where \[
	A_1 = \frac{\Gamma(\gamma)\Gamma(\gamma - \alpha - \beta)}{\Gamma(\gamma - \alpha)\Gamma(\gamma - \beta)}
\]\[
	B_1 = \frac{\Gamma(\gamma)\Gamma(\alpha + \beta - \gamma)}{\Gamma(\alpha)\Gamma(\beta))}
\]

There are various similar formula on page 44 of the notes.
% subsection connection_formulae (end)
% subsection subsection_name (end)
% section hypergeometric_functions (end)

\subsection{Elliptic functions} % (fold)
\label{sec:elliptic_functions}

There are three main approaches to the study of \textbf{elliptic functions}:
\begin{itemize}
	\item Inversion of elliptic integrals;
	\item Nonlinear DE's;
	\item Doubly periodic meromorphic functions.
\end{itemize}

\subsection{Elliptic integrals} % (fold)
\label{sub:elliptic_integrals}
Consider the class of indefinite integrals, \[
	\int R(x, \sqrt{P(x)}) \, dx ,
\] where $R$ is a rational function of two variables and $P$ is a polynomial without square factors.  When $P(x)$ has degree 3 or 4, a M\"obius transform of the polynomial $P(x)$ can be given one of several equivalent normalisations:
\begin{itemize}
	\item[\textbf{Jacobi}] $P(x) = (1-x^2(1-k^2 x^2))$,
	\item[\textbf{Weierstrass}] $P(x) = 4x^3 - g_2 x - g_3$
\end{itemize}

\textbf{Elliptic integral} of the \textbf{first kind:} \[
	F(k, \phi) = \int_0^\phi \frac{d\theta}{\sqrt{1 - k^2 \sin^2 \theta}}
\]
\textbf{Elliptic integral} of the \textbf{second kind:} \[
	E(k, \phi) = \int_0^\phi \sqrt{1 - k^2 \sin^2 \theta} \, d \theta
\]
\textbf{Elliptic integral} of the \textbf{third kind:} \[
	\Pi(n,k \phi) = \int_0^\phi \frac{d \theta}{(1 + n^2 \sin^2 \theta) \sqrt{1 - k^2 \sin^2 \theta}}
\]

\textbf{Complete elliptic integral of the first kind:}\begin{align*}
	K(k) &= F(k, \frac{\pi}{2}) \\
			&= \int_0^1 \frac{dx}{\sqrt{1-x^2}\sqrt{1 - k^2 x^2}} \\
			&= \frac{1}{2} \pi F(\frac{1}{2}, \frac{1}{2}, 1; k^2)
\end{align*}

\textbf{Complete elliptic integral of the second kind:}\begin{align*}
	E(k) &= E(k, \frac{\pi}{2}) \\
			&= \int_0^1 \frac{\sqrt{1-k^2x^2}}{\sqrt{1-x^2}} \, dx \\
			&= \frac{1}{2} \pi F(-\frac{1}{2}, \frac{1}{2}, 1; k^2)
\end{align*}

\subsection{Inversion of elliptic integrals} % (fold)

\newcommand{\sn}{\text{sn}\,}
\newcommand{\cn}{\text{cn}\,}
\newcommand{\dn}{\text{dn}\,}

\label{sub:inversion_of_ellitpic_integrals}
The \textbf{Jacobi elliptic function} $\sn z$ is defined by \[
	\int_0^{\sn z} \frac{dt}{\sqrt{1 - t^2}\sqrt{1 - k^2 t^2}} = z
\] or equivalently, by the differential equation \[
	(w')^2 = (1-w^2)(1 - k^2 w^2) 
\] with $w(0) = 0$ and $w'(0) > 0$.

Then we define $\cn z, \dn z$ by \[
	\cn z = \sqrt{1 - \sn^2 z}
\]\[
	\dn z = \sqrt{1 - k^2 \sn^2 z}
\]

\subsection{Doubly periodic meromorphic functions} % (fold)
\label{sub:doubly_periodic_meromorphic_functions}
 We define \[
 	\wp(z) = \frac{1}{z^2} + \sum_{\substack{m,n \in Z \\ m,n \neq 0}} \left( \frac{1}{z - mw - nw')^2} - \frac{1}{(mw + nw')^2} \right)
 \] where $w$ and $w'$ are nonzero complex numbers with $w'/w$ not real.  It is easy to see that $\wp(z)$ is globally meromorphic with periods $w$ and $w'$.

If $f$ is any elliptic function, then $\int_C f(z) = 0$, where $C$ is the period parallelogram, as the integrals on opposite sides cancel.  Hence, the sum of the residues of all the poles in a period parallelogram.

We have \[
	\wp'(z) = -2 \sum_{m,n \in \Z} \frac{1}{(z - mw - nw')^3}
\]

Consider the \textbf{Laurent expansion} of $\wp$ and $\wp'$ about $z = 0$.  We have \[
	\frac{1}{(z - mw - nw')^2} = \sum_{k = 0}^\infty \frac{(k+1)z^k}{(mw - nw'){k+2}}
\]
Defining $I_{2k} = \sum' \frac{1}{(mw + nw')^{2k}}$ gives \[
	\wp(z) = \frac{1}{z^2} + 3I_4 z^2 + 5 I_6 z^4 + \dots
\]
\[
	\wp'(z) = -\frac{2}{z^3} + 6I_4 z + 20 I_6 z^3
\] where the radius of convergence is the minimum of $|w|$ and $|w'|$.

Letting $g_2 = 60 I_4$, $g_3 = 140 I_6$, we obtain the DE \[
	(w')^2 = 4w^3 - g_2 w - g_3
\] with general solution $\wp(z - z_0)$.

We also have \[
	\int^{\wp(z)} \frac{dt}{\sqrt{4t^3 - g_2 t - g_3}} = z
\]

\subsection{Jacobi elliptic functions} % (fold)
\label{sub:jacobi_elliptic_functions}
 We have \begin{align*}
 	\sn' z = \cn z \dn z \\
	\cn' z = - \sn z \dn z \\
	\dn' z = - k^2 \sn z \cn z
 \end{align*} deduced from $\sn z$ satisfying $w' = \sqrt{1 - w^2}\sqrt{1 - k^2 w^2}$.

For identities of the Jacobi elliptic functions, see pages 57-67 in the notes.

\subsection{Addition theorems} % (fold)
\label{sub:addition_theorems}
See pages 57-67 in the notes

\subsection{Liouville theory} % (fold)
\label{sub:liouville_theory}
Let an elliptic function be defined as any \textbf{doubly periodic meromorphic function}.  
\begin{defn}[Order]
	The order of an elliptic function is the number of poles of $f(z)$ inside a period parallelogram, counting multiplicities (a pole of order $n$ is counted as $n$ poles.)  Then $\wp, \sn, \cn$ are elliptic functions of order $2$.  
\end{defn}

\begin{thm}
	An elliptic function of order zero is constant.
\end{thm}

\begin{thm}
	The sum of the residues of $f(z)$ at all poles in a period parallelogram is zero.
\end{thm}

\begin{thm}
	The transcendental equation $f(z) = a$ where $f(z)$ is an elliptic function of order $m$ has exactly $m$ roots in every period parallelogram, counting multiplicities, for every $a \in \C$.  
\end{thm}

\begin{thm}
	A M\"obius transformation leaves the order and periods of an elliptic function unchanged.  
\end{thm}
\begin{thm}
	Any solution of $(w')^2 = aw^4 + bw^3 + cw^2 +dw + e$ with $a,b$ both not zero, no square factors, is an elliptic function of order 2.
\end{thm}

\subsection{Weierstrass' Theorem} % (fold)
\label{sub:weierstrass_theorem}
	Let $f(z)$ be an elliptic function of any order.  Then it has a unique representation \[
		f(z) = R_1(\wp(z)) + R_2(\wp(z)) \wp'(z)
	\] where $\wp(z)$ is the Weierstrass function having the same periods and $R_i$ rational functions.

% subsection weierstrass_theorem (end)% subsection liouville_theory (end)
% subsection addition_theorems (end)
% subsection jacobi_elliptic_functions (end)
% subsection doubly_periodic_meromorphic_functions (end)
% subsection inversion_of_ellitpic_integrals (end)
% subsection elliptic_integrals (end)

% section elliptic_functions (end)
% chapter analytic_theory_of_differential_equations (end)

\end{document}

### 2011-12-12 MATH 3968 - Differential Geometry/MATH 3968 Lecture Notes.tex
Initial Commit

%
%  untitled
%
%  Created by Andrew Tulloch on 2009-10-30.
%  Copyright (c) 2009 __MyCompanyName__. All rights reserved.
%%!TEX TS-program = xelatex
%!TEX encoding = UTF-8 Unicode

\documentclass[10pt, oneside, reqno]{amsart}
\usepackage{geometry, setspace, graphicx, enumerate}
\onehalfspacing                 
\usepackage{fontspec,xltxtra,xunicode}
\defaultfontfeatures{Mapping=tex-text}

\newcommand{\deriv}[1]{\frac{\partial}{\partial u^{#1}}}
\newcommand{\christ}[3]{\ensuremath{\Gamma^{#1}_{#2#3}}}
\newcommand{\R}{\mathbb{R}}

		\newcommand{\cov}{\text{cov}}
			% AMS Theorems
			\theoremstyle{plain}% default 
			\newtheorem{thm}{Theorem}[section] 
			\newtheorem{lem}[thm]{Lemma} 
			\newtheorem{prop}[thm]{Proposition} 
			\newtheorem*{cor}{Corollary}

			\theoremstyle{definition} 
				\newtheorem{defn}[thm]{Definition}
				\newtheorem{conj}[thm]{Conjecture}
				\newtheorem{exmp}[thm]{Example}

			\theoremstyle{remark} 
				\newtheorem*{rem}{Remark} 
				\newtheorem*{note}{Note} 
				\newtheorem{case}{Case}

\title{MATH 3968 Notes}								% Document Title
\author{Andrew Tulloch}
%\date{}                                           % Activate to display a given date or no date

\begin{document}
\maketitle

\section{Curves in $\mathbb{R}^N$}
\begin{defn}
	A curve is said to be \emph{regular} if $\alpha'(t) \neq 0$ for all $t$
\end{defn}

\begin{defn}
	Let $\alpha$ be a curve parametrized by arc length.  The number $| \alpha''(s)| = k(s)$ is called the \emph{curvature} of $\alpha$ at $s$.
\end{defn}

\begin{defn}[Frenet equations]
	\begin{align*}
		\mathbf{t'} = k \mathbf{n} \\
		\mathbf{n'} = -\mathbf{k}t + \tau \mathbf{b}\\
		\mathbf{b'} = - \tau \mathbf{n}\\
	\end{align*}	
\end{defn}

\begin{defn}
	Total curvature of a unit speed curve $\alpha(s)$ from $a$ to $b$ is defined as \[
		\int_a^b k ds = \theta(b)-\theta(a)
	\]
\end{defn}

\begin{defn}
	The winding number $k$ is defined as the integer such that, for a closed curve, the total curvature \[
		\int_a^b k ds = 2 \pi k
	\]
\end{defn}

\begin{thm}[Rotation Theorem]
	For a piecewise smooth simple closed curve, traversed in an anti-clockwise direction around a bounded region of the plane, the winding number is 1
\end{thm}

\section{General Analysis}
\begin{defn}[Homeomorphism]
	A map is a \emph{homeomorphism} is it is continuous and has a continuous inverse.
\end{defn}

\begin{defn}[Diffeomorphism] 
	A map is a \emph{diffeomorphism} is it is smooth and has a smooth inverse
\end{defn}

\begin{defn}[Differential of a smooth map]
	The differential of a smooth map $\varphi: \R^N \rightarrow \R^M$ is defined as follows.  Let $w \in R^M$, and let $\alpha$ be a differentiable curve such that $\alpha(0) = p$, $\alpha'(0) = w$.  Then the composition $\beta = \varphi \circ \alpha$ is differentiable, and we have \[
		d \varphi_p (w) = \beta'(0) = (\varphi \circ \alpha)'(0)
	\]

\end{defn}

\begin{thm}[Inverse function theorem]
	Let  $\phi : U \subset \R^N \rightarrow \R^M$ be a differentiable mapping and suppose that the differential $d \phi$ is an isomorphism at $p \in U$.  Then there exists a neighbourhood $V \subset U$ and a neighbourhood $W$ of $F(p)$ such that $\phi$ is a diffeomorphism. 
\end{thm}

\begin{thm}[Implicit function theorem]
	%% TODO Find this

\end{thm}

\begin{defn}
	A map $\phi$ is \emph{regular} at a point $p$ if the matrix of $d\phi$ has full rank at $p$.  A critical point has that the matrix of $d\phi$ does not have a full rank at $p$.
\end{defn}

\begin{defn}[Implicitly defined surfaces]
	Let $f: \R^3 \rightarrow \R$ be a smooth function.  Let $p$ be a point in $\R^3$ with $f(p) =0$ and $f'(p) \neq 0$.  Then there exists a neighbourhood of the point $a$ such that there exists a smooth parameterisation $\phi : U \rightarrow \R^3$ of the set of solutions to $f(x) = 0$.
\end{defn}

\begin{rem}
	This allows us to define surfaces implicitly as the set of solutions to some function.  E.g. the sphere in $\R^3$, the sphere $S^2$ is defined as the set of solutions to \[
		f(x,y,z) = x^2 +y^2 + z^2 - 1 = 0
	\]
\end{rem}

\section{Surfaces in $\R^3$}

\begin{defn}
	A \emph{regular surface} is defined intuitively that for every point $p \in S$, there exists a diffeomorphism between some open set $U \in \R^2$ and a neighbourhood $V$ of that point $p \in S$
\end{defn}

\begin{defn}
	Let $\phi: U \rightarrow R^3$ be a parametrization of a surface S.  The tangent space $T_p S$ to the surface at a point $p \in U$ is the image of the mapping \[
		d\phi_p : \R^2 \rightarrow \R^3
	\]
\end{defn}
\begin{thm}[Surface as a graph]
	Let $S \subset \R^3$ be a regular surface and $p \in S$.  Then there exists a neighbourhood $V$ of $p$ in $S$ such that $V$ is the graph of a differentiable function which has one of the following forms: $z = f(x,y), y = g(z,x), x = h(z,y)$.
\end{thm}

\begin{defn}[Surface of revolution]
	Let $C(v) = (\alpha(v),0,\beta(v))$ be a curve in the $xz$ plane.  Then we can parameterise the surface formed by revolving the curve about the $z$ axis by \[
		\phi(u,v) = (\alpha(v)\cos u, \alpha(v) \sin u,\beta v )
	\]
\end{defn}
\begin{defn}[First fundamental form]
	Let $S$ be a surface parameterised by $\phi : (u^1,u^2) \subset \R^2 \rightarrow \R^3$.  Write $\mathbf{E_i} = \frac{\partial \phi}{\partial u^i}$.  Then we can define the matrix $g$ of the second fundamental form $I$  as $g_{ij} = \langle E_i,E_j \rangle$
\end{defn}

\begin{defn}
	The arc length of a parameterised curve $\alpha(t) = \phi(u^1(t),u^2(t))$ is given by \[
		s(t) = \int_0^t |\alpha'(t)| dt = \int_0^t \sqrt{\sum g_{ij} u'^i u'^j }
			\]
\end{defn}

\begin{defn}[Area of a regular surface]
	Let $R \subset S$ be a bounded region of a regular surface, contained in the coordinate neighbourhood of the parameterization $\phi: U \subset \R^2 \rightarrow \R^3$.  The positive number \[
		\iint_Q |\mathbf{E_1} \times \mathbf{E_2} | du^1 du^2 = A(R), \qquad Q = \phi^{-1}(R) 
	\]
\end{defn}

\begin{defn}[Orientation]
	A regular surface is \emph{orientable} if it is possible to cover it with a family of coordinate neighbourhoods in such a way that if a point $p \in S$ belongs to two neighbourhoods of this family, then the change of coordinate matrix has a positive determinant at $p$.  If such a choice is not possible, then the surface is called \emph{non-orientable}.
\end{defn}

\begin{defn}[Normal vector]
	We define the \emph{normal vector} \textbf{N} to a regular surface $S$ as \[
		\mathbf{N}(p) = \frac{E_1 \times E_2}{|E_1 \times E_2|}(p)
	\]
\end{defn}

\begin{defn}[Gauss Map]
	Let $S \subset \R^3$ be a regular surface with an orientation $\mathbf{N}$.  The map $\mathbf{N}: S \rightarrow \R^3$ takes its values in the unit sphere $S^2$, and this map $\mathbf{N}: S \rightarrow S^2$ is called the \emph{Gauss map}.  
\end{defn}

\begin{prop}
	The differential $dN_p : T_p(S) \rightarrow T_p(S)$ is a self adjoint linear map - \emph{i.e.} \[
		\langle dN_p(w),v \rangle = \langle w,dN_p(v) \rangle
	\]
\end{prop}

\begin{defn}
	The matrix of the second fundamental form is given by \[ II_{ij} = h_{ij} = -\langle \mathbf{N_{u_i}},\mathbf{E_j} \rangle = -\langle  \mathbf{dN(E_i)},\mathbf{E_j} \rangle = \langle \mathbf{N}, \mathbf{E_{ij}} \rangle \]

We have \[
	II_p(X) = - \langle dN_p(X),X \rangle
\]
\end{defn}

\begin{defn}[Normal curvature]
	The \emph{normal curvature} of a space curve $k_n$ is defined as $k_n = k cos \theta$, where $\theta$ is the angle between \textbf{N} and \textbf{n}.
\end{defn}

\begin{defn}
	The \emph{principle curvatures} are the eigenvalues of $-dN_p$.     
\end{defn}

\begin{thm}
	If the normal curvatures $	II_p(X) $ with $|X| = 1$ are not all equal, then there is an orthonormal basis $\mathbf{e_1}, \mathbf{e_2}$ of $T_p S$ such that \[
			II_p(\cos \theta \mathbf{e_1} + \sin \theta \mathbf{e_2} = k_1 \cos^2 \theta + k_2 \sin^2 \theta
	\]
	where $k_1,k_2$ are the principle curvatures as defined above.
\end{thm}

\begin{defn}
	A point $p \in S$ where $k_1 = k_2$ is called an umbilical point
\end{defn}

\begin{defn}
	A curve $\alpha(t)$ is \emph{asymptotic} if its normal curvature is zero for all $t$.
\end{defn}

\begin{defn}
	A curve is a \emph{line of curvature} if $\alpha'(t)$ is an eigenvector of $-dN_{\alpha(t)}$ for all $t$
\end{defn}
\begin{defn}
	The \emph{Gauss curvature} of an oriented surface at a point $p$ is \[
		K(p) = \det(-dN_p) = k_1(p) k_2(p)
	\]

	The \emph{mean curvature} of $S$ at $p$ is \[
		H(p) = \frac{1}{2} \text{trace} (-dN_p)
	\]
\end{defn}

\begin{thm}
	\[
	-(dN_p)^T = (II)(I)^{-1}	
	\]
	\[
		K = \frac{h_{11}h_{22}-{h^2_{12}}}{g_{11}g_{22}-g^2_{12}}
	\]
	\[
		H = \frac{h_{11}g_{22}-2h_{12}g_{12}+h_{22}g_{11}}{2 ({g_{11}g_{22}-g^2_{12})}}
	\]
\end{thm}

\begin{defn}[Minimal surface]
	A surface is a \emph{minimal surface} if its mean curvature is identically zero. \[
		H \equiv 0
	\]
\end{defn}

\begin{defn}
	A coordinate chart $\phi: U \subset \R^2 \rightarrow \Sigma \subset \R^3$ is called \emph{isothermal} if $g_{11} = g_{22}, g_{12} = g_{22} = 0$.  This is equivalent to saying $(I)$ is a multiple of the identity.
\end{defn}

\begin{thm}
	If $\phi$ is isothermal, then $\phi(U) = \Sigma$ is a minimal surface if and only if $\nabla^2 \phi = \mathbf{0}$
\end{thm}

\begin{defn}[Isometry]
	A diffeomorphism \[
		\psi: \S_1 \rightarrow S_2 
	\]
	is an isometry if it preserves the inner product, i.e. for all $p \in S_1$ and $X,Y \in T_P S_1$ \[
		\langle X,Y \rangle_p = \langle d \psi_p(X),d \psi_p(Y) \rangle_{\psi(p)}
	\]

	This is equivalent to \[
		I_p(X) = I_{\psi(p)} (d\psi_p(X)
	\]
	If for each $p \in S_1$, there is an open neighbourhood $U_1$ of p and a map $\psi : U_1 \rightarrow S_2$, then we say that $S_1$ is \emph{locally isometric} to $S_2$.
\end{defn}

\begin{thm}
	If two surfaces $\psi: U \rightarrow S_1$ and $\varphi: U \rightarrow S_2 $  have the same coefficients for $g_{ij}$, then $S_1$ and $S_2$ are locally isometric.
\end{thm}
\begin{defn}[Christoffel symbols]
	Define the \emph{Christoffel symbols} $\christ{i}{j}{k}$ by 
	\begin{align*}
		\phi_{u_1 u_1} & = \christ{1}{1}{1} \phi_{u_1} + \christ{2}{1}{1} \phi_{u_2} + L_1 \mathbf{N} \\
		\phi_{u_1 u_2} & = \christ{1}{1}{2} \phi_{u_1} + \christ{2}{1}{2} \phi_{u_2} + L_2 \mathbf{N} \\
		\phi_{u_2 u_1} & = \christ{1}{2}{1} \phi_{u_1} + \christ{2}{2}{1} \phi_{u_2} + \hat{L}_2 \mathbf{N}\\
		\phi_{u_2 u_2} & = \christ{1}{2}{2} \phi_{u_1} + \christ{2}{2}{2} \phi_{u_2} + L_3 \mathbf{N}\\
	\end{align*}

	Taking inner products with $\mathbf{E_i}$ gives

	\begin{align*}
		\christ{1}{1}{1}g_{11} + \christ{2}{1}{1}g_{12} &= \frac{1}{2}(g_{11})_{u_1} \\
			\christ{1}{1}{1}g_{12} + \christ{2}{1}{1}g_{22} &= (g_{12})_{u_1} - \frac{1}{2}(g_{11})_{u_2}\\
				\christ{1}{1}{2}g_{11} + \christ{2}{1}{2}g_{12} &= \frac{1}{2}(g_{11})_{u_2}\\
					\christ{1}{1}{2}g_{12} + \christ{2}{1}{2}g_{22} &= \frac{1}{2}(g_{22})_{u_1} \\
						\christ{1}{2}{2}g_{11} + \christ{2}{2}{2}g_{12} &= (g_{12})_{u_2} - \frac{1}{2}(g_{22})_{u_1}\\
							\christ{1}{2}{2}g_{12} + \christ{2}{2}{2}g_{22} &= \frac{1}{2}(g_{22})_{u_2}
	\end{align*}
\end{defn}

\begin{thm}[Gauss' Theorem]
\[	g_{11}K = (\christ{2}{1}{1})_{u_2} - (\christ{2}{1}{2})_{u_1} + \christ{2}{1}{1} \christ{2}{2}{2} + \christ{1}{1}{1}\christ{2}{1}{2} - \christ{1}{1}{2}\christ{2}{1}{1} - \christ{2}{1}{2}\christ{2}{1}{2}     \]
\end{thm}

\begin{thm}[Mainardi-Codazzi equations]
	\begin{align*}
		(h_{11})_{u_2} - (h_{12})_{u_1} = h_{11} \christ{1}{1}{2} + h_{12}(\christ{2}{1}{2}-\christ{1}{1}{1}) - h_{22}\christ{2}{1}{1} \\
			(h_{12})_{u_2} - (h_{22})_{u_1} = h_{11} \christ{1}{2}{2} + h_{12}(\christ{2}{2}{2}-\christ{1}{1}{2}) - h_{22}\christ{2}{1}{2} 
	\end{align*}
\end{thm}

\begin{thm}[Christoffel symbols in terms of the metric]
	\[
		\christ{l}{i}{k} = \frac{1}{2}g^{jl}\left( g_{ij,k} + g_{jk,i} - g_{ki,j} \right)
	\]
\end{thm}

\begin{defn}
	Let $w$ be a smooth vector field in a neighbourhood $U$ of $p \in S$.  Let $\alpha \subset S$ be a smooth curve with $\alpha(0) = p, \alpha'(0) = v$.  Define \[
		\frac{Dw}{dt}(0) = D_v w(p)
	\]
	to be the projection of $\frac{dw}{dt} = dw_v(p)$ to $T_p S$, and call it the covariant derivative at $p$ of the vector field $w$ in the direction of $v$.
\end{defn}

\begin{defn}
	Let the vector field $w = a^i \mathbf{E_i}$ and $v = b^j \mathbf{E_j}$.  Then we have \[
		\nabla w (v) = \nabla_v(w) = \frac{Dw}{dt} = \frac{da^i}{dt}\mathbf{E_i} + a^i b^j \christ{k}{i}{j} \mathbf{E_k}
	\]
\end{defn}

\begin{defn}[Parallel transport]
	A vector field $w$ along a curve $\alpha \subset S$ is called \emph{parallel} if \[
		\nabla_{\alpha'(t)}w = 0
	\]
	for all $t$.
\end{defn}

\begin{defn}
	A nonconstant parameterised curve $\alpha$ is a parameterised geodesic if \[
	\nabla_{\alpha'(t)} \alpha'(t) = 0	
	\]
	i.e. if its tangent vector field is parallel along the curve.
\end{defn}

\begin{defn}
	Let $w$ be a smooth field of unit vectors along a curve $\alpha$ where $\alpha$ lies in an oriented surface.  Then define the \emph{algebraic value of the covariant derivative} $[	\nabla_{\alpha'(t)}w]$ by \[
			\nabla_{\alpha'(t)}w = [	\nabla_{\alpha'(t)}w] \mathbf{N} \times w(t)
	\]
\end{defn}

\begin{defn}
		Let $\alpha$ be a curve parameterised by arc length.  The algebraic value of  $	\nabla_{\alpha'(t)} \alpha'(t)$ is called the \emph{geodesic curvature of $\alpha$} and is denoted by $k_g$.
\end{defn}

\begin{rem}
	Up to sign, $	\nabla_{\alpha'(t)} \alpha'(t)$ is the tangential projection of $\alpha''(s)$, and so the geodesic curvature is up to sign the tangential component of $a''(s)$.
\end{rem}

\begin{thm}
	For a curve $\alpha$ in an oriented surface $S$, we have \[
		k^2 = k_n^2 + k_g^2
	\]
\end{thm}

\begin{rem}
	Thus, $\alpha$ is a geodesic if and only if its geodesic curvature is identically zero.
\end{rem}

\begin{thm}
	Let $\gamma(t) = \phi(u^1(t), u^2(t))$.  Then $\gamma'(t) = (u^i)'\mathbf{E_i}$.  Then we have $\gamma$ is a geodesic if and only if 
	\begin{align*}
		0 &= 	\nabla_{\gamma'(t)} \gamma'(t) \\
		&= \nabla_{\gamma'(t)} (u^j)' \mathbf{E_j} \\
		&= (u^j)'' \mathbf{E_j} + (u^j)' \nabla_{(u^i)' \mathbf{E_i}} \mathbf{E_j} \\
		&=  [(u^k)'' + (u^j)' (u^i)' \christ{k}{i}{j}] \mathbf{E_k}
	\end{align*}
	if and only if \[
		(u^k)'' + (u^j)' (u^i)' \christ{k}{i}{j} = 0
	\] for $k = 1,2$
\end{thm}

\begin{thm}
A point and a tangent vector to that point uniquely determine a geodesic in a sufficiently small neighbourhood of that point.
\end{thm}

\begin{thm}
	For each $p \in \Sigma$, there is some neighbourhood $U$ of $p$ such that for any $q \in U$, the geodesic from $p$ to $q$ has length less than or equal to the length of any other path from $p$ to $q$.
\end{thm}

\begin{thm}
	Let $v$ and $w$ be smooth \emph{unit} vector fields along a curve $\alpha \subset S$.  Then \[[	\nabla_{\alpha'(t)} w] - [	\nabla_{\alpha'(t)} v] = \frac{d \theta}{dt} \]
	where $\theta$ is a smooth choice of angle from $v$ to $w$.
\end{thm}

\subsection{The Gauss Bonnet Theorem}
We first prove a local version, and then the full global theorem.

The depth of this (global) theorem lies in its topological content rather than in its geometric content, and its most striking feature is that it relates the two.

In fact, a corollary is

\begin{cor}
	For an orientable compact surface $S$, \[ \iint_S K d \sigma = 2 \pi \chi(S) \]
\end{cor}

\begin{thm}[Theorem of Turning Tangents]
	Let $S$ be an oriented surface, and $\phi : U \rightarrow S$ a coordinate chart.  Let $\alpha$ be a simple, closed, parameterised curve, regular in each interval $[ t_i, t_{i+1}]$.

	Choose a smooth angle function $\theta_i$ on each $[t_i, t_{i+1}]$ which measures the angle from $\mathbf{E_1}$ to $\alpha'$.  Denote by $\gamma_i$ the external angle at $t_i$.  Then \[
		\sum_{i=0}^k (\theta_i(t_{i+1})- \theta_i(t_i)) + \sum_{i = 0}^k \gamma_i = \pm 2 \pi
	\]
	and the above sum is $2 \pi$ if $\alpha $ is a positively oriented curve.
\end{thm}

\begin{thm}[Local Gauss-Bonnet Theorem]
	Let $S$ be an oriented surface, $U \subset \R^2$ be homeomorphic to an open disk and $\phi: U \rightarrow S$ be a coordinate chart compatible with the orientation, such that $g_{12} = 0$.

	Let $R \subset \phi(U)$ be a simple region and $\alpha$ a positively oriented piecewise regular curve whose trace equals $\partial R$. Let $\alpha$ be parameterised by arc length, with vertices at $s_0, s_1, \dots, s_k$.

	Then \[
		\sum_{i=0}^k \int_{s_i}^{s_{i+1}} k_g(s) ds + \iint_R K d \sigma + \sum_{i = 0}^k \gamma_i = 2 \pi
	\]
\end{thm}

\begin{cor}
	Let $R \subset \phi(U)$ be a simple region with piecewise regular boundary, and $\alpha: [0,l] \rightarrow S$ be an arc length parametrisation of $\partial R$.   
	Take a unit vector $w_0 \in T_{\alpha(0)}S$ and let $w$ be the vector field along $\alpha$ generated by the parallel transport of $w_0$.  Then we have 
	\[ \Delta \theta = \theta(l) - \theta(0) = \iint_R K d \sigma \]
	where $\theta$ is a continuous choice of angle from $\mathbf{E_1}$ to $w$.
\end{cor}

\begin{defn}
	The \emph{Euler characteristic} $ \chi(S)$ of a surface $S$ is given by $\chi(S) = V - E + F$, where $V,E,F$ are the number of vertices, edges and faces of any triangulation of $S$.
\end{defn}

\begin{thm}
	If two surfaces $S_1$ and $S_2$ are homeomorphic, then $\chi(S_1) = \chi(S_2)$.
\end{thm}

\begin{thm}
	\[\chi(S) = 2 - 2g \]
	where $g$ is the number of holes of $S$, called the \emph{genus} of $S$.
\end{thm}

\begin{thm}[Global Gauss-Bonnet]
	Let $R \subset S$ be a regular region of an oriented surface with positively oriented boundary curves $C_i$. Write $\gamma_j$ for the external angles of the curves $C_i$.  THen \[
		\sum_{i} \int_{C_i} k_g(s) ds + \iint_R K d \sigma + \sum_j \gamma_j = 2 \pi \chi(S)
	\]
\end{thm}

\begin{defn}
	A point $p$ is said to be a \emph{singular point} of the smooth vector field $v$ on $S$ if \[
		v(p) = 0
	\]

	Take a positively oriented simple closed curve $\alpha$ enclosing an isolated singular point of a vector field $v$.  Let $\theta_v$ be a smooth choice of angle from $\mathbf{E_1}$ to $v$.  THen \[
		\theta_v(l) - \theta_v(0) = 2 \pi I
	\]
	for some \textbf{integer} $I$.  We call $I$ the \emph{index} of $v$ at $p$.
\end{defn}

\begin{thm}[Poincar\`{e}-Hopf Theorem]
	Let $v$ be a smooth vector field on $S$ with isolated singular points $p_1,\dots, p_n$.  Then \[
		\sum_{i=1}^n I_i = \chi(S)
	\]
\end{thm}

\begin{cor}[Hairy-Ball theorem]
	If $S$ is not homeomorphic to a torus, then any smooth vector field on $S$ must vanish somewhere.
\end{cor}

\begin{thm}[Morse's Theorem]
	Let $f: S \rightarrow \R$ be a smooth function on a compact oriented surface $S$ such that all critical points ($df_p = 0$) are non-degenerate ($\det A(p) \neq 0$, where $A$ is the \emph{Hessian} matrix of $f$).

Let
\begin{itemize}

	\item 	$M$ = number of local maxima,

	\item 	$m$ = number of local minima,

	\item 	$s$ = number of saddle points.

\end{itemize}

	Then \[
		M - s + m = \chi(S)
	\]
	This is independent of the function $f$, and depends only upon the topology of $S$.
\end{thm}

\section{Abstract Manifolds}

\begin{defn}[Abstract manifolds]

\end{defn}
An abstract surface is a set $\Sigma$ together with a family of maps \[
	\phi_\alpha: U_\alpha \rightarrow \Sigma
\] of open sets $U_\alpha \subset \R^2$ into $\Sigma$ so that

\begin{itemize}

	\item $\cup_\alpha \phi_\alpha(U_\alpha) = \Sigma$

	\item Whenever $\alpha, \beta$ are such that $W_{\alpha \beta} = \phi_\alpha (U_\alpha) \cap \phi_\beta (U_\beta) \neq \emptyset$, then $\phi^{-1}_\alpha (W_{\alpha \beta}), \phi^{-1}_\beta (W)_{\alpha \beta} \subset \R^2$ are open and
  \[
 	\phi^{-1}_\beta \circ \phi_\alpha : \phi^{-1}_\alpha (W_{\alpha \beta}) \rightarrow \phi^{-1}_\beta (W_{\alpha \beta})
 \]

\end{itemize}

\begin{defn}
	A function $f: M \rightarrow \R$ is \emph{smooth} at $p \in M$ if for all coordinate charts $\phi: U \rightarrow M$ about $p$, \[
		f \circ \phi : U \subset \R^N \rightarrow R
	\] is smooth

	A map $f : M_1 \rightarrow M_2$ between smooth manifolds is \emph{smooth} at $p \in M_1$ if for all coordinate charts $\phi : U \rightarrow M_1$ about $p$, and coordinate chart $\psi : V \rightarrow M_2$\[
		\psi^{-1} \circ f \circ \phi : U \subset \R^{N_1} \rightarrow \R^{N_2}
	\]
	It is called smooth if it is smooth at every point.

\end{defn}

\begin{defn}
	Two manifolds $M_1$ and $M_2$ are \emph{diffeomorphic} if there is a smooth map $f: M_1 \rightarrow M_2$ that has a smooth inverse.

	A necessary condition is that the manifolds have the same dimension.
\end{defn}

\begin{defn}
	A smooth parameterised curve in a (smooth) manifold $M$ is a smooth map $\alpha : (a,b) \rightarrow M$
\end{defn}

\begin{defn}[Tangent vectors as equivalence classes of curves]
	A \emph{tangent vector} $v$ to the smooth manifold $M$ at $p \in M$ is an equivalence class $[\alpha]$ of curve $\alpha: (-\epsilon, \epsilon) \rightarrow M$ with $\alpha(0) = p$.

	Fix a coordinate chart $\phi: U \rightarrow M$ around $p$.  If $\alpha_1, \alpha_2$ satisfy $\alpha_1(0) = \alpha_2(0) = p$, then we define \[
		\alpha_1 \sim \alpha_2 \Leftrightarrow (\phi^{-1} \circ \alpha_1)'(0) = (\phi^{-1} \circ \alpha_2)'(0)
	\]

	This is independent of the choice of $\phi$
\end{defn}

\begin{defn}[Derivation]
	A derivation is a map from the real vector space of smooth function on $M$ to $\R$, which is linear and satisfies the product rule at $p$, \emph{i.e.} \[
		D(fg) = fD(g) + gD(f)
 	\]
\end{defn}

\begin{thm}
	THe set of all derivations at $p \in M^N$ forms an $n$-dimensional vector space, with basis \[
		\frac{\partial}{\partial u^i} : \phi \mapsto \frac{\partial \phi(p)}{\partial u^i}
	\]
	where \[
		\phi :(u^1,\dots, u^n) \mapsto \phi(u^1,\dots,u^n)
	\]
\end{thm}

\begin{defn}[Tangent vector as an operator that acts on functions]
	There is a one-to-one correspondence between equivalence classes of curves through $p$ and derivations at $p$.  We can thus alternatively define a tangent vector at $p$ to be a derivation at $p$.  
\end{defn}

\begin{defn}
	A smooth manifold $M$ is orientable if it has an atlas (i.e. may be covered with coordinate neighbourhoods) so that whenever $p ∈\in M$ lies in the image of two local parameterisations $\phi$ and $\psi$, the change of basis matrix at $p$ has positive determinant.

	If this is possible, then the choice of such an atlas is called an orientation of $M$, and we say that $M$ is oriented.

	An orientation of a smooth manifold M can be thought of as an orientation of every tangent plane that “varies smoothly”.
\end{defn}

\begin{defn}[Tangent Bundle]
Define \[
	TM = \{ (p,v): p ]in M, v \in T_p M \}
\]
\end{defn}

\begin{defn}
	A \emph{smooth vector field} on the smooth manifold $M$ is a smooth map $v : M \rightarrow TM$ such that $v_p \in T_pM$ for each $p \in M$.
\end{defn}

\begin{defn}
	Let $v,w$ be smooth vector fields on $M$.  A \emph{covariant derivative} on a smooth manifold $M$ is a map \[
		\nabla : (v,w) \mapsto \nabla_v w
 	\] 
such that 
\begin{itemize}

	\item $\nabla_v$ is linear in $v$ \emph{i.e.} $\nabla_{fu + gv}(w) = f \nabla_u(w) + g \nabla_v(w)$

	\item Each $\nabla_v$ satisfies linearity over $\R$: $\nabla_v(u +w) = \nabla_v(u) + \nabla_v(w)$

	\item Each $\nabla_v$ satisfies the product rule \[
 	\nabla_v (fu) = f \nabla_v(u) + v(f) u
 \]

\end{itemize}
\end{defn}

\begin{defn}[Levi-Civita connection]
	On a Riemannian manifold, there is a unique covariant derivative satisfying the following two conditions:
	\[
		\nabla_{\frac{\partial}{\partial u^i}} \frac{\partial}{\partial u^j} = \nabla_{\frac{\partial}{\partial u^j}} \frac{\partial}{\partial u^i} 
	\]
	and viewing $\frac{\partial}{\partial u^i}$ as a derivation,\[
		\deriv{i} (\langle \deriv{j},\deriv{k}\rangle) = \langle \nabla_{\deriv{i}} \deriv{j},\deriv{k}\rangle + \langle \deriv{j},\nabla_{\deriv{i}} \deriv{k} \rangle
	\]

\end{defn}

\begin{defn}[Christoffel symbols on an abstract manifold]
	Define the Christoffel symbols \christ{k}{i}{j} with respect to a local coordinate chart by \[
		\nabla_\deriv{i} \deriv{j} = \christ{k}{i}{j} \deriv{k}
	\]
	We have \[
		\christ{k}{i}{j} = \frac{1}{2}g^{kl}\left( g_{il,j} + g_{jl,i} - g_{ij,l} \right)
		\] which defines a unique covariant derivative.
\end{defn}

\begin{defn}[Geodesic]
	A geodesic is a curve $\alpha$ satisfying 
	\[ \nabla_\alpha' \alpha' = 0 \]

	This gives us the equations \[
		(u^k)'' + \christ{k}{i}{j}(u^i)'(u^j)' = 0
	\] for $k = 1,\dots,n$
\end{defn}

\begin{defn}
	The \emph{hyperbolic plane} is defined to be the upper half plane with the metric (writing $u^1 = x, u^2 = y$) \[
		g_{11} = \frac{1}{y^2}, g_{12} = 0, g_{22} = \frac{1}{y^2}
 	\]
\end{defn}

\end{document}

### 2011-12-12 MATH 3969 - Measure Theory and Fourier Analysis/MATH 3969 Lecture Notes.tex
Initial Commit

%
%  untitled
%
%  Created by Andrew Tulloch on 2009-11-04.
%  Copyright (c) 2009 __MyCompanyName__. All rights reserved.
%!TEX TS-program = xelatex
%!TEX encoding = UTF-8 Unicode

\documentclass[10pt, oneside, reqno]{amsart}
\usepackage{geometry, setspace, graphicx, enumerate}
\onehalfspacing                 
\usepackage{fontspec,xltxtra,xunicode}
\defaultfontfeatures{Mapping=tex-text}

%\setromanfont[Mapping=tex-text,Contextuals= 
%{NoWordInitial,NoWordFinal,NoLineInitial,NoLineFinal}]{Hoefler Text}
%\setsansfont[Scale=MatchLowercase,Mapping=tex-text]{Hoefler Text}
%\setmonofont[Scale=MatchLowercase]{Andale Mono}

	% AMS Theorems
	\theoremstyle{plain}% default 
	\newtheorem{thm}{Theorem}[section] 
	\newtheorem{lem}[thm]{Lemma} 
	\newtheorem{prop}[thm]{Proposition} 
	\newtheorem*{cor}{Corollary}

	\theoremstyle{definition} 
		\newtheorem{defn}[thm]{Definition}
		\newtheorem{conj}[thm]{Conjecture}
		\newtheorem{exmp}[thm]{Example}

	\theoremstyle{remark} 
		\newtheorem*{rem}{Remark} 
		\newtheorem*{note}{Note} 
		\newtheorem{case}{Case}

		\newcommand{\expc}[1]{\mathbb{E}\left[#1\right]}
		\newcommand{\var}[1]{\text{Var}\left(#1\right)}
		\newcommand{\cov}[1]{\text{Cov}\left(#1\right)}
		\newcommand{\prob}[1]{\mathbb{P}(#1)}
		\newcommand{\given}{ \, | \,}
		\newcommand{\us}{0 \leq u \leq s}
		\newcommand{\ts}[1]{\{ #1 \}}

\newcommand{\al}{\alpha}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\F}{\mathbb{F}}
\newcommand{\Ga}{\mathbb{G}}

\newcommand{\dmu}{\, d \mu}

\newcommand{\aut}[1]{\text{Aut}{(#1)}}

\newcommand{\gener}[1]{\langle #1 \rangle}
\newcommand{\charr}[1]{\text{char}(#1)}
\newcommand{\nth}{n\textsuperscript{th}}

\newcommand{\tworow}[2]{\genfrac{}{}{0pt}{}{#1}{#2}}
\newcommand{\xdeg}[2]{[#1 : #2]}
\newcommand{\gal}[2]{\text{Gal}(#1/#2)}
\newcommand{\minpoly}[2]{m_{#1, #2}(x)}

\renewcommand{\phi}{\varphi}

\newcommand{\mapping}[5]{\begin{align*}
	#1 : \quad     #2 &\rightarrow #3 \\
			#4  &\mapsto #5
\end{align*}	
}

\title{MATH 3969 - Measure Theory and Fourier Analysis \\ Exam Notes}								% Document Title
\author{Andrew Tulloch}
%\date{}                                           % Activate to display a given date or no date

\begin{document}
\maketitle

\section{Measure Theory}
\newcommand{\sig}{$\sigma$-algebra }
\newcommand{\siga}{\mathcal{A}}

\begin{defn}[$\sigma$-algebra]
	Let $X$ be a set.  A collection $\mathcal{A}$ of subsets of $X$ is called a \sig if 
	\begin{itemize}
		\item $\emptyset \in \siga$ 
		\item If $A \subseteq X$ is in $\siga$, then its complement $A^c = X \backslash A$ is in $\siga$
		\item Whenever $A_0,A_1,\dots$ are subsets of $X$ in $\siga$, then their union\[
			\bigcup_{k=0}^\infty A_k
		\] also belongs to $\siga$.
	\end{itemize}
\end{defn}

\begin{defn}[Measure]
	Let $X$ be a set, and let $\siga$ be a \sig of subsets of $X$.   Suppose that $\mu: \siga \rightarrow [0,\infty]$ is a function.  Then $\mu$ is a measure if 
	\begin{itemize}
		\item $\mu(\emptyset) = 0$
		\item Whenever $A_0, A_1,\dots$ are \emph{pairwise disjoint} subsets of $X$ in $\siga$, then \[
			\mu(\bigcup_{k=0}^\infty A_k) = \sum_{k=0}^\infty \mu(A_k)
		\]
	\end{itemize}
\end{defn}

\begin{prop}[Properties of a \sig] Let $\siga$ be a \sig of subsets of a set $X$.  Then
	\begin{itemize}
		\item $X, \emptyset \in \siga$
		\item If $A_k \in \siga$, then $\bigcap_{k=0}^\infty A_k \in \siga$
		\item If $A,B \in \siga$, then $A \cup B, A \cap B \in \siga$
	\end{itemize}

\end{prop}

\begin{defn}[Algebra]
	A collection $\siga$ of subsets $A$ of $X$ which satisfies the first two conditions of a \sig and also 
	\begin{itemize}
		\item If $A, B \in \siga$, then $A \cup B \in \siga$
	\end{itemize}
	is called an algebra.  Every \sig is an algebra, but not every algebra is a \sig
\end{defn}

\begin{defn}[\sig generated by $\mathcal{S}$]
	Let $\mathcal{S}$ be a collection of subsets of $X$.  Let \[
		\siga(\mathcal{S}) = \bigcap \{ \siga : \siga \text{ is a \sig}, \text{ and } \mathcal{S} \subseteq \siga \}
	\]
$	\siga(\mathcal{S})$ is called the \emph{\sig generated by $\mathcal{S}$}
\end{defn}

\begin{defn}[Borel \sig]
	Let $X$ be a metric space and $\mathcal{S}$ the collection of all open sets in $X$.  We call $\mathcal{B} = \siga(\mathcal{S})$ the \emph{Borel \sig}.  Sets in $\mathcal{B}$ are called \emph{Borel sets}.
\end{defn}
\begin{cor}
	We have the following examples of Borel sets.
	\begin{itemize}
		\item Any open set is a Borel set.
		\item If $B$ is a Borel set, then so is $B^c$.
		If $B_0, B_1,\dots$ is a sequence of Borel sets, then so are $\bigcup_{k=0}^\infty B_k$ and $\bigcap_{k=0}^\infty B_k$.
	\end{itemize}
\end{cor}
\subsection{Properties of Measures} % (fold)
\label{sub:properties_of_measures}

\begin{prop}[The Monotonicity Property]
	If $A$ and $B$ are $\mu$-measurable subsets of $X$ with $A \subseteq B$, then $\mu(A)\leq \mu(B)$.
\end{prop}

\begin{prop}[The Countable Subadditivity Property]
	If $A_0,A_1,\dots$ are $\mu$-measurable subsets of $X$, then \[
		\mu(\bigcup_{k=0}^\infty A_k) \leq \sum_{k=0}^\infty \mu(A_k)
	\]
\end{prop}

\begin{prop}[Monotone Convergence Property of Measures]
	Let $A_0 \subseteq A_1 \subseteq A_2$ be an increasing sequence of measurable sets.  Then \[
		\mu(\bigcup_{k=0}^\infty A_k) = \lim_{k \rightarrow \infty} \mu(A_k)
	\]
\end{prop}

\begin{prop}
	Let $A_0 \supseteq A_1 \supseteq A_2 \dots$ be sets from some \sig $\mathcal{A}$.  If $\mu(A_0) < \infty$, then 
	\[
	\mu(\bigcap_{k = 0}^\infty A_k) = \lim_{k \rightarrow \infty} \mu_(A_k)
	\]
\end{prop}

% subsection properties_of_measures (end)

\subsection{Constructing $\sigma$-algebras and measures} % (fold)
\label{sub:constructing_sigma_algebras_and_measures}
\begin{defn}[Lebesgue outer measure]
	If $A \subseteq \R$, let \[
		m^*(A) = \inf \left\{ \sum_{k=0}^\infty (b_k - a_k) \given a_k < b_k, A \subseteq \bigcup_{k=0}^\infty (a_k, b_k) \right\}
	\]
\end{defn}

\begin{prop}[Properties of the Lebesgue outer measure]
	The Lebesgue measure obeys the following properties.
	\begin{itemize}
		\item $m^*(A)$ is defined, and $m^*(A) \in [0,\infty]$ for any subset of $\R$.
		\item $m^*(\emptyset) = 0$
		\item If $A \subseteq B$, $m^*(A) \leq m^*(B)$
		\item For every sequence $A_0, A_1,\dots$, we have \[
			m^*(\bigcup_{k=0}^\infty A_k) \leq \sum_{k=0}^\infty m^*(A_k)
		\]
	\end{itemize}
\end{prop}

\begin{defn}[Outer Measure]
	A function $\mu^* : \mathcal{P} \rightarrow [0,\infty]$ is such that 
	\begin{itemize}
		\item $\mu^*(\emptyset) = 0$
		\item If $A \subseteq B$, $\mu^*(A) \leq \mu^*(B)$
		\item For every sequence $A_0, A_1,\dots$, we have \[
			\mu^*(\bigcup_{k=0}^\infty A_k) \leq \sum_{k=0}^\infty \mu^*(A_k)
		\]
	\end{itemize} 
	Then $\mu^*$ is called an \emph{outer measure} on $X$.
\end{defn}

\begin{thm}[Construction from outer measures]
	Let $\mu^*$ be an outer measure on a set $X$.  Then \[
		\siga = \{ A \subseteq X \given \mu*(S) = \mu^*(S \cap A) + \mu^*(S \cap A^c) \text{ for all } S \subseteq X \}
	\]
	Then $\siga$ is a $\sigma$-algebra.  Let $\mu(A) = \mu^*(A)$ when $A \in \siga$.  Then $\mu : \siga \rightarrow [0,\infty]$ is a measure.
\end{thm}

\begin{prop}
	Let $\mu^*$ be an outer measure, and let $\siga$ be the \sig defined in the last theorem.  Let $A \subseteq X$ satisfy $\mu^*(A) = 0$.  Then $A \in \siga$, and so $\mu(A)$ is defined, and equals 0. 	
\end{prop}

\begin{defn}[Null set]
	If $\mu: \siga \rightarrow [0,\infty]$ is any measure, then a set $A \in \siga$ satisfying $\mu(A) = 0$ is called a null set.
\end{defn}
% subsection constructing_sigma_algebras_and_measures (end)

\subsection{Properties of the Lebesgue measure on $\R$} % (fold)
\label{sub:properties_of_the_lebesgue_measure_on_r_}
Let \[
\mathcal{M} = \{ A \subseteq \R \given m*(S) = m^*(S \cap A) + m^*(S \cap A^c) \text{ for all } S \subseteq \R \}
\]

The sets in $\mathcal{M}$ are called hte \emph{Lebesgue measurable subsets} of $\R$.  If $A \in \mathcal{M}$, the we write $m(A) = m^*(A)$.  This real number is called the \emph{Lebesgue measure of $A$}.

We now show that this \sig $\mathcal{M}$ is very large.

\begin{thm}Let $m^*$ denote the Lebesgue outer measure on $\R$, and let $\mathcal{M}$ be the \sig of Lebesgue measurable sets.  Then 
	\begin{itemize}
		\item If $I \subseteq \R$ is an interval.  Then $m^*(I) = l(I)$.  That is, the outer measure is just its length.
		\item If $I \subseteq \R$ is an interval, then $I \in \mathcal{M}$.
	\end{itemize}

\end{thm}

\begin{prop}
	Any open subset of $\R$ is in $\mathcal{M}$.  Any closed subset of $\R$ is in $\mathcal{M}$.  That is, all open or closed sets in $\R$ are Lebesgue measurable.
\end{prop}

\begin{cor}
	Every Borel subset of $\R$ is contained in $\mathcal{M}$.  
\end{cor}
\begin{proof}
	$\mathcal{M}$ is a \sig which contains every open subset of $\R$.  The \sig $\mathcal{B}$ is by definition the smallest such $\sigma$-algebra.  Thus $\mathcal{B} \subseteq \mathcal{M}$.
\end{proof}

%!TEX root = /Users/ajtulloch/Documents/University 2010/MATH 3969 - Measure Theory and Fourier Analysis/Exam Notes/MATH 3969 - Master Notes.tex

\section{Measurable Functions} % (fold)
\label{sec:measurable_functions}
\newcommand{\Rbar}{\overline{\R}}

\begin{defn}[Measurable function]
	Let $\siga$ be \sig of subsets of a set $X$.  A function $f: X \rightarrow \Rbar$, is called \emph{measurable} (or $\siga$-measurable) if for every $\alpha \in \R$, the set \[
		\{ x \in X \given f(x) > \alpha \}
	\] is in $\siga$.
\end{defn}

\begin{defn}[Indicator function].  Let $S \subset X$.  We define the \emph{indicator function} of $S$ to be the function $1_S : X \rightarrow \R$ given by

	\[ 1_S(x) = \begin{cases}
		1 \quad \text{if $x \in S$}\\
		0 \quad \text{if $x \notin S$}
	\end{cases}\]
\end{defn}

\begin{prop}
	Let $S \subset X$.  Then $1_S$ is measurable if and only if $S \in \siga$.  
\end{prop}
\begin{proof}
	Let $\alpha \in \R$.  Then \[
		\{ x \in X \given 1_S(x) > \alpha \} = \begin{cases}
			\emptyset \quad \text{if $\alpha \geq 1$}\\
			S \quad \text{if $0 \leq \alpha < 1$} \\
			X \quad \text{if $\alpha < 0$}
		\end{cases}
	\]

As $\emptyset, X$ are in $\siga$, then $1_S$ is measurable if and only if $S \in \siga$.
\end{proof}

\begin{prop}[Continuous functions are Lebesgue measurable]
	Let $f: [a,b] \rightarrow \R$ be continuous.  Then $f$ is Lebesgue measurable on $[a,b]$.  More generally, if $X \subset \R$ is in $\mathcal{M}$ and $f: X \rightarrow \R$ is continuous, then $f$ is Lebesgue measurable on $X$.
\end{prop}

\subsection{Basic properties of measurable functions} % (fold)
\label{sub:basic_properties_of_measurable_functions}

\begin{lem}
	Let $\siga$ be a \sig of subsets of a set $X$, and let $f : X \rightarrow \Rbar$ be a function.  Then $f$ is measurable if and only if it satisfies one of the following conditions.
	\begin{itemize}
		\item For each $\alpha \in \R$, the set $\{ x \in X \given f(x) > \alpha \}$ in in $\siga$.
		\item For each $\alpha \in \R$, the set $\{ x \in X \given f(x) < \alpha \}$ in in $\siga$.
		\item For each $\alpha \in \R$, the set $\{ x \in X \given f(x) \leq \alpha \}$ in in $\siga$.
		\item For each $\alpha \in \R$, the set $\{ x \in X \given f(x) \geq \alpha \}$ in in $\siga$.
	\end{itemize}
\end{lem}

\begin{prop}Let $\siga$ be a \sig of subsets of a set $X$, and let $f,g : X \rightarrow \Rbar$ be functions.  Then 
	\begin{itemize}
		\item $f + g$ is measurable (provided that $f(x) = \infty$ and $g(x) = - \infty$ or vice versa holds for no $x \in X$).
		\item $cf$ is measurable for any constant $c \in \R$.
		\item $fg$ is measurable.
		\item $f/g$ is measurable (provided that $g(x)$ is nonzero and not infinity for all $x \in X$).
	\end{itemize}

	Similarly, let $f_0,f_1,\dots : X \rightarrow \Rbar$.  Then 
	\begin{itemize}
		\item $\sup\{f_0,f_1,\dots\}$ and $\inf\{f_0,f_1,\dots\}$ are measurable functions.
	\end{itemize}
\end{prop}
\begin{cor}
	Let $f,g$ be measurable.  Then $\max\{f,g\}$ and $\min\{f,g\}$ are measurable functions.
\end{cor}

\begin{prop}
	Let $\siga$ be a \sig of subsets of a set $X$, and let $f_0, f_1,\dots : X \rightarrow \Rbar$ be measurable functions.  Let $f(x) = \lim_{k \rightarrow \infty} f_k(x)$ for each $x \in X$. Then $f$ is a measurable function.
\end{prop}

% subsection basic_properties_of_measurable_functions (end)

\subsection{Simple functions} % (fold)
\label{sub:simple_functions}
\begin{defn}[Simple function]

	Let $\siga$ be a \sig of subsets of a set $X$.  A function $\phi : X \rightarrow \R$ is called \emph{simple} if it is measurable and only takes a finite number of values.
\end{defn}

\begin{prop}
	Let $\siga$ be a \sig of subsets of a set $X$, and let $f: X \rightarrow [0, \infty]$ be a nonnegative measurable function.  Then there is a sequence $(\phi_n)$ of simple functions such that 
	\begin{itemize}
		\item $0 \leq \phi_1(x) \leq \phi_2(x) \leq \dots \leq f(x)$ for all $x \in X$.
		\item $f(x) = \lim_{n \rightarrow \infty} \phi_n(x)$ for all $x \in X$.
	\end{itemize}
\end{prop}

\begin{proof}
	Define the function $\phi_n$ as follows.

	Let \[
		A_{n,k} = \{ x \in X \given \frac{k}{2^n} \leq f(x) < \frac{k+1}{2^n} \}
	\]
	and let $A_{n,2^n} = \{ x \in X \given f(x) \geq n \}$

	Then the function \[
		\phi_n  = \sum_{k=0}^{n2^n} \frac{k}{2^n}1_{A_{n,k}}
	\]
	obeys the required properties.
\end{proof}

% subsection simple_functions (end)

%!TEX root = /Users/ajtulloch/Documents/University 2010/MATH 3969 - Measure Theory and Fourier Analysis/Exam Notes/MATH 3969 - Master Notes.tex

% \newcommand{\dmu}{\, d \mu}

\section{Integration} % (fold)
\label{sec:integration}

\begin{defn}[Integration of simple functions]
	Let $\phi = \sum_{j=1}^m a_j 1_{A_j}$.  Then the \emph{integral $\int_X \phi \dmu$ of $\phi$ over $X$ with respect to $\mu$} is given by \[
		\int_X \phi \dmu = \sum_{j=1}^m a_j \mu (A_j)
	\]
\end{defn}

\begin{prop}
	Let $\phi$ and $\psi$ be nonnegative simple functions on $X$, and let $c \geq 0$ be constant.  Then \begin{itemize}
		\item $
			\int_X \phi  + \psi \dmu = \int_X \phi \dmu + \int_X \psi \dmu
	$
		\item If $ 0 \leq \psi \leq \phi $, then \[
			0 \leq \int_0 \psi \dmu \leq \int_X \psi \dmu
		\]
		\item $\int_X c \phi \dmu = c \int_X \phi \dmu$
	\end{itemize}
\end{prop}

\begin{defn}[Integral over a subset of $X$]
	Let $\phi$ be a nonnegative simple function, and let $S \subset X$ be measurable.  Then \emph{the integral of $\phi$ over $S$ with respect to $\mu$}, denoted $\int_S \phi \dmu$, is given by \[
		\int_S \phi \dmu = \int_X \phi \cdot 1_s \dmu
	\] 
\end{defn}

\subsection{Integration of nonnegative measurable functions} % (fold)
\label{sub:integration_of_nonnegative_measurable_functions}

\begin{defn}Let $f: X \rightarrow [0, \infty]$ be a nonnegative measurable function.  We define the \emph{integral $\int_X f \dmu$ of $f$ over $X$ with respect to $\mu$} by \[
	\int_X f \dmu = \sup \left\{ \int_X \phi \dmu \given \phi \text{ is simple, and } 0 \leq \phi \leq f \text{ on } X \right\}
	\]
\end{defn}

\begin{lem}
	Suppose that $f,g$ are two nonnegative measurable functions, and $0 \leq g \leq f$
 on $X$.  Then \[
 	0 \leq \int_X g \dmu \leq \int_X f \dmu
 \]\end{lem}

The following is an extremely important theorem in measure theory. 
\begin{thm}[Monotone convergence theorem]
	Let $(f_k)$ be a sequence of nonnegative measurable functions on $X$.  Assume that \begin{itemize}
		\item $ 0 \leq f_0(x) \leq f_1(x) \leq \dots$ for each $x \in X$,
		\item $\lim_{k \rightarrow \infty} f_k(x) = f(x)$ for each $x \in X$.
	\end{itemize}
	When these hold, we write $f_k \nearrow f$ pointwise.

	Then $f$ is measurable, and \[
		\int_X f \dmu = \lim_{ k \rightarrow \infty} \int_X f_k \dmu
	\] 
\end{thm}

\begin{cor}
	Let $f,g$ be measurable on $X$.  THen \[
		\int_X f + g \dmu = \int_X f \dmu + \int_X g \dmu
	\]
\end{cor}

\begin{thm}
	Suppose that $f_k$ is a nonnegative measurable function for $k = 0,1,\dots$.  Then \[
		\int_X \left(\sum_{k=0}^\infty f_k \right) \dmu = \sum_{k=0}^\infty \left( \int_X f_k \dmu \right)
	\]
\end{thm}

\begin{thm}
	Suppose that $X$ is a set, $\siga$ is a \sig of subsets of $X$ and $\mu: \siga \rightarrow [0,\infty]$ is a measure.  Let $f$ be nonnegative and measurable on $X$.  Then define $\mu_f : \siga \rightarrow [0,\infty]$ by \[
		\mu_f(A) = \int_A f \dmu = \int_X f \cdot 1_A \dmu
	\]
	Then $\mu_f$ is a measure.
\end{thm}

\begin{prop}
	Suppose that $f$ is nonnegative and measurable on $X,$ and suppose that $\int_X f \dmu < \infty$.  Then the set $\{ x \in X \given f(x) = \infty \}$ has measure 0.
\end{prop}

\begin{prop}
	Suppose that $X$ is a set, $\siga$ is a \sig of subsets of $X$ and $\mu: \siga \rightarrow [0,\infty]$ is a measure. Suppose that there is a set $N \in \siga$ with $\mu(N) = 0$ and suppose that some property $P$ holds for all $x \in X$ outside $N$.  Then we say that the property $P$ holds \emph{almost everywhere} or \emph{for almost all $x \in X$}.
\end{prop}

\begin{prop}
	Suppose that $f_k$ is a nonnegative measurable function on $X$ for $k = 0,1,\dots$.  Suppose that \[
		\sum_{k=0}^\infty \left(\int_X f_k \dmu \right) < \infty
	\]
Then \[
	\sum_{k=0}^\infty < \infty \text{ for almost all $x \in X$}
\]
\end{prop}

\begin{prop}
	Suppose that $f$ is a nonnegative measurable function on $X$.  Then \[
		\int_X f \dmu = 0 \iff f(x) = 0 \text{ almost everywhere.}
	\]
\end{prop}

\begin{thm}[Fatou's Lemma]
	Suppose that $f_k$ is a nonnegative measurable function on $X$, for $k = 0,1,\dots$, and that $\lim_{k \rightarrow \infty} f_k(x) = f(x)$ for each $x \in X$.  Then \[
		\int_X f \dmu \leq \liminf_{k \rightarrow \infty} \left( \int_X f_k \dmu \right)
	\]
	\end{thm}
% subsection integration_of_nonnegative_measurable_functions (end)

\subsection{Integration of real and complex valued functions} % (fold)
\label{sub:integration_of_real_and_complex_valued_functions}
\begin{lem}

	Let $f: X \rightarrow \Rbar$ be a measurable function, and let $f^+$ and $f^-$ be the positive and negative parts of $f$.  Then $f = f^+ - f^-$, and $|f| = f^+ + f^-$.  Moreover, $|f|$ is a measurable function, and
	\[
		f \text{ is integrable if and only if } \int_X |f| \dmu < \infty
	\]
\end{lem}

\begin{defn}[Integral of a complex valued function]
	Let $f = u + iv,$ where $u,v : X \rightarrow \Rbar$.  Then \[
		\int_X f \dmu = \int_X u \dmu + i \int_X v \dmu
	\]
\end{defn}
\begin{lem}
	Let $f : X \rightarrow \bar{\mathbb{C}}$.  Then $f$ is integrable if and only if $\int_X |f| \dmu < \infty$.
\end{lem}

The next theorem is probably the most important single theorem in these notes.  It has many applications, both of a theoretical and practical nature.
\begin{thm}[Dominated convergence theorem]
	Let $(f_k)$ be a sequence of real or complex valued measurable function on $X$. Assume that \begin{itemize}
		\item $\lim_{k \rightarrow \infty} f_k(x) = f(x)$
	\end{itemize}
	and that there is a measurable function $g: X \rightarrow [0,\infty]$ such that 
	\begin{itemize}
		\item $|f_k(x)| \leq g(x)$ for each $k$ and $x$, and 
	\end{itemize}

	\begin{itemize}
		\item $\int_X g \dmu < \infty$
	\end{itemize}
	Then \[
		\int_X f \dmu = \lim_{k \rightarrow \infty} \int_X f_k \dmu
	\]
\end{thm}

\begin{thm}[Bounded convergence theorem]
	Let $(f_k)$ be a sequence of real or complex valued measurable function on $X$.  Assume that \begin{itemize}
		\item $\lim_{k \rightarrow \infty} f_k(x) = f(x)$ for each $x \in X$,
		\item There exists a constant $M < \infty$ such that $|f_k(x)| \leq M$ for each $k$ and $x$,
		\item $\mu(X) < \infty$.
	\end{itemize}
	Then \[
		\int_X f \dmu = \lim_{k \rightarrow \infty} \int_X f_k \dmu
	\]
	\end{thm}

\begin{thm}
	Suppose that $f_k$ is a measurable real or complex valued function on $X$ for $k = 0,1,\dots$.  Suppose that we have \[
		\sum_{k = 0}^\infty \left(\int_X |f_k| \dmu \right) < \infty
	\] or, equivalently,
	\[
		\int_X \left( \sum_{k = 0}^\infty |f_k| \right) \dmu < \infty
	\]
	Then we have \[
	\int_X \left( \sum_{k = 0}^\infty f_k \right) \dmu =
	\sum_{k = 0}^\infty \left(\int_X f_k \dmu \right)	
	\]
\end{thm}

\begin{defn}[Integrable function]
	We call $f : X \rightarrow \mathbb{K}$ $\mu$ -integrable if $f$ is $\mu$ measurable and

	\[
		\int_X |f| \, d \mu < \infty
	\]

	We set \[
		\mathcal{L}^1(X,\mathbb{K}) = \{ f : X \rightarrow \mathbb{K} \, | \, f \mu \text{-integrable} \}
	\]
\end{defn}

\begin{thm}
	$	\mathcal{L}^1(X,\mathbb{K})$ is a vector space over $\mathbb{K}$
\end{thm}

\begin{defn}[The Lebesgue-Stieltjes integral]
Let $F : \R \rightarrow \R$ be an increasing right continuous function, that is $\lim_{s \rightarrow t^+} F(s) = F(t)$ for all $t \in \R$.  Then for $A \subseteq \R$ let \[
	\mu{^\star}_F (A) = \inf \{ \sum_{k=0}^\infty (F(b_k) - F(a_k)) \, | \, A \subseteq \bigcup_{k \in \mathbb{N}} (a_k, b_k)
\]
The $\mu^\star_F$ is an outer measure inducing an inner measure on $\R$.  Then we have 
	\begin{itemize}
		\item $\mu_F$ is a Borel measure.
		\item $\mu_F((a,b]) = F(b) - F(a)$.
	\end{itemize}

	We then define $\int_A f \, dF = \int_A f \, d \mu_F $ as the Lebesgue-Stieltjes integral.
\end{defn}

\begin{lem}
	If $\mu$ is a finite measure on $\R$, then we define $F(t) = \mu((-\infty, t])$ as the \textbf{distribution} function of $\R$.
\end{lem}

\begin{thm}
	There is a bijection from finite measures and some class of right-continuous increasing functions.
\end{thm}

\begin{defn}[Measures from other measures]
	Let $g: X \rightarrow [0,\infty]$ be a $\mu$-measurable function.  For $A \in \mathcal{A}$ define \[
		\nu(A) = \int_A g \dmu
	\]  
Then using the monotone convergence theorem one can show that $\nu$ is a measure defined on $\siga$.  Moreover, if $f: X \rightarrow \mathbb{K}$ is $\mu$-measurable, then \[
	\int_X f \, d \nu = \int_X fg \dmu
\]
We call $g$ the \textbf{density of $\nu$ with respect to $\mu$}.
\end{defn}

\begin{prop}
	Let $f \in 	\mathcal{L}^1(X,\mathbb{K})$ with respect to the Lebesgue measure.  Then \[
		\int_a^b f(x) \, dx = \lim_{t \rightarrow b^-} \int_a^t f(x) \, dx
	\]
\end{prop}

\subsection{Parameter integrals} % (fold)
\label{sub:parameter_integrals}
\newcommand{\ellone}[1]{\mathcal{L}^1(X,#1)}
\begin{defn}[Parameter integral]
	Let $(X, \siga, \mu)$ be a measure space and $Y$ a metric space.  Suppose that $f : X \times Y \rightarrow \mathbb{K}$ is such that 
	 \begin{itemize}
	 	\item $x \mapsto f(x,y)$ is $\mu$-integrable for all $y \in Y$,
	 	\item $y \mapsto f(x,y)$ is continuous at $y_0$ for almost all $x \in X$,
	 	\item there exists $g \in \ellone{\R}$ such that \[
	 		|f(x,y)| \leq g(x)
	 	\] for almost all $x \in X$.

	 \end{itemize}
		Define $F(y) = \int_X f(x,y) \dmu(x)$.  Then $F$ is continuous at $y_0 \in Y$.

\end{defn}

\begin{thm}[Differentiation of parameter integrals.]
	Let $(X, \siga, \mu)$ be a measure space and $L \subset \R$ an interval.  Suppose that $f: X \times L \rightarrow \R$ is such that 
	\begin{itemize}
		\item $x \mapsto f(x,y)$ is $\mu$-integrable for all $y \in Y$,
		\item $\frac{\partial}{\partial t} f(x,t)$ exists for all $t \in L$, for almost all $x \in X$, and is continuous ,
		\item there exits $g \in \ellone{\R}$ with $|\frac{\partial}{\partial t} f(x,t)| < g(x)$ for almost all $x \in X$ and all $t \in L$.

	\end{itemize}

	Define $F(t) = \int_X f(x,t) \dmu(x)$.  Then $f : L \rightarrow \mathbb{K}$ is differentiable and \[
		F'(t) = \int_X \frac{\partial}{\partial t} f(x,t) \dmu(x)
	\]
\end{thm}

% subsection integration_of_real_and_complex_valued_functions (end)

% section integration (end)

%!TEX root = /Users/ajtulloch/Documents/University 2010/MATH 3969 - Measure Theory and Fourier Analysis/Exam Notes/MATH 3969 - Master Notes.tex

\newcommand{\K}{\mathbb{K}}
\newcommand{\ellp}[1]{\mathcal{L}^{#1}(X)}
\newcommand{\Lp}[1]{L^{#1}(X)}

\section{The $L^p$-spaces} % (fold)
\label{sec:the_l_p_spaces}

\begin{defn}[$L^p$-spaces]
	Let $1 \leq p < \infty$ and $f:X \rightarrow \K$ measurable. We call \[
		\| f \|_p = \left(\int_X |f|^p \dmu\right)^\frac{1}{p}
	\]
	the \emph{$L^p$-norm} of $f$.  We set \[
		\ellp{p} = \{ f: X \rightarrow \K \, | \, f \text{ measurable}, \| f \|_p < \infty
	\]  
\end{defn}

\begin{thm}[H\"older's inequality]
	Let $p,q \in [1,\infty]$ such that $\frac{1}{p}+\frac{1}{q} = 1$.  If $f \in \ellp{p}$ and $g \in \ellp{q}$, then \[
		| \int_X fg \dmu | \leq \|f\|_p \|g\|_q
	\]
\end{thm}

\begin{prop}[Minkowski's inequality]
	If $f,g \in \ell{p}$, $1\leq p \leq \infty$, then \[
		\|f + g \|_p \leq \|f \|_p + \|g \|_p
	\]
\end{prop}

\begin{defn}[$L^p$-spaces]
	Let $f \sim g$ if $f = g$ almost everywhere.  Denote the equivalence class of $f$ by $[f]$.  Then \[
		\Lp{p} = \{ [f] \, | \, f \in \ellp{p} \} 
	\]
\end{defn}

\begin{defn}[Cauchy sequence]
	A sequence $(f_n) \in \Lp{p}$ is called \textbf{Cauchy} if for every $\epsilon > 0$ there exists $n_0 \in \mathbb{N}$ such that \[
		\| f_n - f_m \|_p < \epsilon
	\] for all $n,m > n_0$.
\end{defn}

\begin{thm}[Completeness of $\Lp{p}$]
	Let $(f_n)$ be a sequence in $\Lp{p}$.  Then $(f_n)$ converges in $\Lp{p}$ if and only if $(f_n)$ is a Cauchy sequence.
\end{thm}

\begin{rem}
	Introducing the metric $d(f,g) = \|f-g\|_p$, we have that $\Lp(X)$ is a \textbf{complete normed space} or a \textbf{Banach space}. If $p = 2$, then $\|f \|_2 $ is induced by an inner product - hence $\Lp{2}$ is a \textbf{complete inner product space}, or a \textbf{Hilbert space.}
\end{rem}

\begin{prop}
	Suppose that $f_n, f \in \ellp{p}$ with $\|f_n - f \| \rightarrow 0$. Then there exists a subsequence $(f_{n_k})$ with $f_{n_k}$ converging pointwise to $f$ for almost every $x \in X$.
\end{prop}

\begin{thm}
	The simple functions are dense in $\Lp{p}$ for $1 \leq p < \infty$.

	In $\R^N$ and the Lebesgue measure, we can modify the statement to the simple function with bounded support are dense in $\R^N$. 
\end{thm}

\begin{thm}
	For $1 \leq p < \infty$
	\[
		\text{span} \{1_U \, | \, U \subseteq \R^N \text{ open and bounded} \}
	\] is dense in $L^p(\R^N)$.

	We can also use bounded rectangles in the place of open bounded sets here.
\end{thm}

\begin{defn}[Essential supremum]
	Let $(X, \siga, \mu)$ be a measure space and $f : X \rightarrow \R$ $\mu$-measurable.  We call $\text{ess-sup} f(x) = \inf \{ t \in \R \, | \, \mu(\{ x \in X \, | \, f(x) > t \} ) = 0 \}$ the \emph{essential supremum} of $f$.  The essential supremum of $|f|$ is denoted $\| f \|_\infty$
\end{defn}

\begin{thm}[Completeness of $\Lp{\infty}$]
	$\Lp(\infty)$ is a complete normed space.
\end{thm}

\begin{lem}
	H\"older's inequality holds for $p = 1, q = \infty$.  That is, \[
		| \inf_X fg \dmu | \leq \|f \|_p \|g \|_q
	\]
\end{lem}

\begin{lem}
	If $\mu(X) < \infty$, then $\lim_{p \rightarrow \infty} \|u\|_p = \|f\|_\infty$
\end{lem}

\subsection{Fubini's Theorem} % (fold)
\label{sub:fubini_s_theorem}

\begin{thm}[Tonelli]\label{tonelli}
	Suppose that $f: \R^n \times \R^m \rightarrow [0,\infty]$ is measurable. Then there exist sets $N \subseteq \R^n$ and $M \subseteq \R^m$ of measure zero such that

	\begin{enumerate}[(i)]
		\item $x \mapsto f(x,y)$ is measurable for all $y \in \R^m - M$,
		\item $y \mapsto \int_{\R^n} f(x,y) \, dx$ is measurable,
		\item $y \mapsto f(x,y)$ is measurable for all $x \in \R^m - N$,
		\item $x \mapsto \int_{\R^m} f(x,y) \, dy$ is measurable,
		\item  \begin{align*}
			&\int_{\R^n \times \R^m}f(x,y)\, d(x,y) = \\
			&\int_{R^m-M} \left( \int_\R^n f(x,y) dx \right) dy = \\
			&\int_{R^n-N} \left( \int_\R^m f(x,y) dy \right) dx
		\end{align*}
	\end{enumerate}
\end{thm}

\begin{thm}[Fubini]
	Suppose that $f: \R^n \times \R^m \rightarrow [0,\infty]$ is measurable. Let $N,M$ be the sets from Theorem \ref{tonelli} applied to the function $|f| $ such that (v) holds with $f$ replaced with $|f|$.  Assume that one of these integrals is finite - and hence all of them.  Then there exists sets $N_1$ of $\R^n$ and $M_1$ of $\R^m$ such that $(i) - (v)$ of Theorem \ref{tonelli} hold with $N,M$ replaced with $N_1, M_1$.
\end{thm}

\begin{defn}[Complete measure space]
	Let $(X,\siga, \mu)$ be a measure space.  We call the measure $\mu$ \emph{complete} if whenever $A \in \siga$ has measure $0$, then any subset of $A$ is in $\siga$, (and has measure 0).
\end{defn}
\begin{defn}[$\sigma$-finite measure space]

\end{defn}

% subsection fubini_s_theorem (end)

% section the_l_p_spaces (end)

%!TEX root = /Users/ajtulloch/Documents/University 2010/MATH 3969 - Measure Theory and Fourier Analysis/Exam Notes/MATH 3969 - Master Notes.tex

\section{Convolution} % (fold)
\label{sec:convolution}

\begin{defn}[Translation of a function]
 	Let $f: \R^N \rightarrow \C$ be a function and $t \in \R^N$ a fixed vector. We define the translation operator $\tau_t$ by \[
 		\tau_t f(x) = f(x-t)
 	\]
\end{defn}

\begin{thm}[Continuity of translation]
	Let $1 \leq p < \infty$ and $f \in L^p(\R^N)$.  Then \[
		\lim_{t \rightarrow 0} \| \tau_t f - f \|_p = 0
	\]
\end{thm}
\begin{rem}
	This does not hold if $p = \infty$.
\end{rem}

\begin{lem}
	Let $f : \R^N \rightarrow \C$ be measurable and set 
	\begin{align*}
		F_1(x,y) &= f(x) \\
		F_2(x,y) &= f(y-x)
	\end{align*}

	Then $F_1, F_2 : \R^N \times \R^N \rightarrow \C$ are measurable.
\end{lem}

\begin{defn}[Convolution]
	Let $f,g : \R^N \rightarrow \C$ be measurable.  We define the \textbf{convolution} $f \star g : \R^N \rightarrow \C$ by \[
		(f \star g) (x) = \int_{\R^N} f(x-y) g(y) \, dy
	\]
	wherever the integral exists
\end{defn}

\begin{defn}[Convex function]
	A function $\phi : (a,b) \rightarrow \R$ is called \textbf{convex} if \[
		\phi(\lambda s + (1-\lambda)t) \leq \lambda \phi(s) + (1-\lambda) \phi(t)
	\] for all $s,t \in (a,b)$ and all $\lambda \in (0,1)$
\end{defn}

\begin{lem}
	This is equivalent to the condition\[
		\frac{\phi(t) - \phi(s)}{t-s} \leq \frac{\phi(u) - \phi(t)}{u - t}
	\] whenever $a < s < t < u < b$.
\end{lem}

\begin{thm}[Jensen's inequality in $\ellp{p}$-spaces]
	Let $f \in \ellp{p}$, $1\leq p < \infty$, and let $g \in \ellp{1}$.  Then \[
		\left(\int_X |fg| \dmu \right)^p \leq \|g\|^{p-1}_1 \int_X |f|^p |g| \dmu
	\]
\end{thm}

\begin{thm}[Young's inequality]
	Let $1 \leq p \leq \infty$.  If $f \in L^p(\R^N, \C)$ and $g \in L^1(\R^N, \C)$, then $f \star g$ exists almost everywhere and $f \star g \in L^p(\R^N, \C)$.  Moverover, \[
		\|f \star g \|_p \leq \|f\|_p \|g \|_1
	\]
\end{thm}

\begin{thm}
	Let $1 \leq p, q \leq \infty$ with $\frac{1}{p} + \frac{1}{q} = 1$.  If $f \in L^p(\R^N)$ and $g \in L^q(\R^N)$, then \[
		f \star g \in BC(\R^N)
	\] where $BC$ is the vector space of bounded continuous functions.
\end{thm}

\subsection{Approximate identities} % (fold)
\label{sub:approximate_identities}

\begin{defn}[Approximate identity]
	Let $\phi: \R^N \rightarrow [0,\infty)$ be measurable with \[
		\int_{\R^N} \phi \, dx = 1
	\] and set $\phi_n(x) = n^N \phi(n x)$ for all $x \in \R^N$ and $n \in \mathbb{N}$.  Then $(\phi_n)$ is called an \textbf{approximate identity}
\end{defn}

\begin{thm}
	Let $(\phi_n)$ be an approximate identity and $f \in L^p(\R^N), 1 \leq p < \infty$.  Then \[
		f \star \phi_n \rightarrow f
	\] in $L^p(\R^N)$ as $n \rightarrow \infty$
\end{thm}

\begin{thm}
	Let $f \in L^\infty(\R^N)$ and $(\phi_n)$ an approximate identity.  If $f$ is continuous at $x$, then \[
		f(x) = \lim_{n \rightarrow \infty} (f \star \phi_n) (x)
	\]
\end{thm}

% subsection approximate_identities (end)

\begin{defn}[Test function]
	Let $U \subseteq \R^N$ be open.  We let \[
		C^\infty (U, \K) = \{ f: U \rightarrow \K \, | \, f \text{ has partial derivatives of all orders} \}
	\]
	and \[
		C_c^\infty(U, \K) = \{ f \in C^\infty(U, \K) \, | \, \text{supp}(f) \subseteq U, \text{supp($f$) compact}
	\]

	The functions in $C_c^\infty(U, \K)$ are called \textbf{test functions} on $U$.
\end{defn}

\begin{prop}
	Let $f : \R^N \rightarrow \K$ be measurable such that $f \in \mathcal{L}^1(B)$ for every bounded set $B \subseteq \R^N$.  If $\phi \in C_c^\infty(\R^N)$, then $f \star \phi \in C^\infty(\R^N)$ and \[
		\frac{\partial}{\partial x_i} (f \star \phi) = f \star \frac{\partial \phi}{\partial x_i}
	\]
\end{prop}

\begin{thm}
	Let $U \subseteq \R^N$ open and $1 \leq p < \infty$.  Then $C_c^\infty(U)$ is dense in $L^p(U)$.
\end{thm}
\begin{rem}
	The above proposition does not hold for $p = \infty$.
\end{rem}
% section convolution (end)

%!TEX root = /Users/ajtulloch/Documents/University 2010/MATH 3969 - Measure Theory and Fourier Analysis/Exam Notes/MATH 3969 - Master Notes.tex

\renewcommand{\hat}{\widehat}

\section{The Fourier Transform} % (fold)
\label{sec:the_fourier_transform}

\begin{defn}[Fourier transform]
	Let $f \in L^1(\R^N, \C)$.  We call \[
		\hat{f}(t) = \int_{\R^N} f(x) e^{-2 \pi i x \cdot t} \, dx
	\]
\end{defn}

\begin{thm}
	We have 
	\begin{itemize}
		\item $\hat{f} : \R^N \rightarrow \C$ is continuous,
		\item $\|\hat f \|_\infty \leq \|f \|_1$
	\end{itemize}
\end{thm}

\begin{prop}
	Let $\phi(x) = e^{- \pi |x|^2}$.  Then $\|\phi \|_1 = 1$ and $\hat \phi = \phi$.
\end{prop}

\begin{prop}
	Let $f \in L^1(\R^N, \C), x_0 \in \R^N$ and $\alpha \in R, \alpha > 0$. 
	\begin{enumerate}[(i)]
		\item If $g(x) = f(x - x_0)$, then $\hat g (t) = e^{-2 \pi i x_0 \cdot t} \hat f (t)$,
		\item If $g(x) = f(\alpha x)$, then $\hat g (t) = \frac{1}{\alpha^N} \hat f \left( \frac{x}{\alpha}\right)$,
		\item \vspace{0.1cm} If $g(x) = \overline{f(-x)}$, then $\hat g (t) = \overline{\hat f(t)}$
	\end{enumerate}
\end{prop}

\begin{defn}
	Let $C_0(\R^N, \K) = \{ f \in C(\R^N, K) \, | \, \lim_{|x| \rightarrow \infty} f(x) = 0 \}$, the set of continuous functions vanishing at infinity
\end{defn}

\begin{thm}[Riemann-Lebesgue]
	If $f \in L^1(\R^N, \C)$, then $\hat f \in C_0(\R^N, \C)$ and $\| \hat f \|_\infty \leq \|f \|_1$.
\end{thm}

\begin{thm}
	If $f,g \in L^1(\R^N, \C)$ then $f \star g \in L^1(\R^N, \C)$ and \[
		\hat{f \star g} = \hat f \, \hat g
	\]
\end{thm}
\begin{prop}
	Let $f,g \in L^1(\R^N, \C)$.  Then \[
		\int_{\R^N} \hat f g \, dx = \int_{\R^N} f \hat g \, dx
	\]
\end{prop}

\begin{lem}
	Let $\phi(x) = e^{- \pi |x|^2}$ and $\phi_n(x) = n^N \phi(n x)$.  Then \[
		\int_{\R^N} \hat f (t) e^{2 \pi i x \cdot t} \phi\left(\frac{t}{n} \right) \, dt = (f \star \phi_n) (x)
	\]
	for all $f \in L^1(\R^N,\C), x \in \R^N$, and $n \in \mathbb{N}$.
\end{lem}

\begin{thm}[Fourier inversion formula]
	Let $f \in L^1(\R^N, \C)$. Then 
	\begin{enumerate}[(i)]
		\item \[
			\lim_{n \rightarrow \infty} \int_{\R^N} \hat f (t) e^{2 \pi i x \cdot t} e^{- \pi \frac{|t|^2}{n^2}} \, dt = f
		\] in $L^1(\R^N, \C)$.
		\item If $f$ is continuous at $x$, then \[
			\lim_{n \rightarrow \infty} \int_{\R^N} \hat f (t) e^{2 \pi i x \cdot t} e^{- \pi \frac{|t|^2}{n^2}} \, dt = f(x)
		\]
	\end{enumerate}
\end{thm}

\begin{cor}
	Let $f,g \in L^1(\R^N)$ with $\hat f = \hat g$. Then $f = g$ almost everywhere. 
\end{cor}

\subsection{The Fourier transform on $L^2(\R^N)$} % (fold)
\label{sub:the_fourier_transform_on_l_2_r_n_}
	We have defined the Fourier transform $\hat f$ with $f \in L^1(\R^N)$.  We  have that $C_c^\infty(\R^N)$ is dense in $L^2(\R^N)$ as well as in $L^1(\R^N)$, so in particular $L^2(\R^N) \cap L^1(\R^N)$ is dense in $L^2(\R^N)$. We can use this to extend the Fourier transform to $L^2(\R^N)$.  The key for doing so is the following theorem.  
	\begin{thm}[Plancherel]
		Let $f \in L^2(\R^N) \cap L^1(\R^N)$. Then $\| \hat f \|_2 = \| f \|_2$. 
	\end{thm}

\begin{prop}
	There is a unique coninuous linear operator \[
		\mathcal{F} : L^2(\R^N) \rightarrow L^2(\R^N)
	\] such that $\mathcal{F} f = \hat f$ for all $f \in L^2(\R^N) \cap L^1(\R^N)$.  Moreover, $\| f \|_2 = \| \mathcal{F} f \|_2$ for all $f \in L^2(\R^N)$.
\end{prop}

\begin{rem}
	We use the notation $\hat f = \mathcal{F} f$ for $f \in L^2(\R^N)$.  
\end{rem}
\begin{rem}
	Let $\phi_n : \R^N \rightarrow [0,1]$ such that $\phi_n \in L^2(\R^N)$ and $\phi_n(x) \rightarrow 1$ for all $x \in \R^N$. If $f \in L^2(\R^N)$, then \[
		\hat f = \lim_{n \rightarrow \infty} \int_{\R^N} f(x) \phi_n(x) e^{-2 \pi i x \cdot t} \, dx
	\]
	Common choices for $\phi_n$ are 
	\begin{itemize}
		\item $\phi_n(x) = 1_{B(0,n)},$
		\item $\phi_n(x) = e^{- \pi \frac{|x|^2}{n^2}}$
	\end{itemize}
\end{rem}

\begin{thm}
	Let $f \in L^2(\R^N, \C)$.  Then \[
		\lim_{n \rightarrow \infty} \int_{\R^N} \hat f (t) e^{2 \pi i x \cdot t} e^{- \pi \frac{|t|^2}{n^2}} \, dt = f
	\] in $L^2(\R^N, \C)$.
\end{thm}

\begin{thm}
	$\mathcal{F} : L^2(\R^N) \rightarrow L^2(\R^N)$ is bijective with $(\mathcal{F}^{-1} f) (x) = (\mathcal{F} f) (-x)$ for all $f \in L^2(\R^N)
$.
\end{thm}

\begin{rem}
	Let \[
		\langle f, g \rangle = \int_{\R^N} f(x) \overline{g(x)} \, dx
	\] denote the inner product on $L^2(\R^N)$.  Then the above theorem implies \[
		\langle \hat f, \hat g \rangle = \langle f, g \rangle.
	\]
	Moverover, by approximating $f,g$ by functions in $L^2(\R^N) \cap L^1(\R^N)$ we also have \[
		\langle \hat f, g \rangle = \langle f, \overline{\hat g} \rangle
	\]
\end{rem}

% subsection the_fourier_transform_on_l_2_r_n_ (end)
% section the_fourier_transform (end)

%!TEX root = /Users/ajtulloch/Documents/University 2010/MATH 3969 - Measure Theory and Fourier Analysis/Exam Notes/MATH 3969 - Master Notes.tex

\section{The Radon-Nikodym Theorem} % (fold)
\label{sec:the_radon_nikodym_theorem}

\subsection{The Reisz representation theorem} % (fold)
\label{sub:the_reisz_representation_theorem}

% subsection the_reisz_representation_theorem (end)

Let $H$ be an inner product space with inner product $(x|v)$.  Then $H$ is a normed space with norm \[
	|u \| = \sqrt{(u | u)}
\]
We call $H$ a \textbf{Hilbert space} if $H$ is complete with respect to $\| . \|$, that is, every Cauchy sequence in $H$ converges.

\begin{thm}[Projections]
	Let $H$ be a Hilbert space and $M \subseteq H$ a closed subspace of $H$.  Let $u \in H$.  Then there exists $m_0 \in M$ such that \[
		\| u - m_0 \| = \min_{m \in M} \| u - m \|
	\]
	Moreover, \[
		(u - m_o | m) = 0
	\] for all $m \in M$.
\end{thm}

\begin{rem}
	Fix $g \in H$ and consider the function $\phi_g : H \rightarrow \K$ given by \[
		\phi_g(f) = (f | g)
	\] 
	Then $\phi_g$ is linear, and by the Cauchy-Swartz inequality, \[
		| \phi_g(f) = |(f | g) | \leq \|f \| \|g \|
	\] for all $f \in H$.  We say $\phi_g$ is a bounded linear functional on $H$. 
\end{rem}

\begin{defn}
	Let $H$ be a Hilbert space.  We call a linear operator $\phi: H \rightarrow \K$ a \textbf{bounded linear functional} on $H$ if there exists $M > 0$ such that \[
		| \phi(f) | \leq M \| f \|
	\] for all $f \in H$.
\end{defn}

\begin{thm}[Riesz representation theorem]
	Let $H$ be a Hilbert space over $\K$ and $\phi(H \rightarrow K)$ a bounded linear function.  Then there exists $g \in H$ such that \[
		\phi(f) = (f | g)
	\] for all $f \in H$.
\end{thm}

\subsection{The Radon-Nikodym Theorem} % (fold)
\label{sub:the_radon_nikodym_theorem}
 Suppose that $\mu$ is a measure defined on the $\sigma$-algebra $\siga$ of subsets of $X.$  Given a measurable function $g : X \rightarrow [0,\infty]$ we define \[
 	\nu(A) = \int_A g \dmu
 \]
Then $\nu$ is a measure defined on the $\sigma$-algebra $\siga$.

The converse does not necessarily hold - that is, given two measures $\mu$ and $\nu$ on a $\sigma$-algebra $\siga$, there is not always a measurable function $g : X \rightarrow [0,\infty]$ such that the above equation holds.

\begin{defn}[Absolute continuity]
	Let $\nu, \mu$ be the measures defined on a $\sigma$-algebra $\siga$. We call \textbf{$\nu$ absolutely continuous with respect to $\mu$} if $\nu(A) = 0$ whenever $\mu(A) = 0$.  In that case, we write $\nu << \mu$.
\end{defn}

\begin{prop}
	Let $\mu, \nu$ be measures defined on a $\sigma$-albegra.  Suppose that $\mu(X), \nu(X) < \infty$.  Set $\lambda = \mu + \nu$.  Then there exists a measurable function $h : X \rightarrow [0,\infty] $ such that \[
		\int_X f \, d \nu = \int_X f h \, d \lambda
	\] for all $f \in L^2(X, \lambda)$
\end{prop}

\begin{thm}[Radon-Nikodym]
	Let $\mu,\nu$ be measures defined on a $\sigma$-algebra.  Suppose that $\nu$ and $\mu$ are $\sigma$-finite and that $v << u$.  Then there exits a measurable function $g : X \rightarrow [0,\infty)$ such that $\nu(A) = \int_A g \dmu$ for all $A \in \siga$. 
\end{thm}

Formally we can write \[
	\int_X f \, d \nu = \int_X f \frac{d \nu}{d\mu} \dmu
\] if we define $g = \frac{d \nu}{d\mu}$, where $g$ is the density function from the Radon-Nikodym theorem.

\begin{rem}
	If $g$ is the function in the Radon-Nikodym theorem, it is not hard to show that \[
		\int_X f \, d \nu = \int_X f g \dmu
	\] for all $f \in L^1(X, \nu)$.
\end{rem}
% subsection the_radon_nikodym_theorem (end)

% section the_radon_nikodym_theorem (end)

%!TEX root = /Users/ajtulloch/Documents/University 2010/MATH 3969 - Measure Theory and Fourier Analysis/Exam Notes/MATH 3969 - Master Notes.tex

\section{Probability Theory} % (fold)
\label{sec:probability_theory}

\begin{defn}[Random variable]
	Let $(\Omega, \siga, P)$ be a probability space.  A $\siga$-measurable function \[
		X : \Sigma \rightarrow \R
	\] is called a \textbf{random variable}. 
\end{defn}

\begin{defn}
	Let $X : \Sigma \rightarrow \R$ a random variable.  We say that $X$ has \emph{finite expectation} if $X \in L^1(\Sigma)$ and call \[
		E[X] = \int_\Sigma X \, dP
	\] the \textbf{expectation} of $X$.  
\end{defn}

\begin{defn}[Distrbution]
	For every Borel set $A \subseteq \R$ we define \[
		P_X[A] = P[ \{ \omega \in \Omega | X(\omega) \in A\}] = P[X \in A]
	\]
\end{defn}

Since $X$ is measurable, $X^{-1}[A]$ is measurable for all Borel sets $A \subseteq \R$.

\begin{defn}[Distribution]
	Let $X$ be a random variable. The Borel measure defined above is called the \textbf{distribution} of $X$.  The function \[
		F(t) = P_X\left[(-\infty,t] \right] = P[X \leq t]
	\] is called the \textbf{distribution function} of $X$.
\end{defn}

\begin{lem}
	Let $X$ be a random variable and let $f: \R \rightarrow \R$ be Borel measurable.  Then \[
		\int_\Sigma f \circ X \, dP = \int_\R f \, dP_X
	\]
\end{lem}

\subsection{Conditional expectation} % (fold)
\label{sub:conditional_expectation}

\begin{defn}[Conditional expectation]
	Let $X: \Omega \rightarrow \R$ a random variable with finite expectation.  Let $\mathcal{A}_0$ be a $\sigma$-algebra with $\siga_0 \subseteq \siga$. We call \[
		X_0 : \Sigma \rightarrow \R
	\] a \textbf{conditional expectation} given $\siga_0$ if 
	\begin{itemize}
		\item $X_0$ is $\siga_0$-measurable
		\item $\int_A X_0 \, dP = \int_A X \, dP$ for all $A \subseteq \siga_0$.
	\end{itemize}
	We write $X_0 = E[X | \siga_0]$
\end{defn}

\begin{thm}
	Let $X$ be a random variable with finite expectation.  If $\siga_0$ is a $\sigma$-algebra with $\siga_0 \subseteq \siga$, then the conditional expectation $X_0 = E[X | \siga_0]$ exists and is essentially unique.
\end{thm}

\begin{rem}
	\begin{itemize}
		\item If $X$ is $\siga_0$-measurable, then $X = E[X | \siga_0]$ almost everywhere.
		\item If we set $\siga_0 = \{ \phi, \Omega \}$, then \[
			E[X | \siga_0] = E[X]
		\] 
	\end{itemize}
\end{rem}

% subsection conditional_expectation (end)
% section probability_theory (end)

\end{document}

### 2011-12-12 MATH 3975 - Financial Mathematics/MATH 3975 Lecture Notes.tex
Initial Commit

%
%  untitled
%
%  Created by Andrew Tulloch on 2009-11-04.
%  Copyright (c) 2009 __MyCompanyName__. All rights reserved.
%!TEX TS-program = xelatex
%!TEX encoding = UTF-8 Unicode
\documentclass[10pt, oneside, reqno]{amsart}
\usepackage{geometry, setspace, graphicx, enumerate}
\onehalfspacing                 
\usepackage{fontspec,xltxtra,xunicode}
\defaultfontfeatures{Mapping=tex-text}

%\setromanfont[Mapping=tex-text,Contextuals= 
%{NoWordInitial,NoWordFinal,NoLineInitial,NoLineFinal}]{Hoefler Text}
%\setsansfont[Scale=MatchLowercase,Mapping=tex-text]{Hoefler Text}
%\setmonofont[Scale=MatchLowercase]{Andale Mono}

% \setcounter{section}{3}
% \setcounter{subsection}{1}

	% AMS Theorems
	\theoremstyle{plain}% default 
	\newtheorem{thm}{Theorem}[subsection] 
	\newtheorem{lem}[thm]{Lemma} 
	\newtheorem{prop}[thm]{Proposition} 
	\newtheorem*{cor}{Corollary}

	\theoremstyle{definition} 
		\newtheorem{defn}[thm]{Definition}
		\newtheorem{conj}[thm]{Conjecture}
		\newtheorem{exmp}[thm]{Example}

	\theoremstyle{remark} 
		\newtheorem*{rem}{Remark} 
		\newtheorem*{note}{Note} 
		\newtheorem{case}{Case}

		\newcommand{\expc}[1]{\mathbb{E}\left[#1\right]}
		\newcommand{\expp}[1]{\mathbb{E}_{\rnm}\left[#1\right]}

		\newcommand{\var}[1]{\text{Var}\left(#1\right)}
		\newcommand{\cov}[1]{\text{Cov}\left(#1\right)}
		\newcommand{\prob}[1]{\mathbb{P}(#1)}
		\newcommand{\given}{ \, | \,}
		\newcommand{\us}{0 \leq u \leq s}
		\newcommand{\ts}[1]{\{ #1 \}}

\newcommand{\al}{\alpha}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\F}{\mathbb{F}}
\newcommand{\Ga}{\mathbb{G}}

\newcommand{\sumni}{\sum_{i=1}^n}

\newcommand{\tilp}{\tilde{p}}
\newcommand{\rnm}{\tilde{\mathbb{P}}}

\newcommand{\aut}[1]{\text{Aut}{(#1)}}

\newcommand{\gener}[1]{\langle #1 \rangle}
\newcommand{\charr}[1]{\text{char}(#1)}
\newcommand{\nth}{n\textsuperscript{th}}

\newcommand{\tworow}[2]{\genfrac{}{}{0pt}{}{#1}{#2}}
\newcommand{\xdeg}[2]{[#1 : #2]}
\newcommand{\gal}[2]{\text{Gal}(#1/#2)}
\newcommand{\minpoly}[2]{m_{#1, #2}(x)}

\newcommand{\mapping}[5]{\begin{align*}
	#1 : \quad     #2 &\rightarrow #3 \\
			#4  &\mapsto #5
\end{align*}	
}

\addtolength{\textwidth}{0.9in}
\addtolength{\hoffset}{-0.45in}
\addtolength{\textheight}{0.9in}
\addtolength{\voffset}{-0.9in}

% \includeonly{MATH3975-Chap1, MATH3975-Chap2}

\title{MATH 3975 - Financial Mathematics \\ Exam Notes}								% Document Title
\author{Andrew Tulloch}
%\date{}                                           % Activate to display a given date or no date

\begin{document}
\maketitle

%!TEX root = /Users/ajtulloch/Documents/University 2010/MATH 3975 - Financial Mathematics/Exam Notes/MATH 3975 - Master Notes.tex

\section{Introduction to Markets} % (fold)
\label{sec:introduction_to_markets}
\subsection{Introduction} % (fold)
\label{sub:introduction}

\begin{defn}[Time Value of Money]
	The \textbf{future value} $F(0,t)$ is defined to be the value at time $t > 0$ of \$1 invested at time $0$.

	The \textbf{present value} or \textbf{discount factor} $P(0,t)$ is the amount invested at time $0$ such that its value at time $t$ is equal to \$1.  
\end{defn}

To avoid arbitrage, we must have $P(0,t) = F^{-1}(0,t)$ for all $t$.

\begin{prop}
	We must have the following relationship for $P(t,T)$ and $F(t,T)$.
	\begin{align*}
		F(t,T) &= \frac{F(0,T)}{F(0,t)} \\
		P(t,T) &= \frac{P(0,T)}{P(0,t)}
	\end{align*}
\end{prop}

\begin{defn}[Spot rate]
	The spot rate $r(t)$ is defined as \[
		F(t,t + \Delta) = 1 + r(t) \Delta
	\]
\end{defn}

\begin{prop}
	We have \[
		P(0,T) = \prod_{i=1}^n (1 + r_i)^{-1}
	\] in the discrete case, and in the continuous case, we have \[
		P(t,T) = e^{- \int_t^T r(s) ds}
	\]

\end{prop}

% subsection introduction (end)

\subsection{Riskless Securities and Bonds} % (fold)
\label{sub:riskless_securities_and_bonds}
\begin{thm}[Fundamental Theorem of Riskless Security Pricing]
	If interest rates are deterministic, the arbitrage free price of a riskless security is given by
	\[
		S_0 = \sum_{i=1}^n P(0,t_i) C_i
	\]
\end{thm}

% subsection riskless_securities_and_bonds (end)

% section introduction_to_markets (end)
%!TEX root = /Users/ajtulloch/Documents/University 2010/MATH 3975 - Financial Mathematics/Exam Notes/MATH 3975 - Master Notes.tex

\newcommand{\tst}{(x,\phi)}

\section{Single Period Market Models} % (fold)
\label{sec:single_period_market_models}

A single period market model is the most elementary model.  Only a single period is considered.  At times $t=0$ and $1$ market prices are recorded.

\begin{quote}
\emph{	Single period market models are the atoms of Financial Mathematics
}\end{quote}

We assume that we have a finite sample space \[
	\Omega = \{ \omega_1, \omega_2, \dots, \omega_k \}
\]

\subsection{The most elementary market model} % (fold)
\label{sub:the_most_elementary_market_model}

Assume the sample space consists of two states, $H$ and $T$, with $\prob{H} = p$ and $ \prob{T} = 1-p$.  Define the price of the stock at time $0$ to be $S_0$, and let $S_1$ be a random variable depending on $H$ and $T$.  Let $u = \frac{S_1(H)}{S_0}$ and $d = \frac{S_1(T)}{S_0}$.

\begin{defn}[Trading strategy]
	A trading strategy $(x,\phi)$ is a pair wheere $x$ is the total initial investment at $t=0$, and $\phi$ denotes the number of shares bought at $t = 0$.  Given a strategy $(x,\phi)$, the agent invests the remaining money $x - \phi S_0$ in a money market account.  We note this amount may be negative (borrowing from the money market account).
\end{defn}

\begin{defn}[Value process]
	The value process of the trading strategy $(x,\phi)$ in ou elementary market model is given by $(V_0(x,\phi), V_1(x,\phi))$ where $V_0(x,\phi) = x$ and \[
		V_1 = (x - \phi S_0)(1+r) +\phi S_1
	\]
\end{defn}

\begin{defn}[Arbitrage]
	An arbitrage is a trading strategy that begins with no money, has zero probability of losing money, and has a positive probability of making money.

	More rigorously, we have a trading strategy $(x,\phi)$ is an arbitrage if
	\begin{itemize}
		\item $x = V_0(x,\phi) = 0$, 
		\item $V_1(x,\phi) \geq 0$,
		\item $\expc{V_1(x,\phi)} > 0$. 
	\end{itemize}

\end{defn}

\begin{prop}
	To rule out arbitrage in our model, we must have $d < 1+r < u$.  
\end{prop}

\begin{proof}
	If this inequality is violated, consider the following strategies.
	\begin{itemize}
		\item 	If $d \geq 1+r$, borrow $S_0$ from the money market.
		\item If $u \leq 1+r$, short $S_0$ and invest in the money market.
	\end{itemize}

	These are both arbitrages and the proposition is proven.
\end{proof}

We have the following theorem, giving the converse of the above proposition
\begin{thm}
	The condition $d < 1+r < u$ is a necessary and sufficient no arbitrage condition.  That is,
	\[
		\text{No arbitrage} \iff d < 1+r < u
	\]
\end{thm}

\begin{defn}[Replicating strategy or hedge]
	A \textbf{replicating strategy} or \textbf{hedge} for the option $h(S_1)$ in our elementary single period market model is a trading strategy $(x,\phi)$ satisfying $V_1(x,\phi) = h(S_1)$.  That is,
	\begin{align*}
		(x-\phi S_0)(1+r) + \phi S_1(H) 	&= h(S_1(H))\\
		(x - \phi S_0)(1+r) + \phi S_1(T) 	&= h(S_1(T))
	\end{align*}
\end{defn}

\begin{thm}
	Let $h(S_1)$ be an option in our market model, and let $(x,\phi)$ be a replicating strategy for $h(S_1)$. Then $x$ is the only price for the option at time $t=0$ which does not allow arbitrage.  
\end{thm}

To find a replicating strategy for an arbitrary option, define \begin{align*}
	\phi &= \frac{h(S_1(H)) - h(S_1(T))}{S_1(H) - S_1(T)} \\
	\tilp &= \frac{1 + r - d}{u - d}
\end{align*}

Then by solving the above two equations for $x$, we have \[
	x = \mathbb{E}_{\rnm}\left[\frac{1}{1+r} h(S_1)\right] = \frac{1}{1+r}\left[\tilp h(S_1(H)) + (1-\tilp) h(S_1(T))\right]
\]

% subsection the_most_elementary_market_model (end)

\subsection{A general single period market model} % (fold)
\label{sub:a_general_single_period_market_model}

% subsection a_general_single_period_market_model (end)

\begin{defn}[Trading strategy]
	A \textbf{trading strategy} for an agent in our general single period market model is a pair $(x,\phi)$, where $\phi = (\phi^1, \dots,\phi^n) \in \R^n$ specifying the initial investment in the $i$-th stock.
\end{defn}

\begin{defn}[Value and gains process]
	Let the \textbf{value process} of the trading strategy $(x,\phi)$ is given $(V_0(x,\phi), V_1(x,\phi))$ where $V_0(x,\phi) = x$ and \[
		V_1(x,\phi) = (x - \sum_{i=1}^n \phi^i S_0^i)(1+r) + \sum_{i=1}^n \phi^i S_1^i
	\]

	The \textbf{gains process} is defined as \[
		G(x,\phi) = (x - \sum_{i=1}^n \phi^i S^i_0) r + \sum_{i=1}^n \phi^i \Delta S^i
	\]
	where $\Delta S^i$ is defined as \[
		\Delta S^i = S^i_1 - S^i_0
	\]
\end{defn}

We have the simple result \[
	V_1(x,\phi) = V_0(x,\phi) + G(x,\phi)
\]

To study the prices of the stocks in relation to the money market account, we introduce the \textbf{discounted stock prices} $\hat{S}^i_t$ defined as follows: \begin{align*}
	\hat{S}^i_0 = S^i_0 \\
	\hat{S}^i_1 = \frac{1}{1+r} S^i_1
\end{align*}

\begin{defn}[Discounted value and gains process]
We define the \textbf{discounted value process} $\hat{V}(x,\phi)$ by \begin{align*}
	\hat{V}_0 (x,\phi) &= x \\
	\hat{V}_1 (x,\phi) &= (x - \sumni \phi^i S^i_0) + \sumni \phi^i \hat{S}^i_1
\end{align*}
and the discounted ains process $\hat{G}(x,\phi)$ as \[
	\hat{G}(x,\phi) = \sumni \phi^i \Delta \hat{S}^i
\] with $\Delta \hat{S}^i = \hat{S}^i_1 - \hat{S}^i_0$.
\end{defn}

We then have the relation \[
	\hat{V}_1(x,\phi) = \hat{V}_0(x,\phi) + \hat{G}(x,\phi)
\]

\begin{defn}[Arbitrage]

	A trading strategy $(x,\phi)$ is an arbitrage in our general single period market model if
	\begin{itemize}
		\item $x = V_0(x,\phi) = 0$, 
		\item $V_1(x,\phi) \geq 0$,
		\item $\expc{V_1(x,\phi)} > 0$. 
	\end{itemize}

	Alternatively, if a trading strategy satisfies the first two conditions above, it is an arbitrage if the following condition is satisfied: \[
		\text{There exists } \omega \in \Omega \text{ with } V_1(x,\phi) > 0.
	\]

Alternatively, we can replace all references to $V$ in the above definition with $\hat{V}$.
\end{defn}

\begin{defn}[Risk neutral measure]
	A measure $\rnm$ on $\Omega$ is a \textbf{risk neutral measure} if 
	\begin{itemize}
		\item $\rnm(\omega) > 0$ for all $\omega \in \Omega$
		\item $\mathbb{E}_{\rnm} \left[\Delta \hat{S}^i \right] = 0$ for all $i$
	\end{itemize}
\end{defn}

\begin{thm}[Fundamental Theorem of Asset Pricing]
	In the general single period market model, there are no arbitrages if and only if there exists a risk neutral measure for the market model.
\end{thm}

\begin{defn}[Alternative definition of arbitrage]
	Define the set $\mathbb{W}$ by the following:\[
			\mathbb{W} = \{ X \in \R^k \given X = \hat{G}(x,\phi) \text{ for some trading strategy } (x,\phi) \}
		\]
		Then, letting $\mathbb{A}$ be given as \[
			\mathbb{A} = \{ X \in \R^k \given X \geq 0, X \neq 0 \}
		\]
		Then we have a definition of arbitrage:
		\[
			\text{no arbitrage } \iff \mathbb{W} \cap \mathbb{A} = \emptyset
		\]
\end{defn}

\begin{defn}[Set of risk neutral measures]
	Now, consider the orthogonal component $\mathbb{W}^\perp$, defined as \[
		\mathbb{W}^\perp = \{ Y \in \R^k \given \langle X,Y \rangle = 0 \text{ for all } X \in \R^k \},
	\] the set of vectors in $\R^k$ perpendicular to all elements of $\mathbb{W}$.  Furthermore, defining $\mathcal{P}^+$ as\[
		\mathcal{P}^+  = \{ X \in \R^k \given \sum_{i = 1}^k X_i = 1, X_i > 0 \}
	\] 
\end{defn}

	Then we have the following theorem.
	\begin{thm}A measure $\rnm$ is a risk neutral measure on $\Omega$ if and only if $\rnm \in \mathcal{P}^+ \cap \mathbb{W}^\perp$.
	\end{thm}

	We denote the set of risk neutral measures $\mathbb{M} = \mathcal{P}^+ \cap \mathbb{W}^\perp$.

\begin{defn}[Contingent claim]
	A \textbf{contingent claim} in our general single period market model is a random variable $X$ on $\Omega$ representing a payoff at time $t = 1$.
\end{defn}

\begin{prop}
	Let $X$ be a contingent claim in our general single period market model, and let $(x,\phi)$ be a hedging strategy for $X$, so that $V_1(x,\phi) = X$ then the only price of $X$ which complies with the no arbitrage principle is $x = V_0(x,\phi)$.
\end{prop}

\begin{defn}[Attainable contingent claim]
	A contingent claim is \textbf{attainable} if there exists a trading strategy $(x,\phi)$ which replicates $X$, so that $V_1(x,\phi)$.
\end{defn}

\begin{thm}
	Let $X$ be an attainable contingent claim and $\rnm$ be an arbitrary risk neutral measure.  Then the price $x$ of $X$ at time $t = 0$ can be computed by the formula \[
		x = \expp{\frac{1}{1+r} X}
	\]
\end{thm}
\begin{cor}
	This theorem tells us that in particular, for any risk neutral measure in our model, we get the same value when taking the expectation above.
\end{cor}

\begin{defn}
	We say that a price $x$ for the contingent claim $X$ \textbf{complies with the no arbitrage principle} if the extended model, which consists of the original assets $S^1,\dots,S^n$ and an additional asset $S^{n+1}$ which satisfies $S^{n+1}_0 = x$ and $S^{n+1}_1 = X$ is arbitrage free.
\end{defn}

The following proposition shows that when using the risk neutral measure to price a contingent claim, one obtains a price which complies with the no-arbitrage principle.

\begin{prop}
	Let $X$ be a possibly unattainable contingent claim and $\rnm$ any risk neutral measure for our general single period market model.  Then \[
		x = \expp{\frac{1}{1+r}X}
	\]
	defines a price for the contingent claim at time $t=0$ which complies with the no-arbitrage principle.
\end{prop}

\begin{defn}[Complete market]
A financial market is called \textbf{complete} if for any contingent claim $X$ there exists a replicating strategy $(x,\phi)$.  A model which is not complete is called \textbf{incomplete.}
\end{defn}

\begin{prop}
	Assume a general single period market model consisting of stocks $S^1, \dots, S^n$
 and a money market account modelled on the state space $\Omega = \{ \omega_1,\dots,\omega_k $ is arbitrage free.  Then this model is complete if and only if the $k \times (n+1)$ matrix $A$ given by \[
 	A = \begin{pmatrix}
 		1 +r & S^1_1 (\omega_1) & \ldots & S^n_1(\omega_1) \\
 		1 +r & S^1_1 (\omega_2) & \ldots & S^n_1(\omega_2) \\
		\vdots & \vdots 		& \ddots & \vdots  \\
 		1 +r & S^1_1 (\omega_k) & \ldots & S^n_1(\omega_k) \\

 	\end{pmatrix}
 \] has full rank, that is, $\text{rank}(A) = k$.
\end{prop}

\begin{proof}
	First, a matrix $A$ has full rank if and only if for every $X \in \R^k$, the equation $AZ = X$ has a solution $Z \in \R^{n+1}$.

	Secondly, we have \[
		 \begin{pmatrix}
	 		1 +r & S^1_1 (\omega_1) & \ldots & S^n_1(\omega_1) \\
	 		1 +r & S^1_1 (\omega_2) & \ldots & S^n_1(\omega_2) \\
			\vdots & \vdots 		& \ddots & \vdots  \\
	 		1 +r & S^1_1 (\omega_k) & \ldots & S^n_1(\omega_k) \\

	 	\end{pmatrix} 
		\begin{pmatrix}
			x - \sumni \phi^i S_0^i \\
			\phi^1 \\
			\vdots \\
			\phi^n
		\end{pmatrix}
		= 
		\begin{pmatrix}
			V_1(x,\phi)(\omega_1) \\
			V_1(x,\phi)(\omega_2) \\
			\vdots \\
			V_1(x,\phi)(\omega_k) \\

		\end{pmatrix}
	\]
	This shows that computing a replicating strategy for a contingent claim $X$ is the same as to solve the equation $AZ = X$, and the proposition follows
\end{proof}

\begin{prop}
	A contingent claim $X$ is attainable, if and only if $\expp{\frac{1}{1+r}X}$ takes the same value for all $\rnm \in \mathbb{M}$.
\end{prop}

\begin{thm}
	Under the assumption that the model is arbitrage free, it is complete, if and only if $\mathbb{M}$ consists of only one element - i.e., there is a unique risk measure.
\end{thm}
% section single_period_market_models (end)

\subsection{Single Period Investment} % (fold)
\label{sec:single_period_investment}

\begin{defn}A continuously differentialble function $u: \R^+ \rightarrow \R$ is called a risk averse utility function if it has the following two properties:
	\begin{itemize}
		\item $u$ is strictly increasing - that is, $u'(x) > 0$ for all $x \in \R^+$
		\item $u$ is strictly concave - that is, $u(\lambda x + (1-\lambda)y) > \lambda u(x) + (1-\lambda)u(y)$
	\end{itemize}
		In the case that $u''(x)$ exists, the second condition in the above definition is equivalent to $u''(x) < 0$ for all $x \in \R^+$.  Sometimes one assumes in addition the condition:
		\begin{itemize}
			\item $\lim_{x \rightarrow 0} u'(x) = +\infty$ and $\lim_{x \rightarrow \infty} u'(x) = 0$
		\end{itemize}
\end{defn}

\begin{exmp}[Utility functions]
	The following are all risk averse utility functions:
	\begin{enumerate}
		\item Logarithmic utility: $u(x) = \log(x)$
		\item Exponential utility: $u(x) = 1 - e^{-\lambda x}$
		\item Power utility: $u(x) = \frac{1}{1- \gamma}x^{1-\gamma}$ with $\gamma > 0, \gamma \neq 1$
		\item Square root utility: $u(x) = \sqrt{x}$
	\end{enumerate}
\end{exmp}

\begin{prop}[Principle of expected utility]
	We assume the following axiom of agents behaviour - that of maximising expected utility.\[
		X \text{ is preferred to } Y \iff \expc{u(X)} \geq \expc{u(Y)}
	\]
\end{prop}

From Jensen's inequality, we have that for every risk averse utility function $u$ and risky payoff $X$, \[
	\expc{u(X)} \leq u(\expc{X})
\]

\begin{defn}[Certainty equivalent price]
	The certain value $X_0 \in \R$ that makes an investor indifferent between $X_0$ and a risky payoff $X$ is called the \textbf{certaintly equivalent price} of $X$, that is, \[
		u(X_0) = \expc{u(X)}
	\] or equivalently, \[
		X_0 = u^{-1} \left(\expc{u(X)}\right)
	\]
\end{defn}

\begin{lem}
	The certainty equivalent price of a risky payoff is invariant under a positive linear transformation of the utility function $u(x)$. 
\end{lem}

\begin{defn}[Risk premium]
	We have that $X_0 < \expc{X}$, and the difference between the two is called the \textbf{risk premium} $\rho$, that is, \[
		\rho = \expc{X} - u^{-1} \left( \expc{u(X)} \right) 
	\]
We can write the equation above as \[
	u(\expc{X} - \rho) = \expc{u(X)}
\]
\end{defn}

\begin{defn}[Measures of risk aversion]
	We define the following risk aversion coefficients, as a measure of how risk averse the investor is.
	The \textbf{absolute risk aversion} $\rho_{abs}$, given by \[
		\rho_{abs} = - \frac{u''(x)}{u'(x)}
	\]   and the \textbf{relative risk aversion} $\rho_{rel}$, given by \[
		\rho_{rel} = - \frac{ x u''(x)}{u'(x)}
	\]  
\end{defn}

We now seek to find the optimal investment in a market, which can be translated as finding a trading strategy $\tst$ such that $\expc{u(V_1 \tst)}$ achieves an optimal value.

\begin{defn}
	A trading strategy $(x,\phi^\star)$ is a solution to the optimal portfolio problem with initial investment $x$ and utility function $u$, if \[
		\expc{u(V_1(x,\phi^\star))} = \max_{\phi} \expc{u(V_1 \tst)}
	\]
\end{defn}

\begin{prop}
	If there exists a solution to the optimal portfolio problem, then there can not exist an arbitrage in the market.
\end{prop}

\begin{prop}
	Let $\tst$ be a solution to the optimal portfolio problem with intial wealth $x$ and utility function $u$, then the measure $\Q$ defined by \[
		\Q(\omega) = \frac{\prob{\omega} u'(V_1 \tst(\omega_i))}{\expc{u'(V_1 \tst)}}
	\]
\end{prop}

Assume that our model is complete.  In this case, there is a unique risk neutral measure which we denote by $\rnm$.

\begin{defn}
	We define the set of attainable wealths from initial investment $x > 0$ by \[
		\mathbb{W}_x = \left\{ W  \in \R^k \given \expp{\frac{1}{1+r}W} = x \right\}
	\]

	Our optimisation problem is hence: \begin{align*}
		\text{maximise } \quad &\expc{u(W)} \\
		\text{subject to } \quad &W \in \mathbb{W}_x
	\end{align*}
\end{defn}

To solve this problem, we use the Lagrange multiplier method.  To do this, consider the Lagrange function \[
	\mathcal{L}(W,\lambda) = \expc{u(W)} - \lambda\left(\expp{\frac{1}{1+r}W} - x \right)
\]

By introducing the \textbf{state price density} \[
	L(\omega) = \frac{\Q(\omega)}{\mathbb{P}(\omega)},
\] we can write the Lagrange function as \[
	\mathcal{L}(W,\lambda) = \sum_{i=1}^k \prob{\omega_i} \left[ u(W(\omega_i)) - \lambda\left(L(\omega_i) \frac{1}{1+r} W(\omega_i) - x \right)\right]
\]

Computing partial derivatives with respect to $W_i = W(\omega_i)$ and setting them equal to zero, multiplying with $\prob{\omega_i}$ and summing over $i$, we deduce that \[
	\lambda = \expc{(1+r) u'(W)}
\] and (denoting the inverse function of $u'(x)$ by $I(x)$) \[
	W(\omega) = I \left(\lambda \frac{L(\omega)}{1+r}\right)
\]

Since we have \[
	\expp{\frac{1}{1+r}W} = x
\] and substituting the expression from the above equation into the last equation, we obtain\[
	\expp{\frac{1}{1+r} I \left(\lambda \frac{L}{1+r}\right)} = x
\]
% subsection single_period_investment (end)
%!TEX root = /Users/ajtulloch/Documents/University 2010/MATH 3975 - Financial Mathematics/Exam Notes/MATH 3975 - Master Notes.tex

\renewcommand{\F}{\mathcal{F}}

\newcommand{\ott}{0 \leq t \leq T}

\section{Multi period Market Models} % (fold)
\label{sec:multi_period_market_models}
\subsection{The general model} % (fold)
\label{sub:the_general_model}

\begin{defn}[Specification of the general model]
	The two most important new features of multi period market models are:
	\begin{itemize}
		\item Agents can buy and sell assets not only at the beginning of the trading period, but at any time $t$ out of a discrete set of trading times $t \in \{0,1,2,\dots,T \}$.
		\item Agents can gather information over time, since they can observe prices.  Hence, they can make their investment decisions at time $t=1$ dependent on the prices of the asset at time $t=1$.
	\end{itemize}
\end{defn}

Throughout, we assume we are working on a finite state space $\Omega$ on which there is defined a probability measure $\mathbb{P}$.

\begin{defn}[$\sigma$-algebra]
	A collection $\mathcal{F}$ of subsets of the state space $\Omega$ is called a $\sigma$-algebra if the following conditions hold:
	\begin{itemize}
		\item $\Omega \in \mathcal{F}$
		\item If $F \in \mathcal{F}$, then $F^c \in \F$
		\item If $F_i \in \F$ for $i \in \mathbb{N}$, then $\bigcup_{i=1}^\infty F_i \in \F$.
	\end{itemize}
\end{defn}

\begin{defn}[Partition of a $\sigma$-algebra]
	Let $I$ denote some index set.  A \textbf{partition} of a $\sigma$-algebra $\F$ is a collection of sets $\emptyset \neq A_i \in \F$ for $i \in I$, such that 
	\begin{itemize}
		\item Every set $F \in \F$ can be written as a union of some of the $A_i$.
		\item The sets $A_i$ are pairwise disjoint.
	\end{itemize}
\end{defn}

\begin{defn}
	A random variable $X : \Omega \rightarrow \R$ is called $\F$-\textbf{measurable}, if for every closed interval $[a,b] \subset \R$, the preimage under $X$ belongs to $\F$, that is,
	\[
		X^{-1}([a,b]) \in \F
	\]
\end{defn}

\begin{prop}Let $X : \Omega \rightarrow \R$ be a random variable and $(A_i)$ a partition of the $\sigma$-algebra $\F$, then $X$ is $\F$-measurable if and only if $X$ is constant on each of the sets of the partition, that is, there exist $c_j \in \R$ for all $j \in I$ such that \[
	X(\omega) = c_j \text{ for all } \omega \in A_j
\]	
\end{prop}

\begin{defn}
	A sequence $(\F_t)_{0 \leq t \leq T}$ of $\sigma$-algebras on $\Omega$ is called a \textbf{filtration} if $\F_s \subset \F_t$ whenever $s < t$.
\end{defn}

\begin{defn}
	A family $(X_t)$ with $0 \leq t \leq T$ consisting of random variables, is called a \textbf{stochastic process}.  If $(\F_t)_{0 \leq t \leq T}$ is a filtration, the stochastic process $(X_t)$ is called $(\F_t)$-\textbf{adapted} if for all $t$ we have that $X_t$ is $\F_t$-measurable.
\end{defn}

\begin{defn}Let $(X_t)_{0 \leq t \leq T}$ be a stochastic process on $(\Sigma, \F, \mathbb{P})$.  Define\[
	\F^X_t = \sigma \left(X^{-1}_u ([a,b]) \given 0 \leq u \leq t, a \leq b \right)
\]
This is the smallest $\sigma$-algebra which contains all the sets $X^{-1}_u([a,b])$ where $0 \leq u \leq t$ and $a \leq b$.  Clearly $(\F^X_s)$ is a filtration.  It follows immediately from the definition that $(X_t)$ is $(\F^X_t)$ adapted.  $(F^X_t)$ is called the filtration \textbf{generated} by the process $X$.

\end{defn}

\begin{defn}[Value process]
	The \textbf{value process} corresponding to the trading strategy $\phi = (\phi_t)_{0 \leq t \leq T}$ is the stochastic process $(V_t(\phi))_{0 \leq t \leq T}$ where \[
		V_t(\phi) = \phi^0_t B_t + \sumni \phi^i_t S_t^i
	\]
\end{defn}

\begin{defn}
	A trading strategy $\phi = (\phi_t)$ is called \textbf{self financing} if for all $t = 0,\dot,T-1$,\[
		\phi^0_t B_{t+1} + \sumni \phi^i_t S^i_{t+1} = \phi^0_{t+1} B_{t+1} + \sumni \phi^i_{t+1} S^i_{t+1}
	\]
\end{defn}

\begin{lem}
	For a self-financing trading strategy $\phi = (\phi_t)$, the value process can be alternatively computed via \[
		V_t(\phi) = \phi_{t-1}^0 B_t + \sumni \phi^i_{t-1} S^i_t
	\]
\end{lem}

\begin{defn}[General multi period market model]
		A general multi period market model is given by the following data.
		\begin{itemize}
			\item A probability space $(\Sigma, \F, \mathbb{P})$ together with a filtration $(\F_t)_{0 \leq t \leq T}$ of $\F$.
			\item A money market account $(B_t)$ which evolvs according to $B_t = (1 +r)^t$.
			\item A number of financial assets $(S^1_t), \dots, (S^n_t)$ which are assumed to be $(\F_t)$-adapted stochastic processes.  
			\item A set $\mathcal{T}$ of self financing and $(\F_t)$-adapted trading strategies. 
		\end{itemize}
\end{defn}

\begin{defn}
		Assume we are given a general multi period market model as described above.  The \textbf{increment process} $(\Delta S^i_{t})$ is defined as \[
			\Delta S_t^i = S^i_t - S^i_{t-1}
		\]
		and \[
			\Delta B_t = B_t - B_{t-1} =  r B_{t-1} 
		\]
\end{defn}

\begin{defn}[Gains process]
	Given a trading strategy $\phi$, the corresponding \textbf{gains process} $(G_t(\phi))_{\ott}$ is given by \[
		G_t(\phi) = \sum_{s=0}^{t-1} \phi_s^0 \Delta B_{s+1} + \sumni \sum_{s=0}^{t-1} \phi_s^i \Delta S^i_{s+1}
	\]
\end{defn}

\begin{prop}
	An adapted trading strategy $\phi = (\phi)_{\ott}$ is self financing, if and only if any of the two equivalent statements hold 
	\begin{itemize}
		\item $V_t(\phi) = V_0(\phi) + G_t(\phi)$
		\item $\hat{V}_t(\phi) = \hat{V}_0(\phi) + \hat{G}_t(\phi)$
	\end{itemize}
	for all $\ott$.  Here $\hat{V}_t$ and $\hat{G}_t$ denote the discounted value and gains process, as defined in the following definition.
\end{prop}

\begin{defn}[Discount processes]
	The discounted prices are given by \[
		\hat{S}^i_t = \frac{S^i_t}{B_t}
	\] and discounted gains, discounted value process, and discounted gains process are all defined analogously.  
\end{defn}

% subsection the_general_model (end)

\subsection{Properties of the general multi period market model} % (fold)
\label{sub:properties_of_the_general_multi_period_market_model}
Here, we redefine the general concepts of financial mathematics, such as arbitrage, hedging, in the context of a multi period market model.

\begin{defn}[Arbitrage]
	A (self-financing) trading strategy $\phi = (\phi)_{\ott}$ is called an arbitrage if \begin{itemize}
		\item $V_0(\phi) = 0$
		\item $V_T(\phi) \geq 0$
		\item $\expc{V_T(\phi)} > 0$
	\end{itemize}
\end{defn}

\begin{defn}[Contingent claim] A \textbf{contingent claim} in a multi period market model is an $\F_T$-measurable random variable $X$ on $\Omega$ representing a payoff at terminal time $T$.  A \textbf{hedging strategy} for $X$ in our model is a trading strategy $\phi \in \mathcal{T}$ such that \[
	V_T(\phi) = X,
\] that is, the terminal value of the trading strategy is equal to the payoff of the contingent claim.
\end{defn}

\begin{prop}Let $X$ be a contingent claim in a multi period market model, and let $\phi \in \mathcal{T}$ be a hedging strategy for $X$, then the only price of $X$ at time $t$ which complies with the no arbitrage principle is $V_t(\phi)$.  In particular, the price at the beginning of the trading period at time $t = 0$ is the total initial investment in the hedge.
\end{prop}

\begin{defn}[Attainable contingent claim] A contingent claim $X$ is called \textbf{attainableat} in $\mathcal{T}$, if there exists a trading strategy $\phi \in \mathcal{T}$ which replicates $X$, that is, $V_T(\phi) = X$.
\end{defn}

\begin{defn}[Complete market] 
	A general multi period market model is called \textbf{complete}, if and only if for any contingent claim $X$ there exists a replicating strategy $\phi$.  A model which is not complete is called \textbf{incomplete}.	
\end{defn}

\begin{defn}[Conditional expectation]
	Let $(\Omega, \F, \mathbb{P})$ be a finite probability space and $X$ an $\F$-measurable random variable.  Assume that $\mathcal{G}$ is a $\sigma$-algebra which is contained in $\F$.  Denoting the unique partition of $\mathcal{G}$ with $(A_i)_{i \in I}$, the \textbf{conditional expectation} $\expc{X \given \mathcal{G}}$ of $X$ with respect to $\mathcal{G}$ is defined as the random variable which satisfies \[
		\expc{X \given \mathcal{G}}(\omega) = \sum_{x} x \prob{X = x \given A_i}
	\] whenever $\omega \in A_i$.

	We then have the following identity: \[
		\int_G X \, d\mathbb{P} = \int_G \expc{X \given \mathcal{G}} \, d \mathbb{P}
	\] for any $G \in \mathcal{G}$.
\end{defn}

\begin{prop}
Let $(\Omega, \F, \mathbb{P})$ be a finite probability space and $X$ an $\F$-measurable random variable.  Let $\mathcal{G},\mathcal{G}_1,\mathcal{G}_2$ be sub-$\sigma$-algebras.  Assume furthermore that $\mathcal{G}_2 \subset \mathcal{G}_1$.  Then
\begin{itemize}
	\item \textbf{Tower property.}\[
		\expc{X \given \mathcal{G}_2} = \expc{\expc{X \given \mathcal{G}_1} \given \mathcal{G}_2}
	\]
	\item \textbf{Taking out what is known.}  If $Y:\Omega \rightarrow \R$ is $\mathcal{G}$-measurable, then \[
		\expc{YX \given \mathcal{G}} = Y \expc{X \given \mathcal{G}}
	\]
	\item If $\mathcal{G} = \{ \emptyset, \Omega \}$ is the trivial $\sigma$-algebra, then \[
		\expc{X \given \mathcal{G}} = \expc{X}
	\]
\end{itemize}
\end{prop}

\begin{defn}[Risk neutral measure]
	A measure $\rnm$ on $\Omega$ is called a \textbf{risk neutral measure} for a general multi period market model if 
	\begin{itemize}
		\item $\rnm(\omega) > 0$ for all $\omega \in \Omega$
		\item $\expp{\Delta \hat{S}^i_t \given \F_{t-1}} = 0$ for $i = 1,\dots,n$ and for all $1 \leq t \leq T$.
	\end{itemize}
\end{defn}

An alternative formulation of the second condition in the previous definition is \[
	\expp{\frac{1}{1+r}S^i_{t+1} \given \F_t} = S^i_t
\]

\begin{defn}[Martingale]
	A $\F_t$-adapted process $(X_t)$ on a probability space $(\Omega, \F, \mathbb{P})$ is called a \textbf{martingale} if for all $s < t$, \[
		\expc{X_t \given \F_s} = X_s
	\]
\end{defn}

\begin{lem}
	Let $\rnm$ be a risk neutral measure.  Then the discounted stock prices $(\hat{S}^i_t)$ for $i = 1,\dots,n$ are martingales under $\rnm$.
\end{lem}

\begin{prop}Let $\phi \in \mathcal{T}$ be a trading strategy.  Then the discounted value process $(\hat{V}_t(\phi))$ and the discounted gains process $\hat{G}_t(\phi)$ are martingales under any risk neutral measure $\rnm$.

\end{prop}

\begin{thm}[Fundamental Theorem of Asset Pricing]
	Given a general multi period market model, if there is a risk neutral measure, then there are no arbitrage strategies $\phi \in \mathcal{T}$.  Conversely, if there are no arbitrages among self financing and adapted trading strategies, then there exists a risk neutral measure.
\end{thm}

\begin{defn}We say that an adapted stochastic process $(X_t)$ is a price process for the contingent claim $X$ which \textbf{complies with the no arbitrage principle}, if there is no adapted and self financing arbitrage strategy in the extended model, which consists of the original stocks $(S_t^1),\dots,(S_t^n)$ and an additional asset given by $S^{n+1}_t = X_t$ for $0 \leq t \leq T-1$ and $S^{n+1}_T = X$. 
\end{defn}

\begin{prop}
	Let $X$ be a possibly unattainable contingent claim and $\rnm$ a risk neutral measure for a general multi period market model.  Then \[
		X_t = B_t \expp{\frac{X}{B_T} \given \F_t}
	\] defines a price for the contingent claim consistent with the no arbitrage principle.
\end{prop}

\begin{thm}
	Under the assumption that a general multi period market model is arbitrage free, it is complete, if and only if there is a unique risk neutral measure.
\end{thm}
% subsection properties_of_the_general_multi_period_market_model (end)

% section multi_period_market_models (end)
\input{MATH3975-Chap4}

\end{document}

### 2011-12-12 MSH2 - Probability Theory/MSH2 Lecture Notes.tex
Initial Commit

% Created by Andrew Tulloch

%!TEX TS-program = xelatex
%!TEX encoding = UTF-8 Unicode

\documentclass[10pt, oneside, reqno]{amsart}
\usepackage{geometry, setspace, graphicx, enumerate}
\onehalfspacing                 
\usepackage{fontspec,xltxtra,xunicode}
\defaultfontfeatures{Mapping=tex-text}

	% AMS Theorems
	\theoremstyle{plain}% default 
	\newtheorem{thm}{Theorem}[section] 
	\newtheorem{lem}[thm]{Lemma} 
	\newtheorem{prop}[thm]{Proposition} 
	\newtheorem{cor}[thm]{Corollary}

\newcommand{\res}[2]{\text{Res}(#1,#2)}
	\theoremstyle{definition} 
		\newtheorem{defn}[thm]{Definition}
		\newtheorem{conj}[thm]{Conjecture}
		\newtheorem{exmp}[thm]{Example}
		% \newtheorem{exer}[thm][Exercise]

	\theoremstyle{remark} 
		\newtheorem*{rem}{Remark} 
		\newtheorem*{note}{Note} 
		\newtheorem{case}{Case}

		\newcommand{\expc}[1]{\mathbb{E}\left[#1\right]}
		\newcommand{\var}{\text{Var}}
		\newcommand{\cov}[1]{\text{Cov}\left(#1\right)}
		\newcommand{\prob}[1]{\mathbb{P}(#1)}
		\newcommand{\given}{ \, | \,}
		\newcommand{\us}{0 \leq u \leq s}
		\newcommand{\ts}[1]{\{ #1 \}}

\renewcommand{\phi}{\varphi}
\newcommand{\sigf}{\mathcal{F}}

\newcommand{\dzz}{\, dz}
\newcommand{\bigo}[1]{\mathcal{O}(#1)}

\newcommand{\al}{\alpha}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\N}{\mathbb{N}}

\newcommand{\I}{\mathbb{I}}

\renewcommand{\P}{\mathbb{P}}

\newcommand{\F}{\mathbb{F}}
\newcommand{\Ga}{\mathbb{G}}

\newcommand{\aut}[1]{\text{Aut}{(#1)}}

\newcommand{\gener}[1]{\langle #1 \rangle}
\newcommand{\charr}[1]{\text{char}(#1)}
\newcommand{\nth}{n\textsuperscript{th}}

% \newcommand{\limsup}{\text{limsup}}

\newcommand{\tworow}[2]{\genfrac{}{}{0pt}{}{#1}{#2}}
\newcommand{\xdeg}[2]{[#1 : #2]}
\newcommand{\gal}[2]{\text{Gal}(#1/#2)}
\newcommand{\minpoly}[2]{m_{#1, #2}(x)}

\newcommand{\mapping}[5]{\begin{align*}
	#1 : \quad     #2 &\rightarrow #3 \\
			#4  &\mapsto #5
\end{align*}	
}

\def\cip{\,{\buildrel p \over \rightarrow}\,} 
\def\cid{\,{\buildrel \!d \over \rightarrow}\,} 
\def\cas{\,{\buildrel a.s. \over \rightarrow}\,}

\def\clp{\,{\buildrel L^p \over \rightarrow}\,} 
\def\clone{\,{\buildrel L^1 \over \rightarrow}\,}

\def\eqd{\,{\buildrel \!d \over =}\,} 
\def\eqas{\,{\buildrel a.s. \over =}\,}
\newcommand{\sigg}{\mathcal{G}}		
\newcommand{\indic}[1]{\mathbf{1}_{\{ #1 \}} }
\newcommand{\sumni}{\sum_{i=1}^n}
\newcommand{\sumnj}{\sum_{j=1}^n}
\newcommand{\sumnk}{\sum_{k=1}^n}

\title{MSH2 - Probability Theory}								% Document Title
\author{Andrew Tulloch}
%\date{}                                           % Activate to display a given date or no date

\begin{document}
\maketitle

\section{Lecture 1 - Thursday 3 March} % (fold)
\label{sec:lecture_1_thursday_3_march}

\begin{defn}[$\sigma$-field]
	Let $\Omega$ be a non-empty set.  Let $\mathcal{F}$ be a collection of subsets of $\Omega$.  We call $\mathcal{F}$ a $\sigma$-field if 
	\begin{itemize}
		\item $\emptyset \in \mathcal{F}$,
		\item $A \in \mathcal{F} \Rightarrow A^c \in \mathcal{F}$,
		\item $A, B \in \sigf \Rightarrow A \cup B \in \sigf$
		\item If $(A_i) \in \sigf$, then $\bigcup_{j = 1}^\infty A_j \in \sigf$.
	\end{itemize}
\end{defn}

\begin{defn}[Probability measure]
	Let $\P$ be a function on $\sigf$ satisfying
	\begin{itemize}
		\item If $A \in \sigf$ then $\P(A) \geq 0$,
		\item $P(\Omega) = 1$,
		\item If $(A_j) \in \sigf$ and $A_i \cap A_j = \emptyset$, then $\P \left( \bigcup_{j=1}^\infty A_j \right) = \sum_{j=1}^\infty \P(A_j)$.
	\end{itemize}
	Then we call $\P$ a \textbf{probability measure} on $\sigf$.
\end{defn}

\begin{defn}[$\sigma$-field generated by a set]
	If $A$ is a class of sets, then $\sigma(A)$ is the smallest $\sigma$-field that contains $A$.
\end{defn}

\begin{exmp}
	For a set $B$, $\sigma(B) = \{ \emptyset, \Omega, B, B^c \}$.
\end{exmp}

\begin{defn}[Borel $\sigma$-field] Let $\mathcal{B}$ be the class of all \textbf{finite} unions of intervals of the form $(a, b]$ on $\R$.  The $\sigma$-field $\sigma(\mathcal{B})$ is called the \textbf{Borel $\sigma$-field}.

	Note that $\mathcal{B}$ itself is not a $\sigma$-field - consider $\bigcup_{j=1}^\infty (0, \frac{1}{2} - \frac{1}{j}] = (0, \frac{1}{2}) \notin \mathcal{B}$.
\end{defn}

\subsection{Constructing extensions of functions to form probability measures} % (fold)
\label{sub:constructing_extensions_of_functions_to_form_probability_measures}

% subsection constructing_extensions_of_functions_to_form_probability_measures (end)

\begin{lem}[Continuity property]
	Let $\mathcal{A}$ be a field of subsets of $\Omega$.  Assume $\emptyset \in \mathcal{A}$ and that $\mathcal{A}$ is closed under complements and finite unions.

	If $A_j \in \sigf$ and $A_{j+1} \subset A_j$ with $\bigcap_{j=1}^\infty A_j = \emptyset$, then $\lim_{j \rightarrow \infty} \P(A_j) = 0$.
\end{lem}

\begin{thm}
	Let $\sigma( \mathcal{A})$ be the $\sigma$-field generated by $\mathcal{A}$.  If the continuity property holds, then there is a \textbf{unique} probability measure on $\sigma(\mathcal{A})$ which is an extension of $\P$, i.e. the measures agree on all elements of $\mathcal{A}$.  
\end{thm}

\begin{defn}[Limits of sets] Let $(\Omega, \sigf, \P)$ be a probability space, and assume $(A_i) \in \sigf$.  Then define $\limsup_{m \rightarrow \infty} A_n$ as \[
	\limsup_{n \rightarrow \infty} A_n = \bigcap_{n = 1}^\infty \bigcup_{m \geq n} A_m = \overline{\lim} A_n
\] 
An element $\omega \in \overline{\lim} A_n$ if and only if $\omega \in A_m$ for some $m \geq n$ for all $n$ - that is, $\omega$ is in infinitely many of the sets $A_m$.

Similarly, define $\liminf_{m \rightarrow \infty} A_n$ as \[
	\liminf_{n \rightarrow \infty} A_n = \bigcup_{n = 1}^\infty \bigcap_{m \geq n} A_m = \underline{\lim} A_n
\] 
An element $\omega \in \underline{\lim} A_n$ if and only if $\omega$ is in all but a finite number of sets $A_m$.

Clearly, \[
	\underline{\lim}A_n \subseteq \overline{\lim} A_n
\]
If $\underline{\lim}A_n$ and $\overline{\lim}A_n$ coincide we write it as $\lim A_n$.
\end{defn}

\begin{lem}\label{lem:problim}
	Assume the continuity property holds.  If $A_n \downarrow A$ then $\P(A_n) \downarrow \P(A)$, and if $A_n \uparrow A$ then $\P(A_n) \uparrow \P(A)$.

\end{lem}

\begin{proof}
	If $A_n \downarrow A$, then $A_n \supseteq A_{n+1} \dots$ and $\bigcap_{n=1}^\infty A_n = A$.  We can write $A_n = (A_n - A) \cup A$.  Then we have 
	\begin{align*}
		\P(A_n) = \P(A_n - A) + \P(A) \\
		\P(A_n) \geq \P(A)
	\end{align*}
	By the continuity property, $\P(A_n - A) \rightarrow 0$, and so $\P(A_n) \downarrow \P(A)$.
\end{proof}

% section lecture_1_thursday_3_march (end)

\section{Lecture 2 - Thursday 3 March} % (fold)
\label{sec:lecture_2_thursday_3_march}
\begin{thm}
	\[
		\P( \underline{\lim} A_n ) \leq \underline{\lim} \P(A_n) \leq \overline{\lim} \P(A_n) \leq \P( \overline{\lim} A_n )	
	\]
\end{thm}
\begin{proof}
	We know $A_n \downarrow \underline{\lim} A_n$, and so from Lemma \ref{lem:problim} we have that $a$.
\end{proof}

\begin{defn}[Measurable function]
	Let $(\Omega, \sigf, \P)$ be a probability space.  Let $X : \Omega \rightarrow \R$ be real valued function on $\Omega$.  Then $X$ is \textbf{measurable} with respect to $\sigf$ if $X^{-1}(B)$ is an element of $\sigf$ for every $B$ in the Borel $\sigma$-field of $\R$. 
\end{defn}

\begin{defn}[Random variable]
	A random variable is a measurable function from $\Omega$ to $\R$.  
\end{defn}

\begin{defn}[Expectation]
	If $\int_\Omega | X(\omega) | \, d\P < \infty$ then we can define $\E(X) = \int_\Omega X(\omega) \, d \P$
\end{defn}

\begin{defn}[Distribution]
	$X$ induces a probability measure $\P_X$ on $\R$\[
		\P_X(B) = \P(X^{-1}(S)), S \in \mathcal{B}
	\]
	$P_X$ is called the \textbf{distribution} of $X$.  $(\R, \mathcal{B}, \P_X)$ is a probability space.
	The distribution function $F_X(x) = \P( \{ \omega : X(\omega) \leq x \} ) = \P_X( (-\infty, x] )$.  
	We have $\E(X) = \int_\R x \, dP_X(x) = \int_\R x \, dF_X(x)$.
\end{defn}

\subsection{Key results from Measure Theory} % (fold)
\label{sub:key_results_from_measure_theory}

\begin{thm}[Monotone convergence theorem]
	If $0 \leq X_n \uparrow X a.s$ then $0 \leq \E(X_n) \uparrow \E(X)$ where $\E(X)$ is infinite if $\E(X_n) \uparrow \infty$.
\end{thm}

\begin{thm}[Dominated convergence theorem]
	If $\lim X_n = X a.s.$ and $|X_n | \leq Y$ for all $n \geq 1$, with $\E(|Y|) < \infty$ then $\lim \E(X_n) = \E(X)$.
\end{thm}

\begin{thm}[Fatau's Lemma]
	If $X_n \geq Y$ for all $n$ with $\E(|Y|) < \infty$ then \[
		\E( \liminf X_n) \leq \liminf \E(X_n)
	\]
\end{thm}

% subsection key_results_from_measure_theory (end)

\begin{thm}[Composition]
	Let $(\Omega, \sigf, \P)$ and $(\Omega', \sigf')$ be spaces.  Let $\Phi: \Sigma \rightarrow \Sigma'$ be measurable.  Define $\P_\Phi$ on $\sigf$ by $\P_\Phi(M) = \P(\Phi^{-1}(M))$.  Let $X'$ be a measurable function from $\Sigma'$ to $\R$.  Then $X(\omega) = X'(\Phi(\omega))$ is a measurable function.  Then we have 
	\[
		\E(X) = \int_{\Omega'} X' \, d\P_\phi
	\]
\end{thm}

\begin{proof}
	Suppose $X'$ is an indictor function for $A \in \sigf'$.  Then \[
		\int_{\Omega'} X' \, d\P_\phi = \int_A \, d\P_\phi = \P_\phi(A) = \P(\phi^{-1}(A)) = \E(X)
	\]
	So the result is true for simple functions.

	Now, suppose $X' \geq 0$.  Then there exists a pointwise increasing sequence of simple functions $X'_n$ such that $X_n' \rightarrow X'$.  By the monotone convergence theorem, we know \[
		\lim_{n \rightarrow \infty} \int_{\Omega'} X'_n \, d\P_\phi = \int_{\Omega'} X' \, d\P_\phi
	\] 
	But $X_n(\omega) = X_n'(\Phi(\omega) )$ are also simple functions increasing to $X$.  Hence, we know that $\lim_{n \rightarrow \infty} \E(X_n) = \E(X)$.
\end{proof}

% section lecture_2_thursday_3_march (end)

\section{Lecture 3 - Thursday 10 March} % (fold)
\label{sec:lecture_3_thursday_10_march}
\begin{thm}[Jensen's inequality]
	Let $\phi(x)$ be a convex function on $\R$.  Let $X$ be a random variable.  Assume $\E(X) < \infty$, $\E(\phi(X)) < \infty$.  Then \[
		\phi(E(X)) \leq \E(\phi(X))
	\]
\end{thm}

\begin{thm}[H\"older's inequality]
	Let $1 < p < \infty$ and $\frac{1}{p} + \frac{1}{q} = 1$. Then we have\[
		| \E(XY) | \leq \E(|XY|) \leq \left( \E(|X|^p) \right)^{1/p} \left( \E(|Y|^q) \right)^{1/q}
	\] 
	If $p = q = 2$ we obtain the Cauchy-Swartz inequality $\E(|XY|) \leq \left( \E(|X|^2) \right)^{1/2} \left( \E(|Y|^2) \right)^{1/2}$.

	If $Y = 1$ then $\E(|X|) \leq ( \E(|X|^p))^{1/p}$.
\end{thm}
\begin{proof}
	Let $W$ be a random variable taking values $a_1$ with probability $1/p$, $a_2$ with probability $1/q$, with $1/p + 1/q = 1$.  Applying Jensen's inequality with $\phi(x) = -\log(x)$ gives 
	\begin{align*}
		\E( - \log W) \geq -\log \E(W) \\
		\frac{1}{p} (\log a_1) + \frac{1}{q} ( -\log a_2) \geq -\log( \frac{a_1}{p} + \frac{a_2}{q}) \\
		-\log(a_1^{1/p} \cdot a_2^{1/q}) \geq -\log( \frac{a_1}{p} + \frac{a_2}{q})  \\
		a_1^{1/p} \cdot a_2^{1/q} \leq \frac{a_1}{p} + \frac{a_2}{q}
	\end{align*}
	Where the inequality is trivial if $a_1$ or $a_2$ is zero.

	Setting $a_1 = |x|^p$ and $a_2 = |y|^q$, we obtain \[
		|xy| \leq \frac{|x|^p}{p} + \frac{|y|^q}{q}.
	\]

	Let $x = \frac{X}{(\E(|X|^p))^1/p}$ and $y = \frac{Y}{(\E(|Y|^q))^{1/q}}$
or take expectations across the inequality, we obtain \[
	\E(|XY|) \leq \left( \E(|X|^p) \right)^{1/p} \left( \E(|Y|^q) \right)^{1/q}
\]
\end{proof}

\begin{exmp}
	If $1 < r < r'$ then $\frac{r'}{r} > 1$.  Then \[
		\E(|X|^r) \leq (\E((|X|^r)^{r'/r}))^{1/(r'/r)} = (\E(|X|^{r'}))^{r'/r}
	\]
\end{exmp}

\begin{thm}[Liapounov's inequality]
	\[
		(\E|X|^r)^{1/r} \leq ( \E(|X|^{r'}))^{1/r'}
	\]
\end{thm}
\begin{cor}
	Thus if $\E(|X|^r) < \infty$ then $X$ has all moments of lower order finite i.e. $\E(|X|^p) < \infty$ for all $ 1 \leq p \leq r$
\end{cor}

\begin{thm}[Minkowski's inequality]
	If $p \geq 1$, then \[
		(\E(|X+Y|^p))^{1/p} \leq (\E(|X|^p))^{1/p} + (\E(|Y|^p))^{1/p}
	\]
\end{thm}
\begin{proof}
	\begin{align*}
		\E(|X+Y|^p) \leq \E(|X| \cdot |X+Y|^{p-1} + \E(|Y| \cdot |X+Y|^{p-1}) \\
					= \E(|X|^p)^{1/p} ( \E(|X+Y|^{p-1})^q)^{1/q} + \E(|Y|^p)^{1/p} ( \E(|X+Y|^{p-1})^q)^{1/q}
	\end{align*}
	Let $1/p + 1/q = 1$.  Then from H\"older, 
	\begin{align*}
		\E(|X+Y|^p) \leq (\E(|X+Y|^p))^1/q \cdot ( (\E(|X|^p))^{1/p} + ( \E(|Y|^p))^{1/p}
	\end{align*}
	and so \[
		(\E(|X+Y|^p))^{1/p} \leq (\E(|X|^p))^{1/p} + (\E(|Y|^p))^{1/p}
	\]
\end{proof}

\subsection{Modes of Convergence} % (fold)
\label{sub:modes_of_convergence}
Let $(\Omega, \sigf, \P)$ be a probability space and $X_n(\omega), n \geq 1$ is a sequence of random variables.

\begin{defn}[Almost surely convergence]
	We say $X_n$ converges almost surely if \[
		\P( \{ \omega \, | \, X_n(\omega) \text{ has a limit} \} ) = 1
	\]
	We write $X_n \cas X$ where $X$ denotes the limiting random variable.
\end{defn}

\begin{defn}[Convergence in probability]
	$X_n$ converges in probability to $X$ \[
		X_n \cip X
	\] if for all $\epsilon > 0$, \[
		\P( \{ \omega \, | \, |X_n(\omega) - X(\omega) | > \epsilon \} ) \rightarrow 0
	\]
	or alternatively,\[
		\P( |X_n - X| > \epsilon) \rightarrow 0
	\]
\end{defn}

\begin{defn}[Convergence in mean]
	$X_n$ converges to $X$ in mean of order $p$ (or in $L^p$) if\[
		\E(|X_n - X|^p) \rightarrow 0
	\]
	We write $X_n \clp X$.  We note that for convergence of order $L^p$, we need $\E(|X_n|^p) < \infty$.
\end{defn}

\begin{thm}
	If $X_n \clp X$ then $X_n \cip X$ for any $p > 0$.
\end{thm}
% subsection modes_of_convergence (end)
% section lecture_3_thursday_10_march (end)

\section{Lecture 4 - Thursday 10 March} % (fold)
\label{sec:lecture_4_thursday_10_march}

\begin{lem}
	Let $C_1, C_2, \dots$ be sets in $\sigf$ and $\sum_n \P(C_n) < \infty$.  Then $\P(\overline{\lim} C_n) = 0$
\end{lem}
\begin{proof}
	Since $\overline{\lim} C_n = \bigcap_{n = 1}^\infty \bigcup_{m \geq n} C_m$, we have
	\[
		\P(\overline{\lim} C_n) \leq \P(\bigcup_{m \geq n} C_m) \leq \sum_{m \leq n} \P(C_m) \rightarrow 0 
	\]
\end{proof}

\begin{thm}\label{thm:4}
	If there exists a sequence of positive constants $ \{ \epsilon_n \}$ with $\sum_{n} \epsilon_n < \infty$ and \[
		\sum_{n} \P( |X_{n + 1} - X_n | > \epsilon_n) < \infty
	\] then $X_n$ converges almost surely to some limit $X$.
\end{thm}

\begin{proof}
	Let $A_n = \{ | X_{n+1} - X_n | > \epsilon_n$.  So from the above Lemma, $\P( \overline\lim A_n) = 0$.  We also have that $\omega \in \overline\lim A_n $ if and only if $\omega$ is in infinitely many $A_m$. For $\omega \notin \overline \lim A_n$, then there is a last set containing $\omega$.  Define $N(\omega) = n$ if $\omega \in \bigcup_{m \geq n} A_m - \bigcup_{m > n} A_m$, and zero if $\omega \in ( \bigcup{m \geq 1} A_m)^c$.

	For $\omega \notin \overline \lim A_n$, we have $\sum_{n = 1}^\infty X_{n+1}(\omega) - X_n(\omega)$ exists as $\sum_n \epsilon_n  < \infty$.  Since \[
		X_n(\omega) = X_1(\omega) + ( X_2(\omega) - X_1( \omega) ) + \dots + ( X_n( \omega) - X_{n-1}(\omega))
	\] we know $\lim X_n(\omega)$ exists - i.e. $\P( \lim X_n( \omega)) \text{ exists} ) = 1$.
\end{proof}

\begin{thm}Every sequence of random variables $X_n$ that converges almost surely converges in probability.  Conversely, if $X_n \cip X$ then there exists a subsequence $\{ X_{n_k} \}$ which converges almost surely.  
\end{thm}
\begin{proof}
	Assume $X_n \cas X$.  Let $\epsilon > 0$.  Consider $\overline \lim \P( |X_n - X | > \epsilon) \leq \P( \limsup \{ |X_n - X | > \epsilon \} )$ by a previous theorem (Theorem 2 in Lecture Notes).  We have \begin{align*}
		\limsup \{ | X_n - X | > \epsilon \} &= \{ \omega \, | \, | X_n(\omega) - X(\omega)| > \epsilon \text{ infinitely often} \} \\
											&\subseteq \{ \omega \, | \, \lim X_n (\omega) \neq X(\omega) \}
	\end{align*}
	Hence, we have \[
		\P( \overline \lim | X_n - X | > \epsilon ) \leq 1 - P (\lim X_n(\omega) = X(\omega)) = 0 \quad \text{as $X_n \cas X$}
	\] since $\lim \P( |X_n - X | > \epsilon) = 0$.

	Conversely, assume $X_n \cip X$.  Given $\epsilon > 0$, consider $\P(|X_n - X_m | > \epsilon) \leq \P(|X- X_m | > \epsilon/2 + \P( | X- X_n | > \epsilon/2 )$ (If $|X-X_n| \leq \epsilon/2$ and $|X-X_m| \leq \epsilon/2$, then $|X_n - X_m| \leq \epsilon$ by the triangle inequality). 
	Thus, $\P(|X_m - X_n | > \epsilon) \rightarrow 0$ as $m$ and $n \rightarrow 0$.  Set $n_1 = 1$ and define $n_j$ to be the smallest integer $N > n_{j-1}$ such that \[
		\P(|X_r - X_s| > 2^{-j} ) < 2^{-1} \quad \text{when $r, s > N$}
	\]
	Then apply Theorem \ref{thm:4}, and as \[
		\sum_j \P( |X_{n_{j+1}} - X_{n_j} | > 2^{-j}) < \sum 2^{-j} = 1 < \infty
	\] we know that $X_{n_j}$ converges almost surely.

\end{proof}
\begin{exmp}\label{exmp:cip_does_not_imply_as}
	We now construct an example where $X_n \cip 0$ but $X_n$ does not converge almost surely to $0$.

	Let $\Omega = [0,1], \sigf$ the Borel $\sigma$-field, and $\P$ the Lebesgue measure.  Let $\phi_{kj} = \I_{[j-1/k, j/k]}$ for $j = 1, \dots, k$ and $k = 1, 2, \dots$.

	Let $X_1 = \phi_{11}, X_2 = \phi_{21}, X_3 = \phi_{22}$, etc.  For any $p > 0$, \[
		\E(|X_n|^p) = \int X_n \, d\P = [ j_n - 1/k_n, j_n/k_n] \rightarrow 0
	\] and so $X_n \clp 0$.

	However, for each $\omega \in \Omega$ and each $k$ there are some $j$ such that $\phi_{kj}(\omega) = 1$.  Thus $X_n(\omega) = 1$ infinitely often. Similarly $X_n(\omega) = 0$ infinitely often.  Hence $X_n$ does not converge almost surely to 0.
\end{exmp}
% section lecture_4_thursday_10_march (end)

\section{Lecture 5 - Thursday 17 March} % (fold)
\label{sec:lecture_5_thursday_17_march}

Following from the previous lecture, we now modify the examples to show convergence in probability does not imply convergence in $L^p$ even when $\E(|X_n|^p) < \infty$.

From \ref{exmp:cip_does_not_imply_as}, replace $\phi_{kj}$ by $k^{1/p} \phi_{kj}$.  Then \[
	\P(|X_n| > 0) = 1/{k_n} \rightarrow 0
\] as $n \rightarrow \infty$.  Similar,y \[
	\E(|X_n|^p) = (k_n^{1/p})^p \P(X_n \neq 0) = 1
\] and so \[
	\lim_{n \rightarrow \infty} \E(|X_n|^p) = 1
\] and thus $X_n$ does not converge in $L^p$ to zero.  Thus convergence in probability does not imply convergence in $L^p$.

Next define $X_1 = \phi_{11}$, $X_n = \phi_{n1}n^{1/p}$.  Then \[
	X_n(\omega) \rightarrow 0
\] for $\omega > 0$ so $X_n \cas 0$.  We also have \[
	\E(|X_n|^p) = (n^{1/p})^p \frac{1}{n} = 1 
\] and so $X_n$ does not converge in $L^p$ to zero.

\begin{defn}[Uniform integrability]
	A sequence $\{ X_n \}$ is uniformly integrable if \[
		\lim_{y \rightarrow \infty} \sup_{n} \int_{|X_n| \geq y} |X_n | \, d\P = 0
	\]

\end{defn}

\begin{thm}[Convergence in probability and uniform integrability imply convergence in $L^p$]
	If $X_n \cip X$ and $\{ | X_n | \}$ is uniformly integrable, then $X_n \clp X$.
\end{thm}

\begin{defn}[Independence]
	Let $(\Omega, \sigf, \P)$ be a probability space.  Let $A_1, A_2, \dots, A_n \in \sigf$  The events are said to be independent if \[
		\P(A_{i_1}, \dots, A_{i_k}) = \P(A_{i_1}) \dots \P(A_{i_k})
	\] for all $1 \leq i_1 < \dots < i_k \leq n$, $k = 2, 3, \dots, n$.

	In the infinite case, let $\{A_{\alpha}, \alpha \in I \}$, $I$ an index set, is a set of independent events if each finite subset is independent.

\end{defn}

\begin{defn}[Independence of random variables]
	Let $X_{1}, \dots, X_n$ be random variables on $(\Omega, \sigf, \P)$.  $X_1, \dots, X_n$ are independent if $A_{i} = \{ X_i \in S_i \}$ are independent for every set of Borel sets, $S_i \in \mathcal{B}$.

	Alternatively, let $X$ and $Y$ be random variables.  Let $\mathcal{B}_2$ be the Borel $\sigma$-field on $\R^2$.  $Z(\omega) = (X(\omega), Y(\omega))$ is then a map form $\Omega$ to $\R^2$.  $Z$ is Borel measurable if \[
		Z^{-1}(S) \in \sigf
	\] for all $S \in \mathcal{B}_2$. 
	$\P_{X,Y}$ is the induced measure on $B_2$, and $F_{X,Y}$ is the joint distribution of $(X,Y)$.  Let \[
		F_{X,Y}(x,y) = \P_{X, Y} ( (-\infty,x], (-\infty,y] ) = \P( \{ \omega: X(\omega) \leq x, Y(\omega) \leq y \} ) 
	\] 
\end{defn}

\begin{thm}
	If $X$ and $Y$ are independent then \[
		F_{X,Y}(x,y) = F_X(x) F_Y(y)
	\]
\end{thm}

\begin{thm}
	Let $X$ and $Y$ be independent, with $\E(|X|) < \infty$ and $\E(|Y) < \infty$.  Then \[
		\E(XY) = \E(X) \E(Y)
	\]
\end{thm}

\begin{proof}
	Start with simple functions.  Then \[
		X(\omega) = \sum_{i=1}^n a_i \mathbf{1}_{A_i} (\omega)
	\] with $\{ A_i \}$ disjoint.
	Let \[Y(\omega) = \sum_{j=1}^m b_j \mathbf{1}_{B_j}(\omega) 
	\] with $\{ B_j \}$ disjoint.

	Independence implies $\P(A_i B_j) = \P(A_i) \P(B_j)$.

	Then \begin{align*}
		\E(XY) &= \sum_{i=1}^n \sum_{j=1}^m a_i b_j \E( \mathbf{1}_{A_i} \mathbf{1}_{B_j} ) \\
				&= \sum_{i=1}^n \sum_{j=1}^m a_i b_j \P(A_i B_j) \\
				&= \sum_{i=1}^n \sum_{j=1}^m a_i b_j \P(A_i) \P(B_j)
	\end{align*} by independence.

	Now extend to non-negative random variables $X,Y$ by constructing sequences of simple functions using monotone convergence theorem.  Let \[
		X_n(\omega) = \frac{i}{2^n} \quad \text{if } \frac{i}{2^n} < X(\omega) \leq \frac{i+1}{2^n}, i = 0,1,\dots, n 2^n 
	\]and zero if $X(\omega) > n$.

	For simple functions, we have \[
		\E(X_n Y_n) = \E(X_n) \E(Y_n)
	\] and so by the monotone convergence theorem, \[
		\E(XY) = \E(X) \E(Y)
	\]
\end{proof}

\begin{thm}
	Let $X$ and $Y$ be independent random variables.  Then \[
		\E(|X+Y|^r) < \infty
	\] if and only if \[
		\E(|X|^r) < \infty \text{ and } \E(|Y|^r) < \infty
	\] for any $r > 0$.
\end{thm}

\begin{lem}[$c_r$ inequality]
	We have \[
		|x+y|^r \leq c_r \left( |x|^r + |y|^r \right) 
	\] for $x,y$ real, $c_r$ constant, $r \geq 0$.
\end{lem}
\begin{proof}
	If $r = 0$, trivial.

	If $r = 1$, we obtain the triangle inequality.

	If $r > 1$, we have \begin{align*}
		|x+y|^r &\leq [ 2 \max( |x|, |y|) ]^r \\
				&= 2^r \max(|x|^r, |y|^r) \\
				&\leq 2^r (|x|^r, |y|^r)
	\end{align*} and setting $c_r = 2^r$ proves for $r > 1$.

	If $ 0 < r < 1$, consider $f(t) = 1 + t^r - (1+t)^r$, with $f(0) = 0$.  Differentiating, we have $f'(t) = rt^{r-1} - r(1+t)^{r-1} \geq 0$ for $t > 0$.  Thus $f(t)$ is increasing for $t >0$.  Hence \begin{align*}
		f(t) &> f(0) = 0 \\
		1+t^r &\geq (1+t)^r.
	\end{align*}  Using $t = \frac{|y|}{|x|}$, we obtain \[
		(|x| + |y|)^r \leq |x|^r + |y|^r
	\]

\end{proof}

% section lecture_5_thursday_17_march (end)

\section{Lecture 6 - Thursday 17 March} % (fold)
\label{sec:lecture_6_thursday_17_march}
\begin{lem}
	For any $\alpha > 0$ and distribution function $F$, \[
		\int_0^\infty x^{\alpha} \, dF(x) = \alpha \int_0^\infty x^{\alpha - 1} [ 1 - F(x)] \, dx
	\]
\end{lem}

\begin{proof}
	Consider.  Integrating by parts, we have that this is equal to

\begin{align*}
	\int_0^b x^{\alpha} \, dF(x) &= -\int_0^b x^\alpha \, d(1-F(x)) \\
							&= [-x^\alpha] (1-F(x))|_0^b + \int_0^b \alpha x^{\alpha - 1} (1 - F(x)) \, dx \\
							&= -b^{\alpha}(1-F(b)) + \int_0^b \alpha x^{\alpha - 1} (1 - F(x)) \, dx  \\
\end{align*}

We also have \[
	0 \leq b^\alpha (1-F(b)) \leq \int_b^\infty x^\alpha \, dF(x)
\]
If the LHS converges then $\lim_{b \rightarrow \infty} \int_0^\infty x^\alpha \, dF(x) \rightarrow 0$.  Thus the term $b^\alpha (1-F(b))$ is squeezed to zero.

Conversely, \[
	\int_0^b x^\alpha \, dF(x) \leq \int_0^b \alpha x^{\alpha - 1} (1-F(x)) \, dx
\] and so \[
	\int_0^\infty \alpha x^{\alpha - 1} (1-F(x)) \, dx < \infty \Rightarrow \int_0^\infty x^\alpha \, dF(x) < \infty.
\]
\end{proof}

\begin{thm}
	Let $X,Y$ independent and $r > 0$. Then \[
		\E( |X+Y|^r) <\infty  \iff \E(|X|^r) \infty, \E(|Y|^r) < \infty
	\] 
\end{thm}

\begin{proof}
	If $\E(|X|^r) < \infty$, $\E(|Y|^r) < \infty$.  Then \[
		\E(|X+Y|^r) \leq c_r (\E(|X|^r) + \E(|Y|)^r ) < \infty
	\]

	Assume $\E(|X+Y|^r) < \infty$.  Assume $X$ and $Y$ have median 0 (without loss of generality).  Then \[
		\P(X \leq 0) \geq \frac{1}{2}, \P(X \geq 0) \geq \frac{1}{2}
	\] Similarly for $Y$.

	Now, \begin{align*}
		\P(|X| > t) &= P(X < -t) + P( X > t), t > 0 \\
					&= \frac{P(X  < -t, Y \leq 0)}{P(Y \leq 0)} + \frac{P(X > t, Y \geq 0)}{P(Y \geq 0)} \\
					&= 2 P(X+Y \leq -t) + 2 P(X+Y > t) \\
					&= 2 P(|X+Y| > t)
	\end{align*} by independence.

	Using the previous lemma, we have \begin{align*}
		\E(|X|^r) & \int_0^\infty x^r \, dF(x) = r \int_0^\infty x^{r-1} P(|X| > x) \, dx \\
							&\leq 2r \int_0^\infty x^{r-1} P(|X+Y| > x) \, dx \\
							= 2r \E(|X+Y|^r).
	\end{align*}  So $\E(|X+Y|^r) < \infty \Rightarrow \E(|X|^r) < \infty$.  Similarly for $\E(|Y|^r) < \infty$.

\end{proof}

\begin{thm}
	If $X$ and $Y$ are independent with distribution functions $F$ and $G$ respectively, then 
	\begin{align*}
		P(X+Y \leq x) &= \int_\R F(x-y) \, dG(y) \\
				  	  &= \int_\R G(x-y) \, dF(y)
	\end{align*}
\end{thm}
\begin{proof}
	This is just a simple statement of Fubini's theorem.
\end{proof}

\begin{cor}
	Suppose that $X$ has an absolutely continuous distribution function \[
		F(x) = \int_{-\infty}^x f(u) \, du
	\] for some density function $f$ with $\int_\R f(x) \, dx = 1$ and $f \geq 0$.

	Let $Y$ be independent of $X$.  Then $X+Y$ has an absolutely continuous distribution with density \[
		\int_\R f(x-y) \, dG(y)
	\]

	Thus we have
	\begin{align*}
		P(X+Y \leq x)  	&= \int_\R \int_{-\infty}^x f(t-y) \, dt \, dG(y) \\
						&= \int_{-\infty}^x \int_\R f(t-y) \, dG(y) \, dt
	\end{align*}
\end{cor}

\begin{defn}
	A distribution function $F$ that can be represented in the form \[
		F(x) = \sum_j b_j \mathbf{1}_{[a_j, \infty]}(x)
	\] with $a_j$ real, $b_j \geq 0$, $\sum_{b_j} = 1$ is called \textbf{discrete}.

\end{defn}

If a distribution function is continuous then it may be:
\begin{enumerate}
	\item \textbf{Absolutely continuous}, in which case there is a density function $f \geq 0$ such that $F(b) - F(a) = \int_a^b f(u)\, du$.  $f$ is called the density.
	\item \textbf{Singular}, in which case $F'(x)$ exists and equal zero almost everywhere with respect to the Lebesgue measure (see Chung \S1.3)
\end{enumerate}

\begin{thm}
	Any distribution function $F$ can be written uniquely as a convex combination of a discrete, an absolutely continuous, and a singular distribution.  By convex, we mean a linear combination with non-negative coefficients summing to one.   
\end{thm}

\begin{thm}[Chebyshev's inequality]
	Let $X$ be a random variable and $g$ an increasing, non-negative function.  If $g(a) > 0$, then $$ P(X \geq a) \leq \frac{\E(g(X))}{g(a)}.$$
\end{thm}

\begin{proof}
	We have  \begin{align*}
		\E(g(X)) &= \int_\R g(x) \, dF(x) \\
				&\geq \int_a^\infty g(x) \, dF(x) \\
				&\geq g(a) \int_a^\infty dF(x) \\
				&= g(a) P(X \geq a) 
	\end{align*}
\end{proof}

\begin{cor}
	Let $g(x) = x^2$.  Then\[
		P( | X - \E(X) | > a) \leq \frac{\text{Var}(X)}{a^2}
 	\]

	Let $g(x) = e^{ax}$.  Then \[
		P(X \geq a) \leq \frac{\E(e^{cX})}{e^{ca}} = e^{-ca} \E(e^{cX})
	\]

	Let $g(x) = |x|^k, k > 0$. Then \[
		P(|X| \geq a) \leq \frac{\E(|X|^k)}{a^k}.
	\]
\end{cor}
% section lecture_6_thursday_17_march (end)

\section{Lecture 7 - Thursday 24 March} % (fold)
\label{sec:lecture_7_thursday_24_march}
\begin{defn}[Weak law of large numbers]
	Let $X_1, \dots, X_n \dots$ be IID random variables with $\E(X_i) = \mu$,$ \var(X_i) = \sigma^2 < \infty$.  Then \[
		\overline{X}_n \cip X
	\]
\end{defn}

\begin{proof}
	\begin{align*}
		P(|\overline{X}_n) &\leq \frac{\E(\overline{X}_n - \mu)^2}{\epsilon^2} \\
							&= \frac{\sigma^2/n}{\epsilon^2} \rightarrow 0
	\end{align*} as $n \rightarrow \infty$.

	We have \[
		\E(|\overline{X}_n - \mu|^2) = \sigma^2/n \rightarrow 0
	\] and so $\overline{X}_n$ converges to $\mu$ in $L^2$
\end{proof}

We can relax the assumptions to $E(|X) < \infty$ (no need to have finite variance).  See Chung (1974) p.109, Theorem 5.2.2.

\begin{thm}
	Let $X_i$ be uncorrelated, and $\E(X_i) = \mu_i$, $\var(X_i) = \sigma^2_i < \infty$ with \[
		\frac{1}{n^2} \sum_{i=1}^n \sigma^2_i \rightarrow 0
	\]  then we have \[
		\overline{X}_n - \frac{1}{n} \sum_{i=1}^n \mu_i \cip 0
	\]
\end{thm}
\begin{proof}
	\begin{align*} P(|\overline{X}_n - \frac{1}{n} \sum_{i=1}^n \mu_i | > \epsilon) &= P( \frac{1}{n} \sum_{i=1}^n (X_i - \mu_i)| > \epsilon) \\
		&\leq \frac{\var(\frac{1}{n}\sum_{i=1}^n (X_i - \mu_i))}{\epsilon^2} \rightarrow 0		
	\end{align*} as $\sum_{1}{n^2} \sum_{i=1}^n \sigma_i^2 \rightarrow 0$.  
\end{proof}

\begin{thm}[Borel-Cantelli lemma]
	Let $A_1, \dots$ be events in a probability space.  Let $B = \limsup A_n$ = $\bigcap_{n \geq 1} \bigcup_{m \geq n} A_m$.  Then 
	\begin{enumerate}[(i)]
		\item $\sum_{n} P(A_n) < \infty$ then $P(B) = 0$.  
		\item If $A_i$ are independent and $\sum_n P(A_n) \rightarrow \infty$ then $P(B) = 1$.
	\end{enumerate}
	For (ii) we need independence. Consider $A_i = A$ where $P(A) = \frac{1}{3}$.  Then \[
		B = \limsup A_n = A
	\] and $P(B) = \frac{1}{3}$
\end{thm}

\begin{proof}
	Preliminary lemma - if $0 < x < 1$, then $\log(1-x) < -x$.  We can then show that if $\sum_n a_n \rightarrow \infty$ then $\prod_n (1- a_n) \rightarrow 0$.

	\begin{enumerate}[(i)]
		\item \[ P(B) \leq P(\bigcup_{m \geq n} A_n) \leq \sum_{m  \geq n} P(A_m) \rightarrow 0 \] and so $P(B) = 0$.
		\item We will prove $P(\bigcup_{m \geq n} A_m) = 1$ for all $n$. Take $K > n$.  Then \begin{align*}
			1 - P(\bigcup_{m \geq n} A_m) &\leq 1 - P(\bigcup_{m=n}^K A_m) \\
										&= P( (\bigcup_{m=n}^K A_n)^c) \\
										&= P( \bigcup_{m=n}^K A_m^c) \\
										&= \prod_{m=n}^K ( 1 - P(A_m)) \quad \text{by independence} \\
										&\rightarrow 0
		\end{align*}  as $\sum_{n} P(A_n) \rightarrow \infty$ as $K \rightarrow \infty$.  Thus \[
			P(\bigcup_{m \geq n} A_m) = 1
		\] for all $m$, and so $P(B) = 1$.  
	\end{enumerate}
\end{proof}

\begin{thm}[Strong law of large numbers]  Let $X_1, \dots$ be IID random variables.  Let $\E(X_1) = \mu$, $\E(X_1^4) < \infty$.  Let $S_n = \sum_{j=1}^n X_j$.  Then \[
	\overline X_n = \frac{1}{n} S_n \cas \mu 
\]
\end{thm}

\begin{proof}
	\begin{align*}
		\E(\sum_{i=1}^n (X_i - \mu))^4 &= \sum_{i=1}^n E(X_i - \mu)^4 + 6 {n \choose 2} \sigma^4 \\
		&= n \E(X_1 - \mu)^4 + 3n(n-1) \sigma^4 \\
		&\leq Cn^2. 
	\end{align*}
	From Chebyshev, we have 
	\begin{align*}
		P(| S_n - \mu n| > \epsilon n) &\leq \frac{E(S_n - \mu n)^4}{)(\epsilon n)^4} \\
		 		&\leq \frac{cn^2}{\epsilon^4 n^4} = \frac{k}{n^2} 
	\end{align*}
	and so \[
		\sum_n P(|S_n - n \mu| > n \epsilon) < \infty,	\]
	and so $P(\limsup \{ | \frac{S_n}{n} - \mu | > \epsilon \} ) = 0$.  Letting $A_\epsilon = \{ | \frac{S_n}{n} - \mu | > \epsilon \}$.  Then 
	\begin{align*}
		P(|\frac{S_n}{n} - \mu | \text{ does not converge to zero}) &= P(\bigcup_k A_{1/k}) \\
		&\leq \sum_{k} P(A_{1/k}) \\
		&= 0
	\end{align*} by Borel-Cantelli.
\end{proof}
% section lecture_7_thursday_24_march (end)

\section{Lecture 8 - Thursday 24 March} % (fold)
\label{sec:lecture_8_thursday_24_march}
	Let $X_1, \dots$ be IID random variables with mean $\mu$.  Then \[
		P(\lim_{n \rightarrow \infty} \frac{S_n}{n} = \mu) = 1
	\]
	Conversely, if $\E(|X|)$ does not exist, then \[
		P( \limsup |\frac{S_n}{n}| = \infty) = 1
	\]

\begin{thm}
	If $E(X^2) < \infty$, and $\mu = 0$ (WLOG), 
	\begin{align*}
		P( |n^{-\alpha} S_n | \geq \epsilon) \leq \frac{E(S_n^2)}{n^{2 \alpha} \epsilon^2} \\
		&=n ^{1-2\alpha} \sigma^2/\epsilon^2 \rightarrow 0
 	\end{align*} provided $S \geq \frac{1}{2}, n^{-\alpha} S_n \cip 0$.  
\end{thm}

\begin{thm}[Hausdorff (1913)]   $|S_n | = \mathcal{O}(n^{\frac{1}{2} + \epsilon})$ a.s for any $\epsilon > 0$.

	Assumes $\E(|X_i|^r) < \infty$ for $r = 1, 2, \dots$. 
\end{thm}

\begin{proof}
	Previously, we showed $\E(S_n^4) \leq Cn^2$ for some $C > 0$.  Then we can extend this to \[
		\E(S_n^{2k}) \leq c_k n^k, k = 1, 2, \dots
	\] Then \begin{align*}
		P(n^{-\alpha} |S_n| > a ) &\geq \frac{c_k n^k}{(a n^{\alpha})^{2k}} \\
								&= c_k a^{-2k} n^{k(1-2\alpha)}
	\end{align*} and so \[
		\sum P( n^{-\alpha} |S_n| > a) < \infty 
	\] if $k(1-2\alpha) > -1$ i.e. $\alpha \geq \frac{1}{2} + \frac{1}{2k}$.

	By Borel-Cantelli, $P(|S_n| > an^{\alpha} \, i.o.) = 0$ if $\alpha > \frac{1}{2} + \frac{1}{2k}$.  
\end{proof} 
\begin{thm}[Hardy and Littlewood (1914)] $|S_n| = \mathcal{O}(\sqrt{n \log n})$ a.s.
\end{thm}
\begin{lem}
	Suppose $|X_i| \leq M$ a.s. ($X_i$ is bounded).  Then for any $x \in [0, \frac{2}{M}]$, we have \[
		\E(e^{x S_n}) \leq \exp[ \frac{nx^2 \sigma^2}{2}(1 + xM)]
	\]  
\end{lem}

\begin{proof}
	The random variables $e^{x X_i}$ are independent, so $\E(e^{x S_n}) = \left[ \E(e^{x X_1})\right]^n$.  We can then evaluate 
	\begin{align*}
		\E(e^{x X_1}) &= \E \left[ \sum_{k=0}^\infty \frac{(x X_1)^k}{k!} \right] \\
		&= 1 +0 + x^2 \sigma^2/2 + \E(\sum_{k=3}^\infty \frac{(x X_1)^k}{k!}) \\
		&\leq 1 + x^2 \sigma^2/2 + \sum_{k=3}^\infty \frac{x^k M^{k-2} \sigma^2}{k!} \\
		& \leq 1  + x^2 \sigma^2/2 + \sigma^2 M^{-2}/3! \sum_{k=3}^\infty \frac{x^k M^k}{3^{k-3}} \\
		&= 1 + x^2 \sigma^2/2 + \sigma^2 M^{-2}/6  \frac{ (xM/3)^3}{(1-xM/3)} \\
		&= 1 + x^2 \sigma^2/2 = \frac{\sigma^2 M x^3}{6(1-xM/3)}.
	\end{align*}  If $0 \leq x \leq 2/M$, we have \begin{align*}	
	\E(e^{x X_1}) 	&\leq 1 + \sigma^2 x^2/2 + \sigma^2 x^2 /2(xM) \\
						&= 1 + \sigma^2 x^2/2(1 + xM) \\
						&\leq \exp(\sigma^2 x^2/2(1 + xM))
	\end{align*}

\end{proof}

\begin{cor}
	For $0 < a < \frac{2\sigma^2 n }{M}$, under the conditions of the above Lemma, \[
		P(S_n \geq a) \leq e^{-\frac{a^2}{2n\sigma^2})1 - \frac{Ma}{n\sigma^2}}
	\]
\end{cor}
\begin{proof}
	\begin{align*}
		P(S_n \geq a) &\leq \frac{E(e^{xS_n})}{e^{ax}} \\
					&\leq \exp( \frac{n \sigma^2 x^2}{2}(1 + xM) - ax) \quad 0 < x \leq \frac{2}{M}
	\end{align*} Put $x = \frac{a}{n \sigma^2}$.  THen 
	\begin{align*}
		P(S_n \geq a) &\leq \exp( \frac{a^2}{2 n \sigma^2}(1 = \frac{aM}{n \sigma^2}) - \frac{a^2}{n \sigma^2}) \\
		&= \exp( \frac{-a^2}{2n \sigma^2}(1 - \frac{aM}{n \sigma^2})) 
	\end{align*}
\end{proof}

We can now prove the Hardy-Littlewood result.  
If $|X_i| \leq M$ almost surely then $|S_n| = \mathcal{O}(\sqrt{n \log n})$ a.s.
\begin{proof}
	Put $a = c \sqrt{ n \log n}$.  Then 
	\begin{align*}
		P(S_n \geq c \sqrt{n \log n}) &\leq \exp( \frac{c^2 \log n}{2 \sigma^2}( 1 - \frac{Mc \sqrt{\log n}}{\sqrt{n} \sigma^2})) \\
		&= n^{-c^2/2\sigma^2} \exp( \frac{Mc^3}{2\sigma^4} \frac{\log n \sqrt{\log n}}{\sqrt{n}}) 
	\end{align*}
	If $c^2 > 2 \sigma^2$ then $\sum_{n} P(S_n > c \sqrt{n \log n}) < \infty$.  By Borel-Cantelli, we then have \[
		P(S_n > c \sqrt{n \log n} \, i.o.) = 0
	\] Now apply the argument to $-X_i$.  Then \[
		P(-S_n > c \sqrt{n \log n} \, i.o.) = 0
	\]
\end{proof}

\begin{thm}[Khintchine (1923)] $|S_n | = \mathcal{O}(\sqrt{ n \log \log n})$ a.s.
\end{thm}
\begin{thm}[Khintchine (1924)]Let $X_i = \pm 1$ with probability $\frac{1}{2}$.  Then \[
	\limsup \frac{|S_n|}{\sqrt{n \log \log n}} = \sqrt{2} a.s.
\]
\end{thm}

% section lecture_8_thursday_24_march (end)

\section{Lecture 9 - Thursday 31 March} % (fold)
\label{sec:lecture_9_thursdy_31_march}
\begin{defn}[Induced $\sigma$-field]
	Let $(\Omega, \sigf, \P)$ be a probability space.  Let $Y$ be a set of random variables on $(\Omega, \sigf)$.  Then $\sigma(Y)$ is the smallest $\sigma$-field contained in $\sigf$ with respect to which each $X \in Y$ is measurable.

	That is, for each $B \in \mathcal{B}$, the Borel $\sigma$-field on $\R$, we have \[
		X^{-1}(B) \in \sigma(Y)
	\]  Thus $\sigma(Y)$ is the intersection of all $\sigma$-fields which contain every set of the form $X^{-1}(B)$ for all $B \in \mathcal{B}, X \in Y$.  
\end{defn}

\begin{defn}[Independent $\sigma$-fields]
	If $X_1, \dots$ are independent random variables and $A_i \in \sigma(X_i)$, then \[
		P(\bigcap_{i=1}^n A_i) = \prod_{i=1}^n P(A_i) \tag{$\star$}
	\]
	If $\sigf_1, \sigf_2, \dots$ are $\sigma$-fields contained in $\sigf$ and $(\star)$ holds for any $A_i \in \sigf_i$ then we sa the $\sigma$-fields are independent. 
\end{defn}

\begin{thm}
	Let $\sigf_0, \sigf_1, \dots$ be independent $\sigma$-fields and let $\mathcal{G}$ be $\sigma$-fields generated by any subset of $\sigf_1, \sigf_2, \dots$.  Then $\sigf_0$ is independent of $\mathcal{G}$.  
\end{thm}

\begin{proof}
	\textbf{Outline.} Take $\mathcal{G}$ to be the smallest $\sigma$-field containing $\sigf_1, \sigf_2, \dots$.

	If $A \in \sigf_0$, $B \in \mathcal{G}$, then we need to show \[
		P(A \cap B) = P(A) P(B).
	\]
	\begin{enumerate}[(1)]
		\item Assume $P(A) > 0$. 
		\item If $B = A_1 \cap A_2 \dots A_n$ then the result is true.
		\item Let $\mathcal{G}_a$ be the class of \textbf{finite} unions of $B$.  Then $\mathcal{G}_a$ is a finitely additive field, and $G \in \mathcal{G}_a$ can be written as $G = \bigcup_{i=1}^k G_i$ where $G_i$ has the form of $B$ above.  Then 
		\begin{align*}
			P(A \cap G) &= P( \bigcup_{i=1}^k A \cap G_i) \\
						&= \sum P(A \cap G_i) = \sum_{i=j} P(A \cap G_i \cap G_j) + \dots \\
						&= P(A) P(G)
		\end{align*} by the inclusion-exclusion formula and independence of $A$ and $G_i$.  
		\item Now, let $P_A(B) = \frac{P(A\cap B)}{P(A)}$.  Then $P_A$ and $P$ are measures on $\sigf$, and $P$ and $P_A$ agree on $\mathcal{G}_a$.  Thus by the extension theorem they agree on the $\sigma$-field generated by $\mathcal{G}_a$ which includes $\mathcal{G}$.
	\end{enumerate}
\end{proof}

\begin{defn}[Tail $\sigma$-field]
	Let $X_1, X_2, \dots$ be a sequence of random variables and let \[
		\sigf_n = \sigma(\{ X_n , X_{n+1}, \dots \})
	\] be the $\sigma$ field generated by $X_n, X_{n+1}$.  Then \[
		\sigf_n \supseteq F_{n+1} \supseteq F_{n+2} \dots
	\] and let \[
		\mathcal{T} = \bigcap{n} \sigf_n
	\] be the \textbf{tail $\sigma$-field}.

	$\mathcal{T}$ is the collection of events defined in terms of $X_1, X_2, \dots$ not affected by altering a finite number of the random variables.

\end{defn}

\begin{thm}[The $0-1$ law]
	Any set belonging to the tail $\sigma$-field of a sequence of independent random variables has probability $0$ or $1$.  	
\end{thm}	
\begin{proof}
	We have $\sigma(X_n)$ is independent of $\sigma( \{ X_{n+1}, X_{n+2}, \dots \}) = \sigf_{n+1} \supseteq \mathcal{T}$ and so $\mathcal{T}$ is independent of $\sigma(X_n)$ for every $n$.  By the previous theorem, it follows that $\sigf$ is independent of $\mathcal{G} = \sigma(\{ X_1, X_2, \dots \})$ but as $\mathcal{T} \subseteq \mathcal{G}$, we know that $\mathcal{T}$ is independent of itself.  Thus, for any $A \in \mathcal{T}$, \[
		P(A \cap A) = P(A) P(A)
	\] and so $P(A) = 0$ or $1$.  
\end{proof}

\subsection{Martingales} % (fold)
\label{sub:martingales}

% subsection martingales (end)
\begin{defn}[Martingale]
	Let $(\Omega, \sigf, \P)$ be a probability space.  Let $\{ \sigf_n \}$ be an increasing sequence of $\sigma$-fields.  \[
		\sigf_n \subseteq \sigf_{n+1} \subseteq \dots \subseteq \sigf. 
	\]  Let $\{ S_n \}$ be a sequence of random variables on $\Omega$.  Then $\{ S_n \}$ is a \textbf{martingale} with respect to $\{ \sigf_n \}$ if 
	\begin{enumerate}[(1)]
		\item $S_n$ is measurable with respect to $\sigf_n$. 
		\item $\E(|S_n|) < \infty$.
		\item $\E(S_n \, | \, \sigf_m) = S_m$ almost surely for all $m \leq n$. 
	\end{enumerate}
\end{defn}

% section lecture_9_thursdy_31_march (end)

\section{Lecture 10 - Thursday 31 March} % (fold)
\label{sec:lecture_10_thursday_31_march}
\begin{defn}[Supermartingale]
$\{ S_n \}$ is a \textbf{supermartingale} with respect to $\{ \sigf_n \}$ if 
	\begin{enumerate}[(1)]
		\item $S_n$ is measurable with respect to $\sigf_n$. 
		\item $\E(|S_n|) < \infty$.
		\item $\E(S_n \, | \, \sigf_m) \leq S_m$ almost surely for all $m \leq n$. 
	\end{enumerate}
\end{defn}

\begin{defn}[Submartingale]
$\{ S_n \}$ is a \textbf{submartingale} with respect to $\{ \sigf_n \}$ if 
	\begin{enumerate}[(1)]
		\item $S_n$ is measurable with respect to $\sigf_n$. 
		\item $\E(|S_n|) < \infty$.
		\item $\E(S_n \, | \, \sigf_m) \geq S_m$ almost surely for all $m \leq n$. 
	\end{enumerate}
\end{defn}

\begin{defn}[Regular martingale]
	Let $X$ is a random variable $\E(|X|) < \infty$, $S_n = \E(X \, | \, \sigf_n)$ and assume $\{ S_n \}$ is a martingale with respect to $\{ F_n \}$.

	If a martingale can be written in this way for some $X$ then it is \textbf{regular}.  
\end{defn}

Not every martingale is a regular martingale.

\begin{exmp}
	Assume $P(X_i = 1) = p$, $P(X_i = -1) = 1-p$, and let $S_n = \sum_{i=1}^n X_i$.  If $p \neq \frac{1}{2}$ then \[
		Y_n = \left( \frac{1-p}{p} \right)^{S_n}
	\] is a martingale with respect to $\sigf_n = \sigma(X_1, \dots,X_n)$, since \begin{align*}
		\E(Y_n \, | \, \sigf_{n-1}) &= \E\left(\left( \frac{1-p}{p} \right)^{S_n + X_n} \, | \, \sigf_{n-1} \right) \\
		&= \left( \frac{1-p}{p} \right)^{S_n} \left[\left( \frac{1-p}{p} \right) p + \left(\frac{1-p}{p} \right)^{-1}(1-p) \right] \\
		&= Y_{n-1}
	\end{align*}
\end{exmp}

\subsection{Conditional expectations} % (fold)
\label{sub:conditional_expectations}
If $\sigg \subseteq \sigf$ then \[
	L^2(\sigg) = \{ X \, | \, \E(X^2) < \infty, \text{$X$ is $\sigg$-measurable} \}
\]

If $Y \in L^2$ define $Z = \E(Y \given \sigg)$ to be the projection of $Y$ onto $L^2(\sigg)$, where \[
	\E(Y-Z)^2 = \inf_{U \in L^2(\sigg)} \E(Y - U)^2
\] Then $Y-Z$ will be orthogonal to the elements of $L^2(\sigg)$.  That is, \begin{align*}
	\int (Y-Z) X \, dP = 0 
\end{align*} for all $X \in L^2(\sigg)$.  If $A \in \sigg$, then letting $X = \mathbf{1}_{A}$, we have \[ \boxed{
	\int_A Y \, dP = \int_A \E(Y \given \sigg) \, dP
}
\]

If $Y \geq 0$ construct $\{ Y_n \}$ with $Y_n \in L^2$ such that $Y_n \uparrow Y$.  Define \[
	\E(Y \given \sigg) = \lim_{n \rightarrow \infty} \E(Y_n \given \sigg).
\] The limit exists as \[
	\E(Y_n \given \sigg) \geq \E(Y_m \given \sigg), n \geq m.
\]  We still have 
\begin{enumerate}[(1)]

	\item $\E(Y \given \sigg)$ is $\sigg$-measurable, and 
	\item For all $A \in \sigg$, \[
		\int_A Y \, dP = \int_A \E(Y \given \sigg) \, dP
	\] as \[
		\int_A \E(Y \given \sigg) \, dP= \lim_{n \rightarrow \infty} \int_A \E(Y_n \given \sigg) \, dP = \lim_{n \rightarrow \infty} \int_A Y_n \, dP  = \int_A Y \, dP
	\] by the monotone convergence theorem.  
\end{enumerate}

If $Y \in L^1$, defining $Y = Y^+ - Y^-$, we define \[
	\E(Y \given \sigg) = \E(Y^+ \given \sigg) - \E(Y^- \given \sigg).  
\]
% subsection conditional_expectations (end)

\subsection{Stopping times} % (fold)
\label{sub:stopping_times}

\begin{defn}
	A map \[
		\nu : \Omega \rightarrow \overline{\N} = \{ 0, 1, 2, \dots, \infty \}
	\] is called a \textbf{stopping time} with respect to $\{ \sigf_n \}$, an increasing sequence of $\sigma$-fields, if \[
		\{ \nu = n \} \in \sigf_n.
	\] and thus \[
		\{ \nu \leq n \}, \{ \nu > n \} \in \sigf_n
	\]
\end{defn}

\begin{thm}[Properties of stopping times]
	Let $\sigf_\infty = \vee_{n=1}^\infty \sigf_n$, the $\sigma$-field generated by all $\sigf_n$.   Then we have 
	\begin{enumerate}[(1)]
		\item For all stopping times $\nu$, $\nu$ is $\sigf_\infty$-measurable.\[
			\{ \nu = n \} \in \sigf_n, \{ \nu = \infty \} = \{ \bigcup_{n} \{ v = n \} \}^c \in \sigf_\infty
		\]
		\item The minimum and maximum of a countable sequence of stopping times is a stopping time.  To prove this, let $\{ v_k\}$ be a sequence of stopping times.  Then \begin{align*}
		\{ \max_{k} v_k \leq n \} &= \bigcap_k \{ v_k \leq n \} \in \sigf_n	\\
		\{ \min_{k} v_k > n \} &= \bigcap_k \{ v_k > n \} \in \sigf_n							
		\end{align*}
	\end{enumerate}	
\end{thm}

\begin{lem}
	Let $\{ Y_n^1 \}$ and $\{ Y_n^2 \}$ be two positive supermartingales with respect to $\{ \sigf_n \}$, an increasing sequence of $\sigma$-fields.  Let $\nu$ be a stopping time.  If $Y_n^1 \geq Y_n^2$ on $[\nu = n]$, then \[
		Z_n = Y_n^1 \indic{\nu > n} + Y_n^2 \indic{\nu \leq n}
	\] is a positive supermartingale.  
\end{lem}

\begin{proof}
	We have that $Z_n$ is $\sigf_n$-measurable and positive.  We then have \begin{align*}
		\E(Z_n \given \sigf_{n-1}) &= \E( Y_n^1 \indic{\nu > n} + Y_n^2 \indic{\nu \leq n} \given \sigf_{n-1}) \\
		&= \E( Y_n^1 \indic{\nu > n-1} - Y_n^1 \indic{\nu = n} + Y_n^2 \indic{\nu \leq n-1} + Y_n^2 \indic{\nu = n}\given \sigf_{n-1}) \\
		&\leq Y_n^1 \indic{\nu > n-1} + Y_n^2 \indic{\nu \leq n-1} + \E((Y_n^2 - Y_n^1) \indic{\nu = n} \given \sigf_{n-1}) \\
		&\leq Z_{n-1} 
	\end{align*} as $Y_n^2 - Y_n^1 < 0$ on $\{ \nu = n \}$.  
\end{proof}
% subsection stopping_times (end)
% section lecture_10_thursday_31_march (end)

\section{Lecture 11 - Thursday 7 April} % (fold)
\label{sec:lecture_11_thursday_7_april}

\begin{thm}[Maximal inequality for positive supermartingales]
	Let $\{ Y_n \}$ be a positive supermartingale with respect to $\{ \sigf_n \}$.  Then \[
		\sup_{n } Y_n < \infty a.s
	\] on $[Y_0 < \infty]$ and \[
		P(\sup_{n} Y_n > a \, | \, \sigf_0 ) \leq \min(1, \frac{Y_0}{a})
	\]
\end{thm}
\begin{proof}
	Fix $a > 0$ and let $\nu_a= \inf \{ n : Y_n > a \} = \infty$ if $\sup_n Y_n \leq a$.  Then the sequence $Y_n(2) = a$ is a positive supermartingale, and so \[
		Z_n = Y_n \indic{\nu_a> n} + a \indic{\nu_a\leq n}
	\] is a positive supermartingale by the previous lemma.  Then we have \[
		\E(Z_n \, | \, \sigf_0) \leq Z_0 = \begin{cases}
			Y_0 & Y_0 \leq a \\
			a  & Y_0 > a
		\end{cases}
	\]  Thus $Z_n \geq a \indic{\nu_a\leq n}$ and so \[
		a P({va \leq n \given \sigf_0}) \leq \min(Y_0, a) 
	\] for all $a$.  THus \[
		P(\sup_{n} Y_n > a \given \sigf_0 ) = P(\nu_a< \infty \given \sigf_0) \leq \min(1, \frac{Y_0}{a})
	\]
\end{proof}

Write \begin{align*}
	P(Y_0 < \infty, \sup_{n} Y_n > a) &= \E(\indic{Y_0 < \infty} \indic{\sup_n Y_n > a}) \\
									&= \E(\indic{Y_0 < \infty}) \E(\indic{\sup_n Y_n > a} \given \sigf_0) \\
									&\leq \int_{Y_0 < \infty} \min(1, \frac{Y_0}{a}) \, dP \\
									&\rightarrow 0
\end{align*} as $a \rightarrow \infty$ by the dominated convergence theorem.

Thus, we have \[
	P(Y_0 < \infty, \sup_{n} Y_n < \infty) = 1 a.s.
\]

Fix $a < b \in \R$.  For any process $Y_n$, define the following random variables \begin{align*}
	\nu_1 &= \min(n \geq 0, Y_n \leq a) \\
	\nu_2 &= \min(n > \nu_1, Y_n \geq b) \\
	\nu_3 &= \min(n > \nu_2, Y_n \leq a)
\end{align*} and so on.  If any $v_i$ is undefined it is subsequently set to infinity.

Define $\beta_{ab} = \max\ { p : \nu v_{2p} < \infty}$, equal to the number of upcrossings of $(a,b)$ by $Y_n$.  We have $\beta_{ab} = \infty$ if and only if $\liminf y_n \leq a < b \leq \limsup y_n$.  We also have $Y_n$ converges if and only if $\beta_{ab} < \infty$ for all rationals $a,b$, $a < b$.

\begin{thm}[Dubin's inequality] If $Y_n$ is a positive supermartingale, then $\beta_{ab}(\omega)$ are random variables and for each integer $k \geq 1$, we have \[
	P(\beta_{ab} \geq k \given \sigf_0) \leq (\frac{a}{b})^k \min(1, \frac{Y_0}{a}), 0 < a < b.
\]	
\end{thm} 
\begin{proof}
	The $v_k$ defined above are stopping times with respect to $\sigf_n$, as \[ 
		[\nu_{2p} = n] = \bigcup_{m=0}^{n-1} [\nu_{2p-1} = m, Y_{m+1} \leq b, \dots, Y_{n-1} < b, Y_n \geq b ]
\] and as $\nu_1$ is a stopping time, we then use induction.

	We then have $[\beta_{ab} \geq k] = [\nu_{2k} < \infty]$.  Then define \begin{align*}
		Z_n &= \indic{0 \leq n < \nu_1} + \sum_{k=1}^K (\frac{b}{a})^{k-1} \frac{Y_n}{a} \indic{\nu_{2k-1} \leq n \leq \nu_{2k}} \\
			&\qquad + \left(\frac{b}{a}\right)^k \indic{\nu_{2k} \leq n < \nu_{2k+1}} + \left( \frac{b}{a} \right)^K \indic{n \geq \nu_{2K+1}}
	\end{align*} 
	i.e. $\indic{0 \leq n < \nu_1} + \frac{Y_n}{a} \indic_{\nu_1 \leq n < \nu_2} + \frac{b}{a} \indic{\nu_2 \leq n < \nu_3} + \dots + \left( \frac{b}{a} \right)^K \indic{\nu_{2K} \leq n}$.

	We now apply the previous lemma to show $\{ Z_n \}$ is a positive supermartingale.  We have \[
		\left( \frac{b}{a} \right)^k, \left( \frac{b}{a} \right)^{k-1} \frac{Y_n}{a}
	\] are positive supermartingales.  On $[ \nu_1 = n]$, we have $1 \geq \frac{Y_n}{a}$.  On $[\nu_{2k-1} = n]$ we have \[
		\left(\frac{b}{a} \right)^{k-1} \geq \left(\frac{b}{a} \right)^{k-1} \frac{Y_n}{a}
	\]

	On the even stopping times, we have $\left(\frac{b}{a} \right)^{k-1} \frac{Y_n}{a} \geq \left(\frac{b}{a} \right)^{k}$.  Thus \[
		\E(Z_n \given \sigf_0) \leq Z_0
	\] as $Z_n$ is a positive supermartingale. d

	Since $Z_n \geq \frac{b}{a}^K \indic{\nu_{2k} \leq n}$, we have \[
		P(\nu_{2k} \leq n \, | \, \sigf_0) \leq \frac{a}{b}^K \min(1, \frac{Y_0}{a}) 
	\]  Letting $n \rightarrow \infty$, we have 
	\begin{align*}
		P(\beta_{ab} \geq k \given \sigf_0) &= P(\nu_{2k} < \infty \given \sigf_0) \\
		 		&\leq \left( \frac{a}{b} \right)^K \min(1, \frac{Y_0}{a}).  
	\end{align*}
\end{proof}

% section lecture_11_thursday_7_april (end)

\section{Lecture 12 - Thursday 7 April} % (fold)
\label{sec:lecture_12_thursday_7_april}
\begin{thm}
	Let $\{ Y_n \}$ be a positive supermartingale.  Then there exists a random varaible $Y_\infty$ such that $Y_n \cas Y_\infty$ and $\E(Y_\infty \given \sigf_n) \leq Y_n$ for all $n$.
\end{thm}
\begin{proof}
	From Durbin's inequality, \[
		P(\beta_{ab} \geq k) \leq \left( \frac{a}{b} \right)^k
	\]  By Borel-Cantelli, as we have a summable sequence of probabilities, $\beta_{ab}  < \infty$ almost surely.  Hence \[
		P(\text{$Y_n$ converges}) = P\left(\bigcap_{\substack{a < b \\ a, b \in \Q}} \beta_{ab} < \infty \right) = 1
	\] Let $\lim_{n \rightarrow \infty}Y_n = Y_\infty$. If $p < n$, then \[
		\E\left( \inf_{m \geq n} Y_n \given \sigf_p \right) \leq \E(Y_n \given \sigf_p) \leq Y_p.	\]  Furthermore, $\inf_{m \geq n} Y_m \uparrow Y_\infty$ so by the monotone convergence theorem, we have \[
			\E(Y_\infty \given \sigf_p) = \lim_{n \rightarrow \infty} \E\left( \inf_{m \geq n} Y_m \given \sigf_p \right) \leq Y_p.
		\]  
\end{proof}

\begin{thm}
	Let $Z$ be a positive random variable with $\E Z^p < \infty$, $p \geq 1$. Then \[
		Y_n = \E(Z_n \given \sigf_n) \cas, \clp \E(Z \given \sigf_\infty), 
	\]  Note that almost sure convergence does not, in general, imply $L^p$ convergence, although they both imply convergence in probability.  
\end{thm}
\begin{proof}
	Suppose $Z \leq a$ almost surely.  Then there exists $Y_\infty$ such that $Y_n \cas Y_\infty$ (as $Y_n$ are positive martingales).  Fix $n$ and let $B \in \sigf_n$.  Then \[
		\lim_{n \rightarrow \infty} \int_B Y_{m+n} \, dP = \int_B Z \, dP
	\] by definition of conditional expectation.  Now $0 \leq Y_n \leq a$ so by the dominated convergence theorem, \[
		\int_B Y_\infty \, dP = \int_B Z \, dP
	\] and hence \[
		Y_\infty = \E(Z \given \sigf_\infty)
	\] and so the random variable $Y_\infty$ can be identified as the conditional expectation.

	Since $|Y_n| \leq a$, the $\{ Y_n^p \}$ are uniformly integrable, and so $Y_n \clp Y_\infty$.  This follows from noting that $Y_n \cas Y_\infty$, and using that if $X_n \cip X$ and $\{ |X_n|^p \}$ is uniformly integrable then $X_n \clp X$.

	Now remove the assumption that $Z \leq a$. Taking the $L^p$ norm of the conditional expectations gives \[
		\| E(Z \given \sigf_n) - \E(Z \given \sigf_\infty) \|_p \leq \| E(Z \wedge a \given \sigf_n) - \E(Z \wedge a \given \sigf_\infty) \|_p + 2 \| (Z - a)^+ \|_p.
	\]  Now we know that $\|(Z -a)^+\|_p \rightarrow 0$ as $a \rightarrow \infty$, as $\E(Z^p) < \infty$.  Hence we have \[
		Y_n \clp \E(Z \given \sigf_\infty).
	\]  By uniqueness of limits, we obtain our required result.  
\end{proof}
\begin{cor}
	If $Z \in L^p$ and $Y_n = \E(Z \given \sigf_n)$ then $Y_n \cas, \clp \E(Z \given \sigf_\infty)$
\end{cor}

\begin{thm}{Martingale convergence theorem}{\ }
	\begin{enumerate}[(a)]
		\item 
		If $\{ Y_n \}$ is an integrable submartingale and $\sup_{n} \E(Y_n^+) < \infty$ then there exists an integrable $Y_\infty$ such that \[
		Y_n \cas Y_\infty
		\]  
		\item
		If $\{ Y_n \}$ is an integrable martingale satisfying $\sup_n \E |Y_n| < \infty$ then there exists an integrable $Y_\infty$ such that \[
			Y_n \cas Y_\infty.
		\]
	\end{enumerate}
\end{thm}

\begin{proof}{\ }
	\begin{enumerate}[(a)]
		\item $\{ Y_n^+ \}$ is a positive submartingale as \[
			\E(Y_{n+1}^+ \given \sigf_n) \geq \E(Y_{n+1} \given \sigf_n) \geq Y_n
		\]  If $p > n$, then \begin{align*}
			\E(Y_{p+1}^+ \given \sigf_n) &= \E(\E(Y_{p+1}^+ \given \sigf_p) \given \sigf_n) \\
			&\geq \E(Y_{p^+} \given \sigf_n). 
		\end{align*}   Hence $M_n = \lim_{p \rightarrow \infty}\E(Y_p^+ \given \sigf_n)$ as we have a monotone sequence.

		Now, \begin{align*}
			\E(M_n) &= \E\left(\lim_{p \rightarrow \infty} \E(Y_p^+ \given \sigf_n) \right) \\
			&= \lim_{p \rightarrow \infty} \E(\E( Y_p^+ \given \sigf_n)) \quad \text{MCT} \\
			&= \lim_{p \rightarrow \infty} \E(Y_p^+) < \infty
		\end{align*} so $M_n$ is positive and integrable.  $M_n$ is a martingale as \begin{align*}
			\E(M_{n+1} \given \sigf_n) &= \E\left(\lim_{p \rightarrow \infty} \E(Y_p^+ \given \sigf_{n+1}) \given \sigf_n \right)  \\
			&= \lim_{p \rightarrow \infty} \E(Y_p^+ \given \sigf_n) \quad \text{MCT} \\
			&= M_n.
		\end{align*}

		Let $Z_n = M_n - Y_n$. Then $Z_n$ is integrable as $M_, Y_n$ are, and $Z_n$ is a positive supermartingale, as \begin{align*}
			\E(Z_{n+1} \given \sigf_n) &= \E(M_{n+1} \given \sigf_n) - \E(Y_{n+1} \given \sigf_n) \\
			&\leq M_n - Y_n \quad \text{as $Y_n$ is submartingale} \\
			&= Z_n
		\end{align*} and so $Z_n$ is a positive supermartingale.  Note that $M_n \geq Y_{n}^+$ and so \[
			M_n - Y_n = M_n -(Y_n^+ - Y_n^-) \geq Y_n^+ - (Y_n^+ - Y_n^-) = Y_n^-
		\]  Thus $Z_n$ and $M_n$ converge almost surely to $Z_\infty$ and $M_\infty$ respectively, and so \[
			Y_n = M_n - Z_n \cas M_\infty - Z_\infty = Y_\infty \in L^1.  
		\]
		\item Note that $|Y_n| = 2Y_n^+ - Y_n$, and if $\{ Y_n \}$ is a martingale, then \begin{align*}
			\E |Y_n| &= 2 \E Y_n^+ - \E Y_n \\
			& 2 \E Y_n^+ - \E Y_0 
		\end{align*} and so $\sup \E Y_n^+ < \infty$ if and only if $\sup_n \E |Y_n| < \infty$.  
	\end{enumerate}
\end{proof}
\begin{thm}[Martingale convergence theorem (restated)]
	Let $\{Y_n\}$ be an integrable (sub/super) martingale, that is, $\sup_n \E |Y_n| < \infty$.  Then there exists an almost sure limit \[
		\lim_{n \rightarrow \infty} Y_t = Y_\infty
	\] and $Y_\infty$ is an integrable random variable.
\end{thm}
% section lecture_12_thursday_7_april (end)

\section{Lecture 13, 14 - Thursday 14 April} % (fold)
\label{sec:lecture_13_thursday_14_april}

\begin{defn}[Reverse martingale]
	$\{ Y_n , \sigg_n \}$ is a reverse martingale if $ \{ \sigg_n \}$ is a decreasing sequence of $\sigma$-fields, \[
		\sigg_n \supseteq \sigg_{n+1}
	\] $Y_n$ is $\sigg_n$-measurable, $\E(|Y_n|) < \infty$, and \[
		\E(Y_n \given \sigg_n) = Y_m \, \text{a.s for $m \geq n$}
	\]
\end{defn}

\begin{prop}
	We have \begin{align*}
		\E(|Y_n|) &= \E( \E(|Y_n| \given \sigg_{n+1})) \\
					&\geq \E(| \E(Y_n \given \sigg_{n+1})|) \\
					&= \E(|Y_{n+1}|)
	\end{align*}
	and so $\E(|Y_n|) \leq \E(|Y_0|)$ for all $n$, and \[
		Y_n = \E(Y_0 \given \sigg_n).
	\]
\end{prop}

\begin{thm}
	If $\{ Y_n \}$ is a reverse martingale with respect to $\{\sigg_n \}$, then there exists a random variable $Y_\infty$ such that \[
		Y_n \cas Y_\infty, Y_n \clone Y_\infty = \E(Y_0 \given \sigg_\infty)
	\] where $\sigg_\infty = \bigcap {\sigg_n}$.
\end{thm}

\begin{proof}
	We have $Y_n = \E(Y_0 \given \sigg_n)$ and so $\{ Y_n \}$ is uniformly integrable.  Hence if $Y_n \cas Y_\infty$ it also converges in $L^1$.  Let \[
		Z_n = \E(Y_0^+ \given \sigg_n) - Y_n.
	\]  Note that $Z_n \geq 0$.  Then \[
		\E(Z_n \given \sigg_{n+1}) = Z_{n+1}
	\] and so we only need to consider convergence for positive reverse martingales.

	Let $\beta_{a,b}^{(n)}$ be the number of upcrossings of $[a,b]$ by $\{ Y_0, Y_1, \dots, Y_n \}$.  Applying Dubin's inequality to the martingale \[
		\{ Y_n, Y_{n+1}, \dots, Y_{1}, Y_0 \}
	\] Then \[
		P(\beta_{a,b}^{(n)} \geq k \given \sigg_n) \leq \left( \frac{a}{b} \right)^k
	\] which is independent of $n$, and thus \[
		P(\beta_{a,b}^{(n)} \geq k \given \sigg_\infty) \leq \left( \frac{a}{b} \right)^k
	\] for all $n$, and so \[
		P(\beta_{a,b} \geq k \given \sigg_\infty) \leq \left( \frac{a}{b} \right)^k.
	\] where $\beta_{a,b}$ is the number of upcrossings for $\{ Y_n \}$, which implies \[
		\beta_{a,b} < \infty \, \text{a.s.}
	\]

	Arguing as in the positive supermartingale case, we have $\{ Y_n \}$ converges almost surely, and we have $Y_\infty = \limsup Y_n$ is $\sigg_n$ measurable for all $n$ and so is $\sigg_\infty$ measurable.  
\end{proof}

\begin{thm}[Strong law of large numbers]
	Let $X_1, X_2, \dots$ be IID with $\E(|X_1|) < \infty$.  Let $\E(X_1) = \mu$.  Let $S_n = \sumni X_i$.  Then \[
		\frac{1}{n} S_n \cas \mu.
	\]
\end{thm}

\begin{proof}
	Let $\sigg_n = \sigma \{ S_n , S_{n+1}, S_{n+2}, \dots \} = \sigma \{ S_n, X_{n+1}, X_{n+2}, \dots \}$.  We then have $\sigg_n \supseteq \sigg_{n+1}$. We have \begin{align*}
		\frac{1}{n} S_n &= \E(\frac{1}{n} S_n \given \sigg_n) \\
		 	&= \frac{1}{n} \sumni \E(X_i \given \sigg_n) \\
			&= \E(X_1 \given \sigg_n),
	\end{align*}  as \[
		\E(X_1 \given \sigg_n) = \E(X_2 \given \sigg_n) = \dots \E(X_n \given \sigg_n)
	\] by IID/symmetry.

	Thus $\frac{1}{n} S_n$ is a reverse martingale with respect to $\{ G_n \}$.  From above, we have have \[
		\frac{1}{n} S_n = \overline X_n \cas, \clone \E(X \given \sigg_\infty). 
	\]

	We have $\lim_{ n \rightarrow \infty} \sumni X_i$ is in the tail $\sigma$-field of the sequence of $\{ X_n \}$ and $X_i$ are IID and so the limiting random variable is degenerate.

	Consider $\overline X_\infty = \E(X \given \sigg_\infty)$.  By the Kolmogorov 0-1 law, we have \[
		P(\{ \overline X_\infty \leq a \}) = \text{0 or 1}.   
	\]

	Thus $\overline X_\infty$ is a constant with probability one.  Since \begin{align*}
		\E(X_1 \given \sigg_n) \clone \E(X_1 \given \sigg_\infty)
	\end{align*}  we have \[
		\lim_{n \rightarrow \infty} \E(\frac{1}{n} S_n) = \E( \E(X_1 \given \sigg_\infty)) = \E(X_1) = \mu.
	\]

	Thus $\overline X_\infty = \mu$ almost surely, that is, \[
		\frac{1}{n} S_n \cas, \clone \mu.
	\]
\end{proof}

\subsection{Characteristic functions} % (fold)
\label{sub:characteristic_functions}

Following \textbf{Fallow Volume 2}.
\begin{defn}[Characteristic function]
	Let $X$ be a random variable. Then the characteristic function is defined by \[\phi(t) = \E(e^{itX}).\]

	$\phi(t)$ is always defined (unlike moment generating function (MGF), probability generating function (PGF)).

\end{defn}

\begin{proof}
	Let $\phi(t)$ be the characteristic function of the random variable $X$.  Then 
	\begin{enumerate}[(i)]
		\item  $| \phi(t) | \leq \E(|e^{itX}|) = 1 = \phi(0).$ 
		\item $\phi(-t) = \E(e^{-itX}) = \overline{ \phi(t)}.$
		\item If $X$ is symmetric about 0 then $\phi(t)$ is real. 
		\item $\phi(t)$ is uniformly continuous in $t$.  \begin{proof}
			\begin{align*}
				|\phi(t+h) - \phi(t) &= \left| \int e^{i(t+h)X} - e^{itX} \, dF(x) \right| \\
							&= \left| \int e^{itX}(e^{ihX} - 1) \, dF(x) \right| \\ 
							&\leq \int \left| e^{ihX} - 1 \right|\, dF(x) \\
							& = \int \sqrt{\cos^2(xh - 1) + \sin^2(xh)} \, dF(x) \\
							&= \int\sqrt{2 - 2\cos hx} \, dF(x) \rightarrow 0
 			\end{align*} as $h \rightarrow 0$ by the dominated convergence theorem.
		\end{proof}
		\item If $X$ and $Y$ are independent random variables with characteristic functions $\phi$ and $\psi$ respectively, then $X + Y$ has characteristic function \[
			\chi(t) = \phi(t) \cdot \psi(t)
		\]
		\item If $X$ has a characteristic function $\phi$ then $aX + b$ has a characteristic function $e^{itb} \phi(at)$.
		\item If $\phi$ is a characteristic function the so is $| \phi |^2$.  
		\begin{proof}
			Let $X$ and $Y$ have the same distribution, with $X$ independent of $Y$.  Then $Z = X-Y$ has a characteristic function $\phi(t) \phi(-t) = |\phi(t)|^2$.
		\end{proof}
		\item Let $X$ have a MGF $M(t)$.  Then $\phi(t) = M(it)$.
	\end{enumerate}
\end{proof}

\begin{exmp}
	\begin{enumerate}[(i)]
		\item Let $X \sim N(0,1)$.  Then \[\phi(t) = e^{-\frac{1}{2}t^2}.
		\]  
		\item Let $Y \sim N(\mu, \sigma^2)$.  Then \[
			\phi(t) = e^{it\mu - \frac{1}{2} \sigma^2 t^2}.
		\] as $Y = \mu + \sigma Z$ with $Z \sim N(0,1)$.
		\item Let $X \sim \text{Poisson}(\lambda)$  Then \[
			\phi(t) = e^{\lambda(e^{it} - 1)}.
		\]
		\item Let $P(X= 1) = \frac{1}{2} = P(X = -1)$.  Then \[
			\phi(t) = \frac{1}{2}\left(e^{it} + e^{-it} \right) = \cos t. 
		\]
		\item Let $X \sim \text{Exp}(\lambda)$.  Then \begin{align*}
			\phi(t) &= \int_0^\infty e^{itx} \lambda e^{-\lambda x} \, dx \\
					&= \int_0^\infty \lambda e^{-x(\lambda - it)} \, dx \\
					&= \frac{\lambda}{\lambda - it} 
		\end{align*}
	\end{enumerate}
\end{exmp}

\begin{thm}[Parseval's relation]
	Let $F$ and $G$ be distribution functions with associate characteristic functions $\phi$ and $\psi$.  Then \[
		\int e^{-izt} \phi(z) \, dG(z) = \int \psi(x - t) \, dF(x)
	\]
\end{thm}
\begin{proof}
	\begin{align*}
	\int e^{-izt} \phi(z) \, dG(z)
		  	&= \int e^{-izt} \left(\int e^{izt} \, dF(x) \right) \, dG(z) \\
			&= \int \int e^{iz(x-t)} \, dF(x) \, dG(x) \\
			&= \int \left( \int e^{iz(x-t)} \, dG(z) \right) \, dF(x) \quad \text{by Fubini's theorem}\\
			&= \int \psi(x-t) dF(x)
	\end{align*}
\end{proof}

\begin{cor}
	If $G$ is the distribution function of a $N(0, \frac{1}{\sigma^2})$ random variable.  Then $\psi(t) = e^{-\frac{1}{2\sigma^2} t^2}$, and so the above relationship becomes \[
		\int e^{izt} \phi(z) \frac{\sigma}{\sqrt{2\pi}} e^{-\frac{1}{2}z^2 \sigma^2} \, dz = \int e^{-\frac{1}{2\sigma^2}(x-t)^2} \, dF(x).
	\]  Rearranging, we obtain \[
		\frac{1}{2\pi} \int e^{-izt} \phi(z) e^{-\frac{1}{2} z^2 \sigma^2} \, dz = \int \frac{1}{\sqrt{2 \pi \sigma^2}} e^{- \frac{1}{2 \sigma^2}(x - t)^2} \, dF(x)
	\]  Then the right hand side is the density of the convolution of $F$ and a $N(t, \sigma^2)$ distribution.  Call the convolution distribution $F_\sigma$.  Then \begin{align*}
		F_\sigma(\beta) - F_\sigma(\alpha) &= \int_\alpha^\beta \left( 	\frac{1}{2\pi} \int e^{-izt} \phi(z) e^{-\frac{1}{2} z^2 \sigma^2} \, dz \right) \, dt \\
		&= \frac{1}{2\pi} \int \phi(z) e^{-\frac{1}{2} z^2 \sigma^2} \frac{e^{-iz\beta} - e^{-iz\alpha}}{-iz} \, dz
	\end{align*}

	If $\alpha$ and $\beta$ are continuity points of $F$, then \[
		F(\beta) - F(\alpha) = \lim_{\sigma \rightarrow 0} \frac{1}{2 \pi} \int \phi(z) e^{-\frac{1}{2} z^2 \sigma^2} \frac{e^{-iz\beta} - e^{-iz\alpha}}{-iz} \, dz \tag{$\star$}
	\] for as $\sigma \rightarrow 0$, $F_\sigma \rightarrow F$.

	Since a function has only countably many points of discontinuity, we can then derive the following theorem.
\end{cor}

\begin{thm}
	Let $X$ be a random variable with distribution function $F$ and characteristic function $\phi$.  Assume \[
		\int | \phi(t) | \, dt < \infty.
	\]  Then $F$ has a bounded, continuous density $f$ given by \[
		f(x) = \frac{1}{2\pi} \int e^{-itx} \phi(t) \, dt
	\]
\end{thm}

\begin{proof}
	From $(\star)$ apply DCT.  Then \begin{align*}
		F(\beta) - F(\alpha) = F(\beta) - F(\alpha) &= \lim_{\sigma \rightarrow 0} \int_\alpha^\beta \left( \frac{1}{2 \pi} \int \phi(z) e^{-\frac{1}{2} z^2 \sigma^2} \, dz \right) \, dt \\
		&= \int_\alpha^\beta \left( \frac{1}{2 \pi} \int e^{-izt} \phi(z) \, dz \right) \, dt 
	\end{align*}
\end{proof}

\begin{cor}\label{cor:duality}
	If $\phi(t)$ is non-negative and integrable continuous function associated with a distribution function $F$.  Then $\frac{\phi(t)}{2 \pi F'(0)}$ is a density function with characteristic function $\frac{F'(x)}{F'(0)}$.
\end{cor}

\begin{proof}
	We have \begin{align*}
		F'(x) 	&= \frac{1}{2\pi} \int e^{-izx} \phi(z) \, dz \\
				&= \frac{1}{\pi} \int_0^\infty \cos(xz) \phi(z) \, dz \quad \text{as $\phi(z)$ is real}			\end{align*}
	Thus \begin{align*}
		F'(0)	&= \frac{1}{\pi} \int_0^\infty \phi(z) \, dz  \\
		1		&= \frac{1}{2 F'(0) \pi} \int \phi(z) \, dz 
	\end{align*}
	and thus 
	\begin{align*}
		\frac{F'(x)}{F'(0)} &= \int \cos(xz) \frac{\phi(z)}{2 \phi F'(0)} \, dz
	\end{align*}
\end{proof}
% subsection characteristic_functions (end)
% section lecture_13_thursday_14_april (end)
\section{Lecture 14 - Thursday 14 April} % (fold)
\label{sec:lecture_14_thursday_14_april}

% section lecture_14_thursday_14_april (end)

\section{Lecture 15 - Thursday 21 April} % (fold)
\label{sec:lecture_15_thursday_21_april}
\begin{exmp}
	$X$ has density $f(x) = \frac{1}{2} e^{-|x|}$.  Then \begin{align*}
		\phi(t) &= \frac{1}{2} \int e^{itx} e^{-|x|} \, dx \\
		&= \int_0^\infty \cos tx e^{-x} \, dx \\
		&= \int_0^\infty \frac{1}{2} \left( e^{itx} + e^{-itx} \right) e^{-x} \, dx \\
		&= \frac{1}{2} \int_0^\infty e^{-x(1-it)} + ^{-x(1 +it)} \, dx \\
		&= \frac{1}{2} \left[ \frac{-1}{1-it}e^{-x(1+it)} + \frac{-1}{1+it}e^{-x(1+it)} \right]_0^\infty \\
		&= \frac{1}{1+t^2}
	\end{align*}  Thus $\phi(t) = \frac{1}{1+t^2}$ which is a non-negative, integrable characteristic function.  Thus, \[
		\frac{\phi(t)}{2\pi f(0)} = \frac{1}{\pi(1+t^2)} 
	\] which is the Cauchy distribution.  We then know that the characteristic function of the Cauchy distribution is \[
		\gamma(t) = \frac{F'(x)}{F'(0)} = \frac{f(x)}{f(0)} = e^{-|t|}
	\] from Corollary \ref{cor:duality}.
\end{exmp}

\begin{thm}[Moment theorem]
	Let $F$ be the distribution function of $X$.  Assume $X$ has finite moments up to order $n$, i.e. $\E(|X|^n) < \infty$.  Then the characteristic function $\phi(t)$ has uniformly continuous derivatives up to order $n$, and \[
		\phi^{(k)}(t) = i^k \E(|X|^k), k = 1, 2, \dots, n
	\] and \[
		\phi(t) = 1 + \sum_{k=1}^n \E(X^k) \frac{(it)^k}{k!} + o(t^n)
	\] as $t \rightarrow 0$.

	Conversely, if $\phi$ can be written as \[
	\phi(t) = 1 + \sum_{k=1}^n a_k \frac{(it)^k}{k!} + o(t^n)
	\] as $t \rightarrow 0$, then the associated density function has finite moments up to order $n$ if $n$ is even, and up to order $n-1$ if $n$ is odd, with $a_k = \E(|X|^k)$.   
\end{thm}
\begin{proof}

\begin{lem}\label{lem:taylor}
	For any $t \in \R$, \[
		\left|  e^{it} - 1 - it \dots - \frac{(it)^{n-1}}{(n-1)!} \right| \leq \frac{|t|^n}{n!}.
	\]
\end{lem}
\begin{proof}
	Taylor's Theorem.
\end{proof}

Suppose $\E(|X|^k) < \infty$ for $k = 1, 2, \dots, n$.  Then \[
	|x^k e^{itx} | \leq |x|^k\], so \[
		\int x^k e^{itx} \, dF(x)
	\]  exists.  Now \begin{align*}
		\frac{\phi(t+h) - \phi(t)}{h} &= \left| \int \frac{e^{i(t+h)x} - e^{itx}}{h} \, dF(x) \right| \\
		&= \left| \int e^{itx} \cdot \frac{e^{ihx} - 1}{h} \, dF(x) \right| \\
		&\leq \int |x| \, dF(x) < \infty
	\end{align*} from Lemma \ref{lem:taylor}.

	So by DCT, \begin{align*}
		\phi'(t) = \lim_{h \rightarrow 0} \frac{\phi(t+h) - \phi(t)}{h} &= i \int x e^{itx}  \, dF(x)
	\end{align*} and thus \[
		\phi'(0) = i \E(X).
	\]

	Using induction, we obtain \[
		\phi^{(k)}(t) = i^k \int x^k e^{itx} \, dF(x)
	\] and $\phi^{(k)}(0) = i^k \E(X^k)$ for $k = 1, 2, \dots, n$.

	Arguing as in the proof of characteristic functions uniform continuity.

	Expanding $\phi(t)$ about $t = 0$ in a Taylor series, we have \[
		\phi(t) = 1 + \sumnk \phi^{(k)}(0) \frac[t^k]{k!} + R_n(t), t > 0.
	\] with \[
		R_n(t) = \frac{t^n}{n!} \left[ \phi^{(n)}(\theta t) - \phi^{(n)}(0) \right], 0 < \theta < 1.
	\]

	We then have 
	\begin{align*}
		\left| \frac{R_n(t)}{t^n} \right| &\leq \frac{1}{n!} \int |x|^n | e^{i\theta tx} - 1 | \, dF(x) \\
		&\leq \frac{2}{n!} \int |x|^n \, dF(x).
	\end{align*} and so by the DCT, \[
		\lim_{t \rightarrow 0} \left| \frac{R_n(t)}{t^n} \right| = 0, 
	\] and thus $R_n(t) = o(t^n)$.

	Conversely, suppose $\phi$ has an expansion up to order $2k$.  Then $\phi$ has a finite derivative of order $2k$ at $t = 0$  Then \begin{align*}
		-\phi^{(2)}(0) = -\lim_{h \rightarrow 0} \frac{ \phi(h) - 2 \phi(0) - \phi(-h)}{h^2} \\
		&= \lim_{h \rightarrow 0} 2 \int \frac{1 - \cos hx}{x^2} \, dF(x) \\
		&\geq 2 \int \lim_{h \rightarrow 0} \frac{1 - \cos hx}{h^2} \, dF(x) \, \text{by Fatau} \\
		&= \int x^2 dF(x) = E(X^2)
	\end{align*} and so $\phi^{(2)}(0) < \infty \Rightarrow \E(X^2) < \infty$.

	Using induction, assume finite $2(k-1)$\textsuperscript{th} derivative at $0 \Rightarrow \E(X^{2(k-1)}) < \infty$.  Then from the first part, \[
		\phi^{(2(k-2))}(t) = (-1)^{k-1} \int x^{2k-2} e^{itx} \, dF(x) 
	\]  Suppose $\phi^{2k}(0) < \infty$.  Then let \[
		G(x) = \int_{-\infty}^x y^{2k-2} \, dF(y).  
	\] so $\frac{G(x)}{G(\infty)}$ is a distribution function with characteristic function \begin{align*}
	\psi(t) &= \frac{1}{G(\infty)} \int e^{itx} x^{2k-2} \, dF(x) \\
			&= \frac{(-1)^{k-1} \phi^{(2k-2)}(t)}{G(\infty)}
	\end{align*}
	As $\phi^{(2k-2)}(t)$ is twice differentiable at $t = 0$.  So \[
		\psi^{(2)}(0) \geq \int y^2 y^{2k-2} \, \frac{dF(y)}{G(\infty)} 
	\] and thus $\E(X^{2k}) < \infty$.  as required.
\end{proof}
% section lecture_15_thursday_21_april (end)

\section{Lecture 16 Thursday 21 April} % (fold)
\label{sec:lecture_16_thursday_21_april}
	\begin{cor}
		Let $\phi$ be a characteristic function associated with a random variable $X$.  Then $\phi$ has continuous derivatives of all orders if and only if $X$ has finite moments of all orders.  
	\end{cor}

	\begin{cor}
		The function $\phi(t) = e^{-|t|^\alpha}$ is not a characteristic function if $\alpha > 2$.  Note that $\alpha = 1$ was the Cauchy distribution, $\alpha  = 2$ is the Normal distribution.
	\end{cor}

	\begin{proof}
		If $\alpha > 2$ then \[
			\lim_{t \rightarrow 0 } \phi^{(2)}(t) = 0 \Rightarrow \E(X^2) = 0
		\] which implies $X$ is degenerate.  But if $X$ is degenerate at $b$, then \[
			\phi(t) = e^{itb} \neq e^{-|t|^\alpha}
		\]  Thus by uniqueness of characteristic functions, $e^{-|t|^\alpha}$ is not a characteristic function.
	\end{proof}

\subsection{Lattice distributions} % (fold)
\label{sub:lattice_distributions}

% subsection lattice_distributions (end)
\begin{thm}[Lattice distributions]
	Let $X$ be a random variable with distribution function $F$, characteristic function $\phi$.  If $c \neq 0$ then the following are equivalent.
	\begin{enumerate}[(i)]
		\item $X$ has a lattice distribution whose range is continued in $0, \pm b, \pm 2b, \dots$, $b = \frac{2 \pi}{c}$.
		\item $\phi(t + nc) = \phi(t)$ for $n = \pm 1, \pm 2, \dots$, that is, $\phi$ is periodic with period $c$. 
		\item $\phi(c) = 1$.  
	\end{enumerate}
\end{thm}
\begin{proof}
	$(1) \Rightarrow (2)$.
	\begin{align*}
		\phi(t) &= \sum_{k=-\infty}^\infty P(X = kb) e^{itkb} \\
		&= \sum_{k=-\infty}^\infty P(X=kb) e^{2\pi i t k /c} \\
	\end{align*} which implies \[
		\phi(t + nc) = \phi(t)
	\] as $e^{2 \pi i n c k /c} = 1$.

	$(2) \Rightarrow (3)$.  Simply set $t = 0, n = 1$.  Then $\phi(0) = \phi(c) = 1$.

	$(3) \Rightarrow (1)$.  \begin{align*}
		1 - \E(\cos cX) &= 0 \\
		\E(1 - \cos cX) &= 0 
	\end{align*}
	but as $1 - \cos cX \geq 0$, $X$ must have probability components on points where $\cos cX = 1$, that is, $c X$ takes on the values $0, \pm \pi, \pm 2 \pi, \dots$.  
\end{proof}

\begin{cor}
	$X$ is degenerate if and only if $| \phi(t) | = 1$ for all $t$. 
\end{cor}

\begin{proof}
	If $P(X = b) = 1$, then $\phi(t) = e^{itb},$ and so $| \phi(t) | = 1$ for all $t$.

	If $|\phi(c)| = 1$ for $ c \neq 0$, then $\phi(c) = e^{i\theta}$ for some $\theta$.  Let $\phi_1(t) = \phi(t) e^{-i \theta t/c}$ is characteristic function of $X - \frac{\theta}{c}$.  Then $\phi_1(c) = 1$, thus $X- \frac{\theta}{c}$ is a lattice taking values in $0, \pm \frac{2\pi}{c}, \pm \frac{4\pi}{c}, \dots$.

	Now, pick some $b \in \R$ with $\frac{b}{c}$ irrational.  Then $|\phi(b) | = 1$, and then $X - a_2$ is a lattice taking values in $0, \pm \frac{2 \pi}{b}, \pm \frac{4 \pi}{b}, \dots$.  Then \begin{enumerate}[(i)]
		\item $|\phi(t)| < 1$ for $t \neq 0$ (e.g. Normal, $e^{-\frac{1}{2}t^2}$).
		\item $|\phi(\lambda) | = 1$ and $|\phi(t)| < 1$ on $0 < t < \lambda$ (e.g. discrete $\pm 1$, $\cos t$).
		\item $|\phi(t)| = 1 \forall t$, degenerate distributions.  
	\end{enumerate}
\end{proof}

\begin{exmp}
	We can construct 3 nontrivial distribution functions $\phi_1, \phi_2, \phi_3$ such that 
\begin{enumerate}[(i)]
	\item $\phi_1(t) = \phi_2(t), \forall t \in [-1,1]$.  
	\item $|\phi(t)| = | \phi_3(t) |, \forall t$.  
\end{enumerate}

Consider $g(x) = 1 - | x |, x \in [ -1, 1]$.  This has characteristic function $\phi(t) = \frac{2(1 - \cos t)}{t^2}$.  But the characteristic function is positive and integrable, and so\[
	\phi_1(t) = \begin{cases}
		1 - |t| & |t| \leq 1 \\
		0		& |t| > 1
	\end{cases}
\] is the characteristic function of the density \[
	f(x) = \frac{1 - \cos x}{\pi x^2}.
\]  We can express $\phi_1(t)$ as the trigonometric series, \begin{align*}
	\phi_1(t) = 1 - |t| = \frac{1}{2} + \sum_{k=1}^\infty a_k \cos( k \pi t) 
\end{align*} with \[
	a_k = 2 \int_0^1 ( 1 - t) \cos(k \pi t) \, dt = \begin{cases} 
		\frac{4}{k \pi^2} 	&\text{$k$ odd} \\
		0					&\text{$k$ even}
	\end{cases}
\]
We can thus write \[
	\phi_1(t) = \frac[1]{2} + \sum_{k=1}^\infty \frac{4}{(2k-1)^2 \pi^2} \cos(2k-1) \pi t.
\]

Let $V$ be a random variable, with \[
	P(V = 0) = \frac{1}{2}, P(V = \nu) = \frac{2}{\nu^2}, \nu = \pm \pi, \pm 3 \pi, \pm 5 \pi, \dots
\]    Then $V$ is a lattice distribution, with characteristic function \[
	\phi_2(t) = \frac{1}{2} + \frac{4}{\pi^2} \left( \cos \pi t + \frac{ \cos 3 \pi t}{9} + \frac{\cos 5 \pi t}{25} + \dots \right) 
\] and thus $\phi_1(t) = \phi_2(t)$ on $[-1,1]$, but have different density functions.

Finally, let $U$ be a lattice random variable with distribution \[
	P(U = \pm \frac{(2k+1) \pi}{2}) = \frac{4}{\pi^2(2k+1)^2}, k = 0,1, 2, \dots
\]  Then $U$ has a characteristic function $\phi_3(t) = 2 \left[ \phi_2(\frac{t}{2}) - \frac{1}{2} \right]$.  Thus \[
	| \phi_3(t) | = | \phi_2(t) | \quad \forall t.
	\]
\end{exmp}
% section lecture_16_thursday_21_april (end)

\section{Lecture 17 - Thursday 5 May} % (fold)
\label{sec:lecture_17_thursday_5_may}
\subsection{Sequences of characteristic functions} % (fold)
\label{sub:sequences_of_characteristic_functions}

% subsection sequences_of_characteristic_functions (end)
\begin{lem}[Helly selection theorem]\label{lem:hellysel}
	Given a sequence of distribution functions $\{ F_n \}$ then there exists a sequence $\{ n_k \}$ and a non decreasing right continuous function $F$ such that \[
		F_{n_k}(x) \rightarrow F(x)
	\] at all continuity points $x$ of $F$.  
\end{lem}
\begin{proof}
	First order the rationals to get a sequence $\{ r_k \}$.  From $\{ F_n(r_1)\}$ we choose a subsequence $\{ F_{n_{1k}}(r_1) \}$ which converges.

	Now from the sequence $\{ n_{1k} \}$ choose a subsequence $\{ n_{2k} \}$ such that $\{ F_{n_{2k}}(r_2) \}$ converges, etc.

	Now let $n_k = n_{kk}$.  Then for each rational number $r$, the limit $F_{n_k}(r)$ exists as $n \rightarrow \infty$.  Define $L(R) = \lim F_{n_k}(r), r \in \Q$.  Then $L(r)$ is non-decreasing and takes values in $[0,1]$.  Let $F(x) = \inf_{r \leq x} L(r)$.  Then $F$ is non-decreasing, and right continuous, and $F_{n_k}(x) \rightarrow F(x)$ for all $x \in \Q$ and at all points of continuity of $F$.  
\end{proof}

\begin{lem}[Extended Helly-Bragg theorem] \label{lem:extendedhelly}
	If a sequence of distribution functions $\{ F_n \}$ converges to a function $F$ at all continuity points of $F$ and $g$ is a \textbf{bounded, continuous, real valued function} then \[
		\int_\R g \,dF_n \rightarrow \int_\R g \, dF
	\]
\end{lem}
\begin{proof}
	Let $M = \sup_x | g(x) |$, and let $a,b$ be continuity points of $F$.  Then 
	\begin{align*}
		\left| \int_\R g \, dF_n - \int_\R g \, dF \right| &\leq \left| \int_\R g \, dF_n - \int_a^b g \, dF_n \right| + \left| \int_a^b g \, dF_n - \int_a^b g \, dF \right| + \left| \int_a^b g \, dF - \int_\R g \, dF \right| \\
		&\leq M[F_n(a) - F_n(-\infty) + F_n(\infty) - F_n(b)] + \left| \int_a^b g \, dF_n - \int_a^b g \, dF \right| \\
		&\qquad  + M[F(a) - F(-\infty) + F(\infty) - F(b)] 
	\end{align*} Since \[
		F_n(a) \rightarrow F(a), F_n(b) \rightarrow F(b)
	\] as $a, b$ are continuity points, we can choose $a, b$ large enough to make the 3rd term small $(< \frac{\epsilon}{3}$ for arbitrary $\epsilon > 0$), and then $N$ large enough to make the first term small.

	Now we deal with the middle term.  Let $a = x_{0N} < x_{1N} < \dots < x_{\nu_N, N} = b$ be a sequence of subdivisions of $[a,b]$, such that $\Delta_{n} \rightarrow 0$ (partition width) as $n \rightarrow \infty$.  Then \[
		g_N(x) = \sum_{\nu = 1}^{\nu_N} g(x_{\nu}, N) \indic{x_{\nu-1, N} \leq x \leq x_{\nu, N}} 
	\]  Then $\sup_{x \in [a,b]} | g_N(x) - g(x)| \rightarrow 0$ as $N \rightarrow \infty$ (as $g$ is bounded and continuous.) Then by DCT we have 
	\begin{align*}
		\int_a^b g \, dF_n &= \lim_{N \rightarrow \infty} \int_a^b g_N \, dF_n \\
		\int_a^b g \, dF &= \lim_{N \rightarrow \infty} \int_a^b g_N \, dF
	\end{align*}   Next, we will show \[
		\lim_{ n \rightarrow \infty} \int_a^b g_N \, dF_n = \int_a^b g_N \, dF
	\]  Let $x_{\nu, N}$ be continuity points of $F$ so \[
		F_n(x_{\nu, N}) - F_n(x_{\nu - 1, N}) \rightarrow F(x_{\nu, N}) - F(x_{\nu - 1, N}).
	\] Hence
	\begin{align*}
		\lim_{n \rightarrow \infty} \int_a^B g_N(x) \, dF_n &= \lim_{n \rightarrow \infty} \sum_{\nu = 1}^{\nu_N} g(x_{\nu, N}) (F_n(x_{\nu, N}) - F_n(x_{\nu - 1, N})) \\
		&= \int_a^b g_N(x) \, dF(x) 
	\end{align*} If $M_N = \sup_{x \in [a,b]} | g_N(x) - g(x) |$, then \begin{align*}
		\left| \int_a^b g \, dF_N - \int_a^b g \, dF \right| &\leq \int_a^b |g - g_n | \, dF_n + \left| \int_a^b g_n \, dF_n - \int_a^b g_N \, dF \right| + \int_a^b |g - g_N | \, dF \\
		&\leq M_N [ F_n(b) - F_n(a)] + \left| \int_a^b g_N \, dF_n  \right.\\
		&\qquad \left. \int_a^b g_N \, dF \right| + M_N[F(b) - F(a)] 
	\end{align*} Since $M_N \rightarrow 0$ as $N \rightarrow \infty$.  Then choosing $N$ large enough to make $M_N$ small enough, for a large $N$ fixed, $N_2$ say, we have \[
		\left| \int_a^b g_{N_2} \, dF_n - \int_a^b g_{N_2} \, dF \right| \leq \frac{\epsilon}{9}
	\]  The result then follows.
\end{proof}

\begin{lem}
	Let $\{ F_n \}$ be a sequence of distribution functions with associated characteristic function $\{ \phi_n \}$.  Assume $\phi_n(t) \rightarrow \phi(t)$ as $n \rightarrow \infty$ for all $t \in \R$.  Then there exists a non-decreasing right continuous function $F$ such that $F_n(x) \rightarrow F(x)$ at all continuity points $x$ of $F$. 
\end{lem}
\begin{proof}
	From Lemma \ref{lem:hellysel} there exists a subsequence $\{ n_k \}$ and a non-decreasing continuous function $F$ such that $F_{n_k} (x) \rightarrow F(x)$ at all continuity points of $F$.  Using Parseval's relation on $\{ F_{n_k}, \phi_{n_k} \}$, we have \begin{align*}
		\frac{1}{2 \pi} \int_\R e^{-izt} \phi_{n_k}(z) e^{- \frac{1}{2} \sigma^2 z^2} \, dz = \int_\R \frac{1}{\sqrt{2 \pi}} e^{- \frac{1}{2} \left( \frac{x-t}{\sigma} \right)^2} \, dF_{n_k}(x)
	\end{align*}  Let $k \rightarrow \infty$.  Then the LHS becomes \[
		\frac{1}{2 \pi} \int_\R e^{-i z t} \phi(z) e^{- \frac{1}{2} \sigma^2 z^2} \, dz 
	\] by the dominated convergence theorem.

	The RHS becomes \[
		\int_\R \frac{1}{\sqrt{2 \pi}} e^{- \frac{1}{2} \left( \frac{x-t}{\sigma} \right)^2} \, dF(x)
	\] by an application of Lemma \ref{lem:extendedhelly}.  Thus $\phi$ determines $F$ uniquely (as before), so the limit $F$ must be the same for all convergent subsequences.  
\end{proof}

\begin{thm}[Continuity theorem]
	Let $\{ F_n \}$ be a sequence of distribution functions converging to a distribution function $F$ at all continuity points $x$ of $F$.  This happens if and only if $\phi_n(t) \rightarrow \phi$ pointwise and $\phi$ is continuous in the neighbourhood of the origin.  If this is the case then $\phi$ is the characteristic function associated with $F$, and is continuous everywhere.
\end{thm}

\begin{proof}
	If $\{ F_n \}$ converges to $F$, use Lemma \ref{lem:extendedhelly}, with $g(x) = \cos(xt) + \sin(xt)$.
\end{proof}
% section lecture_17_thursday_5_may (end)

\section{Lecture 18 - Thursday 12 May} % (fold)
\label{sec:lecture_18_thursday_12_may}

\begin{thm}
	Assume $F_n \rightarrow F$ at continuity points of $F$, and associated characteristic function $\phi_n \rightarrow \phi$ pointwise.  If $\phi_n \rightarrow \phi$ and $\phi$ is continuous in a neighbourhood of 0, then $F_n \rightarrow F$ and $F$ is distribution function associated with $\phi$.
\end{thm}

\begin{proof}
	From previous lemma, there exists a non-decreasing, right continuous non-negative function $F$ such that $F_n \rightarrow F$.  We need to show $F$ is a distribution function, that is $F(+ \infty) - F(-\infty) \geq 1$.  By Parseval's relation, we have \[
		\frac{\sigma}{\sqrt{2\pi}} \int_\R e^{-izt} \phi(z) e^{-\frac{1}{2}\sigma^2 z^2}\, dz = \int_\R e^{-\frac{1}{2}\left(\frac{x-t}{\sigma}^2\right)} \, dF(x) \leq F(+\infty) - F(-\infty)
	\]  The left hand side is equal to \begin{align*}
		\E(e^{-i N_\sigma t} \phi(N_\sigma))
	\end{align*} where $N_\sigma \sim N(0, \frac{1}{\sigma^2})$.  Since \[
		\left|e^{-izt} \phi(t) \right| \leq 1
	\] Assume $\phi$ is continuous on $|t| < A$.  Then \begin{align*}
		\E(e^{-iN_\sigma t} \phi(N_\sigma)) &= \E(e^{-iN_\sigma t} \phi(N_\sigma) \given |N_\sigma| \geq A ) \cdot P(|N_\sigma| \geq A) \\
		 				&\qquad + \E(e^{-iN_\sigma t} \phi(N_\sigma) \given |N_\sigma| < A) P(|N_\sigma| < A).
	\end{align*}  The first term tends to zero as $\sigma \rightarrow \infty$, as $P(|N_\sigma| \geq A) \rightarrow 0$ on $|N_\sigma| < A$.  Then the distribution function tends to \[
		G(x) = \begin{cases}
			0 	&x < 0 \\
			1	&x \geq 0 
		\end{cases}
	\] as $\sigma \rightarrow \infty$.

	\[
		\lim_{\sigma \rightarrow \infty}\frac{\sigma}{\sqrt{2\pi}} \int_\R e^{-izt} \phi(z) e^{-\frac{1}{2}\sigma^2 z^2}\, dz = \int_\R e^{-izt} \phi(z) \, dG(z) = 1
	\] by the extended Helly-Bragg theorem.
 \end{proof}

\begin{cor}
	If $X_n$ has distribution function $F_n$ and characteristic function $\phi_n$, and $X$ has distribution function $F$ and characteristic function $\phi$.  Then the following are equivalent.
	\begin{enumerate}[i)]
		\item $F_n(x) \rightarrow F(x)$ at all continuity points $x$ of $F$.  
		\item $\phi_n(t) \rightarrow \phi(t)$ for all $t$, 
		\item $\E(g(X_n)) \rightarrow \E(g(X))$ for all real, bounded, continuous functions $g$. 
	\end{enumerate}
	In these cases we write $X_n \cid X$ ($X_n$ converges in distribution to $X$)
\end{cor}

\begin{cor}
	Suppose $X_n \cid X$.  If $h$ is any continuous real valued function, then $h(X_n) \cid h(X)$.
\end{cor}
\begin{proof}
	$X_n \cid X$ if and only if $\E(g(X_n)) \rightarrow \E(g(X))$.  Then $g(h(x))$ is real, bounded, and continuous.  Then \[
		\E(g(h(X_n))) \rightarrow \E(g(h(X))) \Rightarrow h(X_n) \cid h(x)
	\] for all $g$ real, bounded, continuous.
\end{proof}

\begin{thm}[Slutsky's theorem]
	IF $X_n \cid X$ and $Y_n \cip a$, then \[
		X_n + Y_n \cid X + a
	\]
\end{thm}
\begin{proof}
	Given $\epsilon > 0$, choose $x$ such that $x, x - a \pm \epsilon$ are continuity points of $F(x) = P(X \leq x)$.  Then \begin{align*}
		P(X_n + Y_n \leq x) &= P(X_n + Y_n \leq x, |Y_n - a | > \epsilon) + P(X_n + Y_n \leq x, |Y_n - a| \leq \epsilon) 		\\
		&\leq P(|Y_n - a | > \epsilon) + P(X_n \leq x - a + \epsilon) \\
		P(X_n \leq x - a - \epsilon) &= P(X_n \leq x - a - \epsilon, |Y_n - a| > \epsilon) + P(X_n \leq x - a - \epsilon, |Y_n - a | \leq \epsilon) \\
		&\leq P(|Y_n - a| > \epsilon) + P(X_n + Y_n \leq x)
	\end{align*}  Taking limits as $n \rightarrow \infty$, we have \[
		P(X \leq x - a - \epsilon) \leq \lim_{n \rightarrow \infty} P(X_n + Y_n \leq x) \leq P(X \leq x - a + \epsilon)  
	\]  Since $x - a \pm \epsilon$ are continuity points of $F$, we have \[
		\lim_{n \rightarrow \infty} P(X-n + Y_n \leq x) = P(X \leq x - a).\qedhere
	\]
\end{proof}
\subsection{Central limit theorem} % (fold)
\label{sub:central_limit_theorem}

% subsection central_limit_theorem (end)
\begin{note}[Notation]
	Let $X_1, X_2, \dots$ are independent random variables with characteristic functions $\phi_1, \phi_2, \dots$ and distribution functions $F_1, F_2, \dots$.  Let $\E(X_i) = 0, \var(X_i) = \sigma_i^2 < \infty$, $i = 1, 2, \dots$.  Let \[
		S_n = \sum_{i=1}^n X_i, \qquad s_n^2 = \var(S_n) = \sum_{i=1}^n \sigma_i^2
	\] 
\end{note}

\begin{thm}[Lindeberg conditions]
	Let $\epsilon > 0$.  Then \begin{align*}
		L_n(\epsilon) &= \frac{1}{s_n^2} \sum_{i=1}^n \E \left(X_i^2 \indic(|X_i| > \epsilon s_n) \right)  \\
		&= \frac{1}{s_n^2} \sumni \int_{|x| > \epsilon s_n} x^2 \, dF_i(x) 
	\end{align*}  Then the \textbf{Lindeberg condition} is \[
		\forall \epsilon > 0, \quad L_n(\epsilon) \rightarrow 0 \quad \text{as $n \rightarrow \infty$}
	\]
\end{thm}
\begin{exmp}
	Assume $\E(|X_i|^3) < \infty$.  Then \begin{align*}
		L_n(\epsilon) &\leq \frac{1}{s_n^2} \sumni \E(X_i^2 \frac{|X_i|}{\epsilon s_n}) \\
		 	&= \frac{1}{\epsilon} \frac{1}{s_n^3} \sumni \E(|X_i|^3)
	\end{align*}

\end{exmp}

\begin{thm}[Liapounov's condition]
	\[
		\frac{1}{s_n^3} \sumni \E(|X_i|^3) \rightarrow 0 \quad \text{as $n \rightarrow \infty$}
	\]  From above, Liapounov's condition implies Lindeberg's condition.
\end{thm}

\begin{thm}[Central limit theorem]
	If for all $\epsilon > 0$, $L_n(\epsilon) \rightarrow 0$ as $n \rightarrow \infty$, then \[
		\frac{S_n}{s_n} \cid N(0, 1)
	\]
\end{thm}
\begin{proof}
	Preliminaries.
	\begin{enumerate}[(i)]
		\item If $|a_k| <\leq 1$ and $|b_k| \leq 1$ for all $k$, then \[
			\left| \prod_{i=1}^n a_i - \prod_{i=1}^n b_i \right| \leq \sum_{i=1}^n |a_i - b_i| 
		\] as $a_1 a_2 - b_1 b_2 = (a_1 - b_1) a_2 - b_1(a_1 - b_2)$ and use induction.  
		\item $\left|e^z - 1 - z \right| \leq \delta |z|, \delta > 0$, for $|z|$ sufficiently small.   
	\end{enumerate}

	It is sufficient to prove \[
		\phi_{S_n/s_n}(t) = \prod_{k=1}^n \phi_k(t/s_n) \rightarrow e^{-\frac{1}{2}t^2} \tag{$\ddag$})
	\] for all $t$.

	Now \begin{align*}
		|\phi_k(t/s_n) - 1| &= \left| \int (e^{\frac{itx}{s_n}} - 1 - \frac{itx}{s_n})dF_k(x) \right| \quad \text{as $\E(X_k) = 0$} \\
		&\leq \int \frac{t^2}{x^2}{2 s_n^2} \, dF_k(x) \\
		&= \frac{1}{2}\frac{\sigma_k^2}{s_n^2}t^2 \tag{$\star$}
	\end{align*}  Now \begin{align*}
		\sigma_k^2 &= \E(X_k^2 \indic{|X_k| \leq u s_n}) + \E(X_k^2 \indic{|X_k| >  u s_n}) \\
		&\leq (u s_n)^2 + s_n^2 L_n(u) 
	\end{align*} Hence \[
		\frac{\sigma_k^2}{s_n^2} \leq u^2 + L_n(u)
	\] and since there are no $k$ on the RHS, we have \[
		\max_{k \leq n} \frac{\sigma_k^2}{s_n^2} \leq u^2 + L_n(u)
	\]  By Lindenberg's condition, we have $L_n(u) \rightarrow 0$ as $n \rightarrow \infty$, and as $u$ was arbitrary, we have \[
		\max_{k \leq n} \frac{\sigma^2_k}{s_n^2} \rightarrow 0
	\] From Assignment 5, we know \[
		\exp(\phi_k(t) - 1)
	\] is a characteristic function.  Let $\delta \rightarrow 0$.  Then \begin{align*}
		\left| \exp(\sum_{k=1}^n (\phi_k(t/s_n)) - 1) - \prod_{k=1}^n \phi_k(t/s_n) \right| &\leq \sum_{k=1}^n \left| e^{\phi_k(t/s_n) - 1} - \phi_k(t/s_n) \right| \quad \text{by (i)} \\
		&\leq \delta \sum_{k=1}^n \left|\phi_k(t/s_n) - 1 \right| \quad \text{by (ii)} \\
		&\leq \frac{\delta t^2}{2}{\sum_{k=1}^n \frac{\sigma_k^2}{s_n^2}} \quad \text{by ($\star$)} \\
		&= \frac{\delta t^2}{2}. \quad \text{if $n$ is sufficiently large}
	\end{align*}

	By $(\ddag)$, we must show \begin{align*}
		\sum_{k=1}^n (\phi_k(t/s_n) - 1) + \frac{1}{2} t^2 \rightarrow 0
	\end{align*} that is, \[
		\sum_{k=1}^n \int \left( e^{itx/s_n} - 1 - \frac{itx}{s_n} + \frac{1}{2} \frac{t^2 x^2}{s_n^2} \right)\, dF_k(x) \rightarrow 0  \tag{$\dagger$}
	\]  The modulus of the integral in ($\dagger$) is bounded by \[
		\frac{1}{6} \left|\frac{tx}{s_n} \right|^3 \leq u \frac{|t|^3 x^2}{6s_n^2}
	\] if $|x| \leq u s_n$ and \[
		\frac{x^2 t^2}{2s_n^2} + \frac{x^2 t^2}{2s_n^2}
	\] when $|x| > u s_n$.  Hence the integral of $(\dagger)$ is bounded above by \[
		\frac{u|t|^3}{6} + \frac{t^2}{s_n^2} \sum_{k=1}^n \int_{|x| > u s_n} x^2 \, dF_k(x) = \frac{u |t|^3}{6} + L_n(u) t^2
	\] as the integral is the Lindeberg's condition.

	Given $t, \epsilon > 0$, choose $u$ such that $\frac{u |t|^3}{6} < \frac{\epsilon}{2}$, and $N_0$ large enough such that $L_n(u) t^2 < \frac{\epsilon}{2}$ for $n > N_0$.  So the left hand side of $(\dagger)$ is bounded above by $\epsilon$, and so the result follows.  
\end{proof}

\begin{thm}[Partial converse of the central limit theorem]
	Suppose that $s_n \rightarrow \infty$ and $\frac{\sigma_n}{s_n} \rightarrow 0$ as $n \rightarrow \infty$.  Then the Lindeberg condition is necessary for \[
		\frac{S_n}{s_n} \cid N(0, 1).
	\]
\end{thm}

\begin{proof}
	By assumption, given $\epsilon > 0$ there exists $N_1 > 0$ such that \[
		\frac{\sigma_k}{\sigma_n} < \frac{\sigma_k}{\sigma_k} < \epsilon
	\] for $N_1 \leq k \leq n$ as $s_n^2 \leq s_k^2 (k \leq n)$.  We also have \[
		\frac{\sigma_k}{s_n} < \epsilon, k = 1, 2, \dots, N_1
	\] for $n > N_1$ as $s_n^2 \rightarrow \infty$.  Hence \[
		\max_{1 \leq k \leq n} \frac{\sigma_k}{s_k} \rightarrow 0
	\] as $n \rightarrow \infty$.  Assume $\frac{S_n}{s_n} \cid N(0, 1)$.  If $(5)$ holds then this convergence is equivalent to $(1) \iff (3) ( \Rightarrow (4))$ as $(1) \iff (3)$ requiring $(5)$, to ensure \[
		\left|\phi_{k}(\frac{t}{s_n}) - 1 \right|
	\]  can be made uniformly small.

	The real part of $(4)$, \begin{align*}
		\sum_{k=1}^n \int \left( \cos(\frac{xt}{s_n}) - 1 + \frac{x^2 t^2}{2s_n^2} \right) \, dF_k(x)
		&\geq \sum_{k=1}^n \int_{|x > u s_n} \left( \cos(\frac{xt}{s_n}) - 1 + \frac{x^2 t^2}{2s_n^2} \right) \, dF_k(x)
	\end{align*}  For any $u > 0$, choose $t$ such that $\frac{x^2 t^2}{2 s_n^2} - 2 > 0$ if $|x| > u s_n$ (i.e. $t^2 > \frac{4}{n^2}$).  Continuing, we have \begin{align*}
		&\geq \sum_{k=1}^n \int_{|x| > u s_n} \left( \frac{x^2 t^2}{2 s_n^2} - 2 \right) \, dF_k(x) \\
		&\geq \sum_{k=1}^n \int_{|x| > u s_n} \left(\frac{x^2 t^2}{2s_n^2} - 2\frac{x^2}{u^2 s_n^2} \right) \, dF_k(x) \\
		&= \left(\frac{t^2}{2} - \frac{2}{u^2}\right) \frac{1}{s_n^2} \sum_{k=1}^n \int_{|x| > u s_n} x^2 \, dF_k(x) \\
		&= \left(\frac{t^2}{2} - \frac{2}{u^2}\right) L_n(u)
	\end{align*} Thus $L_n(u) \rightarrow 0$ as $n \rightarrow \infty$, that is, Lindeberg's condition holds.
\end{proof}

\begin{cor}
	Let $X_1, X_2, \dots$ IID with $\E(X_1) = 0, \var(X_1) = \sigma^2$.  Then \[
		\frac{S_n}{\sqrt{n \sigma^2}} \cid N(0, 1)
	\] Let $\bar X_k = \frac{S_n}{n}$.
\end{cor}

\begin{proof}
	 We have $s_n^2 = n \sigma^2$.  For $\epsilon > 0$, we have \begin{align*}
		L_n(\epsilon) &= \frac{1}{n \sigma^2} \sum_{k=1}^n \E( X_k^2 \indic{|X_k| > \epsilon \sigma \sqrt{n}}) \\
		&= \frac{1}{\sigma^2} \E(X_1^2 \indic{|X_1| > \epsilon \sigma \sqrt{n}}) \rightarrow 0
	\end{align*} as $n \rightarrow \infty$ as $\E(X_1^2) < \infty$.
\end{proof}
% section lecture_18_thursday_12_may (end)

\section{Lecture 19 - Thursday 19 May} % (fold)
\label{sec:lecture_19_thursday_19_may}
The central limit theorem is about distribution functions.  It is not an automatic consequence that the derivatives (densities) converge.

If $\frac{S_n}{s_n}$ has density $f_n(x)$ we need further conditions to ensure $f_n(x) \rightarrow \frac{1}{\sqrt{2 \pi}} e^{- \frac{1}{2} x^2}$ as $n \rightarrow \infty$.

\begin{thm}
	If $X_i$ are $IID$ with characteristic functions $\phi(t)$ and $|\phi(t)|$ is integrable then \[
		f_n(x) \rightarrow \frac{1}{\sqrt{2 \pi}} e^{-\frac{1}{2} x^2}
	\]
\end{thm}

\begin{exmp}[Densities not converging]
	Let $X_i$ have density \[
		f(x) = \frac{C}{x (\log x)^2}, \quad 0 < x < \frac{1}{2}.
	\] Then $\E(X^2) < \infty$ but $\sum_{i=1}^n X_i$ has an unbounded density on $(0, \frac{1}{2})$.
\end{exmp}

\subsection{Stable Laws} % (fold)
\label{sub:stable_laws}
\begin{defn}[Stable distribution]
	A distribution $F$ is said to be stable if it is not concentrated at one point, and when $X_1$ and $X_2$ are independent with distribution function $F$ and $a_1, a_2$ are arbitrary constants there exists some $\alpha > 0, \beta$ such that \[
		\frac{\alpha_1 X_1 + \alpha X_2 - \beta}{\alpha}
	\] has distribution function $F$.  	
\end{defn}

\begin{exmp}
	If $X_1$ has a characteristic function $\phi(t)$ then \begin{align*}
		\alpha X_3 + \beta &= a_1 X_1 + a_2 X_2 \\
		e^{i \beta t} \phi(\alpha t) &= \phi(a_1 t) \phi(a_2 t)
	\end{align*}  If $\phi(t) = e^{-c|t|^\gamma}$, $0 < \gamma \leq 2$, then \[
		\phi(a_1 t) \phi(a_2 t) = e^{-c\left( |a_1|^\gamma + |a_2|^\gamma \right) |t|^\gamma}.
		\]

	As these distributions are symmetric, we have $\beta = 0$, and so setting $\alpha = (|a_1|^\gamma + |a_2|^\gamma)$.  Thus distributions with characteristic functions of the form $e^{-c|t|^\gamma}$ are stable.  Hence the Cauchy distribution is stable ($\gamma = 1$), and the normal distribution is stable $(\gamma = 2)$.	
\end{exmp}

\begin{thm}
	If $\phi$ is the characteristic function of a symmetric random variable $(X \eqd -X)$ with a stable distribution then $\phi(t) = e^{-c|t|^\gamma}$ for some $c > 0$, $\gamma \in (0, 2]$.

	Recall that a distribution is symmetric if and only if $\phi$ is real.
\end{thm}
\begin{proof}[Partial]
	$\phi(t) \phi(t) = \phi(\alpha t)$ used to show that $\phi(t) \neq 0$.  (Since $\phi(0) = 1$ and $\phi(t)$ is continuous).

	Then build up properties of $\phi$.
\end{proof}

\begin{thm}[L\`evy]
	Let $X_1, X_2, \dots$ be independent and identically distributed random variables with distribution functions $G$.  Let $S_n = \sum_{i = 1}^n X_i$.  
	Suppose that there exists a sequence of constants $(a_n, b_n)$ with $b_n > 0$, such that \[
		\frac{S_n - a_n}{b_n} \cid X
	\] where $X$ is not a constant.   Then $X$ is stable.
\end{thm}

\begin{defn}[Domain of attraction]
	If $X$ has distribution function $F$ then we say $G$ is in the \textbf{domain of attraction} of $F$.    
\end{defn}
\begin{cor}
	If $X$ has finite variance then $G$ is in the domain of attraction of the normal distribution.  
\end{cor}

\begin{cor}
	If $G$ satisfies $\lim_{x \rightarrow \infty} x(1 - G(x)) = c > 0$ then $G$ is in the domain of attraction of the Cauchy distribution, that is, \[
		x\,\P(X > x) \rightarrow c.
	\]  A necessary and sufficient condition to be in the domain of attraction for the Cauchy distribution is \[
		1 - G(x) = P(X_1 > x) = \frac{L(x)}{x}
	\] where $L(x)$ is a \textbf{slowly varying function}.  $L(x)$ is a slowly varying function if for all $C > 0$, \[
		\lim_{x \rightarrow \infty} \frac{L(C x)}{L(x)} = 1.
	\]  For example, $L(x) = 1, L(x) = \log x$ are slowly varying functions.
\end{cor}

\begin{thm}
	All stable laws are absolutely continuous and the distribution functions have derivatives of all orders.
\end{thm}

\begin{thm}
	The normal distribution is the only stable law with finite variance.
\end{thm}

\begin{thm}
	It can be shown that the canonical form of the characteristic function of a stable law is \[
		\phi(t) = \exp\left[ i \gamma t - c|t|^\gamma \left\{ 1 + \frac{i \beta t}{|t|} \omega(t, \alpha)\right\}\right]
	\]  where \begin{align*}
		\gamma \in \R, \alpha \in (0, 2], c > 0, |\beta | \leq 1, \omega(t, \alpha) = \begin{cases}
			\tan \frac{\pi \alpha}{2} & \alpha \neq 1 \\
			\frac{2}{\pi} \log |t|	&\alpha = 1
		\end{cases}
	\end{align*}

	If $\phi$ is real, then $\beta = \gamma = 0$. 
\end{thm}
% subsection stable_laws (end)  
% section lecture_19_thursday_19_may (end)

\section{Lecture 20 - Thursday 19 May} % (fold)
\label{sec:lecture_20_thursday_19_may}
\subsection{Infinitely divisible distributions} % (fold)
\label{sub:infinitely_divisible_distributions}

% subsection infinitely_divisible_distributions (end)
Consider a triangular array $\{ X_{nk} \}_{k=1}^n$ where for each $n$, $X_{n1}, X_{n2}, \dots, X_{nn}$ are independent random variables.  We assume that the distribution are identically distributed for each $n$.\[  \begin{matrix}
	X_{11} &  		& \\
	X_{21} & X_{22} &	\\
	X_{31} & X_{32} & X_{33} \\
	\vdots &		& 		&\ddots
\end{matrix}
\]

\begin{exmp}
	Let $X_{nk} \sim B(1, p_n)$.  Then $S_n = \sum_{k=1}^n X_{nk} \sim B(n , p_n)$.  We know that if $n p_n \rightarrow \lambda$ as $n \rightarrow \infty$, then \[
		S_n \cid \textsc{Poisson}(\lambda).
	\]

	Note that the Poisson distribution is not continuous, nor is it stable.  Consider $X_1, X_2$ Poisson distributed, and let $Y = 2X_1 + 3X_2$.  Then $Y$ is not in the Poisson family as $P(Y = 1) = 0$.  
\end{exmp}

\begin{defn}[Infinitely divisible]
	A distribution function $F$ is infinitely divisible if for every positive integer $k$, $F$ is the $k$-fold convolution of some distribution $G_k$ with itself.  
\end{defn}

\begin{exmp}
	\begin{enumerate}[(1)]
		\item The Poisson distribution is infinitely divisible, as \[
			\phi(t) = e^{\lambda(e^{it} - 1)} = \left[ e^{\frac{\lambda}{k}(e^{it} - 1)} \right]^k
		\]
		\item Symmetric stable laws are infinitely divisible, as \[
			\phi(t) = e^{-c|t|^\alpha} = \left( e^{-\frac{c}{k}|t|^\alpha} \right)^k
		\]
	\end{enumerate}
\end{exmp}

\begin{lem}
	Assume $X_n \cid X$, $Y_n \cid Y$, $\{ X_n \}, \{ Y_n \}$ independent.  Then \[
		X_n + Y_n \cid X + Y.
		\]  
\end{lem}
\begin{proof}
	$X_n$ has a characteristic function $\phi_n(t) \rightarrow \phi(t)$.  $Y_n$ has characteristic function $\psi(t) \rightarrow \psi(t)$. Then \[
		\phi_n(t) \psi_n(t) \rightarrow \phi(t) \psi(t).
	\]   
\end{proof}

\begin{thm}
	Given the array $\{ X_{nk} \}$, letting $S_n = \sum_{k=1}^n X_{nk}$.  If $P(S_n \leq x) \rightarrow F(x)$ then $F$ is infinitely divisible.  
\end{thm}

\begin{proof}	
	Fix $k$.  We must show that $F$ is the $k$-fold convolution of some $G_k$.  Let $n' = mk$, $m = 1, 2, \dots$, and let \[
	Y_i^{(m)} = X_{n', (i - 1)m + 1} + \dots + X_{n', im}, \quad i = 1, \dots, k.	
	\]  Then \[
		S_{mn} = Y_1^{(m)}  + \dots + Y_k^{(m)} 
	\] and $Y_f^{(m)}$ are IID.

	If $P(Y_1^{(m)} \leq x) \rightarrow G_k(x)$ as $m \rightarrow \infty$ then \[
		G^{*k}_k = F
	\]  So we need to show that $G_k$ is a well defined distribution.  We have $Y_{1}^{(m)}$ is the sum of $m$ iid random variables, and \[
		H_m(x) = P(Y_1^{(m)} \leq x).
	\]  We need to ensure ``no probability escapes to infinity.''  Given a convergent subsequence of distribution functions, we know that the limit satisfies $G_k(x), G_k$ right continuous, non-decreasing.  We need to show $G(+\infty) = 1$.  Suppose that there exits $\epsilon > 0$ such that for any $M> 0$ we can find a subsequence $m'_n$ such that \begin{align*}
		P( |Y_1^{(m'_n)}| > M ) > \epsilon
	\end{align*}

	There is a subsequence of $\{ m'_n \}$, $\{ m''_n \}$ say, such that \[
		P(Y_1^{m_n''} > M) > \frac{\epsilon}{2} \quad \text{or} \quad P(Y_1^{m_n''} < -M) > \frac{\epsilon}{2}
	\] So \[
		P(Y_1^{m_n''} + \dots + Y_k^{m_n''} > kM) > \left(\frac{\epsilon}{2}\right)^k 
	\] and so $F(km) \leq 1 - \left( \frac{\epsilon}{2}\right)^k$ (modulo choosing continuity points $kM$ of $F$).  Now, since we know that our limiting distribution $F$ is a proper distribution function, we obtain our contradiction (no such $\epsilon > 0$ exists).

	Hence $G_k$ is a proper distribution function, and so $G_k^{*k} = F$.    
\end{proof}
\begin{defn}[Compound Poisson distribution]
	Let $X_1, X_2, \dots$ IID random variables.  and let $N \sim \textsc{Poisson}(\lambda)$.  Then let $S_N = X_1 + \dots + X_N$.  Then $S_N$ has a compound Poisson distribution.  If $X$ has characteristic function $\phi$, then $S_N$ has characteristic function \begin{align*}
		\E(e^{itS_N}) &= \sum_{n = 0}^\infty \E(e^{itS_N} \given N = n) P(N = n) \\
					&= \sum_{n = 0}^\infty \phi(t)^n e^{- \lambda} \frac{\lambda^n}{n!} \\
					&= e^{- \lambda} e^{\lambda \phi(t)} \\
					&= e^{\lambda(\phi(t) - 1)}.
	\end{align*}  The compound Poisson distribution is clearly infinitely divisible.  
\end{defn}

\begin{thm}
	A distribution function $F$ is infinitely divisible if and only if it is the weak limit of a sequence of distributions, each of which is compound Poisson.
\end{thm}
% section lecture_20_thursday_19_may (end)

\section{Lecture 21 - Thursday 26 May} % (fold)
\label{sec:lecture_21_thursday_26_may}
\begin{thm}
	A distribution function is infinitely divisible if and only if it is the weak limit (limit in distribution) of a sequence of distribution functions each of which is compound Poisson.
\end{thm}

\begin{lem}
	\label{lem:infd1}
	The weak limit of a sequence of infinitely divisible distributions is infinitely divisible.
\end{lem}

\begin{proof}
	Let $F_n(x)$ be a sequence of distribution functions that are infinitely divisible with \[
		F_n(x) \rightarrow F(x)
	\] at all continuity points $x$ of $F$.  Form an array $\{ X_{nk} \}_{k=1}^n$ where for a given $n$, $X_{n1}, X_{n2}, \dots, X_{nn}$ are IID with distribution function $_nF_n(x)$, the \emph{n\textsuperscript{th} root of $F_n$}.  Then \[
		S_n = \sum_{k=1}^n X_{nk}
	\]   has distribution function $F_n$.

	We know $F_n(x) \rightarrow F(x)$ so from the previous result $F$ is infinitely divisible as it is the limit of the row sums of a triangular arrow of row-wise infinitely divisible random variables. 
\end{proof}

\begin{lem}
	\label{lem:infd2}
	The characteristic function of an infinitely divisible distribution is never zero.
\end{lem}

\begin{proof}
	If $\phi(0) = 1$ and $\phi$ is continuous, without loss of generality assume $\phi$ is real (if not, consider $|\phi|^2 = \phi \overline\phi$ which is real and infinitely divisible.)

	Let $\phi_k(t)^k = \phi(t)$.  Assume $\phi(t) > 0$ for $|t| \leq a$.  Then for $t \in (-a, a)$, $\phi_k(t) \rightarrow 1$ as $k \rightarrow \infty$.

	Now note that \[
		1 - \phi(2t) \leq 4(1 - \phi(t)), \tag{$\star$}
	\] as \begin{align*}
		1 - \phi(2t) 	&= \int (1 - \cos 2tx  \, dF(x) \quad \text{as $\phi$ is real }\\
						&= \int (2 -  2\cos^2 tx) \, dF(x) \quad \cos 2 \theta = 2 \cos^2 \theta - 1 \\
						&= 2  \int(1- \cos tx) (1 + \cos tx) \, dF(x) \\
						&\leq 4\int(1 - \cos tx) \, dF(x) \quad 1 - \cos tx \geq 0 \\
						&= 4(1 - \phi(t))
	\end{align*} as required.

	Then we have $1 - |\phi(2t)| \leq 1 - \phi(2t)|^2 \leq 4(1 - |\phi(t)|^2) \leq 8(1 - |\phi(t)|)$.  If $\phi(t) \neq 0$ on $0 < t < a$ and $\epsilon > 0$ arbitrary, we can find $k$ large enough such that \[
		1 - |\phi_k(t)| < \frac{\epsilon}{8}
	\]  which implies $1 - |\phi_k(2t)| < \epsilon$ and so $|\phi_k(2t)| \neq 0 $ on $|t| < a$.  So $\phi_k(t)| \neq 0$ on $|t| < 2a$, and hence $|\phi(t)| \neq 0$ on $|t| < 2a$.

	Iterating this argument, we have that $|\phi(t)| > 0$ for all $t$.  
\end{proof}

\begin{lem}
	\label{lem:infd3}
	For each $k$, let $\phi_k$ be a characteristic function such that $\phi_k^k(t) = \phi(t)$.  $\phi(t)$ is a characteristic function of an infinitely divisible distribution.  Then $lim_{k \rightarrow \infty} \phi_k(t) = 1$ for all $t$.  
\end{lem}
\begin{proof}
	Since $\phi$ is continuous and $\phi(0) = 1$, we have \[
		| \phi_k(t) | = |\phi(t)|^{1/k} \rightarrow 1 
	\] as $k \rightarrow \infty$.

	We have $k \arg \phi_k(t) = \arg \phi(t) + 2 \pi j, j = 0, 1, \dots, k-1$.  Since \begin{align*}
		\arg \phi_k(0) = \arg(1) = 0 \quad \text{so $j = 0$} \\
		\arg \phi_k(t) = \frac{1}{k} \arg \phi(t) \rightarrow 0 
	\end{align*} as $k \rightarrow \infty$, and so $\phi_k(t) \rightarrow 1$ as $k \rightarrow \infty$.  
\end{proof}

\begin{proof}[Proof of theorem]
	Let $\phi$ be the characteristic function of an infinitely divisible distribution $F$.  Let $\phi_k^k(t) = \phi(t)$.  Then \begin{align*}
		\log \phi(t) 	&= k \log \phi_k(t) \\
						&= k \log ( 1 - (1 - \phi_k(t))) 
	\end{align*}
  Since $ 1 - \phi_k(t) \rightarrow 0$ as $ k \rightarrow \infty$, we have \begin{align*}
		\log \phi(t) &= -k [ 1 - \phi_k(t) + \frac{\left( 1 - \phi_k(t) \right)^2}{2} + \dots] \\
			&= - k[1 - \phi_k(t)]( 1 + \frac{1 - \phi_k(t)}{2} + \dots) \\
			&=  - k[1 - \phi_k(t)] + o(1)
	\end{align*} and so $\phi(t) \sim e^{-k( 1 - \phi_k(t))}$ which is a compound Poisson characteristic function. 
\end{proof}

\begin{exmp}
	Show that the $U([-1, 1])$ distribution is not infinitely divisible.  This has associated characteristic function $\frac{\sin t}{t}$.  Then $\phi(\frac{\pi}{2}) = 0$, and so the distribution is not infinitely divisible.
\end{exmp}
% section lecture_21_thursday_26_may (end)

\section{Exam material} % (fold)
\label{sec:exam_material}
\begin{itemize}
	\item Borel-Cantelli lemma.
	\item Martingales, central limit theorems, strong law of large numbers.
	\item Inequalities of random variables.
\end{itemize}

\begin{exmp}[Q2b) of 2010 Exam]
	Let $(X_j)$ be IID.  Then \[
		\E|X_1| < \infty \iff \P(|X_n| \geq n \text{ i.o }) = 0 
	\]  We have \begin{align*}
		\E|X_1| < \infty &\iff \sum_{j=1}^\infty \P(|X_1| \geq j) < \infty \\
				&\iff \sum_{j=1}^\infty \P(|X_j| \geq j) \quad \text{by IID} \\
				&\iff \P(|X_j| \geq j \text{ i.o }) = 0
	\end{align*} by Borel-Cantelli lemma.
\end{exmp}

\begin{exmp}[Q7 of 2010 Exam]
	Let $\{ X_n \}$ be a sequence of IID random variables on a probability space $(\Omega, \sigf, P)$ with \[
		P(X_1 = 1) = P(X_1 = -1) = \frac{1}{2}.
	\]  Let $\sigf_n = \sigma(X_1, X_2, \dots, X_n)$ and let $\{B_n\}$ be a sequence of events with $B_n \in \sigf_n$, satisfying \[
		B_1 = \Omega, \lim_{n \rightarrow \infty} P(B_n) = 0, P(\limsup B_n) = 1.
	\]  Define $Y_1 = 0$ and \[
		Y_{n+1} = Y_n(1 + X_{n+1}) + \mathbf{1}_{B_n} X_{n+1}, n = 1, 2 \dots.
	\]
	\begin{enumerate}[(a)]
		\item Show that $\{ Y_n \}$ is a martingale.
		\item Show that $Y_n$ converges in probability to 0.
		\item Show that $\limsup B_n \subseteq \limsup \{ Y_n \neq 0 \}$ and hence show that $\{ Y_n \}$ does not converge almost surely.
	\end{enumerate}
\end{exmp} 
\begin{proof}{\ }\begin{enumerate}[(a)]
	\item Note that $Y_1$ is $\sigf_1$-measurable.  By induction, we have that $Y_n+1$ is $\sigf_{n+1}$-measurable.

	We have \begin{align*}
		\E|Y_{n+1}| &\leq 2 \E|Y_n | + P(B_n)  \quad \text{as $|X_{n+1} \leq  1$}\\	
	\end{align*}   as $\E|Y_1| = 0$, $P(B_n) \leq 1$, so by induction, $\E|Y_n| <\infty$ for all $n$.

	Finally, \begin{align*}
		\E(Y_{n+1} \given \sigf_n) &= Y_n \E(1 + X_{n+1} \given \sigf_n) + \mathbf{1}_{B_n} \E(X_{n+1} \given \sigf_n) \\
					&= Y_n \quad \text{as $\E(X_{n+1} \given \sigf_n) = \E(X_{n+1}) = 0$.}
	\end{align*}  Hence $Y_n$ is a martingale.
	\item Let $\epsilon > 0$.  We must show $P(|Y_n| > \epsilon) \rightarrow 0$ as $n \rightarrow \infty$.  Consider $P(Y_{n+1} \neq 0)$. We have \begin{align*}
		P(Y_{n+1} \neq 0) &\leq P(\text{$B_n$ occurs or $Y_{n} \neq 0$ and $X_{n+1} = 1$)} \\
						&= P(B_n) + \frac{1}{2} P(Y_n \neq 0).
	\end{align*}  Hence \[
		\lim_{n \rightarrow \infty} P(Y_{n+1} \neq 0) \leq 2 \lim_{n \rightarrow \infty} P(B_n) = 0
	\] and so $Y_n \cip 0$.  
	\item If $Y_n \cas Y$ almost surely then by uniqueness of limits in probability $Y = 0$ almost surely.  We have \[
		Y_{n+1} = \begin{cases}
			2Y_{n} + \mathbf{1}_{B_n} & X_{n+1} = 1 \\
			-\mathbf{1}_{B_n} & X_{n+1} = -1
		\end{cases}
	\]  Hence $B_n \subseteq \{ \omega : Y_{n+1}( \omega) \neq 0 \}$.   Thus \begin{align*}
		\limsup B_n = \bigcap_{m}^\infty \bigcup_{n=m}^\infty B_{n} \subseteq \bigcap_{m}^\infty \bigcup_{n=m}^\infty \{ Y_{n+1} \neq 0 \} \\
		&= \limsup \{Y_n \neq 0 \}
	\end{align*}  Hence $1 = P(\limsup B_n) \leq P(\limsup \{ Y_n \neq 0 \})$ and so $P(Y_n \neq 0 \text{ i.o.}) = 1$, and so $Y_n$ does not converge almost surely.
\end{enumerate}

\end{proof}
% section exam_material (end)

\end{document}

### 2011-12-12 MSH7 - Applied Probability and SDEs/MSH7 Lecture Notes.tex
Initial Commit

% Created by Andrew Tulloch

%!TEX TS-program = xelatex
%!TEX encoding = UTF-8 Unicode

\documentclass[10pt, oneside, reqno]{amsart}
\usepackage{geometry, setspace, graphicx, enumerate}
\onehalfspacing                 
\usepackage{fontspec,xltxtra,xunicode}
\defaultfontfeatures{Mapping=tex-text}

%\setromanfont[Mapping=tex-text,Contextuals= 
%{NoWordInitial,NoWordFinal,NoLineInitial,NoLineFinal}]{Hoefler Text}
%\setsansfont[Scale=MatchLowercase,Mapping=tex-text]{Hoefler Text}
%\setmonofont[Scale=MatchLowercase]{Andale Mono}

	% AMS Theorems
	\theoremstyle{plain}% default 
	\newtheorem{thm}{Theorem}[section] 
	\newtheorem{lem}[thm]{Lemma} 
	\newtheorem{prop}[thm]{Proposition} 
	\newtheorem*{cor}{Corollary}

\newcommand{\res}[2]{\text{Res}(#1,#2)}
	\theoremstyle{definition} 
		\newtheorem{defn}[thm]{Definition}
		\newtheorem{conj}[thm]{Conjecture}
		\newtheorem{exmp}[thm]{Example}

	\theoremstyle{remark} 
		\newtheorem*{rem}{Remark} 
		\newtheorem*{note}{Note} 
		\newtheorem{case}{Case}

		\newcommand{\expc}[1]{\mathbb{E}\left(#1\right)}
		\newcommand{\var}[1]{\text{Var}\left(#1\right)}
		\newcommand{\cov}[1]{\text{Cov}\left(#1\right)}
		\newcommand{\prob}[1]{\mathbb{P}(#1)}
		\newcommand{\given}{ \, | \,}
		\newcommand{\us}{0 \leq u \leq s}
		\newcommand{\ts}[1]{\{ #1 \}}

\newcommand{\dzz}{\, dz}
\newcommand{\bigo}[1]{\mathcal{O}(#1)}

\newcommand{\al}{\alpha}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}

\renewcommand{\P}{\mathbb{P}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\E}{\mathbb{E}}

\newcommand{\Z}{\mathbb{Z}}
\newcommand{\F}{\mathbb{F}}
\newcommand{\Ga}{\mathbb{G}}

\newcommand{\aut}[1]{\text{Aut}{(#1)}}

\newcommand{\gener}[1]{\langle #1 \rangle}
\newcommand{\charr}[1]{\text{char}(#1)}
\newcommand{\nth}{n\textsuperscript{th}}

\newcommand{\tworow}[2]{\genfrac{}{}{0pt}{}{#1}{#2}}
\newcommand{\xdeg}[2]{[#1 : #2]}
\newcommand{\gal}[2]{\text{Gal}(#1/#2)}
\newcommand{\minpoly}[2]{m_{#1, #2}(x)}

\newcommand{\mapping}[5]{\begin{align*}
	#1 : \quad     #2 &\rightarrow #3 \\
			#4  &\mapsto #5
\end{align*}	
}

\def\cip{\,{\buildrel p \over \rightarrow}\,} 
\def\cid{\,{\buildrel d \over \rightarrow}\,} 
\def\cas{\,{\buildrel a.s. \over \rightarrow}\,}

\def\eqd{\,{\buildrel d \over =}\,} 
\def\eqas{\,{\buildrel a.s. \over =}\,}

\newcommand{\sigf}{\mathcal{F}}

\title{MSH7 - Applied Probability and Stochastic Calculus}								% Document Title
\author{Andrew Tulloch}
%\date{}                                           % Activate to display a given date or no date

\begin{document}
\maketitle

\section{Lecture 1 - Tuesday 1 March} % (fold)
\label{sec:lecture_1_1_march}

\begin{defn}[Finite dimensional distribution]
	The \textbf{finite dimensional distribution} of a stochastic process $X$ is the joint distribution of $(X_{t_1}, X_{t_2}, \dots, X_{t_n})$
\end{defn}

\begin{defn}[Equality in distribution]
	Two random variables $X$ and $Y$ are \textbf{equal in distribution} if $\prob{X \leq \alpha} = \prob{Y \leq \alpha}$ for all $\alpha \in \R$.  We write $X \eqd Y$.
\end{defn}

\begin{defn}[Strictly stationary]
	A stochastic process $X$ is \textbf{strictly stationary} if \[
		(X_{t_1}, X_{t_2}, \dots, X_{t_n}) = (X_{t_1+h}, X_{t_2+h}, \dots, X_{t_n+h})
	\] for all $t_i, h$.
\end{defn}

\begin{defn}[Weakly stationary]
	A stochastic process $X$ is \textbf{weakly stationary} if $\expc{X_t} = \expc{X_{t+h}}$ and $\cov{X_t, X_s} = \cov{X_{t+h}, X_{s+h}}$ for all $t, s, h$.
\end{defn}

\begin{lem}
	If $\expc{X_t^2} < \infty$, then strictly stationary implies weakly stationary.
\end{lem}

\begin{exmp}{\ }
	\begin{itemize}
		\item The stochastic process $\{ X_t \}$ with $X_t$ all IID is strictly stationary.
		\item The stochastic process $W_t$ with $W_t ~ N(0,t)$ and $X_t - X_s$ independent of $X_s$ (for $s < t$) is not strictly or weakly stationary.
	\end{itemize}
\end{exmp}
\begin{defn}[Stationary increments]
	A stochastic process has \textbf{stationary increments} if 
	\[
		X_t - X_s \eqd X_{t-h} - X_{s-h}
	\] for all $s, t, h$.
\end{defn}
% section lecture_1_1_march (end)

\section{Lecture 2 - Thursday 3 March} % (fold)
\label{sec:lecture_2_thursday_3_march}

\begin{exmp}
	Let $X_n, n \geq 1$ be IID random variables.  Consider the stochastic process $\{ S_n \}$ where $S_n = \sum^n_{j=1} X_j$.  Then $\{ S_n \}$ has stationary increments.
\end{exmp}

\subsection{Concepts of convergence} % (fold)
\label{sub:convergence_in_distirbution}
There are three major concepts of convergence of random variables.
\begin{itemize}
	\item Convergence in distribution
	\item Convergence in probability
	\item Almost surely convergence
\end{itemize}

\begin{defn}[Convergence in distribution]
	$X_n \cid X$ if $\prob{X_n \leq x} \rightarrow \prob{ X \leq x}$ for all $x$.
\end{defn}

\begin{defn}[Convergence in probability]
	$X_n \cip X$ if $\prob{ (|X_n - X | \geq \epsilon)} \rightarrow 0$ for all $\epsilon > 0$.
\end{defn}

\begin{defn}[Almost surely convergence]
	$X_n \cas X$ if except on a null set $A$, $X_n \rightarrow X$, that is, $\lim_{n \rightarrow \infty} X_n = X$.  And hence $\prob{ \lim_{n \rightarrow \infty} X_n = X} = 1$
\end{defn}

\begin{defn}[$\sigma$-field generated by $X$]
	Let $X$ be a random variable defined on a probability space $(\Omega, \mathcal{F}, \P)$.  We call $\sigma(X)$ the $\sigma$-field generated by $X$, and we have \[
		\sigma(X) = \left\{ X^{-1}(B): B \in \mathcal{B} \right\}
	\] where $X^{-1}(B) = \{ \omega, X(\omega) \in \mathcal{B} \}$ and $\mathcal{B}$ is the Borel set on $\R$.
\end{defn}

\begin{defn}[Conditional probability]
	We have $\P( A | B) = \frac{\P(A, B)}{\P(B)}$ if $\P(B) \neq 0$.
\end{defn}

\begin{defn}[Naive conditional expectation]
	We have $\expc{X | B} = \frac{ \expc{ X \mathbb{I}_B}}{\P(B)}$.
\end{defn}

\begin{defn}[Conditional density]
	Let $g(x,y)$ be the joint density function for $X$ and $Y$.  Then we have $Y ~ \int_\R g(x,y)\, dx \equiv g_Y(y)$.  We also have \[
		g_{X|Y = y} = \frac{g(x,y)}{g_Y(y)}
	\] which defines the conditional density given $Y = y$.

	Finally, we define $\expc{ X |Y = y} = \int_\R x g_{X|Y = y}(x) \,dx$.
\end{defn}

\begin{defn}[Conditional expectation]
	Let $(\Omega, \mathcal{F}, \P)$ be a probability space.  Let $\mathcal{A}$ be a sub $\sigma$-field of $\mathcal{F}$.  Let $X$ be a random variable such that $\E(|X|) < \infty$.  We define $\expc{X | \mathcal{A}}$ to be a random variable $Z$ such that 
	\begin{enumerate}[(i)]
		\item $Z$ is $\mathcal{A}$-measurable,
		\item $\E(X\mathbb{I}_A) = \E(Z\mathbb{I}_A)$ for all $A \in \mathcal{A}$.
	\end{enumerate}
\end{defn}

\begin{prop}[Properties of the conditional expectation]
	Consider $Z = \E(X|Y) = \E(X|\sigma(Y))$
	\begin{itemize}
		\item If $T$ is $\sigma(Y)$-measurable, then $\E(XT|Y) = T\E(X|Y)$ a.s.
		\item If $T$ is independent of $Y$, then $\E(T|Y) = \E(T)$.
		\item $\E(X) = \E(\E(X|T))$
	\end{itemize}
\end{prop}

% subsection convergence_in_distirbution (end)
% section lecture_2_thursday_3_march (end)

\section{Lecture 3 - Tuesday 8 March} % (fold)
\label{sec:lecture_3_tuesday_8_march}

\begin{defn}[Martingale]
	Let $\{ X_t, t \geq 0 \}$ be a \textbf{right-continuous} with \textbf{left-hand limits}.  \[
		\lim_{t \uparrow t_0} X_t \quad \text{exists}
	\]
	Let $\{\sigf_t, t \geq 0 \} $ be a filtration.

	Then $X$ is called a martingale with respect to $\sigf_t$ if 
	\begin{enumerate}[(i)]
		\item $X$ is \textbf{adapted to $\sigf_t$}, i.e. $X_t$ is $\sigf_t$-measurable
		\item $\E(|X|) < \infty, t \geq 0$
		\item $\E(X_t | \sigf_s) = X_s$ a.s.
	\end{enumerate}
\end{defn}

\begin{exmp}
	Let $X_n$ be IID with $\E(X_n) = 0$.  Then $\{ S_k, k \geq 0 \}$, where $S_k = \sum_{i = 0}^k X_i$, is a martingale.
\end{exmp}

\begin{exmp}
	An independent increment process $\{ X_t, t \geq 0 \}$ with $\E(X_t) = 0$ and $\E(|X_t|) \leq \infty$ is a martingale with respect to $\sigf_t = \sigma\{ X_s, 0 \leq s \leq t \}$
\end{exmp}

\begin{defn}[Gaussian process]
	Let $\{ X_t, t \geq 0 \}$ be a stochastic process.  If the finite dimensional distributions are multivariate normal, that is,\[
		(X_{t_1}, \dots, X_{t_m}) \equiv N(\mu, \Sigma)
	\]
	for all $t_1, \dots, t_m$, then we call $X_t$ a \textbf{Gaussian process}
\end{defn}

\begin{defn}[Markov process]
	A continuous time process $X$ is a \textbf{Markov process} if for all $t$, each $A \in \sigma( X_s, s > t)$ and $B \in \sigma( X_s, s < t)$, we have \[
		\P(A|X_t, B) = \P(A | X_t)
	\]
\end{defn}

\begin{defn}[Diffusion process]
	Consider the stochastic differential equation \[
	dX_t = \mu(t, x) dt + \sigma(t,x) dB_t
	\]
	A diffusion process is \textbf{path-continuous}, \textbf{strong Markov} process such that
	\begin{align*}
		\lim_{h \rightarrow 0} h^{-1} \E( X_{t+h} - X_t | X_t = x) &= \mu(t, x)	\\
		\lim_{h \rightarrow 0} h^{-1} \E([X_{t+h} - X_t - h\mu(t, X)]^2 | X_t = x) &= \sigma^2(t, x)
	\end{align*}

\end{defn}

\begin{defn}[Path-continuous]
	A process is \textbf{path-continuous} if $\lim_{t \rightarrow t_0} X_t = X_{t_0}$.
\end{defn}

\begin{defn}[L\'evy process]
	Let $\{X_t, t \geq 0$ be a stochastic process.  We call $X$ a \textbf{L\'evy process}
	\begin{enumerate}[(i)]
		\item $X_0 = 0$ a.s.
		\item It has stationary and independent increments
		\item $X$ is stochastically continuous, that is, for all $s, t, \epsilon > 0$, we have \[
			\lim_{s \rightarrow t} \P(|X_s - X_t| \geq \epsilon) = 0.
		\]  Equivalently, $X_s \cip X_t$ if $s \rightarrow t$.
	\end{enumerate}
\end{defn}
% section lecture_3_tuesday_8_march (end)

\begin{exmp}[Poisson process]
	Let $(N(t), t \geq 0 )$ be a stochastic process.  We call $N(t)$ a \textbf{Poisson process} if the following all hold: 
	\begin{enumerate}[(i)]
		\item $N(0) = 0$
		\item $N$ has independent increments
		\item For all $s, t \geq 0$, \[
			\P(N(t+s)-N(s) = n) = \frac{e^{-\lambda t}(\lambda t)^n}{n!} \quad n = 0, 1, 2, \dots
		\]
	\end{enumerate}

	The Poisson process is stochastically continuous - that is,
	\begin{align*}
		\P( | N(t) - N(s) | \geq \epsilon) 	&= \P(|N(t - s) - N(0) | \geq \epsilon) \\
		 									&= 1- \P(|N(t-s) | < \alpha ) \\
											&= 1 - \P(|N(t - s) | = 0 ) \\
											&= 1 - e^{-\lambda(t-s)} \rightarrow 0 \quad \text{as $t \rightarrow s$}
	\end{align*}

	The Poisson process is \textbf{not} path-continuous, that is \[
		\P( \lim_{t \rightarrow s} |N(t) - N(s)| = 0) \neq 1
	\] because \[
		\P( \displaystyle\cup_{|t-s| \geq \epsilon} |N(t) - N(s)| > \delta ) \geq \P(|N(s+1) - N(s) | \geq \delta ) > 0
	\]
\end{exmp}

\section{Lecture 4 - Thursday 10 March} % (fold)
\label{sec:lecture_4_thursday_10_march}

\begin{defn}[Self-similar process]
	For any $t_1, t_2, \dots, t_n \geq 0$, for any $c > 0$, there exists an $H$ such that \[
		(X_{ct1}, X_{ct_2}, \dots, X_{ct_n}) \eqd 	(c^H X_{t1}, c^H X_{t_2}, \dots, c^H X_{t_n}).
	\]
	We call $H$ the \textbf{Hurst index}.
\end{defn}

\begin{exmp}[Fractional process]
	\[
		(1-B)^d X_t = \epsilon_t, \quad \epsilon_t \text{ martingale difference}
	\]
	\[
		B X_t = X_{t-1}, \quad 0 < d < 1
	\]
\end{exmp}

\begin{defn}[Brownian motion]
	Let $\{ B_t, t \geq 0 \}$ be a stochastic process.  We call $B_t$ a \textbf{Brownian motion} if the following hold:
	\begin{enumerate}[(i)]
		\item $B_0 = 0$ a.s.
		\item $\{B_t \}$ has stationary, independent increments.
		\item For any $t > 0$, $B_t \equiv N(0,t)$
		\item The path $t \mapsto B_t$ is continuous almost surely, i.e.\[
			\P(\lim_{t \rightarrow t_0} B_t = B_{t_0}) = 1
		\]
	\end{enumerate}
\end{defn}

\begin{defn}[Alternative formulations of Brownian motion]
	A process $\{ B_t, t \geq 0 \}$ is a Brownian motion if and only if 
	\begin{itemize}
		\item $\{ B_t, t \geq 0 \}$ is a Gaussian process
		\item $\E(B_t) = 0, \E(B_s B_t) = \min(s,t)$
		\item The process $\{ B_t \}$ is path-continuous
	\end{itemize}
\end{defn}

\begin{proof}
	$(\Rightarrow)$  For all $t_1, \dots, t_m$, we have \[
		(B_{t_m} - B_{t_{m-1}}, \dots, B_{t_2} - B_{t_1}) \equiv N(0, \Sigma)
	\]
	as $B_t$ has stationary, independent increments.  Hence, $(B_{t_1}, B_{t_2,}, \dots, B_{t_n})$ is normally distributed.  Thus $B_t$ is a Gaussian process.  We can also show that $\E(B_t) = 0$ and $\E(B_s B_t) = \min(t,s)$.

	$(\Leftarrow)$ TO PROVE
\end{proof}

\begin{cor}
	Let $B_t$ be a Brownian motion.  The so are the following:
	\begin{itemize}
		\item $\{ B_{t + t_0} - B_{t_0}, t \geq 0 \}$
		\item $\{ -B_t, t \geq 0 \}$
		\item $\{ c B_{t/c^2}, t \geq 0, c \neq 0 \}$
		\item $\{ t B_{1/t} , t \geq 0 \}$
	\end{itemize}
\end{cor}

\begin{proof} Here, we prove that $\{ X_t \} = \{ t B_{1/t} \}$ is a Brownian motion.  Consider $\sum^n \alpha_i X_{t_i}$.  Then by a telescoping argument, we know that the process is Gaussian (can be written as a sum of $X_{t_1}, X_{t_2} - X_{t_1}$, etc).  We can also easily show that $\E(X_t) = 0$ and $\E(X_t X_s) = \min(s,t)$ as required.

We now show that $\lim_{t \rightarrow 0} X_t = 0$ a.s. Fix $\epsilon \geq 0$.  We must show \[
	\P \left( \bigcap_{n=1}^\infty \bigcup_{m = 1}^\infty \bigcap_{0 < t < \frac{1}{m}} \{ |X_t | \leq \frac{1}{n} \} \right) = 1
\]
However, as $|X_t|$ has the same distribution as $|B_t|$ (as they are both Gaussian with same mean and covariance), we have that this is equivalent to \[
		\P \left( \bigcap_{n=1}^\infty \bigcup_{m = 1}^\infty \bigcap_{0 < t < \frac{1}{m}} \{ |B_t | \leq \frac{1}{n} \} \right)
\] which is clearly one.

\end{proof}
% section lecture_4_thursday_10_march (end)

\section{Lecture 5 - Tuesday 15 March} % (fold)
\label{sec:lecture_5_tuesday_15_march}

\begin{thm}[Properties of the Brownian motion]
	We have \[
		\P( B_t \leq x \, | \, B_{t_0} = x_0, \dots, B_{t_n} = x_n ) = \P(B_t \leq x | B_s = x_s) = \Phi( \frac{x - x_s}{\sqrt{t-s}})
	\] 
\end{thm}

\begin{thm}
	The joint density of $(B_{t_1}, \dots, B_{t_n}$ is given by \[
		g(x_1, \dots, x_n) = \prod_{j=1}^n f(x_{t_j} - x_{t_{j-1}}, t_j - t_1)
	\] where $f(x,t) = \frac{1}{\sqrt{2 \pi t}} e^{-\frac{x^2}{2t}}$
\end{thm}

\begin{thm}[Density and distribution of Brownian bridge]
	Let $t_1  < t < t_2$.  Then \[
		g( B_t | B_{t_1} = a, B_{t_2} = b ) \equiv N\left(a + \frac{(b-a)(t-t_1)}{t_2 - t_1}, \frac{(t_2-t)(t-t_1)}{t_2 - t_1} \right).
	\]

	The density of $B_t | B_{t_1} =a , B_{t_2} = b$ is given as \[
		\frac{g_{t_1, t, t_2}(a,b,t)}{g_{t_1, t_2}(a,b)}
	\]
\end{thm}

\begin{thm}[ Joint distribution of $B_t$ and $B_s$]
	We have 
	\begin{align*}
		\P(B_s \leq x, B_t \leq y) &= \\ \P(B_s \leq x, B_{t} - B_s \leq y - B_s ) \\
		&= \int_{-\infty}^x \int_{-\infty}^{y-z} \frac{1}{\sqrt{2\pi s}} e^{-\frac{x_1^2}{2s}} \frac{1}{\sqrt{2 \pi (t-s)}} e^{-\frac{-z_2^2}{2(t-s)}} \, dz_1 \, dz_2 \\
		&= \int_{-\infty}^x \int_{-\infty}^y \frac{1}{\sqrt{2\pi s}} e^{-\frac{x_1^2}{2s}} 
	\end{align*}
\end{thm}

\subsection{Properties of paths of Brownian motion} % (fold)
\label{sub:properties_of_paths_of_brownian_motion}

\begin{defn}[Variation]
	Let $g$ be a real function.  Then the variation of $g$ over an interval $[a,b]$ is defined as \[
		V([a.b]) = \lim_{\Delta \rightarrow 0} \sum_{j = 1}^n | g(t_j) - g(t_{j-1}) |
	\] where $\Delta$ is the size of the partition $a = t_0, \dots, t_n = b$ of $[a,b]$.
\end{defn}

\begin{defn}[Quadratic variation]
	The quadratic variation of a function $g$ over an interval $[a,b]$ is defined by \[
		[g,g] = \lim_{\Delta \rightarrow 0} \sum_{j = 1}^n | g(t_j) - g(t_{j-1})|^2	\]
\end{defn}

\begin{thm}[Non-differentiability of Brownian motion]
	Paths of Brownian motion are continuous almost everywhere, by definition.  Consider now, the differentiability of $B_t$.  We claim that a Brownian motion is non-differentiable almost surely, that is, \[
		\lim_{t \rightarrow s} \frac{|B_t - B_s|}{|t - s|} = \infty \quad a.s.
	\]

	We claim that
	\begin{enumerate}[(i)]
		\item Brownian motion is not differentiable almost surely for any $t \geq 0$.
		\item Brownian motion has infinite variation on any interval $[a,b]$, that is,\[
			V_B([a,b]) = \lim_{\Delta \rightarrow 0} \sum_{j=1}^n |B(t_j) - B(t_{j-1})| = \infty \quad a.s.
		\]
		\item Brownian motion has quadratic variation $t$ on $[0,t]$, that is \[
			[B,B]([0,t]) = \lim_{\Delta \rightarrow 0} \sum_{j=1}^n |B(t_j) - B(t_{j-1})|^2 = t \quad a.s.
		\] 
	\end{enumerate}
\end{thm}

\begin{proof}
	We know that $(B_{ct_1},\dots, B_{ct_n} ) \eqd (c^H B_{t_1}, \dots, c^H B_{t_n})$.  This is true as $B_{ct} \equiv N(0, ct) = c^{1/2} N(0,t)$ as required.

	Now suppose $X_t$ is $H$-self-similar with stationary increments for some $0 < H < 1$ with $X_0 = 0$.  Then, for any fixed $t_0$, we have \[
		\lim_{t \rightarrow t_0} \frac{|X_t - X_{t_0}|}{t - t_0} = \infty \quad a.s.
	\]

	Consider
	\[
		\P( \limsup_{t \rightarrow t_0} \frac{ |X_t - X_{t_0} |}{t - t_0} \geq M) 
	\] which by stationary increments, is equal to 
	\begin{align*}
		\P( \limsup_{t \rightarrow t_0} \frac{ |X_t |}{t} \geq M) &= \lim_{n \rightarrow \infty}\P( \bigcup_{k \geq n} \frac{|B_{t_n} - B_{t_0}}{|t_n - t_0|} \geq M) \\
		&> \lim_{n\rightarrow \infty} \P( \frac{|B_{t_n} - B_{t_0}}{|t_n - t_0|} \geq M ) \\
		&= \lim_{n \rightarrow \infty} \P( |N(0,1) | \geq M \cdot (t_n - t_0)^{1/2})
	\end{align*}
	and as the RHS goes to zero, we have \[
	\P( |N(0,1) | \geq M \cdot (t_n - t_0)^{1/2}) \rightarrow 1
	\]
	as required.
\end{proof}

Now, Assume $V_B([0,t]) < \infty$ almost surely.  Consider $Q_n = \sum_{j=1}^n | B_{t_j} - B_{t_{j-1}}  |^2$.  Then we have \[
	Q_n \leq \max_{ 0 \leq j \leq n} | B_{t_j} - B_{t_{j-1}} | \cdot \sum_{j=1}^n |B_{t_j} - B_{t_{j-1}} |
\]

Let $\Delta \rightarrow 0$.  Then \begin{align*}
	\lim_{\Delta \rightarrow 0} Q_n &\leq  \lim_{ \Delta \rightarrow 0}\max_{ 0 \leq j \leq n} | B_{t_j} - B_{t_{j-1}} | \cdot \sum_{j=1}^n |B_{t_j} - B_{t_{j-1}} | \\
									&\leq \lim_{ \Delta \rightarrow 0}\max_{ 0 \leq j \leq n} | B_{t_j} - B_{t_{j-1}} | \cdot V_B([0,t]) \\
									&\leq 0 \cdot V_B([0,t])
									&= 0
\end{align*} because $B_s$ is uniformly continuous on $[0,t]$.  This is a contradiction to $V_B([0,t]) < \infty$.

\begin{proof}
	We now show that $\E((Q_n - t)^2) \rightarrow 0$ as $n \rightarrow \infty$.  In fact, we have
	\begin{align*}
		\E(Q_n) &= \sum_{i=1}^n \E(B_{t_j} - B_{t_{j-1}})^2 \\
				&= \sum_{j=1}^n (t_j - t_{j-1}) = t
	\end{align*}

	We now show $L^2$ convergence. We have \begin{align*}
		\E(Q_n - t)^2 = \E((Q_n - \E(Q_n))^2) \\
					&= \E(\sum_{j=1}^n Y_j \quad \text{where $Y_j = |B_{t_j} - B_{t_{j-1}}|^2 - \E(|B_{t_j} - B_{t_{j-1}}|^2)$} \\
					&= \sum_{j=1}^n \E(Y_j^2) ) \\
					&\leq \sum_{j=1}^n | t_{j} - t_{j-1} |^2 \cdot \E | N(0,1) |^4 \\
					&\leq C \cdot \Delta \cdot t \rightarrow 0
	\end{align*} as $\Delta \rightarrow 0$.  Thus we have convergence in $L^2$.  
\end{proof}

% subsection properties_of_paths_of_brownian_motion (end)

% section lecture_5_tuesday_15_march (end)

\section{Lecture 6 - Thursday 17 March} % (fold)
\label{sec:lecture_6_thursday_17_march}
\begin{thm}[Martingales related to Brownian motion]
	Let $B_t, t \geq 0$ be a Brownian motion.  Then the following are martingales with respect to $\sigf_t = \sigma( B_s, 0 \leq s \leq t)$.  
	\begin{enumerate}[(1)]
		\item $B_t, t \geq 0$.
		\item $B_t^2 - t, t \geq 0$.
		\item For any $u$, $e^{u B_t - \frac{u^2 t}{2}}$
	\end{enumerate}
\end{thm}
\begin{proof}
	(1) is simple.

	(2).  We know that $\E( |B_t^2|)$ is finite for any $t$.  We can also easily show $\E(B_t^2 - t \, | \, \sigf_s ) = B_s^2 - s$ a.s.
\end{proof}

\begin{thm}
	Let $X_t$ be a martingale satisfying $X_t^2 - t$ is also a martingale.  Then $X_t$ is a Brownian motion.
\end{thm}

\begin{defn}[Hitting time]
	Let $T_\alpha = \inf_{\{ t \geq 0, B_t = \alpha \}}$

	\begin{enumerate}[(1)]
		\item If $\alpha = 0$, $T_0 = 0$.
		\item If $\alpha > 0$, then \[
			\P(T_\alpha \leq t) = 2 \P(B_t \geq \alpha) = \frac{2}{\sqrt{2 \pi t}} \int_a^\infty e^{-\frac{x^2}{2t}}\, dx 
		\] 	
	\end{enumerate}

	We clearly have $T_\alpha > t \iff \sup_{ 0 \leq s \leq t} < \alpha$

\end{defn}

% section lecture_6_thursday_17_march (end)

\section{Lecture 7 - Tuesday 22 March} % (fold)
\label{sec:lecture_7_tuesday_22_march}

\begin{thm}[Arcsine law]
	Let $B_t$ be a Brownian motion.  Then \[
		\P(\text{$B_t = 0$, for at least once, $t \in [a,b]$}) = \frac{2}{\pi} \text{arcos} \sqrt{\frac{a}{b}}
	\]
\end{thm}

\begin{exmp}
	Processes derived from Brownian motion
	\begin{enumerate}[(1)]
		\item Brownian bridge
		\[
			X_t = B_t - t B_1 \quad t \in [0,1]
		\]
		Consider $X \equiv F(x)$, with $X_1, \dots X_n$ data.  Our empirical distribution \[
			F_n(x) = \frac{1}{n} \sum_{i=1}^n \mathbf{1}_{\{ X_j \leq X}
		\]
		We can then prove that \[
		{}^n\sqrt{F_n(x) - F(x)} \rightarrow X_t \quad t \in (0,1)
		\] 
		\item Diffusion process\[
			X_t = \mu t + \sigma B_t
		\] This is a Gaussian process with $\E(X_t) = \mu t$, $\text{Cov}(X_t, X_s) = \sigma^2 \min(s,t)$.
		\item Geometric Brownian motion\[
			X_t = X_0 e^{\mu t + \sigma B_t}
		\] This is not a Gaussian process.
		\item Higher dimensional Brownian motion\[
			B_t = ( B^1_t, \dots, B^n_t)
		\] where the $B^{i}$ are independent Brownian motions, then

	\end{enumerate}
\end{exmp}

\subsection{Construction of Brownian motion} % (fold)
\label{sub:construction_of_brownian_motion}
Define a stochastic process \[
	\hat B_t^n = \frac{ S_{[nt]} - \E(S_{[nt]})}{\sqrt{n}}
\]

With \[
	\tilde B_t^n = \begin{cases}
		\hat B_t^n &\text{if $t = \frac{i}{n}$} \\
		0			&\text{otherwise}
	\end{cases}
\]
Then we can prove that \[
	\tilde B_t^n \Rightarrow B_t \quad \text{on [0,1]}
\]
% subsection construction_of_brownian_motion (end)

\begin{defn}[Stochastic integral]
	We now turn to defining expressions of the form\[
		\int_0^A X_t \, dY_t
	\] with $X_t, Y_t$ stochastic processes.
\end{defn}

\begin{defn}[$\int_0^t f(B_s) \, ds$]

	We have \begin{align*}
	\int_0^t f(t) \, dt
	\end{align*} exists if $f$ is bonded and continuous, except on a set of Lebesgue measure zero.
	Thus, we can set \[
		\int_0^t f(t) \, dt = \lim_{ \Delta \rightarrow 0} \sum_{j=1}^n f(B_{y_j}) (t_j - t_{j-1})
	\] if $f(x)$ is bounded.

	We now seek to find $\int_0^1 B_s \, ds$.  Consider $Q_n = \sum_{j=1}^n B_{y_j} (t_j - t_{j-1})$.  As the sum of normal variables, we know that $Q_n \equiv N(\mu_n, \sigma^2_n)$.

	Since $\int_0^1 B_s \, ds$ is normally distributed with mean 0, we have 
	\begin{align*}
		\E(X^2) &= \E \int_0^1 B_s \, dx \int_0^1 B_t \, dt  \\
				&= \int_0^1 \E(B_s B_t) \, ds \, dt \\
				&= \int_0^1 \min(s,t) \, ds \, dt \\
				&= 1/3
	\end{align*}
\end{defn}

% section lecture_7_tuesday_22_march (end)

\section{Lecture 8 - Thursday 24 March} % (fold)
\label{sec:lecture_8_thursday_24_march}

\[
	\int_0^1 f(B_s) \, ds = \lim_{\Delta \rightarrow 0} \sum_{j=1}^n f(B_{y_j}) (t_{j+1} - t_{j})
\]
Recall \[
	\int_0^1 B_s \, ds \sim N(0, \frac{1}{3}) = \lim_{\Delta \rightarrow 0} \sum_{i=1}^n {B_{y_j}} (t_{j+1} - t_{j})
\]

Consider $I = \int X_s \, dY_s$.  We have the following:
\begin{thm}
	I exists if 
	\begin{enumerate}[(1)]
		\item The functions $f,g$ are not discontinuous at the same point $x$.  
		\item $f$ is continuous and $g$ has bounded variation or,
		\item[(2)'] $f$ has finite $p$-variation and $g$ has finite $q$-variation, where \[
			1/p + 1/q = 1
		\]  For any $p$, we define $p$-variation by \[
			\lim_{\Delta \rightarrow 0} \sum_{j=1}^n |f(t_j) - f(t_{j-1}) |^p
		\]
	\end{enumerate}
\end{thm}

\begin{thm}
	If $J = \int_0^1 f(t) \, dg(t)$ exists for any continuous $f$, then $g$ must have finite variation.  
\end{thm}

\begin{thm}
	$B_s$ has bounded $p$ variation for any $p \geq 2$ and unbounded $q$-variation for any $q < 2$.
\end{thm}

\begin{proof}
	We can write (for $p \geq 2$), $p = 2 + (p-2)$.  Then we have \[
		\Delta_t = \lim_{\Delta \rightarrow 0} \max |{B_{t_j}} - B_{t_{j-1}} |^{p-2} \sum_{j=1}^n |B_{t_j} - B_{t_{j-1}} |^{2}
	\] and hence $\Delta_t$ exists.
\end{proof}

\begin{cor}
	Thus $\int_0^1 \, dB_t$ is well defined if $f$ has finite variation (as setting $q = 1$, $ p \geq 2$ gives $1/p + 1/q > 1$).
\end{cor}

Consider $\int_0^1 B_s \, dB_s$ - is this an R-S integral?  Consider \[
	\Delta_{1n} = \sum_{j=1}^\infty B_{t_j} ( B_{t_{j+1}} - B_{t_j}), \Delta_{2n} = \sum_{j=1}^\infty B_{t_{j+1}} ( B_{t_{j+1}} - B_{t_j})
\]  We have that \[
	\Delta_{2n} - \Delta_{1n} = \sum_{j=1}^n ( B_{t_{j+1}} - B_{t_j})^2 \rightarrow 1
\] and \[
	\Delta_{1n} + \Delta_{2n} = \sum_{j=1}^n {B_{t_{j+1}}^2} - B_{t_j}^2 ) = B_1^2. 
\]  Thus we know \begin{align*}
	\Delta_{2n} &\rightarrow \frac{1}{2}(B_1^2 + 1) \\
	\Delta_{1n} &\rightarrow \frac{1}{2}(B_1^2 - 1) 
\end{align*}

\begin{defn}[It\^o integral]
	The It\^o integral is defined by evaluating $f(B_{t_j})$, the left-hand endpoint at each partition interval $[t_{j}, t_{j+1})$
\end{defn}
% section lecture_8_thursday_24_march (end)
\section{Lecture 9 - Tuesday 29 March} % (fold)
\newcommand{\ito}{\text{It\^o\ }}
\label{sec:lecture_9_tuesday_29_march}
\begin{defn}[\ito integral]
	Consider $\int_0^1 f(s) \, dB_s$.  Where $f(s)$ is a real function, $B_s$ a Brownian motion.  We define the integral in two steps.  
	\begin{enumerate}[(1)]
		\item If $f(s)$ is a step function, define 
		\begin{align*}
			I(f) = \int_0^1 f(s) \, dB_s 	&= \sum_{j=1}^m \int_{t_j}^{t_{j+1}} f(s) \, dB_s \\
									&= \sum_{j=1}^m c_j (B_{t_{j+1}} - B_{t_j})
		\end{align*}

		\item If $f(s) \in L^2([0,1])$, then let $f_n$ be a sequence in $L^2([0,1])$ such that $f_n \rightarrow f$ in $L^2([0,1])$.

		Then define $I(f)$ to be the limitation in such situations such that \[
			\E(I(f_n) - I(f))^2 \rightarrow 0 
		\] in $L^2([0,1])$ or 
		Let $I(f) = \lim_{n \rightarrow 0} I(f_n)$ in probability.
	\end{enumerate}

	\begin{rem}
		If $f(x), g(x)$ are given step functions then $\alpha I(f) + \beta I(g) = I ( \alpha f + \beta g)$.
	\end{rem}
	\begin{rem}
		If $f(x)$ is a step function, then $I(f) \sim N(0, \int_0^1 f^2(s) \, ds)$
	\end{rem}
	\begin{proof}
		\begin{align*}
			I(f) = \sum_{j=1}^m c_j (B_{t_{j+1}} - B_{t_j}) \sim N(0, \sigma^2) 
		\end{align*} where $\sigma^2 = \sum_{j=1}^m c_j^2 (t_{j+1} - t_j)$.
	\end{proof}

	\begin{thm}
		$I(f)$ is well defined (independent of the choice of $f_n$.)
	\end{thm}

	\begin{proof}
		Let $f_n, g_n \rightarrow f$ in $L^2([0,1])$.  We then need to only compute \[
			\Delta_{n,m} = \E(I(f_n) - I(g_m) )^2 \rightarrow 0
		\] as $m,n \rightarrow \infty$.

		In fact, \begin{align*} 
			\Delta_{n,m} = \E(I(f_n - g_m))^2 = \int_0^1 (f_n - g_m)^2 \, dx \\
						&\leq 2 \int_0^1 ( f_n - f)^2 \, dx + 2 \int_0^1 (g_m - f)^2 \, dx \\
						&\rightarrow 0
		\end{align*} as $n,m \rightarrow \infty$.  
	\end{proof}

	\begin{rem}
		If $f$ is continuous bounded variation, then 
		\begin{align*}
			\int_0^1 f(s) \, dB_s = (R.S.) \int)0^1 f(s) \, dB_s \\
				&= \lim_{\delta \rightarrow 0} \sum_{j=1}^n f(t_j) B(t_{j+1} - B_{t_j})
 		\end{align*}
	\end{rem}
	\begin{proof}
	Case 1).
		\[
			\alpha I(f) + \beta I(g) = \lim_{n \rightarrow \infty} [ \alpha I(f_n) + \beta I(g_n) = \lim_{n \rightarrow \infty} [ I(\alpha f_n + \beta g_n)] = I(\alpha f + \beta g)
		\]
		and thus $\alpha f_n + \beta g_n \rightarrow \alpha f + \beta g \in L^2([0,1])$.   
	\end{proof}

	Case 2).  If $I(f) = \lim_{ \delta \rightarrow 0} I(f_n)$ in probability.  Then 
	\begin{align*}
		I(f_n) &\sim N(0, \sigma_n^2) \\
				&\sim N(0, \int_0^1 f_n^2(s) \,ds) \\
				&\rightarrow N(0, \int_0^1 f^2(s) \, ds)	
	\end{align*} if $\sigma^2_n \rightarrow \int_0^1 f^2(x) \, dx$.  In fact, as 
	\begin{align*}
		\int_0^1 f_n^2 \, dx &= \int_0^1 (f_n - f + f)^2 \, dx \\
								&= \int_0^1 f^2 \, dx + \int_0^1 (f_n - f)^2 \, dx + \int_0^1 (f_n - f) f \, dx\\
								&\rightarrow \int_0^1 f^2 \, dx
	\end{align*} as other terms tend to zero by $L^2$ convergence and H\"older's inequality.

	\begin{rem}
		$(R.S.) \int_0^1 f(s) \, dB_s$ exists if $f$ is of bounded variation.
	\end{rem}

	\begin{rem}
		If $f$ is continuous then $\int_0^1 f^2 \, dx < \infty$ and \[
			f_n(t) = \sum_{j=1}^n f(t_j) I_{[t_j, t_{j+1})} \rightarrow f(t) \, \text{in $L^2([0,1])$}
			\]

		Thus, \begin{align*}
			I(f) 	&= \lim_{\delta \rightarrow 0} I(f_n) \\
					&= \lim_{\delta \rightarrow 0} \sum_{j=1}^m f(t_j) (B_{t_{j+1}} - B_{t_j})
		\end{align*}
		We have to prove \[
			\int_0^1 (f_n - f)^2 \, dx \rightarrow 0
		\] if $f$ is continuous.
	\end{rem}

\end{defn}
% section lecture_9_tuesday_29_march (end)

\section{Lecture 10 - Thursday 31 March} % (fold)
\label{sec:lecture_10_thursday_31_march}
We have \[
	(\text{\ito})\, \int_0^1 f(s) \, dB_s \sim N(0, \int_0^1 f(t)^2 \, dt
\] if $f$ is continuous, and of finite variation.  In this case, we can write \[
	(\text{R.S.}) \, \int_0^1 f(s) \, dB_s = (\text{\ito}) \int_0^1 f(s) \, dB_s = \lim_{\Delta \rightarrow 0} \sum_{j=1}^n f(t_j) (B_{t_{j+1}} - B_{t_j})
\]

\begin{exmp}
	We have \[
		\int_0^1 (1-t) \, dB_t =  (\ito) \int_0^1 (1-t) \, dB_t \sim N(0, \int_0^1 (1-t)^2 \, dt) = N(0, \frac{1}{3})
 	\] and by integrating by parts, we have \[
 		\int_0^1 (1-t) \, dB_t = (1-t) B_t |_0^1 - \int_0^1 B_t \, d(1-t) = 0 + \int_0^1 B_t \, dt \sim N(0, \frac{1}{3})
 	\]
\end{exmp}

Now consider \[
	(\ito) \int_0^1 X_s \, dB_s
\] where we now allow $X_s$ to be a stochastic process.

Write $\sigf_t = \sigma(B_s, 0 \leq s \leq t).$  Let $\pi$ be the collection of $X_s$ such that 
\begin{enumerate}[(1)]
	\item $X_s$ is adapted to $\sigf_s$, that is, for any $s$, $X_s$ is $\sigf_s$ measurable.  
	\item $\int_0^1 X_s^2 \, ds < \infty$ almost surely (R.S)
\end{enumerate}

Let $\pi' = \{ X_s \, | \, X_s \in \pi, \int_0^1 \E(X_s^2) < \infty \}$.  Then $\pi' \subset \pi$.  Let $X_s = e^{B_s^2}$.  Then \[
	\E(X_s^2) = \E(e^{B_s^2}) = \begin{cases}
		\frac{1}{\sqrt{1-4s}} & 0 \leq s < \frac{1}{4} \\
		\infty & s \geq \frac{1}{4}
	\end{cases}
\]

\begin{defn}[\ito integral for stochastic integrands]
	We proceed in two steps.  
	\begin{enumerate}
		\item Let $X_s = \sum_{j=1}^n \zeta_i \mathbf{1}_{[t_j, t_{j+1})}$ where $\zeta_j$ is $\sigf_{t_j}$ measurable.  Then \[
			I(X) = \sum_{j=1}^n \zeta_j (B_{t_{j+1}} - B_{t_j}).
		\]
		\item If $X \in \pi$, there exists a sequence $X^n \in \pi'$ such that $X^n$ are step process with \[
			\int_{0^1} | X^n_s - X_s |^2 \, ds \rightarrow 0
		\] as $n \rightarrow \infty$ in probability or in $L^2([0,1])$ if $X_s \in \pi'$.  
	\end{enumerate}
	\begin{proof}
		We show only for $X_s$ continuous, the general case can be found in Hu-Hsing Kuo (p.65).  As $X_s$ is continuous, then it is in $\pi$.  Then choose \[
			X^n_s = X_0 + \sum{j=1}^n X_{t_j} \mathbf{1}_{[t_j, t_{j+1})}
		\]  Then $X_s \in pi'$ and \[
			\E(|X^n_s - X_s|^2) \rightarrow 0
		\] for any $s \in (0,1)$.

		We can also show that \[
			\E(|X^n_s - X_s|^2) < \infty
		\] and so by the dominated convergence theorem, we have \[
			\lim_{n \rightarrow \infty} \int_0^1 \E(|X^n_s - X_s|^2) \, ds = 0.
		\]
	\end{proof}
	\item Finally if $X_s \in \pi$, the \ito integral $\int_0^1 X_s \, dB_s$ is defined as \[
		I(X) = \lim_{\delta \rightarrow 0} I(X^n)
	\] in probability or in $L^2$ if $X \in \pi'$.  
\end{defn}
% section lecture_10_thursday_31_march (end)

\section{Lecture 11 - Tuesday 5 March} % (fold)
\label{sec:lecture_11_tuesday_5_march}
Let $I(X) = \int_0^1 X_s \, dB_s$ in the \ito sense.  We then require
\begin{enumerate}
	\item $X_s$ is $\sigf_s = \sigma \{ B_t, 0 \leq t \leq s \}$-measurable
	\item $\int_0^1 X_s^2 \, ds < \infty$ almost surely.  
\end{enumerate}

Then $I(X) = \lim_{n \rightarrow \infty} I(X_n)$ where $X_n$ is a sequence of step functions converging to $X$ in $L^2$, that is, \[
	\int_0^1 (X_s^n - X_s)^n \, ds \rightarrow 0
\]

We then show that this definition is independent of the sequence of step functions.  For any $Y_s$ step process, we have \[
	\int_0^1 (Y_s^n - Y_s^m)^2 \, dx \leq 2\int_0^1 (Y_s^m - X_s)^2 \, ds + 2 \int_0^1 (X_s^n - X_s)^2 \, dx \rightarrow 0
\]

\begin{thm}[Properties of the \ito integral]{\ }
	\begin{enumerate}
		\item For any $\alpha, \beta \in \R$, $X, Y \in \pi$, \[
			I(\alpha X + \beta Y) = \alpha I(X) + \beta I(Y)
		\]
		\item For any $X_s \in \pi'$, we have \[
			\E(I(X)) = 0, \quad \E(I^2(X)) = \int_0^1 E(X_s^2) \, ds
		\]  If $X' \in \pi'$, $Y_s \in \pi'$, then \[
			\E(I(X) I(Y) ) = \int_0^1 \E(X_s Y_s) \, ds
		\]
		\item If $X_s$ continuous then \[
			I(X) = \lim_{\delta \rightarrow 0} \sum_{j=1}^n X_{t_j} (B_{t_{j+1}} - B_{t_j})
		\] in probability.
		\item If $X$ is continuous and of finite variation and $\int_0^1 B_s \, dX_s < \infty$ then \[
			(R.S.) \int_0^1 = (\ito) \int_0^1 X_s \, dB_s = X_1 B_1 - X_0 B_0 - \int_0^1 B_s \, dX_s
		\]
	\end{enumerate}
\end{thm}
\begin{prop}
	We now show why we require
	\begin{enumerate}
		\item $X_s$ is $\sigf_s = \sigma \{ B_t, 0 \leq t \leq s \}$-measurable
		\item $\int_0^1 X_s^2 \, ds < \infty$ almost surely.  
	\end{enumerate}
\end{prop}
\begin{proof}
	Motivation for (1)
	\begin{align*}
		X_s &= \sum_{j=1}^n \zeta_j \mathbf{1}_{(t_j, t_{j+1})} \\
		I(X) &= \sum_{j=1} \zeta_j (B_{t_{j+1}} - B_{t_j}) \\
		\E(I(X)) &= \sum_{j=1}^n \E(\zeta_j (B_{t_{j+1}} - B_{t_j})) \\
					&= \sum_{j=1}^n \E(\E(\zeta_j (B_{t_{j+1}} - B_{t_j})\, | \, \sigf_{t_j})) \\
					&= \sum_{j=1}^n \E(\zeta_j  \E(B_{t_{j+1} - B_{t_j}} \, | \, \sigf_{t_j}))  
	\end{align*}

	Motivation for (2)
	\begin{align*}
		\E( I^2(X)) &= \E(\sum_{j=1}^n \zeta_j (B_{t_{j+1}} - B_{t_j}))^2 \\
					&= 2 \sum_{i < j} \E(\E( \zeta_i \zeta_j (B_{t_{j+1}} - B_{t_j})(B_{t_{i+1}} - B_{t_i})) \, | \, \sigf_{t_i} ) + \sum_{j=1}^n \E(\zeta_j^2 (B_{t_{j+1}} - B_{t_{j}})^2 )  \\
					&= \sum_{j=1}^n \E(\zeta_j^2) (t_{j+1} - t_{j}) \\
					&= \int_0^1 \E(X_t^2) \, dt				
	\end{align*}

	We now show that there $X_s^n \in \pi'$.  such that $\int_0^1 \E(X_s^n - X_s)^2 \, ds \rightarrow 0$.  We have \begin{align*}
		\E(I^2(X)) &= \E(I(X)) - I(X^n) + \E(I(X^n_s))^2 \\
				&= \E(I^2(X^n)) + 2 \E(I(X^n)(I(X) - I(X^n)) ) + \E(I(X) - I(X^n))^2 \\ 
	\end{align*}

	By Cauchy-Swartz, the middle term tens do zero, and by definition, the third term tends to zero.  Thus we have 
	\[
		\lim_{n \rightarrow \infty} \E(I^2(X_n)) = \E(I^2(X))
	\]

	Let $X_s^n$ be step processes.  We now show \[
		\E(X_s^n - X_s)^2 \rightarrow 0 \Rightarrow \int_0^1 \E(X_s^n - X_s)^2 \, ds \rightarrow 0
	\] 
	By definition, we have \[
		I(X) = \lim_{n \rightarrow \infty} I(X^n) = \lim_{\delta \rightarrow 0} \sum_{j=1}^n X_{t_j} (B_{t_{j+1}} - B_{t_j}).  
	\] and we proved the required result in lectures.

	We now show that the $(R.S.)$ integral exists if $X$ is continuous and of finite variation, and $\int_0^1 B_s \, dX_s < \infty$, the our integration by parts formula holds.

	  We have 
	\begin{align*}
		(R.S.) \int_0^1 X_s \, dB_s &= \lim_{\delta \rightarrow 0} \sum_{j=1}^n X_{y_j} (B_{t_{j+1}} - B_{t_j}) \\
		&= \lim_{\delta \rightarrow 0} \sum_{j=1}^n X_{t_j}  (B_{t_{j+1}} - B_{t_j})
	\end{align*} 
	which is our \ito integral by definition.

 \end{proof}

\begin{defn}[\ito process]
	Suppose $Y_t = \int_0^t X_s \, dB_s$, $t \geq 0$ is well defined, for $X_s \in \pi$. Then $Y_t$ is an \emph{\ito processes}.  To show a process $Y_t$ is an \ito process, we need to show that $\int_0^t |X_s| \, ds) M \infty$ a.s. and $\int_0^t |X_s|^2 \, ds < \infty$ a.s.
\end{defn}

\begin{thm}
	We have that $Y_t$ is continuous (except on a null set), is of infinite variation, as \begin{align*}
		\sum_{j=1}^n |Y_{t_{j+1}} - Y_{t_j} | &= \sum_{j=1}^n | \int_{t_j}^{t_{j+1}} X_s \, dB_s | \\
		 &\geq	\sum_{j=1}^n \min_{s} |X_s| |B_{t_{j+1} - B_{t_j}} \\
		&\geq C \sum_{j=1^n} |B_{t_{j+1}} - B_{t_j} |
	\end{align*}
\end{thm}
% section lecture_11_tuesday_5_march (end)

\section{Lecture 12 - Thursday 7 March} % (fold)
\label{sec:lecture_12_thursday_7_march}
From before, consider the \ito process $Y_t = \int_0^t X_s \, dB_s$.

\begin{lem}
	$\E(\int_s^t X_u dB_u \given \sigf_s) = 0$. 
\end{lem}
\begin{proof}
	Let $X_u = \sum_{j=1}^n \zeta_j \mathbf{1}_{[t_j, t_{j+1})}$.  Then 
	\begin{align*}
		\E(\int_s^t X_u \, dB_u \given \sigf_s) = \sum_{j=1}^n \E(\zeta_j (B_{t_{j+1}} - B_{t_j}) \given \sigf_s ) \\
					&= \sum_{j=1}^n \E(\E(\zeta_j (B_{t_{j+1}} - B_{t_j}) \given \sigf_{s}) \given \sigf_{t_j}) \\
		 			&= \sum_{j=1}^n \E(\E(\zeta_j (B_{t_{j+1}} - B_{t_j}) \given \sigf_{t_j}) \given \sigf_s) \\
					&= 0
	\end{align*} where the third equality follows from the fact that $(\sigf_t)$ is an increasing sequence of $\sigma$-fields and the final follows from the fact that $B_{t_{j+1}} - B_{t_{j}}$ is independent of $\sigf_{t_j}$.  
\end{proof}

\begin{defn}[Local martingale]
	A process $Y_t$ is a local martingale if there exists a sequence of stopping times $\tau_n, n \geq 1$ such that $Y_{\min(t, \tau), \sigf_t}$ is a martingale.  
\end{defn}

\begin{prop}We have the following.  
	\begin{enumerate}
		\item If $X_s \in \pi'$, that is, $\int_0^t \E(X_s^2) \, ds < \infty$, then $(Y_t, \sigf_t)$ is a martingale.
		\item If $X_s \in \pi$, that is $\int_0^t X_s^2 \, ds < \infty $ a.s., then $(Y_t, \sigf_t)$ is a local martingale.  
		\item For $f(x)$ satisfying $\int_0^t f^2(z) \, dz < \infty$., we have \[
			Y_t = \int_0^t f(s) \, dB_s
		\] is a Gaussian process.  
	\end{enumerate}
\end{prop}

\begin{proof}
	We have $Y_t$ is trivially $\sigf_t$-measurable.

	We have $\E(|Y_t|) < \infty $ a.s. as $\E(Y_t^2) = \int_0^t \E(X_s^2) \, ds < \infty$ by assumption.

	We have \begin{align*}
		\E(Y_t \given \sigf_s) &= \E(\int_0^s X_u \, dB_u + \int_s^t X_u \, dB_u \given \sigf_s ) \\
		&= \E(Y_s \given \sigf_s) + \E( \int_s^t X_u \, dB_u \given \sigf_s) \\
		&= Y_s			
	\end{align*}  from the previous lemma.

	Now assuming $X_s \in \pi'$, the exists an $X^n_s$ a step process such that \[
		\int_0^t \E(X^n_u - X_u)^2 \, du \rightarrow 0
	\] as $n \rightarrow \infty$.

	Set $Y^n_t = \int_0^t X_u^n \, dB_u$.  Then \begin{align*}
		\int_s^t X_u \, dB_u = Y_t - Y_s \\
							&= Y_t - Y_t^n + Y_t^n - Y_s^n + Y_s^N - Y_s \\			
	\end{align*}
	Then for each $n$, we have \begin{align*}
		Z = \E(\int_s^t X_u \, dB_u ) &= \E(Y_t - Y_t^n \given \sigf_s) + \E(Y_s^n - Y_s \given \sigf_s) 
	\end{align*}

	We only need to prove $\E(|Z|^2) \rightarrow 0$ which implies $\E(Z^2) = 0$ and thus $Z = 0$ almost surely.  We have \[
		\E(Z^2) \leq 2\E(Y_t - Y_t^n)^2 + 2\E(Y_s - Y_s^n)^2 \rightarrow 0
	\] by definition of $Y_t^n$.

\end{proof}

% section lecture_12_thursday_7_march (end)

\section{Lecture 13 - Tuesday 12 March} % (fold)
\label{sec:lecture_13_tuesday_12_march}

\begin{thm}
	Let $f$ be continuous.  Then \[
		\lim_{\delta \rightarrow 0 } \sum_{j=1}^n f(\theta_j) (B_{t_{j+1}} - B_{t_j})^2 = \int_0^t f(B_s) \, ds
	\]
\end{thm}
\begin{proof}
	Let \[
		Q_n = \sum_{j=1}^n |f(\theta_j)  - f(B_{t_j}) |B_{t_{j+1}} - B_{t_j}|^2  
	\]
	Note that $\sum_{j=1}^n f(B_{t_j}) (B_{t_{j+1}} - B_{t_j})^2 \rightarrow \int_0^t f(B_s) \, ds$ in probability.

	It only needs to show that $Q_n \rightarrow 0$ in probability.   We have \[
		Q_n \leq \max_{1 \leq j \leq n} | f(\theta_j) - f(B_{t_j})\sum_{j=1}^n |B_{t_{j+1}} - B_{t_j}|^2 \rightarrow 0 \cdot t = 0
	\] in probability from the quadratic variance of the brownian motion 
\end{proof}

\begin{thm}
	Let $f$ be bounded on $[0,1]$.  Then \[
		\lim_{\delta \rightarrow 0} \sum_{j=1}^n f(X_{t_j})(B_{t_{j+1}} - B_{t_j})^2 = \int_0^t f(X_s \, ds)
	\]
\end{thm}
\begin{proof}
	We only prove for cases where $\int_0^t \E(X_s^2) \, ds$ is finite.  In this case, there exists $X_s^n$ step process such that \[
		\int_0^t \E(X_s^n - X_s)^2 \, ds \rightarrow 0
	\] as $n \rightarrow \infty$.

	In fact, we can let $X_s^n = \sum_{j=1}^n X_{t_j}  \mathbf{1}_{[t_j, t_{j+1}]}$.  Let \[
		Y_t^n = \int_0^t X_s^n \, dB_s
	\]  Then $Y_{t_{j+1}}^n - Y_{t_j}^n = \int_{t_j}^{t_{j+1}} X_s^n \, dB_s$.

	Therefore, \[
		\sum_{j=1}^n (Y_{t_{j+1}} - Y_{t_j})^2 = \sum_{j=1}^n (Y_{t_{j+1}} - Y^n_{t_{j}} + Z_{1j} + Z_{2j} )
	\] where $Z_{1j} = Y_{t_{j+1}} - Y_{t_{j+1}}^n$ and $Z_{2j} = Y_{t_j} = Y_{t_{j}}^n$.  Continuing, we have \begin{align*}
		&= \sum_{j=1}^n (Y_{t_{j+1}} - Y_{t_{j}}^n)^2 + \text{error} \\
		&= \sum_{j=1}^n X_{t_j}^2 (B_{t_{j+1}} - B_{t_j})^2 + \text{error} \\
		&\rightarrow \int_0^t X_s^2 \, ds
	\end{align*} if the error term goes to zero.  We have \begin{align*}
		R_n &\leq 2 \sum Z_{1j}^2 + 2 \sum Z_{2j}^2 + 2 \sum |Z_{1j} | \cdot | Y_{t_{j+1}}^n - Y_{t_j}^n | + 2 \sum |Z_{2j} | \cdot | Y_{t_{j+1}}^n - Y_{t_j}^n | + 2 \sum |Z_{1j} | \cdot |Z_{2j} | \\
		&\leq \dots + 2 (\sum Z_{1j}^2 )^{1/2} A_n^{1/2} + 2 (\sum Z_{2j}^2 )^{1/2} A_n^{1/2} + 2  (\sum Z_{1j}^2 )^{1/2} \cdot  (\sum Z_{2j}^2 )^{1/2} 
	\end{align*}
	and as $\sum {Z_{ij}}^2 \rightarrow 0$ in probability, we have our result.
\end{proof}

\newcommand{\itos}{\text{It\^o's\ }}
\begin{thm}[\itos first formula]
	If $f(x)$ is a twice-differentiable function then for any $t$, \[
		f(B_t) - f(B_s) = \int_s^t f'(B_u) \, dB_u + \frac{1}{2 } \int_s^t f''(B_u) \, du
	\]
\end{thm}
\begin{proof}
	Let $s = t_1, \dots, t_{n} = t$.  We have \[
		f(B_t) - f(B_s) = \sum_{j=1}^n \left[ f(B_{t_{j+1}}) - f(B_{t_j}) \right]
	\]

	Applying Taylor's expansion to $f(B_{t_{j+1}}) - f(B_{t_{j}})$ we get \[
		f(B_{t_{j+1}}) - f(B_{t_j} = f'(B_{t_j}) (B_{t_{j+1}} - B_{t_j}) + \frac{1}{2} f''(\theta_j) (B_{t_{j+1}} - B_{t_j})^2 
	\] and so \begin{align*}
		f(B_t) - f(B_s) &= \lim_{\delta \rightarrow 0} \sum_{j=1}^n 	f'(B_{t_j}) (B_{t_{j+1}} - B_{t_j}) + \lim_{\delta \rightarrow 0} \sum_{j=1} \frac{1}{2} f''(\theta_j) (B_{t_{j+1}} - B_{t_j})^2  \\
		&= \int_s^t f'(B_u) \, dB_u + \frac{1}{2} \int_s^t f''(B_u) \, du
	\end{align*}
\end{proof}

\begin{exmp}

\end{exmp}

% section lecture_13_tuesday_12_march (end)
\section{Lecture 14 - Thursday 14 March} % (fold)
\label{sec:lecture_14_thursday_14_march}
\begin{defn}[Covariation]  The covariation of two stochastic processes $X_t, Y_t$ is defined as \[
	[X,Y](t) = \frac{1}{4}( [X+Y, X+Y](t) - [X-Y, X-Y](t))
\] where $[\cdot, \cdot]$ is the quadratic variation previously defined.
\end{defn}

\begin{defn}[Stochastic differential equation]
	Let $X_t$ be an \ito process.  Then \[
		X_t = X_a + \int_a^t \mu(s) \, ds + \int_a^b \sigma(s) \, dB_s
	\]  We write \[
		dX_t = \mu(t) \, dt + \sigma(t) \, dB_s
	\]

	By convention, we write $dX_t \cdot dY_t = d[X,Y](t)$.  In particular, $(dY_t)^2 = d[Y,Y](t)$.  
\end{defn}

\begin{thm}
	Let $Y_t$ be path continuous, and let $X_t$ have finite variation.  Then \[
		[X,Y](t) = 0.
	\]
\end{thm}
\begin{proof}
	We have \begin{align*}
		[X,Y](t)	&= \frac{1}{4}( [X+Y, X+Y] - [X-Y, X-Y]) \\
					&= \lim_{\delta \rightarrow 0} \sum_{j=1}^n (X_{t_{j+1}} - X_{t_j})(Y_{t_{j+1}} - Y_{t_j}) \\
					&\leq \lim_{\delta \rightarrow 0} \max | Y_{t_{j+1}} - Y_{t_j} | \sum_{j=1}^n | X_{t_{j+1}} - X_{t_j} | \\
					&\rightarrow 0 		
	\end{align*} as $Y_t$ is path continuous and $X_t$ has finite variation.
\end{proof}

\begin{cor}
	From this theorem, we then have \[
		dB_t \cdot dt = 0, \quad (dt)^2 = 0, \quad (dB_t)^2 = dt 
	\]
\end{cor}

\begin{cor}
	For an \ito process $X_t$ given above, we then have \[
		d[X,X](t) = dX_t \cdot dX_t = \left(\mu(t) dt + \sigma(t) \, dB_t \right)^2 = \sigma^2(t) \, dt
	\]
\end{cor}
\begin{cor}
	If $f(x)$ has a twice continuous derivative, then \[
	[f(B_t), B_t](t) = \int_0^t f'(B_s) \, ds
	\]
\end{cor}
\begin{proof}
	From \itos formula, we have \[
		df(B_t) = f'(B_t) \, dB_t + \frac{1}{2} f''(B_t) \, dt
	\]  Then from above, we have \[
		d[f(B_t), B_t] (t) = df(B_t) \cdot dB_t	= f'(B_t) dt
	\]
\end{proof}
\begin{thm}[\itos lemma]
	By Taylor's theorem, we have \begin{align*}
		df(t, X_t) &= \frac{\partial f(t, X_t)}{\partial t} dt + \left.\frac{\partial f(t, x)}{\partial x}  \right|_{x = X_t} \cdot dX_t + \left.\frac{1}{2} \frac{\partial^2 f(t, x)}{\partial x^2} \right|_{x = X_t} \cdot (dX_t)^2 \\
		&\qquad + \left.\frac{1}{2} \frac{\partial^2 f(t, x)}{\partial x \partial t} \right|_{x = X_t} \cdot dX_t \cdot dt \\
		&= \frac{\partial f(t, X_t)}{\partial t} dt + \left.\frac{\partial f(t, x)}{\partial x}  \right|_{x = X_t} \cdot (\mu(t) \, dt + \sigma(t) \, dB_t) + \left.\frac{1}{2} \frac{\partial^2 f(t, x)}{\partial x^2} \right|_{x = X_t} \cdot \sigma^2 \, dt \\
		&= \left( \frac{\partial f}{\partial t} + \frac{\partial f}{\partial x} \mu(t) + \frac{1}{2} \frac{\partial^2 f}{\partial x^2} \sigma^2(t) \right) \, dt + \frac{\partial f}{\partial x} \sigma(t) \, dB_t
	\end{align*}
\end{thm}

\begin{exmp}
	Let $f(B_t) = e^{B_t}$. Then \[
		df(B_t) = e^{B_t} \, dB_t + \frac{1}{2} e^{B_t} \, dt
	\]
\end{exmp}

\begin{thm}
	Assume $\int_0^T f^2(t)\, dt < \infty$.  Let $X_t = \int_0^t f(s) \, dB_s - \frac{1}{2}\int_0^t f^2(s) \, ds$.  Then \[
		Y_t = e^{X_t}
	\]  is a martingale with respect to $\sigf_t = \sigma(B_s, 0 \leq s \leq t)$. 
\end{thm}
\begin{proof}
	We have \begin{align*}
		dY_t 	&= e^{X_t} \, dX_t + \frac{1}{2} e^{X_t} (dX_t)^2 \\
				&= e^{X_t}[f(t) \, dB_t - \frac{1}{2} f^2(s) \, dt] + \frac{1}{2} e^{X_t} f^2 \, dt \\
				&= e^{X_t} f(t) \, dB_t
 	\end{align*} and thus $Y_t = \int_0^t e^{X_s} f(s) \, dB_s$ which is a martingale from previous work.
\end{proof}

% section lecture_14_thursday_14_march (end)
\section{Lecture 15 - Tuesday 19 March} % (fold)
\label{sec:lecture_15_tuesday_19_march}

\begin{thm}[Multivariate \itos formula]
	Let $B_j(t)$ be a sequence of independent Brownian motions.  Consider the \ito processes $X_t^i$, with \[
		dX_t^i = b_i(t) \, dt + \sum_{k=1}^m \sigma_{ki}(t) \, dB_k(t)
	\]  Suppose $f(t, x_1, \dots, x_n)$ is a continuous function of its components and has continuous partial derivatives $\frac{\partial f}{\partial t}, \frac{\partial f}{\partial x_i}, \frac{\partial f}{\partial^2 x_i \partial x_j}$.  Then \[
		df(t, X_t^1, \dots, X_t^m) = \frac{\partial f}{\partial t} \, dt + \sum_{i=1}^m \frac{\partial f}{\partial x_i} \, dX_t^i + \frac{1}{2} \sum_{i, j = 1} \frac{\partial^2 f}{\partial x_i \partial x_j} \, dX_t^i \, dX_t^j
	\]

	We have \begin{align*}
		dX_t^i \, dX_t^j 	&= \sum_{k, s = 1}^m \sigma_{ki} \sigma_{sj} \, dB_k(t) \, dB_{s}(t) \\
							&= \sum_{k=1}^n \sigma_{ki}\sigma_{kj} \, dt + 
	\end{align*} as $d[B_i, B_j](t) = 0$ when $B_i, B_j$ are independent 
\end{thm}

\begin{exmp}
	Let \begin{align*}
		dX_t &= \mu_1(t) dt + \sigma_1(t) \, dB_1(t) \\
		dY_t &= \mu_2(t) dt + \sigma_2(t) \, dB_2(t)
	\end{align*} with $B_1, B_2$ independent. Then \begin{align*}
		d(X_t \cdot Y_t) = Y_t \, dX_t + X_t \, dY_t + d[X,Y](t)
	\end{align*} and by independence, $d[X,Y](t) = dX_t \cdot dY_t = 0$.
\end{exmp}

\begin{exmp}
	Let \begin{align*}
		dX_t &= \mu_1(t) dt + \sigma_1(t) \, dB_1(t) \\
		dY_t &= \mu_2(t) dt + \sigma_2(t) \, dB_1(t).
	\end{align*} Then \begin{align*}
		d(X_t \cdot Y_t) = Y_t \, dX_t + X_t \, dY_t + \sigma_1(t) \sigma_2(t) \, dt.
	\end{align*}

	In particular, if $\mu_1 = \mu_2 = 0$, then \[
		X_t \, Y_t = \int_0^t [ \sigma_1(s) Y_s + \sigma_2(s) X_s] \, dB_s + \int_0^t \sigma_1(s) \sigma_2(s) \, dt
	\]

	Thus, $Z_t = X_t Y_t - \int_0^t \sigma_1(s) \sigma_2(s) \, dt$ is a martingale.
\end{exmp}

\begin{thm}[Tanaka's formula]
	We have \[
		|B_t - a | = |a| + \int_0^t \text{sgn}(B_s - a) \, dB_s + \mathcal L(t, a)
	\] where \[\mathcal L(t,a) = \lim_{\epsilon > 0} \frac{1}{2 \epsilon} \int_0^t \mathbf{1}_{|B_s - a| \leq \epsilon} \, ds
 	\]
\end{thm}
\begin{proof}
	Let \[
		f_\epsilon(x) = \begin{cases}
			|x-a| - \frac{\epsilon}{2} &|x-a| > \epsilon \\
			\frac{1}{2 \epsilon}(x-a)^2 &|x-a| \leq \epsilon
		\end{cases}
	\]  Then we have \[
		f'_\epsilon(x) = \begin{cases}
			1 &x > a + \epsilon \\
			\frac{1}{\epsilon}(x-a) &|x-a| \leq \epsilon \\
			-1 &x < a - \epsilon
		\end{cases}
	\] and \[
	f''_\epsilon(x) = \begin{cases}
		0 &|x-a| > \epsilon \\
		\frac{1}{\epsilon} &|x-a| \leq \epsilon
	\end{cases}	
	\]

	Then by \itos formula, we have \[
		f_\epsilon(B_t) = f_\epsilon(0) + \int_0^t f'_\epsilon(B-s) \, dB_s + \frac{1}{2} \int_0^t f''_\epsilon(B_s) \, ds	\]

	Obviously \[
		\lim_{\epsilon \rightarrow 0} f_\epsilon(0) = |a|
	\] and \[
		\lim_{\epsilon \rightarrow 0} \frac{1}{2} \int_0^t f''_\epsilon(B_s) \, ds = \frac{1}{\epsilon} \int_0^t \mathbf{1}_{|B_s - a| \leq \epsilon} \, ds = \mathcal L(t, a)
	\]

	Note that \[
		\int_0^t \left| f'_\epsilon(B_s) - \text{sgn}(B_s  - a) \right|^2 \, ds =  \int_{|B_s - a \leq \epsilon} \left| \frac{1}{\epsilon}(B_s - a) - \text{sgn}(B_s - a) \right|^2 \, ds \rightarrow 0 a.s.
	\]
\end{proof}

\begin{thm}
	\[
		\mathcal L(t,a) = \int_0^t \delta_a(B_s) \,ds
	\]
\end{thm}
\begin{thm}
	If $f$ is integrable on $\R$, then \[
		\int_{-\infty}^\infty \mathcal L(t,s) f(s) \, ds = \int_0^t f(B_s) \ ds
	\]
\end{thm}
% section lecture_15_tuesday_19_march (end)

\section{Lecture 16 - Thursday 21 March} % (fold)
\label{sec:lecture_16_thursday_21_march}
\begin{defn}[Linear cointegration]
	Consider two non stationary time series $X_t$ and $Y_t$.  If there exist coefficients $\alpha$ and $\beta$ such that \[
		\alpha X_t + \beta Y_t = u_t
	\] with $u_t$ stationary, then we say that $X_t$ and $Y_t$ are \textbf{cointegrated}.  
\end{defn}

\begin{defn}[Nonlinear cointegration]
	If $Y_t - f(X_t) = u_t$ is stationary, with $f(\cdot)$ a nonlinear function.   
\end{defn}

% If $\hat f_n(x) = \frac{\sum Y_t K\left( \frac{x_t - x}{h} \right)}{\sum K\left( \frac{x_t - x}{h} \right)}$

% section lecture_16_thursday_21_march (end)

\section{Lecture 17 - Tuesday 3 May} % (fold)
\label{sec:lecture_17_tuesday_3_may}
\subsection{Stochastic integrals for martingales} % (fold)
\label{sub:stochastic_integrals_for_martingales}
We now seek to define stochastic integrals with respect to processes other than Brownian motion.

\begin{exmp}
	Let $X_t = \int_0^t Y_s \, dB_s$, and thus $dX_s = Y_s \, dB_s$.  Then \[
		Z_t = \int_0^t Y'_s \, dX_s = \int_0^t Y'_s \cdot Y_s \, dB_s
	\]
\end{exmp}

Revus and Yov - Continuous Martingale and Brownian Motion

\begin{defn}[Martingale]
A martingale with respect
\begin{enumerate}[(1)]
	\item $M_t$ adapted to $\sigf_t$.
	\item $\E(|M_t|) < \infty$.
	\item $\E(M_t \given \sigf_s) = M_s$ a.s.
\end{enumerate}

A process is a submartingale if (3) is replace with $\E(M_t \given \sigf_s) \geq M_s$.  A process is a supermartingale if (3) is replaced with $\E(M_t \given \sigf_s) \leq M_s$.

\end{defn}

\begin{exmp}
	Let $N_t$ be a poisson process with intensity $\lambda$.  Then 
	\begin{enumerate}
		\item $N_t$ is a submartingale with respect to the natural filtration.
		\item $N_t - \lambda t$ is a martingale with respect to the natural filtration.  
	\end{enumerate}
\end{exmp}

\begin{thm}
	If $\int_0^t \E(H^2(s)) \, ds < \infty$ and $H(s)$ is adapted to $\sigf_s = \sigma( B_t, t \leq s)$, then \[ Y_t = \int_0^t H(s) \, dB_s, t \geq 0
	\]  is a continuous, square integrable martingale - that is, $\E(Y_t^2 < \infty)$.
\end{thm}

\begin{thm}
	Let $M_t$ be a continuous, square integrable martingale with respect to $\sigf_t$.  Then there exists an adapted process $H(s)$ such that $\int_0^t \E(H^2(s)) \, ds < \infty$ and \[
		M_t = M_0 + \int_0^t H(s) \, dB_s
	\] where $B_t$ is a Brownian motion with respect to $\sigf_t$.
\end{thm}

\begin{thm}
	$M_t, t \geq 0$ is a Brownian motion if and only if it is a local continuous martingale with $[M, M](t) = t, t \geq 0$ under some probability measure $Q$.
\end{thm}
\begin{proof}
	A local continuous martingale is of the form $M_t = M_0 + \int_0^t H(s) \, dB_s$.  Then we have \[
		[M, M](t) = \int_0^t H^2(s) \, ds = t \Rightarrow H(s) = 1 a.s. \Rightarrow M_t = B_t.
	\]
\end{proof}

\begin{thm}
	Let $M_t, t \geq 0$ e a continuous local martingale such that $[M, M](t) \uparrow \infty$.  Let \[
		\tau_t = \inf \{ s : [M, M](s) \geq t \}.
	\] Then $M(\tau_t)$ is a Brownian motion.  Moreover, $M(t) = B([M, M](t)), t \geq 0$.

	This is an application of the \textbf{change of time} method.
\end{thm}

\begin{exmp}
	$B_t$ is a Brownian motion - and then $Y_t = B_t^2 - t$ is a martingale.  We have \[
		dY_t = H(s) \, dB_s = 2 B_s \, dB_s.  
	\]  Thus, \[
		B_t^2 - t = 2 \int_0^t B_s \, dB_s.
	\]
\end{exmp}

\begin{defn}[Predictability of a stochastic process]
	A stochastic process $X_t, t \geq 0$ is said to be predictable with respect to $\sigf_t$ if $X_t \in \sigf_{t-}$ for all $t \geq 0$, where \[
		\sigf_{t-} = \bigcap_{h \geq 0}^\infty \sigf_{t + h}, \qquad \sigf_{t-} = \sigma \left( \bigcap_{h > 0}^\infty \sigf_{t - h} \right).
	\] 
\end{defn}

\begin{exmp}
	Let $g_t$ be a step process, with \[
		g_t = \sum_{i=1}^n \zeta_j \mathbf{1}_{[t_j, t_{j+1})}(t)
	\] Then $g_t$ is not predictable.

	Let $g_t$ be a step process, with \[
		g_t = \sum_{i=1}^n \zeta_j \mathbf{1}_{(t_j, t_{j+1}]}(t)
	\] Then $g_t$ is predictable.  
\end{exmp}

\begin{exmp}
	Let $N_t$ be a Poisson process.  Then $N_{t-}$ is predictable, but $N_t$ is not predictable.
\end{exmp}

From now on, assume $M_t, t \geq 0$ is right continuous, square integrable martingale with left hand limits.

\begin{lem}
	$M_t^2$ is a submartingale.
\end{lem}
\begin{proof}
	\begin{align*}
		\E(M_t^2 \given \sigf_s) = \E(M_s^2 + 2(M_t - M_s) + (M_t - M_s)^2 \given \sigf_s) \\
		&= M_s^2 + \E((M_t - M_s)^2 \given \sigf_s) \\
		&\geq M_s^2
	\end{align*}
\end{proof}

\begin{thm}[Doob-Myer decomposition]
	By Doob-Myer we can write \[
		M_t^2 = L_t + A_t
	\] where $L_t$ is a martingale, and $A_t$ is a predictable process, right continuous, and increasing, such that $A_0 = 0, \E(A_t) < \infty, t \geq 0$.

	$A_t$ is called the \textbf{compensator} of $M_t^2$, and is denoted by $\langle M, M \rangle(t)$.
\end{thm}
\begin{exmp}
	Consider $B_t^2 = B_t^2 - t + t$.  Then \[
		\langle B, B \rangle(t) = t = [B, B](t)
	\]
\end{exmp}

\begin{thm}
	If $M_t$ is continuous then \[
		[M, M](t) = \langle M , M \rangle (t).
	\]
\end{thm}

\begin{exmp}
	Let $N_t$ be a Poisson process.  Then we know $\tilde N_t = N_t - \lambda t$ is a martingale.  We may prove \[
		\tilde N_t^2 - \lambda t
	\] is a martingale, that is, $\langle \tilde N, \tilde N \rangle(t)=\lambda t$.  However, \[
		[\tilde N, \tilde N](t) = \lambda t + \tilde N_t \neq \langle \tilde N, \tilde N \rangle (t)
	\]
\end{exmp}
\begin{exmp}
	$X_t = \int_0^t f(s) \, dB_s$ is a continuous martingale.  Thus, \[
		[X, X](t) = \int_0^t f^2(s) \, ds = \langle X, X \rangle(t)
	\]	
\end{exmp}

\begin{thm}
	If $M_t$ is a continuous, square integrable martingale, then \[
		M_t^2 - [M, M](t)
	\] is a martingale, and so \[
		[M, M](t) = \langle M, M \rangle(t) + \text{martingale}
	\] which implies \[
		\E[M, M](t) = \E\langle M , M \rangle(t) = \E M_t^2
	\]
\end{thm}

We now turn to defining integrals such as \[
	\int_0^t X_s \, dM_s
\] where $M_s$ is a martingale.

\begin{defn}[]
	Let $L^2_{pred}$ be the space of all predictable stochastic process $X_s$ satisfying the condition \[
		\int_0^t X_s^2 \, d \langle M, M \rangle (s) < \infty.
	\] Then the integral $\int_0^t X_s \, dMs$ is defined as before in two steps.
	\begin{enumerate}[(1)]
		\item If $X_s \in L^2_{pred}$ and $X_s = \sum_{j=1}^n \zeta_j \mathbf{1}_{[t_j, t_{j+1})}$.  Define \[
			I(X) = \sum_{j=1}^n \zeta_j (M_{t_{j+1}} - M_{t_j}).
		\] 
		\item For all $X_s \in L^2_{pred}$, there exists a sequence of step process $X_s^n$ such that $X_s^n \rightarrow X_s$ in $L^2$.  Define $I(x)$ to be the limit in such situations such that \[
			\E(I(X) - I(X^n))^2 \rightarrow 0.
		\]
	\end{enumerate}
\end{defn}
\begin{prop}
	Properties of the integral.  
	\begin{enumerate}
		\item If $M_s$ is a (local) martingale, then \[
			\int_0^t f(s) \, dM_s
		\] is a (local) martingale.

		\begin{proof}
			\[
				\E(\int_s^t f(u) \, dM_u \given \sigf_s) = 0
			\]
		\end{proof}
		\item If $M_t$ is a square integrable martingale and satisfies \[
			\E( \int_0^t f^2(s)\,  d\langle M, M\rangle (t) ) < \infty
		\] then \[
			I(f) = \int_0^t f(s) \, dM_s 
		\] is square integrable with $E(I(f)) = 0$, and $E(I^2(f)) = \int_0^t f^2(s) \, d\langle M, M \rangle(s)$.

		In particular, if $M(s) = \int_0^s \sigma(u) \, dB_u$, then \[
			\int_0^t X_s \, dM_s = \int_0^t X_s \sigma(s) \, dB_s
		\] provided $\int_0^t X_s^2 \sigma^2(s) \, ds < \infty$ and $\int_0^t \sigma^2(s) \, ds$
		\item If $X_t = \int_0^t f(x) \, dM_s$ and $M_s$ is a continuous, square integrable martingale, then \[
			[X, X](t) = \int_0^t f^2(s) \, d[M, M](s) = \langle X, X \rangle(t)
		\]
	\end{enumerate}
\end{prop}

% subsection stochastic_integrals_for_martingales (end)
% section lecture_17_tuesday_3_may (end)

\section{Lecture 18 - Tudsay 10 May} % (fold)
\label{sec:lecture_18_wednesday_10_may}

Let $M_t$ be a martingale.  Then $M_t^2 - [M, M](t)$ is a martingale, and \[
	M_t^2 = \text{martingale} + \langle M, M \rangle(t)
\]

Recall that if $M_t$ is a continuous square integrable martingale, then \[
	\langle M, M \rangle(t) = [M, M](t)
\]  Generally speaking, \[
	[M, M](t) = \langle M, M \rangle (t) + \text{martingale}
\]

\begin{thm}
	 If \[
		\in_0^t f^2(s) \, d\langle M , M \rangle(s) < \infty
	\] a.s. then \[
		Y_t = \int_0^t f(s) \, dM_s
	\] is well defined, and \begin{align*}
		\E Y_t^2 &= \int_0^t f^2(s) \, d\langle M, M \rangle
	(t) \\
	&= \int_0^t f^2(s) \, d[M,M](t) \quad \text{if $M_t$-continuous}
	\end{align*} and \[
		\langle Y, Y \rangle(t) = [Y, Y](t) = \int_0^t f^2(s) \, d[M, M](s).  
	\]
\end{thm} 
\begin{proof}
	\begin{align*}
		dY_t &= f(t) \, dM_t \\
		d[Y, Y](t) &= dY_t dY_t \\
					&= f^2(t) \, dM_t dM_t \\
					 &= f^2(t) \, d[M, M](t) \\
		[Y, Y](t) &= \int_0^t f^2(s) \, d[M, M](s) = \langle Y, Y \rangle(t)  
	\end{align*}

	By \itos formula, we also have \begin{align*}
		Y_t^2 &= Y_0^2 + 2 \int_0^t Y_s \, dY_s + \int_0^t 1 \cdot [ Y, Y ](t)  \\
				&= 2 \int_0^t Y_s \cdot f(s) \, dM_s + \langle Y, Y \rangle(t) \\
		dY_t^2		&= 2 Y_t f(t) \, dM_t + d \langle Y, Y \rangle(t)
	\end{align*}  Hence \[
		\langle Y, Y \rangle(t) = Y_t^2 - 2 \int_0^t Y_s f(s) \, dM_s = [Y, Y](t) 
	\] since $\int_0^t Y_s f(s) \, dM_s$ is a martingale.
\end{proof}
\subsection{\itos integration for continuous semimartingales} % (fold)
\label{sub:itos_integration_for_continuous_semimartingales}
\begin{defn}
	Let $(X_t, \sigf_t)$ be a continuous semimartingale.  Then \[
		X_t = M_t + A_t 
	\] where $M_t$ is a martingale and $A_t$ is a continuous adapted process of bounded variation ($\lim_{\delta \rightarrow 0} \sum_{j=1}^n |A_{t_{j+1}} - A_{t_j} |< \infty$). 
\end{defn}

\begin{defn}[Integrals for semimartingales]
	\[
		\int_0^t c_s \, dX_s = \int_0^t c_s \, dM_s + \int_0^t c_s \, dA_s	
\] and since $A_t$ has bounded variation, the second integral is defined in the Riemann-Stiljes (R.S.) sense.
\end{defn}

\begin{thm}[\itos formula]
	Let $X_t$ be a continuous semimartingale.  Let $f(x)$ have twice continuous derivatives.  Then \begin{align*}
		f(X_t) &= f(X_0) + \int_0^t f'(X_s) \, dX_s + \frac{1}{2} \int_0^t f''(X_s) \, d[X, X](s)
	\end{align*}
\end{thm}
\begin{proof}
	Partition the interval $[0,t]$, and use a Taylor expansion to express $f(x) = f(x_0) + f'(x_0)(x- x_0) + \frac{1}{2}f''(x_0)(x-x_0)^2$.
\end{proof}

\subsection{Stochastic differential equations} % (fold)
\label{sub:stochastic_differential_equations}
Consider the equation \[
	dX_t = \sigma(t, X_t) \, dB_t = \mu(t, X_t) \, dt
\]  We seek to solve for a function $f(t, x)$ such that \[
	X_t = f(t, B_t).
\]  Such an $f(t, B_t)$ is a \textbf{solution} to the stochastic differential equation.

\begin{defn}[Strong solution]
	$X_t = X_0 + \int_0^t \sigma(s, X_s) \, dB_s + \int_0^t \mu(s, B_s) \, ds$
\end{defn}
% subsection stochastic_differential_equations (end)

% subsection itos_integration_for_continuous_semimartingales (end)
% section lecture_18_wednesday_10_may (end)

\section{Lecture 19 - Thursday 12 May} % (fold)
\label{sec:lecture_19_thursday_12_may}

\begin{thm}
	Let \[
		dX_t = a(t, X_t) \, dt + b(t, X_t) \, dB_t
	\]
	Assume $\E X_0 < \infty$.  $X_0$ is independent of $B_s$ and there exists a constant $c > 0$ such that \begin{enumerate}
		\item $|a(t, x)| + |b(t, x)| \leq C(1 + |x|)$.  
		\item $a(t, x), b(t, x)$ satisfy the Lipschitz condition in $x$, i.e.\[
			|a(t,x)- a(t, y)| + |b(t, x) - b(t, y)| \leq C|x - y|
		\] for all $t \in (0, T).  $
	\end{enumerate}  Then there exists a \textbf{unique (strong) solution}. 
\end{thm}

\begin{exmp}
	Let \[
		dX_t = c_1 X_t \, dt + c_2 X_t \, dB_t,
	\] with $c_1, c_2$ constants.  
\end{exmp}

\begin{exmp}
	Let \[
		dX_t = [ c_1(t) X_t + c_2(t)] \, dt + [ \sigma_1(t) X_t + \sigma_2(t)] \, dB_t
	\]  Let $a(t,x) = c_1(t)x  + c_2(t)x, b(t,x) = \sigma_1(t) x + \sigma_2(t)$.  
	% \begin{cases}[$\sigma_1(t) = 0$.]  We then have \begin{align*}
	% 	dX_t = [ c_1(t) X_t + c_2(t)] \, dt + \sigma_2(t) \, dB_t
	% 	\iff dX_t - c_1(t) X_t \, dt = c_2(t) \, dt + \sigma_2(t) \, dB_t \\
	% 	\iff e^{-\int_0^t c_1(s) \, ds} \left[ \dots \right] - e^{-\int_}
	% \end{align*}
	% 
		Just follow Kuo p. 233.

		Let \[
			H_t = e^{-Y_t}, \quad Y_t = \int_0^t \sigma_1(s) \, ds + \int_0^t c_1(s) \, dB_s - \frac{1}{2} \int_0^t c^2_1(s) \, ds
		\]   Then by the \ito product formula, we have \[
			d(H_t X_t) = H_t \left( dX_t - \sigma_1(t) X_t \, dt - c_1(t) X_t \, dB_t - c_2(t) c_1(t) \, dt \right)
		\] Then by definition of the $X_t$, we obtain \[
			d(H_t X_t) = H_t \left( c_2(t) \, dB_t + \sigma_2(t) \, dt - c_1(t) c_2(t) \, dt \right)
		\] which can be integrated to yield \[
			H_t X_t = C + \int_0^t H_s c_2(s) \, dB_s + \int_0^t H_s(\sigma_2(s) - c_1(t) c_2(t)) \, ds
		\] Dividing both sides by $H_t$ we obtain our solution $X_t$.  
\end{exmp}

\begin{thm}
	The solution to the linear stochastic differential equation \[
		dX_t = [ c_1(t) X_t + c_2(t)] \, dt + [ \sigma_1(t) X_t + \sigma_2(t)] \, dB_t
	\] is given by \[
		X_t = Ce^{-Y_t} + \int_0^t e^{Y_t - Y_s} c_2(t) \, dB_s + \int_0^t e^{Y_t - Y_s}(\sigma_2(s) - c_1(t) c_2(t)) \, dt
	\] where $Y_t = \int_0^t c_1(s) \, dB_s + \int_0^t \left(\sigma_1(s) - \frac{1}{2} c_1^2(s) \right) \, ds$
\end{thm}
% section lecture_19_thursday_12_may (end)

\section{Lecture 20 - Tuesday 17 May} % (fold)
\label{sec:lecture_20_tuesday_17_may}
\subsection{Numerical methods for stochastic differential equations} % (fold)
\label{sub:numerical_methods_for_stochastic_differential_equations}

\begin{thm}[Euler's method]
	For the stochastic differential equation \[
		dX_t = a(X_t) \, dt + b(X_t) \, dB_t,
	\] we simulate $X_t$ according to \[
		X_{t_j} = X_{t_{j-1}} + a(X_{t_{j-1}}) \Delta t_j + b(X_{t_{j-1}}) \, \Delta B_{t_j}
	\]
\end{thm}

\begin{thm}[Milstein scheme]
	For the stochastic differential equation \[
		dX_t = a(X_t) \, dt + b(X_t) \, dB_t,
	\] we simulate $X_t$ according to \[
		X_{t_j} = X_{t_{j-1}} + a(X_{t_{j-1}}) \Delta t_j + b(X_{t_{j-1}}) \, \Delta B_{t_j} + \frac{1}{2}b'(X_{t_{j-1}})(\Delta B^2_{t_j} - \Delta t_j)
	\]
\end{thm}
\subsection{Applications to mathematical finance} % (fold)
\label{sub:applications_to_mathematical_finance}

\subsection{Martingale method} % (fold)
\label{sec:martinag}
Consider a market with risky security $S_t$ and riskless security $\beta_t$.  
\begin{defn}[Contingent claim]
	A random variable $C_T : \Omega \rightarrow \R$, $\sigf_T$-measurable is called a contingent claim.  If $C_T$ is $\sigma(S_T)$-measurable it is \textbf{path-independent.}
\end{defn}
\begin{defn}[Strategy]
	Let $a_t$ represent number of units of $S_t$, and $b_t$ represent number of units of $\beta_t$.  If $a_t, b_t$ are $\sigf_t$-adapted, then they are strategies in our market model.  Our strategy value $V_t$ at time $t$ is \[
		V_t = a_t X_t + b_t \beta_t
	\]
\end{defn}

\begin{defn}[Self-financing strategy]
	A strategy $(a_t, b_t)$ is self financing if \[
		dV_t = a_t \, dS_t + b_t \, d\beta_t
	\]  The intuition is that we make one investment at $t = 0$, and after that only rebalance between $S_t$ and $\beta_t$.  
\end{defn}

\begin{defn}[Admissible strategy]
	$(a_t, b_t)$ is an \textbf{admissible strategy} if it is self financing and $V_t \geq 0$ for all $0 \leq t \leq T$.  
\end{defn}

\begin{defn}[Arbitrage]
	An arbitrage is an admissible strategy such that $V_0 = 0, V_T \geq 0$ and $\P(V_T > 0) > 0$.  Alternatively, an arbitrage is a trading strategy with $V_0 = 0$, and $\E(V_T) > 0$.
\end{defn}

\begin{defn}[Attainable claim]
	A contingent claim $C_T$ is said to be attainable if there exists an admissible strategy $(a_t, b_t)$ such that $V_T = C_T$.  In this case, the portfolio is said to replicate the claim.  By the law of one price, $C_t = V_t$ at all $t$.
\end{defn}

\begin{defn}[Complete]
	The market is said to be \textbf{complete} if every contingent claim is attainable
\end{defn}

\begin{thm}[Harrison and Pliska]
	Let $\P$ denote the real world measure of the underlying asset price $X_t$. If the market is arbitrage free, there exists an equivalent measure $\P^\star$, such that the discounted asset price $\hat X_t$ and every discounted attainable claim $\hat C_t$ are $\P^\star$-martingales.  Further, if the market is complete, then $\P^\star$ is unique.  In mathematical terms,\[
		C_t = \beta_t \E^\star(\beta_T^{-1} C_T \given \sigf_t).
	\]

	$\P^\star$ is called the equivalent martingale measure (EMM) or the risk-neutral measure.  
\end{thm}
% subsection applications_to_mathematical_finance (end)
% subsection numerical_methods_for_stochastic_differential_equations (end)

% section lecture_20_tuesday_17_may (end)

\section{Lecture 21 - Thursday 19 May} % (fold)
\label{sec:lecture_21_thursday_19_may}

For a trading strategy $(a_t, b_t)$, then the value $V_t$ satisfies \[
	V_t = V_0 + \int_0^t \alpha_s \, dS_s + \int_0^t b_s \, d\beta_s
\] where $B_s$ is the riskless asset.

To price an attainable option $X$, let $(a_t, b_t)$ be a trading strategy with value $V_t$ that replicates $X$. Then to avoid arbitrage, the value of $X$ at time $t = 0$ is given by $V_0$.

\subsection{Change of Measure} % (fold)
\label{sec:change_of_measure}
Let $(\Omega, \sigf, \P)$ be a probability space.

\begin{defn}[Equivalent measure]

Let $\P$ and $\Q$ be measures on $(\Omega, \sigf)$.  Then for any $A \in \sigf$, if \[
\P(A) = 0 \iff \Q(A) = 0
\] then we say the measures $\P$ and $\Q$ are equivalent.  If $\P(A) = 0 \Rightarrow \Q(A) = 0$, we write $Q << P$. 
\end{defn}

\begin{thm}[Radon-Nikodyn]
	Let $Q << P$.  Then there exists a random variable $\lambda$ such that $\lambda \geq 0$, $\E_\P(\lambda) = 1$ and \[
		Q(A) = \int_A d\P = \E_p (\lambda \mathbf{1}_A) 
	\] for any $A \in \sigf$.  $\lambda$ is $\P$=almost surely unique.

	Conversely, if there exists $\lambda$ such that $\lambda \geq 1$, $\E_\P (\lambda) = 1$, then defining \[
		\Q(A) = \int_A \lambda \, d\P
	\] and then $\Q$ is a probability measure and $\Q << \P$.  Consequently, if $Q << P$, then \[
		\E_\Q(Z) = \E_\P(\lambda Z)
	\] whenever $\E_\Q(|Z|) < \infty$.

	The random variable $\lambda$ is called the density of $\Q$ with respect to $\P$, and denoted by \[
		\lambda = \frac{d\Q}{d\P}
	\]
\end{thm}

\begin{exmp}
	Let $X \sim N(0, 1)$ and $Y \sim N(\mu, 1)$ under probability $\P$.  Then there exists a $\Q$ such that $\Q$ is equivalent to $\P$ and $Y \sim N(0, 1)$ under $\Q$.  
\end{exmp}
\begin{proof}
	\begin{align*}
		\P(X \in A) &= \frac{1}{\sqrt{2 \pi}} \int_A e^{-\frac{t^2}{2}} \, dt \\
		\text{Define } \Q(A) 	&= \int_A e^{- \mu X - \frac{\mu^2}{2}} \, d\P  \\
							&= \frac{1}{\sqrt{2 \pi}} \int_A e^{- \mu x - \frac{\mu^2}{2}} e^{-\frac{x^2}{2}} \, dx \\
							&= \frac{1}{\sqrt{2 \pi}} \int_A e^{-\frac{(\mu + x)^2}{2}} \, dx
	\end{align*}  
\end{proof}  Then $\Q << \P, \P << \Q$ and let \[
	\lambda = \frac{d\Q}{d\P} = e^{-\mu X - \frac{\mu^2}{2}}
\]  Then $\lambda$ satisfies the conditions of Radon-Nikodyn theorem.

Then we have \begin{align*}
	\E_\Q(Y) &= \E_\P((X + \mu) \lambda) \\
			&= \int (X + \mu) e^{-\mu X - \frac{\mu^2}{2}} \, d\P \\
			&= \frac{1}{\sqrt{2 \pi}} (x + \mu) e^{- \frac{(\mu + x)^2}{2}} \, dx = 0
\end{align*}
% section change_of_measure (end)
% section lecture_21_thursday_19_may (end)
\section{Lecture 22 - Tuesday 24 May} % (fold)
\label{sec:lecture_22_tuesday_24_may}

\begin{thm}
	Let $\lambda(t), 0 \leq t \leq T$ be a positive martingale with respect to $\sigf_t$ such that \[
		\E_\P(\lambda(T)) = 1.  
	\]  Define a new probability measure $\Q$ by \[
		\Q(A) = \int_A \lambda(T) \, d\P
	\]  Then $\Q << \P$ and for any random variable $X$, we have \begin{align*}
		\E_\Q(X) &= \E_\P(\lambda(T) X)\\
		 \quad \E_\Q(X \given \sigf_t) &= \E_\P \left( \frac{\lambda(T) X}{\lambda(t)} \given \sigf_t \right) \tag{$\star$} \\
		&= \frac{\E_\P(\lambda(T) X \given \sigf_t )}{\E_\P(\lambda(T) \given \sigf_t)}  a.s.	
	\end{align*}
	 and if $X \in \sigf_t$, then for any $s \leq t$, we have \[
		\E_\Q(X \given \sigf_s) = \E_\P \left( \frac{\lambda(t) X}{\lambda(s)} \given \sigf_s \right) \tag{$\star \star$}
	\]

	Consequently a process $S(t)$ is a $\Q$-martingale if and only if \[
		S(t) \lambda(t) \tag{$\dagger$}
	\] is a $\P$-martingale
\end{thm}
\newcommand{\indic}[1]{\mathbf{1}_{#1}}

\begin{proof}
	$(\star)$.  We have
	\begin{align*}
		\Q(\E_\P(\lambda(T)) \given \sigf_t = 0) &= \E_p(\lambda(T) \indic{\E_\P(\lambda(T) \given \sigf_t = 0)}) \\
		% &= \E( \indic{\E_\P(\lambda(T) \given \sigf_t) = 0} \E_\P(\lambda(T) \given \sigf_t)) \\
		&= 0
	\end{align*}

	We have \begin{align*}
		\E_\Q \left( \frac{\E_\P(\lambda(T) X \given \sigf_t )}{\E_\P(\lambda(T) \given \sigf_t)} \mathbf{1}_A \right)  &= \E_\Q \left( \lambda(t) \frac{\E_\P(\lambda(T) X \given \sigf_t )}{\E_\P(\lambda(T) \given \sigf_t)} \mathbf{1}_A \right) \\
		&= \E_\P \left( \E_\P( \lambda(T) X \given \sigf_t) \mathbf{1}_A \right) \\
		&= \E_\P (\lambda(T) X \mathbf{1}_A) \\
		&= \E_\Q( X \mathbf{1}_A)
 	\end{align*}

	$(\star \star)$.  We have \begin{align*}
		\E_\P \left( \frac{\lambda(t) X}{\lambda(s)} \given \sigf_s \right) & \frac{1}{\lambda(s)} \E_\P(\lambda(t) X \given \sigf_s) \\
		&= \frac{1}{\lambda(s)} \E_\P\left( \E_\P(\lambda(T) X \given \sigf_t) \given \sigf_s \right) \\
		&= \frac{1}{\lambda(s)} \E_\P(\lambda(T) X \given \sigf_s) \\
		&= \E_\Q(X \given \sigf_s)
	\end{align*} because of $(\star)$.

	$(\dagger)$. 	 We have \begin{align*}
			&\E_\Q(S(t) \given \sigf_u) = S(u)	\\
		\iff &\E_\P \left( \frac{\lambda(t) S(t)}{\lambda(u)} \given \sigf_u \right) = S(u) \\
		\iff &\E_\P(\lambda(t) S(t) \given \sigf_u) = \lambda(u) S(u) 
	\end{align*} as required.
\end{proof}

\begin{thm}
	Let $B_s, 0 \leq s \leq T$ be a Brownian motion under $\P$.  Let $S(t) = B_t + \mu_t, u \neq 0$.  Then there exists a $\Q$ equivalent to $\P$ such that$S(t)$ is a $\Q$-Brownian motion and \[
		\lambda(T) = \frac{d\Q}{d\P} = e^{- \mu B_T - \frac{1}{2} \mu^2 T}.
	\]   Note that $S(t)$ is not a martingale under $\P$, but it is a martingale under $\Q$.
\end{thm}

\begin{proof}
	Under $\Q$, \[
		\Q(B_0 = 0) = \int_{B_0 = 0} \lambda(T) \, d\P = 1
	\]  $S(t)$ is a $\Q$-martingale if and only if $S(t) \lambda(t)$ is a $\P$-martingale.  But we have \[
		X_t = S(t) \lambda(t) = \left( B_t + \mu t \right) e^{- \mu B_t - \frac{1}{2} \mu^2 t}
	\] is a martingale.

	Finally, note that \begin{align*}
		[S, S](t) = [ B, B](t) = t
	\end{align*}
\end{proof}

\subsection{Black-Scholes model} % (fold)
\label{sub:black_scholes_model}
\begin{defn}[Black-Scholes model]
	The Black-Scholes model assumes the risky asset $S_t$ follows the diffusion process given by \[
		\frac{dS_t}{S_t} = \mu \, dt + \sigma \, dB_t
	\] and the riskless asset follows the diffusion \[
		\frac{d\beta_t}{\beta_t} = r \, dt
	\]

Define the discounted process as follows:\[
	\hat S_t = \frac{S_t}{\beta_t}, \quad \hat V_t = \frac{V_t}{\beta_t}, \quad \hat C_t = \frac{C_t}{\beta_t}.
\]  
\end{defn}

% subsection black_scholes_model (end)
% section lecture_22_tuesday_24_may (end)

\section{Lecture 23 - Thursday 26 May} % (fold)
\label{sec:lecture_23_thursday_26_may}

\begin{lem}{\ }
	\begin{enumerate}[(a)]
	\item By a simple application of \itos lemma, \[
		\frac{d\hat S_t}{\hat S_t} = (\mu - r) \, dt + \sigma \, dB_t.
	\]
	\item $\hat S_t$ is a $\Q$-martingale with \[
		\lambda = \frac{d \Q}{d \P} = e^{- q B_T - \frac{1}{2} q^2 T}
	\] with $ q = \frac{\mu - r}{\sigma}$.

	\item Note that \[
		\frac{d \hat S_t}{\hat S_t} = \sigma d(B_t + \frac{\mu - r}{\sigma} t) = \sigma d \hat B_t
	\] where $\hat B_t = B_t + qt$ is a Brownian motion under $\Q$.

	\item $d\hat S_t = \sigma \hat S_t \, d \hat B_t$.  
	\item In a finite market, where $S_t$ takes only finitely many values, $\hat S_t$ is a $\Q$-martingale is a necessary condition for no-arbitrage.
	\item Note that \[
		\frac{dS_t}{S_t} = \mu \, dt + \sigma \, dB_t = r \, dt + \sigma \, d \hat B_t
	\] 
	\end{enumerate}  
\end{lem}

\begin{thm}
	A value process $V_t$ is self financing if and only if the discounted value process $\hat V_t$ is a $\Q$-martingale.
\begin{align*}
		d \hat V_t &= a_t \, d\hat S_t &\iff dV_t = \alpha dS_t + b_t \, d\beta_t \\
						&\iff V_t = V_0 + \int_0^t a_s \, dS_s + \int_0^t b_s \, d\beta_s \\
						&\iff \text{$V_t$ is self financing.  }
\end{align*}
\end{thm}

\begin{proof}
	By \itos formula, we have \begin{align*}
		d \hat V_t 	&= e^{-rt} \, d V_t - re^{-rt} V_t \, dt \\
					&= e^{-rt} \left( a_t \, dS_t + b_t \, d\beta_t \right) - re^{-rt} \left( a_t S_t + b_t \beta_t \right) \, dt \\
					& a_t (e^{-rt} \, dS_t - re^{-rt} \, dt) \\
					&= a_t \, d \hat S_t
	\end{align*} 
\end{proof}

\begin{thm}
	In the Black-Scholes model, there are no arbitrage opportunities.
\end{thm}

\begin{proof}
	For any admissible trading strategy $(a_t, b_t)$, we have that the discounted value process $\hat V_t$ is a $Q$-martingale.  So if $V_0 = 0$, then $\E(\hat V_0) = 0$, and we have \[
		\E_\Q(\hat V_T) = \E_\Q(\hat V_T \given \sigf_0) = \hat V_0 = 0
	\] which implies $\Q(\hat V_T > 0) = 0$, which implies $\P(V_T > 0) = 0$, which implies that $\E_\P(V_T) = 0$, which then implies no arbitrage.     
\end{proof}

\begin{thm}
	For any self financing strategy, \[
		V_t = a_t S_t + b_t \beta_t = V_0 + \int_0^t a_u \, dS_u + \int_0^t b_u \, d\beta_u
	\]  And so a strategy is self financing if \[
		S_t \, da_t + \beta_t \, db_t + d[a, S](t) = 0
	\]  We now consider several cases.
	\begin{enumerate}
		\item   If $a_t$ is of bounded variation, then $[a, S](t) = 0$.  Hence \[
			S_t \, da_t + \beta_t \, db_t = 0,
		\] which implies \[
			db_t = -\frac{S_t}{\beta_t} \, da_t
		\]  Hence $da_t \cdot db_t < 0$.  
		\item If $a_t$ is a semi-martingale $a_t = a_t^2 + A_t$, then $b_t$ must be a semi-martingale, where $b_t = b_t^2 + B_t$ where $a_t^2, b_t^2$ are the martingale parts and $A_t, B_t$ are of bounded variation.
		% \item
	\end{enumerate}
\end{thm}
% section lecture_23_thursday_26_may (end)

\section{Lecture 24 - Tuesday 31 May} % (fold)
\label{sec:lecture_24_tuesday_31_may}
\begin{thm}
	Given a claim $C_T$ under the self-financing assumption, there exists a $\Q$-martingale such that \[
		V_t = \E_\Q\left( e^{-r(T-t) C_T \given \sigf_t} \right), \sigf_t = \sigma(B_s, 0 \leq s \leq t).
	\] In particular, we have \[
		V_0 = \E_\Q(e^{-rT}C_T)
	\]
\end{thm}

\begin{proof}
	For any $\Q$-martingale $\hat V_t$, we have $\hat V_t = \E_\Q(\hat V_T \given \sigf_t)$.  Then \[
		V_t = \E_\Q(e^{-r(T-t)}C_T \given \sigf_t),
	\] as $V_T = C_T$.  
\end{proof}

\begin{thm}
	A claim is attainable (there exists a trading strategy replicating the claim), that is, \begin{align*}
		V_t = V_0 + G(t) \\
		G(t) = \int_0^t a_u \, dS_u + \int_0^t b_u \, d\beta_u \\
		V_T \geq 0, V_T = C_T 
	\end{align*}
\end{thm}

\begin{thm}
	Suppose that $C_T$ is a non-negative random variable, $C_t \in \sigf_T$ and $\E_\Q(C_T^2f < \infty)$, where $\Q$ is defined as before.  Then \begin{enumerate}[(a)]
		\item The claim is replicable.
		\item \[
			V_t = \E_\Q\left( e^{-r(T-t)} C_T \given \sigf_t \right) \iff \hat V_t = \E_\Q(\hat C_T \given \sigf_t)
		\] where $\hat C_T = e{-rT} C_T$.

		In particular, $V_0 = \E_\Q(e^{-rT} C_T) = \E_\Q(\hat C_T)$.
	\end{enumerate}
\end{thm}

\begin{thm}
	Assume $\hat V_t = \E_\Q(\hat C_T \given \sigf_t)$.  Using the martingale representation theorem, there exists an adapted process $H(s)$ such that \begin{align*}
		\hat V_t = \hat V_0 \int_0^t H(s) \, d \hat B_s \iff d \hat V_t = H(t) \, d\hat B_t.
	\end{align*}  On the other hand, $d \hat V_t = a_t \, d\hat S_t = a_t \cdot \sigma \hat S_t \, d \hat B_t$.  Hence we obtain our required result, \begin{align*}
		a_t = \frac{H(t)}{\sigma \hat S_t},
	\end{align*} and then solve for $b_t$.
\end{thm}

\begin{exmp}
	Let $C_T = f(S_T)$.  Then $V_t = \E_\Q \left( e^{-r(T- t) f(S_T) \given \sigf_t} \right)$.    Since $\sigf_t = \sigma(B_s, 0 \leq s \leq t) = \sigma(\hat B_s, 0 \leq s \leq t)$, and $\hat S_t$ is a $\Q$-martingale, we have \[
		\hat S_t = \hat S_0 e^{- \frac{\sigma^2}{2}t + \sigma \hat B_t}
	\] and so \[
		S_T =  e^{rT} \hat S_T = \hat S_t e^{rT} e^{- \frac{\sigma^2}{2}(T- t) + \sigma(\hat B_T - \hat B_t)}. 
	\]  Then \begin{align*}
		V_t = \E_\Q \left[ e^{-r(T-t)} f\left( e^{rT}\hat S_t e^{-\frac{\sigma^2}{2}(T-t) + \sigma( \hat B_T - \hat B_t)} \right) \right]
	\end{align*} and so \[
		V_t = F(t, S_t)
	\] where \begin{align*}
		F(t, x) &= \E_\Q\left[ e^{-r(T-t)} f\left(e^{(-\frac{\sigma^2}{2})(T-t) + \sigma \sqrt{T-t} Z} \right) \right]\\
				&= e^{-r(T-t)} \int_\R f\left(xe^{-\frac{\sigma^2}{2}(T-t) + \sigma z \sqrt{T-t}} \right) \phi(z) \, dz
	\end{align*}  where $\phi(z) = \frac{1}{\sqrt{2 \pi}} e^{- \frac{z^2}{2}}$.
\end{exmp}

\begin{exmp}
	In particular, if $f(y) = (y - K)^+$, then we obtain \begin{align*}
		F(t, x) &= e^{-r\theta} \int_{-d_1'}^\infty xe^{-\frac{\sigma^2}{2} \theta + xz \sqrt{\theta} - \frac{z^2}{2}} \, dz - K \int_{-d_1'}^\infty e^{-r\theta - \frac{z^2}{2}} \, dz \\
		&= x \Phi(d_1' + \sigma \sqrt{\theta}) - K e^{-r \theta} \Phi(d_1') \\
		&= x \Phi(d_1) - K e^{-r \theta} \Phi(d_2), 
	\end{align*} where \[
		d_1 = \frac{ \log \left( \frac{x}{K} \right) + (r + \frac{\sigma^2}{2}) \theta}{\sigma \sqrt{\theta}}, \quad d_2 = d_1 - \sigma \sqrt{\theta} 
	\]
\end{exmp}

\begin{thm}[Black-Scholes model summary]
	$V_t = a_t S_t + b_t \beta_t$, where \begin{align*}
		\frac{dS_t}{S_t} = \mu \, dt + \sigma \, dB_t \\
		\frac{d\beta_t}{\beta_t} = r \, dt 
	\end{align*}

	\begin{enumerate}[(a)]
		\item $\hat S_t = e^{-rt} S_t$ is a $\Q$-martingale, where \[
			\frac{d\Q}{d\P} = e^{-qB_T - \frac{1}{2} q^2 T}
		\] and $q = \frac{\mu - r}{\sigma}$.

		Then $\hat B_t = B_t + \frac{\mu - r}{\sigma} t$ is a $\Q$-Brownian motion, and $d \hat S_t = \sigma \hat S_t \, d\hat B_t$.
		\item $V_t$ is self-financing if $(a_t, b_t)$ satisfies \[
			S_t \, da_t + \beta_t \, db_t + d[a, S](t) = 0
		\] which then implies $\hat V_t$ is a $\Q$-martingale, $\hat V_t = e^{-rt}V_t$.  		
		\item
		There are no arbitrage opportunities in the Black-Scholes model.
		% \item Under t
		% \item
	\end{enumerate}
\end{thm}
% section lecture_24_tuesday_31_may (end)

\section{Lecture 25 - Thursday 2 June} % (fold)
\label{sec:lecture_25_thursday_2_june}
\begin{thm}[Feyman-Kac formula]{\ }
	\begin{enumerate}[(1)]
		\item Suppose the function $F(x, t)$ solves the boundary value problem \[
			\frac{\partial F(t, x)}{\partial t} + \mu(t, x) \frac{\partial F(t, x)}{\partial x} + \frac{1}{2} \sigma^2(t, x) \frac{\partial^2 F(t, x)}{\partial x^2} = 0
	\] such that $F(T, x) = \Psi(x)$.  
		\item Let $S_t$ be a solution of the SDE \[
			dS_t = \mu(t, S_t) \, dt + \sigma(t, S_t) \, dB_t \tag{$\star$}
		\] where $B_t$ is a $\Q$-Brownian motion
		\item Assume \[
			\int_0^T \E(\sigma(t, S_t) \frac{\partial^2 F(t, S_t)}{\partial x^2}) \, dt < \infty
		\]
	\end{enumerate}

	Then \[
		F(t, S_t) = \E_\Q\left( \Psi(S_T) \given \sigf_t \right) = \E_\Q\left(F(T, S_T) \given \sigf_t \right).
	\] where $\sigf_t = \sigma(B_s, 0 \leq s \leq t)$. 
\end{thm}

\begin{proof}
	It is enough to show that $F(t, S_t)$ is a martingale with respect to $\sigf_t$ under $\Q$.  By \itos lemma, we have \begin{align*}
		dF(t, S_t) &= \frac{\partial f(t, S_t)}{\partial t} \, dt + \frac{\partial F(t, S_t)}{\partial x} \, dS_t + \frac{1}{2} \frac{\partial^2 F(t, S_t)}{\partial x^2} \cdot (dS_t)^2 \\
		&= \left[ \frac{\partial F}{\partial t} + \mu(t, S_t) \frac{\partial F}{\partial x} + \frac{\sigma^2(t, S_t)}{2} \frac{\partial^2 F(t, S_t)}{\partial x^2} \right] \, dt + \frac{\partial F(t, S_t)}{\partial x} \sigma(t, S_t) \, dB_t \\
		&= \frac{\partial F(t, S_t)}{\partial x} \sigma(t, S_t) \, dB_t
	\end{align*} which is a $\Q$-martingale.
\end{proof}

\begin{thm}[General Feyman-Kac formula]
	Let $S_t$ be a solution of the SDE $(\star)$.  Assume that there is a solution to the PDE \[
		\frac{\partial F(t, x)}{\partial t} + \mu(t, x) \frac{\partial F(t, x)}{\partial x} + \frac{1}{2} \sigma^2(t, x) \frac{\partial^2 F(t, x)}{\partial x^2} = r(t, x) F(t, x).
	\]  Then \[
		F(t, S_t) = \E_\Q\left( e^{-\int_t^T r(u, S_u) \, du} F(T, S_T) \given \sigf_t \right)
	\]
\end{thm}

\begin{proof}
	Again by \itos lemma, \begin{align*}
		dF(t, S_t) &= \left( \frac{\partial F}{\partial t} + \mu \frac{\partial F}{\partial x} + \frac{\sigma^2}{2} \frac{\partial^2 F}{\partial x^2} \right) \, dt + \frac{\partial F}{\partial x} \cdot \sigma(t, S_t) \, dB_t \\
		&= r(t, x) F(t, S_t)\, dt + dM_t
	\end{align*}  
	where $M_t = \int_0^t \frac{\partial F}{\partial x} \sigma(u, S_u) \, dB_u$.  Hence we have \begin{align*}
		dF(t, S_t) 	&= r(t, S_t) X_t \, dt + dM_t \\
\tag{$\Rightarrow$}		d\left[ e^{-\int_t^\nu r(u, s_u)\, du} X_\nu \right] &= e^{-\int_t^\nu r(u, S_u) \, du} dM_\nu \\
\tag{$\Rightarrow$}		e^{-\int_t^T r(u, S_u) \, du} F(T, S_T) &= F(t, S_t) + \int_t^T e^{-\int_t^\nu r(u, S_u)\, du} \, dM_\nu  \\
\tag{$\Rightarrow$}	 	\E_\Q \left(	e^{-\int_t^T r(u, S_u) \, du} F(T, S_T) \given \sigf_t \right) &= F(t, S_t) \\
			&\qquad +  \underbrace{\E_\Q \left( \int_t^T e^{-\int_t^\nu r(u, S_u)\, du} \, \frac{\partial F}{\partial x} \sigma(\nu, S_\nu) \, dB_\nu \given \sigf_t \right)}_{= 0} 
	\end{align*} and so we obtain our result, \[
		F(t, S_t) = \E_\Q\left( e^{-\int_t^T r(u, S_u) \, du} F(T, S_T) \given \sigf_t \right)
	\]
\end{proof}
% section lecture_25_thursday_2_june (end)

\end{document}

### 2011-12-12 PMH3 - Functional Analysis/PMH3 Lecture Notes.tex
Initial Commit

% Created by Andrew Tulloch

%!TEX TS-program = xelatex
%!TEX encoding = UTF-8 Unicode

\documentclass[10pt, reqno, oneside]{amsart}
\usepackage{amsthm, amsmath, amssymb}
\usepackage{setspace, graphicx, enumerate}
\onehalfspacing
% \usepackage{mathspec}
% \setallmainfonts{Minion Pro}

\usepackage{fontspec,xltxtra,xunicode}
\defaultfontfeatures{Mapping=tex-text}

% \setlength{\columnsep}{50pt}
% \setlength{\columnseprule}{2pt}
% \setromanfont[Mapping=tex-text,Contextuals=w {NoWordInitial,NoWordFinal,NoLineInitial,NoLineFinal}]{Adobe Caslon Pro}
%\setsansfont[Scale=MatchLowercase,Mapping=tex-text]{Hoefler Text}
%\setmonofont[Scale=MatchLowercase]{Andale Mono}

	% AMS Theorems
	\theoremstyle{plain}% default 
		\newtheorem{thm}{Theorem}[section] 
		\newtheorem{lem}[thm]{Lemma} 
		\newtheorem{prop}[thm]{Proposition} 
		\newtheorem*{cor}{Corollary}

	\theoremstyle{definition} 
		\newtheorem{defn}[thm]{Definition}
		\newtheorem{conj}[thm]{Conjecture}
		\newtheorem{exmp}[thm]{Example}
		\newtheorem{exer}[thm]{Exercise}

	\theoremstyle{remark} 
		\newtheorem*{rem}{Remark} 
		\newtheorem*{note}{Note} 
		\newtheorem{case}{Case}

\newcommand{\expc}[1]{\mathbb{E}\left[#1\right]}
\newcommand{\var}[1]{\text{Var}\left(#1\right)}
\newcommand{\cov}[1]{\text{Cov}\left(#1\right)}
\newcommand{\prob}[1]{\mathbb{P}(#1)}
\newcommand{\given}{ \, | \,}
\newcommand{\us}{0 \leq u \leq s}
\newcommand{\ts}[1]{\{ #1 \}}

\newcommand{\dzz}{\, dz}
\newcommand{\bigo}[1]{\mathcal{O}(#1)}

% Maths Field Symbols
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\K}{\mathbb{K}}
\newcommand{\N}{\mathbb{N}}
\renewcommand{\H}{\mathcal{H}}

\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\F}{\mathbb{F}}
\newcommand{\Ga}{\mathbb{G}}

\newcommand{\aut}[1]{\text{Aut}{(#1)}}

\newcommand{\gener}[1]{\langle #1 \rangle}
\newcommand{\charr}[1]{\text{char}(#1)}
\newcommand{\nth}{n\textsuperscript{th}}

\newcommand{\tworow}[2]{\genfrac{}{}{0pt}{}{#1}{#2}}
\newcommand{\xdeg}[2]{[#1 : #2]}
\newcommand{\gal}[2]{\text{Gal}(#1/#2)}
\newcommand{\minpoly}[2]{m_{#1, #2}(x)}
\renewcommand{\Re}{\text{Re}}

\newcommand{\mapping}[5]{\begin{align*}
	#1 : \,     #2 &\rightarrow #3 \\
			#4  &\mapsto #5
\end{align*}	
}

\newcommand{\hilb}{(\H, \langle \cdot, \cdot, \rangle )}
\renewcommand{\phi}{\varphi}

\newcommand{\iprod}[1]{\left\langle #1 \right\rangle}
\newcommand{\sumkn}{\sum_{k=1}^n}
\newcommand{\sumin}{\sum_{i=1}^n}
\newcommand{\sumkinf}{\sum_{k=1}^\infty}
\newcommand{\sumiinf}{\sum_{i=1}^\infty}
\newcommand{\spans}{\text{span}\ }
\newcommand{\im}{\textsc{Im\ }}
\renewcommand{\ker}{\textsc{Ker\ }}

\title{PMH3 - Functional Analysis \\ Lecture Notes}								% Document Title
\author{James Parkinson}
%\date{}                                           % Activate to display a given date or no date

\begin{document}
\maketitle

\section{Lecture 1 - Monday 28 February} % (fold)
\label{sec:lecture_1_28_february}

% Norms, metrics, Lp spaces, Banach Space, Inner Product Space, Normed Space, etc.  
\begin{defn}[Norm]
	Let $X$ be a vector space.  A norm on $X$ is a function $\| \cdot \| : X \mapsto \R$ satisfying 
	\begin{itemize}
		\item $\| x \| \geq 0$ with equality if and only if $x = 0$.  
		\item $\| \alpha x \| = | \alpha | \| x \|$.
		\item $\| x + y \| \leq \| x \| + \| y \|$ for all $x, y \in X$.  
	\end{itemize}

	We call the pair $(X, \| \cdot \|)$ a \textbf{normed vector space.}
\end{defn}

\begin{thm}[Reverse triangle inequality]
	Let $X$ be a normed vector space.  For any $x, y \in X$, we have \[
		\left| \|x \| - \| y \| \right| \leq \| x - y \|  
	\]
\end{thm}

\begin{defn}[Complete space]
	Let $X$ be a normed vector space.  Then $X$ is \textbf{complete} if every Cauchy sequence in $X$ converges to some $x \in X$.  
\end{defn}

\begin{defn}[Banach space]
	A \textbf{Banach space} is a complete normed vector space.
\end{defn}

% section lecture_1_28_february (end)

\section{Lecture 2 - Wednesday 2 March} % (fold)
\label{sec:lecture_2_2_march}

\begin{prop}[Convergence] Let $(V, \| \cdot \| )$ be a normed vector space.  A sequence $(x_n)$ in $V$ converges to $x \in V$ if given $\epsilon > 0$, there exists $N$ such that $\| x - x_n \| < \epsilon$ whenever $n < N$. 
\end{prop}

\begin{lem}
	If $x_n \rightarrow x$, then $\| x_n \| \rightarrow \| x \| \in \R$.
	\begin{proof}
		$\left| \| x_n \| - \| x \| \right| \leq \| x - x_n \| \rightarrow 0$.
	\end{proof}
\end{lem}

\begin{prop}
	Every convergent sequence is Cauchy.
\end{prop}

\begin{defn}[Banach space]
A complete, normed, vector space is called a \textbf{Banach space}
\end{defn}

\begin{prop}
	$( \K, |\cdot |)$ is complete.
\end{prop}

\begin{prop}
	$(\ell^p, \| \cdot \|_p)$ is a Banach space for all $1 \leq p \leq \infty$
\end{prop}

\begin{proof}
	A general proof outline follows.
	\begin{itemize}
		\item Use completeness of $\R$ to find a candidate for the limit.
		\item Show this limit function is in $V$.
		\item Show that $x_n \rightarrow x$ in $V$.
 	\end{itemize}
	Let $x^{(n)}$ be a Cauchy sequence in $\ell^p$.  Since $|x_j^{(n)} - x_{j}^{(n)}| \leq \|x^{(n)} - x^{(m)}\|$, we know that $x^{(n)}_j$ is a Cauchy sequence in $\K$.  Hence, $\lim_{n \rightarrow \infty} x^{(n)}_j := x_j$ exists, and is our limit candidate.

	We now show that $\sum_{j =1}^\infty |x_j|^p < \infty$.  We have 
\end{proof}

\begin{prop}
	($\ell([a,b]), \|\cdot \|_\infty$) is a Banach space
\end{prop}

\begin{prop}
	If $1 \leq p < \infty$, then $(\ell([a,b]), \| \cdot \|_p)$ is \textbf{not} a Banach space.
\end{prop}
\begin{proof}
	Consider a sequence of functions that is equal to one on $[0, \frac{1}{2}]$, zero on $[\frac{1}{2}+ \frac{1}{n}, 1]$, and linear between.  This is a Cauchy sequence that does not converge to a continuous function. 
\end{proof}
% section lecture_2_2_march (end)

\section{Lecture 3 - Monday 7 March} % (fold)
\label{sec:lecture_3_monday_7_march}
We've seen that $(\ell([a,b]), \| \cdot \|_p)$ is not complete for $1 \leq p < \infty$.

\begin{thm}[Completion] Let $(V, \| \cdot \| )$ be a normed vector space over $\K$.  There exists a Banach space ($V_1, \| \cdot \|_1$) such that $( V, \| \cdot \|)$ is isometrically isomorphic to a dense subspace of $(V_1, \| \cdot \|_1)$.

	Furthermore, the space $( V_1, \| \cdot \|_1)$ is unique up to isometric isomorphisms.	
\end{thm}

\begin{proof}
	Rather straightforward - construct Cauchy sequences, append limits, quotient out (as different sequences may converge to the same limit).
\end{proof}

\begin{defn}[] $(V_1, \| \cdot \|_1)$ is called \textbf{the completion} of $(V, \| \cdot \|)$. 
\end{defn}

\begin{defn}[Dense]
	If $X$ is a topological space and $Y \subseteq X$, then $Y$ is \textbf{dense} in $X$ if the closure of $Y$ in $X$ equals $X$, that is, $\overline{Y} = X$.

Alternatively, for each $x \in X$, there exists $(y_n)$ in $Y$ such that $y_n \rightarrow x$.
\end{defn}

\begin{defn}[Isomorphism of vector spaces]
	Two normed vector spaces $(X, \| \cdot \|X)$ and $(Y, \| \cdot \|Y)$ are \textbf{isometrically isomorphic} if there is a vector space isomorphism $\Psi: X \rightarrow Y$ such that \[
		\| \Psi(x) \|_Y = \| x \|_X \quad \forall x \in X
	\]
\end{defn}

\begin{exmp}
	Let $\ell_0 = \{ (x_i) \, | \, \# \{ i, x_i \neq 0 \} < \infty \}$.  The completion of $\ell_0, \| \cdot \|_p$ is $(\ell^p, \| \cdot \|_p )$, because,
	\begin{itemize}
		\item $\ell_0$ is a subspace of $\ell^p$,
		\item It is dense, since we can easily construct a sequence in $\ell_0$ converging to arbitrary $x \in \ell^p$.
	\end{itemize}
\end{exmp}
% % 
\begin{exmp}[ $L^p$ spaces]
	Let $\mu$ be the Lebesgue measure on $\R$.  Let \[
		\mathcal{L}^p([a,b]) = \{ \text{measurable } f: [a,b] \rightarrow \K \, | \, \int_a^b |f|^p \, d \mu < \infty \}
	\]

	Let $\| f \|_p = \left( \int_a^b |f|^p \, d \mu \right)^{1/p}$.  Since $\| f \|_p = 0 \iff f = 0 \, a.e$, we quotient out by the rule $f \equiv g \iff f - g = 0 \, a.e.$, and then our space of equivalence classes forms a normed vector space, denoted $L^p([a,b])$.
\end{exmp}

\begin{thm}[Riesz-Fischer] $(L^p([a,b]), \| \cdot \|_p )$ is the completion of $(\mathcal{C}[a,b], \| \cdot \|_p )$, and is a Banach space.
\end{thm}
\begin{proof}
	Properties of the Lebesgue integral.
\end{proof}

\begin{rem}{\ }
	\begin{itemize}
		\item Let $X$ be any compact topological space, let $\mathcal{C}(X) = \{ f : X \rightarrow \K \, | \, \text{$f$ is continuous} \}$, and let $\|f \|_\infty = \sup_{x \in X} \|f(x)|$.  Then $\mathcal{C}(X, \| \cdot \|_\infty)$ is Banach.  
		\item Let $X$ be any topological space.  Then the set of all continuous and bounded functions with the supremum norm forms a Banach space.
		\item Let $(S, \mathcal{A}, \mu)$ be a measure space.  Then we can define the $\mathcal{L}^p$ and $L^p$ analogously, and they are also Banach.
	\end{itemize}
\end{rem}

\begin{defn}[Linear operators on normed vector spaces]
	Let $X,Y$ be vector spaces over $\K$.  A linear operator is a function $T:X \rightarrow Y$ such that 
	\begin{align*}
		T(x+y) &= T(x) + T(y) \\
		T(\alpha x) &= \alpha T(x)
	\end{align*}
	for all $x,y,\alpha$.

	We write $\text{Hom}(X,Y) = \{ T: X \rightarrow Y \, | \, \text{$T$ is linear} \}$
\end{defn}

\begin{defn}[]
	$T: X \rightarrow Y$ is continuous at $x \in X$ if for all $\epsilon > 0$, there exists $\delta > 0$ such that \[
			\| x - y \|_X < \delta \Rightarrow \| Tx - Ty \|_y < \epsilon 
	\]
\end{defn}

\begin{defn}[]
	\[
	\mathcal{L}(X,Y) = \{ T:  X \rightarrow Y \, | \, \text{$T$ is linear and continuous} \}
	\]
\end{defn}

\begin{rem}
	If $\text{dim}(X) < \infty$ then $\text{Hom}(X,Y) = \mathcal{L}(X,Y)$.  This is \textbf{not} true if $X$ has infinite dimension.
\end{rem}

\begin{defn}[Bounded linear operator]
	Let $T: X \rightarrow Y$ be linear, then $T$ is \textbf{bounded} if $T$ maps bounded sets in $X$ to bounded sets in $Y$.  That is: for each $M > 0$ there exists $M' > 0$ such that \[
		\| x \|_X \leq M \Rightarrow \|Tx\|_Y \leq M'
	\]
\end{defn}
% section lecture_3_monday_7_march (end)

\section{Lecture 4 - Wednesday 9 March } % (fold)
\label{sec:lecture_4_}
Consider the space $\mathcal{L}(X, Y)$, the set of all linear and continuous maps between two normed vector spaces $X$ and $Y$.

\begin{thm}[Fundamental theorem of linear operators]
	Let $( X, \| \cdot \|_X)$ and $Y, \| \cdot \|_Y$ be normed vector spaces.  Let $ T \in \text{Hom}(X,Y)$, the set of all linear maps from $X$ to $Y$.  Then the following are all equivalent.
	\begin{enumerate}[1)]
		\item $T$ is uniformly continuous
		\item $T$ is continuous
		\item $T$ is continuous at 0
		\item $T$ is bounded
		\item There exists a constant $c > 0$ such that \[
			\| Tx \|_Y \leq c \| x \|_X \quad \forall x \in X
		\]
	\end{enumerate}
\end{thm}

\begin{proof}
	$1) \Rightarrow 2) \Rightarrow 3)$ is clear.

	$3) \Rightarrow 4)$.  Since $T$ is continuous at 0, given $\epsilon = 1 > 0$, there exists $\delta$ such that \[
		\| Tx - T0 \| \leq 1 \quad \text{whenever} \quad \| X - 0 \| \leq \delta,
	\] i.e. that $\| x \leq \delta \Rightarrow \| Tx \| \leq 1$.  Let $y \in X$.  The $\| \frac{\delta y}{\| y \|} \| \leq \delta$, and so $\| T\left( \frac{\delta y}{\| y \|} \right) \| < \leq 1$.  Hence, \[
		\frac{\delta}{\|y \|} \|T y \| \leq 1
	\] and so \[
		\|Ty \| \leq \frac{ \| y \|}{\delta}
	\] for all $y \in X$.  Thus, for all $\| y \| \leq M$, we have $\| Ty \| \leq M'$, where $M' = \frac{M}{\delta}$, and so $T$ is \textbf{bounded.}

	$4) \Rightarrow 5)$.  If $T$ is bonded, given $M = 1 > 0$, there exists $c \geq 0$ such that $\| x \| \leq 1 \Rightarrow \|T x \| \leq c$.  Then \[
		 \|T \left( \frac{x}{\|x \|} \right) \| \leq c	\] 
		Hence, $\|Tx \| \leq c \| x \|$.

	$5) \Rightarrow 1)$.  If $5)$ holds, then  \[
		\| Tx - Ty \| = \| T(x-y) \| \leq c \| x - y \|.
	\]  So if $\epsilon$ is given, taking $\delta = \frac{\epsilon}{c}$, we have \[
		\|Tx - Ty \| \leq c \| x - y \| < c \frac{\epsilon}{c} = \epsilon. \qedhere
	\]
\end{proof}

\begin{cor}
	If $T \in \text{Hom}(X,Y)$, then $T$ continuous $\iff$ $T$ bounded $\iff$ $\|Tx \| \leq c \| x \|$ for all $x \in X$.
\end{cor}

\begin{defn}[Operator norm]
	The \textbf{operator norm} of $T \in \mathcal{L}(x,y)$, $\| T\|$ is defined by any one of the following equivalent expressions.
	\begin{enumerate}[(a)]
		\item $\|T \| = \inf \{ c > 0 \, | \, \| Tx \| < c \| x \| \}$.
		\item $\| T\| = \sup_{x \neq 0} \frac{ \|Tx \|}{\| x \|}$.
		\item $\| T \| = \sup_{ \|x \| \leq 1} \| Tx \|$.
		\item $\| T \| = \sup_{\| x \| = 1} \|T x \|$.
	\end{enumerate}
\end{defn}

\begin{prop}
	The operator norm is a norm on $\mathcal{L}(x,y)$.  
\end{prop}

\begin{proof}  The following are simple to verify.
	\begin{enumerate}[(a)]
		\item $\| T \| \geq 0$, with equality if and only if $ T = 0$. 
		\item $\| \alpha T \| = | \alpha | \|T \|$.
		\item $\| S + T \| \leq  \| S \| + \| T \|$.
	\end{enumerate}
\end{proof}

\begin{exmp}[Calculating $\| T \|$]
	To calculate $\| T \|$, try the following. 
	\begin{enumerate}[1)]
		\item Make sensible calculations to find $c$ such that \[
			\| Tx \| \leq c \| x \|
		\] for all $x \in X$.
		\item Find $x \in X$ such that $\|Tx \| = c \| x \|$.
	\end{enumerate}
\end{exmp}
% section lecture_4_ (end)

\section{Lecture 5 - Tuesday 15 March} % (fold)
\label{sec:lecture_5_tusday_15_march}
\begin{rem}
	Ignore !2, Q3(b), Q8 on the practice sheet, as we will be ignoring Hilbert space theory for the time being.
\end{rem}

\begin{defn}[Algebraic dual]
	Let $(X, \| \cdot \|)$ be a normed vector space over $\K$.  The \textbf{algebraic dual} of $X$ is \[
		X^\star = \text{Hom}(X, \K) = \{ \phi: X \rightarrow \K \, | \, \text{$\phi$ is linear} \}.
	\]  Elements of $X^\star$ are called linear functionals.
\end{defn}

\begin{defn}[Continuous dual (just \textbf{dual})]
	The \textbf{continuous dual} (just dual) of $X$ is \[
		X' = \mathcal{L}(X, \K) = \{ \phi : X \rightarrow K \, | \, \text{$\phi$ is linear and continuous} \}.
	\]
\end{defn}

\begin{rem}
	$X^\star \supsetneq X'$ if $\dim(X) = \infty$.
\end{rem}

\begin{exmp}
	Let $(\wp([a,b]), \| \cdot \|_\infty)$ be the normed vector space of polynomials $p: [a,b] \rightarrow \K$. 				
	\begin{enumerate}[(a)]
		\item The functional $D : \wp([0,1]) \rightarrow \K$ given by $D(p) = p'(1)$ is linear, but \textbf{not} continuous.
		\item The functional $I: \wp([0,1]) \rightarrow \K$ given by $I(p) = \int_0^1 p(t) \, dt$ is linear \textbf{and} continuous.
	\end{enumerate}
\end{exmp}

\begin{proof}
\begin{enumerate}[(a)]
		\item Linearity is clear.  The $p_n(t) = t^n$ for all $t \in [0,1]$.  Then $|D(p_n)| =   n \| p_n \|_\infty$.  So $D$ is not continuous, as continuity implies that there exists $c$ such that \[
			\| Tx \| \leq c \| x \|.
		\]
		\item Exercise: Show $\| I \| = 1$.
\end{enumerate}
\end{proof}

Describing the continuous dual space $X'$ is one of the first things to do when trying to understand a normed vector space.  It is generally pretty difficult to describe $X'$.

\begin{prop}[Dual of the $\ell^p$ space for $(1 < p < \infty)$]
	Let $1 < p < \infty$.  Let $q$ be the ``dual'' of $p$, defined by $\frac{1}{q} + \frac{1}{p} = 1$.  Then $(\ell^p)'$ is isometrically isomorphic to $\ell^q$.  
\end{prop}
\begin{rem}[Observation before proof]
	Let $1 \leq p < \infty$.  Let $e_i = (0,0,\dots, 1, 0, \dots)$ where $1$ is in the $i$-th place.  
\begin{enumerate}[1)]
		\item If $x = (x_i) \in \ell^p$, then \[
			x = \sum_{i=1}^\infty x_i e_i
 		\] in the sense that the partial sums converge to $x$.
 		\item If $\phi : \ell^p \rightarrow \K$ is linear and continuous, then \[
 			\phi(x) = \sum_{i = 1}^\infty x_i \phi(e_i)
 		\]
\end{enumerate}

\begin{proof}[Proof of observations.]
	Let $S_n = \sum_{i=1}^n x_i e_i$.  Then \begin{align*}
		\| x - S_n \|_p^p &= \| (0,0,\dots, x_{n+1}, x_{n+2}, \dots) \|_p^p \\
		&= \sum_{i=n+1}^\infty |x_i|^p \\
		&\rightarrow 0 \quad \text{as it is the tail of a convergent sum.}
	\end{align*}

	Write $\phi(x)$ as \begin{align*}
		\phi(x) &= \phi( \lim_{n \rightarrow \infty} S_n)  \quad \text{(continuity)}\\
				&= \lim_{n \rightarrow \infty} \left( \phi(S_n ) \right) \\
				&= \lim_{n \rightarrow \infty} \phi \left( \sum_{i=1}^n x_i e_i \right) \\
				&= \lim_{n \rightarrow \infty} \sum_{i=1}^n x_i \phi(e_i) \quad \text{(linearity)} 	\\
				&= \sum_{i=1}^\infty x_i \phi(e_i) \qedhere
	\end{align*}
\end{proof}
\end{rem}

\begin{proof}
	Define a map $\theta$ by \mapping{\theta}{\ell^q}{(\ell^p)'}{y}{\phi_y} where $\phi_y(x) = \sum x_i y_i$ for all $x \in \ell^p$.

	\begin{enumerate}[(1)]
		\item $\phi_y$ is linear, as $\phi_y( x + x') = \phi_y(x) + \phi_y(x')$ (valid as sums converge absolutely.)
		\item $\phi_y$ is continuous, as \[
			| \phi_y(x) | = | \sum x_i y_i | \leq \sum | x_i y_i | \leq  \| x \|_p \| y \|_q
		\] by H\"older's inequality.  From the fundamental theorem of linear operators, as $| \phi_y(x) | \leq \| x \|_p \| y \|_q$, we have that $\phi_y$ is continuous, and that 
		\begin{equation}
			\| \phi_y \| \leq \| y \|_q \tag{$\star$}
		\end{equation}  
		\item $\theta$ is linear. 
		\item $\theta$ is injective, as \[
			\theta(y) = \theta(y') \Rightarrow \phi_y = \phi_{y'} \Rightarrow \phi_y(x) = \phi_{y'}(x)\quad \forall x \in \ell^p
		\] \[
			\Rightarrow \phi_y(e_i) = \phi_{y'}(e_i) \quad\forall i \in \N \Rightarrow y_i = y_i' \quad\forall i \in \N \Rightarrow y = y'
		\]
		\item $\theta$ is surjective.  Let $\phi \in (\ell^p)$.  Let $y = ( \phi(e_1), \dots, \phi(e_n), \dots) = (y_1, \dots, y_n, \dots)$.  We now show $y \in \ell^q$.

		Let $x^{(n)} \in \ell^q$  be defined by \[
			x_i^{(n)} = \begin{cases}
				\frac{|y_i|^q}{y_i} &\text{if $i \leq n$ and $y_i \neq 0$} \\
				0 &\text{otherwise}
			\end{cases}
		\]
		Then \begin{equation}
			\phi( x^{(n)}) = \sum_{i=1}^\infty x_i^{(n)} \phi(e_i) = \sum_{i=1}^n |y_i|^q
			\tag{$\dagger$}
		\end{equation} by Observation 2) above.

		On the other hand, we know \begin{align*}
			\| \phi( x^{(n)}) &\leq \| \phi \| \| x^{(n)} \|_p \\
							&= \| \phi \| \left( \sum_{i=1}^\infty |x_i^{(n)}|^p \right)^{1/p} \\
							&= \| \phi \| \left( \sum_{i = 1}^n |y_i |^{(q-1)p} \right)^{1/p} \\
							&= \| \phi \| \left( \sum_{i=1}^n |y_i |^q \right)^{1/p} \quad \text{as $1/p + 1/q = 1$.} \tag{$\star \star$}
		\end{align*}
		Now, using $(\dagger)$ and $(\star \star)$, we have \[
			 \sum_{i=1}^n |y_i|^q  \leq \| \phi \| \left( \sum_{i=1}^n | y_i |^q \right)^{1/p} 
		\] and so we must have \[
			\| y \|_q \leq \| \phi \| \tag{$\star \star \star$}
		\]
		and so $y \in \ell^q$.

		We also have, by $(\star \star)$,\[
			\| y \|_q \leq \| \phi_y \|
		\] 
		\item Finally, we show that $\theta$ is an isometry.  By $(\star)$ and $(\star \star \star)$, we have \[
			\| \theta(y) \| = \| \phi_y \| = \| y \|_q 
		\] as required.\qedhere

	\end{enumerate} 
\end{proof}	
% section lecture_5_tusday_15_march (end)

\section{Lecture 6 - Wednesday 16 March} % (fold)
\label{sec:lecture_6_wednesday_16_march}
How big is $X'$?  When is $X' \neq \{ 0 \}$?  Examples suggest that $X'$ is big with a rich structure.

\subsection{The Hahn-Banach theorem} % (fold)
\label{sub:the_hahn_banach_theorem}
The Hahn-Banach theorem is a cornerstone of functional analysis.  It is all about extending linear functionals defined on a subspace to linear functionals on the whole space, while preserving certain properties of the original functional.

\begin{defn}[Seminorm]
A let $X$ be a vector space over $\K$.  A seminorm on $X$ is a function $p : X \rightarrow \R$ such that 
\begin{enumerate}[(1)]
	\item $p(x+y) \leq p(x) + p(y) \quad \forall x,y \in X$
	\item $p(\lambda x) = | \lambda | p(x) \quad \forall x \in X, \lambda \in \K$
\end{enumerate}
\end{defn}

\begin{thm}[General Hahn-Banach]
	Let $X$ be a vector space over $\K$.  Let $p: X \rightarrow \R$ be a seminorm on $X$.  Let $Y \subseteq X$ be a subspace of $X$.  If $f : Y \rightarrow \K$ is a linear functional such that \[
		| f(y) | \leq p(y) \quad \forall y \in Y
	\] then there is an extension $\tilde{f} : X \rightarrow \K$ such that 
	\begin{itemize}
		\item $\tilde{f}$ is linear
		\item $\tilde f (y) = f(y) \quad \forall y \in Y$
		\item $|f(x)| \leq p(x) \quad \forall x \in X$
	\end{itemize}
\end{thm}

\begin{rem}
	This is great.
	\begin{itemize}
		\item $Y$ can be finite dimensional (and we know about linear functionals on finite dimensional spaces)
		\item If $p(x) = \| x \|$, then \[
			|\tilde f (x) | \leq \|x \| \quad \forall x \in X
		\] and so $\tilde f \in X'$
	\end{itemize}
\end{rem}

\begin{cor}
	Let $(X, \| \cdot \|)$ be a normed vector space over $\K$.  For each $y \in X$, with $y \neq 0$, there is $\phi \in X'$ such that \[
		\phi(y) = \| y \| \quad \text{and} \quad \| \phi \| = 1
	\]
\end{cor}
\begin{proof}
	Fix $y \neq 0$ in $X$.  Let $Y = \{ \K y \} = \{ \lambda y | \lambda \in \K \}$, a one-dimensional subspace.

	Define $f : Y \rightarrow \K$, $f( \lambda y) = \lambda \| y \|$.  This is linear.  Set $p(x) = \| x \|$.  Then \[
		| f(\lambda y ) = p(\lambda y)
	\] and so by Hahn-Banach, there exists $\tilde f : X \rightarrow \K$ such that
	\begin{itemize}
		\item $\tilde f$ is linear
		\item $\tilde f(\lambda y) = f( \lambda y) \quad \forall \lambda \in \K$
		\item $| \tilde f (x) | \leq \| x \| \quad \forall x \in X$
	\end{itemize}
	Then we  have $\tilde f \in X'$ and $\| f \| = 1$ as required.
\end{proof}

% subsection the_hahn_banach_theorem (end)

\subsection{Zorn's Lemma} % (fold)
\label{sub:zorn_s_lemma}

\begin{thm}[Axiom of Choice is equivalent to Zorn's Lemma]
	See handout for proof that \[
		A.C. \Rightarrow Z.L.
	\]
\end{thm}

\begin{defn}[Partially ordered set]
	A \textbf{partially ordered set} (poset) is a set $A$ with a relation $\leq$ such that 
	\begin{enumerate}[(1)]
		\item $ a \leq a$ for all $a \in A$,
		\item If $a \leq b$ and $b \leq a$then $a = b$,
		\item If $a \leq b$ and $b \leq c$, then $a \leq c$
	\end{enumerate}
\end{defn}

\begin{defn}[Totally ordered set]
	A \textbf{totally ordered set} is a poset $(A, \leq)$ such that if $a, b \in A$ then either $a \leq b$ or $b \leq a$.
\end{defn}

\begin{defn}[Chain]
	A \textbf{chain} in a poset $(A, \leq)$ is a totally ordered subset of $A$.
\end{defn}

\begin{defn}[Upper bound]
	Let $(A, \leq)$ be a poset.  An \textbf{upper bound} for $B \subseteq A$ is an element $u \in A$ such that $b \leq u$ for all $ b \in B$.  
\end{defn}

\begin{defn}[Maximal element]
	A \textbf{maximal element} of a poset $(A, \leq)$ is an element $m \in A$ such that $m \leq x$ implies $x = m$, that is, \[
		m \leq x \Rightarrow x = m
	\]
\end{defn}

\begin{exmp}
	Let $S$ be any set.  Let $\mathcal{P}(S)$ be the power set of $S$ (the set of all subsets of $S$).  Define $a \leq b \iff a \subseteq b$.  Maximal element is $S$
\end{exmp}

\begin{thm}[Zorn's Lemma]
	Let $(A, \leq)$ be a poset.  Suppose that every chain in $A$ has an upper bound.  Then $A$ has (at least one) maximal element.  
\end{thm}

\begin{exmp}[Application - all vector spaces have a basis]
	\begin{defn}[Linearly independent]
		Let $X$ be a vector space over $\F$.  We call $B \subseteq X$ \textbf{linearly independent} if \[
			\lambda_1 x_1 + \dots + \lambda_n x_n = 0 \Rightarrow \lambda_1 = \dots = \lambda_n = 0
		\] for all finite $\{ x_1, \dots, x_n \} \subseteq B$. 
	\end{defn}

	\begin{defn}[Span]
		We say $B \subseteq X$ \textbf{spans} $X$ if each $x \in X$ can be written as \[
			x = \lambda_1 x_1 + \dots + \lambda_n x_n
		\] for some $\lambda_1, \dots, \lambda_n \in \F$ and $\{ x_1, \dots, x_n \} \subseteq B$. 
	\end{defn}

	\begin{defn}[Hamel basis]
		A Hamel basis is a linearly independent spanning set. Equivalently, $B \subseteq X$ is a Hamel basis if and only if each $x \in X$ can be written in exactly one way as a finite linear combination of elements of $B$.
	\end{defn}

	\begin{thm}
		Every vector space has a Hamel basis
	\end{thm}

	\begin{proof}
		Let $L = \{ \text{linearly independent subsets} \}$, with subset ordering.  Let $C$ be a chain in $L$.  Let $u = \bigcup_{a \in C} a$.  Then 
		\begin{enumerate}[(1)]
			\item $u \in L$,
			\item $u$ is an upper bound for $C$.
		\end{enumerate}  
		So Zorn's Lemma says that $L$ has a maximal element $\mathbf{b}$.

		Then $\mathbf{b}$ is a Hamel basis. 
		\begin{itemize}
			\item $\mathbf{b}$ is linearly independent.  
			\item If $\text{Span}(\mathbf{b}) \neq X$, there exists $X \in X \backslash \text{Span}(\mathbf{b})$, and $\mathbf{b'} = \mathbf{b} \bigcup \{ x \} \in L$ is linearly independent, contradicting maximality of $\mathbf{b}$.
		\end{itemize}  
	\end{proof}

	\begin{rem}
		If $ X, \| \cdot \| )$ is Banach, every Hamel basis is uncountable.
	\end{rem}
\end{exmp}
% subsection zorn_s_lemma (end)

% section lecture_6_wednesday_16_march (end)

\section{Lecture 7 - Monday 21 March} % (fold)
\label{sec:lecture_7_monday_21_march}
Proof of Hahn-Banach Theorem
Discussion of Dual operators

\begin{thm}[Hahn-Banach theorem over $\R$]
	Let $X$ be a real linear space and let $p(x)$ be a seminorm on $X$.  Let $M$ be a real linear subspace of $X$ and $f_0$ a real-valued linear functional defined on $M$.  Let $f_0$ satisfy $f_0(x) \leq p(x)$ on $M$.  Then there exists a real valued linear functional $F$ defined on $X$ such that 
	\begin{enumerate}[(i)]
		\item $F$ is an extension of $f_0$, that is, $F(x) = f_0(x)$ for all $x \in M$, and 
		\item $F(x) \leq p(x)$ on $X$.  
	\end{enumerate}
\end{thm}

\begin{proof}
	We first show that $f_0$ can be extended if $M$ has codimension one.  Let $x_0 \in X \backslash M$ and assume that $\text{span}(M \cup \{ x_0 \}) = X$. As $x_0 \notin M$ be can write $x \in X$ uniquely in the form \[
		x = m + \alpha x_0
	\] for $\alpha \in \R$.  Then for every $c \in \R$, the map $f_c \in \text{Hom}(X, \R)$ given by $f_c(m + \alpha x) = f_0(m) + c \alpha$ is well defined, and $f_c(m) = f_0(m)$ for all $m \in M$. We now show that we can choose $c \in \R$ such that $f_c(x) \leq p(x)$ for all $ x \in X$.  Equivalently, we must show 
	\[
		f_0(m) + c \alpha \leq p(m + \alpha x_0)
	\] for all $m \in M$ and $\alpha \in \R$.  By positive homogeneity of $p$ and linearity of $f$ we have \begin{align*}
		f_0(m / \alpha) + c &\leq p(x_0 + m/\alpha) \quad \alpha > 0 \\
		f_0(-m/\alpha) - c &\leq p(-x_0 -m/\alpha) \quad \alpha < 0
	\end{align*}  Hence we need to choose $c$ such that \begin{align*}
		c &\leq p(x_0 + m) - f_0(m) \\
		c &\geq -p(-x_0 + m) + f_0(m).
	\end{align*}  This is possible if \[
		-p(-x_0 + m_1) + f_0(m_1) \leq p(x_0 + m_2) - f_0(m_2) 
	\] for all $m_1, m_2 \in M$.  By subadditivity of $p$ we can verify this condition since \[
		f_0(m_1 + m_2) \leq p(m_1 m_2) = p(m_1 - x_0 + m_2 - x_0) \leq p(m_1 - x_0) + p(m_2 + x_0)
	\] for all $m_1, m_2 \in M$.  Hence $c$ can be chosen as required.

	Hence $D(F) = X$, and the theorem is proven.
\end{proof}

\begin{thm}[Hahn-Banach over $\C$]
	Suppose that $c$ is a seminorm on a complex vector space $X$ and let $M$ sub a subspace of $X$.  If $f_0 \in \text{Hom}(M, \C)$ is such that $|f_0(x) | \leq p(x)$ for all $x \in M$, then there exists an extension $f \in \text{Hom}(X, \C)$ such that $f|_M = f_0$ and $|f(x)| \leq p(x)$ for all $x \in X$.  
\end{thm}

\begin{proof}
	Split $f_0$ into real and imaginary parts \[
		f_0(x) = g_0(x) + ih_0(x).
	\]  By linearity of $f_0$ we have \begin{align*}
		0 	&= if_0(x) - f_0(ix) = ig_0(x) - h_0(x) - g_0(ix) - ih_0(ix) \\
			&= -(g_0(ix) + h_0(x) ) + i(g_0(x) - h_0(ix))
	\end{align*} and so $h_0(x) = -g_0(ix)$.  Therefore, \[
		f_0(x) = g_0(x) - ig_0(ix)
	\] 
	for all $x \in M$.  We now consider $X$ as a vector space over $\R$, $X_\R$.  Now considering $M_\R$ as a subspace of $X_\R$.  GSince $g_0 \in \text{Hom}(M_\R, \R)$ and $g_0(x) \leq |f_0(x)| \leq p(x)$ and so by the real Hahn-Banach, there exists $g \in \text{Hom}(X_\R, \R)$ such that $g|_{M_\R} = g_0$ and $g(x) \leq p(x)$ for all $x \in X_\R$.  Now set $F(x) = g(x) - ig(ix)$ for all $x \in X_\R$.  Then by showing $f(ix) = if(x)$, we have that $f$ is linear.

	We now show $|f(x)| \leq p(x)$.  For a fixed $x \in X$ choose $\lambda \in \C$ such that $\lambda f(x) = |f(x)|$.  Then since $|f(x)| \in \R$ and by definition of $f$, we have \[
		|f(x)| = \lambda f(x) | = f(\lambda x) = g(\lambda x) \leq p(\lambda x) = |\lambda p(x) = p(x)
	\] as required.  
 \end{proof}
% section hahn_banac (end)
% section lecture_7_monday_21_march (end)

\section{Lecture 8 - Wednesday 23 March} % (fold)
\label{sec:lecture_8_wednesday_23_march}
\begin{defn}[Inner product]
	Let $X$ be a vector space over $\K$.  
	An \textbf{inner product} is a function \[
	\langle \cdot, \cdot \rangle : X \times X \rightarrow \K 
	\] such that 
	\begin{enumerate}[(1)]
		\item $\langle x + y, z \rangle = \langle x, z \rangle + \langle y, z \rangle$
		\item $\langle \alpha x, z \rangle = \alpha \langle x, z \rangle$
		\item $\langle x, y \rangle = \overline{ \langle y, x \rangle}$
		\item $\langle x, x \rangle \geq 0$ with equality if and only if $x = 0$
	\end{enumerate}
	We then have \[
		\langle x, y + z \rangle = \langle x, y \rangle + \langle x, z \rangle
	\] and \[
		\langle x, \alpha z \rangle = \overline \alpha \langle x, z \rangle
	\]
\end{defn}

\begin{defn}[Inner product space]
	Let $(X, \langle \cdot, \cdot \rangle)$ be an \textbf{inner product space}.  Defining $\| x \| = \sqrt{ \langle x, x \rangle}$ turns $X$ into a normed vector space.  To prove the triangle inequality, we use the Cauchy-Swartz theorem.
\end{defn}

\begin{thm}[Cauchy-Schwarz]
	In an inner product space $(X, \langle \cdot, \cdot \rangle)$, we have \[
		| \langle x, y \rangle | \leq \|x \| \| y \| \quad \forall x, y  \in X
	\] 
\end{thm}
\begin{proof}
	\begin{align*}
		0 	&\leq \langle x - \lambda y, x - \lambda y \rangle \\
			&= \langle x, x \rangle - \langle x, \lambda y \rangle - \langle \lambda y, x \rangle +\langle \lambda y, \lambda y \rangle  \\
			&= \| x \|^2 - \bar \lambda \langle x, y \rangle - \lambda \langle y, x \rangle + |\lambda|^2 \| y \|^2 \\
			&= \| x \|^2 - 2 \Re ( \lambda \langle y, x \rangle ) + |\lambda|^2 \| y \|^2 \\
	\end{align*}  Set $\lambda = \frac{\langle x, y \rangle}{\| y \|^2}$.  Then
	\begin{align*}
		0 &\leq \| x \|^2 - 2 \Re ( \frac{|\langle x, y \rangle|^2}{\| y \|^2} ) + \frac{|\langle x, y \rangle|^2}{\| y \|^2} \\
			&= \| x \|^2 - \frac{|\langle x, y \rangle|^2}{\| y \|^2}
	\end{align*} as required.
\end{proof}

\begin{cor}
	\[
		\| x + y \| \leq \| x \| + \| y \|
	\]
\end{cor}

\begin{defn}[Hilbert space]
	If $(X, \langle \cdot, \cdot \rangle)$ is complete with respect to $\| \cdot \|$ then it is called a \textbf{Hilbert space.} 
\end{defn}

\begin{exmp}
	\begin{enumerate}[(a)]
		\item $\ell^2$, where $\langle x, y \rangle = \sum_{i=1}^\infty x_i \overline{y_i}$.

		Cauchy-Schwarz then says \[
			| \sum_{i=1}^\infty x_i \overline{y_i} | \leq \sqrt{ \sum_{i=1}^\infty |x_i|^2} \sqrt{\sum_{i=1}^\infty |y_i|^2}
		\]
		\item $L^2([a,b])$, where $\langle f, g \rangle = \int_a^b f(x) \overline{g(x)} \, dx$.

		Cauchy-Swartz then says \[
			| \int_a^b f(x) \overline{ g(x)} \, dx \leq .....
		\]
	\end{enumerate}
\end{exmp}

\begin{defn}[Orthogonality]
	Let $(X, \langle \cdot , \cdot \rangle )$ be inner product spaces.  Then $x, y \in X $ are orthogonal if $\langle x, y \rangle = 0$ where $x, y \neq 0$.  
\end{defn}

\begin{thm}
	Let $x_i, \dots, x_n$ be pairwise orthogonal elements in $(X, \langle \cdot , \cdot \rangle)$.  Then \[
		\| \sum_{i=1}^n x_i \|^2 = \sum_{i=1}^n \|x_i\|^2
	\]
\end{thm}

\begin{thm}[Parallelogram identity]
	In $(X, \langle \cdot, \cdot \rangle)$ we have \[
		 \| x+ y \|^2 + \| x-y \|^2 = 2( \| x \|^2 + \|y \|^2) \tag{$\star$}
	\] for all $x, y \in X$.
\end{thm}

\begin{rem}
	If $(X, \| \cdot \|)$ is a normed vector space which satisfies parallelogram identity then $X$ is an inner product space with inner products defined by the polarisation equation \[
	\langle x, y \rangle =	\begin{cases} 
			\frac{1}{4} \left( \| x + y \|^2 - \| x - y \|^2 \right) 	& \K = \R \\
			\frac{1}{4}\left( \| x + y\|^2 - \| x - y \|^2 + i\| x + iy \|^2 - i\|x - iy\|^2 \right)										 		& \K = \C
		\end{cases}
	\]
\end{rem}

\begin{defn}[Projection]
	Let $X$ be a vector space over $\K$.  A subset $M$ of $X$ is convex if for any $x, y \in M$, then \[
		tx + (1-t) y \in M \quad \forall t \in [0,1]
	\]
\end{defn}

\begin{thm}[Projection]
	Let $(\H, \langle \cdot, \cdot, \rangle )$ be a Hilbert space.  Let $M \subseteq \H$ be closed and convex.  Let $x \in \H$.  THen there exists a unique point $m_x \in M$ which is closest to $x$, i.e. \[
		\| x - m_x \| = \inf_{m \in M} \| x - m \| = d
 	\]  
\end{thm}

\begin{proof}
	For each $k \geq 1$ choose $m_k \in M$ such that \[
		d^2 \leq \| x - m_k \|^2 \leq d^2 + \frac{1}{k} 
	\] Each $m_k$ exists as $d$ is defined as the infimum over all $m$.

	Then \begin{align*}
		\| m_k - m_l \|^2 	&= \| (m_k - x) - (m_k - x) \|^2 \\
							&= 2 \| m_k - x \|^2 + 2 \| m_l - x \|^2 - \| m_k + m_l - 2x \|^2 \\
							&\leq 2d^2 + \frac{2}{l} + 2d^2 + \frac{2}{k} - 4 \| \frac{ m_k + m_l}{2} - x \|^2 \\
	\end{align*} and as $m_k/2 + m_l/2 \in M$, we have $\| \frac{ m_k + m_l}{2} - x \|^2  \geq d^2$.  Then 
	\begin{align*}
			\| m_k - m_l \|^2 \leq 2 ( \frac{1}{k} + \frac{1}{l})
	\end{align*} Thus $(m_k)$ is Cauchy.  So $m_k \rightarrow m_x \in M$ as $\H$ is complete and $M$ is closed.  We then have \[
		\| x - m_x \| = d 
	\] and so now we show that $m_x$ is unique.

	Suppose that there exists $m'_x \in M$ with $\| x - m'_x \| = d$.  Then by the above inequality, we have \[
		\| m_x - m'_x \|^2 = 2 \| m_x - x \|^2 + 2 \| m'_x - x \|^2 - 4 \| \frac{m_x - m'_x}{2} - x \|^2 \leq 0	
	\] from above. 
\end{proof}

\begin{defn}[Projection operator]
Let $\hilb$ be a Hilbert space.  Let $M \subseteq \H$ be closed and convex.  Define \[
	P_M : \H \rightarrow \H
\] by $P_M(x) = m_x$ from above. This is the projection of $\H$ onto $M$.
\end{defn}

\begin{defn}[Orthogonal decomposition]	
	If $S \subseteq \H$, let \[
		S^\perp = \{ x \in \H
 \, | \langle x, y \rangle = 0 \quad \forall y \in S. 
\] We call $S^\perp$ the orthogonal component.
\end{defn}

% section lecture_8_wednesday_23_march (end)

\section{Lecture 9 - Monday 28 March} % (fold)
\label{sec:lecture_9_monday_28_march}

\begin{thm}[From previous lecture]
	If $M \subseteq \H$, then the projection of $\H$ onto $M$ is \mapping{P_m}{\H}{\H}{x}{m_x} where $m_x \in M$ is the unique element with $\| x - m_x \| = \inf_{m \in M} \| x - m \|$.  
\end{thm}

\begin{lem}
	Let $M \subseteq \H$ be closed subspace.  Then $x - P_M x \in M^{\perp}$ for all $x \in \H$.
\end{lem}

\begin{proof}
	Let $m \in M$.  We need to show $\langle x - P_M x , m \rangle = 0$.  This is clear if $m = 0$.  Without loss of generality, assuming $ m \neq 0$, we can assume $ \| m \| = 1$.  Then write \[
		x - P_M x = x - \left( P_M x + \langle x - P_M x, m \rangle m \right) + \langle x - P_M x, m \rangle m.
	\]   Let the bracketed term be $m'$.  Then $ x- m' \perp \langle x - P_M x, m \rangle m$ because 
	\begin{align*}
		\langle x - m', \langle x - P_M x, m \rangle m \rangle &= \overline{\langle x - P_M x, m \rangle} \langle x - m', m \rangle \\
		&= C \langle x - P_M x - \langle x - P_M x, m \rangle m, m \rangle \\
		&= C( \langle x - P_M x, m \rangle - \langle x - P_M x, m \rangle \| m \|) \\
		&= 0.
	\end{align*}

	So $\| x - P_M x \|^2 = \| x - m' \|^2 + | \langle x - P_M x, m \rangle |^2$.  So $\| x - P_M x \|^2 \geq \| x - P_M x\|^2 + | \langle x - P_M x, m \rangle |^2$ by definition of $P_M x$.  Thus, \[
		\langle x - P_M x, m \rangle = 0 
	\] and thus $x - P_M x \in M^\perp$.
\end{proof}

\begin{thm}
	The following theorem is the key fundamental result.  
	Let $\hilb$ be a Hilbert space.  Let $M$ be a closed subspace of $\H$.  Then \[
		 \H = M \oplus M^\perp.
	\]  That is, each $x \in \H$ can be written in exactly one way as $x = m + m^\perp$ with $m \in M$, $m^\perp \in M^\perp$. 
\end{thm}

\begin{proof}
	\textbf{Existence} - Let $x = P_m x + ( x - P_M x)$.

	\textbf{Uniqueness} - Let $x = x_1 + x_1^\perp$, $x = x_2 + x_2^\perp$ with $x_1, x_2 \in M, x_1^\perp, x_2^\perp \in M^\perp$
.  Then \[
	x_1 -x_2 = x^\perp_2 - x_1^\perp \in M^\perp
\] Then \[
	\langle x_1 - x_2, x_1 - x_n \rangle = 0 \Rightarrow x_1 = x_2 .  
\] Thus $x_1^\perp = x_2^\perp$.
\end{proof}

\begin{cor}
	Let $M \subseteq \H$ be a closed subspace.  Then we have
	\begin{enumerate}[(a)]
		\item $P_M \in \mathcal{L}(\H, \H)$.
		\item $\| P_M \| \leq 1$.
		\item $\text{Im}P_m = M, \ker P_M = M^\perp$.
		\item $P^2_M = P_M$.
		\item $P_{M^\perp} = I - P_M$.
	\end{enumerate} 
\end{cor}
\begin{proof}
	(c), (d), (e) exercises.

	(a).  Let $x,y \in H$.  Write $x = x_1 + x^\perp_1$ and $y = y_1 + y_1^\perp$ with $x_1, y_1 \in M$ and $x_1^\perp, y_1^\perp \in M^\perp$.  Then \[
		x = y = (x_1 + y_1) + (x^\perp_1 + y^\perp_1)
	\] and so \[
		P_M(x+y) = x_1 + y_1
	\] and similarly $P_M(\alpha x) = \alpha P_M x$.  We also have 
	\begin{align*}
		\| x \|^2 	&= \|P_M x + (x - P_M x) \|^2 \\
					&= \| P_M x \|^2 + \| x - P_M x \|^2 \\
					\geq  \| P_M x \|^2 
	\end{align*} and so $\| P_M \| \leq 1$.  
\end{proof}

\subsection{The dual of a Hilbert space} % (fold)
\label{sub:the_dual_of_a_hilbert_space}

If $y \in \H$ is fixed, then the map \mapping{\phi_y}{\H}{\K}{x}{\langle x, y \rangle} is in $\H'$.  Linearity is clear, and continuity is proven by Cauchy-Swartz, \[
	| \phi_y(x) | = | \langle x, y \rangle | \leq \| y \| \| x \|.
\] So $\| \phi_y \| \leq \| y \|$. Since $|\phi_y(y)| = \| y \|^2$, we then have \[
	\| \phi_y \| = \| y \|.
\]

\begin{thm}[Riesz Representation Theorem]
	Let $\H$ be a Hilbert space. The map \mapping{\theta}{\H}{\H'}{y}{\phi_y} is a conjugate linear bijection, and $\| \phi_y \| = \| y \|$.  
\end{thm}
\begin{proof}
	Conjugate linearity is clear.

	\textbf{Injectivity} 
	\begin{align*}
		\phi_y = \phi_{y'} \Rightarrow \phi_y(x) = \phi_{y'}(x) \quad \forall x
	\end{align*} so \[
		\langle x, y = \langle x, y' \rangle  = 0 \quad \Rightarrow \langle y - y', y - y' \rangle = 0
	\] and so $y = y'$.

	\textbf{Surjectivity}
	Let $\phi \in H'$.  We now find $y \in \H$ with $\phi = \phi_y$.  If $\phi = 0$, take $y = 0$.  Suppose $\phi \neq 0$.  Then $\ker \phi \neq \H$.  But $\ker \phi$ is a closed subspace of $\H$.  So \[
		H = (\ker \phi) \oplus (\ker \phi)^\perp.
	\] Hence $(\ker \phi)^\perp \neq \{ 0 \}$. Pick $z \in (\ker \phi)^\perp, z \neq 0$.  For each $x \in \H$, the element \[
		x - \frac{\phi(x)}{\phi(z)} z \in \ker \phi
	\] Note that $\phi(z) \neq 0$ since $z \notin \ker \phi$.  Then \begin{align*}
		0 	&=  \langle x - \frac{\phi(x)}{\phi(z)}z, z \rangle \\
			&= \langle x, z - \frac{\phi(x}{\phi(z)} \| z \|^2
	\end{align*}
	and so \[
	\phi(x) = \langle x, \frac{\overline{\phi(z)}}{\| z \|^2} z \rangle \quad \forall x \in \H,\]
	and so letting $y = \frac{\overline{\phi(z)}}{\| z \|^2} z$, we have $\phi = \phi_y$.  
\end{proof}

\begin{exmp}
	From Hahn-Banach given $y \in \H$ there exists $\phi \in \H'$ such that \[
		\| \phi \| = 1
	\] and $\phi(y) = \| y \|$. We can be very constructive in the Hilbert case, and let \[
		\phi(x) = \langle x, \frac{y}{\| y \|} \rangle
	\] 
\end{exmp}
\begin{exmp}
	All continuous linear functionals on $L^2([a,b])$ are of the form \[
		\phi(f) = \int_a^b f(x) \overline{g(x)} \, dx 
	\] for some $g \in L^2([a,b])$.
\end{exmp}

\begin{exmp}[Adjoint operators]
	Let $\H_1, \H_2$ be Hilbert spaces.  Let $ T \in \mathcal{L}(\H_1, \H_2)$.  The \textbf{adjoint} of $T$ is $T^\star \in \mathcal{L}(\H_2, \H_1)$ given by \[
		\langle T x, y \rangle_2 = \langle x, T^\star y \rangle_1
	\] for all $x \in \H_1, y \in \H_2$
\end{exmp}

\begin{exer}
	Check all of the above.
\end{exer}

\begin{exer}
	Prove $T^\star = \overline{T^t}$ where $T^t$ is the transpose.   
\end{exer}
% subsection the_dual_of_a_hilbert_space (end)
% section lecture_9_monday_28_march (end)

\section{Lecture 10 - Wednesday 30 March} % (fold)
\label{sec:lecture_10_wednesday_30_march}

\begin{defn}[Orthonormal system]
	As subset $S \subseteq \H$ is an \textbf{orthonormal system} (orthonormal) if \[
		\langle e, e' \rangle = \delta_{e, e'} \quad \forall e, e' \in S
	\]
\end{defn}

\begin{defn}[Complete orthonormal system or Hilbert basis]
	An orthonormal system $S$ is \textbf{complete} or a \textbf{Hilbert basis} if \[
		\overline{\spans S} = \H
	\]
\end{defn}

\begin{rem}
	By Gram-Schmidt and Zorn's Lemma, every Hilbert space has a complete orthonormal system.   
\end{rem}

\begin{exmp}
	\begin{enumerate}
		\item $\ell^2$.  Then \[
			S = \{ e_i \, | \, i \geq 1 \}
		\] is orthonormal and is complete.  
		\item $L^2_\C([0, 2 \pi])$.  Then \[
			S = \{ \frac{1}{2\pi} e^{i n t} \, | \, n \in \Z \}
		\] is orthonormal and is complete.  Completeness follows from Stone-Weierstrass theorem.
		\item $L^2_\R([0, 2 \pi])$.  Then \[
			S= \{ \frac{1}{\sqrt{2\pi}}, \frac{1}{\sqrt{\pi}} \cos nt, \frac{1}{\sqrt{\pi}} \sin nt \, | \, n \geq 1 \}
		\] is orthonormal and is complete, again by Stone-Weierstrass.  
	\end{enumerate}
\end{exmp}

	We want to look at series $\sum_{e \in S} ...$, which is tricky if $S$ is not countable.

\begin{lem}\label{lem:tod}
	If $\{ e_k \, | \, k \geq 0 \}$ is orthonormal, then \[
		\sum_{k=0}^\infty a_l e_k
	\] converges in $\H$ if and only if \[
		\sum_{k=0}^\infty |a_k |^2
	\] converges in $\K$.

	If either series converges, then \[
		\left\| \sum_{k=0}^\infty a_k e_k \right\|^2 = \sum_{k=0}^\infty | a_k |^2
	\]
\end{lem}

\begin{note}
	If $x_n \rightarrow x,y_n \rightarrow y$, then \[
		\langle x_n, y_n \rangle \rightarrow \langle x, y \rangle
	\]
\end{note}

\begin{proof}
	If $\sum_{k=0}^\infty a_k e_k$ converges to $x$, then \begin{align*}
		\langle x, x \rangle &= \lim_{n \rightarrow \infty} \langle \sum_{k=0}^n a_k e_k, \sum_{k=0}^n a_k e_k \rangle \\
		&= \lim_{n \rightarrow \infty} \sum_{k=0}^n | a_k |^2
	\end{align*}

	Conversely, if $\sum_{k=0}^\infty |a_k |^2$ converges, then writing $x_n = \sum_{k=0}^n a_k e_k$, we have \begin{align*}
		\| x_m - x_n \|^2 &= \| \sum_{k=n+1}^m a_k e_k \|^2 \\
						&= \sum_{k=n+1}^m \| a_k e_k \|^2 \, \text{by Pythagoras} \\
						&= \sum_{k=n+1}^m |a_k |^2 \rightarrow 0
	\end{align*} and so $(x_n)$ is Cauchy, and hence converges by completeness of $\H$.  
\end{proof}

\begin{lem}
	Let $\{ e_1, \dots, e_n \}$ be orthonormal. Then \[
		\sumkn | \iprod{x, e_k} |^2 \leq \|x\|^2 
	\] for each $x \in \H$.  
\end{lem}
\begin{proof}
	Let $y = \sumkn \iprod{x, e_k} e_k$.  Let $z = x-y$.  We claim that $z \perp y$.  We have 
	\begin{align*}
		\iprod{x,y} &= \iprod{x-y, y} \\
					&= \iprod{x,y} - \| y \|^2 \\
					&= \sumkn \overline{\iprod{x,e_k}} \iprod{x, e_k} - \sumkn | \iprod {x, e_k} |^2 \\
					&= 0.
	\end{align*}
	So \begin{align*}
		\| x \|^2 	&= \|y + z \|^2 \\
					&= \| y \|^2 + \| z \|^2 \, \text{Pythagoras} \\
					&\geq \|y \|^2 = \sumkn | \iprod{x, e_k} |^2
	\end{align*} 
\end{proof}

We want to write expressions like $\sum_{e \in S} \iprod{x, e} e$.
\begin{cor}
	Let $x \in \H$ and $S$ orthonormal.  Then \[
		\{ e \in S \, | \, \iprod{x, e} \neq 0 \}
	\] is countable.  
\end{cor}

\begin{proof}
	\[
		\{ e \in S \, | \, \iprod{x, e} \neq 0 \} = \bigcup_{k \geq 1} \{ e \in S \, | \, |\iprod{x, e} | > \frac{1}{k}
	\] From the lemma, \[
		\# \{ e \in S \, | \, | \iprod{x,e} | > \frac{1}{k} \} \leq k^2 \| x ^2 \|
	\] For if this number were greater than $k^2 \| x \|^2$, then the LHS in Lemma is greater than $\frac{1}{k^2}k^2 \| x \|^2$.  
\end{proof}

Therefore:
\begin{cor}[Bessel's Inequality]
	If $S$ is orthonormal, then \[
		\sum_{e \in S} | \iprod{x,e}|^2 \leq \| x \|^2
	\] for all $x \in \H$
\end{cor}
\begin{proof}
	$\sum_{e \in S} | \langle x, e \rangle|^2$ is a sum of countably many positive terms, and so order is not important.
\end{proof}

We want to write $\sum_{e \in S} \iprod{x,e} e$.  This sum is over a countable set, but is the order important?

\begin{thm}
	Let $S$ be orthonormal.  Let $M = \overline{\spans S}$.  Then \[
		P_M x = \sum_{e \in S} \iprod{x,e} e
	\] where the sum can be taken in any order.
\end{thm}
\begin{proof}
	Fix $x \in H$.  Choose an enumeration $$\{ e_k \, | \, k \geq 0 \} = \{ e \in S \, | \, \iprod{x,e} \neq 0 \}.$$

	By Bessel's inequality, we have \begin{align*}
		\sum_{k=0}^\infty | \iprod{x, e_k} |^2 \leq \| x \|^2
	\end{align*} and so the LHS converges.  By Lemma~\ref{lem:tod}, we know \[
		y = \sum_{ k = 0}^\infty \iprod{x,e_k} e_k \in M
	\] converges in $\H$.

	Write $x = y + (x-y) = M + M^\perp$.  We claim $(x-y) \in M^\perp$.  Then $P_Mx = y$ from characterisation of projection operator.  Let $e \in S$.  Then \begin{align*}
		\iprod{x-y, e} &= \lim_{n \rightarrow \infty} \iprod{ x - \sum_{k=0}^n \iprod{x, e_k} e_k, e} \\	
		&= \lim_{n \rightarrow \infty} ( \iprod{x,e} - \sum_{k=}^n \iprod{x, e_k} \iprod{e_k, e}) \\
		&= \iprod{x,e} - \sum_{k=0}^\infty \iprod{x, e_k} \iprod{e_k, e}.
	\end{align*}
	If $e \in \{ e' \in S \, | \, \iprod{x, e'} \neq 0 \}$, then $e = e_j$ for some $j$, and so \[
		\iprod{x-y, e} = \iprod{x, e_j} - \iprod{x, e_j} = 0
	\]  
	If $\iprod{x,e} = 0$, then $e \neq e_j$ for all $j$, and so $\iprod{e_j, e} = 0$, and so \[
		\iprod{x-y, e} = 0 - 0 = 0.
	\]

	Thus $x-y \in (\spans S)^\perp$.

	\begin{exer}
		Show that \[
			x-y \in \overline{(\spans S)}^\perp = M^\perp.
		\]
	\end{exer}
\end{proof}

% section lecture_10_wednesday_30_march (end)

\section{Lecture 11 - Monday 4 April} % (fold)
\label{sec:lecture_11_monday_4_april}
Recall that if $\{ x_1. \dots \}$ is a countable orthonormal system in a Hilbert space $\H$.  Then \[
	\sum_{k=1}^\infty a_k e_k < \infty \iff \sum_{k=1}^\infty |a_k|^2 < \infty 
\] and \[
	\| \sum_{k=1}^\infty a_k e_k \|^2 = \sum_{k=1}^\infty |a_k|^2 \tag{$\star$}
\] We also had the following.

\begin{thm}\label{thm:fourier}
	Let $S$ be orthonormal in $\H$.  Let $M = \overline{\spans S}$.  Then \[
		P_Mx = \sum_{e \in S} \iprod{x, e} e \quad \forall x \in \H
	\] where the sum has only countable many terms and convergence is unconditional.
\end{thm}

\begin{thm}
	Let $S$ be orthonormal in $\H$.  Then following are equivalent.
	\begin{enumerate}[(a)]
		\item $S$ is a complete orthonormal system ($\overline{\spans S} = \H$).
		\item $x = \sum_{e \in S} \iprod{x,e} e$ for all $x$ (Fourier series).
		\item $\| x \|^2 = \sum_{e \in S} |\iprod{x,e}|^2$ for all $x$ (Parseval's formula).
	\end{enumerate}
\end{thm}

\begin{proof}
	(a) $\Rightarrow$ (b).  If $M = \overline{\spans S} = \H$, then \[
		P_M x = x = \sum_{e \in S} \iprod{x,e} e
	\] by Theorem~\ref{thm:fourier}.

	(b) $\Rightarrow$ (c).  By the infinite Pythagoras theorem $(\star)$.

	(c) $\Rightarrow$ (a).  Let $M = \overline{\spans S}$. Suppose that $z \in M^\perp$.  Then $z = 0 + z \in M + M^\perp$.  Hence \[
		0 = \| P_M z \|^2 = \|\sum_{e \in S} \iprod{z,e} e \|^2 = \sum_{e \in S} |\iprod{z,e} |^2 = \| z \|^2 
	\] which implies $z = 0$, so $M = \H$, and so $S$ is complete.
\end{proof}

\begin{rem}
	Consider $L^2([0,2\pi])$, and let $S = \{ e_n \, | \, n \in Z \}$.  Then we can write \[
		 f = \sum_{ n \in \Z} c_n e_n
	\] where $c_n = \iprod{f,e_n} = \frac{1}{\sqrt{2 \pi}} \int_0^{2\pi} f(t) e^{-in t} \, dt$.

	We do not claim that convergence is pointwise, what we have proven is convergence is in $L^2$, \[
		\| f - \sum_{|n| \leq N} c_n e_n \|_2 \rightarrow 0
	\] as $N \rightarrow \infty$.  This is not the same as pointwise or uniform convergence ($\| \cdot \|_\infty$).  
\end{rem}

\subsection{Stone-Weierstrass theorem} % (fold)
\label{sub:stone_weierstrass_theorem}
This is a useful tool to show an orthonormal system is complete.  In fact, this theorem is about uniformly approximating elements of $\mathcal{C}(X)$, where $X$ is a compact Hausdorff space.  it is a generalisation of the Weierstrass approximation theorem.

\begin{thm}[Weierstrass approximation theorem]
	Let $f \in \mathcal{C}([a,b])$ and let $\epsilon > 0$ be given.  Then there exists a polynomial $p(x)$ such that \[
		|f(x) - p(x) | < \infty \quad \forall x \in [a,b],
	\] that is, $\| f - p \|_\infty < \epsilon$.

\end{thm}
\begin{cor}  This implies the following important results:
	\begin{itemize}
		\item Continuous functions can be uniformly approximated by polynomials.  
		\item $\mathcal{P}([a,b])$, the space of polynomials on $[a,b]$, is dense in $\mathcal{C}([a,b])$.
		\item $\overline{\mathcal{P}([a,b])} = \mathcal{C}([a,b])$.   
	\end{itemize}
\end{cor}

We now prove Stone's 1930's generalisation.

\textbf{First some setup:} Let $X$ be a compact Hausdorff space throughout.  We then know that $\mathcal{C}(X)$ is a vector space.  It also has sensible vector multiplication, \[
	(fg)(x) = f(x) g(x).
\] Thus $\mathcal{C}(X)$ is a unital, commutative, associative ring.  As we have \[
	f(\lambda g) = \lambda (fg)
\] then $\mathcal{C}(X)$ is a unital, commutative, associative algebra over $\K$.

\begin{defn}[Subalgebra] A subalgebra of $\mathcal{C}(X)$ is a subset $\mathcal{A}$ which is closed under scalar multiplication, vector addition, and vector multiplication.  $\mathcal{A}$ is unital if it contains the constant function $f(x) = 1$.  
\end{defn}

\begin{exmp}
	$\mathcal{P}([a,b])$ is a subalgebra of $\mathcal{C}([a,b])$.  
\end{exmp}

When is $\mathcal{A}$ dense in $\mathcal{C}(X)$?

\begin{thm}[Stone-Weierstrass theorem]
	Let $X$ be a compact Hausdorff space, and let $\mathcal{A}$ be a subalgebra of $\mathcal{C}(X)$. If 
	\begin{enumerate}[(1)]
		\item $\mathcal{A}$ is unital,
		\item $f \in \mathcal{A} \Rightarrow f^\star \in \mathcal{A}$, where $f^\star(x) = \overline{f(x)}$,
		\item $\mathcal{A}$ separates points of $X$.
	\end{enumerate}
	Then $\overline{\mathcal{A}} = \mathcal{C}(X)$. 
\end{thm}

\begin{defn}
 	$\mathcal{A}$ separates points of $X$ if, given $x \neq y$, there is a function $f \in \mathcal{A}$ with $f(x) \neq f(y)$.  
\end{defn}
\begin{cor}
	\begin{enumerate}[(a)]
		\item $\mathcal{P}([a,b])$ is dense in $\mathcal{C}([a,b])$, as $f(x) = x$ separates points.  
		\item Trigonometric polynomials are dense in \[
			\{ f \in \mathcal{C}([0,2 \pi]) \, | \, f(0) = f(2 \pi) \}. 
		\]
		\item Trigonometric polynomials are dense in $L^2([0,2\pi])$, and \[
			S = \{ e_n \given n \in \Z \} 
		\] is complete.
	\end{enumerate}
\end{cor}

\textbf{Setup}
\begin{lem}
	The function $f(t) = |t|$ can be uniformly approximated by polynomials on $[-1,1]$  
\end{lem}
\begin{proof}
	The binomial theorem says \[
		(1 + x)^{1/2} = \sum_{n=0}^\infty {\frac{1}{2} \choose n} x^n \quad \forall x \in [-1,1]
 	\]  We then have 
\begin{align*}
	|t | = \sqrt{t^2} = \sqrt{1 + (t^2 -1)} = \sum_{n=0}^\infty {\frac{1}{2} \choose n } (t^2 - 1)^n \quad t \in [-\sqrt{2}, \sqrt{2}]
\end{align*}

Now let $p_N(t) = sum_{n=0}^N {\frac{1}{2} \choose n } (t^2 - 1)^n$, and \[
	| |t| -p_N(t) | = | \sum_{n = N+1}^\infty { \frac{1}{2} \choose n}(t^2 - 1)^n | \leq \sum_{n = N+1}^\infty | {\frac{1}{2} \choose n} |
\] and so $\| |t| - p_n \|_\infty \rightarrow 0$ as $N \rightarrow \infty$ on $[-1,1]$.
\end{proof}

% subsection stone_weierstrass_theorem (end)
% section lecture_11_monday_4_april (end)

\section{Lecture 12 - Wednesday 6 April} % (fold)
\label{sec:lecture_12_wednesday_6_april}
\begin{thm}[Stone-Weierstrass theorem]
	Let $X$ be a compact Hausdorff space, and let $\mathcal{A}$ be a subalgebra of $\mathcal{C}(X)$. If 
	\begin{enumerate}[(1)]
		\item $\mathcal{A}$ is unital,
		\item $f \in \mathcal{A} \Rightarrow f^\star \in \mathcal{A}$, where $f^\star(x) = \overline{f(x)}$,
		\item $\mathcal{A}$ separates points of $X$.
	\end{enumerate}
	Then $\overline{\mathcal{A}} = \mathcal{C}(X)$. 
\end{thm}
\begin{proof}
	We first prove for $\mathcal{C}_\R(X)$.

	\begin{lem}
		Let $\mathcal{A}$ be a unital subalgebra of $\mathcal{C}_\R(X)$ . Then \begin{enumerate}[(a)]
			\item $|f| \in \overline{\mathcal{A}}$,
			\item $\min(f_1, \dots, f_n), \max(f_1, \dots, f_n) \in \overline{\mathcal{A}}$ 
		\end{enumerate} for all $f, f_1, \dots, f_n \in \mathcal{A} \subseteq \mathcal{C}_\R(X)$. 
	\end{lem}
	\begin{proof}
		\begin{enumerate}[(a)]
			\item Replace $f$ by $\frac{f}{\| f \|_\infty}$ so we can assume that $\|f \|_\infty = 1$.  From the previous lemma, we know for each $n \geq 1$ there is a polynomial $p_n: [-1,1] \rightarrow \R$  such that $\left| |t| - p_n(t) \right| < \frac{1}{n}$ for all $t \in [-1,1]$.

			Since $|f(x)| \leq \|f \|_\infty = 1$ for all $x \in X$, we have \[
				\left\| |f| - p_n(f) \right\| \leq \frac{1}{n}
			\] But $p_n(f)$ is a finite linear combination of $1, f, f^2, f^3, \dots$ and so in in $\mathcal{A}$, as $\mathcal{A}$ is unital.  Thus $|f| \in \overline{\mathcal{A}}$.  
			\item Use the formulas \[
				\max(f,g) = \frac{f+g - |f-g|}{2}, \quad \min(f,g) = \frac{f + g - |f-g|}{2} \in \overline{\mathcal{A}}
			\] and induction.  
		\end{enumerate}
	\end{proof}

	\begin{proof}[Proof of Stone-Weierstrass for $\mathcal{C}_\R(X)$]
		Let $f \in \mathcal{C}_\R(X)$ and let $\epsilon > 0$ be given.  We need to find $g \in \mathcal{A}$ such that \[
			|f(z) - g(z) | < \epsilon \quad \forall z \in X 
		\]

		\textbf{Step 0.}  We can assume that $\mathcal{A}$ is closed.  
		\begin{exer}
			Why?
		\end{exer}

		\textbf{Step 1.}  Let $x,y \in X$ be fixed.  
		\begin{prop}
			There exists $f_{xy} \in \mathcal{A}$ with \[
				f_{xy}(x) = f(x), \quad f_{xy}(x) = f(y)
			\]
		\end{prop}
		\begin{proof}
			If $x = y$ then trivial (take $f_{xy}(z) = f(x) \mathbf{1}(z)$).

			If $x \neq y$, since $A$ separates points, there is $h \in \mathcal{A}$ with $h(x) \neq h(y)$.  Then take \[
				f_{xy} = a h + b 1 \in \mathcal{A}
			\] we can invert the coefficient matrix to find our coefficients $a$ and $b$.  
		\end{proof}

		\textbf{Step 2.} Let $x \in X$ be fixed.  
		\begin{prop}
			There exists $f_x \in \mathcal{A}$ such that \begin{itemize}
				\item $f_x(x) = f(x)$.  
				\item $f_x(z) < f(z) + \epsilon$
			\end{itemize}
		\end{prop}
		\begin{proof}
			For each $y \in X$,let \[
				O_y = \{ z \in X \, | \, f_{xy}(z) < f(z) + \epsilon \}
			\] where $f_{xy}$ is the function from Step 1.  These are all open sets (why?) and thus \[
				X = \bigcup_{y \in X} O_y
			\] since $y \in O_y$.

			By compactness of $X$, we have \[
				X = \bigcup_{i=1}^m O_{y_i}
			\]  Letting $f_x = \min(f_{xy_1}, \dots, f_{xy_n})$.  Then 
			\begin{itemize}
				\item Since $f_{xy_i}(x) = f(x)$ for all i, \[
					f_x(x) = f(x)
				\]
				\item If $z \in X$, then $z \in O_{y_i}$ for some $i$, and so \[
					f_x(z) \leq f_{x y_i}(z) < f(z) + \epsilon
				\] as required.
			\end{itemize} 
		\end{proof}

		\textbf{Step 3.}  \begin{prop}
			There exists a function $g \in \mathcal{A}$ such that \[
				|f(z) - g(z) | < \epsilon
			\] for all $z \in X$.  
		\end{prop}
		\begin{proof}
			For each $x \in X$, let \[
				U_x = \{ z \in X \, | \, f_x(z) > f(x) - \epsilon \} 
			\] where $f_x$ is from Step 2.  These sets $U_i$ are open and since $x \in U_x$, for an open cover, we can write \[
				X = \bigcup_{x \in X} U_x = \bigcup_{j=1}^n U_{x_j}.
			\]  
			Define $g = \max(f_{x_1}, \dots, f_{x_n})$.  If $z \in X$, 
			\begin{itemize}
				\item $g(z) = f_{x_i}(z)$ for some $i$, which is less than $f(z) + \epsilon$ from Step 2.
				\item If $z \in U_{x_j}$ for some $j = 1, \dots, n$, then \[
					g(z) \geq f_{x_j}(z) > f(x) - \epsilon.  
				\]
			\end{itemize}
		\end{proof} 
	\end{proof}
	\begin{exer}
		Where did we use the Hausdorff property?
	\end{exer}

	We now prove for $\mathcal{C}_\C(X)$.

	Let \[ 
	\mathcal A_\R = \{ f \in \mathcal A \given \text{$f$ is real valued} \}. 
	\]  THen $\mathcal A_\R$ is an $\R$-subalgebra of $\mathcal C_\R(X)$.  It is unital, as $1 \in \mathcal A$ and it is real valued.

	We now show $\mathcal A_\R$ separates points.  If $x \neq y$, there is $f \in \mathcal A$ such that $f(x) \neq f(y)$.  Write $f = u + iv$ with $u,v$ real valued.  Either $u(x) \neq u(y)$ or $v(x) \neq v(y)$, and so $\mathcal A_\R$ separates points.

	Hence $\mathcal A_\R$ is dense in $\mathcal C_\R$.

	Now, let $f \in \mathcal C_\C(X)$.  Then write $f = u + iv$.  Then $u,v \in \mathcal C_\R(X)$.  Then given $\epsilon > 0$, there exists $u_1,v_1 \in \mathcal A_\R$ such that \[
		\| u - u_1 \|_\infty \leq \frac{\epsilon}{2}, \quad \| v - v_1 \|_\infty \leq \frac{\epsilon}{2} 
	\]  Writing $f_1 = u_1 + i v_1 \in \mathcal{A}$, we have \[
		\| f - f_1 \|_\infty \leq \| (u - u_1) + i(v-v_1) \|_\infty \leq \| u - u_1 \|_\infty + \| v - v_1 \|_\infty < \epsilon
 	\] and thus $\mathcal{A}$ is dense in $\mathcal C_\C(X)$.
\end{proof}

% section lecture_12_wednesday_6_april (end)

\section{Lecture 13 - Monday 11 April} % (fold)
\label{sec:lecture_13_monday_11_april}

\subsection{Applications of Stone-Weierstrass theorem} % (fold)
\label{sub:applications_of_stone_weierstrass_theorem}
\begin{cor}
	Polynomials are dense in $\mathcal C([a,b])$.  
\end{cor}
\begin{proof}
	$\mathcal A = \mathcal P([a,b])$ is an algebra, is unital, is closed under complex conjugation, and separates points.  Thus, $\mathcal A$ is dense in $\mathcal C([a,b])$.
\end{proof}

\begin{defn}[Trigonometric polynomials]
	A \textbf{trigonometric polynomail} is an expression \[
		\sum_{ n \in \Z} c_n e^{i n t}
	\] with finitely many $c_n \neq 0$. So these are polynomials in $s = e^{it}$ and $s^{-1} = \overline s = e^{-it}$.
\end{defn}

\begin{cor}
	The space $\mathcal A$ of all trigonometric polynomials is dense in $\mathcal C(\Pi)$, where $\Pi = \{ z \in \C \, | \, |z | = 1 \}$
\end{cor}
\begin{proof}
	$\mathcal A$ is a sub-algebra of $\mathcal C(\Pi)$, it is unital, closed under complex conjugation, \[
		\overline{\sum_{n \in \Z} c_n e^{int}} = \sum_{n \in \Z} \overline{c_{-n}} e^{int}
	\] and separates points.  $T$ is a compact Hausdorff space, and thus Stone-Weierstrass states that $\mathcal{A}$ is dense in $\mathcal C(\Pi)$.  
\end{proof}

\begin{cor}
	The orthonomal system \[
		S = \{ \frac{1}{\sqrt{2\pi}} e^{int} \given n \in \Z \}
	\] is complete in $L^2([0, 2 \pi])$.  
\end{cor}
\begin{proof}
	$\spans{S} = \mathcal A$
 is a space of trigonometric polynomials, which is dense in $\mathcal C(\Pi)$.  Define \mapping{\Phi}{\mathcal C_p([0, 2 \pi])]}{\mathcal{C}(\Pi)}{f}{\tilde f}
where $\mathcal C_p([0, 2 \pi]) = \{ f \in \mathcal C([0, 2\pi] ) \given f(0 = f(2 \pi)) \}$.  Then $\Phi$ is an isometric isomorphism, and therefore functions of the form $f(t) = \sum c_n e^{i nt}$ is dense in $\mathcal C_p([0, 2\pi])$.

By the construction of the Lebesgue integral, \textbf{simple} functions \[
	\sum_{i=1}^n a_i \mathbf{1}_{A_i}
\] are dense in $L^2([0, 2\pi])$.

\begin{exer}
	Given $f \in L^2([0, 2 \pi])$ and $\epsilon > 0$, there exists $g \in \mathcal C_p([0, 2\pi]) $ such that $\| f - g \|_2 < \epsilon$.  
\end{exer}

Thus $\mathcal A$ is dense in $L^2([0, 2 \pi])$.
\end{proof}

\begin{cor}
	The following are separable (have a countable dense subset):
	\begin{enumerate}[(a)]
		\item $ \mathcal C([a,b])$,
		\item $ L^p([a,b])$ , $ 1 \leq p < \infty$
	\end{enumerate}
\end{cor}
\begin{proof}
	\begin{enumerate}[(a)]
		\item We have $\mathcal P([a,b])$ is dense in $\mathcal C([a,b])$ and set $\mathcal P_\Q([a,b])$ with rational coefficients is dense in $\mathcal P([a,b])$.  Clearly, $\mathcal P_\Q([a,b])$ is countable, and thus is dense in $\mathcal{C}([a,b])$.
		\item Use the fact that $\mathcal C([a,b])$ is dense in $L^p([a,b])$.  
	\end{enumerate}
\end{proof}

\begin{cor}
	Let $X$ be a compact metric space.  Then $\mathcal C(X)$ is separable.
\end{cor}
\begin{proof}
	As $X$ is a compact metric space, then $X$ is separable.  
	\begin{exer}
		Why?
	\end{exer}
	Let $\{ x_n \given n \geq 1 \}$ be a countable dense subset of $X$.  For each $n \geq 1$ and $m \geq 1$ define \[
		f_{n,m} : X \rightarrow \K
	\] by \[f_{n,m}(x) = \inf_{z \notin B(x_n, \frac{1}{m})} d(x,z)\]
	We then claim $f_{n,m}$ is continuous.  Now, let $\mathcal{A}$ be the space of all $\K$-linear combinations of \begin{equation*}
		f^{k_1}_{n_1,m_1}, \dots, f_{n_l, m_l}^{k_l}, k_1, \dots, k_l \in \mathcal{N}. \tag{$\star$}
	\end{equation*}
 This is a sub-algebra of $\mathcal C(X)$, as $\mathcal A$ is unital, closed under conjugation, and separates points - if $z_1, z_2 \in X$ with $z_1 \neq z_2$, Choose $n,m$ such that $z_1 \in B(x_n, \frac{1}{m})$, $z_n \notin B(x_n, \frac{1}{m})$.  Thus the sub-algebra $\mathcal A$ is dense by Stone-Weierstrass.

	The subset of $\Q$-linear combinations of $(\star)$ is countable and dense.
\end{proof}

\begin{lem}
	If $X$ is compact metric space then $X$ is separable.  
\end{lem}
\begin{proof}
	For each $m \geq 1$, \[
		X = \bigcup_{x \in X} B(x ; \frac{1}{m})
	\]  has a finite subcover \[
		X = \bigcup_{n=1}^{N_m} B(x_{m,n} \frac{1}{m})
	\] and thus the subset of all $\{x_{m,n} \}$ is a countably dense subset. 
\end{proof}

\begin{cor}
	\[
	\frac{pi^2}{6} = \sum_{n = 1}^\infty \frac{1}{n^2}.
	\] 
\end{cor}

\begin{proof}
	$S = \{ \frac{1}{\sqrt{2 \pi}} e^{i nt } \given n \in \Z \}$ is complete, and so Parseval's formula holds, \[
		\| f \|_2^2 = \sum_{ n \in \Z} | \iprod{f, e_n} |^2.
	\]  Apply to $f(x) = x$.  
\end{proof}

A common strategy is to prove for polynomials, and then Stone-Weierstrass proves it for continuous functions.  
\begin{cor}
	If $ f \in \mathcal C([a,b] \times [c,d])$ then \[
		\int_a^b \int_c^d f(x,y) \, dy dx = \int_c^d \int_a^b f(x,y) \, dx dy
	\]
\end{cor}

\begin{proof}
	By direct calculation, the result is true for two-variable polynomials.  Let $f \in \mathcal C([a,b] \times [c,d])$ and $\epsilon > 0$ be given.  By Stone-Weierstrass, the space of polynomials in 2 variables is dense in $\mathcal C([a,b] \times [c,d])$ and so there exists a polynomial $p(x,y)$ with \[
		|f(x,y) - p(x,y)| < \frac{\epsilon}{(b-a)(d-c)}.
	\]  The result then follows by direct calculation.
\end{proof}
% subsection applications_of_stone_weierstrass_theorem (end)
% section lecture_13_monday_11_april (end)

\section{Lecture 14 - Wednesday 13 April} % (fold)
\label{sec:lecture_14_wednesday_13_april}
The following is at the core of two of the cornerstone theorems of functional analysis - the uniform boundedness principle and the open mapping theorem.

\begin{thm}[Baire's theorem]
	Let $X$ be a complete metric space.  If $U_1, U_2, \dots$ are open dense subsets of $X$, then \[
		U = \bigcap_{n=1}^\infty U_n
	\] is dense in $X$.  
\end{thm}

\begin{proof}
	Let $x \in X$ and $\epsilon > 0$ be given.  We need to show that \[
		B(x, \epsilon) \cap U \neq \emptyset.
	\]

	\begin{lem}
		There exists sequences $(x_n)$ in $X$ and $(\epsilon_n)$ in $\R^+$ with the property that 
		\begin{enumerate}[(a)]
			\item $x_1 = x$, $\epsilon_1 = \epsilon$.  
			\item $\epsilon_n \downarrow 0$
			\item $\overline{ B(x_{n+1}, \epsilon_{n+1})} \subseteq B(x_n, \epsilon_n) \cap U_n$ for all $n \geq 1$.  
		\end{enumerate} 
	\end{lem}
	\begin{proof}
		Let $x_1, \dots, x_n$ and $\epsilon_1, \dots, \epsilon_n$ be chosen.  By density of $U_n$, \[
			B(x_n, \epsilon_n) \cap U_n \neq \emptyset.
		\]
		Choose $x_{n+1} \in B(x_n, \epsilon_n) \cap U_n$.  Choose $\epsilon'_{n+1} > 0$ such that $B(x_{n+1}, \epsilon'_{n+1}) \subseteq B(x_n, \epsilon_n) \cap U_n$ (openness).  We have $\epsilon_{n+1}' \leq \epsilon_n$.  Choose $0 \leq \epsilon_{n+1} \leq \min(\frac{\epsilon'_{n+1}}{2}, \frac{1}{n+1})$, then we have \begin{align*}
			\overline B(x_{n+1}, \epsilon_{n+1}) &\subseteq B(x_{n+1}, \epsilon'_{n+1}) \\
			&\subseteq B(x_n, \epsilon_n) \cap U_n
		\end{align*}
		and $\epsilon_{n+1} < \epsilon_n$ with $\epsilon_{n+1} < \frac{1}{n+1}$.
	\end{proof}

	Given the lemma, the theorem follows.  If $m \geq n$, then by (c), \[
		B_(x_m, \epsilon_m) \subseteq B(x_n, \epsilon_n) \cap U_n \tag{$\star$}	
		\]
	In particular, $x_m \in B(x_n, \epsilon_n)$.  Thus, $d(x_n, x_m) < \epsilon_n$ for all $m \geq n$.  Thus $(x_n)$ is Cauchy, and so $x_n \rightarrow \zeta$ in $X$ by completeness.  By $(\star)$, we then have $d(x_n, \zeta) \leq \epsilon_n$ for all $n \geq 1$.  So $\zeta \in \overline{B(x_n, \epsilon_n)}$.  So by (c), $\zeta \in \overline{B(x_{n+1}, \epsilon_{n+1})} \subseteq B(x_n, \epsilon_n) \cap U_n$.

	Thus $\zeta \in B(x, \epsilon)$ and thus $\zeta \in U = \bigcap_{n=1}^\infty U_n$.       	
\end{proof}

The following corollary is often used
\begin{cor}
	Let $X$ be a complete metric space. If $C_1, C_2, \dots$ are closed with $X = \bigcup_{n=1}^\infty$ then $\text{Int} (C_n) \neq \emptyset$ for some $n$.
\end{cor}
\begin{proof}
	If $\text{Int}(C_n) = \emptyset$ for all $n$ then $U_n = X \backslash C_n$ are open and dense.  So by Baire's theorem, $\cap_{n=1}^\infty U_n$ is sense, and in particular, $\bigcap{n=1}^\infty U_n \neq \emptyset$. We have 
	\begin{align*}
		X  = \bigcup_{n=1}^\infty C_n &= \bigcup_{n=1}^\infty (X \backslash U_n) \\
		&= X \backslash (\bigcap_{n=1}^\infty U_n) \\
		&\subsetneq X,
	\end{align*} a contradiction.
\end{proof}

There are three cornerstone theorems.

\begin{itemize}
	\item Hahn-Banach,
	\item Uniform Boundedness,
	\item Open Mapping.
\end{itemize}

\begin{thm}[Uniform boundedness]
	Let $X, Y$ be Banach spaces.  Let $T_\alpha$, $\alpha \in A$, a family of continuous linear operators $T_\alpha : X \rightarrow Y$.  Then if \[
		\sup_{\alpha \in A} \| T_\alpha x \| < \infty
	\] for each fixed $x \in X$, then \[
		\sup_{\alpha \in A} \| T_\alpha \| < \infty
	\]
\end{thm}
\begin{rem}
	Rather amazing - you get a global bound from pointwise bounds.
\end{rem}
\begin{proof}
	For each $n \geq 1$, let \[
		X_n = \{ x \in X \given \| T_\alpha x \| \leq n \, \forall \alpha \in A \}
	\]  These are \textbf{closed} ($T_{\alpha}$ is continuous) and \[
		X = \bigcup_{n=1}^\infty X_n
	\] by the hypothesis.

	By the corollary to Baire's theorem, we know there exists $n_0 \geq 1$ with $\text{Int}(X_{n_0}) \neq \emptyset$.  Choose $x_0 \in \text{Int}(X_{n_0})$, and let $r > 0$ such that \[
		B(x_0, r) \subseteq \text{Int}(X_{n_0}).  
	\]

	If $\| z \| \leq 1$ then $x_0 + rz \in \overline B (x_0, r)$. So $x_0 + rz \in X_{n_0}$, and \[
		\| T_\alpha(x_0 + rz) \| \leq n_0 \, \forall \alpha \in A,
	\]  but $| \| a \| - \| b \| | \leq \| a + b \|$, so \[
		\| T_\alpha (rz) \| - \| T_\alpha (x_0) \| \leq \| T_\alpha (x_0 + rz) \| \leq n_0.
	\]
	So $r \| T_\alpha z \| \ \leq n_0 + n_0$, and \[
		\| T_\alpha z \| \leq \frac{2 n_0}{r} \, \forall \| z \| \leq 1, \forall \alpha \in A
	\]

	For a general $x \in X$, \[
		\| T_\alpha x \| = \| T_\alpha ( \frac{x}{\| x \|}) \| x \| \leq \frac{2 n_0}{r} \| x \|
	\] and thus $\|T_\alpha \| \leq \frac{2 n_0}{r}$, which implies \[
		\sup_{\alpha \in A} \| T_\alpha \| < \infty \qedhere
	\]

\end{proof}
% section lecture_14_wednesday_13_april (end)

\section{Lecture 15 - Monday 18 April} % (fold)
\label{sec:lecture_15_monday_18_april}
Recall, the Fourier series of $f \in L^2([-\pi, \pi])$ is \[
	\sum_{k \in \Z} \iprod{f, e_k} e_k
\] where $e_k(t) = \frac{e^{ikt}}{\sqrt{2 \pi}}$.  This converges to $f$ in the $L^2$ norm.

\begin{exer}
	If $f$ is $2\pi$-periodic and continuous, does the Fourier series converge pointwise?  
\end{exer}

There are explicit (complicated) examples, but the easiest existence is using the uniform boundedness principle.

\begin{prop}
	There is a $2\pi$ periodic continuous function whose Fourier series does not converge at 0.  
\end{prop}
\begin{proof}
	Let $\mathcal{C}_p([-\pi, \pi]) = \{ f \in \mathcal C([-\pi, \pi]) \given f(-\pi) = f(\pi) \}.$  This is a Banach space with $\| \cdot \|_\infty $.  If $f \in \mathcal C_p$, let \[
		f_n = \sum_{|k | \leq n} \iprod{f, e_k} e_k.
	\]

\begin{rem}
	We can now define, for each $n \geq 1$, a linear operator $T_n : \mathcal C_p \rightarrow \K$ by \[
		T_n(f) = f_n(0). 
	\]  If $f_n(0)$ converges (as $ n \rightarrow \infty$) for each $f \in \mathcal C_p$, then \[
		\sup_{n \geq 1} |T_n f | = \sup_{n \geq 1} |f_n(0)| < \infty
	\] for all $f \in \mathcal C_p$, which by uniform boundedness implies \[
		\sup_{n \geq 1} \| T_n \| \leq \infty \tag{$\star$}.
	\]  
	We now show that $(\star)$ is false.
\end{rem}
We have \begin{align*}
	f_n(x) 	&= \sum_{| k \leq n} \frac{1}{2 \pi} \int_{-\pi}^\pi f(t) e^{-ikt} \, dt e^{ikx}  \\
			&= \frac{1}{2\pi} \int_{-\pi}^\pi f(t) \left( \sum_{|k| \leq n} e^{-ik(x-t)} \right) \, dt \\
			&= \frac{1}{2 \pi} \int_{-\pi}^\pi f(t) D_n(x-t) \, dt 
\end{align*} where $D_n(t) = \sum_{|k| \leq n} e^{ikt}$ is the \textbf{Dirichlet Kernel}.  The Dirichlet kernel is real, and even, with \[
	D_n(t) = \frac{\sin(n+\frac{1}{2}) t}{\sin \frac{t}{2}}.  
\]

\begin{note}
	$T_n$ is continuous, with norm $\| T_n \| = \frac{1}{2\pi} \int_{-\pi}^\pi |D_n(t) | \, dt$.
\end{note}
\begin{proof}
	\begin{align*}
		|T_n(f)| &\leq \frac{1}{2\pi} \int_{-\pi}^\pi |f(t)| |D_n(t)| \, dt \\
				&\leq \left( \frac{1}{2\pi} \int_{-\pi}^\pi |D_n(t)| \, dt \right) \| f\|_\infty
	\end{align*} and so $\|T_n \| \leq \frac{1}{2\pi} \int_{-\pi}^\pi |D_n(t)| \, dt$.

	Going the other way, let \[
		s_t = \begin{cases}
		1 &D_n(t) \geq 0 \\
		-1 &D_n(t) < 0
		\end{cases}.
	\]  We have seen hat set functions can be approximated in $L^1$-norm by continuous (periodic) functions.  So if $\epsilon > 0$ is given, there is a $g \in \mathcal C_p$ such that \[
		\left| \frac{1}{2\pi} \int_{-\pi}^\pi (g(t) - s(t))D_n(t) \, dt \right| < \epsilon. 
	\] $g$ can be chosen with $\| g \|_\infty = 1$.

	  So \[
		\left| T_n(g) - \frac{1}{2\pi} \int_{-\pi}^\pi |D_n(t)| \, dt \right| < \epsilon.
	\]  Thus \[
		\frac{1}{2\pi} \int_{-\pi}^\pi |D_n(t) | \, dt - |T_n(g)| < \epsilon.  
	\]  So \[
		|T_n(g)| \geq \frac{\| g \|_\infty}{2 \pi} \int_{-\pi}^\pi |D_n(t)| \, dt - \epsilon. 
	\]  Since $\epsilon > 0$ was arbitrary, \[
		\|T_n \| \geq \frac{1}{2 \pi} \int_{-\pi}^\pi |D_n(t) | \, dt. \qedhere
	\]
\end{proof}

All that remains is to show that \[
	\|T_n \| = \frac{1}{2\pi} \int_{-\pi}^\pi |D_n(t)| \, dt \rightarrow \infty
\]  We have 
\begin{align*}
	\|T_n \| 	&= \frac{1}{\pi}\int_0^\pi |D_n(t) | \, dt \\
				&= \frac{1}{\pi} \int_0^\pi \frac{|\sin(n+\frac{1}{2})t|}{|\sin \frac{t}{2}|} \, dt \\
				&\geq \frac{2}{\pi} \int_0^\pi \frac{|\sin(n+\frac{1}{2}) t |}{t} \, dt \\
				&= \frac{2}{\pi} \int_0^{(n+\frac{1}{2})\pi} \frac{\sin v}{v} \, dv \\
				&\geq \frac{2}{\pi} \int_0^{n\pi} \frac{\sin v}{v} \, dv \\
				&= \frac{2}{\pi} \sum_{k=1}^n \int_{(k-1)\pi}^{k \pi} \frac{|\sin v|}{v} \, dv \\
				&\geq \sum_{k=1}^n \frac{1}{k \pi} \int_{(k-1)\pi}^{k \pi} |\sin v | \, dv \\
				&= \frac{4}{\pi^2} \sum_{k =1 }^n \frac{1}{k} \rightarrow \infty
\end{align*} as $n \rightarrow \infty$.

Thus there exists $f \in \mathcal C_p$ such that the Fourier series of $f$ diverges at $x = 0$.  
\end{proof}

\subsection{The open mapping theorem} % (fold)
\label{sub:the_open_mapping_theorem}
This theorem is \emph{tailor-made} to deal with inverse operators.

\begin{defn}[Open mapping]
	Let $X, Y$ be metric spaces.  A function $f : X \rightarrow Y$ is \textbf{open} if open sets in $X$ are mapped to open sets in $Y$.  
\end{defn}

\begin{thm}[Open mapping theorem]
	Let $X, Y$ be Banach spaces.  If $T \in \mathcal{L}(X,Y)$ is surjective then $T$ is open. 
\end{thm}

\begin{cor}[Bounded inverse theorem]
	Let $X, Y$ be Banach spaces.  If $T \in \mathcal L(X,Y)$ is bijective, then \[
		T^{-1} \in \mathcal L(Y,X).
	\]

\end{cor}
\begin{proof}
	Let $O \subseteq X$ be open.  Then $(T^{-1})^{-1}(O) = T(O)$ is open (by the open mapping theorem).  Thus $T^{-1}$ is continuous.
\end{proof}

\begin{cor}
	Let $(X, \| \cdot \|_1)$ and $(X, \| \cdot \|_2)$ be Banach spaces.  If \[
		\| x \|_1 \leq C \| x \|_2 \quad \forall x \in X
	\] then $\| \cdot \|_1$ and $\| \cdot \|_2$ are equivalent.  
\end{cor}
\begin{proof}
	\mapping{i}{(X, \| \cdot \|_2)}{(X, \| \cdot \|_1)}{ x}{x} is linear, surjective and injective, and also continuous, as \[
		\|i(x) \| = \| x \|_1 \leq C \| x \|_2.
	\]  So the bounded inverse theorem gives \[
		i^{-1} : (X, \| \cdot \|_1) \rightarrow (X, \| \cdot \|_2)
	\] is continuous.  Thus there exists $A > 0$ such that $\| i^{-1}(x)\|_2 \leq A \| x \|_1$, which implies $\| x \|_2 \leq A \| x \|_1$.  So \[
		\frac{1}{A} \| x \|_2 \leq \| x \|_1 \quad \forall x \in X \qedhere
	\]  More generally, if $T \in \mathcal L(X,Y)$ is bijective, then by the bounded inverse theorem, \[
		c \| x \| \leq \| Tx \| \leq C \| x \|
	\] where $c  = \frac{1}{ \| T^{-1} \|}$, $C = \| T \|$.  
\end{proof} 
% subsection the_open_mapping_theorem (end)
% section lecture_15_monday_18_april (end)

\section{Lecture 16 - Wednesday 20 April} % (fold)
\label{sec:lecture_16_wednesday_20_april}

\begin{lem}
	Let $X$ be a Banach space and $Y$ a normed space.  Then for $T \in \mathcal L(X, Y)$, the following are equivalent.
	\begin{enumerate}[(a)]
		\item $T$ is open
		\item There exists $r > 0$ such that $B(0, r) \subseteq T(\overline{B(0, 1)})$
		\item There exists $r > 0$ such that $B(0, r) \subseteq \overline{T(\overline{B(0, 1)})}$.
	\end{enumerate} 
\end{lem}

\begin{proof}
	$(a) \Rightarrow (b), (c)$.  As $B(0, 1)$ is open, the set $T(B(0,1))$ is open in $Y$.  Since $0 \in T(B(0,1))$ there exists $ > 0$ such that the set 
	\[
	B(0, r) \subseteq T(B(0,1)) \subseteq T(\overline{B(0,1)}) \subseteq \overline{T(\overline{B(0,1)})}.
	\]

	$(c) \Rightarrow (b)$.  Assume that there exists $r > 0$ such that \[
		B(0 ,r) \subseteq \overline{T(\overline{B(0,1)})}.
	\] We now show that $B(0, \frac{r}{2}) \subseteq T(\overline{B(0,1)})$ which proves $(b)$.  Let $y \in B(0, \frac{r}{2})$.  Then $2y \in B(0, r)$ and since $B(0, r) \subseteq \overline{T(\overline{B(0,1)})}$ there exists $x_1 \in \overline{B(0,1)}$ such that \[
		\| 2y - Tx_1 \| \leq \frac{r}{2}
	\]  Hence $4y - 2Tx_1 \in B(0, r) $ and by the same argument as before there exists $x_2 \in \overline{B(0,1)}$ such that \[
		\| 4y  - 2Tx_1 - Tx_2 \| \leq \frac{r}{2}
	\] Continuing this way we construct a sequence $(x_n) \in \overline{B(0, 1)}$ such that \[
		\| 2^n y - 2^{n-1} T x_1 - \dots - 2T x_{n-1} - T x_n\| \leq \frac{r}{2}
	\]  for all $n$.  Dividing by $2^{n}$ we obtain \[
		\| y - \sum_{k=1}^n 2^{-k} T x_k \| \leq \frac{r}{2^{n+1}}
	\]  Hence $y = \sum_{k=1}^\infty 2^{-k} T x_k$. 
	Since $\| x_k \| \leq 1$ for all $ k \in \N$ we have that \[
		\sum_{k=1}^\infty 2^{-k} \| x_k \| \leq \sum_{k=1}^\infty 2^{-k} = 1
	\] and so the series \[
		x = \sum_{k=1}^\infty 2^{-k} x_k
 	\] converges absolutely in $X$ as $X$ is Banach and hence complete.  We have also that $\| x \| \leq 1$ and so $x \in \overline{B(0,1)}$.  Because $T$ is continuous we have \[
 		Tx = \lim_{n \rightarrow \infty} \sum_{k=1}^n 2^{-k} T x_k = y
 	\] by construction of $x$.  Hence $y \in T(\overline{B(0,1)})$ and $(b)$ follows.

	$(b) \Rightarrow (a)$.  By $(b)$ and the linearity of $T$ we have \[
		T(\overline{B(0, \epsilon)}) = \epsilon T(\overline{B(0, 1)})
	\] for all $\epsilon > 0$.  Since the map $x \mapsto \epsilon x$ is a homeomorphism on $Y$ the set $T(\overline{B(0, \epsilon)})$ is a neighbourhood of zero for all $\epsilon > 0$.  Now let $U \subseteq X$ be open and $y \in T(U)$.  As $U$ is open there exists $\epsilon > 0$ such that \[
		\overline{B(x, \epsilon)} = x + \overline{B(0, \epsilon)} \subseteq U
	\] where $y = Tx$.  Since $z \mapsto x + z$ is a homeomorphism and $T$ is linear we have \[
		T(\overline{B(x, \epsilon)}) = Tx + T(\overline{B(0, \epsilon)}) = y + T(\overline{B(0, \epsilon)}) \subseteq T(U).
	\]  Hence $T(\overline{B(x, \epsilon)})$ is a neighbourhood of $y$ in $T(U)$.  As $y$ was arbitrary in $T(U)$ it follows that $T(U)$ is open.  
\end{proof}

\begin{lem}
	Let $X$ be a normed vector space and $S \subseteq X$ convex with $S = -S$. If $\overline{S}$ has a non-empty interior, then $\overline{S}$ is a neighbourhood of zero.
\end{lem}

\begin{proof}
	First note that $\overline{S}$ is convex.  If $x, y \in S$ and $x_n, y_n \in S$ with $x_n, y_n \rightarrow x, y$ then $tx_n + (1-t y_n) \in S$ for all $n$ and $t \in [0,1]$.  Letting $n \rightarrow \infty$ we get $tx + (1-t) y \in \overline S$ for all $t \in [0,1]$ and so $\overline S$ is convex.  We also easily have $\overline S = - \overline S$.  If $\overline S$ has a non-empty interior, there exists $z \in \overline S$ and $\epsilon > 0$ such that $B(z, \epsilon) \subseteq \overline S$.  Therefore $z \pm h \in \overline S$ whenever $\| h \| < \epsilon$ and since $\overline S = - \overline S$ we also have $-(z \pm h) \in \overline S$. By the convexity of $\overline S$ we have \[
		y = \frac{1}{2}( (x + h) + (-x + h)) \in \overline S
	\] whenever $\| h \| < \epsilon$.  Hence $B(0, \epsilon) \subseteq \overline S$, and so $\overline S$ is a neighbourhood of zero.
\end{proof}

\begin{thm}[Open mapping theorem]
	Suppose that $X$ and $Y$ are Banach spaces.  If $T \in \mathcal L(X, Y)$ is surjective, then $T$ is open.
\end{thm}
\begin{proof}
	As $T$ is surjective we have \[
		Y = \bigcup_{n \in \N} \overline{T(\overline{B(0, n)})} 
	\] with $\overline[T(\overline{B(0, n)})]$ closed for all $n \in \N$.  Since $Y$ is complete, by a corollary to Baire's theorem, there exists $n \in \N$ such that $\overline{T(\overline{B(0, n)})}$ has non-empty interior. Since the map $x \mapsto nx$ is a homeomorphism and $T$ is linear, the set $\overline{T(\overline{B(0, 1)})}$ has non-empty interior as well.  Now $\overline{B(0, 1)}$ is convex and $\overline{B(0, 1)} = - \overline{B(0, 1)}$.  By linearity of $T$ we have that \[
		T(\overline{B(0, 1)}) = - T(\overline{B(0, 1)})
	\] is convex as well.  Since we know that $\overline{T(\overline{B(0, 1)})}$ has non-empty interior, the previous lemma implies that $\overline{T(\overline{B(0, 1)})}$ is a neighbourhood of zero, and thus there exists $r > 0$ such that \[
		B(0, r) \subseteq \overline{T(\overline{B(0, 1)})}
	\] and since $X$ is Banach the previous lemma shows that $T$ is open.
\end{proof}
% section lecture_16_wednesday_20_april (end)

\section{Lecture 17 - Monday 1 May} % (fold)
\label{sec:lecture_17_monday_1_may}

\begin{exer}
	If $X, Y$ are vector spaces, and if $T: X \rightarrow Y$ is linear, then $\Gamma(T)$ is a subspace of $X \times Y$.  Moreover, if $X, Y$ are normed vectors paces, with \[
		\| (x, Tx) \|_\Gamma = \| x \| + \| Tx \|.
	\]	
\end{exer}

\begin{thm}[Closed Graph theorem]
	Let $X, Y$ be Banach spaces, and $T \in \text{Hom}(X,Y)$.  Then $T \in \mathcal L(X,Y)$ if and only if $\Gamma(T)$ is closed in $X \times Y$.  
\end{thm}

\begin{proof}
	Suppose $T \in \mathcal L(X,Y)$.  If $x_n \rightarrow x$ in $X$, then \[
		(x_n, Tx_n) \rightarrow (x, Tx)
	\] by continuity of $T$, and so $\Gamma(T)$ is closed.

	Conversely, suppose that $\Gamma(T)$ is closed in $X \times Y$.  Define a norm $\| \cdot \|_\Gamma$ on $X$ by $\| x \|_\Gamma = \| x \| + \| Tx \|$.  Since $\Gamma(T)$ is closed, and since $(X, \| \cdot \|)$ is Banach, then $(X, \| \cdot \|_\Gamma)$ is also a Banach space (exercise).  Note that $\| x \| \leq \| x \|_\Gamma$.  So by a corollary to the Open Mapping theorem, $\| \cdot \|$ and $\| \cdot \|_\Gamma$ are equivalent.  So there is $c > 0$ with \[
		\| x \|_\Gamma \leq c \| x \| \quad \forall x \in X.
	\]    So $\| x \| + \| Tx \| \leq c \| x \|$, and so $\| Tx \| \leq (c-1) \| x \|$, and so $T$ is continuous.
\end{proof}

\subsection{Spectral Theory} % (fold)
\label{sub:spectral_theory}
The eigenvalues of an $n \times n$ matrix $T$ over $\C$ are the $\lambda \in \C$ with \[
	\det(\lambda I - T) = 0
\] that is, $\lambda I - T$ is not invertible.

\begin{rem}
	Showing existence of eigenvalues is equivalent to the fundamental theorem of algebra.  
\end{rem}
\begin{rem}
	We need our base field to be $\C$ to get reasonable spectral theory.
\end{rem}

\begin{defn}
Write $\mathcal L(X) = \mathcal L(X,X)$. 
\end{defn}

\begin{defn}
	Let $X$ be a Banach space over $\K$, and let $T \in \mathcal L(X)$.  Then the spectrum of $T$ is \[
		\sigma(T) = \{ \lambda \in \K \given \text{$\lambda I - T$ is not invertible} \}.
	\]
\end{defn}  
\begin{rem}
	$\lambda I - T$ is non invertible if either $\lambda I - T$ is not injective, or $\lambda I - T$ is not surjective.  
\end{rem}
\begin{rem}
	If $\dim(X) < \infty$, then $X \backslash \ker(T) \simeq \text{im}(T)$, and so $T$ is injective if and only if $T$ is surjective. This fails in the infinite dimensional case - consider the left and right shift operators on $\ell^2$.   
\end{rem}

\begin{defn}[Eigenvalue]
$\lambda \in \K$ is an eigenvalue of $T \in \mathcal L(X)$ if there is $ x \neq 0$ with $Tx = \lambda x$, i.e. $\lambda$ is an eigenvalue if and only if $\lambda I - T$ is not injective.
\end{defn}

\begin{thm}
	Let $X \neq \{ 0 \}$ be a Banach space over $\C$, and let $T \in \mathcal L(X)$.  Then $\sigma(T)$ is a non-empty, compact (closed and bounded) subset of \[
		\{ \lambda \in \C \given | \lambda | \leq \| T \| \}
	\]
\end{thm}

\begin{exmp}
	Let $L, R : \ell^2 \rightarrow \ell^2$ be the left and right shift operators.

	Then $\| L \| = 1$, and so $\sigma(L) \subseteq \overline D(0,1)$.  If $| \lambda | < 1$, then \[
		L(\lambda, \lambda^2, \lambda^3, \dots) = (\lambda^2, \lambda^3, \lambda^4, \dots) = \lambda(\lambda, \lambda^2, \lambda^3,\dots)
	\] 
	and so $\lambda$ is an eigenvalue.  Thus $D(0,1) \subseteq \sigma(L) \subseteq \overline D(0,1)$.  But $\sigma(L)$ is closed, and so $\sigma(L) = \overline D(0,1)$.  Are the $\lambda$ with $| \lambda | = 1$ eigenvalues?  No - suppose $| \lambda | = 1$ and $x \neq 0$ with $Lx = \lambda x$.

	Then \[
		L^n (x) = \lambda^n x.  
	\]  Thus, $x_{n+1} = \lambda^n x_1$.  Then $x = (x_1, \lambda, x_1, \lambda^2 x_1, \dots)$ which is not in $\ell^2$.

	Then $\| R \| = 1$, and so $\sigma(R) \subseteq \overline D(0,1)$.

	\begin{note}
		$LRx = L(0, x_1, \dots) = (x_1, x_2, \dots)$, so \[
			LR = I \tag{$\star$}
		\]
	\end{note}
	\begin{rem}
		Unlike $\dim(X) < \infty$, $(\star)$ does NOT say that $R$ is invertible $(RL = I)$.
	\end{rem}

	Consider the operator $L(\lambda I - R) = \lambda L - I = -\lambda(\lambda^{-1} I - L)$.  If $0 < | \lambda | < 1$, then we know that $\lambda^{-1}I - L$ is invertible (as $\lambda^{-1} \notin \sigma(L)$).  So if $\lambda I - R$ were invertible, then $L$ is invertible, which is false. Thus $\lambda \in \sigma(R)$. Hence \[
		D(0,1) \backslash \{ 0 \} \subseteq \sigma(R) \subseteq \overline D(0,1).  
	\]  Since $\sigma(R)$ is closed, $\sigma(R) = \overline D(0,1)$. 
\end{exmp} 
% subsection spectral_theory (end)
% section lecture_17_monday_1_may (end)

\section{Lecture 18 - Wednesday 4 May} % (fold)
\label{sec:lecture_18_wednesday_4_may}
\begin{thm}
	Let $X \neq \{ 0 \}$ be a Banach space over $\C$.  Let $T \in \mathcal L(X)$.  Then $\sigma(T)$ is a nonempty, compact subset of \[
		\{ \lambda \in \C \given | \lambda | \leq \| T \| \}.
	\]
\end{thm}

\begin{lem}
	With above assumptions $\sigma(T) \subseteq \{ \lambda \in \C \given | \lambda | \leq \| T \| \}$.  
\end{lem}

\begin{proof}
	We need to show that if $|\lambda | > \| T \|$ then $\lambda I - T$ is invertible.

	\textbf{Technique: Geometric series}.  We guess \[
		(\lambda I - T)^{-1} = \frac{1}{\lambda I - T} = \sum_{k = 0}^\infty \frac{T^k}{\lambda^{k+1}}.
	\]  We now verify this guess.  Since $$\sum_{k=0}^{\infty} \frac{\| T^k \|}{|\lambda|^{k+1}} \leq \sum_{k=0}^\infty \frac{\| T \|^k}{|\lambda|^{k+1}} < \infty,$$ the series $S = \sum_{k=0}^\infty \frac{T^k}{\lambda^{k+1}}$ converges in X.

	We now show that $S$ is the inverse of $\lambda I - T$.  As we are working in infinite dimensions, we ned to check left and right inverses. Let $S_n = \sum_{k=1}^{n-1} \frac{T^k}{\lambda^{k+1}}$. Then 
	\begin{align*}
		S_n (\lambda I - T) &= \left( \sum_{k=0}^{n-1} \frac{T^k}{\lambda^{k+1}} \right) \left( \lambda I - T \right) \\
		&= I - \frac{T^n}{\lambda^n} \rightarrow I \\
		(\lambda I - T) S_n &= I - \frac{T^n}{\lambda^n} \rightarrow I
	\end{align*} and so $S(\lambda I - T) = (\lambda I - T) S$ and so $\lambda I - T$ is invertible.  
\end{proof}

\begin{exer}
	Show that if $\| I - T \| < 1$ then $T$ is invertible with inverse $\sum_{k=0}^\infty (I-T)^k$ \emph{Hint: Consider \[
		\frac{1}{T} = \frac{1}{I - (I-T)}.
	\]}
	In particular, the ball $B(I, 1)$ in $\mathcal L(X)$ consists of invertible elements.  
\end{exer}

The following is used to show $\sigma(T)$ is closed and nonempty, it is also interesting in its own right.  \begin{prop}
	Let $X$ be Banach over $\K$.  Let $\text{GL}(X) = \{ T \in \mathcal L(X) \given \text{$T$ invertible}$.  Then \begin{enumerate}[(a)]
		\item $\text{GL}(X)$ is a group under composition of operators.
		\item $\text{GL}(X)$ is open in $\mathcal L(X)$.  
		\item The map \mapping{\phi}{\text{GL}(X)}{\text{GL}(X)}{T}{T^{-1}} is continuous.
	\end{enumerate}
\end{prop}

\begin{proof}
	\begin{enumerate}[(a)]
		\item The open mapping theorem tells us that if $T \in \text{GL}(X)$ then $T^{-1} \in \mathcal L(X)$, and so $T^{-1} \in \text{GL}(X)$.  The rest is clear.
		\item Let $T_0 \in \text{GL}(X)$.  We claim \[
			B\left(T_0, \frac{1}{\| T_0^{-1}} \|\right) \subseteq \text{GL(X)}.
		\]  We have \begin{align*}
			\| I - T_0^{-1} T \| 	&= \| T_0^{-1}(T_0 - T) \| \\ 
									&\leq \| T_0^{-1} \| \| T_0 - T \| \\
									&< 1 \quad \text{as $T \in B\left(T_0, \frac{1}{\| T_0^{-1} \|}\right) $}
		\end{align*}
		\item We have \begin{align*}
			\| T_0^{-1} - T^{-1} \| &= \| T^{-1} ( T - T_0) T_0^{-1} \ | \\
									&\leq \| T^{-1} \| T - T_0 \| \| T_0^{-1} \| \tag{$\star$}
		\end{align*}  If $\| T - T_0 \| \leq \frac{1}{2 \| T_0^{-1} \|}$, then 
		\begin{align*}
			\| I - T T_0^{-1} \| &= \| (T_0 - T) T_0^{-1} \| \\
								&\leq \| T_0 - T \| \| T_0^{-1} \| \\
								&\leq \frac{1}{2}.
		\end{align*}  We then have 
		\begin{align*}
			\| T_0 T^{-1} \| 	&= \| (T T_0^{-1})^{-1} \| \\
								&= \| \sum_{k=0}^\infty ( I - T T_0^{-1})^k \| \\
								&\leq \sum_{k=0}^\infty \| I - T T_0^{-1} \|^k \\
								&\leq 2
		\end{align*}

		Hence $\|T^{-1}\| = \|T_0^{-1}( T_0 T^{-1}) \| \leq \| T_0^{-1} \| \| T_0 T^{-1} \| \leq 2 \| T_0^{-1} \|$, and from $(\star)$, we have \[
			\| T_0^{-1} - T^{-1} \| \leq 2 \| T_0^{-1} \|^2 \| T - T_0 \|
		\] and so $T \mapsto T^{-1}$ is continuous.  
	\end{enumerate}
\end{proof}

\begin{cor}
	$\sigma(T)$ is closed.
\end{cor}

\begin{proof}
	Let \mapping{f}{\C}{\mathcal L(X)}{\lambda}{\lambda I - T}  This is continuous, as \begin{align*}
		\| f(\lambda) - f(\lambda_0) \| &= \| (\lambda - \lambda_0) I \| \\
										&= | \lambda - \lambda_0 |
	\end{align*} and \begin{align*}
			\sigma(T) = f^{-1}\left(\mathcal L(X) \backslash \text{GL}(X) \right)
	\end{align*} which is the inverse image of a closed set, and hence is closed.
\end{proof}

So $\sigma(T)$ is a compact subset of $\{ \lambda \in \C \given | \lambda | \leq \| T \| \}$.  Write $\rho(T) = \C \backslash \sigma(T)$ (the \textbf{resolvent set}), and let $R_T = R : \rho(T) \rightarrow \mathcal L (X)$ with $R_T(\lambda) = (\lambda I - T)^{-1}$.

\begin{thm}
	Let $\K = \C$ and $X \neq \{ 0 \}$ and $T \in \mathcal L(X)$.  Then $\sigma(T) \neq \emptyset$.
\end{thm}

\begin{proof}
	We use Lioville's theorem - a bounded entire function must be constant.

	Let $\phi = \mathcal L(X)'$ (hence $\phi : \mathcal L(X) \rightarrow C$.)  Let \mapping{f_\phi}{\rho(T)}{\C}{\lambda}{\phi(R(\lambda))} 
	\begin{lem}
	 $f_\phi$ is analytic on $\rho(T)$.  		
	\end{lem}
	\begin{proof}
		We show $f_\phi$ is differentiable.  Consider \begin{align*}
			\frac{f_\phi(\lambda) - f_\phi(\lambda_0)}{\lambda - \lambda_0} &= \phi\left(\frac{R(\lambda) - R(\lambda_0)}{\lambda - \lambda_0} \right)  \\
			&= \phi \left( \frac{(\lambda I - T)^{-1} - (\lambda_0 I - T)^{-1}}{\lambda - \lambda_0}  \right) \\
			&= \phi \left( \frac{(\lambda_0 I - T)^{-1}( (\lambda_0 - \lambda) I)(\lambda I - T)^{-1}}{\lambda - \lambda_0} \right) \\
			&= -\phi \left( (\lambda_0 I - T)^{-1} (\lambda I - T)^{-1} \right) \\
			&\rightarrow -\phi \left( (\lambda_0 I - T)^{-2} \right)
		\end{align*} 
		as $\lambda \rightarrow \lambda_0$, where we use the fact that $\phi$ is continuous and $T \rightarrow T^{-1}$ is continuous.    So $f_\phi$ is analytic on $\rho(T)$ for all $\phi \in \mathcal L(X)'$.  
	\end{proof}

	Now suppose that $\sigma(T) = \emptyset$.  Then $f_\phi : \C \rightarrow \C$ is analytic.

	\begin{lem}
		$f_\phi$ is bounded.  
	\end{lem}
	\begin{proof}
		If $|\lambda > \| T \|$, then \begin{align*}
			f_\phi(\lambda) &= \left| \phi \left( ( \lambda I - T)^{-1} \right) \right| \\
							&= \left| \phi \left( \sum_{k=0}^\infty \frac{T^k}{\lambda^{k+1}} \right) \right| \\
							&\leq \| \phi \| \left \| \sum_{k=0}^\infty \frac{T^k}{\lambda^{k+1}} \right \| \\
							&\leq \| \phi \| \sum_{k=0}^\infty \frac{\| T \|^k}{| \lambda|^{k+1}} \\
							&= \frac{ \| \phi \|}{ |\lambda | - \| T \|} \rightarrow 0
		\end{align*} as $|\lambda| \rightarrow \infty$.  So $f_\phi$ is bounded, entire, and thus $f_\phi = c$ by Lioville's theorem.  By the above, $f_\phi(\lambda) = 0$ for all $\lambda$.  Hence $\phi(R(\lambda)) = 0$ for all $\lambda, \phi$.

		Thus from Hahn-Banach, $R(\lambda) = 0$ for all $\lambda$ which is a contradiction, as the zero operator is not invertible if $X \neq \{ 0 \}$.   
	\end{proof}
\end{proof}

% section lecture_18_wednesday_4_may (end)
\section{Lecture 19 - Monday 9 May} % (fold)
\label{sec:lecture_19_monday_9_may}
\begin{thm}[Spectral mapping theorem (polynomials)]
	Let $T$ be an $n \times n$ matrix over $\C$.  IF we know all the eigenvalues of $T$, then we know the eigenvalues of every polynomial $p(T) = a_0 + a_1 T + \dots + a_n T^n$.  Specifically, \[
		\{ \text{eigenvalues of $p(T)$} \} = \{ p(\lambda) \given \text{$\lambda$ is an eigenvalue of $T$} \}
	\] Therefore \[
		\sigma(p(T)) = p(\sigma(T)).
	\]  This is called the \textbf{spectral mapping theorem} (for matrices/polynomials).

	This also holds for $X$ Banach over $\C$, and $T \in \mathcal L(X)$.
\end{thm}

\begin{lem}
	Let $\C[t]$ be the algebra of polynomials in $t$ with complex coefficients.  Multiplication is defined as usual.
\end{lem}

\begin{lem}
	Let $X$ be Banach over $\C$.  Let $T \in \mathcal L(X)$.  Then \mapping{\phi}{\C[t]}{\mathcal L(X)}{p}{p(T)} is an algebra homomorphism (multiplication corresponds to composition in $\mathcal L(X)$.)
\end{lem} 
\begin{proof}
	Simply check \begin{align*}
		\phi(p_1 + p_2) = \phi(p_1) + \phi(p_2) \\
		\phi(p_1 p_2) = \phi(p_1)\phi(p_2) \\
		\phi(\alpha p) = \alpha \phi(p)
	\end{align*} for all $p_1, p_2, p \in \C[t], \alpha \in \C$.  
\end{proof}
\begin{thm}
	Let $X$ be Banach over $\C$, and let $T \in \mathcal L(X)$.  Then \[
		\sigma(p(T)) = p(\sigma(T)).
	\]
\end{thm}
\begin{proof}
	If $p = c$ is constant, then $p(T) = cI$ has spectrum \[
		\sigma(p(T)) = \sigma(cI) = \{ c \}
	\]  On the other hand, \[
		p(\sigma(T)) = \{ c \}
	\]  
	Now, suppose that $p$ is non constant.  Let $\mu \in \C$ fixed.  By the fundamental theorem of algebra, we can factorise $\mu - p(t)$ as \[
		\alpha( t - \lambda_1)^{m_1} \dots (t - \lambda_n)^{m_n}
	\] where $\lambda_1, \dots, \lambda_n$ are the distinct roots of $\mu - p(t)$.  Note that $\mu = p(\lambda_i)$ for each $i$.  Applying $\psi: \C[t] \rightarrow \mathcal L(X)$ from above, we have \begin{align*}
		\mu I - p(T) = \alpha(T- \lambda_1 I)^{m_1} \dots (T - \lambda_n I)^{m_n}
	\end{align*}
	\begin{exer}
		If $T_1, \dots, T_n \in \mathcal L(X)$ which commute with each other, then $T_1 \dots T_n$ is invertible if and only if the individual elements are invertible.  
	\end{exer}

	We know \begin{align*}
		\mu \in \sigma(p(T)) &\iff \mu - p(T) \, \text{is not invertible} \\
			&\iff T - \lambda I \, \text{non invertible for some $i$} \\
			&\iff \lambda \in \sigma(T) \, \text{for some $i$} \\
			&\iff \mu = p(\lambda_i) \in p(\sigma(T))
	\end{align*} and so \[
		\sigma(p(T)) = p(\sigma(T))
	\] 
\end{proof}

\begin{defn}[Spectral radius]
Let $X \neq \{ 0 \}$ be a Banach space over $\C$.  The \textbf{spectral radius} of $T \in \mathcal L(X)$ is
\begin{align*}
	r(T) &= \sup \{ | \lambda | \given \lambda \in \sigma(T) \} \\
	&= \max \{ |\lambda | : \lambda \in \sigma(T)\}
\end{align*}
\end{defn}
\begin{note}
	\[
		r(T) \leq \| T \|
	\] since $\sigma(T) \subseteq \{ \lambda \in \C \given | \lambda | \leq \| T \| \}$.  Strict inequality can (and often does) occur.
\end{note}

\begin{exmp}
	Let \[
		T = \begin{pmatrix}
			0 & 1 \\
			0 & 0 
		\end{pmatrix}.
	\]  Then consider $T : \C^2 \rightarrow \C^2$ where $\| (x, y) \|_2 = \sqrt{|x|^2 + |y|^2}$.  Then \begin{align*}
		\| T \| &= \sup \{ \| Tx \|_2 \given x \in \C^2 \} \\
				&= \sqrt{\lambda_{max} (T^* T)} 
	\end{align*} where \[
		T^* = \begin{pmatrix}
			0 & 0 \\
			1 & 0
		\end{pmatrix}
	\] is conjugate transpose.  Then $\| T \| = 1$.  But $\sigma(T) = \{ 0 \}$, and so $r(T) = 0 < 1 = \| T \|$.  
\end{exmp}

\begin{thm}[Gelfand, 1941]
	Let $X \neq \{ 0 \}$ be Banach over $\C$, and let $T \in \mathcal L(X)$.  Then \begin{align*}
		r(T) = \lim_{n \rightarrow \infty} \| T^n \|^{1/n}.
	\end{align*} In particular, the limit exists.  
\end{thm}
\begin{proof}
	By the spectral mapping theorem, \[
		\sigma(T^n) = \{\sigma(T)\}^n = \{ \lambda^n \given | \lambda \in \sigma(T) \}.
	\]  So \begin{align*}
		r(T) &= r(T^{n})^{1/n} \\
			&\leq \| T^n \|^{1/n}.  
	\end{align*} So \[
		r(T) \leq \liminf_{n \rightarrow \infty} \| T^n \|^{1/n}
	\]

	Now, we must show that \[
		\limsup_{n \rightarrow \infty} \| T^n \|^{1/n} \leq r(T).
	\]
	Let $\phi \in \mathcal L(X)$ and let \mapping{f_\phi}{\rho(T)}{\C}{\lambda}{\phi((\lambda I - T)^{-1})}

	We saw that $f_\phi$ is analytic on $\rho(T)$.  We also have \[
		f_\phi(\lambda) = \sum_{n = 0}^\infty \frac{1}{\lambda^{n+1}} \phi(T^n) \tag{$\star$}
 	\] if $| \lambda | > \| T \|$.  By general theory of Laurent series, ($\star$) actually holds for all $\lambda \in \rho(T)$.  In particular, it holds if $| \lambda | > r(T)$.

Thus, \begin{align*}
	\lim_{n \rightarrow \infty} \frac{1}{\lambda^{n+1}}\phi(T^n) = 0 \quad \boxed{|\lambda| > r(T)}
\end{align*}  Sp for each $\phi \in \mathcal L(X)'$, and each $|\lambda| > r(T)$, there is $C_{\lambda, \phi}$ such that \[
	\left| \phi\left( \frac{1}{\lambda^{n+1}} T^n \right) \right| \leq C_{\lambda, \phi} \quad \forall n \geq 0
\]  Then by the principle of uniform boundedness, there exists a constant $C_\lambda$ such that \[
	\left\| \frac{1}{\lambda^{n+1}} T^n \right\| \leq C_\lambda \quad \forall n \geq 0
\]  So $\| T^n \|^{1/n} \leq |\lambda | (C_\lambda | \lambda|)^{1/n}$, which gives \[
	\limsup_{n \rightarrow \infty} \| T^n \|^{1/n} \leq \lambda
\] for all $| \lambda | > r(T)$.  So \[
	\limsup_{n \rightarrow \infty} \| T^n \|^{1/n} \leq r(T)
\]  We used the following lemma.

\begin{lem}
	Let $X$ be a normed vector space, $A \subseteq X$ a subset.  We say that \begin{enumerate}[(1)]
		\item $A$ is \textbf{bounded} if there exists $C > 0$ with $\| x \| \leq C$, for all $x \in A$.
		\item $A$ is \textbf{weakly bounded} if for each $\phi \in X'$, there exists $C_\phi > 0$ such that \[
			| \phi(x) | \leq C_\phi
		\] for all $x \in A$.
	\end{enumerate}

	Then we have \[
		\text{$A \subseteq X$ is bounded}  \iff \text{weakly bounded}
	\]
\end{lem}
\begin{proof}
	$A$ bounded $\Rightarrow$ $\| x \| \leq C$ for all $x \in A$ $\Rightarrow$ $|\phi(x)| \leq \| \phi \| \| x \| \leq \|\phi \| C$.  So $A$ is weakly bounded.

	Now, suppose $A$ is weakly bounded.  For each $x \in X$, let $\hat x \in X''$ with \[
		\hat x(\phi) = \phi(x).
	\]  So $| \hat x(\phi) | \leq C_\phi$ for all $x \in A$.  By the principle of uniform boundedness, \[
		\| \hat x \| \leq C
	\] for all $x \in A$, and since $\| \hat x \| = \| x \|$.  Thus $A$ is bounded.  
\end{proof}
\end{proof}  
% section lecture_19_monday_9_may (end)

\section{Lecture 20 - Wednesday 11 May} % (fold)
\label{sec:lecture_20_wednesday_11_may}

We now turn to compact operators.  In general, calculating $\sigma(T)$ is difficult, but for compact operators on a complex Banach space, we have a fairly explicit theory.

\begin{thm}
	Let $X$ be a complex Banach space, with $\dim(X) = \infty$.  Let $T : X \rightarrow X$ be a compact operator.  Then \begin{enumerate}[(1)]
		\item $0 \in \sigma(T)$.  
		\item $\sigma(T) \backslash \{ 0 \} = \sigma_p(T) \backslash \{ 0 \}$, that is, each $\lambda \in \sigma(T) \backslash \{ 0 \}$ is an eigenvalue of $T$ ($0$ may or may not be an eigenvalue.)
		\item We are in exactly one of the cases:
		 \begin{itemize}
		 	\item $\sigma(T) = \{ 0 \}$.
		 	\item $\sigma(T) \backslash \{ 0 \}$ is finite (nonempty).
		 	\item $\sigma(T) \backslash \{ 0 \}$ is a sequence of points converging to $0$.  
		 \end{itemize}
		 \item Each $\lambda \in \sigma(T) \backslash \{ 0 \}$ is isolated, and the eigenspace $\ker(\lambda I - T)$ is finite dimensional.  
	\end{enumerate}  
	where $\sigma_p(T)$ is the \textbf{point spectrum of $T$}, where \begin{align*}
		\sigma_p(T) &= \{ \lambda \in \K \given \text{$\lambda I - T$ is not injective}\} \\
		 			&= \{ \lambda \in \K \given \text{there exits nonzero vector $x$ with $(\lambda I - T)x = 0$}\} \\
					&= \{ \text{eigenvalues of $T$} \}
	\end{align*}

\end{thm}
\begin{proof}
	We shall prove these results next week.
\end{proof}

\begin{defn}
	Let $X, Y$ be normed vector spaces.  An operator $T : X \rightarrow Y$ is \textbf{compact} if $T$ is linear, and if $B \subseteq X$ is bounded then $T(B) $ is relatively compact (a set is relatively compact if its closure is compact.)  Symbolically, \[
		\text{$B\subseteq X$ bounded} \Rightarrow \text{$\overline{T(B)}$ compact}
	\]
\end{defn}

\begin{lem}
	IF $T$ is compact, then $T$ is continuous.
\end{lem}
\begin{proof}
	The closed ball $B  = \{ x \in X \given \| x \| \leq 1 \}$ is bounded, and so if $T$ is a compact operator, then $\overline{T(B)}$ is compact, and hence bounded.  Hence $\| Tx \| \leq M$ for all $\| x \| \leq 1$, so $T$ is continuous, with $\| T \| \leq M$.  
\end{proof}

We now recall definitions of compactness
\begin{thm}[Characterisations of compactness]
	Let $X$ be a metric space. The following are equivalent.  
	\begin{enumerate}[(1)]
		\item $X$ is \textbf{compact} (every open cover has a finite subcover).
		\item $X$ is \textbf{sequentially compact} (every sequence in $X$ has a convergent subsequence)
	\end{enumerate}
\end{thm}

\begin{lem}
	Let $X$ be a compact set.  Let $Y \subseteq X$. If $Y \subseteq X$ is closed, then $Y$ is compact.
\end{lem}

\begin{lem}
	Let $V$ be a finite dimensional vector space.  If $X \subseteq V$ is closed and bounded, then $X$ is compact.
\end{lem}

\begin{thm}[Characterisations of compact operators]
	\label{thm:charcompact}
	Let $X, Y$ be normed vector spaces over $\K$.  Let $T \in \mathcal L(X,Y)$.  Then the following are equivalent.
	\begin{enumerate}[(a)]
		\item $T$ is compact.
		\item $\overline{T(B)}$ is compact, where $B = \{ x \in X \given \| x \| \leq 1 \}$.  
		\item If $(x_n)_{n \geq 1}$ is bounded in $X$, then $(Tx_n)_{n \geq 1}$ has a convergent subsequence (\textbf{sequentially compact}).
	\end{enumerate}
\end{thm}

\begin{proof}
	$(a) \Rightarrow (b)$ by definition.

	$(b) \Rightarrow (a)$.  Suppose $(b)$ holds.  Let $B_1 \subseteq X$ be bounded.  THen $B_1 \subseteq \alpha B$ for some $\alpha > 0$.  So \[
		\overline{T(B)} \subseteq \overline{T(\alpha B)} = \alpha \overline{T(B)}
	\] which is a closed subset of a compact set, and hence compact.

	$(a) \Rightarrow (c)$.  Suppose $T$ is compact.  Let $(x_n)_{n \geq 1}$ be bounded sequence in $X$.  Then $T(B) = \{ Tx_n \given n \geq 1 \}$ is relatively compact.  So $\overline{T(B)}$ is compact, and hence is sequentially compact, and so has a convergence subsequence.

	$(c) \Rightarrow (a)$.  Let $B \subseteq X$ be bounded.  Let $(y_n)_{n \geq 1}$ be a sequence in $T(B)$.  Then there is $x_n \in B$ with $Tx_n = y_n$.  So $(x_n)_{n \geq 1}$ is a bonded sequence.  By assumption $(Tx_n)_{n \geq 1}$ has a convergent subsequence.  So $\overline{T(B)}$ is sequentially compact, and hence compact.
\end{proof}

\begin{cor}
	The set $\{ \text{compact operators $T : X \rightarrow Y$}\}$ is a vector space.  That is, if $T_1, T_2$ are compact, then $T_1 + T_2$ and $\alpha T_1$ are compact.  
\end{cor}

\begin{proof}
	Exercise.  Use $(c)$ from the characterisation of compact operators.
\end{proof}

\begin{cor}
	\[
		\mathcal K(X,Y) \subseteq \mathcal L(X, Y) \subseteq \text{Hom}(X,Y)
	\] where $\mathcal K(X,Y)$ is the set of compact operators $T : X \rightarrow Y$. 
\end{cor}

\begin{exmp}[Finite rank operators]  
	Let $X, Y$ be normed vector spaces, and let $T \in \mathcal L(X,Y)$.  If $\dim(\im T) < \infty$, then $T$ is said to have \textbf{finite rank.}  Then if $T$ has finite rank, then $T$ is compact.
\end{exmp}

\begin{proof}
	Let $(x_n)$ be a bounded sequence in $X$.  Then $\| Tx_n \| \leq \| T \| \| x_n \|$ so $(Tx_n)$ is a bounded sequence in $\im T$.  But $\im T$ is finite dimensional, and so $\overline{ \{ Tx_n \given n \geq 1 \}}$ is compact (closed and bounded), and so $(Tx_n)_{n \geq 1}$ has a convergent subsequence.  By $(c)$ in Theorem \ref{thm:charcompact}, $T$ is compact. 
\end{proof}

\begin{lem}
	Let $X, Y$ be normed vector spaces.  If $T \in \mathcal L(X,Y)$ has finite rank, then there exists $y_1, \dots, y_n \in \im T$ and $\phi_1, \dots, \phi_n \in X'$ with $Tx = \sum_{j=1}^n \phi_j(x) y_j$ for all $x \in X$, with $n = \dim( \im T)$. 
\end{lem}

\begin{proof}
	Choose a basis $y_1, \dots, y_n$ of $\im T$.  For each $j = 1, \dots, n$, define $\alpha_j \in (\im T)'$ by \[
		\alpha_j(a_1 y_1 + \dots + a_n y_n) = a_j
	\] i.e. coordinate projection.  By Hahn-Banach, we can extend $a_j$ to a continuous linear functional $\tilde a_j \in Y'$.  Let $\phi_j = \tilde a_j \circ T : X \rightarrow \K$.  So $\phi_j \in X'$.  Since \[
		 y = \sum_{j=1}^n \tilde a_j(y) y_j \quad \forall y \in \im T
	\]

	we have \begin{align*}
		Tx 	&= \sum_{j= 1}^n \tilde a_j (Tx) y_j \\
			&= \sum_{j=1}^n (\alpha_j \circ T)(x) y_j	 \\
			&= \sum_{j=1}^n \phi_j(x) y_j \quad \forall x \in X.
	\end{align*}
\end{proof}
% section lecture_20_wednesday_11_may (end)

\section{Lecture 21 - Monday 16 May} % (fold)
\label{sec:lecture_21_monday_16_may}

Recall that the closed unit ball in $X$ is compact if and only if $\dim(X) < \infty$.  Then it follows that the identity map $I : X \rightarrow X$ is compact if and only if $\dim(X) < \infty$.  Hence, \[
	\mathcal K(X) \subsetneq \mathcal L(X) \subsetneq \text{Hom}(X, X) 
\] when $\dim(X) = \infty$.

Consider a sequence of compact operators $T_n$.  If $T_n$ is compact and $T_n \rightarrow T$, then $T$ is compact.

\begin{lem}[Riesz's Lemma]
	Let $X$ be a normed vector space.  Let $Y \subsetneq X$ be a proper \textbf{closed} subspace.  Let $\theta \in (0, 1)$ be given. THen there exists $x$ with $\| x \| = 1$ such that $\| x - y \| \geq \theta$ for all $y \in Y$.  
\end{lem}
\begin{proof}
	Pick any $z \in X \backslash Y$.  Let $\alpha = \inf_{y \in Y} \| z - y \| > 0$ since $Y$ is closed.  Then by the definition of the infimum, there is $y_0 \in Y$ with $\alpha \leq \| z - y_0 \| \leq \frac{\alpha}{\theta}$.  Now let $x = \frac{x - y_0}{\| z - y_0 \|}$.  Then $\| x \| = 1$.

	Now, \begin{align*}
		\| x - y \| &= \left \| \frac{z - y_0}{\| z - y_0 \|} - y \right \| \\
					&= \frac{1}{\| z - y_0 \|} \left \| z - y_- + \| z - y_0 \| y \right \| \\
					&\geq \frac{\theta}{\alpha} \alpha = \theta
	\end{align*}
\end{proof}

\begin{cor}
	Let $X$ be a normed vector space.  The closed unit ball $\overline B(0, 1)$ is compact if and only if $\dim(X) < \infty$.  
\end{cor}
\begin{proof}
	If $\dim(X) < \infty$ then $\overline B(0, 1)$ is compact (since closed and bounded if and only if compact in finite dimensions).  Now suppose $\dim(X) = \infty$.  Build a sequence $(x_n)$ with $\| x_n \| = 1$ with no convergent subsequence.  Choose finite dimensional subspaces \[
		\{ 0 \} = X_0 \subsetneq X_1 \subsetneq X_2 \subsetneq \dots
	\]

	These are all closed (finite dimensional spaces are complete, and hence closed).  Use the lemma to choose $x_k \in X_k$ with $\| x_k \| = 1$, $\| x_k - x \| \geq \frac{1}{2}$ for al $ x \in X_{k-1}$.  So $x_k - x \| \geq \frac{1}{2}$ for all $x \in X_j (j \leq k - 1)$.  So $\| x_n - x_m \| \geq \frac{1}{2}$ for all $m , n \geq 1$.  So $(x_n)$ has no convergent subsequence, and so $\overline B(0, 1)$ is not compact.  
\end{proof}
\begin{cor}
	$I : X \rightarrow X$ is compact if and only if $\dim(X) < \infty$.  
\end{cor}
\begin{proof}
	Recall $T$ is compact if and only if $T(\overline B(0, 1))$ is relatively compact.	
\end{proof}

\subsection{Limits of compact operators} % (fold)
\label{sub:limits_of_compact_operators}
One way to show that an operator is compact is to apply the following.  
\begin{prop}
	Let $X$ be a normed vector space, and let $Y$ be Banach.  Suppose that $T_n \in \mathcal K(X, Y)$ for each $n \geq 1$.  If $T_n \rightarrow T$ (in operator norm, $\| T_N - T \| \rightarrow 0$) then $T$ is compact.  
\end{prop}
\begin{proof}
	Let $(x_n)$ be a bounded sequence in $X$.  We now construct a subsequence $(x'_n)$ for which $(Tx'_n)$ converges.  \begin{itemize}
		\item Since $T_1$ is compact, $(x_n)$ has a subsequence $x^{(1)}_n$ such that $(T_1 x^{(1)}_n)$ converges. 
		\item Since $T_2$ is compact and $x^{(1)}_n$ is bounded, there is a subsequence $x_n^{(2)}$ such that $T_2 x^{(2)}_n$ converges. 
		\item Continuing, we can form a subsequence $x^{(k)}_n$ such that $T_k x^{k}_n$ converges. 
 	\end{itemize}

Let $x_n' = x_n^{(n)}$.  Then $(x'_n)$ is a subsequence of $(x_n^{(1)})$, and $(x'_n)_{n \geq 2}$ is a subsequence of $(x_n^{(2)})$, etc.  So for each fixed $k \geq 1$, $(T_k x'_n)$ converges.

We now show $T x'_n$ is Cauchy, and hence converges.  We have \begin{align*}
	\| Tx'_m - T x'_n \| \leq \| T x'_m - T_k x'_m \| + \| T_K x'_m - T_k x'_n \| + T_k x'_n - T x'_n \|
\end{align*} where $k$ is to be chosen.  Suppose $\| x_n \| \leq M$ for all $n \geq 1$.  THen \begin{align*}
	\| Tx'_m - T x'_n \| \leq 2 M \| T - T_k \| + \| T_k x'+m - T_k x'_n \|
\end{align*} Let $\epsilon > 0$ be given. Since $\| T - T_k \| \rightarrow 0$ as $ k \rightarrow \infty$, fix a $k$ for which $\| T - T_k \| \leq \frac{\epsilon}{3M}$.  For this fixed $k$, we know $(T_k x'_n)$ converges, and so is Cauchy.  So there exists $N < 0$ such that $\| T_k x'm - T_k x'_n\| < frac{\epsilon}{3}$ for all $m,n < N$.  Hence $\|T x'_m - Tx'_n \| \leq \frac{2M}{\epsilon}{3M} + \frac{\epsilon}{3} = \epsilon$ for all $m , n > N$, so is Cauchy, and so converges. 
\end{proof}

\begin{exmp}
	Let $K(x, y) \in L^2(\R^2)$.  Define $T : L^2(\R) \rightarrow  L^2(\R)$ by \[
		Tf(x) = \int_\R K(x, y) f(y) \, dy
	\] (Hilbert-Schimidt Integral operator)
\end{exmp}
\begin{prop}
	$T$ is compact.  
\end{prop}
\begin{proof}
	Note that $\| Tf \|_2 \leq \| K \|_2 \| f \|_2$ for all $f \in L^2(\R)$, where $\| K \|_2 = \left( \iint_{\R^2} |K(x, y)|^2 \, dx \, dy \right)^{1/2}$.  So $T$ is continuous, with $\| T \| \leq \| K \|_2$.  We now exhibit $T$ as a limit of finite rank (hence compact) operators, with $T_n : L^2(\R) \rightarrow L^2(\R)$.  Once can see that there is a sequence $K_n \in L^2(\R^2)$ of the form \[
		K_n(x, y) = \sum_{k=1}^{N_n} \alpha_k^{(n)}(x) \beta_{k}^{(n)}(y)
	\] with $K_n \rightarrow K$ in $L^2(\R^2)$.  Then $\| T_n - T \| \leq \| K_n - K \|_2 \rightarrow 0$, and so $T_n \rightarrow T$.  Hence 
	\begin{align*}
		T_n f(x) \sum_{k=1}^{N_n} \int_\R \alpha_k^{(n)}(x) \beta_k^{(n)}(y) f(y) \, dy \\
		&= \sum_{k=1}^{N_n} \iprod{f, \overline{\beta_k^{(n)}}} \alpha_k^{(n)} (x) 
	\end{align*}  and so $T_nf = \sum_{k=1}^{N_n} \iprod f, \overline{\beta_k^{(n)}} \alpha_k^{(n)}$ from which we use that $T_n$ has finite rank.  
\end{proof}

% subsection limits_of_compact_operators (end)

% section lecture_21_monday_16_may (end)

\section{Lecture 22 - Wednesday 18 May} % (fold)
\label{sec:lecture_22_wednesday_18_may}

\begin{thm}
Let $X$ be a complex Banach space, with $\dim(X) = \infty$.  Let $T : X \rightarrow X$ be a compact operator.  Then \begin{enumerate}[(1)]
	\item $0 \in \sigma(T)$.  
	\item $\sigma(T) \backslash \{ 0 \} = \sigma_p(T) \backslash \{ 0 \}$, that is, each $\lambda \in \sigma(T) \backslash \{ 0 \}$ is an eigenvalue of $T$ ($0$ may or may not be an eigenvalue.)
	\item We are in exactly one of the cases:
	 \begin{itemize}
	 	\item $\sigma(T) = \{ 0 \}$.
	 	\item $\sigma(T) \backslash \{ 0 \}$ is finite (nonempty).
	 	\item $\sigma(T) \backslash \{ 0 \}$ is a sequence of points converging to $0$.  
	 \end{itemize}
	 \item Each $\lambda \in \sigma(T) \backslash \{ 0 \}$ is isolated, and the eigenspace $\ker(\lambda I - T)$ is finite dimensional.  
\end{enumerate}  
where $\sigma_p(T)$ is the \textbf{point spectrum of $T$}, where \begin{align*}
	\sigma_p(T) &= \{ \lambda \in \K \given \text{$\lambda I - T$ is not injective} \\
	 			&= \{ \lambda \in \K \given \text{there exits nonzero vector $x$ with $(\lambda I - T)x = 0$} \\
				&= \{ \text{eigenvalues of $T$} \}
\end{align*}
\end{thm}

Compact operators are very well behaved with respect to composition.  
\begin{prop}
	Let $X, Y, Z$ be normed vector spaces.  
	\begin{enumerate}[(a)]
		\item If $T \in \mathcal K(X, Y)$ and $S \in \mathcal L(Y, Z)$, then $
			ST \in \mathcal K(X, Z)
		$.
		\item If $S \in \mathcal L(X, Y)$ and $T \in \mathcal K(Y, Z)$, then $
			TS \in \mathcal K(X, Z)
		$.
	\end{enumerate}
\end{prop}
\begin{proof}
	\begin{enumerate}[(a)]
		\item Let $(x_n)$ be a bounded sequence in $X$.  Since $T$ is compact, $Tx_n$ has a convergent subsequence, say $T_{x_{n_k}} \rightarrow y \in Y$. Then $(STx_n)$ has a convergent subsequence, namely $STx_{n_k} = S(T_{n_k}) \rightarrow Sy$  by continuity of $S$.  So $ST$ is compact. 
		\item Let $B \subseteq X$ be bounded.  Then $S(B)$ is bounded in $Y$, as $S$ is continuous.  So $TS(B) = T(S(B))$ is relatively compact since $T$ is compact. Hence $TS$ is compact.
	\end{enumerate}
\end{proof}

\begin{cor}[Part (1) of theorem]
	If $X$ is infinite dimensional Banach space, then $0 \in \sigma(T)$.  
\end{cor}
\begin{proof}
	If $0 \notin \sigma(T)$ then $T$ is invertible.  By bounded inverse theorem $T^{-1}$ is continuous, and then $I = T T^{-1}$ is compact, which is a contradiction.
\end{proof}

\begin{thm}[Part (3) of theorem]
	Let $X$ be a normed vector space.  Let $T \in \mathcal K(X)$.  Then $T$ has at most countably many eigenvalues.  If $T$ has infinitely many eigenvalues, then they can be arranged in a sequence converging to zero.    
\end{thm}
\begin{proof}
	We show that for each $N > 0$, we have \[
		\#\{ \lambda \in \sigma_p(T) \given | \lambda | \geq N \} < \infty \tag{$\star$}
	\] 
	 Suppose that there is $N > 0$ such that $(\star)$ fails.  So $\lambda_1, \lambda_2, \dots$ are distinct eigenvalues with $|\lambda_n | \geq N$ for $n = 1, 2, \dots$.  Let $x_n \neq 0$ be an eigenvector.  $Tx_n = \lambda_n x_n$, $n = 1, 2, \dots$.  Let $X_n = \spans\{ x_1, \dots, x_n \}$.  Since $\{ x_n \given \geq 1 \}$ are linearly independent, we have \[
		X_1 \subsetneq X_2 \subsetneq \dots
	\] and each $X_n$ is closed (finite dimensional).

	 By Reisz's Lemma from previous lecture, choose $y_n \in X_n$ such that $\| y_n \| = 1$ and $\| y_n - x \| \geq \frac{1}{2}$ for all $x \in X_{n-1}$.  So $(y_n)$ is bounded in $X$.  We show that $Ty_n$ has no convergence subsequence, contradicting compactness of $T$.

	Let $ m  > n$.  Then \begin{align*}
		\| Ty_m - Ty_n \| 	&= \left\| \lambda_m y_m - \left( \lambda_m y_m - Ty_m + Ty_n \right) \right| \\
							&= |\lambda_m | \left\| y_m - (\text{something in $X_{m-1}$}) \right\| \\
							&\geq \frac{1}{2}|\lambda_m| \geq \frac{1}{2}N
	\end{align*} as required.

	Note that $y_m = a_1 x_1 + \dots + a_m x_m$.  Then \begin{align*}
		\lambda_m y_m - T y_m &=  \lambda_m a_1 x_1 + \dots + \lambda_m a_m x_m - (a_1 \lambda_1 x_1 + \dots + a_m \lambda_m x_m) \\
		&= a_1(\lambda_m - \lambda_1) x_1 + \dots + a_{m-1}(\lambda_m - \lambda_{m-1}) x_{m-1} \in X_{m-1}
	\end{align*} and $T y_n \in X_{m-1}$ since $n< m$.

\end{proof}

\subsection{Projections} % (fold)
\label{sub:projections}

\begin{defn}[Projection operator]
	Let $X$ be a vector space.  A linear operator $P : X \rightarrow X$ is called a projection if $P^2 = P$.
\end{defn}
\begin{prop}
	If $P : X \rightarrow X$ is a projection then $I - P$ is a projection, and \[
		\im I - P= \ker P, \quad \ker I - P = \im P
	\]
\end{prop}

\begin{proof}
	If $P^2 = P$ then $(I - P)^2 = I - 2P + P^2 = I - P$ and so $I -  P$ is a projection.  Let $x \in \im I-P$  Then $x = (I - P)y$ for some $y \in X$.  So $P x = P(I - P)y = (P - P^2) y = 0$.  So $x \in \ker P$ and $\im I - P \subseteq \ker P$.  If $x \in \ker P$ the $Px = 0$.  So $(I - P) x = x$, and $x \in \im(I - P)$. 
\end{proof}

\begin{defn}[Direct sum]
	Let $X$ be a vector space, and let $X_1, X_2$ be subspaces.  Then $X = X_1 \oplus X_2$ (direct sum) if \[
		X = X_1 + X_2
	\] and $X_1 \cap X_2 = \{ 0 \}$.  Equivalently, $X = X_1 \oplus X_2$ if and only if each $x \in X$ can be written in exactly one way as $x = x_1 + x_2$ with $x_1 \in X_1, x_2 \in X_2$.  
\end{defn}
\begin{thm}[Equivalence of direct sums and projections]
	Let $X$ be a vector space.  \begin{enumerate}[(a)]
		\item If $P : X \rightarrow X$ is a projection, then \[
			X = (\im P) \oplus (\ker P) 
		\] 
		\item If $X = X_1 \oplus X_2$, there exists a unique projection with \[
			\im P = X_1, \quad \ker P = X_2.
		\] Specifically, $Px = x_1$ if $x = x_1 + x_2$. 
	\end{enumerate}
\end{thm}

\begin{proof}
	\begin{enumerate}[(a)]
		\item Let $P : X \rightarrow X$ be a projection.  Then we show $X = (\im P) \oplus (\im I - P)$, $x = Px + (I - P) x$.  This shows that $X = \im P + \im I - P$.  If $x \in \im P \cap \ker P$ then $x = Py$ and $Px = 0$.  Hence, $Px = P^2 y = P^y = 0$ and so $ x = 0$.
		\item Exercise.  
	\end{enumerate}
\end{proof}

\begin{prop}
	Let $X$ be Banach.  Let $X = X_1 \oplus X_2$.  Let $P : X \rightarrow X$ be the corresponding projection operator.  Then \[
		P \in \mathcal L (X) \iff \text{$X_1, X_2$ closed}
	\]
\end{prop}
\begin{proof}
	$(\Rightarrow)$.  Suppose $P$ is continuous.  Then $X_1 = \im P = \ker I - P$ and $X_2 = \ker P$ are both closed.  For example, if $x_n \in \ker P$ and $x_n \rightarrow x$, then $0 = P x_n \rightarrow Px$ and so $x \in \ker P$.

	$(\Leftarrow)$.  Suppose that $X_1, X_2$ are closed.  Since $X = X_1 \oplus X_2$, we can define a new norm $\| \cdot \|'$ by $\| x \|' = \| x_1 \| + \| x_2 \|$ where $x = x_1 + x_2$.  
	\begin{exer}{\ }
		\begin{enumerate}[(a)]
			\item Show that $\| \cdot \|'$ is a norm.
			\item Show that $(X, \| \cdot \|')$ is Banach.  This relies on the fact that $(X, \| \cdot \|)$ is Banach and $X_1, X_2$ are closed. 
		\end{enumerate}
	\end{exer}

	Note that $\| x \| = \| x_1 + x_2 \| \leq \| x_1 \| + \| x_2 \| = \| x \|'$, and so by a corollary to the open mapping theorem, there is a $c > 0$ with $\| x \|' \leq c \| x \|$ for all $x \in X$, and so \[
		\| Px \| = \| x_1 \| \leq \| x_1 \| + \| x_2 \| = \| x \|' \leq c \| x \|
	\] and hence $P$ is continuous.  
\end{proof}
% subsection projections (end)

% section lecture_22_wednesday_18_may (end)

\section{Lecture 23 - Monday 23 May} % (fold)
\label{sec:lecture_23_monday_23_may}

\begin{cor}
	Let $X$ be Banach, and let $M$ be a finite dimensional subspace.  Then there exists a \textbf{closed} $N$ with \[
		X = M \oplus N.
	\]
\end{cor}
\begin{proof}
	Let $v_1, \dots, v_n$ be a basis of $M$.  Define, for each $ j = 1, \dots, n$, $\phi_j \in M'$ by $\phi_j(a_1 v_1 + \dots + a_n v_n) = a_j$.  Then using Hahn-Banach to extend $\tilde \phi_j \in X'$.  Let $P : X \in X$ be defined by \[
		Px = \sum_{j =1 }^n \tilde \phi_j(x) v_j.
	\]  Then we need only check that $P$ is linear and continuous, $\im P = M$, and $P^2 = P$. Now take $N = \ker P$ and then $X = M \oplus N$.
\end{proof}.

We are now ready to prove the following theorem.

\begin{thm}
	Let $X$ be Banach, and let $T \in \mathcal K(X)$, and let $\lambda \in \K \backslash \{ 0 \}$.  For all $k \in \N$, we have \begin{enumerate}[(a)]
		\item $\underbrace{\ker(\lambda I - T)^k}_{\text{generalised eigenspace}}$ is finite dimensional.
		\item $\im(\lambda I - T)^k$ is closed.   
	\end{enumerate}
\end{thm}
\begin{proof}
	\textbf{Reductions}.  Since $\ker(\lambda I - T)^k = \ker(I - \lambda^{-1}T)^k$, and similarly for the image, by replacing $T \in \mathcal K(X)$ by $\lambda T \in \mathcal K(X)$, we can assume that $\lambda = 1$.

	Also, we have \begin{align*}
		(I - T)^k 	&= \sum_{n=0}^k {k \choose n} (-1)^n T^n \\
					&= I - T \underbrace{\sum_{n = 1}^k {k \choose n} (-1)^{n-1} T^{n-1}}_\text{continuous}  \\
					&= I - \tilde T.
	\end{align*} where $\tilde T$ is the composition of compact and continuos operators, and so is compact.  So we can take $\lambda = 1, k = 1$.

	\begin{enumerate}[(a)]
		\item The closed unit ball in $\ker I - T$ is \begin{align*}
		\{ x \in \ker I - T \given \| x \| \leq 1 \} &= \{ Tx \given x \in \ker I - T, \ | x \| \leq 1 \} \\
		&\subseteq \overline{T (\overline B(0, 1))}
		\end{align*} which is compact as $T$ is compact.  Hence, the closed unit ball in $\ker I - T$ is compact, and thus $\ker I - T$ is finite dimensional.
		\item Let $S = I - T$.  We then need to show that $\im S$ is closed.  Since $\ker S$ is finite dimensional from above, there is a \textbf{closed} subspace $N$ with \[
			X = (\ker S) \oplus N
		\]  Note that $\im S = S(X) = S(N)$, and that $S |_N : N \rightarrow X$ is injective.

		Suppose that $S(N)$ is not closed.  So there is a sequence $(x_n)$ in $N$ such that $Sx_n \rightarrow y \in X \backslash S(N)$. Then there are two cases
		\begin{case}[$\| x_n \| \rightarrow \infty$]  Let $y_n = \frac{1}{\| x_n \|} x_n$.  Then $Sy_n = \frac{1}{\| x_n \|} S x_n \rightarrow 0$.  But $(y_n)_{ n \geq 1}$ is bounded in $X$, and so there exists a subsequence $y_{n_k}$ such that $T y_{n_k} \rightarrow z$ (as $T$ is compact).  Hence $y_{n_k} = S y_{n_k} + T y_{n_k} \rightarrow 0 + z$.  Thus $z \in N$ (as $y_{n_k} \in N$, and $N$ is closed), and $\| z \| = 1$.

			So $S y_{n_k} \rightarrow 0$, but $S y_{n_k} \rightarrow Sz$ with $z \in N \backslash \{ 0 \}$, by the continuity of $S$.  This contradicts the injectivity of $S|_N$.  
		\end{case}   
		\begin{case}[$ \| x_n \|$ does not tend to infinity]
			So $(x_n)$ has a bounded subsequence $(x_{n_k})$.  Since $T$ is compact, $(x_{n_k})$ has a subsequence such that $(T x_{n_{k_l}})$ converges, to $z_1$ say.  By replacing $x_n$ by this subsequence we can assume that $Sx_n \rightarrow y$, and that $Tx_n \rightarrow z$.  A before, we can write \begin{align*}
			x_n = S x_n + T x_n &\rightarrow y + z.
			\end{align*}  So $x_n$ converges to $x \in N$.  So $S x_n \rightarrow Sx \in S(N)$ by continuity, but we assume that $S x_n \rightarrow y \in X \backslash S(N)$, which achieves our contradiction.
		\end{case}
	\end{enumerate}  
\end{proof}
% section lecture_23_monday_23_may (end)

\section{Lecture 24 - Wednesday 25 May} % (fold)
\label{sec:lecture_24_wednesday_25_may}
Let $T: \C^n \rightarrow \C^n$ be a linear operator.  Then in the simplest case, $T$ has $n$ distinct eigenvalues, and the corresponding eigenvectors are linearly independent, forming a basis for $\C^n$.

Hence, $\C^n = \C x_1 \oplus \dots \oplus \C x_n$ and the matrix of $T$ relative to this basis is simply diagonal with $\lambda_1, \dots, \lambda_n$.

This is not always possible, because there is not always a basis of eigenvectors.  Instead look at the generalised eigenspace, \[
	\{ x \in \C^n \given \text{$(\lambda I - T)^k x = 0$ for some $k \geq 1$}.  
\] 
But $\{ 0 \} \subseteq \ker(\lambda I - T)^1 \subseteq \ker(\lambda I - T)^2 \subseteq \dots$
 and since $\dim(\C^n) < \infty$ this must stabilise.  Let $r \geq 1$ be the fist time that $\ker (\lambda I - T)^r = \ker( \lambda I - T)^{r+1}$.  Then the generalised $\lambda$-eigenspace is just $\ker(\lambda I - T)^r$.  There \textbf{is} a basis of $\C^n$ consisting of generalised eigenvectors, and the matrix of $T$ relative to this basis is in block form.

\begin{defn}[Complete reduction]
	Let $T: X \rightarrow X$ be linear.  If $X = X_1 \oplus X_2$ be can write \[
		Tx = \begin{pmatrix}
			T_{11} & T_{12} \\
			T_{21} & T_{22}
		\end{pmatrix} \begin{pmatrix}
			x_1 \\
			x_2
		\end{pmatrix}
	\] where we identify $x_1 + x_2 \iff (x_1, x_2)$.  Here, \begin{align*}
		T_{11}: X_1 \rightarrow X_1 \\
		T_{12}: X_2 \rightarrow X_1 \\
		T_{21} : X_2 \rightarrow X_2 \\
		T_{22} : X_2 \rightarrow X_2
	\end{align*} we say that $X = X_1 \oplus X_2$ \textbf{completely reduces} $T$ (well adapted to $T$) if \[
		Tx = \begin{pmatrix}
			T_1 & 0 \\
			0	& T_2 
		\end{pmatrix} \begin{pmatrix}
			x_1 \\
			x_2
		\end{pmatrix}
	\]  We write $T = T_1 \oplus T_2$.
\end{defn}

\begin{exer}
	If $X = X_1 \oplus X_2$ completely reduces $T = T_1 \oplus T_2$, then \begin{enumerate}[(a)]
		\item $\ker T = \ker T_1 \oplus \ker T_2$ 
		\item $\im T = \im T_1 \oplus \im T_2$
		\item $T$ is injective if and only if $T_1, T_2$ are injective
		\item $T$ is surjective if and only if $T_1, T_2$ are surjective
		\item If $T$ is bijective, then $X = X_1 \oplus X_2$ completely reduces $T^{-1} = T_1^{-1} \oplus T^{-1}_2$.
	\end{enumerate}
\end{exer}

\begin{cor}
	Let $X = X_1 \oplus X_2$ be Banach, with $X_1, X_2$ closed subspaces.  If $X = X_1 \oplus X_2$ completely reduces $T = T_1 \oplus T_2 \in \mathcal L(X)$, then \begin{enumerate}[(a)]
		\item $T_1 \in \mathcal L(X_1), T_2 \in \mathcal L(X_2)$
		\item $\sigma(T) = \sigma(T_1) \cup \sigma(T_2)$ 
		\item $\sigma_p(T) = \sigma_p(T_1) \cup \sigma_p(T_2)$
	\end{enumerate}  
\end{cor}

\begin{proof}
	Exercise.
\end{proof}

Consider the following chains \begin{align*}
	\{ 0 \} \subseteq \ker S^1 \subseteq \ker S^2 \subseteq \dots \\
	X \supseteq \im S^1 \supseteq \im S^2 \supseteq \dots
\end{align*} where $X$ is a vector space and $S \in \text{Hom}(X, X)$.  It is easy to see that if $\ker S^r = \ker S^{r+1}$ then $\ker S^r = \ker S^{r+k}$.  Similarly for images (p. 109 in Daners.)

There is no reason that these should stabilise in general.
\begin{thm}
	Let $X$ be Banach, $T \in \mathcal K(X), \lambda \neq 0$.  Then both chains (with $S = \lambda I - T$) stabilise.  
\end{thm}
\begin{proof}
	 Without loss of generality, assume $\lambda = 1$, so we can write $S = I - T$.  Suppose that the kernel chain does not stabilise.  Since we assume \[
	 	\ker S^1 \subsetneq \ker S^2 \subsetneq \ker S^3 \subsetneq
	 \]  We know that these are closed (being finite dimensional) subspaces.  So Reisz's Lemma gives $x_n \in \ker S^n$ with $\| x_n \| = 1$, $\|x_n - x \| \geq \frac{1}{2}$ for all $x \in \ker S^{n+1}$. This is a bounded sequence.  We claim that $T x_n$ has no convergent subsequence.

	Let $m > n$.  Then \begin{align*}
		\| Tx_m - Tx_n \| 	&= \| ( I - T) x_n - (I-T) x_m + x_m - x_n \| \\
							&= \| Sx_n - Sx_m - x_m - x_n \|\\
							&= \| x_m - \underbrace{(S x_m  - S x_n  + x_n)}_\text{in $\ker S^{m-1}$} \| \\
							&\geq \frac{1}{2}
	\end{align*}  The image argument is similar - using the fact that the images are closed - proved in the previous lecture.
\end{proof}

\begin{thm}
	Let $X$ be a vector space, $S \in \text{Hom}(X, X)$.  Suppose that \begin{align*}
		\alpha(S) &= \inf \{ r \geq 1 \given \ker S^r = \ker S^{r+1} \} \\
		\delta(S) &= \inf \{ r \geq 1 \given \im S^r = \im S^{r+1} \},
	\end{align*} the \textbf{ascent} and \textbf{descent} of $S$ respectively, are both finite.

	Then \begin{enumerate}[(a)]
		\item $\alpha(S) = \delta(S) = r$, say
		\item $X = \ker S^r \oplus \im S^r$
		\item The direct sum in (b) completely reduces $S$. 
	\end{enumerate} 
\end{thm}
\begin{proof}
	Daner's notes, p. 109.
\end{proof}

\begin{cor}
	Let $X$ be Banach, $T \in \mathcal K(X)$, $\lambda \neq 0$.  Let $r = \alpha(\lambda I - T) = \delta(\lambda I - T)$.  Then $X = \ker(\lambda I - T)^r \oplus \im(\lambda I - T)^r$ and this completely reduces $\mu I - T, \mu \in \K$.
\end{cor}

\begin{cor}
	If $X$ is Banach, $T \in \mathcal K(X), \lambda \neq 0$ then $\lambda I - T$ is injective if and only if $\lambda I - T$ is surjective.  
\end{cor}

\begin{proof}
	\begin{align*}
		&\qquad 		\text{$\lambda I - T$ injective} \\
		&\Rightarrow 0 \in \ker(\lambda I - T)^1 = \ker(\lambda I - T)^2 \\
		&\Rightarrow \alpha(\lambda I - T) = 1 \\
		&\Rightarrow \delta(\lambda I - T) = 1 \\
		&\Rightarrow X = \underbrace{\ker(\lambda I - T)}_{= \{ 0 \}} \oplus \im(\lambda I - T) \\
		&\Rightarrow X = \im (\lambda I - T) \\
		&\Rightarrow \text{$X$ is surjective}
	\end{align*}  The other direction is similar.
\end{proof}

\begin{cor}
	Let $X$ be Banach, $T \in \mathcal K(X)$.  Thus each $\lambda \in \sigma(T) \backslash \{ 0 \}$ is an eigenvalue.  
\end{cor}
\begin{proof}
	Immediate from the previous corollary.
\end{proof}

% section lecture_24_wednesday_25_may (end)

\section{Lecture 25 - Monday 30 May} % (fold)
\label{sec:lecture_25_monday_30_may}
Recall the following.
\begin{cor}
	Let $X$ be Banach, $T \in \mathcal K(X)$, $\lambda \neq 0$.  Let $r = \alpha(\lambda I - T) = \delta(\lambda I - T)$.  Then $X = \ker(\lambda I - T)^r \oplus \im(\lambda I - T)^r$ and this completely reduces $\mu I - T, \mu \in \K$.
\end{cor}

Also note that $\im \ker(\lambda I - T)^r$ is closed, and $\ker (\lambda I - T)^r$ is finite dimensional.  
\begin{exer}
	Let $\lambda 1, \dots, \lambda_n \in \sigma(T) \backslash \{ 0 \}$.  Let $N_j = \ker(\lambda_j I - T)^r_j$ be the generalised $\lambda_j$-eigenspace.  Show that there exists closed subspaces $M$ with \[
		X = N_1 \oplus N_2 \oplus \cdots \oplus M
	\] with $T = T_1 \oplus T_2 \oplus \cdots \oplus T_M $, and so spectral theory tells us how to \emph{diagonalise} $T$.
\end{exer}

In Hilbert spaces we can say even more.  Recall that the adjoint of $T \in \mathcal L(\mathcal H)$ is defined by \[
	\iprod{Tx, y} = \iprod{x, T^* y} \quad \forall x, y \in \mathcal H
\]  Then $T^* \in \mathcal L(\mathcal H)$.  
\begin{defn}[Self-adjoint]
$T \in \mathcal L(\mathcal H)$ is \begin{enumerate}[(a)]
	\item \textbf{Hermitian (self-adjoint)} if $T^* = T$.
	\item \textbf{Unitary} if $T^* T = T T^* = I$.  
	\item \textbf{Normal} if $T^* T = T T^*$.  
\end{enumerate}
\end{defn}

\begin{rem}
	For matrices, we have \begin{enumerate}[(a)]
		\item Hermitian if and only if $\overline{A^T} = A$. 
		\item Unitary if and only if the columns of $A$ are orthonormal.
		\item Hermitian and unitary operators are normal.
	\end{enumerate}
\end{rem}

\begin{prop}
	Let $\mathcal H$ be Hilbert over $\C$.  IF $T \in \mathcal L(\mathcal H)$ is normal, then $r(T) = \| T \|$.  
\end{prop}

\begin{proof}
	For Hermitian operators it is easy.  We have \[
		\| T \|^2 = \| T^* T \| = \| T^2 \|.
	\]  By induction ,we then have $\| T \|^{2^n} = \| T^{2^n} \|$.  So \begin{align*}
		r(T) &= \lim_{n \rightarrow \infty} \| T^n \|^{1/n} \\
			&= \lim_{n \rightarrow \infty} \| T^{2^n} \|^{1/{2^n}} \\
			&= \| T \|.
	\end{align*}

	For normal operators, we have \begin{align*}
		\| T^2 \|^2 &= \| (T^2)^* T^2 \| \\
					&= \| T^*(T^* T) T \| \\
					&= \| T^* T T^* T \| \quad \text{normal} \\
					&= \| (T^* T)^* (T^* T) \|  \\
					&= \| T^* T \|^2 \\
					&= \| T^4 \|   
	\end{align*} and then we have $\| T^2 \| = \| T \|^2$ and the proof follows by induction.
\end{proof}

\begin{cor}
	Let $\mathcal H$ be a Hilbert space over $\C$.  \begin{enumerate}[(a)]
		\item If $T \in \mathcal L(\mathcal H)$ is unitary, then \[
			\sigma(T) \subseteq \mathbb{T} = \{ \lambda \in \C \given | \lambda | = 1 \}\]
		\item If $T \in \mathcal L(\mathcal H)$ is Hermitian, then \[
			\sigma(T) \subseteq \R.
		\]
	\end{enumerate}
\end{cor}
\begin{proof}{\ }\begin{enumerate}[(a)]
	\item On practice sheet.  Use the fact that $\sigma(T^*) = \overline{\sigma(T)}$.
	\item Let $\lambda = a + ib \in \sigma(T)$.  So $\lambda I - T$ is not invertible.  Hence, $(\lambda + it)I - (T + it I)$ is not invertible for all $t \in \R$. 
  Then \begin{align*}
		\| \lambda + it \|^2 	&\leq r(T + itI)^2 \\
								&\leq \| T + itI \|^2 \\
								&= \| (T + itI)^* ( T + itI) \| \\
								&= \| (T - itI)(T + itI) \| \\
								&= \| T^2 + t^2 I \| \\
								&\leq \| T^2 + t^2
	\end{align*}  However, the left hand side is equal to \[
		a^2 + b^2 + 2bt + t^2,
	\] and so we obtain \[
		a^2 + b^2 + 2bt \leq \| T \|^2 \quad \forall t \in \R
	\] and so $b = 0$.  
\end{enumerate}
\end{proof}

\begin{lem}
	Let $\mathcal H$ be Hilbert over $\C$.  Let $T \in \mathcal L(\mathcal H)$, and let \[
		M_\lambda = \{ x \in \mathcal H \given Tx = \lambda x \} = \ker \lambda I - T
	\]  be the $\lambda$-eigenspace of $T$.  Then \begin{enumerate}[(a)]
		\item $M_\lambda \perp M_\mu$ if $\lambda \neq \mu$.  
		\item If $T$ is normal, each $M_\lambda$ is $T$ and $T^*$ invariant.  That is, \[
			T(M_\lambda) \subseteq M_\lambda, \quad T^*(M_\lambda) \subseteq M_\lambda.
		\]
	\end{enumerate}
\end{lem}

\begin{proof}{\ }
	\begin{enumerate}[(a)]
		\item Let $u \in M_\lambda, v \in M_\mu$.  Then \begin{align*}
			(\lambda - \mu) \iprod{u, v} &= \iprod{\lambda u, v} - \iprod{u, \overline \mu v} \\
			&= \iprod{Tu, v} - \iprod{u, T^* v} \\
			&= \iprod{Tu, v} - \iprod{Tu, v} \\
			&= 0
		\end{align*} and so $\iprod{u, v} = 0$.  
		\item If $T$ is normal, then $\ker T = \ker T^*$ as \begin{align*}
			\| Tx \|^2 &= \iprod{Tx, Tx} = \iprod{x, T^* T x} \\
					&= \iprod{x, T T^* x} = \iprod{T*x, T^* x} \\
					&= \| T^* x \|^2.
		\end{align*}  Similarly, if $T$ is normal then $\lambda I - T$ is normal.  Then \begin{align*}
			M_\lambda 	&= \ker \lambda I  - T \quad \text{($T$ invariant)} \\
						&= \ker \overline \lambda I  - T^* \quad \text{($T^*$ invariant)}.
		\end{align*}
	\end{enumerate}
\end{proof}

The spectral theory for compact normal operators in a Hilbert space is particularly nice, as the following theorem demonstrates.
\begin{thm}
	Let $T \in \mathcal L(\mathcal H)$ be compact and normal.  Then \[
		\mathcal H = \overline{\bigoplus_{\lambda \in \sigma(T)} M_\lambda},
	\] the closure of the span of the eigenspaces, and $\mathcal H$ has an orthonormal basis consisting of eigenvectors.  Moreover, $T$ acts diagonally with respect to this basis.   
\end{thm}
\begin{proof}
	Let \[
		M = \overline{\bigoplus_{\lambda \in \sigma(T)} M_\lambda},
	\] a closed subspace.  Hence $H = M \oplus M^\perp$, where \[
	M^\perp = \{ x \in \mathcal H \given \iprod{x, m} = 0 \, \forall m \in M \}.
	\]   We must show that $M^\perp = \{ 0 \}$.  Assume the contrary.  Then consider $\tilde T = M^\perp \rightarrow \mathcal H$ be the restriction of $T$ to $M^\perp$.  Then we have \[
		\tilde T: M^\perp \rightarrow M^\perp
	\] is compact and normal (Exercise). Then  \begin{enumerate}[(a)]
		\item $\sigma(\tilde T) = \{ 0 \}$.  Then $r(\tilde T) = 0$, and so $\| \tilde T \| = 0$, and so $\tilde T = 0$.  Then each $x \in M^\perp \backslash \{ 0 \}$ satisfies $\tilde T x = 0 = 0x$, and so $x \in M_0$ with $M^\perp \subseteq M_0 \subseteq M$, a contradiction (from direct sum decomposition).  Hence $M = \{ 0 \}$.  
		\item $\sigma(\tilde T) \neq \{ 0 \}$.  So there is an eigenvalue $\lambda \in \sigma(T) \backslash \{ 0 \}$.  So there is $x \in M^\perp \backslash \{ 0 \}$ with $\tilde T x = \lambda x$.  o $Tx = \lambda x$, and so $x \in (M_\lambda \cap M^\perp) \backslash \{ 0 \}$, a contradiction.  Hence $M^\perp = \{ 0 \}$.    
	\end{enumerate}

	Choose an orthonormal basis for each $M_\lambda$, and combine to get an orthonormal basis of $\mathcal H$, using $M_\lambda \perp M_\mu$.  
\end{proof}

% section lecture_25_monday_30_may (end)

\section{Lecture 26 - Wednesday 1 June} % (fold)
\label{sec:lecture_26_wednesday_1_june}
\subsection{Fredholm alternative for compact operators on a Hilbert space} % (fold)
\label{sub:fredholm_alternative_for_compact_operators_on_hilbert_space}

% subsection fredholm_alternative_for_compact_operators_on_hilbert_space (end)
Recall that for matrices, we have the following result, known as the Fredholm alternative.

\begin{thm}[Fredhold alternative (Finite dimensional spaces)]
	Let $A: \C^n \rightarrow \C^n$ be linear.  Then exactly one of the following two things occur:
	\begin{enumerate}[(1)]
		\item $Ax = 0$ has only the trivial solution $x = 0$, in which case $Ax = b$ has a unique solution for each $b \in \C^n$.  
		\item $Ax = 0$ has a non-trivial solution, in which case $Ax = b$ has either no solutions, or infinitely many solutions.
	\end{enumerate}
\end{thm}

\begin{defn}[Hilbert-Schmidt integral operators]
	\mapping{T}{L^2([a,b])}{L^2([a,b])}{(Tf)(x)}{\int_a^bK(x, y) f(y) \, dy} where $\| K \|_2$ is finite.  These are compact operators.    
\end{defn}

Consider equations of the following form\[
	\lambda f(x) - \int_a^b K(x, y) f(y) \, dy = g(x),
\] where $\lambda \neq 0$ and $g \in L^2$ are given. This can be rewritten in the form \[
	(\lambda I - T) f = g.
\]  Then we have the following theorem, due to Fredholm.

\begin{thm}[Fredholm alternative (Hilbert spaces)]
	Let $\mathcal H$ be Hilbert over $\C$, and let $T \in \mathcal K(\mathcal H)$.  Then exactly one of the following occurs.
	\begin{enumerate}[(a)]
		\item $(\lambda I - T) = 0$ has only the trivial solution, in which case $(\lambda I - T)x = b$ has a unique solution for each $b \in \mathcal H$.  
		\item $(\lambda I - T) x = 0$ has a non trivial solution, in which case $(\lambda I - T) x = b$ has a solution if and only if $b \perp y$ for every solution $y$ of the equation \[
			(\overline \lambda I - T^*) y = 0
		\]  This is finite dimensional, as it is the kernel of $(\lambda I - T)^*$.   
	\end{enumerate}
\end{thm}

\begin{proof}
	{\ } \begin{enumerate}[(a)]
		\item If $(\lambda I - T) x = 0$ has only the trivial solution, then $\ker \lambda I - T = \{ 0 \}$ and so it is injective.  Hence $\lambda$ is not an eigenvalue, and so $\lambda$ is not a spectral value.  So $\lambda I - T$ is invertible, and so $(\lambda I - T) x = b$ has a unique solution $x = (\lambda I - T)^{-1} b$, which can be expanded into a series expression if $|\lambda| > r(T)$.  
		\item Suppose $(\lambda I - T) x = 0$ has a non-trivial solution.  Then
		\begin{align*}
			&\qquad\text{$(\lambda I - T) x = b$ has a solution} \\
			&\iff b \in \im \lambda I - T \quad \text{which is closed} \\
			&\iff b \in ((\im \lambda I - T)^\perp)^\perp \\
			&\iff b \in (\ker \overline \lambda - T^*)^\perp \\
			&\iff b \perp y \quad \forall y \in \ker \overline \lambda I - T^*.\qedhere
		\end{align*}
	\end{enumerate}
\end{proof}

\begin{prop}[Miscelaneous]{\ }\begin{enumerate}[(a)]
	\item If $M$ is a closed subspace of $\mathcal H$, then $M = M^{\perp \perp}$. 
	\item IF $S: \mathcal H \rightarrow \mathcal H$ and $S \in \mathcal L(\mathcal H)$, then $(\im S)^\perp = \ker S^*$.  
\end{enumerate}
\end{prop}

\begin{proof}{\ }
	\begin{enumerate}[(a)]
		\item Let $m \in M$, then $\iprod{m, x} = 0$ for all $x \in M^\perp$, and so $m \in (M^\perp)^\perp = M^{\perp \perp}$, and so $M \subseteq M^{\perp \perp}$.

		Let $x \in M^{\perp \perp}$.  Since $M$ is closed, $\mathcal H = M \oplus M^{\perp}$, and so $x = m + m^\perp$.  So $x - m \in M^{\perp \perp} + M \subseteq M^{\perp \perp}$, and so $x - m = m^\perp \in M^{\perp \perp}$.  But $M^\perp$ is closed, and so $\mathcal H = M^\perp \oplus M^{\perp \perp}$.  So $x - m - 0$, and $x = m \in M$.  
		\item \begin{align*}
			(\im S)^\perp &= \{ x \in \mathcal H \given \iprod{x, sy} = 0 \quad \forall y \in \mathcal H \} \\
							&= \{ x \in \mathcal H \given \iprod{S^* x, y} = 0 \quad \forall y \in \mathcal H \} \\
							&= \{ x \in H \given S^* x = 0 \} \\
							&= \ker S^*
		\end{align*} 
	\end{enumerate}
\end{proof}
% section lecture_26_wednesday_1_june (end)

\end{document}

### 2011-12-12 PMH8 - Spectral Theory and PDEs/PMH8 Lecture Notes.tex
Initial Commit

% Created by Andrew Tulloch

%!TEX TS-program = xelatex
%!TEX encoding = UTF-8 Unicode

\documentclass[10pt, oneside, reqno]{amsart}
\usepackage{geometry, setspace, graphicx, enumerate}
\onehalfspacing                 
\usepackage{fontspec,xltxtra,xunicode}
\defaultfontfeatures{Mapping=tex-text}

	% AMS Theorems
	\theoremstyle{plain}% default 
	\newtheorem{thm}{Theorem}[section]
	\newtheorem{lem}[thm]{Lemma} 
	\newtheorem{prop}[thm]{Proposition} 
	\newtheorem*{cor}{Corollary}

	\numberwithin{equation}{section}

\newcommand{\res}[2]{\text{Res}(#1,#2)}
	\theoremstyle{definition} 
		\newtheorem{defn}[thm]{Definition}
		\newtheorem{conj}[thm]{Conjecture}
		\newtheorem{exmp}[thm]{Example}
		\newtheorem{exer}[thm]{Exercise}

	\theoremstyle{remark} 
		\newtheorem*{rem}{Remark} 
		\newtheorem*{note}{Note} 
		\newtheorem{case}{Case}

		\newcommand{\expc}[1]{\mathbb{E}\left[#1\right]}
		\newcommand{\var}{\text{Var}}
		\newcommand{\cov}[1]{\text{Cov}\left(#1\right)}
		\newcommand{\prob}[1]{\mathbb{P}(#1)}
		\newcommand{\given}{ \, | \,}
		\newcommand{\us}{0 \leq u \leq s}
		\newcommand{\ts}[1]{\{ #1 \}}

% \renewcommand{\phi}{\varphi}
\newcommand{\sigf}{\mathcal{F}}

\newcommand{\dzz}{\, dz}
\newcommand{\bigo}[1]{\mathcal{O}(#1)}

\newcommand{\al}{\alpha}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\K}{\mathbb{K}}

\newcommand{\Z}{\mathbb{Z}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\N}{\mathbb{N}}

\newcommand{\El}{\mathcal{L}}
\newcommand{\I}{\mathbb{I}}

\renewcommand{\P}{\mathbb{P}}

\newcommand{\F}{\mathbb{F}}
\newcommand{\Ga}{\mathbb{G}}

\newcommand{\aut}[1]{\text{Aut}{(#1)}}

\newcommand{\gener}[1]{\langle #1 \rangle}
\newcommand{\charr}[1]{\text{char}(#1)}
\newcommand{\nth}{n\textsuperscript{th}}

% \newcommand{\limsup}{\text{limsup}}

\newcommand{\tworow}[2]{\genfrac{}{}{0pt}{}{#1}{#2}}
\newcommand{\xdeg}[2]{[#1 : #2]}
\newcommand{\gal}[2]{\text{Gal}(#1/#2)}
\newcommand{\minpoly}[2]{m_{#1, #2}(x)}

\newcommand{\mapping}[5]{\begin{align*}
	#1 : \quad     #2 &\rightarrow #3 \\
			#4  &\mapsto #5
\end{align*}	
}
\newcommand{\im}{\textsc{Im\ }}
\renewcommand{\ker}{\textsc{Ker\ }}

\def\cip{\,{\buildrel p \over \rightarrow}\,} 
\def\cid{\,{\buildrel d \over \rightarrow}\,} 
\def\cas{\,{\buildrel a.s. \over \rightarrow}\,}

\def\clp{\,{\buildrel L^p \over \rightarrow}\,}

\def\eqd{\,{\buildrel d \over =}\,} 
\def\eqas{\,{\buildrel a.s. \over =}\,}
\newcommand{\sigg}{\mathcal{G}}		
\newcommand{\indic}[1]{\mathbf{1}_{\{ #1 \}} }
\newcommand{\iprod}[1]{\left\langle #1 \right\rangle}
\renewcommand{\Re}{\text{Re}}

\usepackage{hyperref}		
\title{PMH8 - Spectral Theory and PDEs}								% Document Title
\author{Andrew Tulloch}
%\date{}                                           % Activate to display a given date or no date

\begin{document}
\maketitle

% \tableofcontents
% \clearpage
\section{Preliminaries} % (fold)
\label{sec:preliminaries}

In general, we cannot solve arbitrary PDEs.  We generally seek to prove \textbf{existence} of solutions and various \textbf{properties} of these solutions.

% section preliminaries (end)
\textbf{Assessment Schedule:}
\begin{enumerate}[(i)]
	\item Assignments - 2 or 3 (40\%)
	\item Exam - (60\%) 
	% Exam on Friday Nov 4
\end{enumerate}

\textbf{References}
\begin{enumerate}[(i)]
	\item M. Protter and Weinberger -- \emph{Maximum Principle ...}
	\item M. Renardy -- \emph{Elliptic PDEs}
	\item A. Friedman -- \emph{Elliptic PDEs}
	\item F. John -- \emph{PDEs}
\end{enumerate}

\section{Introduction to Functional Analysis} % (fold)
\label{sec:introduction_to_functional_analysis_}

% section introduction_to_functional_analysis_ (end)
\begin{defn}[Quotient space]
	If $M$ is a closed subspace of a normed vector space $E$, then we define another normed space $E / M$, the \textbf{quotient space}.  Elements of $E / M$ are of the form  $\{ u + m \given m \in M \}$ where $u \in E$.

	We now define the vector space operations.  Define $(u_1 + M) + (u_2 + M) = (u_1 + u_2) + M$.  If $\lambda \in \K$, define $\lambda(u + M) = \lambda u + M$.  These operations make $E/M$ a vector space.
\end{defn}

\begin{exer}
	Show these operations are well defined.  
\end{exer}

\begin{defn}[Normed quotient space]
	Define \[
	\| u + M \| = \inf_{m \in M} \| u + m \|. 
	\]  If $u \notin M$, $\| u + M \| > 0$.  This is because if there exists $(m_n) \in M$ with $\| u + m_n \| \rightarrow 0$, then $m_n \rightarrow -u$, and so $-u \in M$, which implies $u \in M$.

	We can also show that  $\| \lambda u + M \| = |\lambda| \| u + m \|$, and \[
		\| (u_1 + u_2) + M \| \leq \| u_1 + M \| + \| u_2 + M \|.
	\]

	With this norm, $E/M$ is a normed space.
\end{defn}

\begin{exer}
	Check the triangle and scaling inequalities.
\end{exer}

\begin{lem}
	Define an operator $P$ by \mapping{P}{E}{E/M}{x}{x + M}  Then $P$ is linear and bounded.  
\end{lem}

\begin{proof}
	\begin{align*}
		\| Px \| = \| x + M \| = \inf_{m \in M} \| x + M \| \leq \| x \|
	\end{align*}  Hence $\| Px \| \leq \| x \|$ and so $P$ is bounded with $\| P \| \leq 1$.  
\end{proof}

\begin{thm}
	If $E$ is a Banach space, then so is $E / M$, where $M$ is a closed subspace of $E$. 
\end{thm}

\begin{thm}
	If $M$ is a closed subspace of a normed space $E$ and $z \in E \backslash M$, there exists $f \in E'$ such that $f(m) = 0$ for all $m \in M$, and $f(z) \neq 0$.  
\end{thm}

\begin{proof}
	$z + M$ is not zero in $E/M$, and so by the Hahn-Banach theorem, there exists $h \in (E/M)'$ such that $h(z+M) \neq 0$.  Then define $f: E \rightarrow \K$ by $f(x) = h(Px)$ where $P : E \rightarrow E/M$ is the projection operator defined previously.

	As $f$ is the composition of two continuous maps, we have that $f \in E'$.  Now, not that $f(m) = 0$ if $m \in M$, as $m + M$ is the zero coset.  If $z \in E\backslash M$, then $f(z) = h(z+M) \neq 0$ by definition.         
\end{proof}

\begin{thm}
	\label{thm:rangeclosed}
	If $T \in \El(X, Y)$ and $\im T$ is closed, then \[
		\im T = \{ y \in Y \given f(y) = 0 \text{ for all } f \in \ker T' \}
	\] 
\end{thm}

\begin{rem}{\ } 
	\begin{enumerate}[(i)]
		\item In fact, if $\im T$ is not closed, the above theorem holds with $\overline{\im T}$. 
		\item This gives a solution to the inverse problem, i.e. given $y \in Y$, does there exists $x \in X$ such that $Tx = y$.   
	\end{enumerate}
\end{rem}

\begin{defn}[Dual mapping]
	Let $T \in \El(X, Y)$. Define the dual mapping $T' \in \El(Y', X')$ with $(T'f)(x) = f(Tx)$ for all $f \in Y'$.  
\end{defn}

\begin{proof}[Proof of Theorem \ref{thm:rangeclosed}]
	Let $A = R(T)$, if $z \in A$ there exists$f \in Y'$ such that $f(y) = 0$ for all $y \in A$ and $f(z) \neq 0$.  Let $B = \{ y \in Y \given f(y) = 0 \, \forall f \in N(T') \}$.

	Hence \begin{align*}
		f(y) = 0 &\forall y \in A \\
		f(Tx) = 0 &\forall x \in X \\
		(T'f)(x) = 0 &\forall x \in X
	\end{align*} so that $T'f = 0$, and so $f \in N(T')$.  But $f(z) \neq 0$, so $z \notin B$, and so $B \subseteq A$.

	If $v \in R(T)$, then $v = Tx$.  If $f \in N(T')$, then $f(v) = f(Tx) = (T'f)(x) = 0$, and so $v \in B$.  Hence $A \subseteq B$.  \qedhere
\end{proof}

\begin{rem}
	If $H$ is a Hilbert space, and $T \in \mathcal L(H)$, then \[
		\iprod{Tx, y} = \iprod{x, T^\star y} \quad \forall x, y \in H
	\] where $T^\star$ is the adjoint.  NOte that $T^\star = J^{-1} T' J$ where $J: H \rightarrow H'$ and $T'$ is the conjugate operator.  In this case, if $R(T)$ is closed, then \[
		R(T) = \{ x \in H \given \iprod{x, y} = 0 \quad \forall y \in N(T^\star) \}.
	\]
\end{rem}

\begin{rem}When is $R(T)$ closed?

	\begin{enumerate}[(i)]
		\item If $\lambda \neq 0$ and $K \in \mathcal K(X)$, $\lambda I - K$ has closed range.
		\item If $K \in \mathcal K(X)$, $R(K)$ is closed if and only if $R(K)$ is finite dimensional.  
		\item If $N(T) = \{ 0 \}$, $X, Y$ are Banach spaces, and $T \in \mathcal L(X, Y)$, then $R(T)$ is closed if and only if there exists $c > 0$ such that $\| Tx \| \geq c \| x \|$ for all $x \in X$.  Note that if $R(T)$ is closed, it is a Banach space.  
	\end{enumerate}

\end{rem}
% subsection functional_analysis (end)

% section lecture_1 (end)

\begin{cor}[Corollary to Theorem \ref{thm:rangeclosed}]
	If $X$ and $Y$ are Banach spaces and $T \in \mathcal L(X, Y)$, then $T$ is invertible if and only if $\ker T = \{ 0 \}$, $\ker T' = \{ 0 \}$ and $\im T$ is closed.

	Note that the open mapping theorems shows that $T$ is invertible if and only if $\im T = Y$ and $\ker T = \{ 0 \}$. 
\end{cor}

\begin{proof}
	If $\im T$ is closed, then by Theorem \ref{thm:rangeclosed}, \[
		\im T = Y \iff \ker T' = \{ 0 \},
	\] as $\im T = \{ y \in Y \given f(y) = 0 \text{ for all } f \in \ker T' \}$.  
\end{proof}

In a Hilbert space $\mathcal H$, if $T \in \mathcal L(\mathcal H)$, then $T^* = J^{-1} T' J$ where $J: \mathcal H \rightarrow \mathcal H'$ is an isomorphism of Hilbert spaces.

\begin{defn}[Weak convergence]
	Let $(x_n)$ be a sequence in $X$.  We say that $x_n \rightharpoonup x$ \textbf{weakly} if $f(x_n) \rightarrow f(x)$ for all $f \in X'$.
\end{defn}

\begin{lem}
	If $x_n \rightarrow x$ in the usual sense, then $x_n \rightharpoonup x$ \textbf{weakly}. 
\end{lem}

\begin{lem}
	If $x_n \rightharpoonup x$ weakly, then $\{ x_n \}$ is bounded. Furthermore, $\| x \| \leq \liminf_{n \rightarrow \infty} \| x_n \|$.    
\end{lem}

\begin{proof}
	By Hahn-Banach, there exists $f \in X'$ such that $\| f \| = 1$ and $f(x) = \| x \|$.  So $\| x \| = f(x) = \lim_{n \rightarrow \infty} f(x_n) = \liminf_{n \rightarrow \infty} f(x_n)$.  But \[
		\| f(x_n) \| \leq \| f \| \| x_n \| \leq \| x_n \| 
	\] as $\| f \| = 1$.  So \begin{align*}
		\| x \| \leq \liminf_{n \rightarrow \infty} \| x_n \|
	\end{align*}

\end{proof}

\begin{exer}
	If $(x_n)$ is bounded, then $x_n \rightharpoonup x$ weakly if and only if $f(x_n) \rightarrow f(x)$ as $n \rightarrow \infty$ for all functions in a dense subset of $X'$.

	In fact, if $(x_n)$ is bounded, we only need prove that if $f(x_n) \rightarrow f(x)$ for a subset $M$ of $X'$, then $f(x_n) \rightarrow f(x)$ for all finite linear combinations of elements of $M$.     
\end{exer}

\begin{exmp}
	Let $1 < p < \infty$, and consider the Banach space $\ell^p$.  Then let $e_n = (\overbrace{0, 0, 0, \dots}^{p-1}, 1, \dots)$.  Then $\| e_n \|_p = 1$ and $e_n \rightharpoonup 0$ weakly in $\ell^p$ as $n \rightarrow \infty$.  Let $\left(\ell^p \right)' = \ell^{p'}$ where $\frac{1}{p} + \frac{1}{p'} = 1$.  If fact, every $f \in \left( \ell^p \right)'$ can be uniquely written as \[
		f(x) = \sum_{i=1}^\infty x_i y_i
	\] where $(y_i) \in \ell^{p'}$.  In $\ell^{p'}$, the set of finite linear combinations of the $e_n$ are dense in $\ell^{p'}$, since we can approximate $(x_n)$ by $(x_1, x_2,\dots, x_m, 0, 0 \dots)$, which is a finite linear combination of the $(e_n)$.

	Hence a \textbf{bounded} sequence in $\ell^p$, say $x^1 = (x^1_1, x^1_2, \dots), x^2 = (x^2_1, x^2_1, \dots)$ converges weakly if and only if $e_i(x^n) = x^n_i$ converges as $n \rightarrow \infty$ for each $i$.

	In particular, $e_i \rightharpoonup 0$ weakly in $\ell^p$.  
\end{exmp}

\begin{thm}
	If $x_n \rightharpoonup x$ weakly in $X$ and $T \in \El(X)$, then $Tx_n \rightharpoonup Tx$ weakly in $X$.

	Note that this is \textbf{not true} for continuous non-linear maps. 
\end{thm}
\begin{proof}
	Let $f \in X'$.  Then \begin{align*}
		f(Lx_n) = (L'f)(x_n) \rightharpoonup (L'f)(x) = f(Lx)
	\end{align*} weakly as $x_n \rightarrow x$ weakly and $(L'f)$ is a bounded linear operator.  
\end{proof}

\begin{defn}[Bidual]
	Let $X$ be a normed vector space. Then $X'$ is a Banach space.  The dual of the dual space, $(X')' = X''$ is known as the \textbf{bidual} of $X$.  
\end{defn}

There is a natural map \mapping{c}{X}{X''}{x}{\hat x} of $X$ into $X''$, defined as folllows.  Let $\hat x(f) = f(x)$ for all $f \in X'$.  Then we can see that $\hat x$ is a linear mapping, and we must show that it is a bounded map from $X'$ to $\R$.

We have \begin{align*}
	\| \hat x \| = \sup_{\| f \| \leq 1} |\hat x(f)| = \sup_{\| f \| \leq 1} |f(x)| \leq \sup_{\|f \| \leq 1} \| f \| \| x \| \leq \| x \|. 
\end{align*}  Thus $\| \hat x \| \leq \| x \|$.  (By Hahn-Banach, we can show $\| \hat x \| = \| x \|$. )

\begin{exer}
	Show that $\ker c = \{ 0 \}$.  
\end{exer}

Thus $c$ is a bounded linear map with a zero null-space.

\begin{defn}[Reflexive]
	A Banach space is refexive if this map of $X$ onto $X''$ is bijective.  
\end{defn}

\begin{exmp}{\ }
	\begin{enumerate}[(i)]
		\item Finite dimensional spaces are reflexive (as the bidual has the same dimension as the base space).
		\item $\ell^p, L^p$ are reflexive if $1 < p < \infty$, and are not reflexive otherwise.
		\item Hilbert spaces $\mathcal H$ are reflexive.
		\item $\mathcal C(\Omega)$, the set of continuous operators on a compact set in $\R^n$.  
	\end{enumerate}
\end{exmp}

\begin{thm}[Compactness property]
	A Banach space $X$ is reflexive if and only if every bounded sequence in $X$ has a subsequence that converges weakly in $X$.
\end{thm}

\begin{rem}[Closeness property]
	If $C$ is a closed and convex subset of a Banach space $X$, and $x_n$ is a sequence in $C$ with $x_n \rightharpoonup y \in X$ weakly, then $y \in C$.   
\end{rem}

\begin{proof}
	Uses the geometric version of the Hahn-Banach theorem.

	\begin{thm}[Geometric Hahn-Banach]
		If $C$ is a closed and convex subset in $X$ and $z \notin C$, there exists $f \in X'$ and $m \in \R$, such that $f(x) \leq m$ for all $x \in C$ and $f(z) > m$. 
	\end{thm}

	If $y \notin C$, there exists $f \in X'$ and $m \in \R$ such that $f(x) \leq m$ for all $x \in C$ and $f(y) > m$.  But as $f(x_n) \leq m$ and $f(x_n) \rightarrow f(y)$ (by weak convergence), we must have $f(y) \leq m$.  Thus we achieve our required result, $y \in C$.  
\end{proof}
% section lecture_4 (end)

\section{Linear Operators on Hilbert Spaces} % (fold)
\label{sec:linear_operators_on_hilbert_spaces}

% section linear_operators_on_hilbert_spaces (end)
\begin{thm}[Lax-Milgram theorem]
	\label{thm:lax_milgram}
	If $T \in \mathcal L(\mathcal H)$ and there exists $\mu > 0$ such that $\Re \iprod{Tx, x} \geq \mu \| x \|^2$ for all $x \in \mathcal H$, then $T$ is invertible.  
\end{thm}

\begin{proof}
	If suffices to prove that $\ker T = \{ 0 \}$, $\im T$ is closed, and $\ker T^\star = \{ 0 \}$, by a corollary to Theorem \ref{thm:rangeclosed}.

	By Cauchy-Swartz, we have \begin{align*}
		\mu \| x \|^2 \leq \Re\iprod{Tx, x} \leq |\iprod{Tx, x}| \leq \| Tx \| \| x \|.
	\end{align*}  If $x \neq 0$, then $\mu \| x \| \leq \| Tx \|$, so $\ker T = \{ 0 \}$.

	Secondly, $\| Tx \| \geq \mu \| x \|$ for $\mu > 0$ implies $\im T$ is closed.

	\begin{exer}
		Prove this proposition.
	\end{exer}

	Then finally, we have \begin{align*}
		\Re \iprod{Tx, x} = \Re \iprod{x, T^\star x} \leq | \iprod{x, T^\star x} | \leq \| x \| \|T^\star x \|
	\end{align*} by Cauchy-Swartz.  So \[
		\mu \| x \|^2 \leq \| x \| \| T^\star x \|
	\] and so $\mu \| x \| \leq \| T^\star x \|$ and so $\ker T^\star = \{ 0 \}$.
\end{proof}

\begin{defn}[Coercive]
$T$ is coercive if there exists $\mu > 0$ such that $\Re \iprod{Tx, x} \geq \mu \| x \|^2$.  
\end{defn}

\begin{defn}[Spectral radius]
	Let $X$ be a complex Banach space.  If $T \in \mathcal L(X)$, then we can define the spectral radius $r(T)$ by the formula \[
		r(T) = \limsup_{n \rightarrow \infty} \| T^n \|^{1/n}.
	\]
\end{defn}

\begin{thm} 
	We have \[
		r(T) = \sup \{ |\lambda | \given \lambda \in \sigma(T) \}.
	\]

	Note that $r(T) \leq \| T \|$.  
\end{thm}

\begin{thm} 
	If $T \in \mathcal L(\mathcal H)$ and $T$ is a self-adjoint operator then $r(T) = \| T \|$.   
\end{thm}

\begin{proof}
	We have \[
		\| T \|^2 = \| T^\star T \| = \| T^2 \|,
	\] since for any linear operator $T$, we have \[
		\| T \|^2 = \| T^\star T \|. 
	\]

	Then by induction, we have \[
		r(T) = \limsup_{n \rightarrow \infty} \| T^{2^n} \|^{1/2^n} = \| T \|
	\]
\end{proof}

\begin{thm}[Raleigh-Rety algorithm]
	\label{thm:raleigh_rety}
	If $T \in \mathcal L(\mathcal H)$ is self-adjoint, then \begin{align*}
		\sup \sigma(T) &= \sup \{ \iprod{Tx, x} \given \| x \| = 1 \} \\
		\inf \sigma(T) &= \inf \{ \iprod{Tx, x} \given \| x \| = 1 \}
	\end{align*}
\end{thm}

\begin{proof}
	If suffices to prove the first statement (and then apply to $-T$).  We first show $\sup \sigma(T) \leq \sup \{ \iprod{Tx, x} \given \| x \| = 1 \} \equiv \mu$.

	If $\lambda > \mu$, then \[
		\lambda \| x \|^2 - \iprod{Tx, x} \geq \lambda - \mu > 0
	\] if $\| x \| = 1$.  Hence \begin{align*}
		\lambda - \mu 	&\leq \iprod{(\lambda I - T)x, x} \quad \| x \| = 1 \\
						&\leq \| (\lambda I - T)x \| \| x \| \\
					\Rightarrow \| (\lambda I - T)\| x \| \geq (\lambda - \mu) \| x \|
	\end{align*} and hence $\ker \lambda I - T = \{ 0 \}$, and as $\im \lambda I - T$ is closed by Exercise \ref{exer:closed_range_condition}, we have that $\lambda I - T$ is invertible.  Thus $\sup \sigma(T) \geq \sup \{ \iprod{Tx, x} \given \| x \| = 1 \}$.

	Consequently, it suffices to assume $\sigma(T)$ is non-negative (replace $T$ with $T + rI$).  Then if $\mu \in \sigma(T_1)$ with $T_1$ self-adjoint, then there exists a sequence $x_n$ with $\| x_n \| = 1$ such that \[
		\| T_1 x_n - \mu x_n \| \rightarrow 0 
	\] as $n \rightarrow \infty$.  Existence of such a sequence is proven as if $\| T_1 x - \mu x \| \geq \alpha \| x \|$, then $\mu \notin \sigma(T_1)$.

	Thus \begin{align*}
		\iprod{T_1 x_n, x_n} &\rightarrow \mu \\
		\iprod{T_1 x_n, x_n} &= \underbrace{\iprod{ (T_1 - \mu I) x_n, x_n}}_{\rightarrow 0} + \underbrace{\mu \iprod{x_n, x_n}}_{\mu}.
	\end{align*}

	Thus \[
		\sup \{ \iprod{Tx, x} \given \| x \| = 1\} \geq \sup \sigma(T).
	\]
\end{proof}

\begin{exer}
	\label{exer:closed_range_condition}
	If $\| Tx \| \geq m \| x \|$ for all $x$, then $\im T$ is closed.
\end{exer}

\section{Generalised Derivatives} % (fold)
\label{sec:generalised_derivatives}

% section generalised_derivatives (end)
\begin{defn}[$L^1_{loc}(\Omega)$]
	Let $\Omega \subset \R^n$ be open.  Then $u \in L^1_{loc}(\Omega)$ if $u$ is measurable and $u|_K \in L^1(K)$ for every compact $K \subseteq \Omega$. 
\end{defn}

\begin{defn}[Generalised derivative]
	We say $u \in L^1_{loc}(\Omega)$ has a (weak) generalised $j$-th partial derivative if there exists $g \in L^1_{loc}(\Omega)$ such that \begin{equation}
	\label{eq:generalised_derivative}
		\int_\Omega u \frac{\partial \phi}{\partial x_j} = - \int_\Omega g \phi
	\end{equation}
	for all $\phi \in C_c^\infty(\Omega)$.

	Note that $g$ is defined only up to sets of measure zero.
\end{defn}

\begin{note}
	The motivation comes from the integration by parts formula, where if $u$ is $C^1(\Omega)$, then \[
		\int_\Omega u \frac{\partial \phi}{\partial x_j} = - \int_\Omega \frac{\partial u}{\partial x_j} \phi
	\] for all $\phi \in C^1_c(\Omega)$ by integration by parts. Thus we can write $g = \frac{\partial u}{\partial x_j}$.   
\end{note}

\begin{lem}
	\label{lem:uniqueness_of_generalised_derivative}
	The function $g$, if it exists, is unique (up to sets of measure zero).  
\end{lem}

\begin{proof}
	If $g_1, g_2$ both satisfy \eqref{eq:generalised_derivative}, then \[
		-\int_\Omega u \frac{\partial \phi}{\partial x_j} = \int_\Omega g_1 \phi = \int_\Omega g_2 \phi
	\] for all $\phi \in C_c^\infty(\Omega)$.  Thus \[
		\int_\Omega (g_1 - g_2) \phi = 0
		\tag{$\star$}
	\] for all $\phi \in C_c^\infty(\Omega)$.

	Suppose $B$ is a ball with $\overline B \subseteq \Omega$.  Then \[
		(g_1 - g_2) |_B \in L^1(B).  
	\]   Since ($\star$) holds for all $\phi \in C_c^\infty(B)$, consider the measurable function \[
		\text{sgn}(g_1 - g_2) = \begin{cases}
			1 & (g_1 - g_2)(x) \geq 0 \\
			-1 & (g_1 - g_2)(x) < 0.
		\end{cases}  
	\]  
	We assume that there exists $(\phi_n) \in C_c^\infty(B)$ such that $\phi_n$ are uniformly bounded and $\phi_n(x) \rightarrow \text{sgn}(g_1 - g_2)$ almost everywhere.  This can be justified by Young's inequality, where if \[
		f_n(x) = \int_B \psi_n(x-y) f(y) \, dy
	\] then $\| f_n \|_\infty \leq \| \psi_n \| \ | f \|_\infty$, so our approximating function $f_n$ are uniformly bounded.

	Then \[
		0 = \int_\Omega (g_1 - g_2) \phi_n \rightarrow \int_B (g_1 - g_2) \text{sgn}(g_1 - g_2) = \int_B |g_1 - g_2 |
	\] as $n \rightarrow \infty$ by the dominate convergence theorem.

	Thus $g_1 - g_2 = 0$ almost everywhere on $B$.  By the Lindeloff property (Lemma \ref{lem:lindeloff}), $\Omega$ is a countable union of balls, and so we can extend this result to the result, \[
		g_1 - g_2 = 0
	\] almost everywhere on $\phi$.  
\end{proof}

\begin{lem}[Lindeloff property]
	\label{lem:lindeloff}
	A separable metric space, such as $R^n$, any open set is a countable union of open balls.
\end{lem}

\begin{rem}{\ }
	\begin{enumerate}[(i)]
		\item If $g$ is the generalised $j$-th partial derivative of $u$ on $\Omega$ and $\Omega_1 \subset \Omega$ is open, then $g |_{\Omega_1}$ is the $j$-th generalised partial derivative of $u|_{\Omega_1}$.
		\item Assume $A \subseteq \Omega$, $u$ has a generalised $j$-th partial derivative on $\Omega$, $A$ is open, and $u$ is $C^1$ on $A$.

		Then the generalised $j$-the partial derivative of $u$ is equal to the classical partial derivative almost everywhere on $A$.
	\end{enumerate}
\end{rem}

\begin{exmp}
	Consider the function \[
		u(x, y) = \begin{cases}
			1 & y \geq 0	\\
			0 & y < 0		
		\end{cases}
	\]  If the generalised derivative $\frac{\partial u}{\partial x}$ exists, it must be zero when $y > 0$ and when $y < 0$.

	It turns out that $\frac{\partial u}{\partial x}$ exists but $\frac{\partial u}{\partial y}$ does not.
\end{exmp}

\begin{exmp}
	$f: \R \rightarrow \R$ define by $f(x) = |x|$ has a generalised derivative $g$ defined by \[
		g(x) = \begin{cases}
			1 & x > 0 \\
			-1 & x > 0
		\end{cases}
	\]  Note that $f$ is $C^1$ if $x \neq 0$ so if the generalised derivative exists it must be equal to $g$. 
	\end{exmp}

\begin{exmp}
	If $B_1$ is the open unit ball in $\R^2$ and \[
		f(x) = \begin{cases}
			\ln(x^2 + y^2) & (x, y) \neq (0, 0) 
		\end{cases}
	\] - thus $f(x) = 2 \ln r$ in polar coordinates.

	Then \begin{align*}
		\frac{\partial u}{\partial x} = \frac{2x}{x^2 + y^2}  \\
		\frac{\partial u}{\partial y} = \frac{2y}{x^2 + y^2} 
	\end{align*} are the generalised partial derivatives on $\R^2$.
\end{exmp}

\begin{defn}[Generalised derivative]
	We say that $u \in L^1_{loc}(\Omega)$ has a generalised derivative on $\Omega$ if all the generalised partial derivatives $\frac{\partial u}{\partial x_j}$ exist for $1 \leq j \leq n$ (where $\Omega$ is an open set in $\R^n$).
\end{defn}

\begin{rem}{\ }
	\begin{enumerate}[(i)]
		\item If $u_1$ and $u_2$ have generalised derivatives on $\Omega$ and $C_1, C_2$ are constant, then $C_1u_1 + C_2 u_2$ has a generalised derivative on $\Omega,$ given by the appropriate linear combination.  
		\item If $u$ has a generalised derivative on $\Omega$ and $\Psi \in C^\infty(\Omega)$, then  $u\Psi$ has a generalised derivative on $\Omega$ and \[
			\frac{\partial}{\partial x_j} (u \Psi) = \frac{\partial u}{\partial x_j} \Psi + u \frac{\partial \Psi}{\partial x_j}
		\]
	\end{enumerate}
\end{rem}

\begin{lem}
	\label{lem:convergence_of_gen_deriv}
	If $u_k$ has a generalised derivative on $\Omega$ and $u_k \rightarrow u$ in $L^1_{loc}(\Omega)$ as $k \rightarrow \infty$ and if $\frac{\partial u_k}{\partial x_l} \rightarrow g_l$ in $L^1_{loc}(\Omega)$ then $u$ has a generalised derivative on $\Omega$ and \[
		\frac{\partial u}{\partial x_l} = g_l.
	\]
\end{lem}

\begin{proof}
	\[
		\int_\Omega u_k \frac{\partial \phi}{\partial x_j} = -\int_\Omega \frac{\partial u}{\partial x_j} \phi
		\tag{$\star$}
	\] if $\phi \in C_c^\infty(\Omega)$.  Fix $\phi$ and choose $K$ compact so the support of $\phi$ is contained in $K$.  Then $u_k {\frac{\partial \phi}{\partial x_j}} \rightarrow u \frac{\partial \phi}{\partial x_j}$ in $L^1$ on $K$.

	Then letting $k \rightarrow \infty$ in $(\star)$ we obtain \[
		\int_\Omega u \frac{\partial \phi}{\partial x_j} = - \int_\Omega g_j \phi.
	\]
\end{proof}

\begin{rem}
	If $g_j \in L^p(\Omega)$ and $g_j \rightarrow g$ in $L^p(\Omega), (1 \leq p \leq \infty)$, then $g_j \rightarrow g$ in $L^1_{loc}(\Omega)$.  
\end{rem}

\begin{proof}
	If $K$ is compact, then \[
		\int_K (g_j - g) \leq \| g_j - g\|_{p, K}^\frac{1}{p} \| 1 \|^{1/p'}_{p', K}
	\] by H\"older's inequality.
\end{proof}

\section{Sobolev Spaces} % (fold)
\label{sec:sobolev_spaces}

% section sobolev_spaces (end)
\begin{defn}[Sobolev spaces]
	If $1 \leq p \leq \infty$ and $\Omega$ is open in $\R^n$, then the space \[
		W^{1, p}(\Omega) = \{ u \in L^p(\Omega) \given \underbrace{\frac{\partial u}{\partial x_i} \in L^p(\Omega)}_{\substack{\text{generalised} \\ \text{derivatives}}} \text{ for } 1 \leq i \leq n \}
	\] 
	equipped with the norm \[
		\| u \|_{1, p} = \| u \|_p + \sum_{i=1}^n \| \frac{\partial u}{\partial x_i} \|_p
	\] is a Banach space.  We call $W^{1, p}$ a Sobolev space.

	It is a linear space by linearity of the generalised derivatives.  Similarly, the triangle inequality holds as all components of the norm $\| \cdot \|_{1, p}$ satisfy the triangle inequality.  It can be shown that $W^{1, p} \subseteq L^p(\Omega)^{N+1}$ and $W^{1, p}$ is a closed subspace, which shows that $W^{1, p}$ is Banach, being the closed subspace of a Banach space.
\end{defn}

\begin{prop}
	$W^{1, p}$ is a Banach space.  In fact, $W^{1,p}$ is a closed subspace of $L^p(\Omega)^{n+1}$.
\end{prop}

\newcommand{\parder}[2]{\frac{\partial #1}{\partial #2}}
\begin{proof}
	Consider the the map \begin{align*}
		(u_j, \parder{u_j}{x_1}, \dots, \parder{u_j}{x_n}) \rightarrow (w_0, w_1, \dots, w_n)
	\end{align*}  If $u_j \rightarrow w_0 \in L^p(\Omega)$, then $\parder{u_j}{x_1} \rightarrow w_1$ in $L^p(\Omega)$ which implies that $\parder{u_j}{x_1} \rightarrow w_1$ in $L^1_{loc}(\Omega)$.

	By Lemma \ref{lem:convergence_of_gen_deriv}, $\parder{w_0}{x_1}$ exists on $\Omega$ and equals $w^1$.  Similarly, $\parder{w_0}{x_l}$ exists and equals $w_l$.  Then since $w_0 \in W^{1, p}(\Omega)$ and $\parder{w_0}{x_l} = w_l$ the closure property holds.  Hence we have a Banach space.
\end{proof}

\begin{note}
	Recall that all norms on a finite dimensional vector space are equivalent.  For example, \[
		\left( \| u \|_p^p + \sum_{j=1}^n \| \parder{u}{x_j}\|_p^p)^{1/p} \right)
	\] and \[
		\| u \|_p + \left( \sum_{j=1}^n \| \parder{u}{x_j} \|_p^p \right)^{1/p}
	\] are equivalent.
\end{note}

\begin{defn}[Higher Sobolev spaces]
	We have \[
	W^{2, p}(\Omega) = \{ u \in L^p(\Omega) \given \frac{\partial u}{\partial x_i}, \parder{u}{x_i x_j} \in L^p(\Omega) \text{ for } 1 \leq i \leq n, 1 \leq j \leq n \}
	\]
\end{defn}

\begin{defn}
	$\dot W^{1, p}(\Omega)$ is the closure of $C^\infty_c(\Omega)$ in $W^{1, p}(\Omega)$ in the norm $\| \cdot \|_{1, p}$.  	
	In general, $\dot W^{1, p}(\Omega) \subseteq W^{1, p}(\Omega)$.
\end{defn}

\begin{prop}
	$\dot W^{1, p}(\R^n) = W^{1, p}(\R^n)$.  
\end{prop}

\begin{prop}
	$\dot W^{1, 2}(\Omega), W^{1, 2}(\Omega)$ are Hilbert spaces under the inner product $\iprod{\cdot, \cdot}$ defined by \[
		\iprod{u, v} = (u, v) + \sum_{i=1}^n \left(\parder{u}{x_i}, \parder{v}{x_i}\right)
	\] where $(u, v) = \int_\Omega u(x) \overline{v(x)} \, dx$.
\end{prop}

\section{Convolutions and Approximations} % (fold)
\label{sec:convolutions_and_approximations}
Recall that there exists $\phi  \in C^\infty_c(\R^n)$ such that $\phi(x) > 0$ if $\| x \| < 1$ and $\phi(x) = 0$ if $\| x \| \geq 1$.  We can assume that $\int_{\R^n} \phi = 1$.

If $f \in L^p_{loc}(\R^n)$ and $1 \leq p < \infty$, we define $T_\epsilon f$ by \[
	(T_\epsilon f)(x) = \epsilon^{-N} \int \phi \left( \frac{x-y}{\epsilon} \right) f(y) \, dy = \phi_\epsilon \star f
\] where $\phi_\epsilon = \epsilon^{-N} \phi\left( \frac{x}{\epsilon}\right)$.

\begin{prop}
	\label{prop:convergence_of_covolution}
	If $f \in L^p(\R^n)$ where $1 \leq p < \infty$, then \[
		T_\epsilon f \rightarrow f
	\] in $L^p(\R^n)$ as $\epsilon \rightarrow 0$.  
\end{prop}

\begin{lem}
	If $f$ has support in a compact set $K$, then $T_e f$ has support in $\{ x \in \R^n \given d(x, K) \leq \epsilon \}$. 
\end{lem}

\begin{lem}
	By Proposition \ref{prop:convergence_of_covolution}, if $f \in L^p(\R^n)$, there exists $\epsilon_l \rightarrow 0$ such that $T_{\epsilon_l} f \rightarrow f$ almost everywhere as $l \rightarrow \infty$.
\end{lem}

\begin{lem}
	 \[
		\| T_\epsilon f \|_{\infty} \leq \| f \|_\infty
	\] if $f \in L^\infty(\R^n)$.
\end{lem}

\begin{proof}
	If $-1 \leq f \leq 1$ on $\R^n$, then as \[
		T_\epsilon(-1) \leq T_\epsilon f \leq T_\epsilon 1
	\] that is, \[
		-1 \leq T_\epsilon f(x) \leq 1 \quad \forall x
	\] then since $T_\epsilon f$ is linear we have \[
		\| T_\epsilon f \|_{\infty} \leq \| f \|_\infty
	\] if $f \in L^\infty(\R^n)$.
\end{proof}

\begin{prop}
	$C_c^\infty(\R^n)$ is dense in $W^{1, 2}(\R^n)$ that is, if $f \in W^{1, 2}(\R^n)$, there exists $(f_n) \in C^\infty_c(\R^n)$ such that $\| f - f_n \|_{1, 2} \rightarrow 0$ as $n \rightarrow \infty$.

	Note that this is non-trivial as if $\Omega$ is bounded the corresponding result is false.
\end{prop}

\begin{proof}
	Let $f \in W^{1, 2}(\R^n)$ and $\delta > 0$.  By a previous exercise, there exists $\tilde f \in W^{1, 2}(\R^n)$ of compact support such that \[
		\| f - \tilde f \| \leq \frac{\delta}{2}.
	\]  Hence it suffices to find $f_n \in C^\infty_c(\R^n)$ such that $\| f_n - \tilde f\|_{1, 2} \rightarrow 0$ as $n \rightarrow \infty$ (as this would imply $\| f_n - f \| \leq \delta$ for large enough $n$).

	We prove that $T_\epsilon \tilde f \in W^{1, 2}(\R^n)$ and $T_\epsilon \tilde f \rightarrow \tilde f$ in $W^{1, 2}(\R^n)$ as $\epsilon \rightarrow 0$.  Recall that $T_\epsilon \tilde f \in C^{\infty}(\R^n)$.  Suppose that $T_\epsilon \tilde f \subseteq B(\epsilon)\{ \text{supp}(\tilde f) \} = \{ x \in \R^n \given d(x, \text{supp}(\tilde f)) \leq \epsilon \}$.  Recall that $T_\epsilon \tilde f \rightarrow \tilde f$ in $L^2(\R^n)$ as $\epsilon \rightarrow 0$ from \textsc{math 3969}.

	We thus need to prove \begin{equation}
		\parder{}{x_l} T_\epsilon \tilde f = T_\epsilon \underbrace{\left( \parder{\tilde f}{x_l} \right)}_{\substack{\text{generalised} \\ \text{derivative}}}
	\end{equation}
  If we prove that \[
		\parder{}{x_l}\left(T_\epsilon \tilde f \right) = T_\epsilon \left(\parder{\tilde f}{x_l} \right) \rightarrow \parder{\tilde f}{x_l}
	\] in $L^2(\R^n)$.

	We have \begin{align*}
		\parder{}{x_l} T_\epsilon \tilde f(x) &= \parder{}{x_l} \left( \epsilon^{-n} \int \phi\left(\frac{x-y}{\epsilon} \right) \tilde f(y) \, dy \right) \\
		&= \epsilon^{-n} \int \parder{}{x_l} \phi\left(\frac{x-y}{\epsilon} \right) \tilde f(y) \, dy \\
		&= e^{-n} \int -\parder{}{y_l} \phi\left( \frac{x-y}{\epsilon} \right) \tilde f(y) \, dy 
	\end{align*} where we use the fact that \[
		\parder{}{x_l}g(x-y) = - \parder{}{y_l} g(x-y).
	\]  Continuing, we obtain \begin{align*}
		\parder{}{x_l} T_\epsilon \tilde f(x)	&= -\epsilon^{-n} \int \parder{}{y_l} \left( \phi\left(\frac{x-y}{\epsilon} \right) \right) \tilde f(y) \, dy \\
		\tag{$\star$}
		 &= \epsilon^{-n} \int \phi \left(\frac{x-y}{\epsilon} \right) \parder{}{y_l} \tilde f(y) \, dy \\
		&= T_\epsilon \left( \parder{\tilde f}{ y_l}\right)
	\end{align*} as $\phi\left(\frac{x-y}{\epsilon} \right)$ is a smooth function of compact support.  Thus the generalised derivative exists as $\tilde f \in W^{1, 2}(\R^n)$, and so the manipulation in $(\star)$ is justified.

\end{proof}
% section convolutions_and_approximations (end)

\section{Fourier Transforms and Weak Derivatives} % (fold)
\label{sec:fourier_transforms_and_weak_derivatives}

\begin{defn}[Fourier transform]
	If $f \in L^1(\R^n), \lambda \in \R^n$, then the Fourier transform $\hat f$ of $f$ is defined by \begin{equation}
		\hat f(\lambda) = \int_{\R^n} f(t) e^{i \lambda t} \, dt.
	\end{equation}
\end{defn}

\begin{thm}
	The map $f \mapsto \hat f$ is a bijection on $L^2(\R^n)$. 
\end{thm}

\begin{thm}[Parseval's theorem]
	If $f \in L^2(\R^n)$, then $(2 \pi)^n \| f \|_2^2 = \| \hat f \|_2^2$.  This can be generalised slightly to if $f, g \in L^2(\R^n)$, then \[
		(2 \pi)^n \iprod{f, g} = \iprod{\hat f, \hat g} = \int_{\R^n} \hat f(x) \overline{\hat g(x)} \, dx
	\]
\end{thm}

\begin{thm}
	\label{thm:his_thm_4.1}
	If $f \in L^2(\R^n)$, the following are equivalent:
	\begin{enumerate}[(i)]
		\item $f \in W^{1, 2}(\R^n)$,
		\item $-i \lambda_j \hat f(\lambda) \in L^2(\R^n)$ for $1 \leq j \leq n$, 
		\item $1 + |\lambda| \hat f(\lambda) \in L^2(\R^n)$.  
	\end{enumerate}

	If any of these hold, the generalised derivative $\parder{f}{x_j}$ exists and $\hat{ \parder{f}{x_j}}(\lambda) = -i \lambda_j \hat f (\lambda)$ for $1 \leq j \leq n$.
\end{thm}

\begin{proof}
	$(ii) \iff (iii)$ $|\lambda_j \hat f(\lambda) | \leq |\lambda | | \hat f(\lambda)|$ and hence (ii) $\iff$ (iii).

	$(i) \Rightarrow (ii)$  The only thing left to prove is that \[
		\parder{\hat f}{x_j}(\lambda) = -i \lambda_j \hat f(\lambda) 
	\] for $f \in W^{1, 2}(\R^n)$.  We have \begin{align*}
		f \in W^{1, 2}(\R^n) &\Rightarrow \parder{f}{x_j} \in L^2(\R^n) \\
			&\Rightarrow \hat{\parder{f}{x_j}}(\lambda) \in L^2(\R^n)
	\end{align*} so to prove the previous result we choose $f_n \in C^\infty_c(\R^n)$ such that $\| f_n - f\|_{1, 2} \rightarrow 0$ as $n \rightarrow \infty$.  Since \[
		\hat{\parder{f_n}{x_j}}(\lambda) = -i \lambda_j \hat f_n (\lambda),
	\] and $f_n \rightarrow f$ in $W^{1, 2}(\Omega)$, we have \begin{align*}
		&\Rightarrow f_n \rightarrow f \text{ in } L^2(\R^n) \\
		&\Rightarrow \hat f_n \rightarrow \hat f \text{ in } L^2(\R^n) \\
		&\Rightarrow \hat f_n(\lambda) \rightarrow \hat f(\lambda) \text{ a.e. (taking subsequences)} \\
		&\Rightarrow \parder{f_n}{x_j} \rightarrow \parder{f}{x_j} \text{ in } L^2(\R^n) \\
		&\Rightarrow \hat{\parder{f_n}{x_j}} \rightarrow \hat{\parder{u}{x_j}} \text{ in} L^2(\R^n) \\
		&\Rightarrow -i \lambda_j \hat f_n(\lambda) \rightarrow \hat{\parder{f}{x_j}}(\lambda) \text{ in } L^2(\R^n) \\
		&\Rightarrow \hat{\parder{f}{x_j}}(\lambda) = -i \lambda_j \hat f(\lambda) \text{ a.e.}
	\end{align*}

	$(ii) \Rightarrow (i)$ As $-i \lambda_j \hat f(\lambda) \in L^2(\R^n)$, and so there exists $g_j \in L^2(\R^n)$ such that \[
		\hat g_j = -i \lambda_j \hat f(\lambda).
	\]  Thus we have

	\begin{align*}
		(2\pi)^n \left(f, \parder{\phi}{x_j} \right) &=  \left(\hat f, \hat{ \parder{\phi}{x_j}} \right) \quad \phi \in C^\infty_c(\R^n)\\
		&= \int_{\R^n} \hat f(\lambda) \overline{-i \lambda_j \hat \phi(\lambda)} \\
		&= \int_{\R^n} \hat f(\lambda) i \lambda_j \overline{\hat \phi(\lambda)} \\
		&= \int_{\R^n} i \lambda_j \hat u(\lambda) \overline{\hat \phi(\lambda)} \\
		&= \int_{\R^n} \hat g_j \overline{\hat \phi(\lambda)} \\
		&= -(2\pi)^n \left(g_j, \phi \right) \\
		&\Rightarrow \int_{\R^n} f \parder{\phi}{x_j} = -\int g_j \phi
	\end{align*} and so $g_j$ is the $j$-th generalised derivative of $f$ and $g_j \in L^2(\R^n)$, thus $f \in W^{1, 2}(\R^n)$   
\end{proof}

\begin{rem}
	As a consequence, \[
		u \in W^{2, 2}(\R^n) \iff \left(1 + |\lambda|^2 \right) \hat u(\lambda) \in L^2(\R^n)
	\] and a similar result can be obtained for $W^{k, 2}(\R^n)$.  This follows from the fact that \[
		C_2 \leq \frac{1 + |\lambda|^2}{(1 + |\lambda|)^2} \leq C_1
	\] on $\R^n$ where $C_1, C_2 > 0$.
\end{rem}

\begin{exmp}
	Consider the PDE \begin{equation}
		\label{eq:poisson_pde}
		-\Delta u + u = f
	\end{equation}
	on $\R^n$, where $f \in L^2(\R)$ and we look for $u \in W^{2, 2}(\R^n)$.  Taking Fourier transformations, we have \begin{align*}
		\hat{\frac{\partial^2 u}{\partial x_j \partial x_k}} &= (-i \lambda_k)(-i\lambda_j) \hat u(\lambda) \\
		&= - \lambda_k \lambda_j \hat u(\lambda) \\
		- \left(-\sum_{k=1}^n \lambda_k^2 \right) \hat u(\lambda) + \hat u(\lambda) &= \hat f(\lambda) \\
		\left(1 + |\lambda|^2 \right) \hat u(\lambda) &= \hat f(\lambda). 
	\end{align*}  So \[
		\hat u(\lambda) = \frac{\hat f(\lambda)}{1 + |\lambda|^2}
	\] and $u \in W^{2, 2}(\R^n)$ (since $\left(1 + |\lambda|^2\right) \hat u (\lambda) = \hat f(\lambda) \in L^2(\R^n)$.  This is the unique solution in $W^{2, 2}$.
\end{exmp}

\begin{exmp}
	Consider a slightly modified version of \eqref{eq:poisson_pde} \[
		-\Delta u = f. 
	\] we obtain $\hat u (\lambda) = \frac{\hat f(\lambda)}{|\lambda|^2}$ and this is not well defined for $\lambda$ near zero.
\end{exmp}

\newcommand{\grad}{\nabla}
\begin{exmp}
	Considering the equation \eqref{eq:poisson_pde}, we take $u \in W^{1, 2}(\R^n)$ such that \[
		\int \left( \grad u \grad \phi + u \phi \right) = \int f \phi \quad \forall \phi \in C_c^\infty(\R^n).
		\tag{$\star$}
	\] If this holds, it follows that $\phi \in W^{1, 2}(\R^n)$.  By Parseval's theorem, we have \[
		\int \sum_{j} -i \lambda_j \hat u(\lambda) \overline{-i \lambda_j \hat \phi(\lambda)} + \int \hat u(\lambda) \overline{\hat \phi(\lambda)} = \int \hat f(\lambda) \overline{ \hat\phi(\lambda)}
	\] and this is solved by \[
		\hat u(\lambda) = \frac{\hat f(\lambda)}{1 + |\lambda|^2}
	\]

	Note that $(\star)$ has at most one solution in $W^{1, 2}(\R^n)$.  If $u_1, u_2$ are solutions then we have \begin{align*}
		\int \grad u_1 \grad \phi  + u_1 \phi &= \int f \phi \quad \forall \phi \in W^{1, 2}(\R^n). \\
		\int \grad u_2 \grad \phi + u_2 \phi &= \int f \phi.
	\end{align*}  Subtracting these obtains \[
		\int \grad{u_1 - u_2} \grad \phi + (u_1 - u_2 \phi) = 0.
	\]  Letting $\phi = u_1 - u_2 \in W^{1, 2}(\R^n)$, we have \[
		\int \underbrace{\grad(u_1 - u_2) \grad{u_1 - u_2}}_{\geq 0} + \underbrace{(u_1 - u_2)^2}_{\geq 0} = 0.
	\]
\end{exmp}
% section fourier_transforms_and_weak_derivatives (end)

\newcommand{\poincare}{Poincar\'e\ }
\section{\poincare Inequality and Applications} % (fold)
\label{sec:poincar'e_inequality_and_applications}

\begin{lem}
	\label{lem:poincare_lemma}
	If $v \in C^1(\R)$, $a \neq b$ and $v(a) = v(b) = 0$, then \begin{equation}
		\int_a^b v^2(t) \, dt \leq (b-a)^2 \int_a^b \left(v'(t) \right)^2 \, dt.
	\end{equation}	
\end{lem}

\begin{proof}
	We have $v(x) = v(a) + \int_a^x v'(t) \, dt = \int_a^x v'(t) \, dt$ for $a < x < b$.  So \begin{align*}
		|v(x)| 	&\leq \left| \int_a^x v'(t) \, dt \right| \\
				&\leq \int_a^b |v'(t)| \, dt \\
				&\leq (b-a)^{1/2} \left( \int_a^b \left(v'(t) \right)^2 \, dt \right)^{1/2}.  
	\end{align*}  Squaring and integrating from $a$ to $b$, we obtain our result, \[
		\int_a^b v^2(t) \, dt \leq (b-a)^2 \int_a^b \left(v'(t) \right)^2 \, dt. \qedhere
	\]
\end{proof}

\begin{thm}[\poincare inequality]
	\label{thm:poincare_inequality}
	If $\Omega$ is a domain in $\R^n$ with $\Omega \subseteq C$ where $C$ is a cube of side $d$, then 
	\begin{equation}
		\label{eq:poincare_inequality}
		\| w \|_{2, \Omega} \leq d \| \grad w \|_{2, \Omega}
	\end{equation}
	for $w \in \dot W^{1, 2}(\Omega)$.
\end{thm}

\begin{rem}
	Recall that $\cdot W^{1, 2}(\Omega) \subset W^{1, 2}(\Omega)$ if $\Omega$ is a bounded domain.  Note that the identity function $1 \in W^{1, 2}(\Omega)$ does not satisfy this inequality.
\end{rem}

\begin{proof}
	First assume $u \in C^\infty_c(\Omega)$.  We can extend $u$ to $\tilde u$ in $C^\infty_c(C)$ by defining $\tilde u(x) = 0$ if $x \in C \, \backslash \Omega$.  Assume $C = [a, b]^n$.  Then (identifying $u$ with $\tilde u$), \begin{align*}
		\int_a^b  u(x_1, \dots, x_n)^2 \, dx_1 \leq (b-a)^2 \int_a^b \left(\parder{u}{x_1} \right)^2 \, dx_1
	\end{align*} by Lemma \ref{lem:poincare_lemma}.  Integrating over the entire $n$-cube, we then have \begin{align*}
		\int_a^b \dots \int_a^b u(x_1, \dots, x_n)^2 \, dx_1 \dots dx_n &\leq (b-a)^2 \int_a^b \dots \int_a^b \left(\parder{u}{x_1} \right)^2 \, dx_1 \dots dx_n \\
		&\leq (b-a)^2 \int_C |\grad u|^2 
	\end{align*} as $|\grad u |^2 = \sum_{i=1}^n \left(\parder{u}{x_1} \right)^2$.  As $u$ is zero on $C \, \backslash \, \Omega$ we have the result \[
	\tag{$\star$}
		\int_\Omega u^2 \leq d^2 \int_\Omega |\grad u|^2
	\] for $u \in C^\infty_c(\Omega)$.

	Now, if $u \in \dot W^{1, 2}(\Omega)$, there exists $u_n \in C^\infty_c(\Omega)$ such that $\| u_n - u \|_{1, 2} \rightarrow 0$ as $n \rightarrow \infty$.  For each $n$, we then have \[
		\int_\Omega u_n^2 \leq d^2 \int_\Omega |\grad u_n |^2
	\] by $(\star)$.  Taking the limit, we obtain our required result, \[
		\int_\Omega u^2 \leq d^2 \int_\Omega |\grad u|^2. \qedhere
	\] 
\end{proof}

Intuitively, $\dot W^{1, 2}(\Omega)$ is the set of functions in $W^{1, 2}(\Omega)$ vanishing on $\partial \Omega$.  If $\Omega$ is a domain with a smooth boundary, then it can be proven there is a map $T$, known as the trace map, \[
	T : W^{1, 2}(\Omega) \rightarrow L^2(\partial \Omega)
\] such that $\dot W^{1, 2}(\Omega) = \ker T$.   The key difficulty in the proof is showing the inequality \[
	\int_{\partial \Omega} \left(v|_{\partial \Omega}\right)^2 \leq K \| v \|_{1, 2}^2
\] if $v \in W^{1, 2}(\Omega)$.  By the \poincare inequality, we can use $\| \grad u \|_2$ as a norm on $\dot W^{1, 2}(\Omega)$ if $\Omega$ is abounded domain.  This is equivalent to $\|u \|_2 + \| \grad u \|_2$.

Note that this norm is induced by the scalar product (assuming real $u, v$) \[
	\iprod{u, v} = \int_\Omega \grad u \grad v =  \sum_{j=1}^n \int_\Omega \parder{u}{x_j} \parder{v}{x_j}
\]

\begin{prop}
	Consider the equation 
	\begin{equation}
	-\Delta u = f
	\end{equation} in $\Omega$, with boundary conditions $u = 0$ on $\partial \Omega$ and $f \in L^2(\Omega)$.  If $\Omega$ is bounded, then this has a unique weak solution in $\dot W^{1, 2}(\Omega)$.

	That is, there exists a unique $u \in \dot W^{1, 2}(\Omega)$ such that \[
		\int_\Omega - \Delta u \phi = \int_\Omega \grad u \cdot \grad \phi = \int_\Omega f \phi
	\]  for all $\phi \in C^\infty_c(\Omega).$  This equation follows from multiplying by a smooth function $\phi$ and integrating by parts. 
\end{prop}

\begin{proof}
	Let $\iprod{u, v} = \int_\Omega \grad u \grad v$ is a scalar product on $\dot W^{1, 2}(\Omega)$ generalising the norm.  The map $\phi \mapsto \int_\Omega f \phi$ is linear in $\phi$.  Our equation then reduces to \[
		\iprod{u, \phi} = \left( f, \phi \right)
	\] where the right hand side is the $L^2$ inner product.  Then we have \begin{align*}
		\left| (f, \phi) \right| &\leq \| f \|_2 \| \phi \|_2 \\
					&\leq C \| f \|_2 \| \grad \phi \|_2 \quad \text{by \poincare inequality}  
	\end{align*} and so $(f, \phi)$ is a bounded linear functional on $\dot W^{1, 2}(\Omega)$.

	So $(f, \phi) = \iprod{F, \phi}$ where $F \in \dot W^{1, 2}(\Omega)$ by the Reisz representation theorem.  Thus setting $u = F$ we obtain our solution.

	Uniqueness is clear from $\iprod{u - F, \phi} = 0 \Rightarrow u - F = 0$.
\end{proof}

\begin{note}
	Consider looking for a solution $u \in C^2(\Omega) \wedge C(\overline\Omega)$ for the equation \begin{align*}
		-\Delta u &= f \quad \text{in $\Omega$}\\
			u &= 0 \quad \text{on $\partial \Omega$}
	\end{align*} with $f \in L^2(\Omega)$.
\end{note}

	We can prove existence of a weak solution quite generally. If $a(u, v): \mathcal H \oplus \mathcal H \rightarrow k$ which is linear in $v$ for fixed $u$ and linear in $u$ for fixed $v$ (bilinear) and there exists $K$ such that \[
		|a(u, v)| \leq K \| u \| \| v \|,
	\] then we can write $a(u, v) = \iprod{Lu, v}$ where $L : \mathcal H \rightarrow \mathcal H$ is bounded and linear.

	If $a$ is of this class on on $\dot W^{1, 2}(\Omega)$ and $f \in L^2(\Omega)$, then the equation 
	\[ 
		a(u, v) = (f, v)
		\tag{$\star$}
	\] for all $v \in \dot W^{1, 2}(\Omega)$ can be written as \[
		\iprod{Lu, v} = \iprod{F, v}
	\] where $L: \dot W^{1, 2}(\Omega) \rightarrow \dot W^{1, 2}(\Omega)$ is bounded linear.  Thus $Lu = F$.  Thus the equation is uniquely soluble if $L$ is invertible.  By the Lax-Milgram result (Theorem \ref{thm:lax_milgram}), $L$ is invertible if \[
	\Re \iprod{Lu, u} \geq c \| u \|^2_{1, 2}
	\tag{$\star \star$}
	\]
	on $\dot W^{1, 2}(\Omega)$ where $c > 0$.  Thus $(\star)$ has a unique solution if $(\star \star)$ holds and $\iprod{Lu, v}$ is bilinear and bounded on $\dot W^{1, 2}(\Omega)$.  Notice that $(\star \star)$ can be written as $\Re a(u, u) \geq c \| u \|_{1, 2}^2$.

Recall that the equation \begin{align*}
	-\Delta u &= f \quad \text{in $\Omega$}\\
		u &= 0 \quad \text{on $\partial \Omega$}
\end{align*} with $f \in L^2(\Omega)$.
has a unique weak solution if $\Omega$ is bounded.  We prove that $u \in W^{2, 2}_{loc}(\Omega)$.  To prove thius, suppose that $x_0 \in \Omega$ and choose $\phi \in C^\infty_c(\Omega)$ and that $\phi = 1$ in a neighbourhood of $x_0$.  We prove that $u \phi$ is the weak solution of the problem \[
	- \Delta (u \phi) + (u \phi) = w 
	\tag{$\star \star \star$}
\]  on $\R^n$ where \[
	w = \underbrace{f \phi - 2 \grad u \grad \phi - u \Delta \phi + u \phi}_{L^2(\R^n)}
\]  But the solution of $(\star \star \star)$ is $W^{2,2}(\R^n)$, which can be derived by Fourier transforms.

We now seek to prove $(\star \star \star)$.  Choose $\psi \in C^\infty_c(\R^n)$.  Then \begin{align*}
	\int \grad(u \phi) \cdot \grad \psi &= \int \left( \psi \grad u  + u \grad \phi \right) \cdot \grad \psi \\
	&= \int u \grad \phi \cdot \grad \psi + \int \phi \grad u \cdot \grad \psi \tag{$\dagger$} 
\end{align*} and similarly,
\begin{align*}
\int \grad u \cdot \grad(\phi \psi)  &= \int \grad u \cdot (\grad \phi) \psi + \int \grad u (\grad \psi) \grad \phi \\
&= \int_\Omega(f \phi \psi)  \\
&= \int f \phi \psi
\end{align*} and so \[
	\int \grad u \cdot (\grad \phi) \psi = \int f \phi \psi - \int \grad u \cdot (\grad \psi ) \phi \tag{$\ddag$}
\]  Then we have \begin{align*}
	\int \grad(u \phi) \cdot \grad \psi &= \int u (\grad \phi) \cdot \grad \psi + \int \phi \grad u \cdot \grad \psi &\text{by $(\dagger)$} \\
	&=  \int u \grad \phi \cdot \grad \psi + \int f \phi \psi - \int \grad u \cdot (\grad \phi) \psi &\text{by $(\ddag)$} \\
	&= - \int \psi \grad(u \grad \phi) + \int f \phi \psi - \int \grad u \cdot (\grad \phi) \psi \\
	&= - \int \psi \left( \grad u \cdot \grad \phi + u \Delta \phi \right) + \int f \phi \psi - \int \grad u \cdot (\grad \phi) \psi \\
	&= \int (f \phi) \psi - 2 (\grad u \cdot \grad \phi) - u (\Delta \phi) \psi.
\end{align*}  So \[
	\int \left( \grad(u \phi) \cdot \grad \psi + u \phi \psi \right) = \int \left(f \phi \psi - (2 \grad u \cdot \grad \phi) \psi - u (\Delta \phi) \psi + u \phi \psi\right).
\]
Hence $u \phi$ is a weak solution on $\R^n$ of $-\Delta z + z = w$.  We can use similar arguments to show that $f \in W^{k, 2}(\Omega)$, which then implies that $u \in W^{k+2, 2}_{loc}(\Omega)$.

It can be show that if $u \in L^p(\Omega)$ and $-\Delta u + u = f$ in $\Omega$, then $u \in W^{2, p}_{loc}(\Omega)$.

Now, consider the weak solutions of \begin{equation}
	-\parder{}{x_l} (a_{ij}(x) \parder{u}{x_j}) + b_l \parder{u}{x_l} + cu = f.
\end{equation} on $\Omega$, with $u \in \dot W^{1, 2}(\Omega)$.   We implicitly use the repeated index summation convention.

We seek to find $u$ such that \[
	\int_\Omega \left( a_{ij}(x) \parder{u}{x_j} \parder{\phi}{x_l}\right) + \int_\Omega b_l \parder{u}{x_l} \phi + \int_C c u \phi = \int_\Omega f \phi 
 \] for all $\phi \in \dot W^{1, 2}(\Omega)$.  The left hand side is a bilinear operator $A(u, \phi)$ where
 \[
 	A: \dot W^{1, 2}(\Omega) \times \dot W^{1, 2}(\Omega) \rightarrow \R
 \] is bounded if $a_{ij}, b_j$, $c \in L^\infty(\Omega)$ and hence are bounded on $\Omega$.  In this case, there is a generalised theorem that \[
 	A(u, \phi) = \iprod{Lu, \phi} 
 \] where $L$ is a bounded linear map $\dot W^{1, 2}(\Omega) \rightarrow W^{1, 2}(\Omega)$.  Then our equation becomes \[
 	\iprod{Lu, \phi} = \int_\Omega f \phi = \iprod{F, \phi}
 \] by the Reisz representation theorem.  That is, $Lu = f$.  Then our problem has a unique solution if $L$ is invertible.  By Lax-Milgram (Theorem \ref{thm:lax_milgram}), this is true if \[
 	A(u, u) \geq c \| u \|^2_{1, 2}.
\]

We now seek to find assumptions such that $A$ satisfies these conditions.  We assume that there exists $c_1 > 0$ such that \[
	\iprod{a_{ij}(x)\eta_i, \eta_j} \geq c_1 |\eta|^2 \tag{$\dagger \dagger$}
\] for all $\eta \in \R^n, x \in \Omega$.

Consider the operator \[
	A(u, v) = \int_\Omega a_{ij}(x) \parder{u}{x_i} \parder{v}{x_j} + \int_\Omega b_l \parder{u}{x_l} v + \int_\Omega cuv.
 \]

To bound the second term, we have have \begin{align*}
	\left| \int_\Omega b_l \parder{u}{x_l} |v| \right| &\leq \int_\Omega |b_l| \left| \parder{u}{x_l} \right| |v| \\
	&\leq K \int \left| \parder{u}{x_l} \right| |v| \\
	&\leq K \left\| \parder{u}{x_l} \right\|_2 \| v \|_2  &\text{by Cauchy-Swartz} \\
	&\leq K \left( \epsilon \left\| \parder{u}{x_l} \right \|_2^2 +  \frac{1}{\epsilon} \|v\|_2^2 \right) &\text{by the inequality $|st| \leq \epsilon s^2 + \frac{t^2}{\epsilon}$}
\end{align*}

Other terms are similar but easier to bound.  Thus we have a bounded bilinear map.

With the above assumptions, we have \begin{align*}
	A(u, u) &= \underbrace{\int_\Omega a_{ij}\parder{u}{x_i}\parder{u}{x_j}}_{\geq \mu \int_\Omega \left| \grad u \right|^2} + \int_\Omega b_l u \parder{u}{x_l} + \int c u^2			
\end{align*}

Estimating the final term, we have \[
	\int c u^2 \geq \inf c \| u \|_2^2.
\]

The first term is bounded by the assumption $(\dagger \dagger)$.

Coercivity is then given, $A(u, u) \geq \alpha \| u \|_{1, 2}^2$ for $\alpha > 0$ if $b_l = 0$ and $\inf c \geq 0$.  If $b_l = 0$ and $\Omega$ is bounded, then $\inf c \geq 0$ is sufficient.

If $b_l$ does not vanish on $\Omega$, then we have the estimate \[
	A(u, u) \geq \mu \| \grad u \|_2^2 - K \left( \epsilon \| \grad u \|_2^2 + \frac{1}{\epsilon} \| u \|_2^2 \right) + \inf c \| u \|_2^2.
\]  Choose $\epsilon$ such that $K \epsilon < \mu$ and $\inf c > \frac{K}{\epsilon}$.  Then we get \[
	A(u, u) \geq \tilde c \left( \| \grad u \|_2^2 + \| u \|_2^2 \right),
\] and we obtain coercivity.

\begin{lem}
	If $A(u, v) = \int_\Omega f v$ for all $v \in \dot W^{1, 2}(\Omega)$, then there is a unique weak solution in $\dot W^{1, 2}(\Omega)$ if $A$ is bounded and bilinear, $A$ is coercive, and $f \in L^2(\Omega)$. 
\end{lem}
% section poincar'e_inequality_and_applications (end)

\section{Compactness in Sobolev Spaces} % (fold)
\label{sec:compactness_in_sobolev_spaces}

\begin{thm}
	\label{thm:compactness_of_w12_inclusion}
	If $\Omega$ is bounded and open in $\R^n$ the natural inclusion $i: \dot W^{1, 2}(\Omega) \rightarrow L^2(\Omega)$ is compact.  That is, bounded sets in $\dot W^{1, 2}(\Omega)$ are contained in a compact set of $L^2(\Omega)$.  
\end{thm}

\begin{rem}
	The theorem does not hold for $\Omega = R^n$, but true for $W^{1, 2}(\Omega)$ under minor assumptions on $\partial \Omega$.  There is a similar result for $\dot W^{1, p}(\Omega)$ for $1 < p < \infty$.
\end{rem}

\begin{lem}
	\label{thm:fourier_bound_for_compact_sobolev}
	For $\epsilon$ sufficiently small, \begin{equation}
		\label{eq:fourier_bound_for_compact_sobolev}
		\left| \left( \hat \phi(\epsilon s) - 1 \right) \right| \leq r \sqrt{1 - |s|^2}
	\end{equation}
	on $\R^n$. 
\end{lem}
\begin{proof}
	We have $\hat \phi(0) -= 1$, $\hat \phi$ is continuous and bounded, and so $\left|\hat \phi(s) \right| \leq K$ on $\R^n$.  
	So \begin{align*}
		\left| \hat \phi(\epsilon s) - 1 \right| &\leq K + 1 \leq r\sqrt{1 + |s|^2}
	\end{align*} if \[
		|s|^2 \geq \underbrace{\left( \frac{K+1}{r} \right)^2 - 1}_{\mu^2}.
		\]  And so this is true if $|s| \geq \mu$ (uniformly in $\epsilon$).  Thus \eqref{eq:fourier_bound_for_compact_sobolev} holds if $|s| \geq \mu$.

	If $|s| \leq \mu$, $\epsilon s$ is small, and so $\left|\hat \phi(\epsilon s) - 1 \right|$ is close to $\hat \phi(0) - 1 = 0$.  Note that $\left| \hat \phi(\epsilon s) - 1 \right| \leq r$ if $\epsilon$ is small and $|s| \leq \mu$.  Hence \[
		\left| \hat \phi(\epsilon s) - 1 \right| \leq r \sqrt{1 + |s|^2}
	\]  if $|s| \leq \mu$ and $\epsilon$ is small.  Hence \eqref{eq:fourier_bound_for_compact_sobolev} holds and our lemma is proven.
\end{proof}

\newcommand{\interior}{\textsc{Int\ }}

\begin{lem}
	\label{lem:bound_on_continuity_of_convolution}
	Given $r > 0$, there exists $\epsilon_0 > 0$ such that $\| T_\epsilon u - u \| \leq r \| u \|_{1, 2}$ if $0 < \epsilon \leq \epsilon_0$ and $u \in \dot W^{1, 2}(\Omega)$.
\end{lem}

\begin{proof} 
	We choose a cube $C$ such that $\overline \Omega \subset \interior C$.  Notice that $\dot W^{1, 2}(\Omega)$ can be extended to $\dot W^{1, 2}(C)$ by letting $u = 0$ on $C \backslash \Omega$.

	Choose $\phi \in C^\infty_c(\R^n)$ such that $\phi \geq 0$, $\int \phi = 1$, and $\phi$ even.  Let \[
		T_\epsilon u = \epsilon^{-n} \int_\Omega \phi\left(\frac{x-y}{\epsilon} \right) u(y) \, dy = \phi_\epsilon \star u
	\] where $\phi_\epsilon(x) = \epsilon^{-n} \phi \left( \frac{x}{\epsilon} \right)$.

	Taking Fourier transforms, we have
	\begin{align*}
		\hat \phi_\epsilon(S) &= \int_{\R^n} e^{its} \phi_\epsilon(t) \, dt \\
			&= \epsilon^{-n} \int e^{its} \phi \left( \frac{t}{\epsilon} \right) \, dt \\
			&= \hat \phi(\epsilon s).
	\end{align*}  Then estimating $\| T_\epsilon u - u \|_2^2$ by Fourier transforms, we have \begin{align*}
		A = \| T_\epsilon u - u \|_2^2 &= (2 \pi)^{-n} \left\| \hat{T_\epsilon u - u}\right\|_2^2 \\
		&= \| \hat{T_\epsilon u } - \hat u \|_2^2.  
	\end{align*}  But $\hat{T_\epsilon u} = \hat \phi_\epsilon \hat u = \hat \phi (\epsilon s) \hat u(s)$, and so \[
		A = (2 \pi)^{-n} \int \left| \left( \left( \hat \phi(\epsilon s) - 1 \right) \hat u(s) \right)^2 \right| \, ds
	\]

	 From Lemma \ref{thm:fourier_bound_for_compact_sobolev}, we have \begin{align*}
		A &\leq (2 \pi)^{-n} \int r^2 \left(1 - |s|^2 \right)^2 \left| \hat u (s) \right|^2 \, ds \\
		&\leq r^2 (2 \pi)^{-n} \int \left( 1 + |s|^2 \right) \left| \hat u(s)\right|^2 \, ds \\
		&= r^2 \| u \|_{1, 2}^2
	\end{align*} using the definition of the $\|u \|_{1, 2}^2$  as $\| u \|_2^2 + \| \grad u \|_2^2$.

	Hence $\|T_\epsilon u - u \|_2^2 \leq r^2 \| u \|_{1, 2}^2$.
\end{proof}

\begin{defn}[Finite $\epsilon$-net]
A finite set $\{ a_i \}_{i=1}^n$ in a metric space $Y$ is a finite $\epsilon$-net if $Y \subseteq \bigcup_{i=1}^n B_\epsilon(a_i)$.
\end{defn}

\begin{thm}
	A closed net $Y$ in a compact metric space is compact if and only if it has a finite $\epsilon$ -net for every $\epsilon > 0$. 
\end{thm}

\begin{defn}[Precompact]
	A subset $T$ in a complete metric space is said to be precompact if $\overline T$ is compact.

	$T$ is precompact if and only if $T$ has a finite $\epsilon$ net for every $\epsilon > 0$.
\end{defn}

\begin{proof}[Proof of Theorem \ref{thm:compactness_of_w12_inclusion}]
	It suffices to show that for any $\delta > 0$, the set \[
		\{ u \in \dot W^{1, 2}(\Omega) \given \| u \|_{1, 2} \leq 1 \}
	\]  lies in a compact set of $L^2(\Omega)$ and hence it suffices to prove if $\delta > 0$ it haas a finite $\delta$-net in $L^2(\Omega)$.   Recall that \[
		\| T_\epsilon u - u \| \leq \frac{1}{2}\delta
	\]  if $u \in \dot W^{1, 2}(\Omega)$, $\| u \|_{1, 2} \leq 1$ by Lemma \ref{lem:bound_on_continuity_of_convolution}.  Thus it suffices to get a finite $\frac{\delta}{2}$-net in $L^2(\Omega)$ for \[
		\{ T_\epsilon u \given \| u \|_{1, 2} \}
	\] for $\epsilon$ small.

	There are $C^1$ function on $\R^n$, and so \[
		(T_\epsilon u)' (x) = \epsilon^{-n-1} \int \phi' \left(\frac{x-y}{\epsilon} \right) u(y) \, dy.
		\tag{$\star$}
	\] It suffices to prove for a fixed $\epsilon$, \[
		\{ T_\epsilon u \given \| u \|_{1, 2} \leq 1 \}
	\] is precompact in $C(C)$ (the set of continuous functions on the cube $C$.)

	The map $i : C(C) \rightarrow L^2(C)$ is continuous and so maps compact sets to compact sets.  For \emph{fixed} $\epsilon$, $\left| (T_\epsilon u)' (x) \right| \leq K$ if $\| u \|_{1, 2} \leq 1$, as \begin{align*}
		|T'(u)(x)| \leq K \int |u(y) | \, dy &\leq K \| u \|_1\\
		&\leq K_1 \|u \|_2 \\
		&\leq K_1 \| u \|_{1, 2} \\
		&\leq K_1.
	\end{align*}

	On $C$, 
	\begin{align*}
		|T_\epsilon u(x_1) - T_\epsilon u(x_2)| &\leq \sup |(T_\epsilon u)'(x)| |x_1 - x_2| \\
		&\leq K_1 |x_1 - x_2|
	\end{align*} for any $x_1, x_2 \in C$.  So $T_\epsilon u$ is uniformly bounded. This shows that $\{ T_\epsilon u \given \| u \|_{1, 2} \leq 1 \}$ is \emph{equicontinuous}, in the sense that given $\mu > 0$, there exists $\tau > 0$ such that $\| T_\epsilon u(x_1) - T_\epsilon(x_2) \| \leq \mu$ if $|x_1 - x_2| \leq \tau$ for all $u$ such that $\| u \|_{1, 2} \leq 1$.

\begin{lem}[Anzela-Anscoli]
	A bounded set in $C(C)$ is precompact if and only if it is equicontinuous.
\end{lem}
\begin{proof}
	See Simmond's book on Modern Analysis.
\end{proof}

Applying Anzela-Anscoli to our set $\{ T_\epsilon u \given \| u \|_{1, 2} \leq 1 \}$ then proves that it is precompact in $C(C)$.  As a set that is precompact in $C(C)$ is precompact in $L^2(C)$, our theorem is proven.
\end{proof}

\begin{rem}
	There are similar results for $i : \dot W^{1, p}(\Omega) \rightarrow L^p(\Omega)$ if $1 < p < \infty$ if $\Omega$ is bounded open.  
\end{rem}

Recall that $\dot W^{1, 2}(\Omega) \rightarrow L^2(\Omega)$ is compact if $\Omega$ is bounded.

Consider the equation \begin{align}
	\label{eq:eigenvalue_equation}
	-\Delta u = \lambda u + f \quad \text{in $\Omega$} \\
	u = 0 \quad \text{on $\partial \Omega$} \notag
\end{align} with $\Omega$ a \textbf{bounded} domain in $\R^n$.

For a weak solution, we seek to find $u \in \dot W^{1, 2}(\Omega)$ such that \[
	\int \grad u \cdot \grad v = \lambda \int u v + \int fv 
\] for all $v \in \dot W^{1, 2}(\Omega)$.  This is equivalent to asking that \[
	\iprod{u, v} = \lambda \iprod{Bu, v} + \iprod{F, v} \tag{$\star$}
\]  where $\iprod{Bu, v} = (u, v)$ is a bounded bilinear form on $\dot W^{1, 2}(\Omega)$ and $\int fv = \iprod{F, v}$.  Note that $(\star)$ is equivalent to \[
	u = \lambda Bu + F \tag{$\star \star$}
\] whern $u \in \dot W^{1, 2}(\Omega)$.

Recall that \begin{align*}
	\left|\iprod{Bu, v}\right| &= \left|\int uv\right| \\
	&\leq \| u \|_2 \| v \|_2 \\
	&\leq C \| \grad u \|_2 \| \grad v \|_2 
\end{align*} by \poincare.  Moreover, $B$ is compact, as $\Omega$ is bounded.  This is true as supposing that $u_n$ is a bounded sequence in $\dot W^{1, 2}(\Omega)$..  Then $\{ u_n \}$ has a convergent subsequence in $\dot W^{1, 2}(\Omega)$.  But by Theorem \ref{thm:compactness_of_w12_inclusion}, $\{ u_n \}$ has a subsequence which converges in $L^2(\Omega)$.  Restricting now to the subsequence, for any $u_n, u_m$, we have \begin{align*}
	\| B u_n - B u_m \|_{1, 2} &= \sup_{\| v \|_{1, 2} \leq 1} \left| \iprod{Bu_n - Bu_m, v} \right| \\
	&= \sup_{\| v \|_{1, 2} \leq 1} \left| \iprod{B(u_n - u_m), v} \right| \\
	&\leq \sup_{\| v \|_{1, 2} \leq 1} \left| (u_n - u_m, v) \right| \\
	&\leq \sup_{\| v \|_{1, 2} \leq 1} \underbrace{\| u_n - u_m \|_2}_{\rightarrow 0} \underbrace{\| v \|_2}_{\leq C \| v \|_{1, 2}} 
\end{align*} by convergence in $L^2(\Omega)$, Cauchy-Swartz and the \poincare inequality.

So $\| Bu_n - Bu_m \|_{1,2} \rightarrow 0$ as $n,m \rightarrow \infty$, and so $\{ B u_n \}$ converges in $\dot W^{1, 2}(\Omega)$ as required.

$B$ is also self adjoint as $\iprod{Bu, v} = \int uv$.

\begin{thm}
	The problem $u = \lambda B u $ on $\dot W^{1, 2}(\Omega)$ has an infinite sequence of eigenvalues $\{ \lambda_n \}$ which are all real and $\lambda_n > 0$ and $\lambda_n \rightarrow \infty$ as $n \rightarrow \infty$.  Moreover, $I- \lambda B$ is invertible if $\lambda \neq \lambda_n$ for all $n$.
\end{thm}
\begin{proof}{\ }

\begin{enumerate}[(i)]
	\item The eigenvalues are all real as $B$ is self-adjoint.
	\item
	Note that the null-space of $B$ is $\{ 0 \}$, since \[
		\iprod{Bu, u} = (u, u) = \int_\Omega u^2 > 0
	\] if $u \neq 0$.

	Hence \begin{align*}
		u = \lambda Bu \iff \underbrace{\iprod{u, u}}_{> 0} &= \iprod{\lambda Bu, u} \\
										&= \lambda \iprod{Bu, u} \\
										&= \lambda \underbrace{\int u^2}_{> 0}
	\end{align*} and so all eigenvalues are greater than zero.

	\item 
		The smallest eigenvalue is $\inf_{u \neq 0} \frac {\int |\grad u |^2}{\int u^2}$. By Theorem \ref{thm:raleigh_rety}, for any operator $T$ we have \begin{align*}
			\sup \sigma(T) 	&= \sup_{\| x \| = 1} \iprod{Tx, x} \\
							&= \sup_{\| x \neq 0 \|} \frac{\iprod{Tx, x}}{\iprod{x, x}} \\
							\inf \sigma(B) &= \inf_{x \neq 0} \frac{\iprod{x, x}}{\iprod{Bx, x}} \\
							&= \inf_{u \neq 0} \frac{\iprod{u, u}}{\iprod{Bu, u}}  \\
							&= \inf_{u \neq 0} \frac{\int |\grad u |^2}{\int u^2}.
		\end{align*} 
	\item If $\lambda \neq \lambda_n$, \eqref{eq:eigenvalue_equation} has a unique weak solution for all $f \in L^2(\Omega)$.  If $\lambda = \lambda_n$, \eqref{eq:eigenvalue_equation} has a solution if and only if $(f, \phi_n) = 0$ for all eigenfunctions $\phi_n$ corresponding to $\lambda = \lambda_n$.

	Recall from Theorem \ref{thm:rangeclosed}, $Tx = y$ has a solution if and only if $f(y) = 0$ for all $f \in \ker T'$.  Note that the this is satisfied if and only if $\iprod{F, \phi_n} = (f, \phi_l) = 0$ for all eigenfunctions $\phi_n$.
	\item The set of eigenfunctions are an orthogonal basis for $L^2(\Omega)$ and $\dot W^{1, 2}(\Omega)$.

	This is true for any compact self-adjoint operator.

	\qedhere
\end{enumerate}
\end{proof}

Consider the equation \begin{align}
    \label{eq:eigenvalue_pde}
    -\Delta u &= \lambda u + f \quad \text{in $\Omega$} \\
    u &= 0 \quad \text{on $\partial \Omega$} \notag \\
\end{align} with $u \in \dot W^{1, 2}(\Omega)$ and $f \in L^2(\Omega)$.

Consider the equation \begin{align*}
    \parder{}{x_i} \left(a_{ij} \parder{u}{x_j} \right) + b_i \parder{u}{x_l} + cu &= \lambda u  + f \quad \text{in $\Omega$} \\
    u = 0 \quad \text{on $\partial \Omega$}
\end{align*} with $\Omega$ bounded.  We can apply the previous theory to this case (modulo some complications.)

Let $f = 0$.  Then if $\tilde \lambda$ is the least eigenvalue of \eqref{eq:eigenvalue_pde} then there is a non-negative eigenfunction of \eqref{eq:eigenvalue_pde} corresponding to $\lambda = \tilde \lambda$.

\begin{thm}
    \label{lem:minimal_eigenvalues}
    Recall that \begin{equation}
        \tilde \lambda = \inf_{\substack{u \in \dot W^{1,2}(\Omega) \\ u \neq 0}} \frac{\int \left| \grad u \right|^2}{\int u^2} 
        \label{eq:minimal_eigenvalue}
    \end{equation}

    If $\tilde u \in \dot W^{1, 2}(\Omega)$ achieves this minimum, then \[
        -\Delta \tilde u = \tilde \lambda \tilde u.
    \]
\end{thm}

\begin{proof}
    Consider test functions of the form $\tilde u + \epsilon \phi$ where $\phi \in \dot W^{1, 2}(\Omega)$. Then \[
        \frac{\int \left| \grad(\tilde u + \epsilon \phi) \right|^2}{\int (\tilde u + \epsilon \phi)^2} \geq \tilde \lambda.
    \]  and \[
        \frac{d}{d\epsilon} \left.\frac{\int \left| \grad(\tilde u + \epsilon \phi) \right|^2}{\int (\tilde u + \epsilon \phi)^2} \right|_{\epsilon = 1} = 0.
    \]  This implies that \[
        \int \grad u \tilde \grad \phi - \tilde \lambda \tilde u \phi = 0
    \]  As this is true for all $\phi \in \dot W^{1, 2}(\Omega)$, we have that $\tilde u$ is a weak solution of \eqref{eq:eigenvalue_pde} for $\lambda = \tilde \lambda$ and $f = 0$.
\end{proof}

\begin{lem}
    If $\tilde u$ is an eigenfunction corresponding to $\tilde \lambda$ then $|\tilde u |$ is in $\dot W^{1, 2}(\Omega)$ and $|\tilde u$ is a minimiser of \eqref{eq:minimal_eigenvalue}, and hence, as in Lemma \ref{lem:minimal_eigenvalues}, $|\tilde u|$ is also an eigenfunction corresponding to $\tilde \lambda$.   
\end{lem}

\begin{proof}

Recall that \[
    |\tilde u |(x) = \begin{cases}
        \tilde u(x)     & \tilde u \geq 0 \\
        -\tilde u(x)    & \tilde u(x) < 0
    \end{cases}
\]  By the next section, \[
    \parder{}{x_i}|\tilde u|(x) = \begin{cases}
        \parder{\tilde u}{x_i}  & \tilde u(x) \geq 0 \\
        -\parder{\tilde u}{x_i} & \tilde u(x) < 0 \\
    \end{cases}
\]  Then \[
    \left| \parder{}{x_i} |\tilde u|(x) \right| = \left| \parder{\tilde u}{x_i} \right|
\] and so \[
    \frac{\int \left(\grad|\tilde u| \right)^2}{\int |\tilde u |^2}  = \frac{\int \left| \grad \tilde u \right|^2}{\int \tilde u^2} = \tilde \lambda. \qedhere
\]
\end{proof}

\begin{thm}
    If $f \in L^2(\Omega)$ is non-negative and \begin{align*}
        -\Delta u &= f \quad \text{in $\Omega$} \\
        u &= 0 \quad \text{on $\partial \Omega$}
    \end{align*} for $u \in \dot W^{1, 2}(\Omega)$, then $u \geq 0$ on $\partial \Omega$.  
\end{thm}

\begin{proof}
    Consider $u^-$ as a test function in the definition of the weak solution \[
        u^-(x) = \begin{cases}
            0       & u(x) \geq 0 \\
            u(x)    & u(x) < 0.
        \end{cases}
    \] and \[
     \parder{}{x_i}u^-(x) = \begin{cases}
            0               & u(x) \geq 0 \\
            \parder{u}{x_i} & u(x) < 0.
        \end{cases}
    \]  Since \[
        \int \grad u  \cdot \grad \phi = \int f \phi
    \] letting $\phi = u^-$, we have \[
    \underbrace{\int f u^{-}}_{\leq 0} = \int \grad u \grad u^- = \underbrace{\int \left| \grad u^- \right|^2 }_{\leq 0}
    \]  Thus $\grad u^- = 0$ and so $u^- = 0$ by \poincare inequality (Theorem \ref{thm:poincare_inequality}).
\end{proof}

% section compactness_in_sobolev_spaces (end)

\section{Further Properties of $\dot W^{1,2}(\Omega)$} % (fold)
\label{sec:further_properties_of_dot_w_1_2_omega_}
\begin{thm}
    \label{thm:his_thm_7.1}
    If $u \in \dot W^{1, 2}(\Omega)$ where $\Omega$ is bounded and open then $u^+ \in \dot W^{1, 2}(\Omega)$ and \[
        \parder{}{x_i} u^+ = \begin{cases}
            \parder{u}{x_i} & u(x) > 0 \\
            0 & u(x) \leq 0
        \end{cases}
    \]  
\end{thm}

\begin{proof}
    If $f \in C^1(\Omega)$, $f(0) = 0$, and $f'$ is bounded on $\R$, then $f(u) \in \dot W^{1,2}(\Omega)$ and $\parder{}{x_i} f(u) = f'(u) \parder{u}{x_i}$ if $u \in C^\infty_c(\Omega)$ by the chain rule.

    If $u \in \dot W^{1, 2}(\Omega)$, choose $u_n \in C^\infty_c(\Omega)$ so $u_n \rightarrow u$ in $\dot W^{1, 2}(\Omega)$ as $n \rightarrow \infty$.  Then \begin{align*}
        -\int f(u_n) \parder{\phi}{x_l} = \int f'(u_n) \parder{u_n}{x_i} \phi
    \end{align*}
    Since $u_n \rightarrow u$ and $\parder{u_n}{x_i} \rightarrow \parder{u}{x_i}$ in $L^2$, taking subsequences gives \[
        - \int f(u) \parder{\phi}{x_i} = \int f'(u) \parder{u}{x_i} \phi \qedhere.
    \]  This can be shown as $|f(0) - f(t)| \leq K |s-t|$ by the mean value theorem, and so $|f(u_n(x)) - f(u(x))| \leq K | u_n(x) - u(x) |$.  On the left hand side, it suffices to show that $f(u_n) \rightarrow f(u)$ in $L^1(\Omega)$, then we use the dominated convergence theorem.  We have \begin{align*}
    	\|f(u_n(x)) - f(u(x)) \|_1 \leq K \| u_n - u \|
    \end{align*}

On the right hand side, we have $L^2$ convergence if we prove that each term ($\parder{u_n}{x_i}$, $f'(u_n) \phi$) converges in $L^2(\Omega)$,  We have \[
	\parder{u_n}{x_i} \rightarrow \parder{u}{x_i}
\] by a lemma of generalised derivatives, and \[
	f'(u_n) \phi \rightarrow f'(u) \phi
\] since they are both uniformly bounded and converge pointwise.

    More explicitly, we have \begin{align*}
        \int f(u) \parder{\phi}{x_l} - \int f(u_n) \parder{\phi}{x_l} &= \int \left(f(u_n) - f(u) \right) \parder{\phi}{x_i} \\
        &\leq \underbrace{\left\| f(u) - f(u_n) \right|_2}_{\rightarrow 0} \left\| \parder{\phi}{x_i} \right\|_2
    \end{align*}

	Now, consider the function $f_\epsilon(y)$ given by \[
		f_\epsilon(y) = \begin{cases}
			\sqrt{y^2 + \epsilon^2} & y \geq 0 \\
			0 & y < 0.
		\end{cases}
	\]  Then $f_\epsilon \in C^1$, $f'_\epsilon (y) = \frac{y}{\sqrt{y^2 + \epsilon^2}}$, and $|f'_\epsilon (y)| \leq 1$.

	If $u \in \dot W^{1, 2}(\Omega)$, then \begin{align*}
		-\int f_\epsilon (u) \parder{\phi}{x_i} = \int f'_\epsilon(u) \parder{u}{x_i} \phi
	\end{align*} by the previous step.

	Note that $f_\epsilon (y) \rightarrow y^+$ uniformly in $\R$  as $\epsilon \rightarrow 0$, and hence converges in $L^2$.  Thus \[
		\int f_\epsilon(u) \phi = \int u^+ \phi.
	\]

	Next, note that $f'_\epsilon (y)$ is uniformly bounded and converges pointwise and in $L^2(\Omega)$ to $\mathbf{1}_{y > 0}$.  Thus by Cauchy-Swartz, \[
		\int f'_\epsilon(u) \parder{u}{x_i} \phi \rightarrow \int \mathbf{1}_{u > 0} \parder{u}{x_i} \phi.
	\] Hence \[
		\int u^+ \parder{\phi}{x_l} = \int \mathbf{1}_{u > 0} \parder{u}{x_i} \phi
	\] and so $\parder{u^+}{x_i}$ exists and is $\mathbf{1}_{u > 0} \parder{u}{x_i}$.
\end{proof}

\begin{rem}
	This tells us that $\parder{u^+}{x_i} = 0$ a.e. where $u = 0$, and also that $\parder{u^-}{x_i} = -\mathbf{1}_{u < 0} \parder{u}{x_i}$.
\end{rem}

\begin{rem}
	If $u \in  W^{1, 2}(\Omega)$m then $u^+ \in W^{1, 2}(\Omega)$ with the same formula for $\parder{u}{x_i}$.

	If $N = 1$ and $\dot W^{1, 2}([a, b])$ then there exists $\tilde u$ such that $\tilde u  = u$ almost everywhere and $\tilde u \in C[a, b]$.

	If $N = 2$ and $u \in \dot W^{1, 2}(\Omega)$, $u \in L^p(\Omega)$ for all $p \geq 1$.

	If $N \geq 3$ and $u \in \dot W^{1, 2}(\Omega)$, then $u \in L^{2^*}(\Omega)$ where $2^\star = \frac{2N}{N-2}.$
\end{rem}

\begin{exer}
	If $u \in \dot W^{1, 2}(\Omega)$ and $a > 0$, then \[
		(u-a)^+ \in \dot W^{1, 2}(\Omega).
	\]
\end{exer}

\begin{rem}
	In general, if $F$ is Lipschitz on $\R$ with $F(0) = 0$ and $u \in \dot W^{1, 2}(\Omega)$, then \[
		F(u) \in \dot W^{1, 2}(\Omega).
	\]
\end{rem}

\begin{thm}
	\label{thm:his_thm_7.2}
	If $n = 1$ and $u \in \dot W^{1, 2}(\Omega)$, then $u \in C(\overline \Omega)$ and $u = 0$ on $\partial \Omega$ (assuming $\Omega$ is bounded and open).  More precisely, there exists $K > 0$ such that if $u \in \dot W^{1, 2}(\Omega)$ there exists $v \in C(\Omega)$ such that $v = 0$ on $\partial \Omega$, $v = u$ almost everywhere, and $\| v \|_\infty \leq K \| u \|_{1, 2}$.

	This is true for $\dot W^{1, p}(\Omega)$ if $n = 1$ and $p > 1$.
\end{thm}

\begin{proof}
	We prove for $\Omega = (a, b)$, as any open set in $\R$ is a countable union of disjoint intervals.  We prove that there exists $K > 0$ such that if $u \in C^\infty_c((a, b))$, then 
	\[
		\| u \|_\infty \leq K \| u \|_{1, 2}.
		\tag{$\star$}
	\]

	This will be sufficient to prove the theorem.  To show this, suppose that $w \in \dot W^{1, 2}(\Omega)$ and $u_n \in C^\infty_c((a,b))$ such that $u_n \rightarrow w$ in $\dot W^{1, 2}([a,b])$ as $n \rightarrow \infty$, then \begin{align*}
		\| u_n - u_m \|_\infty &\leq K \| u_n - u_m \|_{1, 2} \quad \text{by ($\star$)}\\
		&\leq K \left( \| u_n - w \|_{1, 2} + \| u_m - w \|_{1, 2} \right) \\
		&\rightarrow 0 
	\end{align*} as $m, n \rightarrow \infty$.  Hence $\sup_{x \in [a, b]} |u_n(x) - u_m(x)| \rightarrow 0$ as $m, n \rightarrow \infty$, and so $\{ u_n \}$ is Cauchy in $C([a, b])$.  Hence there exists $v \in C([a, b])$ such that $u_n \rightarrow v$ uniformly as $n \rightarrow \infty$ and $v(a) = v(b) = 0$.  Since $u_n \rightarrow w$ in $L^2([a,b])$ as $n \rightarrow \infty$, then $u_n(x) \rightarrow w(x)$ almost everywhere as $n \rightarrow \infty$, and so $v = w$ almost everywhere, with $w$ continuous and $w(a) = w(b) = 0$.

	So we have \begin{align*}
		\| u_n \|_\infty &\leq K \| u_n \|_{1, 2} \\
		\| v \|_\infty &\leq K \| v \|_{1, 2}
	\end{align*} and this is what we need.  It suffices to prove $(\star)$ for $u \in C^\infty_c((a,b))$.  Let $x, y \in (a, b)$, with $x < y$.  THen \[
		u(x) - u(y) = \int_x^y u'(t) \, dt
	\]  So \begin{align*}
		|u(x) - u(y) | &\leq \left| \int_x^y u'(t) \, dt \right| \\
		&\leq \left( \int_x^y \, dt \right)^{1/2} \left( \int_x^y |u'(t)|^2 \, dt \right)^{1/2} \\
		&\leq \left(y-x\right)^{1/2} \left( \int_a^b |u'(t)|^2 \, dt \right)^{1/2} \\
		&\leq (b-a)^{1/2} \| \grad u \|_2
	\end{align*} 
\end{proof}

\begin{thm}
	\label{thm:his_thm_7.3}
	If $n > 2$ and $\Omega$ is bounded and open, there exists $C$ depending only on $n$ such that if $u \in \dot W^{1, 2}(\Omega)$ hen $u \in L^{2^\star}(\Omega)$ and $\| u \|_{2^\star} \leq C \| \grad u \|_{1, 2}$.  Here $2^\star = \frac{2n}{n-2}$.   
\end{thm}

\begin{rem}
	There is a similar theorem for $\dot W^{1, 2}(\Omega)$ if $1 \leq p < n$.  Here $p^\star = \frac{np}{n - p}$.
\end{rem}

\begin{rem}
	$u \in L^2(\Omega)$ and $u \in L^{2^\star}(\Omega)$ implies that $u \in L^{q}(\Omega)$ for all $2 \leq q \leq 2^\star$ by H\"older's inequality.
\end{rem}

\begin{proof}[Proof of Theorem \ref{thm:his_thm_7.3}]{\ }
	It suffices to assume that $u \in C^\infty_c(\Omega)$.

	\textbf{Step 1.} 
	\begin{quote}
	\begin{lem}
		If $u \in \dot W^{1, 1}(\Omega)$, then $\| u \|_{1^\star} \leq \| \grad u \|_{1}$, where $\mathbf{1}^\star = \frac{n}{n-1}$.
	\end{lem}

	\begin{proof}
		By H\"older, we have \begin{align*}
			\int |g_1| \times |g_2| \dots  |g_{n-1}| \leq \left( \int |g_1|^{n-1}\right)^{\frac{1}{n-1}} \dots \left( \int |g_{n-1}|^{n-1} \right)^{\frac{1}{n-1}} \\
			\left( \int |g_1| \times |g_2| \dots |g_{n-1} \right)^{n-1} \leq \prod_{i=1}^{n-1} \left( \int |g_i|^{n-1} \right) \quad \text{by induction.}
		\end{align*}
		For $f \in C^{\infty}_c(\R^n)$, we have \[
			f(x_1, \dots, x_n) = \int_{-\infty}^{x_i} \parder{f}{x_i}(x_1, \dots, x_{i-1}, t_i, x_{i+1}, \dots, x_n) \, dt_i
		\]  Then we can estimate $|f|$ by \begin{align*}
			|f(x_1, \dots, x_n)| &\leq \int_{-\infty}^{x_i} \left| \parder{f}{x_i}(x_1, \dots, x_{i-1}, t_i, x_{i+1}, \dots, x_n) \right| \, dt_i \\
			&\leq \int_{-\infty}^{\infty} \left| \grad f (x_1, \dots, x_{i-1}, t_i, x_{i+1}, \dots, x_n) \right| \, dt_i
		\end{align*} and so \[
			|f|^n \leq \prod_{i=1}^n \int_{-\infty}^{\infty} \left| \grad f (x_1, \dots, x_{i-1}, t_i, x_{i+1}, \dots, x_n) \right| \, dt_i
		\]  Then taking the $(n-1)$-th root, we obtain \begin{align*}
			|f|^{1^\star} \leq \prod_{i=1}^n \left( \int_{-\infty}^{\infty} \left| \grad f (x_1, \dots, x_{i-1}, t_i, x_{i+1}, \dots, x_n) \right| \, dt_i \right)^{\frac{1}{n-1}}.
		\end{align*}  Now, integrating in $x_1$, we have \begin{align*}
			\int |f|^{1^\star} \, dx_1 \leq \left( \int_{-\infty}^{\infty} |\grad f | \, dt_1 \right)^{\frac{1}{n-1}} \prod_{i=2}^n \left(\int \left|\grad f(x_1, \dots, x_{i-1}, t_i, x_{i+1}, \dots, x_n) \right| \, dt_i \, dx_1 \right)^{\frac{1}{n-1}}
		\end{align*}

		 Now integrating in $x_2$, we have \begin{align*}
		 			\int |\grad f |^{1^\star} \, dx_1 \, dt_2 \leq \left( \iint \left|\grad f \right| \, dt_1 \, dx_2 \right)^{\frac{1}{n-1}}  \left( \iint \left|\grad f \right| \, dt_1 \, dx_2 \right)^{\frac{1}{n-1}}  \left(\prod_{i=3}^n \iint |\grad f| \, dx_1 \, dx_2 \, dt_i \right)^{\frac{1}{n-1}}
		 		\end{align*}
		 % 		
		By induction, we obtain \begin{align*}			
			\int |f|^{1^\star} \, dx_1 \dots dx_n  &\leq \left( \prod_{i=1}^n \int |\grad f | \, dx_1 \dots dx_n \right)^{\frac{1}{n-1}} \\
						&\leq \left( \int |\grad f| \, dx_1 \dots dx_n \right)^{\frac{n}{n-1}}
		\end{align*}	 and taking $\frac{n-1}{n}$-th roots obtains the required result.
		\end{proof}

		We can also show \begin{enumerate}[(i)]
			\item If $p > n$, then $\dot W^{1, p}(\Omega) \subseteq C(\overline \Omega)$ and $\| u \|_{\infty} \leq C \| \grad u \|_p$.
			\item If $n \leq 3$, function in $W^{2, 2}(\Omega)$ are continuous on the interior.
			\item If $u \in W^{1, 2}(\Omega)$ and $u(x) \rightarrow 0$ as $x \rightarrow \partial \Omega$, then $u \in \dot W^{1, 2}(\Omega)$.
		\end{enumerate}

	\end{quote}

	\textbf{Step 2.}
	\begin{quote}
 Complete the proof.  As before, we construct $u \in C^\infty_c(\R^n)$.  Applying \textbf{Step 1.} to $|u|^\gamma$ where $\gamma > 1$ and $\gamma$ is to be chosen, then \begin{align*}
		\left( \int \left( |u|^\gamma \right)^{\frac{n}{n-1}} \right)^{\frac{n-1}{n}} &\leq \left\| \gamma |u|^{\gamma - 1} (\grad u) \right\|_1 \\
		&\leq \gamma \left\| |u|^{\gamma - 1} \grad u \right\|_1 \\
		&\leq \gamma \left( \int u^{(\gamma - 1) \cdot \frac{p}{p-1}}\right)^{\frac{p-1}{p}} \left( \int \left| \grad u \right|^p \right)^{1/p} \quad \text{by H\"older}
	\end{align*} 
	 Then choosing $\gamma$ such that $\gamma \frac{n}{n-1} = \left( \gamma - 1 \right) \frac{p}{p-1} = p^\star$,  we have \begin{align*}
		\left( \int |u|^{p^\star} \right)^{\frac{n-1}{n}} \leq \gamma \left( |u|^{p^\star} \right)^{\frac{p-1}{p}} \left( \int |\grad u |^p \right)^{\frac{1}{p}}
	\end{align*}  Then dividing both sides by $\left(\int |u|^{p^\star} \right)^{\frac{p-1}{p}}$ , we have \begin{align*}
		\left(\int|u|^{p^\star} \right)^{\frac{1}{p^\star}} \leq C(\gamma) \left(\int |\grad u |^p \right)^{\frac{1}{p}}
	\end{align*}

	\end{quote}

\end{proof}

\begin{lem}
	Consider the differential equation \begin{align*}
		\grad u &= f \quad \text{on $\Omega$} \\
		u &= 0	\quad \text{on $\partial \Omega$}
	\end{align*} with $\Omega$ bounded.  This has a weak solution in $\dot W^{1, 2}(\Omega)$.  We claim that any classical solution is also a weak solution.
\end{lem}

Consider a classical solution $u \in C^2(\Omega) \cap C(\overline \Omega)$.  Then a classical solution (if it exists) is a weak solution $u \in \dot W^{1, 2}(\Omega)$ and \[
	\int \grad u \cdot \grad \phi = \int f \phi
\]  for all $\phi \in C^\infty_c(\Omega)$.  If $u$ is smooth and $\phi$ has compact support, then \[
	\int \grad u \cdot \grad \phi = - \int \Delta u \phi = \int f \phi
\] if $\grad u = f$.

We need to check $\grad u \in L^2(\Omega)$ and $u \in \dot W^{1, 2}(\Omega)$.  Consider the function $(u-a)^+ \in W^{1, 2}(\Omega)$ if $a > 0$ that vanishes near $\partial \Omega$.  Then $u - a \in W^{1, 2}(\Omega)$ and so $(u-a)^+ \in W^{1,2}(\Omega)$ on compact sets, as \[
	\parder{}{x_i}(u-a)^+ = \parder{u}{x_i} \mathbf{1}_{\{ u > a \}}.
\]     Then $(u-a)^+ \in \dot W^{1, 2}(\Omega)$ (by an exercise.)

Using $(u-a)^+$ as a text function, we have \begin{align*}
	\int \grad u \cdot \grad (u-a)^+ &= \int f (u-a)^+  \\
									&\leq \leq \| f \|_2 \| (u-a)^+ \|_2 \quad \text{by Cauchy-Swartz} \\
									&= K
\end{align*} but \begin{align*}
	\int \grad u \cdot \grad (u-a)^+ &= \int \left| \grad u \right|^2 \mathbf{1}_{\{ u > a \}} \leq K \\
		&\rightarrow \int_{u \geq 0} \left| \grad u \right|^2
\end{align*} by monotone convergence theorem as $a \rightarrow 0$, and so $u^+ \in L^2$.  Similarly, $\grad u^- \in L^2(\Omega)$, and so $\grad u \in L^2(\Omega)$.
% section further_properties_of_dot_w_1_2_omega_ (end)

\section{Applications to Nonlinear Equations} % (fold)
\label{sec:applications_to_nonlinear_equations}
Consider the differential equation \begin{align*}
	-\Delta u = g(u) \quad \text{in $\Omega$} \\
	u = 0 \quad \text{on $\partial \Omega$}
\end{align*} where $g: \R \rightarrow \R$ is continuous - so $-\Delta u(x) = g(u(x))$.

We look for weak solutions, that is $u \in \dot W^{1, 2}(\Omega)$ satisfying \begin{equation}
	\label{eq:his_eq_8.1}
	\int \grad u \cdot \grad \phi = \int g(u) \phi \quad \text{for all $\phi \in C^\infty_c(\Omega)$.}
\end{equation}

\section{Variational Methods} % (fold)
\label{sub:variational_methods}

% subsection variational_methods (end)
Assume that $\Omega$ is bounded and $g$ is continuous and satisfies \[
		\left| g(y) \right| \leq K_1 | y | + K_2
	\] on $\R$, and if $G' = g$ we assume that there exists $\mu < \lambda_1$ such that\footnote{Here, $\lambda_1$ is the minimal eigenvalue of the eigenvalue equation \begin{align*}
			-\Delta u &= \lambda u \quad \text{ on $\Omega$.} \\
			u &= 0 \quad \text{on $\Omega$}.
		\end{align*}  Indeed, \[
			\lambda_1 = \inf \frac{\int |\grad u |^2}{\int u^2}.
		\]
		} \[
		G(y) \leq \frac{1}{2} \mu y^2 
	\]  for $|y|$ large.   Equivalently, $G(y) \leq \frac{1}{2} \mu y^2 + K_3$.

Consider the \emph{energy} function $E: \dot W^{1, 2}(\Omega) \rightarrow \R$ defined by \begin{equation}
	\label{eq:energy}
	\int_\Omega \left( \frac{1}{2} \left| \grad u \right|^2 - G(u) \right).
\end{equation}  We prove that there exists $w \in \dot W^{1, 2}(\Omega)$ such that \[
	E(u) \geq E(w)
\] for all $u \in \dot W^{1, 2}(\Omega)$ and that such a $w$ is a weak solution of our equation.

\begin{quote}\textbf{Step 1.}
	We prove that there exists $C_1 > 0$ such that $E(u) \geq - C_1$ for all $u \in \dot W^{1, 2}(\Omega)$.  From \eqref{eq:his_eq_8.1}, we have \begin{align*}
		E(u) &\geq \int_\Omega \left( \frac{1}{2} \left| \grad u \right|^2 - \frac{1}{2} \mu u^2 - K_3 \right) \tag{$\star \star$} \\
		&\geq \frac{1}{2} \underbrace{\int \left( \left| \grad u \right|^2 - \mu u^2 \right)}_{\geq 0} - \tilde K_3 \\
		& \geq - \tilde K_3 \\
		& \equiv \gamma.
	\end{align*} since $\lambda_1 = \inf \frac{\int|\grad u |^2}{\int u^2}$ and so $\int |\grad u|^2 \geq \lambda_1 \int u^2$.  Hence \[
		\int \left( |\grad u |^2 - \mu u^2 \right) \geq (\lambda_1 - \mu) \int u^2 \geq 0.
	\]  We get a little more, \[
		E(u) \geq (\lambda_1 - \mu) \left(\int u^2 \right)
 - K_3	\] so if $E(u) \leq K_4$, we have that $\int u^2$ is bounded.  Thus by $(\star \star)$, \[
 	\int |\grad u|^2
 \] is bounded.  Thus if \[
	E(u_n) \rightarrow \inf \left\{E(u) \given u \in \dot W^{1, 2}(\Omega) \right\},
\] then $\{ u_n \}$ is bounded in $\dot W^{1 ,2}(\Omega)$.

\begin{lem}
	 The sequence $\{ u_n \}$ has a subsequence which converges weakly to $w  \in \dot W^{1, 2}(\Omega)$ and $w$ is a minimiser of $E$.
\end{lem}
\begin{proof}
	Recall from \S \ref{sec:linear_operators_on_hilbert_spaces} that every bounded sequence in a Hilbert space $\mathcal H$ has a subsequence which converges weakly. Thus our sequence $\{ u_n \}$ has a subsequence that converges weakly to $w$.

	We now need only show that $w$ is a minimiser of $E$.   Let $u_n \rightharpoonup w$ in $\dot W^{1, 2}(\Omega)$.  Let $i: \dot W^{1, 2}(\Omega) \rightarrow L^2(\Omega)$ be the inclusion mapping.  Then $i$ is a bounded linear operator, and \[
		i(u_n) \rightharpoonup i(w)
	\] in $L^2(\Omega)$.  That is, $u_n \rightarrow w$ in $L^2(\Omega)$.  Since bounded sets in $\dot W^{1, 2}(\Omega)$ are precompact sets in $L^2(\Omega)$, we can choose a subsequence such that $u_n \rightarrow w$ (strongly) in $L^2(\Omega)$.  Hence the weak convergence in $\dot W^{1, 2}(\Omega)$ can be ``converted" into strong convergence in $L^2(\Omega)$.

	We now need to show that $w$ minimises $E$ and $w$ is a solution to our equation.  We need to show that $E(u_n) \rightarrow E(w) = \gamma$. 
	Recall that in a Banach space, if $u_n \rightharpoonup u$ weakly, then \[
		\| u \| \leq \| \liminf_{n \rightarrow \infty} \| u_n \|.
	\]
	Now, we then have \[
		\| w \|_{1, 2} \leq \liminf_{n \rightarrow \infty}  \| u_n \|_{1, 2}
	\]  Taking squares, we obtain \[
		\| \grad w \|_2^2 \leq \liminf_{n \rightarrow \infty} \| \grad u_n \|_2^2.
	\]  We also need to prove that \[
		\tag{$\star \star \star$}
		\int G(u_n) \rightarrow \int G(w) \quad \text{as $n \rightarrow \infty$}
	\]  Then we can show that \[
		E(u_n) = \frac{1}{2} \int \left| \grad u_n \right|^2 - \int G(u_n)  \rightarrow \gamma 
	\] and hence \[
		E(w) = \frac{1}{2} \int \left| \grad w \right|^2 - \int G(w)  \leq \gamma.
	\] But $E(w) \geq \gamma$, and so $E(w) = \gamma$, that is, $w$ is a minimiser.

	It thus remains to prove $(\star \star \star)$.  Since $u_n \rightarrow w$ in $L^2(\Omega)$, we can show that $u_n \rightarrow w$ a.e by taking subsequences.  By a result in analysis (Ergerov's theorem), there exists sets $V_k$ of arbitrarily small measure such that \[
		u_n(x) \rightarrow w(x)
	\] uniformly on $\Omega \, \backslash \, V_k$ as $n \rightarrow \infty$, again taking subsequences.  We know that $w$ is bounded off a set of  small measure and hence we can find a set $Z$ of small measure so $u_n \rightarrow w$ uniformly on $\Omega \, \backslash \, Z$ and $w$ is bounded on $\Omega \, \backslash \, Z$.  This implies that \[
		G(u_n) \rightarrow G(w)
	\]  uniformly on $\Omega \, \backslash \, Z$ the fact that a continuous function on $\R$ is uniformly continuous on bounded sets.  Hence, \[
		\int_{\Omega \, \backslash \, Z} G(u_n) \rightarrow \int_{\Omega \, \backslash \, Z} G(w).
	\]  We now prove $\int_Z G(u_n), \int_Z G(w)$ are uniformly small in $Z$ if $Z$ has small measure.  We have \begin{align*}
		\int_Z G(u_n) &\leq \int_Z \left(\frac{1}{2} \mu u_n^2 + K_3 \right) \\
					&\leq  \frac{1}{2} \mu \int_Z u_n^2 + K_3 m(Z) 
	\end{align*} where $m(Z)$ is the measure of $Z$.

	 Since $u_n$ is bounded in $\dot W^{1, 2}(\Omega)$, by Sobolev's embedding theorem, we can show that $\| u_n \|_{p^\star}$ is bounded for $p^\star > 2$.  So the first term is less than or equal to $\frac{1}{2}\mu \| u_n \|_{2, Z}^2$.  Since \begin{align*}
		\int_Z u_n^2 &\leq \left( \int_Z \left(|u_n|^2\right)^{q} \right)^\frac{1}{q} \left( \int_Z 1^q \right)^\frac{1}{q'} \\
	\end{align*}  for $q, q'$ H\"older pairs, so letting $p^\star = 2q$ for $q > 1$, we have \begin{align*}
		\int_Z u_n^2 &\leq \left( \int_Z |u_n|^{p^\star}\right)^{\frac{1}{q}} \left(m(Z \right)^{\frac{1}{q'}} \\
		&\leq  \left( \| u_n \|_{p^\star} \right)^{\frac{p^\star}{2}} \left( m(Z) \right)^{\frac{1}{q'}}
	\end{align*} as required.

	Recall that since $w$ is a minimizer, we have \begin{align*}
		E(w + t \phi) \geq E(w) \quad \forall \phi \in C^\infty_c(\Omega)\, \quad \forall t \\
		\left.\frac{d}{dt} E(w + t \phi) \right|_{t = 0} = 0
	\end{align*}  if it exists.  We will now prove that the derivative exists and equals \[
		\int_\Omega \grad w \cdot \grad \phi - g(w) \phi. 
	\]  In this case, \[
		\int \grad w \cdot \grad \phi = g(w) \phi \quad \forall \phi \in C^\infty_c(\Omega),
	\] and so $-\Delta w = g(w)$.

	We have \begin{align*}
		E(w + t \phi) &= \frac{1}{2} \int_\Omega \grad(w + t \phi) \cdot \grad (w + t \phi) - \int_\Omega G(w + t \phi) \\
		&= \frac{1}{2} \int_\Omega |\grad w|^2 + 2t\grad w \cdot \grad \phi + t^2 |\grad \phi |^2 - \int G(w + t \phi).
	\end{align*} Therefore \begin{align*}
		\frac{d}{dt} E(w + t \phi) = \int_\Omega \grad w \cdot \grad \phi + t \int_\Omega |\grad \phi|^2 - \frac{d}{dt} \int G(w + t \phi) \\
		\frac{d}{dt} E(w + t\phi) |_{t = 0} = \int_\Omega \grad w \cdot \grad \phi - \frac{d}{dt} \int G(w + t \phi). 
	\end{align*}  We thus need only prove that \[
		\frac{d}{dt}(\int G(w + t\phi))|_{t = 0} = \int g(w) \phi.
	\]  Now \begin{align*}
		\frac{  \int G(w + t \phi) - G(w)}{t} = \int G'(w + \theta(x) t \phi(x)) \phi(x)
	\end{align*} where $0 \leq \theta(x) \leq 1$.  We need to prove (remembering $G' = g$), that \[
		\int G'(w + \theta(x) t \phi(x)) \phi(x) \rightarrow \int g(w) \phi(x)
	\]

	Choose a set $T$ so that $\mu(\Omega - T)$ is small and $w, \phi$ are bounded on $T$. Then \[
		g(w + t \theta(x) \phi(x)) \rightarrow g(w(x)) \phi(x)
	\] uniformly on $T$ as $t \rightarrow 0$ as $g$ is uniformly continuous on bounded sets.  We need only prove that \[
		\int_{\Omega \backslash T} g(w + t \theta(x) \phi(x)) \phi(x)
	\] is small for all $t$ small.

	.... \emph{CBF finishing this}.

	\begin{rem}{\ }
		\begin{enumerate}[(i)]
			\item If $g(0) = 0$, our minimum may be $u(x) = 0$.  
			\item If $g(0) = 0$ and $g'(0) > \lambda$, $0$ may not be the minimum and we must have a non-trivial solution.  We only need to find $z \in \dot W^{1, 2}(\Omega)$ with $E(Z) < 0$.  We choose $z = t \phi$, where $t$ is small and positive and $\phi_1$ is the eigenfunctino corresponding to $\lambda_1$.  Then \[
				G(s) = \frac{1}{2} g'(0) s^2 + m(s),
			\] where $\frac{m(s)}{s^2} \rightarrow 0$ as $s \rightarrow 0$.  Then \begin{align*}
			E(t \phi_1) = \frac{1}{2} t^2 \left( \lambda_1 - g'(0) \right) \int_\Omega \phi_1^2 + o(t^2)  < 0	
			\end{align*} if $t$ is small.
		\end{enumerate}
	\end{rem}
\end{proof}

\end{quote}

\section{Fixed Point Methods} % (fold)
\label{sec:fixed_point_methods}
\begin{thm}[Brower]
	$B^n$ is the closed ball in $\R^n$ and $f: B^n \rightarrow B^n$ is continouus then there exists $x \in B^n$ such that $f(x) = x$.  
\end{thm}

\begin{defn}[Completely continuous]
	$A: E \rightarrow E$ is completely continuous (cc) if $A$ is continuous and if $D$ is bounded in $E$, then $A(D)$ is compact in $E$.  
\end{defn}

\begin{lem}
	If $E$ is an infinite dimensional Banach space hen $I: E \rightarrow E$ is not cc.

	If $A$ is linear, $A : E \rightarrow E$, then $A$ is cc if and only if $A$ is compact.

	(Shauder).  If $D$ is closed, bounded and convex in a Banach space $E$ and $A: D \rightarrow E$ is cc and $A(D) \subseteq D$, then there exists $x \in D$ such that $A(x) = x$ (fixed point).
\end{lem}

\begin{exmp}[Example of fixed point methods]
	Let $g : \R \rightarrow \R$ be continuous and $\frac{g(y)}{y} \rightarrow \tau$ as $|y| \rightarrow \infty$ where $\tau$ is not an eigenvalue of \begin{align*}
		-\Delta u &= \lambda u \quad \text{on $\Omega$} \\
		u &= 0 \quad \text{on $\partial \Omega$}
	\end{align*}  We prove the problem \begin{align*}
		-\Delta u &= g(u) \quad \text{on $\Omega$} \\
		u &= 0 \quad \text{on $\partial \Omega$}
	\end{align*} has a \emph{weak} solution \[
		g(y) = \tau y + h(y) 
	\] where $\frac{h(y)}{y} \rightarrow 0$ as $|y| \rightarrow \infty$.

	Note that if such a solution exists, then we have \begin{align*}
		& &- \Delta u &= \tau u + h(u) \\
		&(\Rightarrow) &(-\Delta - \tau I) u &= h(u) \\
		&(\Rightarrow) &u &= \left(-\Delta - \tau I \right)^{-1} h(u) \equiv H(u).
 	\end{align*}
\end{exmp}

\begin{proof}
	For simplicity, assume $\tau = 0$.  We prove that for large $M$, $H$ maps the set $ Z = \{ u \in L^2(\Omega) \given \| u \|_2 \leq M \}$ into itself and is cc.

	If we do this then by the Schauder theorem, we can show that $H$ has a fixed point which is our solution.

	\textbf{Aside.}  Consider \begin{align*}
		-\Delta u &= f(x) \quad \text{on $\Omega$} \\
		u &= 0 \quad \text{on $\partial \Omega$}
	\end{align*}

	Then a weak solution satisfies $u \in \dot W^{1, 2}(\Omega)$ and \[
		\int_\Omega \grad u \cdot \grad \phi = \underbrace{\int_\Omega f \phi}_{\substack{\text{bounded linear functional} \\ \text{on $\dot W^{1,2}(\Omega)$ if $f \in L^2(\Omega)$}}} \quad \forall \phi \in \dot W^{1, 2}(\Omega).
	\]  Thus \[
		\iprod{u, \phi} = \iprod{F, \phi}
	\] and so our solution is $u = F$.

	  If $n \geq 3$, and if $f \in L^{\frac{2n}{n+2}}(\Omega)$ with $\Omega$ bounded, then it suffices to prove $\int_\Omega f \phi$ is a bounded linear functional on $\dot W^{1,2}(\Omega)$.  We have \begin{align*}
	  	\left| \int f \phi \right| \leq \| f \|_{\frac{2n}{n+2}} \| \phi \|_{\frac{2n}{n-2}} \quad \text{by H\"older} \\
	&\leq K \|  \|_{\frac{2n}{n+2}} \| \grad \phi \|_2 \quad \text{by Sobolev embedding}
	  \end{align*} and so \begin{align*}
	  \| \grad u \|_2^2 =	\int \left| \grad u \right|^2 \leq C \| f \|_{\frac{2n}{n+2}} \| \grad u \|_2
	  \end{align*} and hence \[
	  	\| \grad u \|_2 \leq C \| f \|_{\frac{2n}{n+2}}
	  \]

	\textbf{Proof of example.}  We now show that $H$ has the desired properties.  Let $\epsilon > 0$.  Then there exists $K > 0$ such that \[
		|h(y) | \leq \epsilon |y| + K
	\]  So we have \begin{align*}
		\| h(u) \|_2 	&\leq \| \epsilon|u| + K \|_2 \\
						&\leq \| \epsilon u \|_2 + \| K \|_2 \\
						&\leq \epsilon \| u \|_2 + K m(\Omega)^{1/2}. \tag{$\star$}
	\end{align*} Then we have \begin{align*}
		\| H(u) \|_2 	&= \| (-\Delta^{-1}) h(u) \| \\
		 				&\leq K_1 \| h(u) \|_2 \\
						&\leq K_1 \left( \epsilon \| u \|_2 + K m(\Omega)^{1/2} \right) \\
						&\leq \frac{1}{2} \| u \|_2 + \underbrace{K_2}_{= K_1 K m(\Omega)^{1/2}} \quad \text{letting $\epsilon = \frac{1}{2K_1}$}
	\end{align*}  Then $H$ maps the set $ Z = \{ u \in L^2(\Omega) \given \| u \|_2 \leq 2 K_2 \}$ into itself (that is, $H(Z) \subseteq Z$.)

	Secondly, the image under $H$ of this ball lies in a compact set in $L^2(\Omega)$.  It suffices to prove $H$ of this set lies in a bounded set in $\dot W^{1,2}(\Omega)$ and then use the result that the inclusion mapping $i : \dot W^{1, 2}(\Omega) \rightarrow L^2(\Omega)$ is compact.

	This is easy since $\{ h(u) \given u \in Z \}$ lies in a bounded set in $L^2(\Omega)$ by $(\star)$ and $(-\Delta)^{-1}$ maps bounded sets in $L^2(\Omega)$ to bounded sets in $\dot W^{1, 2}(\Omega)$.

	Finally, $H$ is continuos. We prove that the map $u \rightarrow h(u)$ is continuous and $L^2(\Omega) \rightarrow L^{\frac{2n}{n+2}}(\Omega)$.  This suffices since $H = (-\Delta)^{-1} \circ h$.

	Suppose that $u_n \rightarrow u$ in $L^2(\Omega)$.  As before, there exists $T$ a set such that $\Omega - T$ has small measure such that $u$ is bounded on $T$ and $u_n \rightarrow u$ uniformly on $T$.  Hence $h(u_n) \rightarrow h(u)$ uniformly on $T$ and so \[
		\int_T \left| h(u_n) - h(u) \right|^{\frac{2n}{n+2}} \rightarrow 0.
	\]  We now need only prove \begin{align*}
		&\int_{\Omega \, \backslash \, T} \left| h(u_m) - h(u) \right|^{\frac{2n}{n+2}} \quad \text{is small for large $m$} \\
		&= \| h(u_m) - h(u) \|_{\frac{2n}{n+2}, \Omega \, \backslash \, T} \\
		&\leq \| h(u_m) - h(u) \|_{2, \Omega \, \backslash \, T}^\alpha \left( \int_{\Omega \, \backslash \, T} 1 \right)^\beta \quad \text{by H\"older}
	\end{align*} where $\frac{1}{\alpha} + \frac{1}{\beta} = 1$.  We need only then bound \begin{align*}
		\| h(u_m) - h(u) \|_{2, \Omega \, \backslash \, T} &\leq \| h(u_m)\|_2  + \| h(u) \|_2 \\
		&\leq K_1 \quad \text{by ($\star$)}.
		\end{align*}

		This result can also be shown using the result that if $u \in L^1(\Omega)$, $\Omega$ bounded, then given $\epsilon > 0$ there exists $\delta > 0$ such that \[
			\int_A | u | \leq \epsilon
		\] if $m(A) \leq \delta$.
\end{proof}

Consider the equation \begin{align*}
	-\Delta u &= g(u, \grad u) \quad \text{on $\Omega$} \\
		u &= 0 \quad \text{on $\partial \Omega$.}
\end{align*}  This has a weak solution if $g$ is continuous and bounded on $\R \times \R^n$ and $\Omega$ is bounded (by Schauder).  It is possible to show that this equation is a mapping of \[
	\{ u \in \dot W^{1, 2}(\Omega) \given \| u \|_{1, 2} \leq K \}
\] into itself.  We need to show that this mapping is compact, as above.

\begin{lem}[Schauder]
	If $A$ is a Banach \begin{enumerate}[(i)]
		\item $A: E \times [0, 1] \rightarrow E$ is completely continuous, and 
		\item $A(x, 1) = L$ where $L$ is linear and $I - L$ is invertible, and 
		\item if $x = A(x, t)$ where $0 \leq t \leq 1$, then $\| x \| \leq M$,
	\end{enumerate}
	then the equation $x = A(x, 0)$ has a solution.
\end{lem} 
% section fixed_point_methods (end)
% section lecure_24 (end)
% section applications_to_nonlinear_equations (end)

\section{Other Types of Problems} % (fold)
\label{sec:other_types_of_problems}
If $\Omega$ is a bounded domain with smooth boundary, and consider the equation \begin{align*}
	\frac{\partial u}{\partial t} &= \Delta u \quad \text{on $\Omega$} \\
	u(x, t) &= 0 \quad \text{if $x \in \partial \Omega$}
\end{align*} with $u(x, 0) = u_0(x) \in L^2(\Omega)$ given.

Suppose $\phi_i$ are the weak eigenfunctions of $-\Delta$ for the Dirichlet Boundary condition $u(x, t) = 0$ for $x \in \partial \Omega$.  Then $\| \phi_i \|_2 = 1$ and they form a complete orthonormal basis for $L^2(\Omega)$.  Then we can write \[
	u(x, 0) = \sum_{i=1}^\infty c_i \phi_i(x)
\] where $\sum c_i^2 < \infty$.

The solution can be then be uniquely written as \[
	u(x, t) = \sum_{i=1}^\infty c_i e^{-\lambda_i t} \phi_i(x)
\]  We can trivially see that \begin{align*}
	\| u(x, t) - u(x, 0) \|_2 \rightarrow 0 
\end{align*} as \begin{align*}
	\| u(x, t) - u(x, 0) \|_2^2 &= \left\| \sum c_i \left( e^{-\lambda_i t} - 1 \right) \phi_i(x)  \right\|_2^2 \\
	&= \sum c_i^2 \left( e^{-\lambda_i t} - 1 \right)^2 \rightarrow 0.
\end{align*}  Note that $u_0 \in L^2$, but $u(x, t) \in C^\infty$ for all $t > 0$.

Consider now the differential equation \begin{align*}
	\frac{\partial u}{\partial t} &= -\Delta u \quad \text{on $\Omega$} \\
	u(x, t) &= 0 \quad \text{if $x \in \partial \Omega$}
\end{align*} for $t \geq 0$.  This is equivalent to running the heat equation backwards in time.  Formally, the solution is \[
	\sum c_i e^{\lambda_i t} \phi_i(x) 
\] for $t \geq 0$, which does not converge in $L^2$.

It can be shown that there is at most one solution.  This is an \emph{ill-posed} problem.

% section other_types_of_problems (end)

\section{Various Other Results} % (fold)
\label{sec:various_other_results}

% section various_other_results (end)
\begin{thm}
	Eigenfunctions of a compact self-adjoint operator form a complete set 
\end{thm}

\begin{thm}
	The inverse of the Laplacian is a compact, self-adjoint operator.
\end{thm}

Comments on the exam.

\begin{enumerate}[(i)]
	\item Asked some definitions.
	\item Asked some simple proofs.
	\item Asked some problem questions, possibly similar to assignments.
	\item Look at the assignments for questions.
\end{enumerate}

\end{document}

### 2011-12-12 STAT 3911 - Stochastic Processes and Time Series Analysis/STAT 3911 Lecture Notes.tex
Initial Commit

%
%  untitled
%
%  Created by Andrew Tulloch on 2009-11-04.
%  Copyright (c) 2009 __MyCompanyName__. All rights reserved.
%!TEX TS-program = xelatex
%!TEX encoding = UTF-8 Unicode

\documentclass[10pt, oneside, reqno]{amsart}
\usepackage{geometry, setspace, graphicx, enumerate}
\onehalfspacing                 
\usepackage{fontspec,xltxtra,xunicode}
\defaultfontfeatures{Mapping=tex-text}

%\setromanfont[Mapping=tex-text,Contextuals= 
%{NoWordInitial,NoWordFinal,NoLineInitial,NoLineFinal}]{Hoefler Text}
%\setsansfont[Scale=MatchLowercase,Mapping=tex-text]{Hoefler Text}
%\setmonofont[Scale=MatchLowercase]{Andale Mono}

	% AMS Theorems
	\theoremstyle{plain}% default 
	\newtheorem{thm}{Theorem}[section] 
	\newtheorem{lem}[thm]{Lemma} 
	\newtheorem{prop}[thm]{Proposition} 
	\newtheorem*{cor}{Corollary}

	\theoremstyle{definition} 
		\newtheorem{defn}[thm]{Definition}
		\newtheorem{conj}[thm]{Conjecture}
		\newtheorem{exmp}[thm]{Example}

	\theoremstyle{remark} 
		\newtheorem*{rem}{Remark} 
		\newtheorem*{note}{Note} 
		\newtheorem{case}{Case}

		\newcommand{\expc}[1]{\mathbb{E}\left[#1\right]}
		\newcommand{\var}[1]{\text{Var}\left(#1\right)}
		\newcommand{\cov}[1]{\text{Cov}\left(#1\right)}
		\newcommand{\prob}[1]{\mathbb{P}(#1)}
		\newcommand{\given}{ \, | \,}
		\newcommand{\us}{0 \leq u \leq s}
		\newcommand{\ts}[1]{\{ #1 \}}

\newcommand{\al}{\alpha}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{Q}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\F}{\mathbb{F}}
\newcommand{\Ga}{\mathbb{G}}

\newcommand{\aut}[1]{\text{Aut}{(#1)}}
\newcommand{\ol}{\overline}
\newcommand{\gener}[1]{\langle #1 \rangle}
\newcommand{\charr}[1]{\text{char}(#1)}
\newcommand{\nth}{n\textsuperscript{th}}

\newcommand{\tworow}[2]{\genfrac{}{}{0pt}{}{#1}{#2}}
\newcommand{\xdeg}[2]{[#1 : #2]}
\newcommand{\gal}[2]{\text{Gal}(#1/#2)}
\newcommand{\minpoly}[2]{m_{#1, #2}(x)}

\newcommand{\mapping}[5]{\begin{align*}
	#1 : \quad     #2 &\rightarrow #3 \\
			#4  &\mapsto #5
\end{align*}	
}

\title{STAT 3911 - Stochastic Processes}								% Document Title
\author{Andrew Tulloch}
%\date{}                                           % Activate to display a given date or no date

\begin{document}
\maketitle
\section{Stochastic Processes} % (fold)
\label{sec:stochastic_processes}
\begin{defn}[Markov process]
	A process $X = \{X_t : t \in T\}$ is a \textbf{Markov process} if 
	\[
		\prob{X(t) \leq x \, | \, X(t_1) = x_1, X(t_2) = x_2, \dots, X(t_n) = x_n} = \prob{X(t) \leq x \, | \, X(t_n) = x_n}
	\]
	whenever $t_1 < t_2 < \cdots < t_n < t$.

	In other words, a Markov process is a stochastic process satisfying the Markov property: the conditional distribution of future values is independent of the past.

\end{defn}

\begin{defn}[Markov chain]
	A \emph{Markov chain} is a stochastic process with a finite or countable state space is a Markov chain if it satisfies the Markov property.  Define the transition probabilities $P_{ij}$ by \[
		P_{ij} = \prob{X_{n+1} = j \, | \, X_n = i}
	\]
\end{defn}

\begin{thm}[Chapman-Kolmogorov equation]
	Let $P^n_{ij}$ denote the probability of going from state $i$ to state $j$ in $n$ steps, that is, \[
		P_{ij}^n = \prob{X_{n+m} = j \, | \, X_m = i} = \prob{X_n = j \, | \, X_0 = i}
	\]

	Then we have, for any $m,n = 0,1,2,\dots$ and $i,j \in S$, \[
		P_{ij}^{n+m} = \sum_{k=0}^\infty P^m_{ik} P^n_{kj}.
	\]
	In terms of matrix multiplication, this is equivalent to \[
		P^{n+m} = P^{n}P^{m}
	\]
\end{thm}

\begin{prop}[Initial distribution]
	At time $t$, let $\pi_n$ denote the column vector such that $\pi_n(j) = \prob{X_n = j}$ for all $j \in S$.
	\[
		\pi_n = \pi_0 P^n
	\]
\end{prop}

\section{Classification of states} % (fold)
\label{sub:classification_of_states}

\begin{defn}[Classification of states]
	Define an equivalence relation $\sim$ on the set of states such that $i \leftrightarrow j$ if and only if $P^n_{ij} > 0$ for some $n \geq 0$.  Two states $i$ and $j$ which are accessible to each other are said to communicate and we write $i \leftrightarrow j$
\end{defn}

\begin{lem}
	Communication is an equivalence relation on the set of states.
\end{lem}

If two states communicate with each other we say they belong to the same \textbf{communicating class}.

\begin{defn}[Irreducible]
	A Markov chain is said to be \textbf{irreducible} if there is only one class, i.e., all states communicate with each other.  A class $C$ is closed if no state outside $C$ can be reached from any state in $C$.  A state $i$ is called \textbf{absorbing} if the set $\{i \}$ is a closed class, i.e., $P_{ii} = 1$.
\end{defn}

\begin{defn}[Period of a state]
	A state $i$ is said to have \textbf{period} $d_i$ if $P^{n}_{ii} = 0$ except when $n = d_i, 2d_i, 3d_i, \dots$.
\end{defn}

\begin{prop}
	If $i \leftrightarrow j$ then $d_i = d_j$.  
\end{prop}

\begin{defn}[First transition time]
	For any states $i$ and $j$, define $f^n_{ij}$ as the probability that starting in $i$, the first transition into $j$ occurs at time $n$, that is \begin{align*}
		f^0_{ij} = 0 \\
		f^n_{ij} = \prob{X_n = j, X_k \neq j, k = 1,2,\dots, n-1 \, | \, X_0 = i}
	\end{align*}
Let $T_j$ be the random time of the \textbf{first transition} into state $j$.  Thus we have \[
	\prob{T_j = n \given X_0 = i} = f^n_{ij}
\]
If $X_0 = j$ then $T_j$ is the \textbf{first return} time to $j$.  If $X_0 = i \neq j$ then $T_j$ is the \textbf{first hitting} time of $j$.

We define the quantity $f_{ij}$ by the equality \[
	f_{ij} = \sum_{n=1}^\infty f^n_{ij}
\]

$f^{ij}$ is the probability of ever making a transition into state $j$, given that the process starts in state $i$.  We also have that $f_{ij} > 0$ if and only if state $j$ is accessible from state $i$.
\end{defn}

\begin{defn}[Recurrent states]
	A state is \textbf{recurrent} if \[
		\prob{X_n = i \text{ for some } n \geq 1 \given X_0 = i} = 1
	\]

	A state $i$ is recurrent if and only if $f_{ii} = 1$.  A state $i$ is transient if and only if $f_{ii} < 1$.
\end{defn}

\begin{prop}
	A state $i$ is recurrent if and only if \[
		\sum_{n=1}^\infty P^n_{ii} = \infty
	\]
A state $i$ is transient if and only if \[
	\sum_{n=1}^\infty P^n_{ii} < \infty
\]
\end{prop}

\begin{cor}
	If $i$ is recurrent and $i \leftrightarrow j$ then state $j$ is recurrent.
\end{cor}

\begin{cor}
	If $i \leftrightarrow j$ and $j$ is recurrent then $f_{ij} = 1$.
\end{cor}
% subsection classification_of_states (end)

\section{Limit theorems for Markov chains} % (fold)
\label{sub:limit_theorems_for_markov_chains}

Let $m_{jj}$ be the \textbf{mean recurrence time}, that is, the expected number of transitions until a Markov chain starting in $j$ returns to that state, so \[
	m_jj = \begin{cases}
		 \infty &\text{if $j$ is transient} \\
							\sum_{n=1}^\infty n f^n_{jj} &\text{if $j$ is recurrent}
	\end{cases}
\]

\begin{defn}[Ergodic state]
	Let $j$ be recurrent.  Then a state $j$ is called \textbf{positive recurrent} if $m_{jj} < \infty$ and \textbf{null recurrent} if $m_{jj} = \infty$.  A positive recurrent, aperiodic state is called \textbf{ergodic}.
\end{defn}

We have $\frac{1}{m_{jj}} = 0$ if a state $j$ is either transient of null recurrent.  If $j$ is positive recurrent then $0 < \frac{1}{m_{jj}} < \infty$.

\begin{thm}
	Assume $i \leftrightarrow j$. If a state $j$ is aperiodic, then \[
		\lim_{n \rightarrow \infty} P^n_{ij} = \frac{1}{m_{jj}}
	\]

	If a state $j$ has period $d_j$, then \[
			\lim_{n \rightarrow \infty} P^n_{ij} = \frac{d_j}{m_{jj}}
	\]
\end{thm}

% subsection limit_theorems_for_markov_chains (end)

\begin{defn}[Stationary distributions]
	A probability distribution $\pi^\star$ is called \textbf{stationary} for for a Markov chain $X$ with transition matrix $P$ if \[
		\pi^\star P = \pi^\star
	\]
We note that the stationary distribution $\pi^\star$ associated with a Markov chain is not necessarily unique, nor does it always exist.
\end{defn}

\begin{thm}[Irreducible aperiodic Markov chains]
		Let $X$ be an irreducible aperiodic Markov chain.  Then
		\begin{itemize}
			\item Either all the states are transient or null recurrent; in that case we have that $\lim_{n \rightarrow \infty} P^n_{ij} = 0$ for all $i,j$ and there exists no stationary distribution.
			\item or else all states are positive recurrent - in that case we have $\Pi_j = \lim_{n \rightarrow \infty} P^n_{ij} > 0$ for all $i,j$; moreover, $\Pi_j$ is a stationary distribution and the stationary distribution is unique.  Moreover, for all $j$, \[
				\Pi_j = \frac{1}{m_{jj}}
			\]
		\end{itemize}

\end{thm}
	We note that in a \textbf{finite-state}, irreducible, aperiodic Markov chain all states are positive recurrent/ergodic, and so the stationary distribution exists and is unique.

In a \textbf{finite state}, irreducible, \textbf{periodic} Markov chain all states are positive recurrent and the unique stationary distribution exists and is unique.

\section{First step analysis} % (fold)
\label{sub:first_step_analysis}
Consider the Markov chain $X$ with transition probability matrix 
\[
	P = \begin{bmatrix}
			1 & 0 & 0 \\
			\alpha & \beta & \gamma \\
			0 & 0 & 1
	\end{bmatrix}	
\]

We ask the following questions.
\begin{itemize}
	\item What state will the process ultimately be absorbed in?
	\item How long, on average, will it take to reach one of these states?
\end{itemize}

Let \[T = \min \{n \geq 0 \given X_n = 0 \text{ or } X_n = 2 \}\]
be the random \textbf{time of absorption} for the process $X$.  We ask the questions \[
	u = \prob{X_T = 0 \given X_0 = 1} 
\] and \[
	v = \expc{T \given X_0 = 1}
\]

Thus $u$ is the probability of begin absorbed in state $0$ given we start in state $1$, and $v$ is the expected time of being absorbed given we start in state $1$.

Now we have the following.\[
	u = 1\cdot \alpha + u \cdot \beta + 0 \cdot \gamma
\]

Similarly, we have 
\[
	v = 1 + a \cdot 0 + \beta \cdot v + \gamma \cdot 0
\]
% subsection first_step_analysis (end)

\section{Branching processes} % (fold)
\label{sub:branching_processes}

\begin{defn}[Branching process]
	let $X_0 = 1$ and for every $n = 0,1,2,\dots$ define\[
		X_{n+1} = \sum_{i=1}^{X_n} Z_i
	\]
	where $Z_i$ are independently, identically distributed random variables with probability distribution $\prob{Z=k} = P_k, k = 0,1,2,\dots$.  Then $X = \{ X_n, n \geq 0 \}$ is a branching process
\end{defn}

We have the following formulas.

\begin{lem}
	We have \[
		\expc{E(X_n)} = \mu \expc{X_{n-1}} = \mu^n
 	\]
and \[
	\var{X_n} = \begin{cases}
		\sigma^2 \mu^{n-1} \left( \frac{\mu^n - 1}{\mu - 1}\right) &\mu \neq 1 \\
		n \sigma^2 &\mu = 1
	\end{cases}
\]
\end{lem}

\begin{prop}[Extinction probabilities]
	Define $u_n$ to be the probability of extinction at or prior to the $\nth$ generation, assuming $X_0 = 1$.  Then \[
		u_n = \prob{X_n = 0 \given X_0 = 1}
	\]

Then we have \[
	u_n = \sum_{k=0}^\infty P_k (u_{n-1})^k
\] where $P_k = \prob{Z = k}$.
\end{prop}

\begin{prop}
	Let $\Pi_0 = \lim_{n \rightarrow \infty} \prob{X_n = 0}$ be the probability of eventual extinction.  Then, definining the probability generating functions $G_n(s) = \expc{s^{X_n}} = \sum_{j=0}^\infty s^j \prob{X_n = j}$, we have that 
	\begin{itemize}
		\item $\Pi_0 = 1$ if and only if $\mu \leq 1$,
		\item If $\mu > 1$ then $\Pi_0 < 1$ is the smallest positive number $p$ satisfying $G(p) = p$ - that is, \[
			\Pi_0 = G(\Pi_0)
		\] where $G(p) = \sum_{j=0}^\infty p^j P_j$.
	\end{itemize}
\end{prop}
% subsection branching_processes (end)

\section{Poisson process} % (fold)
\label{sub:poisson_process}
\begin{defn}[Poisson process]
	A process $N = \{N(t), t \geq 0\}$ is a \textbf{Poisson process} if 
	\begin{itemize}
		\item $N(0) = 0$,
		\item $N$ is a process of independent increments.
		\item The number of events in any interval of length $t$ has a Poisson distribution with parameter $\lambda t$, that is \[
			\prob{N(t+s) - N(s) = n}  = \frac{e^{-\lambda t} (\lambda t)^n}{n!}
		\]
	\end{itemize}

	The positive paramenter $\lambda > 0$ is called the \textbf{intensity} of the process.  We have that $\expc{N(t)} = \lambda t$ and $\var{N(t)} = \lambda t$
\end{defn}

\begin{lem}
	The sum of two Poisson processes of intensities $\lambda$ and $\mu$ is a Poisson process of intensity $\lambda + \mu$.
\end{lem}

% subsection poisson_process (end)

\section{Sojourn and waiting times} % (fold)
\label{sub:sojourn_and_waiting_times}

\begin{defn}[Soujourn time]
	For $n = 0,1,2,\dots$, the \textbf{sojourn time} in state $n$ equals \[
		S_n = \inf \{t \in [0, \infty) \given N(t + S_{n-1}) - N(S_{n-1}) = 1 \}
	\]
\end{defn}

\begin{lem}
	The sojourn times $S_n$ are independently, identically distributed exponential variables with parameter $\lambda$.
\end{lem}

\begin{defn}[Waiting time]
	For $n = 0,1,2,\dots$, the \textbf{waiting time} until the $\nth$ event equals \[
		W_n = \inf \{ t \in [0,\infty) \given N(t) = n\}
	\]
\end{defn}

\begin{lem}
	If $X_i$ are independently, identically distributed exponential random variables with parameter $\lambda$ then $\sum_{i = 1}^n X_i$ is distributed with a gamma distribution with parameters $n, \lambda$, that is, $\Gamma(n,\lambda)$.
\end{lem}

\begin{prop}
	Given $N(t) = 1$, the waiting time $W_1 = S_0 = T_1$ has uniform distribution on $[0,t]$
\end{prop}

\begin{thm}
	Given that $N(t) = 1$, the waiting times $(W_1, \dots, W_n)$ have the same conditional joint probability density of $n$ independently, identically distributed uniformly distributed random variables on the interval $[0,t]$.
\end{thm}

\begin{lem}
	If $X_i, i = 1,2$ are independent exponential random variables with parameters $\lambda_i$, then \[
		\prob{X_1 < X_2} = \frac{\lambda_1}{\lambda_1 + \lambda_2}
	\]
\end{lem}
% subsection sojourn_and_waiting_times (end)

\section{Continuous-time Markov chains} % (fold)
\label{sub:continuous_time_markov_chains}
Consider a continuous time Markov process $X = \{X(t), t \geq 0 \}$ where transition probabilities \[
	P_{ij}(t) = \prob{X(t+s) = j \given X(s) = i}
\] are independent of $s$.  This is a \textbf{time-homogenous}, or \textbf{stationary}, process.

\begin{defn}[Pure birth process]
	A \textbf{pure birth process} is a Markov process satisfying the following conditions.
	\begin{itemize}
		\item $P_{k,k+1}(h) = \lambda_k h + o(h)$,
		\item $P_{k,k}(h) = 1 - \lambda_k h + o(h)$,
		\item $\prob{X(t+h) - X(t) < 0 \given X(t) = k} = 0$,
		\item $X(0) = 0$ or $X(0) = 1$
	\end{itemize}
\end{defn}

\begin{exmp}
	Let us review some interesting special cases for birth intensities $\lambda_k$.

	\begin{itemize}
		\item \textbf{Poisson process:} $\lambda_k = \lambda$ for all $k$.
		\item \textbf{Simple birth:} $\lambda_k = k \beta$.  This model gives the probabiities \[
			\prob{X(t+h) - X(t) = m \given X(t) =k} = \begin{cases}
				1 - k\beta h + o(h) & m = 0 \\
				k\beta h + o(h) &m=1\\
				o(h) &\text{otherwise}
			\end{cases}
		\]
		\item \textbf{Simple birth with immigration:} $\lambda_k = k \beta + \nu$.  
	\end{itemize}
\end{exmp}

% subsection continuous_time_markov_chains (end)

\section{Differential equations for marginal probabilities} % (fold)
\label{sub:differential_equations_for_marginal_probabilities}

Consider a pure birth process.  Let $P_n(t) = \prob{X(t) = n}$.  Assume $P_0(0) = 1$.
\begin{thm}
	The functions $P_n$ satisfy the following system of ordinary differential equations.
	\begin{align*}
		P'_0(t) &= -\lambda_0 P_0(t) \\
		P'_n(t) &= -\lambda_n P_n(t) + \lambda_{n-1} P_{n-1}(t)
	\end{align*}
\end{thm}

\begin{lem}
	The marginal probability functions $P_n$ satisfy \[
		P_0(t) = e^{-\lambda_0 t}
	\] and \[
		P_n(t) = \lambda_{n-1} e^{-\lambda_n t} \int_0^t e^{\lambda_n s} P_{n-1}(s) ds
	\]
\end{lem}

As above, define $S_k$ to be the time between the $k$th and $(k+1)$th birth.  Then we have the following lemma.

\begin{lem}
	The random variables $S_k$ are independent and $S_k$ has an exponential distribution with parameter $\lambda_k$
\end{lem}
% subsection differential_equations_for_marginal_probabilities (end)

\section{Birth and death processes} % (fold)
\label{sub:birth_and_death_processes}
\begin{defn}[Pure birth process]
	A \textbf{birth and death process} is a Markov process satisfying the following conditions.
	\begin{itemize}
		\item $P_{k,k+1}(h) = \lambda_k h + o(h)$,
		\item $P_{k,k}(h) = 1 - (\lambda_k + \mu_k h + o(h)$,
		\item $P_{k,k-1}(h) \mu_k h + o(h)$
		\item $\prob{X(t+h) - X(t) < 0 \given X(t) = k} = 0$,
		\item $X(0) = 0$ or $X(0) = 1$
	\end{itemize}
\end{defn}

\begin{prop}[Chapman-Kolmogorov equation]
	\[
		P_{ij}(t+s) = \sum_{k=0}^\infty P_{ij}(t) P_{kj}(s).
	\]
\end{prop}

\begin{prop}[Sojourn times]
	We wish to calculate the \textbf{sohourn time} in state $i$, denoted $S_i$.  Defining $\prob{S_i \geq t} = D_i(t)$. We have \[
		D_i(t) = e^{-(\lambda_i + \mu_i)t}
	\]

	Thus $S_i$ is exponential distributed with parameter $\lambda_i + \mu_i$.
\end{prop}

\begin{prop}
	The process stays in a given state $i$ for a random length of time whose distribution function is exponential with parameter $\lambda_i + \mu_i$.  The process leaves state $i$ by entering state $i+1$ or $i -1 $ with probability $\frac{\lambda_i}{\lambda_i + \mu_i}$ and $\frac{\mu_i}{\lambda_i + \mu_i}$.
\end{prop}

\begin{prop}
	Transition probabilities $P_{ij}(t)$ satisfy a system of ordinary differential equations, known as the \textbf{backward Kolmogorov equations.}  For a fixed $j$, 
	\begin{align*}
		P'_{0j}(t) &= -\lambda_0 P_{0j}(t) + \lambda_0 P_{1j}(t) \\
		P'_{ij}(t) &= \mu_i P_{i-1,j}(t) - (\lambda_i + \mu_i)P_{ij}(t) + \lambda_i P_{i+1,j}(t)
	\end{align*} with initial conditions $P_{ij}(0) = \delta_{ij}$.  In matrix notation, \[
		P'(t)  = AP(t),
	\] with $P(0) = I$.
\end{prop}

\begin{prop}
	Transition probabilities $P_{ij}(t)$ satisfy a system of ordinary differential equations, known as the \textbf{forward Kolmogorov equations.}  For a fixed $j$, 
	\begin{align*}
		P'_{i0}(t) &= -\lambda_0 P_{i,0}(t) + \mu_1 P_{i,1}(t) \\
		P'_{ij}(t) &= \lambda_{j-1} P_{i,j-1}(t) - (\lambda_i + \mu_i)P_{ij}(t) + \mu_{j+1} P_{i,j+1}(t)
	\end{align*} with initial conditions $P_{ij}(0) = \delta_{ij}$.  In matrix notation, \[
		P'(t)  = P(t)A,
	\] with $P(0) = I$.
\end{prop}

From the forward Kolmogorov equation, we can find a system of ordinary differential equations for the marginal probabilities $P_k = \prob{X(t) = j}$.  We have \begin{align*}
	P'_0(t) &= - \lambda_0 P_0(t) + \mu_1 P_1(t) \\
	P'_j(t) &= \lambda_{j-1}P_{j-1}(t) - (\lambda_j + \mu_j) P_j(t) + \mu_{j+1}P_{j+1}(t)
\end{align*} with $P_j(0) = \pi_0(j)$.  In matrix notation, \[
	\pi'_t = \pi_t A
\]
% subsection birth_and_death_processes (end)
% section stochastic_processes (end)
\clearpage

\section{Time Series Analysis} % (fold)
\label{sec:time_series_analysis}

\begin{defn}[Time series]
	Let $\ts{X_t}$ be a time series with $\expc{X_t^2} < \infty$.  The \textbf{mean function} for $\ts{X_t}$ is \[
		\mu_X(t) = \expc{X_t}
	\]
	The \textbf{covariance function} of $\ts{X_t}$ is \[
		\gamma_X(r,s) = \cov(X_r, X_s) = \expc{(X_r - \mu_X(r)(X_s - \mu_X(s)))}
	\]
\end{defn}

\begin{defn}[Stationary]
	A time series $\ts{X_t}$ is \textbf{weakly stationary} if 
	\begin{itemize}
		\item $\mu_X(t)$ is independent of $t$,
		\item $\gamma_X(t+h,t)$ is independent of $t$ for each $h$.
	\end{itemize}
\end{defn}

\begin{defn}[Autocovariance function]
	Let $\ts{X_t}$ be a stationary time series.  The \textbf{autocovariance function} (ACVF) of $\ts{X_t}$ at lag $h$ is \[
		\gamma_X(t) = \cov{X_{t+h}, X_t}
	\]
	The \textbf{autocorrelation function} (ACF) of $\ts{X_t}$ at lag $h$ is \[
		\rho_X(h) = \frac{\gamma_X(h)}{\gamma_X(0)} = \text{Cor}({X_{t+h}, X_t})
	\]
\end{defn}

\begin{defn}[Sample autocovariance function]
	The \textbf{sample autocovariance function} is \[
		\hat{\gamma}(h) = \frac{1}{n} \sum_{t=1}^{n-|h|} (x_{t+|h|} - \bar{x})(x_t - \bar{x})
	\]
	The \textbf{sample autocorrelation function} is \[
			\ol{\rho}_X(h) = \frac{\ol{\gamma}_X(h)}{\ol{\gamma}_X(0)}
	\]
\end{defn}

\section{ARMA($p,q$) Models} % (fold)
\label{sec:arma_model}

\begin{defn}[ARMA$(p,q)$ process]
	A time series $\ts{X_t}$ is an \textbf{ARMA($p,q$)} process if $\ts{X_t}$ is stationary and if for every $t$,\[
		X_t - \phi_1 X_{t-1} - \cdots - \phi_p X_{t-p} = Z_t + \theta_1 Z_t + \cdots + \theta_q Z_{t-q}
	\] where $\ts{Z_t} \sim WN(0,\sigma^2)$ and the polynomials $\theta(z)$ and $\phi(z)$ have no common factors.
\end{defn}

\begin{thm}[Stationary]
	A stationary solution $\ts{X_t}$ of the equations exists, and is the unique stationary solution if and only if \[
		\phi(z) = 1 - \phi_1 z - \cdots - \phi_p z^p \neq 0
	\] for all $|z| = 1$
\end{thm}

\begin{thm}[Causality]
	An ARMA($p,q$) process is \textbf{causal}, if there exists constants $\ts{\psi_j}$ with \[
		X_t = \sum_{j=0}^\infty \psi_j Z_{t-j}
	\] for all $t$.

	Causality is equivalent to the condition \[
		\phi(z) = \phi(z) = 1 - \phi_1 z - \cdots - \phi_p z^p \neq 0
	\] for all $|z| \leq 1$.
\end{thm}

The coefficients $\psi_j$ can be determined by the identity \[
	\psi(z) = \psi_0 + \psi_1 z + \dots = \frac{\theta(z)}{\phi(z)}
	\]

\begin{defn}[Invertibility]
	An ARMA($p,q$) process $\ts{X_t}$ is \textbf{invertible} if there exist constants $\pi_j$ such that \[
		Z_t = \sum_{j=0}^\infty \pi_j X_{t-j}
	\]
	Invertibility is equivalent to the condition \[
		\theta(z) = 1 + \theta_1 z + \cdots + \theta_q z^q
	\] for all $|z| \leq 1$.
\end{defn}

The coeffients $\pi_j$ can be determined by the identity \[
	\pi(z) = \pi_0 + \pi_1 z + \dots = \frac{\phi(z)}{\theta(z)}
	\]

\begin{defn}[Partial autocorrelation function]
	The \textbf{partial autocorrelation function} $\alpha(\cdot)$ satisfies the following equations.
	\begin{align*}
		\alpha(0) &= 1\\
		\alpha(1) &= \rho(1) \\
		\alpha(2) &= \frac{\rho(2) - \rho(1)^2}{1 - \rho(1)^2}
	\end{align*}
	and in general, is the last component of \[
		\phi_h  =\Gamma_h^{-1} \gamma_y
	\] where $\Gamma_h$ is the covariance matrix and $\gamma_y$ is a vector of covariances. 
\end{defn}

\begin{thm}
	The PACF of an AR($p$) process is given by \[
		\alpha(p) = \phi_p
	\]
	and \[
		\alpha(h) = 0
	\] for $h > p$
\end{thm}

\begin{thm}[Stationary AR($2$) process]
	An AR($2$) process $X_t = \alpha_1 X_{t-1} + \alpha_2 X_{t-2} + Z_t$ is stationary if and only if the following all hold.
	\begin{align*}
		\alpha_1 + \alpha_2 < 1 \\
		\alpha_1 - \alpha_2 > -1 \\
		\alpha_2 > -1
	\end{align*}
\end{thm}
% section arma_model (end)

\section{ARIMA Models} % (fold)
\label{sec:arima_models}

\begin{defn}[ARIMA process]
	If $d$ is a nonnegative integer, then $\{X_t\}$ is an \textbf{ARIMA$(p,d,q)$ process} if $Y_t = (1-B)^d X_t$ is a causal ARMA($p,q$) process.
\end{defn}

% section arima_models (end)

\section{Model selection} % (fold)
\label{sec:model_selection}
	To select a model, we must first transform the data until we are examining a stationary process. 
	Then, we consider the ACF and PACF functions.  A few general points.
	\begin{itemize}
		\item A slowly decaying (slower than exponential decay) ACF indicates the data needs to be differenced, i.e., that an ARIMA($p,d,q$) model should be considered.
		\item A PACF that cuts out sharply after lag $p$ indicates an AR($p$) model.  This can be quantified by Quenoille's Test.  \begin{thm}[Quenouille's Test]
			If a process is AR($m$), and if $N$ is large, then \[
				\hat \pi_k \sim \mathcal{N}(0, \frac{1}{N}), \quad k > m
			\]
		Thus any $\hat \pi_k$ satisfying $|\hat \pi_k| > \frac{2}{\sqrt{N}}$ for $k > m$, indicates disagreement with $H_0$ at the $5\%$ significance level.
		\end{thm}
		\item Any MA process must have $\rho(1) < 0.5$
		\item An MA($q$) process must have $\rho(k) = 0$ for all $k > q$.
		\item An AR($p$) process will have the ACF function exhibiting exponential decay.
		\end{itemize}

To test the fit of a model, we use the Box-Pierce test and consider the AIC statistic.

\begin{thm}[Box-Pierce test]
	Assume an ARIMA$(p,d,q)$ model has been fitted to time series.  Consider the series of residual $\ts{s_t}$.  Then, for a fixed maximum lag $K$, we consider the statistic \[
		BP = N \sum_{k = 1}^K \hat r_k^2	\] where $\hat r_k^2$ is the squared sample autocorrelation coefficient for the residual time series.

This is distributed with according to a $\chi^2$-distribution with $K - p - q$ degrees of freedom.  
This statistic is then used to test the hypothesis that the none of the autocorrelation coefficients for the residuals are different from zero against the alternate that at least one autocorrelation coefficient is greater than zero.
\end{thm}

\begin{thm}[AIC]
	The AIC statistic for a fitted model is a sum of the likelihood (as estimate of model fit) and the number of parameters ($p$ and $q$).  The \emph{best} model is the one that minimizes the value of the AIC.
\end{thm}
% section model_selection (end)

\section{Prediction using ARIMA($p,d,q$) models} % (fold)
\label{sec:prediction_using_arima_p_d_q_models}

\begin{defn}[Best linear predictor]
	The best linear predictor of $Y$ in terms of $X_i, i = 1,2,\dots, n$ in the least squares sense, is the linear function of $\mathbf{X}$\[
		\lambda(\mathbf{X}) = \beta_0 + \sum_{i = 1}^n \beta_i X_i
	\] where $\beta_i, i = 0,1,2,\dots,n$ are chosen to minimise \[
		\expc{\left(Y - \beta_0 - \sum_{i=1}^n \beta_i X_i\right)^2}
	\]
\end{defn}

\begin{thm}[Properties of the best linear predictor]
	If $\hat \lambda(\mathbf{X}) = \hat \beta_0 + \sum_{i=1}^n \hat \beta_i X_i$ is the best linear predictor for $Y$ in terms of $X_i, i = 1,2,\dots,n$, the the residual \[
		U = Y - \hat \lambda(\mathbf{X})
	\] has expectation 0 and is uncorrelated with \textbf{every} linear function $\lambda(\mathbf{X})$.
\end{thm}

\begin{thm}[Uniqueness of the best linear predictor]
	Let $\hat \lambda(\mathbf{X}) = \hat \beta_0 + \sum_{i=1}^N \hat \beta_i X_i $ be the best linear predictor, and let $U  = Y - \hat \lambda(\mathbf{X})$.

	Suppose $\lambda^\star(\mathbf{X}) = \beta_0^\star + \sum_{i=1}^N \beta_i^\star X_i$ and $U^\star = Y - \lambda^\star (\mathbf{X})$ has the properties:
	\begin{enumerate}
		\item $\expc{U^\star} = 0$
		\item $U^\star$ is uncorrelated with every linear function of $X_1, X_2, \dots X_n$.  Then
		\[
			\lambda^\star(\mathbf{X}) = \hat \lambda(\mathbf{X})
		\]
	\end{enumerate}
\end{thm}

\begin{exmp}[Best linear predictor for AR($p$)]
	Let $\ts{X_t}$ be an AR$(p)$ process.  Then we have \[
	 X_{t-1}(1) = \mu +	\alpha_1 X_{t-1} + \cdots + \alpha_p X_{t-p}
	\]
	is the \textbf{best linear predictor} for $X_t$ in terms of $X_{t-1}, \dots, X_{t-p}$
\end{exmp}

\begin{exmp}[Best linear predictor for ARMA($p,q$)]
	Let $\ts{X_t}$ be an ARIMA($p,q$) process.  Then we have \[
		X_{t-1}(1) = \mu + \alpha_1 X_{t-1} + \cdots + \alpha_p X_{t-p} + \beta_1 Z_{t-1} + \cdots + \beta_q Z_{t-q}
	\]

	As we have \[
		X_{N+1} - X_N(1) = Z_t \sim \mathcal{N}(0,\sigma_Z^2),
	\]
	we then have a $(1- \alpha)100 \%$ confidence interval for $X_{N+1}$ is given by \[
		X_N(1) \pm \Phi\left(1 - \frac{\alpha}{2}\right) \sigma_Z
	\]
\end{exmp}

\begin{exmp}[Two step predictors for an AR($p$) process]
	We have the best linear predictor for $X_{t-2}(2)$ is given as \[
		X_{t-2}(2) = \mu +  \alpha_1 X_{t-2}(1) + \alpha_2 X_{t-2} + \cdots + \alpha_p X_{t-p}
	\]
	where $X_{t-2}(1)$ is the one step predicted value for the AR($p$) process above.

We can calculate that the $(1- \alpha)100\%$ confidence intervals for $X_{t}$ are given by \[
	X_{t-2}(2) \pm \Phi\left(1 - \frac{\alpha}{2}\right) \sqrt{(1 + \alpha_1^2) \sigma_Z^2}
\]
\end{exmp}
% section prediction_using_arima_p_d_q_models (end)
% section time_series_analysis (end)

\end{document}

### 2011-12-16 MATH 3964 - Complex Analysis/MATH 3964 Lecture Notes.tex
Added hyperref package to generate links within PDFs
fixed rakefile

@@
 \newcommand{\R}{\mathbb{R}}
-\newcommand{\C}{\mathbb{C}}
+\newcommand{\Com}{\mathbb{C}}
 \newcommand{\Z}{\mathbb{Z}}
@@
 \newcommand{\nth}{n\textsuperscript{th}}
+
+\usepackage{hyperref}
 
@@
 Thus the derivative $f'(z)$ is a function defined on $D$.
-a function is analytic at a particular point $z_0 \in \C$ if it is differentiable on some open neighbourhood of $z_0$.
+a function is analytic at a particular point $z_0 \in \Com$ if it is differentiable on some open neighbourhood of $z_0$.
 \end{defn}
@@
 	Suppose that $f(z)$ is analytic on an unbounded set and let $g(z) = f(1/z)$.
-Then the point $\infty \in \C^{\star}$ is  point of analyticity of $f(z)$ if $z=0$ is a point of analyticity of $g(z)$.
+Then the point $\infty \in \Com^{\star}$ is  point of analyticity of $f(z)$ if $z=0$ is a point of analyticity of $g(z)$.
 Similarly, $z = \infty$ is a singularity of a particular type of $f(z)$ if $g(z)$ has a singularity of that same type at $z=0$.
@@
 \begin{defn}
-	A function is \textbf{meromorphic} if it is analytic in the whole complex plane $\C$ except for poles.  
+	A function is \textbf{meromorphic} if it is analytic in the whole complex plane $\Com$ except for poles.  
 \end{defn}
@@
 	\] where $f$ is a locally analytic function of its $n$ complex arguments.  A singular point of the differential equation is a point \[
-		(z_0, w_0, w'_0, \dots, w_0^{(n-1)}) \in \C^n
+		(z_0, w_0, w'_0, \dots, w_0^{(n-1)}) \in \Com^n
 	\]
@@
 	F(\alpha, \beta; \gamma; z) = \frac{\Gamma(\gamma)}{\Gamma(\gamma - \beta)\Gamma(\beta)} \int_0^1 t^{\beta - 1} (1-t)^{\gamma - \beta - 1} (1- zt)^{- \alpha} \,dt
-\] valid for $\Re \gamma > \Re \beta > 0$, $|z| < 1$. but can be extended to $z \in \C = [1,\infty)$.
+\] valid for $\Re \gamma > \Re \beta > 0$, $|z| < 1$. but can be extended to $z \in \Com = [1,\infty)$.
 
@@
 \begin{thm}
-	The transcendental equation $f(z) = a$ where $f(z)$ is an elliptic function of order $m$ has exactly $m$ roots in every period parallelogram, counting multiplicities, for every $a \in \C$.  
+	The transcendental equation $f(z) = a$ where $f(z)$ is an elliptic function of order $m$ has exactly $m$ roots in every period parallelogram, counting multiplicities, for every $a \in \Com$.  
 \end{thm}

### 2011-12-16 MATH 3969 - Measure Theory and Fourier Analysis/MATH 3969 Lecture Notes.tex
Added hyperref package to generate links within PDFs
fixed rakefile

@@
 }
+
+\usepackage{hyperref}
 
@@
 \begin{defn}[Translation of a function]
- 	Let $f: \R^N \rightarrow \C$ be a function and $t \in \R^N$ a fixed vector. We define the translation operator $\tau_t$ by \[
+ 	Let $f: \R^N \rightarrow \Com$ be a function and $t \in \R^N$ a fixed vector. We define the translation operator $\tau_t$ by \[
  		\tau_t f(x) = f(x-t)
@@
 \begin{lem}
-	Let $f : \R^N \rightarrow \C$ be measurable and set 
+	Let $f : \R^N \rightarrow \Com$ be measurable and set 
 	\begin{align*}
@@
 
-	Then $F_1, F_2 : \R^N \times \R^N \rightarrow \C$ are measurable.
+	Then $F_1, F_2 : \R^N \times \R^N \rightarrow \Com$ are measurable.
 \end{lem}
@@
 \begin{defn}[Convolution]
-	Let $f,g : \R^N \rightarrow \C$ be measurable.  We define the \textbf{convolution} $f \star g : \R^N \rightarrow \C$ by \[
+	Let $f,g : \R^N \rightarrow \Com$ be measurable.  We define the \textbf{convolution} $f \star g : \R^N \rightarrow \Com$ by \[
 		(f \star g) (x) = \int_{\R^N} f(x-y) g(y) \, dy
@@
 \begin{thm}[Young's inequality]
-	Let $1 \leq p \leq \infty$.  If $f \in L^p(\R^N, \C)$ and $g \in L^1(\R^N, \C)$, then $f \star g$ exists almost everywhere and $f \star g \in L^p(\R^N, \C)$.  Moverover, \[
+	Let $1 \leq p \leq \infty$.  If $f \in L^p(\R^N, \Com)$ and $g \in L^1(\R^N, \Com)$, then $f \star g$ exists almost everywhere and $f \star g \in L^p(\R^N, \Com)$.  Moverover, \[
 		\|f \star g \|_p \leq \|f\|_p \|g \|_1
@@
 \begin{defn}[Fourier transform]
-	Let $f \in L^1(\R^N, \C)$.  We call \[
+	Let $f \in L^1(\R^N, \Com)$.  We call \[
 		\hat{f}(t) = \int_{\R^N} f(x) e^{-2 \pi i x \cdot t} \, dx
@@
 	\begin{itemize}
-		\item $\hat{f} : \R^N \rightarrow \C$ is continuous,
+		\item $\hat{f} : \R^N \rightarrow \Com$ is continuous,
 		\item $\|\hat f \|_\infty \leq \|f \|_1$
@@
 \begin{prop}
-	Let $f \in L^1(\R^N, \C), x_0 \in \R^N$ and $\alpha \in R, \alpha > 0$. 
+	Let $f \in L^1(\R^N, \Com), x_0 \in \R^N$ and $\alpha \in R, \alpha > 0$. 
 	\begin{enumerate}[(i)]
@@
 \begin{thm}[Riemann-Lebesgue]
-	If $f \in L^1(\R^N, \C)$, then $\hat f \in C_0(\R^N, \C)$ and $\| \hat f \|_\infty \leq \|f \|_1$.
-\end{thm}
-
-\begin{thm}
-	If $f,g \in L^1(\R^N, \C)$ then $f \star g \in L^1(\R^N, \C)$ and \[
+	If $f \in L^1(\R^N, \Com)$, then $\hat f \in C_0(\R^N, \Com)$ and $\| \hat f \|_\infty \leq \|f \|_1$.
+\end{thm}
+
+\begin{thm}
+	If $f,g \in L^1(\R^N, \Com)$ then $f \star g \in L^1(\R^N, \Com)$ and \[
 		\hat{f \star g} = \hat f \, \hat g
@@
 \begin{prop}
-	Let $f,g \in L^1(\R^N, \C)$.  Then \[
+	Let $f,g \in L^1(\R^N, \Com)$.  Then \[
 		\int_{\R^N} \hat f g \, dx = \int_{\R^N} f \hat g \, dx
@@
 	\]
-	for all $f \in L^1(\R^N,\C), x \in \R^N$, and $n \in \mathbb{N}$.
+	for all $f \in L^1(\R^N,\Com), x \in \R^N$, and $n \in \mathbb{N}$.
 \end{lem}
@@
 \begin{thm}[Fourier inversion formula]
-	Let $f \in L^1(\R^N, \C)$. Then 
+	Let $f \in L^1(\R^N, \Com)$. Then 
 	\begin{enumerate}[(i)]
@@
 			\lim_{n \rightarrow \infty} \int_{\R^N} \hat f (t) e^{2 \pi i x \cdot t} e^{- \pi \frac{|t|^2}{n^2}} \, dt = f
-		\] in $L^1(\R^N, \C)$.
+		\] in $L^1(\R^N, \Com)$.
 		\item If $f$ is continuous at $x$, then \[
@@
 \begin{thm}
-	Let $f \in L^2(\R^N, \C)$.  Then \[
+	Let $f \in L^2(\R^N, \Com)$.  Then \[
 		\lim_{n \rightarrow \infty} \int_{\R^N} \hat f (t) e^{2 \pi i x \cdot t} e^{- \pi \frac{|t|^2}{n^2}} \, dt = f
-	\] in $L^2(\R^N, \C)$.
+	\] in $L^2(\R^N, \Com)$.
 \end{thm}

### 2011-12-16 PMH3 - Functional Analysis/PMH3 Lecture Notes.tex
Added hyperref package to generate links within PDFs
fixed rakefile

@@
 \newcommand{\N}{\mathbb{N}}
-\renewcommand{\H}{\mathcal{H}}
-
-\newcommand{\C}{\mathbb{C}}
+\newcommand{\Hil}{\mathcal{H}}
+
+\newcommand{\Com}{\mathbb{C}}
 \newcommand{\Z}{\mathbb{Z}}
@@
 
-\newcommand{\hilb}{(\H, \langle \cdot, \cdot, \rangle )}
+\newcommand{\hilb}{(\Hil, \langle \cdot, \cdot, \rangle )}
 \renewcommand{\phi}{\varphi}
@@
 \renewcommand{\ker}{\textsc{Ker\ }}
+
+\usepackage{hyperref}
 
@@
 
-\begin{thm}[Hahn-Banach over $\C$]
+\begin{thm}[Hahn-Banach over $\Com$]
 	Suppose that $c$ is a seminorm on a complex vector space $X$ and let $M$ sub a subspace of $X$.
-If $f_0 \in \text{Hom}(M, \C)$ is such that $|f_0(x) | \leq p(x)$ for all $x \in M$, then there exists an extension $f \in \text{Hom}(X, \C)$ such that $f|_M = f_0$ and $|f(x)| \leq p(x)$ for all $x \in X$.
+If $f_0 \in \text{Hom}(M, \Com)$ is such that $|f_0(x) | \leq p(x)$ for all $x \in M$, then there exists an extension $f \in \text{Hom}(X, \Com)$ such that $f|_M = f_0$ and $|f(x)| \leq p(x)$ for all $x \in X$.
 \end{thm}
@@
 
-	We now show $|f(x)| \leq p(x)$.  For a fixed $x \in X$ choose $\lambda \in \C$ such that $\lambda f(x) = |f(x)|$.  Then since $|f(x)| \in \R$ and by definition of $f$, we have \[
+	We now show $|f(x)| \leq p(x)$.  For a fixed $x \in X$ choose $\lambda \in \Com$ such that $\lambda f(x) = |f(x)|$.  Then since $|f(x)| \in \R$ and by definition of $f$, we have \[
 		|f(x)| = \lambda f(x) | = f(\lambda x) = g(\lambda x) \leq p(\lambda x) = |\lambda p(x) = p(x)
@@
 			\frac{1}{4} \left( \| x + y \|^2 - \| x - y \|^2 \right) 	& \K = \R \\
-			\frac{1}{4}\left( \| x + y\|^2 - \| x - y \|^2 + i\| x + iy \|^2 - i\|x - iy\|^2 \right)										 		& \K = \C
+			\frac{1}{4}\left( \| x + y\|^2 - \| x - y \|^2 + i\| x + iy \|^2 - i\|x - iy\|^2 \right)										 		& \K = \Com
 		\end{cases}
@@
 \begin{thm}[Projection]
-	Let $(\H, \langle \cdot, \cdot, \rangle )$ be a Hilbert space.
-Let $M \subseteq \H$ be closed and convex.
-Let $x \in \H$.
+	Let $(\Hil, \langle \cdot, \cdot, \rangle )$ be a Hilbert space.
+Let $M \subseteq \Hil$ be closed and convex.
+Let $x \in \Hil$.
 THen there exists a unique point $m_x \in M$ which is closest to $x$, i.e.
@@
 			\| m_k - m_l \|^2 \leq 2 ( \frac{1}{k} + \frac{1}{l})
-	\end{align*} Thus $(m_k)$ is Cauchy.  So $m_k \rightarrow m_x \in M$ as $\H$ is complete and $M$ is closed.  We then have \[
+	\end{align*} Thus $(m_k)$ is Cauchy.  So $m_k \rightarrow m_x \in M$ as $\Hil$ is complete and $M$ is closed.  We then have \[
 		\| x - m_x \| = d 
@@
 \begin{defn}[Projection operator]
-Let $\hilb$ be a Hilbert space.  Let $M \subseteq \H$ be closed and convex.  Define \[
-	P_M : \H \rightarrow \H
-\] by $P_M(x) = m_x$ from above. This is the projection of $\H$ onto $M$.
+Let $\hilb$ be a Hilbert space.  Let $M \subseteq \Hil$ be closed and convex.  Define \[
+	P_M : \Hil \rightarrow \Hil
+\] by $P_M(x) = m_x$ from above. This is the projection of $\Hil$ onto $M$.
 \end{defn}
@@
 \begin{defn}[Orthogonal decomposition]	
-	If $S \subseteq \H$, let \[
-		S^\perp = \{ x \in \H
+	If $S \subseteq \Hil$, let \[
+		S^\perp = \{ x \in \Hil
  \, | \langle x, y \rangle = 0 \quad \forall y \in S. 
@@
 \begin{thm}[From previous lecture]
-	If $M \subseteq \H$, then the projection of $\H$ onto $M$ is \mapping{P_m}{\H}{\H}{x}{m_x} where $m_x \in M$ is the unique element with $\| x - m_x \| = \inf_{m \in M} \| x - m \|$.  
+	If $M \subseteq \Hil$, then the projection of $\Hil$ onto $M$ is \mapping{P_m}{\Hil}{\Hil}{x}{m_x} where $m_x \in M$ is the unique element with $\| x - m_x \| = \inf_{m \in M} \| x - m \|$.  
 \end{thm}
@@
 \begin{lem}
-	Let $M \subseteq \H$ be closed subspace.  Then $x - P_M x \in M^{\perp}$ for all $x \in \H$.
+	Let $M \subseteq \Hil$ be closed subspace.  Then $x - P_M x \in M^{\perp}$ for all $x \in \Hil$.
 \end{lem}
@@
 	The following theorem is the key fundamental result.  
-	Let $\hilb$ be a Hilbert space.  Let $M$ be a closed subspace of $\H$.  Then \[
-		 \H = M \oplus M^\perp.
-	\]  That is, each $x \in \H$ can be written in exactly one way as $x = m + m^\perp$ with $m \in M$, $m^\perp \in M^\perp$. 
+	Let $\hilb$ be a Hilbert space.  Let $M$ be a closed subspace of $\Hil$.  Then \[
+		 \Hil = M \oplus M^\perp.
+	\]  That is, each $x \in \Hil$ can be written in exactly one way as $x = m + m^\perp$ with $m \in M$, $m^\perp \in M^\perp$. 
 \end{thm}
@@
 \begin{cor}
-	Let $M \subseteq \H$ be a closed subspace.  Then we have
+	Let $M \subseteq \Hil$ be a closed subspace.  Then we have
 	\begin{enumerate}[(a)]
-		\item $P_M \in \mathcal{L}(\H, \H)$.
+		\item $P_M \in \mathcal{L}(\Hil, \Hil)$.
 		\item $\| P_M \| \leq 1$.
@@
 
-If $y \in \H$ is fixed, then the map \mapping{\phi_y}{\H}{\K}{x}{\langle x, y \rangle} is in $\H'$.  Linearity is clear, and continuity is proven by Cauchy-Swartz, \[
+If $y \in \Hil$ is fixed, then the map \mapping{\phi_y}{\Hil}{\K}{x}{\langle x, y \rangle} is in $\Hil'$.  Linearity is clear, and continuity is proven by Cauchy-Swartz, \[
 	| \phi_y(x) | = | \langle x, y \rangle | \leq \| y \| \| x \|.
@@
 \begin{thm}[Riesz Representation Theorem]
-	Let $\H$ be a Hilbert space. The map \mapping{\theta}{\H}{\H'}{y}{\phi_y} is a conjugate linear bijection, and $\| \phi_y \| = \| y \|$.  
+	Let $\Hil$ be a Hilbert space. The map \mapping{\theta}{\Hil}{\Hil'}{y}{\phi_y} is a conjugate linear bijection, and $\| \phi_y \| = \| y \|$.  
 \end{thm}
@@
 	\textbf{Surjectivity}
-	Let $\phi \in H'$.  We now find $y \in \H$ with $\phi = \phi_y$.  If $\phi = 0$, take $y = 0$.  Suppose $\phi \neq 0$.  Then $\ker \phi \neq \H$.  But $\ker \phi$ is a closed subspace of $\H$.  So \[
+	Let $\phi \in H'$.
+We now find $y \in \Hil$ with $\phi = \phi_y$.
+If $\phi = 0$, take $y = 0$.
+Suppose $\phi \neq 0$.
+Then $\ker \phi \neq \Hil$.
+But $\ker \phi$ is a closed subspace of $\Hil$.
+So \[
 		H = (\ker \phi) \oplus (\ker \phi)^\perp.
-	\] Hence $(\ker \phi)^\perp \neq \{ 0 \}$. Pick $z \in (\ker \phi)^\perp, z \neq 0$.  For each $x \in \H$, the element \[
+	\] Hence $(\ker \phi)^\perp \neq \{ 0 \}$. Pick $z \in (\ker \phi)^\perp, z \neq 0$.  For each $x \in \Hil$, the element \[
 		x - \frac{\phi(x)}{\phi(z)} z \in \ker \phi
@@
 	and so \[
-	\phi(x) = \langle x, \frac{\overline{\phi(z)}}{\| z \|^2} z \rangle \quad \forall x \in \H,\]
+	\phi(x) = \langle x, \frac{\overline{\phi(z)}}{\| z \|^2} z \rangle \quad \forall x \in \Hil,\]
 	and so letting $y = \frac{\overline{\phi(z)}}{\| z \|^2} z$, we have $\phi = \phi_y$.  
@@
 \begin{exmp}
-	From Hahn-Banach given $y \in \H$ there exists $\phi \in \H'$ such that \[
+	From Hahn-Banach given $y \in \Hil$ there exists $\phi \in \Hil'$ such that \[
 		\| \phi \| = 1
@@
 \begin{exmp}[Adjoint operators]
-	Let $\H_1, \H_2$ be Hilbert spaces.  Let $ T \in \mathcal{L}(\H_1, \H_2)$.  The \textbf{adjoint} of $T$ is $T^\star \in \mathcal{L}(\H_2, \H_1)$ given by \[
+	Let $\Hil_1, \Hil_2$ be Hilbert spaces.  Let $ T \in \mathcal{L}(\Hil_1, \Hil_2)$.  The \textbf{adjoint} of $T$ is $T^\star \in \mathcal{L}(\Hil_2, \Hil_1)$ given by \[
 		\langle T x, y \rangle_2 = \langle x, T^\star y \rangle_1
-	\] for all $x \in \H_1, y \in \H_2$
+	\] for all $x \in \Hil_1, y \in \Hil_2$
 \end{exmp}
@@
 \begin{defn}[Orthonormal system]
-	As subset $S \subseteq \H$ is an \textbf{orthonormal system} (orthonormal) if \[
+	As subset $S \subseteq \Hil$ is an \textbf{orthonormal system} (orthonormal) if \[
 		\langle e, e' \rangle = \delta_{e, e'} \quad \forall e, e' \in S
@@
 	An orthonormal system $S$ is \textbf{complete} or a \textbf{Hilbert basis} if \[
-		\overline{\spans S} = \H
+		\overline{\spans S} = \Hil
 	\]
@@
 		\] is orthonormal and is complete.  
-		\item $L^2_\C([0, 2 \pi])$.  Then \[
+		\item $L^2_\Com([0, 2 \pi])$.  Then \[
 			S = \{ \frac{1}{2\pi} e^{i n t} \, | \, n \in \Z \}
@@
 		\sum_{k=0}^\infty a_l e_k
-	\] converges in $\H$ if and only if \[
+	\] converges in $\Hil$ if and only if \[
 		\sum_{k=0}^\infty |a_k |^2
@@
 						&= \sum_{k=n+1}^m |a_k |^2 \rightarrow 0
-	\end{align*} and so $(x_n)$ is Cauchy, and hence converges by completeness of $\H$.  
+	\end{align*} and so $(x_n)$ is Cauchy, and hence converges by completeness of $\Hil$.  
 \end{proof}
@@
 		\sumkn | \iprod{x, e_k} |^2 \leq \|x\|^2 
-	\] for each $x \in \H$.  
+	\] for each $x \in \Hil$.  
 \end{lem}
@@
 \begin{cor}
-	Let $x \in \H$ and $S$ orthonormal.  Then \[
+	Let $x \in \Hil$ and $S$ orthonormal.  Then \[
 		\{ e \in S \, | \, \iprod{x, e} \neq 0 \}
@@
 		\sum_{e \in S} | \iprod{x,e}|^2 \leq \| x \|^2
-	\] for all $x \in \H$
+	\] for all $x \in \Hil$
 \end{cor}
@@
 		y = \sum_{ k = 0}^\infty \iprod{x,e_k} e_k \in M
-	\] converges in $\H$.
+	\] converges in $\Hil$.
 
@@
 \label{sec:lecture_11_monday_4_april}
-Recall that if $\{ x_1. \dots \}$ is a countable orthonormal system in a Hilbert space $\H$.  Then \[
+Recall that if $\{ x_1. \dots \}$ is a countable orthonormal system in a Hilbert space $\Hil$.  Then \[
 	\sum_{k=1}^\infty a_k e_k < \infty \iff \sum_{k=1}^\infty |a_k|^2 < \infty 
@@
 \begin{thm}\label{thm:fourier}
-	Let $S$ be orthonormal in $\H$.  Let $M = \overline{\spans S}$.  Then \[
-		P_Mx = \sum_{e \in S} \iprod{x, e} e \quad \forall x \in \H
+	Let $S$ be orthonormal in $\Hil$.  Let $M = \overline{\spans S}$.  Then \[
+		P_Mx = \sum_{e \in S} \iprod{x, e} e \quad \forall x \in \Hil
 	\] where the sum has only countable many terms and convergence is unconditional.
@@
 \begin{thm}
-	Let $S$ be orthonormal in $\H$.  Then following are equivalent.
+	Let $S$ be orthonormal in $\Hil$.  Then following are equivalent.
 	\begin{enumerate}[(a)]
-		\item $S$ is a complete orthonormal system ($\overline{\spans S} = \H$).
+		\item $S$ is a complete orthonormal system ($\overline{\spans S} = \Hil$).
 		\item $x = \sum_{e \in S} \iprod{x,e} e$ for all $x$ (Fourier series).
@@
 \begin{proof}
-	(a) $\Rightarrow$ (b).  If $M = \overline{\spans S} = \H$, then \[
+	(a) $\Rightarrow$ (b).  If $M = \overline{\spans S} = \Hil$, then \[
 		P_M x = x = \sum_{e \in S} \iprod{x,e} e
@@
 		0 = \| P_M z \|^2 = \|\sum_{e \in S} \iprod{z,e} e \|^2 = \sum_{e \in S} |\iprod{z,e} |^2 = \| z \|^2 
-	\] which implies $z = 0$, so $M = \H$, and so $S$ is complete.
+	\] which implies $z = 0$, so $M = \Hil$, and so $S$ is complete.
 \end{proof}
@@
 
-	We now prove for $\mathcal{C}_\C(X)$.
+	We now prove for $\mathcal{C}_\Com(X)$.
 
@@
 
-	Now, let $f \in \mathcal C_\C(X)$.  Then write $f = u + iv$.  Then $u,v \in \mathcal C_\R(X)$.  Then given $\epsilon > 0$, there exists $u_1,v_1 \in \mathcal A_\R$ such that \[
+	Now, let $f \in \mathcal C_\Com(X)$.  Then write $f = u + iv$.  Then $u,v \in \mathcal C_\R(X)$.  Then given $\epsilon > 0$, there exists $u_1,v_1 \in \mathcal A_\R$ such that \[
 		\| u - u_1 \|_\infty \leq \frac{\epsilon}{2}, \quad \| v - v_1 \|_\infty \leq \frac{\epsilon}{2} 
@@
 		\| f - f_1 \|_\infty \leq \| (u - u_1) + i(v-v_1) \|_\infty \leq \| u - u_1 \|_\infty + \| v - v_1 \|_\infty < \epsilon
- 	\] and thus $\mathcal{A}$ is dense in $\mathcal C_\C(X)$.
+ 	\] and thus $\mathcal{A}$ is dense in $\mathcal C_\Com(X)$.
 \end{proof}
@@
 \begin{cor}
-	The space $\mathcal A$ of all trigonometric polynomials is dense in $\mathcal C(\Pi)$, where $\Pi = \{ z \in \C \, | \, |z | = 1 \}$
+	The space $\mathcal A$ of all trigonometric polynomials is dense in $\mathcal C(\Pi)$, where $\Pi = \{ z \in \Com \, | \, |z | = 1 \}$
 \end{cor}
@@
 \label{sub:spectral_theory}
-The eigenvalues of an $n \times n$ matrix $T$ over $\C$ are the $\lambda \in \C$ with \[
+The eigenvalues of an $n \times n$ matrix $T$ over $\Com$ are the $\lambda \in \Com$ with \[
 	\det(\lambda I - T) = 0
@@
 \begin{rem}
-	We need our base field to be $\C$ to get reasonable spectral theory.
+	We need our base field to be $\Com$ to get reasonable spectral theory.
 \end{rem}
@@
 \begin{thm}
-	Let $X \neq \{ 0 \}$ be a Banach space over $\C$, and let $T \in \mathcal L(X)$.  Then $\sigma(T)$ is a non-empty, compact (closed and bounded) subset of \[
-		\{ \lambda \in \C \given | \lambda | \leq \| T \| \}
+	Let $X \neq \{ 0 \}$ be a Banach space over $\Com$, and let $T \in \mathcal L(X)$.  Then $\sigma(T)$ is a non-empty, compact (closed and bounded) subset of \[
+		\{ \lambda \in \Com \given | \lambda | \leq \| T \| \}
 	\]
@@
 \begin{thm}
-	Let $X \neq \{ 0 \}$ be a Banach space over $\C$.  Let $T \in \mathcal L(X)$.  Then $\sigma(T)$ is a nonempty, compact subset of \[
-		\{ \lambda \in \C \given | \lambda | \leq \| T \| \}.
+	Let $X \neq \{ 0 \}$ be a Banach space over $\Com$.  Let $T \in \mathcal L(X)$.  Then $\sigma(T)$ is a nonempty, compact subset of \[
+		\{ \lambda \in \Com \given | \lambda | \leq \| T \| \}.
 	\]
@@
 \begin{lem}
-	With above assumptions $\sigma(T) \subseteq \{ \lambda \in \C \given | \lambda | \leq \| T \| \}$.  
+	With above assumptions $\sigma(T) \subseteq \{ \lambda \in \Com \given | \lambda | \leq \| T \| \}$.  
 \end{lem}
@@
 \begin{proof}
-	Let \mapping{f}{\C}{\mathcal L(X)}{\lambda}{\lambda I - T}  This is continuous, as \begin{align*}
+	Let \mapping{f}{\Com}{\mathcal L(X)}{\lambda}{\lambda I - T}  This is continuous, as \begin{align*}
 		\| f(\lambda) - f(\lambda_0) \| &= \| (\lambda - \lambda_0) I \| \\
@@
 
-So $\sigma(T)$ is a compact subset of $\{ \lambda \in \C \given | \lambda | \leq \| T \| \}$.
-Write $\rho(T) = \C \backslash \sigma(T)$ (the \textbf{resolvent set}), and let $R_T = R : \rho(T) \rightarrow \mathcal L (X)$ with $R_T(\lambda) = (\lambda I - T)^{-1}$.
+So $\sigma(T)$ is a compact subset of $\{ \lambda \in \Com \given | \lambda | \leq \| T \| \}$.
+Write $\rho(T) = \Com \backslash \sigma(T)$ (the \textbf{resolvent set}), and let $R_T = R : \rho(T) \rightarrow \mathcal L (X)$ with $R_T(\lambda) = (\lambda I - T)^{-1}$.
 
 \begin{thm}
-	Let $\K = \C$ and $X \neq \{ 0 \}$ and $T \in \mathcal L(X)$.  Then $\sigma(T) \neq \emptyset$.
+	Let $\K = \Com$ and $X \neq \{ 0 \}$ and $T \in \mathcal L(X)$.  Then $\sigma(T) \neq \emptyset$.
 \end{thm}
@@
 
-	Let $\phi = \mathcal L(X)'$ (hence $\phi : \mathcal L(X) \rightarrow C$.)  Let \mapping{f_\phi}{\rho(T)}{\C}{\lambda}{\phi(R(\lambda))} 
+	Let $\phi = \mathcal L(X)'$ (hence $\phi : \mathcal L(X) \rightarrow C$.)  Let \mapping{f_\phi}{\rho(T)}{\Com}{\lambda}{\phi(R(\lambda))} 
 	\begin{lem}
@@
 
-	Now suppose that $\sigma(T) = \emptyset$.  Then $f_\phi : \C \rightarrow \C$ is analytic.
+	Now suppose that $\sigma(T) = \emptyset$.  Then $f_\phi : \Com \rightarrow \Com$ is analytic.
 
@@
 \begin{thm}[Spectral mapping theorem (polynomials)]
-	Let $T$ be an $n \times n$ matrix over $\C$.  IF we know all the eigenvalues of $T$, then we know the eigenvalues of every polynomial $p(T) = a_0 + a_1 T + \dots + a_n T^n$.  Specifically, \[
+	Let $T$ be an $n \times n$ matrix over $\Com$.  IF we know all the eigenvalues of $T$, then we know the eigenvalues of every polynomial $p(T) = a_0 + a_1 T + \dots + a_n T^n$.  Specifically, \[
 		\{ \text{eigenvalues of $p(T)$} \} = \{ p(\lambda) \given \text{$\lambda$ is an eigenvalue of $T$} \}
@@
 
-	This also holds for $X$ Banach over $\C$, and $T \in \mathcal L(X)$.
+	This also holds for $X$ Banach over $\Com$, and $T \in \mathcal L(X)$.
 \end{thm}
@@
 \begin{lem}
-	Let $\C[t]$ be the algebra of polynomials in $t$ with complex coefficients.  Multiplication is defined as usual.
+	Let $\Com[t]$ be the algebra of polynomials in $t$ with complex coefficients.  Multiplication is defined as usual.
 \end{lem}
@@
 \begin{lem}
-	Let $X$ be Banach over $\C$.  Let $T \in \mathcal L(X)$.  Then \mapping{\phi}{\C[t]}{\mathcal L(X)}{p}{p(T)} is an algebra homomorphism (multiplication corresponds to composition in $\mathcal L(X)$.)
+	Let $X$ be Banach over $\Com$.
+Let $T \in \mathcal L(X)$.
+Then \mapping{\phi}{\Com[t]}{\mathcal L(X)}{p}{p(T)} is an algebra homomorphism (multiplication corresponds to composition in $\mathcal L(X)$.)
 \end{lem} 
@@
 		\phi(\alpha p) = \alpha \phi(p)
-	\end{align*} for all $p_1, p_2, p \in \C[t], \alpha \in \C$.  
+	\end{align*} for all $p_1, p_2, p \in \Com[t], \alpha \in \Com$.  
 \end{proof}
 \begin{thm}
-	Let $X$ be Banach over $\C$, and let $T \in \mathcal L(X)$.  Then \[
+	Let $X$ be Banach over $\Com$, and let $T \in \mathcal L(X)$.  Then \[
 		\sigma(p(T)) = p(\sigma(T)).
@@
 	\]  
-	Now, suppose that $p$ is non constant.  Let $\mu \in \C$ fixed.  By the fundamental theorem of algebra, we can factorise $\mu - p(t)$ as \[
+	Now, suppose that $p$ is non constant.  Let $\mu \in \Com$ fixed.  By the fundamental theorem of algebra, we can factorise $\mu - p(t)$ as \[
 		\alpha( t - \lambda_1)^{m_1} \dots (t - \lambda_n)^{m_n}
@@
 Note that $\mu = p(\lambda_i)$ for each $i$.
-Applying $\psi: \C[t] \rightarrow \mathcal L(X)$ from above, we have \begin{align*}
+Applying $\psi: \Com[t] \rightarrow \mathcal L(X)$ from above, we have \begin{align*}
 		\mu I - p(T) = \alpha(T- \lambda_1 I)^{m_1} \dots (T - \lambda_n I)^{m_n}
@@
 \begin{defn}[Spectral radius]
-Let $X \neq \{ 0 \}$ be a Banach space over $\C$.  The \textbf{spectral radius} of $T \in \mathcal L(X)$ is
+Let $X \neq \{ 0 \}$ be a Banach space over $\Com$.  The \textbf{spectral radius} of $T \in \mathcal L(X)$ is
 \begin{align*}
@@
 		r(T) \leq \| T \|
-	\] since $\sigma(T) \subseteq \{ \lambda \in \C \given | \lambda | \leq \| T \| \}$.  Strict inequality can (and often does) occur.
+	\] since $\sigma(T) \subseteq \{ \lambda \in \Com \given | \lambda | \leq \| T \| \}$.  Strict inequality can (and often does) occur.
 \end{note}
@@
 		\end{pmatrix}.
-	\]  Then consider $T : \C^2 \rightarrow \C^2$ where $\| (x, y) \|_2 = \sqrt{|x|^2 + |y|^2}$.  Then \begin{align*}
-		\| T \| &= \sup \{ \| Tx \|_2 \given x \in \C^2 \} \\
+	\]  Then consider $T : \Com^2 \rightarrow \Com^2$ where $\| (x, y) \|_2 = \sqrt{|x|^2 + |y|^2}$.  Then \begin{align*}
+		\| T \| &= \sup \{ \| Tx \|_2 \given x \in \Com^2 \} \\
 				&= \sqrt{\lambda_{max} (T^* T)} 
@@
 \begin{thm}[Gelfand, 1941]
-	Let $X \neq \{ 0 \}$ be Banach over $\C$, and let $T \in \mathcal L(X)$.  Then \begin{align*}
+	Let $X \neq \{ 0 \}$ be Banach over $\Com$, and let $T \in \mathcal L(X)$.  Then \begin{align*}
 		r(T) = \lim_{n \rightarrow \infty} \| T^n \|^{1/n}.
@@
 	\]
-	Let $\phi \in \mathcal L(X)$ and let \mapping{f_\phi}{\rho(T)}{\C}{\lambda}{\phi((\lambda I - T)^{-1})}
+	Let $\phi \in \mathcal L(X)$ and let \mapping{f_\phi}{\rho(T)}{\Com}{\lambda}{\phi((\lambda I - T)^{-1})}
 
@@
 \label{sec:lecture_24_wednesday_25_may}
-Let $T: \C^n \rightarrow \C^n$ be a linear operator.
-Then in the simplest case, $T$ has $n$ distinct eigenvalues, and the corresponding eigenvectors are linearly independent, forming a basis for $\C^n$.
-
-Hence, $\C^n = \C x_1 \oplus \dots \oplus \C x_n$ and the matrix of $T$ relative to this basis is simply diagonal with $\lambda_1, \dots, \lambda_n$.
+Let $T: \Com^n \rightarrow \Com^n$ be a linear operator.
+Then in the simplest case, $T$ has $n$ distinct eigenvalues, and the corresponding eigenvectors are linearly independent, forming a basis for $\Com^n$.
+
+Hence, $\Com^n = \Com x_1 \oplus \dots \oplus \Com x_n$ and the matrix of $T$ relative to this basis is simply diagonal with $\lambda_1, \dots, \lambda_n$.
 
 This is not always possible, because there is not always a basis of eigenvectors.  Instead look at the generalised eigenspace, \[
-	\{ x \in \C^n \given \text{$(\lambda I - T)^k x = 0$ for some $k \geq 1$}.  
+	\{ x \in \Com^n \given \text{$(\lambda I - T)^k x = 0$ for some $k \geq 1$}.  
 \] 
 But $\{ 0 \} \subseteq \ker(\lambda I - T)^1 \subseteq \ker(\lambda I - T)^2 \subseteq \dots$
- and since $\dim(\C^n) < \infty$ this must stabilise.
+ and since $\dim(\Com^n) < \infty$ this must stabilise.
 Let $r \geq 1$ be the fist time that $\ker (\lambda I - T)^r = \ker( \lambda I - T)^{r+1}$.
 Then the generalised $\lambda$-eigenspace is just $\ker(\lambda I - T)^r$.
-There \textbf{is} a basis of $\C^n$ consisting of generalised eigenvectors, and the matrix of $T$ relative to this basis is in block form.
+There \textbf{is} a basis of $\Com^n$ consisting of generalised eigenvectors, and the matrix of $T$ relative to this basis is in block form.
 
@@
 \begin{prop}
-	Let $\mathcal H$ be Hilbert over $\C$.  IF $T \in \mathcal L(\mathcal H)$ is normal, then $r(T) = \| T \|$.  
+	Let $\mathcal H$ be Hilbert over $\Com$.  IF $T \in \mathcal L(\mathcal H)$ is normal, then $r(T) = \| T \|$.  
 \end{prop}
@@
 \begin{cor}
-	Let $\mathcal H$ be a Hilbert space over $\C$.  \begin{enumerate}[(a)]
+	Let $\mathcal H$ be a Hilbert space over $\Com$.  \begin{enumerate}[(a)]
 		\item If $T \in \mathcal L(\mathcal H)$ is unitary, then \[
-			\sigma(T) \subseteq \mathbb{T} = \{ \lambda \in \C \given | \lambda | = 1 \}\]
+			\sigma(T) \subseteq \mathbb{T} = \{ \lambda \in \Com \given | \lambda | = 1 \}\]
 		\item If $T \in \mathcal L(\mathcal H)$ is Hermitian, then \[
@@
 \begin{lem}
-	Let $\mathcal H$ be Hilbert over $\C$.  Let $T \in \mathcal L(\mathcal H)$, and let \[
+	Let $\mathcal H$ be Hilbert over $\Com$.  Let $T \in \mathcal L(\mathcal H)$, and let \[
 		M_\lambda = \{ x \in \mathcal H \given Tx = \lambda x \} = \ker \lambda I - T
@@
 \begin{thm}[Fredhold alternative (Finite dimensional spaces)]
-	Let $A: \C^n \rightarrow \C^n$ be linear.  Then exactly one of the following two things occur:
+	Let $A: \Com^n \rightarrow \Com^n$ be linear.  Then exactly one of the following two things occur:
 	\begin{enumerate}[(1)]
-		\item $Ax = 0$ has only the trivial solution $x = 0$, in which case $Ax = b$ has a unique solution for each $b \in \C^n$.  
+		\item $Ax = 0$ has only the trivial solution $x = 0$, in which case $Ax = b$ has a unique solution for each $b \in \Com^n$.  
 		\item $Ax = 0$ has a non-trivial solution, in which case $Ax = b$ has either no solutions, or infinitely many solutions.
@@
 \begin{thm}[Fredholm alternative (Hilbert spaces)]
-	Let $\mathcal H$ be Hilbert over $\C$, and let $T \in \mathcal K(\mathcal H)$.  Then exactly one of the following occurs.
+	Let $\mathcal H$ be Hilbert over $\Com$, and let $T \in \mathcal K(\mathcal H)$.  Then exactly one of the following occurs.
 	\begin{enumerate}[(a)]

### 2013-01-06 PMH3 - Functional Analysis/PMH3 Lecture Notes.tex
Unified title naming
Changed author on PMH3
Added ToC to all notes
Changed tabs to spaces

@@
 
-\documentclass[10pt, reqno, oneside]{amsart}
-\usepackage{amsthm, amsmath, amssymb}
-\usepackage{setspace, graphicx, enumerate}
+\documentclass[10pt, oneside, reqno]{amsart}
+\usepackage{geometry, setspace, graphicx, enumerate, amssymb}
 \onehalfspacing                 
-% \usepackage{mathspec}
-% \setallmainfonts{Minion Pro}
-
 \usepackage{fontspec,xltxtra,xunicode}
 \defaultfontfeatures{Mapping=tex-text}
-
-% \setlength{\columnsep}{50pt}
-% \setlength{\columnseprule}{2pt}
-% \setromanfont[Mapping=tex-text,Contextuals=w {NoWordInitial,NoWordFinal,NoLineInitial,NoLineFinal}]{Adobe Caslon Pro}
-%\setsansfont[Scale=MatchLowercase,Mapping=tex-text]{Hoefler Text}
-%\setmonofont[Scale=MatchLowercase]{Andale Mono}
 
@@
 
-\title{PMH3 - Functional Analysis \\ Lecture Notes}								% Document Title
-\author{James Parkinson}
+\title{PMH3 - Functional Analysis}                              % Document Title
+\author{Andrew Tulloch}
 %\date{}                                           % Activate to display a given date or no date
@@
 \begin{document}
-\maketitle
+\maketitle \tableofcontents \clearpage
 
@@
 \begin{enumerate}[(a)]
-		\item Linearity is clear.  The $p_n(t) = t^n$ for all $t \in [0,1]$.  Then $|D(p_n)| =   n \| p_n \|_\infty$.  So $D$ is not continuous, as continuity implies that there exists $c$ such that \[
+        \item Linearity is clear.
+The $p_n(t) = t^n$ for all $t \in [0,1]$.
+Then $|D(p_n)| =   n \| p_n \|_\infty$.
+So $D$ is not continuous, as continuity implies that there exists $c$ such that \[
             \| Tx \| \leq c \| x \|.
@@
 Let $x \in \Hil$.
-THen there exists a unique point $m_x \in M$ which is closest to $x$, i.e.
+Then there exists a unique point $m_x \in M$ which is closest to $x$, i.e.
 \[
@@
     \mathcal A_\R = \{ f \in \mathcal A \given \text{$f$ is real valued} \}. 
-	\]  THen $\mathcal A_\R$ is an $\R$-subalgebra of $\mathcal C_\R(X)$.  It is unital, as $1 \in \mathcal A$ and it is real valued.
+    \]  Then $\mathcal A_\R$ is an $\R$-subalgebra of $\mathcal C_\R(X)$.  It is unital, as $1 \in \mathcal A$ and it is real valued.
 
@@
 
-	$(b) \Rightarrow (a)$.  Suppose $(b)$ holds.  Let $B_1 \subseteq X$ be bounded.  THen $B_1 \subseteq \alpha B$ for some $\alpha > 0$.  So \[
+    $(b) \Rightarrow (a)$.  Suppose $(b)$ holds.  Let $B_1 \subseteq X$ be bounded.  Then $B_1 \subseteq \alpha B$ for some $\alpha > 0$.  So \[
         \overline{T(B)} \subseteq \overline{T(\alpha B)} = \alpha \overline{T(B)}
@@
 Let $\theta \in (0, 1)$ be given.
-THen there exists $x$ with $\| x \| = 1$ such that $\| x - y \| \geq \theta$ for all $y \in Y$.
+Then there exists $x$ with $\| x \| = 1$ such that $\| x - y \| \geq \theta$ for all $y \in Y$.
 \end{lem}
@@
 Then by the definition of the infimum, there is $y_0 \in Y$ with $\alpha \leq \| z - y_0 \| \leq \frac{\alpha}{\theta}$.
-Now let $x = \frac{x - y_0}{\| z - y_0 \|}$.
+Now let $x = \frac{z - y_0}{\| z - y_0 \|}$.
 Then $\| x \| = 1$.
@@
         \| x - y \| &= \left \| \frac{z - y_0}{\| z - y_0 \|} - y \right \| \\
-					&= \frac{1}{\| z - y_0 \|} \left \| z - y_- + \| z - y_0 \| y \right \| \\
+                    &= \frac{1}{\| z - y_0 \|} \left \| z - y_0 + \| z - y_0 \| y \right \| \\
                     &\geq \frac{\theta}{\alpha} \alpha = \theta
@@
     \| Tx'_m - T x'_n \| \leq \| T x'_m - T_k x'_m \| + \| T_K x'_m - T_k x'_n \| + T_k x'_n - T x'_n \|
-\end{align*} where $k$ is to be chosen.  Suppose $\| x_n \| \leq M$ for all $n \geq 1$.  THen \begin{align*}
+\end{align*} where $k$ is to be chosen.  Suppose $\| x_n \| \leq M$ for all $n \geq 1$.  Then \begin{align*}
     \| Tx'_m - T x'_n \| \leq 2 M \| T - T_k \| + \| T_k x'+m - T_k x'_n \|
@@
             x_n = S x_n + T x_n &\rightarrow y + z.
-			\end{align*}  So $x_n$ converges to $x \in N$.  So $S x_n \rightarrow Sx \in S(N)$ by continuity, but we assume that $S x_n \rightarrow y \in X \backslash S(N)$, which achieves our contradiction.
+            \end{align*}  So $x_n$ converges to $x \in N$.
+So $S x_n \rightarrow Sx \in S(N)$ by continuity, but we assume that $S x_n \rightarrow y \in X \backslash S(N)$, which achieves our contradiction.
         \end{case}
